### 2023-06-07 frontend/merci.tex
First set-up (biblio, symbols)

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

Thank you, thank you, far too kind

### 2023-12-21 chap3_myopic/chap3_myopic.tex
First set-up

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode
In this chapter, I highlight the advantages of the myopic pathway, the methodological adaptations needed to have a myopic optimisation from EnergyScope Pathway perfect foresight and detail the difference with the perfect foresight for the case study.

\section{Why and how myopic}
\label{sec:chap3_why_how}
Besides the fact that it is a necessary framework for the application of RL,  explain here why it is interesting, as is, to consider myopic optimisation (time saving, more appropriate to mimic policymakers' shortsightedness, etc.)

Detail the little tweeks to go from a perfect foresight to a myopic optimisation, in terms of implementation.

\section{Case study and results: Belgian energy system under \ce{CO2} trajectory}
\label{sec:chap3_case_study_results}

\subsection{Case study}
{\color{red}Question to HJ and FC : J'hésite sur le cas de référence pour la comparaison en termes de trajectoire \ce{CO2}. Voici ce à quoi je pense:
\begin{enumerate}
\item Comme ce qu'on a présenté dans le papier Pathway, une décroissance linéaire entre les émissions de 2020 et la neutralité carbone en 2050
\item Sur base du budget \ce{CO2} identique à celui de la décroissance linéaire (~1.9Gt\ce{CO2}), imposer au myopique la trajectoire \ce{CO2} issue de l'optimisation perfect foresight. Dans cette option, il y a deux sous-choix:
\begin{itemize}
\item Imposer la neutralité carbone en 2050, comme dans le cas 1.
\item Ne pas imposer la neutralité carbone en 2050
\end{itemize}
\item Sur base du budget \ce{CO2} que je donne implicitement comme objectif à mon agent (~1.2Gt\ce{CO2}), imposer au myopique la trajectoire \ce{CO2} issue de l'optimisation perfect foresight. Dans cette option, il y a deux sous-choix:
\begin{itemize}
\item Imposer la neutralité carbone en 2050, comme dans le cas 1.
\item Ne pas imposer la neutralité carbone en 2050
\end{itemize}
\end{enumerate}
Perso, je choisirais de me rapprocher le plus possible de ce que je fais ensuite avec l'agent. Du coup, ce serait le cas 3  sans imposer la neutralité carbone 2050. Dans ce cas, pour l'avoir observé, le modèle n'atteint pas la neutralité carbone.

Quel est votre avis sur la question?
}

Define the case study and the \ce{CO2} trajectory.

\subsection{Results and comparison with Perfect foresight}

\subsection{Discussion and perspective with the literature}

### 2023-12-21 chap4_RL/chap4_RL.tex
First set-up

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

\section{Reinforcement learning fundamentals}
\label{sec:RL_fundamentals}
As RL is framework that is not really applied to the optimisation of the whole-energy system transition, I find it necessary to introduce the general concepts of reinforcement learning and say that, even if it is usually used in other fields, there is a point to use it here.

\section{Definition of the actions, states and rewards}
\label{sec:act_states_rew}
After pointing out that the environment is basically the EnergyScope myopic model at the different steps of the transition, define the actions, the states as well as the shape of the reward.

\subsection{Actions}
\subsection{States}
\subsection{Reward}

\section{Uncertainties in the learning}
Remind here that uncertainties considered in the learning of the agent are those presented in Chapter \ref{chap:chap1_ES_PCE} and already applied and screened in the work/results presented in Chapter \ref{chap:chap2_electro_uq}. Present as well how we affect the value of the parameters based on the value of the sample.

\section{Results}

\subsection{Training}
Here present results RL-oriented (actions, states, reward)

Then, present results energy system-oriented (installed capacities, costs, generation)

\subsection{Testing}
See how the different policies saved successively during the training behave when facing to new samples. This way, we can pick an optimal policy (the one giving the maximum average (or another metric) reward).

We then compare the results of RL with the perfect foresight-TD deterministic and the perfect foresight-monthly with uncertainties (by setting the actions of the agent as variables in these two optimisations). The comparison would go over different aspects:
\begin{itemize}
\item Over-cost
\item Over-change, a bit like what Paolo defined as design error/change (in terms in consumed resources, installed capacities,...)
\end{itemize}

### 2023-12-21 chap_intro_ccl/conclusion.tex
First set-up

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode
I took here the same sections as in Gauthier's thesis
\section*{Thesis contribution}
\addcontentsline{toc}{section}{Thesis contribution}
Insist here on the methodological added value of the thesis

\section*{Application outcomes}
\addcontentsline{toc}{section}{Application outcomes}
What this new methodology has brought over when applied to the case of Belgium

\section*{Recommendations and guidance}
\addcontentsline{toc}{section}{Recommendations and guidance}
What to do then for policymakers, how to use the tool

\section*{Perspectives}
\addcontentsline{toc}{section}{Perspectives}
List the future works to build upon the thesis

### 2023-12-21 chap_intro_ccl/intro.tex
First set-up

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

Our society In their 2030 Agenda, United Nations have worked on identifying 17 \gls{SDGs} as a plan of action for society (or people), environment (or planet) and economy (or prosperity) \cite{un_sdgs}. 
%
%As Fatih
%
%On top of the depletion of (economically and environmentally) easy to harvest conventional fossil resources, the most urgent motivation for energy transition is the growing threat of climate change.  This change jeopardizes most of the \gls{SDGs} listed by the United Nations . Besides their environmental pillars (\ie 13-climate action, 14-life below water and 15-life on land)\cite{vinuesa2020role}, succeeding such a transition would contribute in reaching others of these goals like affordable

General introduction to the need of energy transition\\

How electrofuels will be part of the solution. Here include part of the terminology paper \url{https://www.frontiersin.org/articles/10.3389/fenrg.2021.660073/full} to clearly set the context and make sure the readers understand what's behind electrofuels and emphasize at the end that in our case, we consider electrofuels as the ones from renewable sources (with a 0-gwp).\\

From this need, develop a thread like I did in my FRIA application, we need whole-energy system model to give more insights to policymakers\\

Listing some examples of situations where uncertainties have not been considered and ended up in over-cost/waste of time/waste of..., highlight the need as well to consider uncertainties when we want to advise policymakers.\\

As the question of policymakers is not only what to do but how to do it, we need to address the optimisation of policies. This would introduce the RL part

### 2023-12-21 options/symbols_and_cmd.tex
First set-up

%!TEX encoding = UTF-8 Unicode
%!TEX root = ../thesis_main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	MATH SYMBOLS							%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\glsxtrsetgrouptitle{Abreviation}{Abreviation}
\glsxtrsetgrouptitle{Notation}{Notation}

% \newcommand{\sdot}{{\bullet}}
\newcommand{\sdot}{(\cdot)}

% ---------------------------------------------------------------------------- %
% ACRONYMS
%% A

%% P
\newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}

%% S
\newabbreviation[group={Acronyms}]{SDGs}{SDGs}{Sustainable Development Goals}

%% U
\newabbreviation[group={Acronyms}]{UQ}{UQ}{Uncertainty Quantification}

% ---------------------------------------------------------------------------- %
% ROMAN
%% A
%\glsxtrnewsymbol[group={rs},description={ith ambient particle}]{ai}{\ensuremath{A_i}}

% ---------------------------------------------------------------------------- %
% GREEK
%\glsxtrnewsymbol[group={gs},description={Yaw angle [$\deg$]}]{beta}{\ensuremath{\beta}}

% ---------------------------------------------------------------------------- %
% Sub- Superscripts
%% A
%\glsxtrnewsymbol[group={ss},description={Relative to the ambient}]{uA}{\ensuremath{\sdot_{A}}}

\renewcommand\glstreegroupheaderfmt[1]{\begingroup\textbf{#1}\vspace{-.2cm}\par\endgroup}
\glsfindwidesttoplevelname
\glsxtrsetgrouptitle{Acronyms}{Acronyms}
\glsxtrsetgrouptitle{rs}{Roman Symbols}
\glsxtrsetgrouptitle{gs}{Greek Symbols}
\glsxtrsetgrouptitle{ss}{Sub- Superscripts}
\glsxtrsetgrouptitle{o}{Operators}

%------------------------
%% A
%\newcommand{\ABL}[2]{$U_\mABL=\SIms{#1}$ - $TI_\mABL={#2}\%$}
%------------------------
%% B

%------------------------
%% C

%------------------------
%% D
\newcommand{\degree}{^\circ}
\newcommand{\dd}{\mathrm{d}}

%------------------------
%% E
\newcommand{\etal}{\textit{et~al.}}
\newcommand{\eg}{e.g.\ }

%------------------------
%% F

%------------------------
%% G

%------------------------
%% H

%------------------------
%% I
\newcommand{\ie}{i.e.\ }

%------------------------
%% K

%------------------------
%% L

%------------------------
%% M

%------------------------
%% N

%------------------------
%% O

%------------------------
%% P

%------------------------
%% Q

%------------------------
%% R

%------------------------
%% S

%------------------------
%% T

%------------------------
%% U
\newcommand{\uh}{\hat{u}}
\newcommand{\uv}{\mathbf{u}}

\newcommand{\ub}{\boldsymbol{u}}

\newcommand{\uconvw}{\tilde{U}_W}

%------------------------
%% V
\newcommand{\vv}{\mathbf{v}}
\newcommand{\vb}{\boldsymbol{v}}

%------------------------
%% W
\newcommand{\wv}{\mathbf{w}}
\newcommand{\wb}{\boldsymbol{w}}
\newcommand{\WT}{{\scriptscriptstyle \mathrm{WT}}}

%------------------------
%% X
\newcommand{\xv}{\mathbf{x}}
\newcommand{\xb}{\boldsymbol{x}}

%------------------------
%% Y
\newcommand{\yv}{\mathbf{y}}
\newcommand{\yb}{\boldsymbol{y}}

%------------------------
%% Z
\newcommand{\zv}{\mathbf{z}}
\newcommand{\zb}{\boldsymbol{z}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	OPERATORS 								%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\der}[2]{\frac{d #1}{d #2}}
\newcommand{\pder}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\pdern}[3]{\frac{\partial^#3 #1}{\partial #2^#3}}

\newcommand{\avgt}[1]{\left\langle {#1}\right\rangle}
\newcommand{\avgs}[1]{\overline{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	REFERENCES 							%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\eqqref}[1]{Eq.~\eqref{#1}}
\newcommand{\seqqref}[1]{Eqs.~\eqref{#1}}
\newcommand{\eqqrefs}[2]{Eqs.~\eqref{#1} and \eqref{#2}}
\newcommand{\eqqrefc}[2]{Eqs.~\eqref{#1}, \eqref{#2}}
\newcommand{\eq}[1]{Eq.~\eqref{#1}}
% \newcommand{\eqs}[1]{Eqs.~(\ref{#1})}

\newcommand{\fig}{Fig.~\ref}
\newcommand{\figs}{Figs.~\ref}
\newcommand{\tab}{Table~\ref}
\newcommand{\sect}{Section~\ref}
\newcommand{\chap}{Chapter~\ref}
\newcommand{\chaps}{Chapters~\ref}
\newcommand{\app}{Appendix~\ref}
\newcommand{\apps}{Appendices~\ref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	CAPTIONS							%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Color palette

\usepackage{xcolor}
\definecolor{myBlue}{rgb}{0.25, 0.25, 0.75}
\definecolor{myRed}{rgb}{0.8, 0.1, 0.3}
\definecolor{myGrey0}{rgb}{0.25, 0.25, 0.25}
\definecolor{myGrey1}{rgb}{0.50, 0.50, 0.50}
\definecolor{myGrey2}{rgb}{0.75, 0.75, 0.75}

\definecolor{Blue0}{rgb}{0.7752402921953095, 0.8583006535947711, 0.9368242983467897}
\definecolor{Blue1}{rgb}{0.41708573625528644, 0.6806305267204922, 0.8382314494425221}
\definecolor{Blue2}{rgb}{0.1271049596309112, 0.4401845444059977, 0.7074971164936563}
\definecolor{Blue3}{rgb}{0.03137254901960784, 0.18823529411764706, 0.4196078431372549}

\definecolor{BluePy}{rgb}{0.69140625, 0.8164, 0.8868}
\definecolor{RedC}{rgb}{0.9453, 0.7266, 0.6602}

\definecolor{gw}{rgb}{1, 0, 0}
\definecolor{jb}{rgb}{0, 1, 0}
\definecolor{tg}{rgb}{0, 0, 1}
\definecolor{jh}{rgb}{1, 1, 0}

\newcommand{\gw}[1]{{#1}}
% \newcoxmmand{\gw}[1]{\textcolor{gw}{#1}}
\newcommand{\jb}[1]{\textcolor{jb}{#1}}
\newcommand{\tg}[1]{\textcolor{tg}{#1}}
\newcommand{\jh}[1]{\textcolor{jh}{#1}}

### 2023-12-21 appendices/CO2_budget.tex
Drop Chapter on case_study
From atom_vs_molecule paper

\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed for each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead. Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050. The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat. In the former, this is linked to the extended use of oil products through naphtha-cracking. The latter is produced by industrial coal boilers for longer, until 2040. Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.49\textwidth]{GWP_per_sector_REF.pdf}
\includegraphics[width=0.49\textwidth]{GWP_per_sector_lin.pdf}
\caption{Respecting the \ce{CO2}-budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
\label{fig:app_CO2_REF_lin}
\end{figure}

### 2023-12-21 appendices/belgium_2020.tex
Drop Chapter on case_study
From atom_vs_molecule paper

The Belgian whole-energy system of 2020 was largely based (88.6\% of the primary energy mix) on \og conventional fuels\fg (\ie oil and oil products (38.2\%), natural gas (29.5\%), uranium (16.3\%) and solid fossil fuels (4.6\%) while the rest mainly accounts for 26.7\,TWh of lignocellulosic and wet biomass, 12.8\,TWh of wind and 5.1\,TWh of solar \cite{spf_economy_2022}. Given the data available in the literature (mostly for the power sector) and, when not available, following the assumptions made by \citet{Limpens2020}, \Cref{tab:Belgium_2020} gives the major technologies used in 2020 to supply the different demands of \Cref{fig:cs_demands}.

\begin{table}[htbp!]
\caption{Major technologies used to supply the 2020-demands of \Cref{fig:cs_demands} in terms of share of production and installed capacity. $^{(a)}$ The decentralised heating units provide 98\% of the low-temperature heat demand. $^{(b)}$ The private mobility accounts for 80\% of the passengers mobility.}
\label{tab:Belgium_2020}
\begin{minipage}{\linewidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l c c c}
\toprule
\multirow{2}{*}{\textbf{End-use demand}} & \textbf{Major} & \textbf{Share of} & \textbf{Installed}\\
    &	 \textbf{technologies} 	& \textbf{supply} & \textbf{capacity}	 \\ 	
\midrule							
\multirow{3}{*}{Electricity}
 & Nuclear & 39\% & 5.9\,GW\\
 & CCGT & 21\% & 3.9\,GW\\
 & Wind turbines & 14\% & 5.0\,GW\\
\midrule
\multirow{3}{*}{Heat High-Temp.}
 & Gas boiler & 36\% & 3.3\,GW\\
 & Coal boiler & 30\% & 2.3\,GW\\
 & Oil boiler & 20\% & 1.5\,GW\\
\midrule
\multirow{3}{*}{Heat Low-Temp. (DEC)$^{(a)}$}
 & Oil boiler & 48\% & 21.4\,GW\\
 & Gas boiler & 40\% & 17.5\,GW\\
 & Wood boiler & 10\% & 4.4\,GW\\
\midrule
\multirow{3}{*}{Heat Low-Temp. (DHN)}
 & Gas CHP & 59\% & 0.3\,GW\\
 & Gas boiler & 15\% & 0.3\,GW\\
 & Waste CHP & 15\% & 0.1\,GW\\
\midrule
\multirow{2}{*}{Private mobility$^{(b)}$}
 & Diesel car & 49\% & 93.5\,Mpass.-km/h\\
 & Gasoline car & 49\% & 94.7\,Mpass.-km/h\\
 & HEV & 2\% & 5.9\,Mpass.-km/h\\
\midrule
\multirow{3}{*}{Public mobility}
 & Diesel bus & 43\% & 3.6\,Mpass.-km/h\\
 & Train & 43\% & 3.9\,Mpass.-km/h\\
 & CNG bus & 5\% & 0.8\,Mpass.-km/h\\
\midrule
\multirow{3}{*}{Freight mobility}
 & Diesel truck & 74\% & 62.7\,Mt.-km/h\\
 & Diesel boat & 15\% & 10.8\,Mt.-km/h\\
 & Train & 11\% & 2.5\,Mt.-km/h\\
\midrule
HVC & Naphtha/LPG cracking & 100\% & 4.6\,GW\\
Ammonia & Haber-Bosch & 100\% & 1\,GW\\
Methanol & Import & 100\% & -\\
\bottomrule							
\end{tabular}}
\end{minipage}
\end{table}

### 2023-12-21 chap_ES_PCE/chap_ES_PCE.tex
Drop Chapter on case_study
From atom_vs_molecule paper

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode
In this chapter, we introduce the two main tools, not exhaustively of course (because already presented, in Gauthier and Diederik's theses) on which I based my work: EnergyScope and RHEIA/PCE. I'd like too to talk about the case study (Belgian energy system) that will be addressed all along the manuscript but I'm afraid it'd be too many stuff in one chapter. On the other side, I would not dedicate an entire chapter to any of these three parts. What's your opinion?
\section{EnergyScope: To optimise the energy transition pathway}
\label{sec:energyscope}

\subsection{EnergyScope TD}
\label{subsec:estd}
Presentation of the capacity to model a whole-energy system,with a hourly resolution, and of the main equations of the snapshot model (similarly to what I've done in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}).

\subsection{EnergyScope Pathway}
\label{subsec:espathway}
From snapshot to pathway optimisation, presentation of the main equations to link the representative years, focus on the salvage value (not presented in Gauthier's thesis but well in our Pathway paper)

\section{RHEIA: To quantify the uncertainties}
\label{sec:rheia}

\subsection{Uncertainty characterisation}
\label{subsec:uncert_charac}
Presentation here of the uncertainty characterisation from S. Moret and adding some parameters specific to the pathway model (\eg $\Delta_{\text{change}}$ linked to change speed) or specific to the main case study (\eg possibility to have nuclear SMR from 2040)

\subsection{Polynomial Chaos Expansion}
\label{subsec:pce}
Similarly to what we did in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), presentation of the PCE and Sobol' sequence (which is a more optimised way to explore the ranges of uncertainties, compared, for instance, to random exploration).

\subsection{Preliminary screening and selection}
\label{subsec:screening}
Similarly to what we did in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), emphasise that we only consider a limited amount of uncertain parameters to keep a reasonable computation time while capturing the impact of (almost) all the uncertainties.

\section{Case study: The Belgian energy system}
\label{sec:case_study}
Similarly to what we did in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}),I'd present generally here the case study, much shorter than what Gauthier did in his thesis. It'd be presenting the different sets of resources, technologies and demands.

Here, I could integrate the information on the NED and how important it is to consider it as it represents 10\% worldwide (and even 20\% in Belgium) of the energy consumption, especially because it is usually a sector that is overlooked in whole-energy system optimisation.

### 2023-12-21 chap_case_study/chap_case_study.tex
Drop Chapter on case_study
From atom_vs_molecule paper

As detailed by \citet{limpens2023pathway}, the analysis carried out in this work can be applied to any regional whole-energy system. As a densely-populated and highly-industrialised country with limited local renewable potentials (mainly solar and wind), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon-neutrality in 2050 makes it an intricate case study. Moreover, this case study and the subsequent analyses can be transferred - to some extent - to other industrialised countries highly dependent on fossil fuels with limited local renewable potentials (\eg the Netherlands or Germany) \cite{dommisse2020modelling}. This section presents the different demands to satisfy as well as the resources available and the conversion technologies to supply those. The focus here is given to the renewable molecules to import from abroad as well as the techno-economic details of \gls{SMR}. For a comprehensive understanding and detailed descriptions of the technologies, please refer to the documentation \cite{readthedocs_pathway}. Finally, the \ce{CO2}-budget over the 2020-2050 transition is presented.

\section{Demands}
\label{sec:cs:demand}

End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative years of the transition, in order to account for their daily or seasonal variability. In this work, the yearly end-use demands (EUD) for all sectors are calculated from the rather slightly increasing forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}). Although, given a significant and unsubstantiated discrepancy in the non-energy use forecasts compared to their previous report (\ie +80\% over the 2020-2030 time window), the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}. Looking at \Cref{fig:cs_demands}, between 2020 and 2050, one observes a noteworthy increase of the electricity (+40\%), passenger (+45\%) and freight mobility (+35\%) demands. The rise of the non-energy demand is more limited, \ie +6\%, whereas the heating demands is forecast to decrease: -11\% and -3\% respectively for the low and high-temperature heat demands. Regarding the center graph of \Cref{fig:cs_demands}, it is the aggregation of the same data as in the left graph but per category, rather than per sector, with the non-energy demand being associated with the industry. This illustrates how industrialised is Belgium, compared to households and services, and, consequently, highly energy-intensive. The right graph of \Cref{fig:cs_demands} gives the passenger and the freight mobility. The sharp increase from 2020 to 2025 is due the COVID-crisis that has significantly reduced these demands in 2020. As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for parts of electricity and low-temperature heating demands \cite{Limpens2020}. A daily time series is used for the passenger mobility and applied similarly to every typical days. Finally, for the other demands, the yearly demand is distributed uniformly over the different hours of the year.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.32\textwidth]{EUD_sec.pdf}
\includegraphics[width=0.32\textwidth]{EUD_cat.pdf}
\includegraphics[width=0.32\textwidth]{EUD_mob.pdf}
\caption{EnergyScope splits the whole-energy system end-use demands (EUD) into two sets: (non)-energy and transport-related. This figure presents the nominal values of each of these demands. In the center graph, the non-energy demand has been fully associated with the industrial demand. As detailed by \citet{rixhon2022integration}, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh. Graphs have been adapted from \cite{limpens2023pathway}.}
\label{fig:cs_demands}
\end{figure}

\section{Resources}
\label{sec:cs:resources}
To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$, their global warming potential, $\mathit{gwp}_{\mathrm{op}}$, as well as their 
availability, as detailed by \citet{limpens2023pathway}. \Cref{fig:cs_resources_cost} depicts the evolution the respective costs of purchasing. Regarding ``renewable electrofuels'', these are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis\cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''. Then, besides their cost, the resources are either limited or unlimited in terms of availability and either renewable or not. The limitation in terms of availability can be direct or indirect. On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use. On the other hand, wind, solar, hydro and uranium are limited by the technical potentials respectively, of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and nuclear power plants (6\,GW). As \gls{SMR} are foreseen, if installed, to be around the same locations (\ie Thiange and Doel) as the conventional nuclear power plants and using the same area in kW/ha, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}. This is even without considering the potential limit due to the local availability, in terms of volume and flow rate, of enough water that would be a more socially-accepted solution than cooling towers exhausting a dense plume. Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050). In the current work, the electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``sustainable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}. In practice, it means that their \gls{GWP} is assumed to be zero in the model. Regarding specifically these electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco). As the amount to import from each of these locations is hard to forecast, the current work considers the average cost between the different locations. Besides these, every other resource has its specific \gls{GWP} like coal ($\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), natural gas ($\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh) or the fossil-based molecules equivalent to the electrofuels (\eg $\mathit{gwp}_{\mathrm{op,ammonia}}=0.46$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,methanol}}=0.41$\,kt$_{\ce{CO2},\text{eq}}$/GWh).

\begin{figure}[htbp!]
\centering

\includegraphics[width=0.32\textwidth]{Res_ren.pdf}
\includegraphics[width=0.32\textwidth]{Res_foss.pdf}
\includegraphics[width=0.32\textwidth]{Res_others.pdf}
\caption{Cost of purchasing the different resources. Besides the free local renewables (\ie sun, wind and hydro) limited by technical potentials, EnergyScope accounts for renewable energy carriers and their respective fossil counterparts (left and center graphs). These fuels can be imported from abroad without limitation on their availability. Other carriers are limited either by their local potentials (\ie biomass and waste) or other considerations like the power grid interconnections or the capacity of nuclear power plants.}
\label{fig:cs_resources_cost}
\end{figure}

\section{SMR and other conversion technologies}
\label{sec:cs:technologies}
Finally, as the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, technologies are implemented to convert these resources into the end-use demands. Besides their CAPEX, OPEX and lifetime defined in Section \ref{subsec:meth:ES_Pathway}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) exhibit their own charge/discharge losses. Eventually, there are also infrastructure technologies like the grid or the \gls{DHN} that allow to account for the investment necessary, respectively, to integrate more intermittent renewables in the power sector and the expand the use of centralised heating systems.

A specific attention is to put on the implementation of \glsxtrfull{SMR} whereas the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035. Similarly to the analysis of \citet{PATHS2050}, a Belgian consortium for energy research, and in line with the Belgian Nuclear Research Centre (SCK-CEN) \cite{SCK-CEN_SMR}, \gls{SMR} are implemented with the features listed in \Cref{tab:SMR_features}. Where most of the features are similar to conventional nuclear power plants, it differs from these on two main points: their potential year start, 2040, and their flexibility. Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity independently at any hour of each representative year. Here, the authors simplify SMRs as only producing electricity and assume that the after-heat is lost to the atmosphere anyway.

\begin{table}[htbp!]
\caption{Nominal features of the SMRs in EnergyScope. \gls{SMR} exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year. $^{(a)}$ This annual availability accounts for yearly maintenance where the reactors might not operate or, at least, not at their maximum capacity. $^{(b)}$ 2040 is the soonest year at which \gls{SMR} could be available, optimistically assuming industrial prototypes being completed by 2035 and 5 additional years for their commercial installation.}
\label{tab:SMR_features}
\centering
\begin{tabular}{l c c|c}
\toprule
\multirow{2}{*}{\textbf{Feature}} & \multirow{2}{*}{\textbf{Value}} & \multirow{2}{*}{\textbf{Unit}} & \textbf{Similarity with}\\
 & & & \textbf{conventional nuclear}\\
\midrule
CAPEX & 4850 & €/kW & \checkmark\\
Annual OPEX & 103 & €/kW/year & \checkmark\\
Lifetime & 60 & year & \checkmark\\
Efficiency & 40\% & -& \checkmark\\
Maximum capacity & 6 & GW & \checkmark\\
Annual availability & 85\%$^{(a)}$ & -& \checkmark\\
\midrule
Operational year & 2040$^{(b)}$ & - & \xmark\\
Flexibility & Full & - & \xmark\\
\bottomrule

\end{tabular}
\end{table}

For the sake of comparison, \Cref{fig:LCOE} gives the \gls{LCOE} of the principal technologies to produce electricity, based on the computation used by \citet{limpens2021generating}. Compared to the other flexible generation units, \gls{SMR} is significantly more cost-effective. In addition, we see that \gls{CCGT} supplied by e-ammonia outcompetes its e-methane equivalent, unlike their respective fossil-based equivalent.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.6\textwidth]{LCOE.pdf}
\caption{Levelised cost of energy (LCOE) for the main technologies in the power sector. Where \gls{SMR} is cheaper than the other flexible options, CCGT running on e-ammonia is, a priori, cheaper than its e-methane alternative.}
\label{fig:LCOE}
\end{figure}

\section{\ce{CO2}-budget for the transition}
\label{sec:cs:CO2-budget}
In most of the studies carried out on the pathway optimisation of a whole-energy system, a \ce{CO2}-trajectory is \textit{a priori} set to reach carbon-neutrality by 2050. \citet{nerini2017myopic} used the emission trajectory indicated by the UK's Committee on Climate Change in their analysis of the impact of limited foresight to achieve the target of 80\% reduction of \gls{GHG} by 2050 in the United Kingdom. In their assessment of the impacts of economy-wide emissions policies in the water-energy-land nexus, \citet{licandeo2023assessing} analysed different \ce{CO2}-trajectories considering more or less severe water scarcity for the US. \citet{poncelet2016myopic} with LUSYM (Leuven University SYstem Model) and \citet{PATHS2050} with TIMES-BE also set decreasing emission trajectories in their analysis of respectively the Belgian power sector and whole-energy system.  Others only set the objective as the carbon-neutrality by 2050. For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of ``unicorn technologies'' or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.

In this work, the effect of greenhouse gases is cumulative over time and a constraint is set on the overall emissions of the transition---a \ce{CO2}-budget for the transition. This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}. Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in
Eq.\,(\ref{eq:limit_gwp_trans}). This is another sign of the urgency to act to mitigate climate change as this 30-year budget represents only 10 years of the current emissions. Compared to a linear decrease from the current emissions, as done by \citet{limpens2023pathway}, this budget represents a 60\% reduction of the cumulative emissions over the transition.  Appendix \ref{app:CO2_budget} compares the emissions trajectory between the REF case and a case (without \gls{SMR}) where the linear decrease is imposed between 2020 and carbon-neutrality in 2050.

\textbf{GRANDFATHERING dans la vision d'attribuer le budget CO2 donné à la Belgique. Defining the share of the safe operating space from How to bring absolute sustainability into decision-making: An industry case study using a Planetary Boundary-based methodology. Sharing the safe operating space}

### 2023-12-21 chap_electro_uq/chap_electro_uq.tex
Drop Chapter on case_study
From atom_vs_molecule paper

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

In this chapter aims at showing the first step of uncertainty quantification and highlight a first insight in terms of impacting parameters. It'd also highlight that depending on the gwp\_limit 
\section{From a cost-optimised to a carbon-neutral Belgian energy system: progressive defossilisation}
\label{sec:chap_2_ses_sec2}
In this section, I'd include the section 2 of the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027})(without 2.2 where I presented EnergyScope that I'd include in Chapter 1).

\subsection{Electrofuels}
\label{subsec:chap2_electrofuels}
This section would be as the section 2.1 of the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), where I present more specficially how electrofuels are implemented in the model

\subsection{Reference Case Study: The Belgian Energy System in 2050}
\label{subsec:chap2_case_study}
Basically section 2.3 of the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), where I present more specficially how electrofuels are implemented in the model where I present what a cost-optimised (without gwp-constraint) Belgian energy system would look like

\subsection{A step by step defossilisation of the snapshot system}
\label{subsec:chap2_defossilisation}
As in section 2.4, quickly show how we implement the defossilisation without having an entire pathway.

\section{Results}
\label{sec:chap2_results}
Similar to Section 4 of the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}). There'll be work to generate up-to-date results as the model evolved (no more FC\_cars but BEV, for instance)
\subsection{Statistical analysis of the cost}
\label{subsec:chap_2_stat_analysis}

\subsection{Critical parameters}
\label{subsec:chap_2_crit_param}

\section{Discussion and perspective with the literature}
Similar to section 5.1 of the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027})

### 2023-12-21 options/symbols_and_cmd.tex
Drop Chapter on case_study
From atom_vs_molecule paper

@@
 
+%% B
+\newabbreviation[group={Acronyms}]{BEV}{BEV}{battery electric vehicle}
+
+%% C
+\newabbreviation[group={Acronyms}]{CCGT}{CCGT}{combined cycle gas turbine}
+
+%% D
+\newabbreviation[group={Acronyms}]{DHN}{DHN}{district heating network}
+
+%% E
+\newabbreviation[group={Acronyms}]{EUD}{EUD}{end-use demand}
+
+%% F
+\newabbreviation[group={Acronyms}]{FC}{FC}{fuel cell}
+
+%% G
+\newabbreviation[group={Acronyms}]{GWP}{GWP}{global warming potential}
+
+%% H
+\newabbreviation[group={Acronyms}]{HP}{HP}{heat pump}
+\newabbreviation[group={Acronyms}]{HVC}{HVC}{high value chemicals}
+
+%% L
+\newabbreviation[group={Acronyms}]{LCOE}{LCOE}{levelised cost of energy}
+
+%% N
+\newabbreviation[group={Acronyms}]{NED}{NED}{non-energy demand}
+
 %% P
 \newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
+\newabbreviation[group={Acronyms}]{PV}{PV}{photovoltaic}
 
@@
 \newabbreviation[group={Acronyms}]{SDGs}{SDGs}{Sustainable Development Goals}
+\newabbreviation[group={Acronyms}]{SMR}{SMR}{small modular reactor}
 
@@
 \newcommand{\etal}{\textit{et~al.}}
-\newcommand{\eg}{e.g.\ }
 
@@
 %% I
-\newcommand{\ie}{i.e.\ }
 
@@
 \newcommand{\xb}{\boldsymbol{x}}
+\newcommand{\xmark}{\ding{55}}
 

### 2024-01-10 chap_case_study/chap_case_study.tex
Case study: 1st version

@@
-As detailed by \citet{limpens2023pathway}, the analysis carried out in this work can be applied to any regional whole-energy system.
+%\begin{flushright}
+%%%\emph{``For the things we have to learn before we can do them, we learn by doing them.''}\\%Each wrong scenario you have analysed, is a wrong decision you won't made.''}\\
+%%%Aristote
+%\emph{All models are wrong, but some are useful.}\\
+%George P. Box
+%\end{flushright}
+
+%\medskip
+%
+%\begin{mybox}{Chapter overview}
+%\begin{itemize}[left=0em]%[leftmargin=0cm,itemindent=.5cm,labelwidth=\itemindent,labelsep=0cm,align=left]
+%\setlength\itemsep{-0.3em}
+%\item Belgian energy system overview%Case studies: the Belgian energy systems during the transition (2015-2050).
+%\end{itemize}
+%\vspace{-0.3cm}
+%
+%\emph{This chapter is an improved and extended case study version of \citet{Limpens_belgian_2020}}.
+%\end{mybox}
+%
+%\medskip
+
+As detailed by \citet{limpens2024pathway}, the analysis carried out in this work can be applied to any regional whole-energy system.
 As a densely-populated and highly-industrialised country with limited local renewable potentials (mainly solar and wind), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon-neutrality in 2050 makes it an intricate case study.
 Moreover, this case study and the subsequent analyses can be transferred - to some extent - to other industrialised countries highly dependent on fossil fuels with limited local renewable potentials (\eg the Netherlands or Germany) \cite{dommisse2020modelling}.
-This section presents the different demands to satisfy as well as the resources available and the conversion technologies to supply those.
-The focus here is given to the renewable molecules to import from abroad as well as the techno-economic details of \gls{SMR}.
+This chapter presents the different demands to satisfy, with a particular focus on the non-energy demand, as well as the resources available and the conversion technologies to supply those.
 For a comprehensive understanding and detailed descriptions of the technologies, please refer to the documentation \cite{readthedocs_pathway}.
@@
 
-\section{Demands}
+\section{Contributions}
+\label{sec:cs:contributions}
+First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heart and mobility), the \gls{NED} is often not considered.
+The latter is defined as ‘\textit{energy products used as raw materials in the different sectors; that is not consumed as a fuel or transformed into another fuel}’ \cite{Eurostat2019}.
+The previous analyses carried out with \gls{ESTD} considered the non-energy demand as the related primary energy needs, \ie either natural gas or \gls{LFO}.
+To minimise the total cost of the system, the model simply selected the cheapest between the two resources (\ie natural gas).
+This work goes one step further and accounts for the \gls{NED} as a demand of three commodities (\ie \gls{HVC}, ammonia and methanol) as well as with associated production technologies.
+This allows bringing the non-energy sector to a similar level of details as the other sectors.
+Regarding the end-use demands,  keeping the same methodology to define \gls{EUD} in EnergyScope as \citet{Limpens2020}, this work considers updated values given the latest release of the \og EU reference scenario 2020 : energy, transport and GHG emissions: Trends to 2050 \fg by the European Commission \cite{EuropeanCommission2021}.\\
+
+Second, given the focus of this work on the electrofuels, the case study includes a more explicit representation of e-ammonia and e-methanol (as well as their fossil-based equivalents), on top of e-hydrogen and e-methane, already included in the previous definition of the case study \cite{limpens2021generating}.
+As detailed later on, these electrofuels are considered as renewable in the sense that their \gls{GWP} is zero.
+This more explicit representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.\\
+
+Third, as nuclear energy could be a real game-changer in the energy transition worldwide \cite{IEA2022nuclear}, and especially in Belgium, this thesis has integrated the potential to install \gls{SMR} from 2040 onward.\\
+
+Fourth, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.\\
+
+Finally, to a smaller extent, this work includes updated values for some parameters compared to the work of \citet{limpens2021generating}.
+The main change concerns the cost and performance of private mobility vehicles, which is specifically a key components in the Belgian energy transition.
+The previous version of the case study was excessively favouring fuel cell car versus \gls{BEV}.
+Comparing with other works \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
+Regarding \gls{BEV}, while the CAPEX has been kept unchanged, the efficiency and the battery capacity, \ie the range, have been increased.
+As seen in the results, this change of data made \gls{BEV} often more competitive than its hydrogen-based equivalent.
+
+\section{End-use demands}
 \label{sec:cs:demand}
+End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative years of the transition, in order to account for their daily or seasonal variability \cite{Limpens2020,limpens2021generating}.
+In this work, the yearly end-use demands (EUD) for all sectors are calculated from the rather slightly increasing forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}).
 
-End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative years of the transition, in order to account for their daily or seasonal variability.
-In this work, the yearly end-use demands (EUD) for all sectors are calculated from the rather slightly increasing forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}).
+\subsection{Non-energy demand}
+\label{subsec:cs:NED}
+Where previously published works of \citet{rixhon2021comprehensive,rixhon2022integration} investigated more extensively the integration of the \gls{NED} in the case study of Belgium, this section summarises the rationale of doing so as well as the methodology used to quantify this demand.\\
+
+\textbf{Definition and historical trend}\vspace{-0.3cm}\\
+
+The \gls{NED} can be split into four main categories of final molecules \cite{IEA2018_petrochemicals}: (i) \gls{HVC} (worldwide production of $\sim$365Mt/year); (ii) ammonia ($\sim$185Mt/year); (iii) methanol ($\sim$100Mt/year) and (iv) the other products.
+\Gls{HVC} gather the light olefins (e.g.
+ethylene, propylene) and aromatics (benzene, toluene, xylene – BTX), mainly for the production of plastics, synthetic fibers or rubber.
+Their production today relies mainly on petroleum products such as naphtha, ethane or liquified petroleum gas.
+Ammonia is  mainly used for the production of fertilizers ($\sim$80\% of global ammonia consumption).
+Its production is dominated by \gls{NG} via steam reforming to produce hydrogen, used as feedstock in the Haber-Bosch process.
+Methanol is mainly converted to formaldehyde (resin) but also used for the production of other chemicals (e.g.
+solvents and gasoline-blends).
+Currently, its synthesis, like ammonia, is mainly relying on natural gas via steam reforming.
+Finally, the other products gather all chemicals not mentioned in the other categories such as bitumen, lubricants and other heavy products from oil refineries \cite{daioglou2014energy}.\\
+
+The \gls{NED} currently represents around 20\% of the final energy consumption in Belgium \cite{FPSEconomy2021}.
+Over the recent history, there has been a relatively constant evolution of three main categories of the final consumption for non-energy use in Belgium, \cite{statbel_NED_2019}: (i) naphtha and \gls{LPG} (between 59\% and 67\% of the total final consumption, around 59.4\,TWh in 2019), (ii) \gls{NG} (between 9\% and 14\%, 11.8\,TWh in 2019), and (iii) others (\ie bitumen, coal tar and other oil products) (between 21\% and 28\%).
+Naphtha and \gls{LPG} are consumed in a naphtha cracker, which results in ethylene and propylene, what will be considered as \gls{HVC} in the rest of this work.
+Similarly, \gls{NG}, as non-energy carrier, is used in steam methane reformer to produce the required hydrogen to the synthesis process of ammonia.
+The small shares of bitumen and coal tar are respectively used for roadworks and to produce synthetic gas through gasification.
+Finally \og other oil products\fg take into account, indistinguishably, tar and sulphur as well as by-products of the refineries (\eg \gls{BTX}).
+About methanol, there is currently no production plant in Belgium even if the country plays a role in trading this commodity between its neighbouring countries and consumes part of what it imports.\\
+
+\textbf{Methodology of quantification}\vspace{-0.3cm}\\
+
+The non-energy demand studied in this analysis focuses on the chemical industry (more than 90\% of the non-energy use in Belgium) and, similarly to other studies \cite{IEA2018_petrochemicals, daioglou2014energy}, is split between the three aforementioned main groups of products (\ie \gls{HVC}, ammonia and methanol).
+Before describing these three demands, this study excludes bitumen, coal tar and \og other oil products\fg.
+The first two represent marginal shares of the current non-energy use in such a way that they should not affect the big trends provided by this study.
+As described previously, the latest are mostly by-products from refineries that the system uses because they are available.
+However, in a perspective of defossilisation, since the future of fossil-based refineries is unclear, they have not be implemented in this study nor their by-products.\\
+
+Regarding \gls{HVC}, the future of their production is highly uncertain.
+One of the reasons is new regulations and strategies promoting recycling and limitation of single-use plastics \cite{EU_plastics}.
+Besides this uncertainty, Belgium stays a major exporter as approximately 2/3 of plastic raw materials produced locally are exported abroad \cite{agoria_plastics}.
+Even if a significant part of \gls{HVC} produced in Belgium is not locally consumed, this demand has been set based on the assumption that Belgium will keep its industrial activity in this sector.
+This assumption does not consider the net imports of \gls{HVC}, unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
+Therefore, the actual demand of \gls{HVC} is inferred from the consumption of naphtha and \gls{LPG} as non-energy use as well as energy-carrier in the chemical and petrochemical industries \cite{statbel_NED_2019}.
+This assumption is based on the fact that, in the conversion processes to produce \gls{HVC} from naphtha or \gls{LPG}, these fuels also serve as energy-carrier to supply the process itself.
+Then, given the respective efficiencies (1.83t$_{\text{naphtha}}$/t$_{\text{HVC}}$ and 1.67t$_{\text{LPG}}$/t$_{\text{HVC}}$) \cite{IEA2018_petrochemicals}, the current demand of \gls{HVC} is estimated equal to 3069\,kt, without making distinctions between the different chemicals (\ie ethylene, propylene and \gls{BTX}).
+\\
+
+The ammonia sector in Belgium is quite different: the country locally produces and imports ammonia much more than it exports it.
+Thanks to a database from the United Nations \cite{UN_statistics} and the National Bank of Belgium, it has been identified that, over the last ten years, Belgium has imported, exported and locally produced, on average, respectively, 1010\,kt, 105\,kt and 990\,kt of ammonia.
+Therefore, on top of the local production, the net import (\ie import minus export) is also included in this non-energy demand.
+This gives a current demand of 1895\,kt of ammonia.\\
+
+Concerning the demand of methanol, similarly to ammonia, this work solely considers the net imports as there is no local production in Belgium.
+To define the actual non-energy demand of methanol, only a 51\%-share of this net import is kept since, according to the Methanol Institute and \gls{MMSA}, this share is used for formaldehyde production in Belgium \cite{MMSA51}.
+The rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
+This methodology gives a current non-energy demand of methanol of 269\,kt.\\
+
+Finally, after converting these masses of products into energy content (\ie LHV: \gls{HVC} - 47\,MJ/kg, ammonia - 18.8\,MJ/kg and methanol - 19.9\,MJ/kg), this work assumed constant shares between the three categories within the \gls{NED} and the same growing rate as presented by \citet{EuropeanCommission2016}.
+
+\subsection{Forecast of the demands over the transition}
+\label{subsec:cs:EUD_forecast}
 Although, given a significant and unsubstantiated discrepancy in the non-energy use forecasts compared to their previous report (\ie +80\% over the 2020-2030 time window), the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}.
@@
 In the center graph, the non-energy demand has been fully associated with the industrial demand.
-As detailed by \citet{rixhon2022integration}, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh.
-Graphs have been adapted from \cite{limpens2023pathway}.}
+As detailed previously, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh.
+Graphs have been adapted from \cite{limpens2024pathway}.}
 \label{fig:cs_demands}
@@
 To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$, their global warming potential, $\mathit{gwp}_{\mathrm{op}}$, as well as their
-availability, as detailed by \citet{limpens2023pathway}.
+availability, as detailed by \citet{limpens2024pathway}.
 \Cref{fig:cs_resources_cost} depicts the evolution the respective costs of purchasing.
@@
 
-\section{SMR and other conversion technologies}
+\section{Conversion technologies: SMR, \gls{NED}-related and others}
 \label{sec:cs:technologies}
-Finally, as the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, technologies are implemented to convert these resources into the end-use demands.
+As the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, for instance, technologies are implemented to convert these resources into the end-use demands.
 Besides their CAPEX, OPEX and lifetime defined in Section \ref{subsec:meth:ES_Pathway}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) exhibit their own charge/discharge losses.
 Eventually, there are also infrastructure technologies like the grid or the \gls{DHN} that allow to account for the investment necessary, respectively, to integrate more intermittent renewables in the power sector and the expand the use of centralised heating systems.
+The exhaustive list of these technologies have already been presented in previous works \cite{limpens2021generating}.\\
 
@@
 \centering
-\includegraphics[width=0.6\textwidth]{LCOE.pdf}
+\includegraphics[width=0.8\textwidth]{LCOE.pdf}
 \caption{Levelised cost of energy (LCOE) for the main technologies in the power sector.
@@
 \label{fig:LCOE}
+\end{figure}
+
+\autoref{fig:NED_tech} illustrates the different paths to produce the final molecules of the NED integrated within the model.
+Similarly to \cite{tsiropoulos2018emerging}, naphtha, here considered as \gls{LFO}, resulting from refinery operation is modeled as an imported commodity.
+Presented here for the specific year of 2035, all data and related references can be found in \cite{GIT_NED}.
+To keep the same level of details with other sectors of the model, the implementation of the conversion technologies consists of a single kind of technology per type of resource to produce a certain product.
+For instance, in the model, there is only one technology to produce \gls{HVC} either from naphtha or from LPG, two liquid fossil hydrocarbons, \ie \gls{NSC}.
+For ammonia and methanol, the molecules can either be produced locally from other resources or directly imported (with distinction between non-renewable and renewable molecules).
+Some other technologies included in the model (not represented here for the sake of clarity), are able to turn some resources presented here into others (\eg \gls{NG} to hydrogen, woody biomass to hydrogen or to synthetic natural gas, more information in \cite{Limpens2020}).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{NED_tech.pdf}
+\caption{Schematic view of the different resources able to produce \gls{HVC}, ammonia and methanol with their related conversion technologies (including energy efficiency and their CAPEX - in €/kW of final molecules).
+Values stand for 2035.
+Graph from \cite{rixhon2021comprehensive}}
+\label{fig:NED_tech}
 \end{figure}
@@
 Others only set the objective as the carbon-neutrality by 2050.
-For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of ``unicorn technologies'' or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.
+For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of \og unicorn technologies\fg or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.\\
 
 In this work, the effect of greenhouse gases is cumulative over time and a constraint is set on the overall emissions of the transition---a \ce{CO2}-budget for the transition.
+The arbitrarily chosen way of attributing emissions-budget to Belgium is usually called \og grandfathering\fg.
+Even though this approach has his pros and cons out of discussion within the scope of this work, it consists in \og maintaining that prior emissions increase future emission entitlements\fg \cite{knight2013grandfathering}.
 This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.
-Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in
-Eq.\,(\ref{eq:limit_gwp_trans}).
+Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in Eq.\,(\ref{eq:limit_gwp_trans}).
 This is another sign of the urgency to act to mitigate climate change as this 30-year budget represents only 10 years of the current emissions.
-Compared to a linear decrease from the current emissions, as done by \citet{limpens2023pathway}, this budget represents a 60\% reduction of the cumulative emissions over the transition.
+\\
+
+Compared to a linear decrease from the current emissions, as done by \citet{limpens2024pathway}, this budget represents a 60\% reduction of the cumulative emissions over the transition.
 Appendix \ref{app:CO2_budget} compares the emissions trajectory between the REF case and a case (without \gls{SMR}) where the linear decrease is imposed between 2020 and carbon-neutrality in 2050.
-
-\textbf{GRANDFATHERING dans la vision d'attribuer le budget CO2 donné à la Belgique.
-Defining the share of the safe operating space from How to bring absolute sustainability into decision-making: An industry case study using a Planetary Boundary-based methodology.
-Sharing the safe operating space}

### 2024-01-10 options/symbols_and_cmd.tex
Case study: 1st version

@@
 \newabbreviation[group={Acronyms}]{BEV}{BEV}{battery electric vehicle}
+\newabbreviation[group={Acronyms}]{BTX}{BTX}{benzene, toluene and xylene}
 
@@
 %% E
+\newabbreviation[group={Acronyms}]{ESTD}{EnergyScope TD}{EnergyScope Typical Days}
 \newabbreviation[group={Acronyms}]{EUD}{EUD}{end-use demand}
@@
 %% G
+\newabbreviation[group={Acronyms}]{GHG}{GHG}{greenhouse gases}
 \newabbreviation[group={Acronyms}]{GWP}{GWP}{global warming potential}
@@
 \newabbreviation[group={Acronyms}]{LCOE}{LCOE}{levelised cost of energy}
+\newabbreviation[group={Acronyms}]{LFO}{LFO}{light fuel oil}
+\newabbreviation[group={Acronyms}]{LPG}{LPG}{liquefied petroleum gas}
+
+%% M
+\newabbreviation[group={Acronyms}]{MMSA}{MMSA}{Methanol Market Services Asia}
+\newabbreviation[group={Acronyms}]{MTBE}{MTBE}{methyl tert-butyl ether}
 
@@
 \newabbreviation[group={Acronyms}]{NED}{NED}{non-energy demand}
+\newabbreviation[group={Acronyms}]{NG}{NG}{natural gas}
+\newabbreviation[group={Acronyms}]{NRE}{NRE}{non-renewable energy}
+\newabbreviation[group={Acronyms}]{NSC}{NSC}{naphtha steam cracker}
 
@@
 \newabbreviation[group={Acronyms}]{PV}{PV}{photovoltaic}
+
+%% R
+\newabbreviation[group={Acronyms}]{RE}{RE}{renewable energy}
 
@@
 \newcommand{\jh}[1]{\textcolor{jh}{#1}}
+
+% Boxes at the beginning of the chapter
+\usepackage{enumitem}
+\definecolor{dark_blue}{RGB}{0,33,78}
+\definecolor{very_light_blue}{RGB}{227, 241, 255}
+\usepackage{tcolorbox} % to create boxes in the text
+\newtcolorbox{mybox}[1]{colback=very_light_blue,colframe=dark_blue,fonttitle=\bfseries,title=#1}

### 2024-01-10 Figures.py

@@ -1,0 +1,753 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Jan 10 14:47:49 2024
+
+@author: xrixhon
+"""
+
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jul 20 10:52:57 2023
+
+@author: xrixhon
+"""
+
+import rheia.POST_PROCESS.post_process as rheia_pp
+import rheia.UQ.uncertainty_quantification as rheia_uq
+import matplotlib.pyplot as plt
+import plotly.express as px
+import plotly.graph_objects as go
+
+import pandas as pd
+import plotly.io as pio
+import numpy as np
+import os, sys
+from pathlib import Path
+import pickle as pkl
+import hashlib
+
+pio.templates.default = 'simple_white'
+pio.kaleido.scope.mathjax = None
+
+pio.renderers.default = 'browser'
+
+dir_to_save = '/Users/xrixhon/Documents/PhD_Thesis/Papers/5 - Nuke_vs_Mol/Images/'
+
+def dict_color_full():
+    categories = ['Sectors','Electricity','Heat_low_T','Heat_high_T','Mobility','Freight','Ammonia',
+               'Methanol','HVC','Conversion','Storage','Storage','Storage_daily','Resources',
+               'Infrastructure','Years_1','Years_2','Phases']
+    color_dict_full = dict.fromkeys(categories)
+
+    for c in categories:
+        color_dict_full.update(dict_color(c))
+    
+    return color_dict_full
+
+def dict_color(category):
+    color_dict = {}
+    
+    if category == 'Electricity':
+        color_dict = {"NUCLEAR":"deeppink", "NUCLEAR_SMR": "deeppink", "CCGT":"darkorange", "CCGT_AMMONIA":"slateblue", "COAL_US" : "black", "COAL_IGCC" : "dimgray", "PV" : "yellow", "WIND_ONSHORE" : "lawngreen", "WIND_OFFSHORE" : "green", "HYDRO_RIVER" : "blue", "GEOTHERMAL" : "firebrick", "ELECTRICITY" : "dodgerblue"}
+    elif category == 'Heat_low_T':
[…]

### 2024-01-12 appendices/app_case_study.tex
Case study + UQ ranges

\section{Belgian energy system in 2020} 
\label{app:bel_2020}
The Belgian whole-energy system of 2020 was largely based (88.6\% of the primary energy mix) on \og conventional fuels\fg (\ie oil and oil products (38.2\%), natural gas (29.5\%), uranium (16.3\%) and solid fossil fuels (4.6\%) while the rest mainly accounts for 26.7\,TWh of lignocellulosic and wet biomass, 12.8\,TWh of wind and 5.1\,TWh of solar \cite{spf_economy_2022}. Given the data available in the literature (mostly for the power sector) and, when not available, following the assumptions made by \citet{Limpens2020}, \Cref{tab:Belgium_2020} gives the major technologies used in 2020 to supply the different demands of \Cref{fig:cs_demands}.

\begin{table}[htbp!]
\caption{Major technologies used to supply the 2020-demands of \Cref{fig:cs_demands} in terms of share of production and installed capacity. $^{(a)}$ The decentralised heating units provide 98\% of the low-temperature heat demand. $^{(b)}$ The private mobility accounts for 80\% of the passengers mobility.}
\label{tab:Belgium_2020}
\begin{minipage}{\linewidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l c c c}
\toprule
\multirow{2}{*}{\textbf{End-use demand}} & \textbf{Major} & \textbf{Share of} & \textbf{Installed}\\
    &	 \textbf{technologies} 	& \textbf{supply} & \textbf{capacity}	 \\ 	
\midrule							
\multirow{3}{*}{Electricity}
 & Nuclear & 39\% & 5.9\,GW\\
 & CCGT & 21\% & 3.9\,GW\\
 & Wind turbines & 14\% & 5.0\,GW\\
\midrule
\multirow{3}{*}{Heat High-Temp.}
 & Gas boiler & 36\% & 3.3\,GW\\
 & Coal boiler & 30\% & 2.3\,GW\\
 & Oil boiler & 20\% & 1.5\,GW\\
\midrule
\multirow{3}{*}{Heat Low-Temp. (DEC)$^{(a)}$}
 & Oil boiler & 48\% & 21.4\,GW\\
 & Gas boiler & 40\% & 17.5\,GW\\
 & Wood boiler & 10\% & 4.4\,GW\\
\midrule
\multirow{3}{*}{Heat Low-Temp. (DHN)}
 & Gas CHP & 59\% & 0.3\,GW\\
 & Gas boiler & 15\% & 0.3\,GW\\
 & Waste CHP & 15\% & 0.1\,GW\\
\midrule
\multirow{2}{*}{Private mobility$^{(b)}$}
 & Diesel car & 49\% & 93.5\,Mpass.-km/h\\
 & Gasoline car & 49\% & 94.7\,Mpass.-km/h\\
 & HEV & 2\% & 5.9\,Mpass.-km/h\\
\midrule
\multirow{3}{*}{Public mobility}
 & Diesel bus & 43\% & 3.6\,Mpass.-km/h\\
 & Train & 43\% & 3.9\,Mpass.-km/h\\
 & CNG bus & 5\% & 0.8\,Mpass.-km/h\\
\midrule
\multirow{3}{*}{Freight mobility}
 & Diesel truck & 74\% & 62.7\,Mt.-km/h\\
 & Diesel boat & 15\% & 10.8\,Mt.-km/h\\
 & Train & 11\% & 2.5\,Mt.-km/h\\
\midrule
HVC & Naphtha/LPG cracking & 100\% & 4.6\,GW\\
Ammonia & Haber-Bosch & 100\% & 1\,GW\\
Methanol & Import & 100\% & -\\
\bottomrule							
\end{tabular}}
\end{minipage}
\end{table}

\section{\ce{CO2}-budget versus linear decrease of emissions}
\label{app:CO2_budget}

\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed for each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead. Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050. The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat. In the former, this is linked to the extended use of oil products through naphtha-cracking. The latter is produced by industrial coal boilers for longer, until 2040. Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.49\textwidth]{GWP_per_sector_REF.pdf}
\includegraphics[width=0.49\textwidth]{GWP_per_sector_lin.pdf}
\caption{Respecting the \ce{CO2}-budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
\label{fig:app_CO2_REF_lin}
\end{figure}

\section{Uncertainty characterisation for the 5-year steps transition} 
\label{app:UC_full}
Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025. Refer to \cite{Moret2017, Moret2017PhDThesis} for the methodology and sources. As the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III, based on the work of \citet{Moret2017PhDThesis}. For type III uncertainties (\ie uncertainty ranges increasing with time), a 50\% increase has been set arbitrarily between the ranges for 2025 and these same ranges for 2050. In other words, for these specific uncertainties, the ranges for 2050 are 50\% larger than for 2025.

\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits. Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}. The uncertainty characterisation gives the uncertainty ranges per parameter or group of parameters (category).

This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors. For the specific case of \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ will influence the maximum capacity (\ie 6~GW) to install to translate somehow the readiness of this technology. If it is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.

% \textcolor{red}{\textbf{See low-wind speed in 2021 from energy key data edition february 2023 (wind-based production decreased 6.4\% despite additional wind farm installations), to justify the -22\% for c,p,t}}

\begin{table}[htbp!]
\caption{Application of the uncertainty characterization method to the EnergyScope Pathway model for the year 2025. $^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg. $^{(b)}$ The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model. $^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}
\label{tab:UC_full}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l l l c c c}
\toprule
\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}$^{(a)}$}  & \multicolumn{2}{c}{\textbf{Relative variation$^{(b)}$}}\\
    & & & &	 min 	&	 max \\ 	
\midrule		
\multirow{4}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
& $c_{\mathrm{op,elec}}$ & Purchase electricity & II & -64.3\% & 179.8\% \\
& $\bm{c_{\mathrm{op,electrofuels}}}$ & \textbf{Purchase electrofuels} & \textbf{II} & \textbf{-64.3\%} & \textbf{179.8\%} \\
& $c_{\mathrm{op,biofuels}}$ & Purchase biofuels & II & -64.3\% & 179.8\% \\
\midrule
\multirow{9}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
& $c_{\mathrm{inv,bus}}$ & CAPEX bus & I & -21.6\% & 25.0\% \\
& $c_{\mathrm{inv,ic\_prop}}$ & CAPEX ICE & I & -21.6\% & 25.0\% \\
& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
& $c_{\mathrm{inv,efficiency}}$ & CAPEX efficiency measures & I & -39.3\%  & 39.3\% \\
& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
& $c_{\mathrm{inv,grid}}$ & CAPEX power grid & I & -39.3\% & 39.3\% \\
& $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}$^{(c)}$ & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
\midrule
\textbf{Maintenance cost} & $c_{\mathrm{maint,var}}$ & Variable OPEX of technologies & I & -48.2\% & 35.7\% \\
\midrule
\multirow{2}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
& $\eta_{\mathrm{fc\_prop}}$ & Consumption fuel cell vehicles & I & -28.7\% & 28.7\% \\
\midrule
\multirow{3}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
& $f_{\mathrm{max,windoff}}$ & Max capacity offshore wind & I & -24.1\% & 24.1\% \\
\midrule
\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
\midrule
\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
\midrule

\multirow{4}{*}{\textbf{End-use demand}} &$HH\_EUD$ & Households EUD & III & -13.8\% & 11.2\% \\
& $services\_EUD$ & Services EUD & III & -14.3\% & 11\% \\
& $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
\midrule

\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
& $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\
& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
& $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
& $\bm{f_{\mathrm{max,SMR}}}$ & \textbf{Potential capacity \gls{SMR}} & \textbf{-} & \textbf{0} & \textbf{1} \\

\bottomrule

\end{tabular}}
\end{table}

### 2024-01-12 chap_case_study/chap_case_study.tex
Case study + UQ ranges

@@
 For a comprehensive understanding and detailed descriptions of the technologies, please refer to the documentation \cite{readthedocs_pathway}.
+Then, the uncertainty ranges considered for some of the parameters are detailed.
 Finally, the \ce{CO2}-budget over the 2020-2050 transition is presented.
@@
 Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity independently at any hour of each representative year.
-Here, the authors simplify SMRs as only producing electricity and assume that the after-heat is lost to the atmosphere anyway.
+Here, we simplify SMRs as only producing electricity and assume that the after-heat is lost to the atmosphere anyway.
 
@@
 
+\section{Uncertainty ranges}
+\label{sec:cs:uncertainty}
+\Cref{tab:UC_short} gives the uncertainty ranges of some key parameters.
+Like other works \cite{li2019renewables,coppitters2021robust}, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_short}.
+As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040.
+Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
+As SMRs are foreseen, if installed, to be around the same locations (\ie Tihange and Doel) as the conventional nuclear power plants and using the same area in kW/ha, the same 6\,GW are assumed to be the maximum capacity for SMRs.
+If it is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about the availability.
+Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
+
+The exhaustive list of the parameters accounted in this work is presented in Appendix \ref{app:UC_full}.
+
+\begin{table}[htbp!]
+\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025.
+$^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
+$^{(b)}$ The nominal value of each parameter is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+$^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}
+\label{tab:UC_short}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{l l l c c c}
+\toprule
+\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}$^{(a)}$}  & \multicolumn{2}{c}{\textbf{Relative variation$^{(b)}$}}\\
+    & & & &	 min 	&	 max \\ 	
+\midrule		
+\multirow{2}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+& $\bm{c_{\mathrm{op,electrofuels}}}$ & \textbf{Purchase electrofuels} & \textbf{II} & \textbf{-64.3\%} & \textbf{179.8\%} \\
+\midrule
+\multirow{5}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
+& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
+& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
+& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}$^{(c)}$ & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
+\midrule
+\multirow{1}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+\midrule
+\multirow{2}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
+& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
+\midrule
+\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
+& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
+\midrule
+\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
+& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
+\midrule
+
+\multirow{2}{*}{\textbf{End-use demand}} & $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
+& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
+\midrule
+
+\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
+& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
+& $\bm{f_{\mathrm{max,SMR}}}$ & \textbf{Potential capacity \gls{SMR}} & \textbf{-} & \textbf{0} & \textbf{1} \\
+
+\bottomrule
+
+\end{tabular}}
+\end{table}
+
 \section{\ce{CO2}-budget for the transition}

### 2024-01-16 appendices/app_EnergyScope.tex
Drop publications, first part of methodo + app. Pathway choice

\section{EnergyScope Pathway: The right model to choose} 
\label{app:ESPathway_choice}

Energy system models of varying complexity are valuable tools for guiding policymakers and projecting future trends. These models enable the exploration of different energy scenarios and the assessment of their consequences based on the underlying assumptions. Specifically, techno-economic models play a crucial role in identifying technically feasible pathways for the energy transition while considering the associated economic costs. These models can be classified based on two key factors: technical resolution and simulation horizon, as illustrated in Figure \ref{fig:energy_models_classification}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{Models_classification.png}
\caption{Model can be classified by their core focus: \textbf{Operation} or \textbf{Design}. These categories can be broken down into subcategories. This paper focuses on the system planning and scenario analysis models. Inspired from \cite{palmintier2013incorporating}. }
\label{fig:energy_models_classification}
\end{figure}

Increasing the technical resolution of energy system models often comes at the expense of a shorter simulation horizon, and vice versa. For instance, day-ahead grid operation models prioritise accurate grid resolution and capacity reserves for uncertainties, but they may not incorporate long-term market trends. Different model classes cater to various needs, with decreasing technical resolution. These include machine-level control, network dispatch, unit commitment, maintenance, power plant expansion, planning for new infrastructure, and scenario analysis. Each class serves a specific purpose, from fine-grained control within a machine to the exploration of multiple assumptions across different scenarios.

% Specific need & criteria

In accordance with the previous classification, models aimed at aiding decision-makers in the energy transition primarily fall under the categories of planning and scenario analysis, with a relatively lower technical resolution. Nonetheless, ensuring technical accuracy is of paramount importance to ensure the effective functionality of future energy systems. Hence, these models should meet the following requirements as a minimum:
(i) assessment of intermittent renewable energy integration;
(ii) accounting for all energy flows in different sectors, including the measurement of greenhouse gas emissions in the energy sector;
(iii) exploration of all available options;
(iv) consideration of investments throughout the transition process; and
(v) ensuring a reasonable computation time for analysing different trajectories.
Additionally, to enhance result reproducibility and user understanding, it is advantageous for such models to:
(vi) maintain transparency and preferably be open-source, accompanied by collaborative documentation.

These requirements can be transposed into criteria that a model should match:
(i) it should have an hourly resolution spanning a one-year time horizon;
(ii) it should encompass the entire energy system, including all types of demands (such as heat, electricity, mobility, and non-energy
%\footnote{Non-energy demand is often omitted in models while it represents up to 10\%  of worldwide energy demand and up to 20\% in some countries \cite{rixhon2022integration}. \citet{eurostat2019energy} defines the non-energy demand as `\emph{energy products used as raw materials in different sectors, not consumed as a fuel or transformed into another fuel}´.}
), as well as all resources, conversion processes, and storage technologies;
(iii) it should optimise the system design, accounting for all the options;
(iv) it should have a long-term investment horizon, spanning several decades;
(v) its computational time should be reasonable, typically less than one hour on a personal laptop;
(vi) it should be open-source, with accessible data and comprehensive documentation.
These requirements are commonly found in reviews of energy system models. In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria: (i-iv) and (vi), along with others such as the number of users and market equilibrium. In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv). This review was followed by a classification where criteria (i-iv) were taken into account \cite{prina2020classification}. In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
Based on these reviews, Table \ref{tab:intro_soa} compares models based on all the previous criteria except the computational time (v). Indeed, the latter is hard to compare as models are not apply to the same case study and the information is rarely given. The table includes only the models that achieved partially at least four out of the five criteria. The authors endeavored to refresh the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.

\begin{table}[!htbp]
\caption{Comparison of existing models that partially satisfy at least four of the six criteria (in alphabetical order). Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied. Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition}.}
\label{tab:intro_soa}
\begin{minipage}{\textwidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{Ref.} & \textbf{Hourly} & \textbf{\parbox{2cm}{\centering Whole-energy}}& \textbf{\parbox{2cm}{\centering Optimis. invest. \& operation}} & \textbf{Pathway} & \textbf{\parbox{2cm}{\centering Open-source}} \\
\midrule
Calliope & \cite{Atlason2014,Calliope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{Topic is being discussed in the chat of their repository but not yet included in their documentation.} & \checkmark \\
COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted.}\footnote{\label{foot:industrynotaccounted} Industry not accounted} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{gray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
DIETER & \cite{zerrahn2017long} & \checkmark & {\color{gray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
E2M2 & \cite{E2M2} & \checkmark & {\color{gray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} individual heating not accounted.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark}\footref{foot:freeundersomespecial} \\
Ener. Trans. Model & \cite{ETM} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{The ETM is a simulation model with a simple merit order 'optimisation' for electricity, flex and heat.} & \checkmark & \checkmark \\
EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{gray} \checkmark}\footref{foot:freeware} \\
energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only, while iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & \checkmark \\
EnergyScope TD & \cite{limpens2019energyscope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:notpathwaystated} & \checkmark \\
Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{ Only for internal use.} \\
ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & \checkmark \\
GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year. \cite{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years. However, the time resolution is 16 time slices representing 4 hours per day and one day per season.} & \checkmark \\
H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \cite{prina2021bottom} classified H2RES as a simulation model on power sector only. In their work \cite{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors. Their study applied the model to a transition pathway for Croatia. In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\ 
iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \cite{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\cite{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes. Using a myopic approach, the model optimises the energy transition with a 3-hours resolution). } & \checkmark \\
RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
\bottomrule
\end{tabular}}
\end{minipage}
\end{table}

From Table \ref{tab:intro_soa}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. 
The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models. It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration. This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute in their research endeavors. Notably, TIMES was reported commercial in 2010 \cite{Connolly2010}. A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}. Recent developments by the IEA-ETSAP have resulted in a version that is compatible with open-source solver CBC. In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}. It is noteworthy that \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using 288 time slices, equivalent to a 12-day time resolution, highlighting the sensitivity of results to time resolution. Regarding data accessibility, while some publications partially present the dataset used, the overall accessibility of TIMES data remains an area of ongoing inquiry \cite{openmod_times_description}. 
 Calliope is a `\emph{tool that makes it easy to build energy system models}´ at different geographical scale. Even if the framework offers the possibility of modelling multi-year systems, the authors didn't find a relevant publication on this topic. In fact, the model is typically employed for scenario analysis with a specific focus on the electricity system. Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}. More recently, the model has been applied to analyse a scenario of a multi-energy district in Switzerland \cite{pickering2021quantifying}. Moreover, the model has been used with decades of weather data. However, its application has been limited to assessing the impact of inter-year variability in wind and PV on the results, rather than evaluating a transition pathway \cite{pfenninger2017dealing}. 
Similarly GENeSYS-MOD presents some limitations. This model is an application of the open-source energy modelling system (OSeMOSYS), itself represented as a model with a poor time discretisation and a heavy computational burden according to 
\cite{prina2019transition}. \citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors. The time disaggregation can be chosen by the user, for their application they used representative years with three days and two time slice per day. 
%\cite{bartholdsen2019pathways} performed a similar study for the case of Germany with 16 time slices per representative years and motivated this choice as sufficient as \cite{welsch2014incorporating} demonstrates that it gives consistent results compared to using hourly time resolution over a year.  However, with limited number of time slices, storage technologies, especially the ones for inter-month storage, cannot be optimised and thus simplifies the non-yet solved challenge of intermittent renewable energy integration. 
%PyPSA, one of the most promising model thanks to its strong and active community, has also been used to perform a myopic transition \cite{pedersen2022long}.  
Among the open-source models with an active community, PyPSA is one of the best-performing, with a large and active community, development at the state of the art, worldwide applications, and usage not only limited to academia. A study conducted by \citet{bartholdsen2019pathways} centered on Germany employed a representation comprising 16 time slices for representative years. This choice was substantiated by the work of \citet{welsch2014incorporating}, which demonstrated that this level of temporal granularity yields consistent results in comparison to hourly time resolution over a year. However, it is noteworthy that the utilization of a limited number of time slices may simplify the optimization of storage technologies, especially those designed for inter-month energy storage. This simplification can be viewed as a pragmatic approach to reduce the computational burden while over-simplifying the challenge of accurately integrating intermittent renewable energy sources. Furthermore, PyPSA, a modeling framework recognized for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}. 
Hence, it is worth noting that while Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, the authors have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices.

% SOA

Hence, it appears that none of the models of Table \ref{tab:intro_soa} fully meet all five criteria outlined in the table, topped with the additional consideration of acceptable computational time. This observation is consistent with the findings presented in \cite{prina2019transition} who identified two approaches for optimising the energy transition pathway based on the six criteria. The first approach involves running a snapshot model multiple times using an algorithm that optimises the transition path and validates the system's operability. The second approach aims to extend a snapshot model to represent the entire transition pathway. However, they excluded this option due to the lack of models that met the requirements of being fast enough and easily adaptable. 
Therefore, they developed a new model based on the first methodology, named EPLANoptTP. It uses a multi-objective evolutionary algorithm to optimise the EnergyPLAN model \cite{lundenergyplan}. To manage computational time, the number of decision variables is limited to three: \gls{PV}, wind turbine and battery capacities. Thus, the model does not investigate all the options (\ie criteria (iii)).

### 2024-01-16 appendices/app_case_study.tex
Drop publications, first part of methodo + app. Pathway choice

@@
 \begin{table}[htbp!]
-\caption{Major technologies used to supply the 2020-demands of \Cref{fig:cs_demands} in terms of share of production and installed capacity.
-$^{(a)}$ The decentralised heating units provide 98\% of the low-temperature heat demand.
-$^{(b)}$ The private mobility accounts for 80\% of the passengers mobility.}
+\caption{Major technologies used to supply the 2020-demands of \Cref{fig:cs_demands} in terms of share of production and installed capacity.}
 \label{tab:Belgium_2020}
@@
 \midrule
-\multirow{3}{*}{Heat Low-Temp. (DEC)$^{(a)}$}
+\multirow{3}{*}{Heat Low-Temp. (DEC)\footnote{\label{foot:DHN_98}The decentralised heating units provide 98\% of the low-temperature heat demand. }}
  & Oil boiler & 48\% & 21.4\,GW\\
@@
 \midrule
-\multirow{2}{*}{Private mobility$^{(b)}$}
+\multirow{2}{*}{Private mobility\footnote{\label{foot:privatemob_80}The private mobility accounts for 80\% of the passengers mobility.}}
  & Diesel car & 49\% & 93.5\,Mpass.-km/h\\
@@
 
+%$^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
+$^{(b)}$ The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+$^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.
+
 \begin{table}[htbp!]
-\caption{Application of the uncertainty characterization method to the EnergyScope Pathway model for the year 2025.
-$^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
-$^{(b)}$ The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
-$^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}
+\caption{Application of the uncertainty characterization method to the EnergyScope Pathway model for the year 2025. }
 \label{tab:UC_full}
+\begin{minipage}{\linewidth}
 \centering
@@
 \toprule
-\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}$^{(a)}$}  & \multicolumn{2}{c}{\textbf{Relative variation$^{(b)}$}}\\
+\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
+}}  & \multicolumn{2}{c}{\textbf{Relative variation}\footnote{\label{foot:nom_val_uncert}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+}}\\
     & & & &	 min 	&	 max \\ 	
@@
 & $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
-& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}$^{(c)}$ & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
+& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
 \midrule
@@
 \end{tabular}}
+\end{minipage}
 \end{table}

### 2024-01-16 chap_methodo/chap_methodo.tex
Drop publications, first part of methodo + app. Pathway choice

%\begin{flushright}
%%%\emph{``For the things we have to learn before we can do them, we learn by doing them.''}\\%Each wrong scenario you have analysed, is a wrong decision you won't made.''}\\
%%%Aristote
%\emph{All models are wrong, but some are useful.}\\
%George P. Box
%\end{flushright}

%\medskip
%
%\begin{mybox}{Chapter overview}
%\begin{itemize}[left=0em]%[leftmargin=0cm,itemindent=.5cm,labelwidth=\itemindent,labelsep=0cm,align=left]
%\setlength\itemsep{-0.3em}
%\item Belgian energy system overview%Case studies: the Belgian energy systems during the transition (2015-2050).
%\end{itemize}
%\vspace{-0.3cm}
%
%\emph{This chapter is an improved and extended case study version of \citet{Limpens_belgian_2020}}.
%\end{mybox}
%
%\medskip

\section{Contributions}
\label{sec:meth:contributions}
\begin{itemize}
\item Apply Stefano's method on the pathway model with a similar approach as Guevara et al.
\item Check that PCE was appropriate as a method for such a system (ECOS2020)
\end{itemize}

\section*{Other authors' main contribution statement}
On top of the main contributions of this thesis that are aforementioned, three main authors are to be mentioned for having brought a significant part of the methodological work. Based on Stefano Moret's monthly whole-energy system model (\ie EnergyScope) \cite{moret2016strategic}, Gauthier Limpens has developed the hourly version of the snapshot model (\ie EnergyScope TD) \cite{limpens2019energyscope}, as well as the perfect foresight pathway model \cite{limpens2024pathway}, to which I personally contributed too. Diederik Coppitters has developed the RHEIA framework allowing to quantify the impact of uncertainties and carry out robust optimisation of energy systems \cite{coppittersthesis}. The current work used this framework for the first of these functionalities. Finally, Stefano Moret extensively assessed the uncertainty characterisation on the Swiss energy system \cite{Moret2017}. This thesis follows the same methodology, updating the uncertainty ranges for the pathway model.

\section{Whole-energy system transition model optimisation: EnergyScope Pathway}
\label{sec:meth:ES}

On the contrary, this work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}. According to pathway models review, EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is open-source documented model. Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation). Presenting here only the main constraints of the model and the main input data in Section \ref{chap:case_study}, the reader is invited to refer to the extended paper \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway} for more extensive information.

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{ES_Pathway.pdf}
\caption{Illustration of the pathway methodology based on an existing energy system model. The methodology spans from 2020 to 2050, with one representative year every five years. The model \acrfull{ESTD} is applied in 7 representative years (light blue boxes). The formulation includes additional constraints (black boxes) that link the years together. The pathway's initialisation assumes that all capacities installed in 2020 were built during the pseudo-phase of 2015-2020 (grey box). The overall problem is defined as the pathway model. Graph adapted from \cite{limpens2024pathway}.}
\label{fig:meth_path_methodology}
\end{figure}
%
%We used the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology})---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050. Even though this lacks realism compared to the myopic approach \cite{babrowski2014reducing,fais2016impact,heuberger2018impact} where the pathway is optimised via a sequence of shorter time windows, the perfect foresight approach allows to optimise the transition with a \ce{CO2}-budget (Section \ref{subsec:cs:CO2-budget}) rather than along a prescribed \ce{CO2}-trajectory. However, the work of \citet{limpens2024pathway} has shown that PF provides similar results for the study case compared to the myopic approach with a prescribed linear decrease of the \ce{CO2}-emissions . All the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}, are kept as is with an extra-dimension to relate them to a specific representative year, $y$, of the pathway. For instance, the energy balance is guaranteed at every hour of each of these years. 
%
%The optimised objective of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as follows: 
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign} 
%% Objective function + investment 
%\label{eq:obj_func_v2}%1
%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
%\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
%\label{eq:CAPEX_v2}
%& \textbf{C\textsubscript{tot,capex}} =
%\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
%\textbf{C\textsubscript{inv,phase}}(p)
%-
%\sum_{\mathclap{j \in \emph{TECH}}} 
%\textbf{C\textsubscript{inv,return}}(j)\\
%  \label{eq:Copex_tot_v2}%5
%& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
%+ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}} 
% \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
%\label{eq:path_annu_factor}
%& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
%%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
%\end{flalign}
%\endgroup
%
%\noindent
%where $\emph{t\textsubscript{phase}}=5\,\text{years}$ and $\emph{diff\_2015\_year(p)}$ are respectively the duration of a phase between two representative years and the number of years between the middle of a phase and 2015 for a correct annualisation. $\textbf{C\textsubscript{inv,return}}$ accounts for the residual value, also called \textit{salvage value}, of the technologies installed during the transition and having not reached the end of their lifetime by 2050. This last variable is crucial to avoid penalising heavy (and potentially long-lifetime) investments at the end of the transition as these assets would still be operational beyond 2050. The interested reader will find more information about its implementation and the formulation choices related to it in the work of \citet{limpens2024pathway}. The other variables in Eq.\,(\ref{eq:CAPEX_v2}-\ref{eq:Copex_tot_v2}) are detailed here below:
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign} 
%\label{eq:opex_yearly}
%&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i)&\forall y\in \emph{YEARS}\\
%\label{eq:PhaseInv}%5
%&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
%\end{flalign}
%\begin{flalign}
%&&\forall p \in \emph{PHASE} | y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)
%\end{flalign}
%\endgroup
%
%\noindent
%where $\textbf{F\textsubscript{new}}$ are the capacities newly installed. In Eq. (\ref{eq:opex_yearly}-\ref{eq:PhaseInv}), the costs related to each representative year are:
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign} 
%% Objective function + investment 
%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
%\hspace{0pt} 
% \label{eq:c_inv}%3
% &\textbf{C\textsubscript{inv}}(y,j) = c_{\text{\emph{inv}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\
% \label{eq:c_maint}%4
% &\textbf{C\textsubscript{maint}}(y,j) = c_{\text{\emph{maint}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\ 
%  \label{eq:c_op}%5
% &\textbf{C\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t) t_{op} (t)  
% & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
% \end{flalign}
% \endgroup
%
%\noindent where the variable $\textbf{F}$ represents the size of the installed capacities (for all technologies $j$) and the variable $\textbf{F\textsubscript{t}}$ is the hourly consumption of the resources; the parameters $c_{\text{\emph{inv}}}$ and $c_{\text{\emph{maint}}}$ are the CAPEX and the OPEX of the technologies, and the parameter $c_{\text{\emph{op}}}$ is the cost of purchasing resources. For the sake of simplicity, as done by \citet{limpens2024pathway}, the sum over the 8760 hours of the year is written as the sum over $t \in T $. 
%
%Then, as detailed in section \ref{subsec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign} 
%\label{eq:gwp_tot_transition}
%&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}\left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)/2 &
%\\
%\label{eq:limit_gwp_trans}
%& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
%\end{flalign}
%\endgroup
%
%\noindent
%where the computation of the yearly emissions are based on the \acrfull{GWP} of the resources:
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign}
%\hspace{0pt}
% \label{eq:GWP_tot}%8
% & \textbf{GWP\textsubscript{tot}}(y)  =    \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{GWP\textsubscript{op}}(y,i) 
% & \forall y \in \text{\emph{YEARS}}\\
%  \label{eq:GWP_op}%7
% & \textbf{GWP\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t)  t_{op} (t) & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
%\end{flalign}
%\endgroup
%
%\noindent
%where $gwp_{\text{\emph{op}}}$ is the specific emissions (\ie in kt$_{\ce{CO2},\text{eq}}$/GWh) of each resource. Based on an approach developed by the Intergovernmental Panel on Climate Change (IPCC) \cite{stocker2014climate}, this work considers the indicator ``GWP100a - IPCC2013'' to compute the emissions related to the use of resources. This includes the emissions due to the extraction, the transportation and the combustion of the energy carrier. EnergyScope proposes to account for the embodied emissions of the technologies based on a \gls{LCA}. These stand for extraction of materials, refining, construction and end of life \cite{schnidrig2023integration}. However, this work is still in progress and the database is not yet complete. Consequently, it is not included in this work and not accounted for.
%
%Besides this constraint on the emissions, the main constraint to link years with each other is the one dictating the installed capacities at the end of each year:
%
%\begingroup
%\belowdisplayskip=2pt
%\abovedisplayskip=2pt
%\begin{flalign} 
%\label{eq:F_newBuilt}%5
%&\textbf{F}(y\textsubscript{stop},j) = \textbf{F}(y\textsubscript{start},j)
% + \textbf{F\textsubscript{new}}(p,j)
% - \textbf{F\textsubscript{old}}(p,j)
% - \sum_{\mathclap{p2 \in \text{\emph{PHASE}} \cup \{2015\_2020\}}} \textbf{F\textsubscript{decom}}(p,p2,j)& \notag \nonumber 
% \end{flalign}
%\begin{flalign} 
% &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), j \in \text{\emph{TECH}}
% \end{flalign}
%\endgroup
%
%\noindent
%where the variables $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$ are the capacities respectively having reached the end of their lifetime and prematurely decommissioned. Moreover, to account for the society inertia and to prevent unrealistically fast modal share change, constraints limit this change for the sectors of the low-temperature, the passenger mobility and freight mobility demands. The parameters $\Delta_{\mathrm{change,LT\_heat}}$, $\Delta_{\mathrm{change,pass}}$ and $\Delta_{\mathrm{change,freight}}$ respectively limit their respective modal share change up to 33\%, 50\% and 50\% per phase of 5 years.

\section{Uncertainty quantification}
\label{sec:meth:UQ}

\subsection{Uncertainty characterisation}
\label{subsec:uncert_charac}
%Presentation here of the uncertainty characterisation from S. Moret and adding some parameters specific to the pathway model (\eg $\Delta_{\text{change}}$ linked to change speed) or specific to the main case study (\eg possibility to have nuclear SMR from 2040)

\subsection{Polynomial Chaos Expansion}
\label{subsec:pce}

We used \gls{PCE}, an approach for surrogate-assisted \gls{UQ}, to propagate uncertainties in input parameters through the system model. This allowed us to assess statistical moments on the quantity of interest and determine Sobol' indices~\cite{coppitters2020robust}. To construct a PCE of the EnergyScope Pathway model, we employed the open-source Python framework RHEIA~\cite{coppitters2022rheia,readthedocs_rheia}. Where the first part of this section is dedicated to the mathematical definition of this approach, the second details its choice and summarises the comparison made with another approach (\ie Morris method) in a previous work \cite{limpens2020impact}.\\

\vspace{0.1cm}\textbf{Definition}\vspace{-0.3cm}\\

The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie results) in the EnergyScope Pathway model ($M$). This representation is constructed as a truncated series of multivariate orthonormal polynomials $\bm{\Psi}$, weighted by coefficients $u$:

\begin{equation}
\hat{M} \left( \bm{\xi} \right) = \sum_{\bm{\alpha} \in \mathcal{A}^{d,p}} u_{\bm{\alpha}} \bm{\Psi}_{\bm{\alpha}} \left( \bm{\xi} \right) \approx M \left( \bm{\xi} \right), 
\end{equation}

where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (\autoref{app:UC_full}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$. As uniform distributions are considered, the Legendre polynomials are adopted, as they are the associated family of polynomials that are orthogonal with respect to standard uniform distributions~\cite{Sudret2014}.

A truncation scheme is implemented to restrict the number of multivariate polynomials in the series. This is done based on two factors: a specified limiting polynomial order ($p$) and the number of uncertain parameters ($d$) involved. The multivariate polynomial order $|\bm{\alpha}|$ is the summation of the orders for each univariate polynomial in the multivariate polynomials space. Thus, only the multi-indices corresponding to an order that is less than or equal to the specified limiting order are retained and stored in the truncated series denoted as $\mathcal{A}^{d,p}$:

\begin{equation}
\mathcal{A}^{d,p} = \left \{ \bm{\alpha} \in \mathbb{N}^d : |\bm{\alpha}| \leq p \right \}. 
\end{equation}

The number of multi-indices satisfying this condition is as the cardinality of $\mathcal{A}$, \ie the number of its elements:
\begin{equation}
\mathrm{card} \left( \mathcal{A}^{d,p} \right) = {p + d \choose p} = \dfrac{\left( d + p \right) !}{d! p!} = P + 1.
\label{eq:pce:nterms}
\end{equation}

The coefficients ($u_0, u_1, \dots, u_{P+1}$) are quantified using a regression method applied to orthonormal polynomials~\cite{Sudret2014}. To ensure a well-posed least-square minimisation, it is recommended to have a number of training samples at least twice the number of coefficients~\cite{Sudret2014}. Therefore, $2 \left( P+1 \right)$ samples are evaluated in the system model, and the model response for each quantity of interest is recorded. To generate the training samples, the quasi-random Sobol' sampling technique is employed~\cite{bratley2003implementing}. As a low-discrepancy sequence, this technique exhibits the main advantage to investigate efficiently and (almost) uniformly the hypercube of uncertainties, unlike uniformly distributed random numbers.

The process of defining the polynomial degree includes incrementally increasing it until a desired level of accuracy is achieved~\cite{coppitters2022rheia}. Starting with $p=1$, a PCE is constructed and the \gls{LOO} error is evaluated. If the \gls{LOO} error is below a specified threshold, the corresponding polynomial order is considered sufficient for generating an accurate PCE. However, if the error exceeds the threshold, the order is increased, and additional samples are generated following the rule of Eq.\,(\ref{eq:pce:nterms}).

For the specific study of this work, a polynomial order of 2 is necessary (with 1260 training samples as per Eq.\,(\ref{eq:pce:nterms})) to achieve a \gls{LOO} error below \SI{1}{\%} for the total transition cost.

Lastly, the statistical moments can be analytically derived from the PCE coefficients, eliminating the need for further model evaluations. The mean $\mu$ and standard deviation $\sigma$ are obtained as follows:
\begin{align}
\mu &= u_0,\\
\sigma^2 &= \sum_{i \neq 0 } u_{i}^2 .
\label{eq:pce:statmom}
\end{align}

Furthermore, the Sobol' indices can also be determined analytically. The total-order Sobol' indices ($S_i^{T}$) assess the overall influence of a stochastic input parameter on the performance indicator, encompassing all possible interactions:

\begin{equation}
S_i^{T} = \sum_{\bm{\alpha} \in A_i^T}^{} u_{\bm{\alpha}}^2/\sum_{i=1}^P u_i^2 ~~~~~~ A_i^T = \{\bm{\alpha} \in A | \alpha_i > 0\}.
\end{equation}

Here, $A$ denotes the collection of all PCE coefficients, and $\alpha_i$ corresponds to the coefficient associated with the uncertain parameter $i$.\\

\vspace{0.1cm}\textbf{Comparison with a proven method}\vspace{-0.3cm}\\

Besides being an in-house used method, an early step of this thesis consisted in assessing \gls{PCE} with similar approach used in the literature \cite{limpens2020impact}. \\

After characterising the uncertainty ranges, \citet{Moret2017} quantified the impact of these uncertainties on the snapshot model of EnergyScope,  \ie ranking them, using the Morris method \cite{morris_factorial_1991}.  This method, as a statistical analysis, relies on individually randomized one-factor-at-a-time designs. Given the $d$ model parameters $\vec{\xi}=(\xi_1,\xi_2,...,\xi_d)$, the first step of the method consists in generating independent random samples of $\vec{\xi}$ in a standardised and discretised $p$-level \textit{region of experimentation}, $\omega$. In this \textit{region of experimentation}, each $\xi_i$, varying in the interval $[\xi_{i,min},~\xi_{i,max}]$, can take a random discrete value as follows :
\begin{equation}
  \xi_{i}=\xi_{i,min}+j\cdot\frac{1}{p-1}\left(\xi_{i,max}-\xi_{i,min}\right)~~\text{with }j\in\{0,1,...,p-1\}
\end{equation}

Then, given these random one-factor-at-a-time samples, Morris method defines, for a given set of $\vec{\xi}$, the elementary effect of the $i$th parameter ($EE_i$) as :
\begin{equation}
  EE_{i}=\frac{M(\xi_1,\xi_2,...,\xi_i+\Delta,...,\xi_d)-M(\vec{\xi})}{\Delta}
\end{equation}
where $M$ is the objective function, $\vec{\xi}\in\omega$, except $\xi_i\leq1-\Delta$ and $\Delta$ is a set multiple of $1/(p-1)\left(\xi_{i,max}-\xi_{i,min}\right)$. As in other studies \cite{Sin2009,Moret2017,Moret2017PhDThesis}, we consider $p$ as even and $\Delta=p/[2(p-1)]\left(\xi_{i,max}-\xi_{i,min}\right)$.\\

Finally, in order to evaluate the importance of the $i$th parameter over an output, Morris method relies on $F_i$, the distribution of $r$ elementary effects. Computing the mean, $\mu_{i}=\mu(F_{i})$, and the standard deviation, $\sigma_{i}=\sigma(F_{i})$, of the $F_{i}$ distribution, allows ranking the parameters based on their influence on the concerned output. Usually, in Morris method, $p$ and $r$ respectively get values as follows : $p\in \{4,6,8\}$ and $r\in [15;100]$ depending on, $d$, the number of uncertain parameters. The higher this number is,  the higher shall be, simultaneously, $p$ and $r$. In the following comparative analysis, we set $p$ and $r$ to their maximum values, respectively $8$ and $100$ in order to get the most reliable parameters ranking.\\

Beyond the original Morris method, we used the standardized elementary effects, $SEE_{i}$, formulation \cite{Sin2009}, given by
\begin{equation}
    SEE_{i}=EE_{i}\cdot\frac{\sigma(\xi_i)}{\sigma(M)}.
\end{equation}
Among other things, the $SEE$ allows comparing the influence of different inputs on the same output or compare the influence of a same parameter on different outputs, even if these parameters or outputs are significantly different in terms of variation range or average amplitude. Moreover, this standardized analysis does not require any additional model evaluations.\\
Therefore, in the following results, we rather use
\begin{equation}
\mu^*_{i}=\mu(\vert SF_{i}\vert)
\label{eq:mu*}
\end{equation}

to rank parameters among each other. In (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.\\

In \cite{limpens2020impact}, we have assessed the \gls{PCE} approach, comparing the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$. Even if the output of each method does not have the same physical meaning, both methods can rank the parameters by their impact on the total annual  cost  of  the  energy  system. Both rankings were very similar which validates the use of \gls{PCE} in the rest of this work.

\subsection{Preliminary screening and selection}
\label{subsec:screening}
%Similarly to what we did in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), emphasise that we only consider a limited amount of uncertain parameters to keep a reasonable computation time while capturing the impact of (almost) all the uncertainties.

\section{Reinforcement learning}
\label{sec:meth:RL}

\section{Principal Components Analysis}
\label{sec:meth:PCA}

### 2024-01-16 frontend/publications.tex
Drop publications, first part of methodo + app. Pathway choice

%\begingroup
%\let\cleardoublepage\clearpage

% This ensures that only the main file is compiled
% !TEX root = ../my_thesis.tex

% English abstract
\chapter*{List of publications}
\markboth{List of publications}{List of publications}
%\addcontentsline{toc}{chapter}{List of publications} % adds an entry to the table of contents
\vskip0.5cm
% put your text here

%First part of Chapter \ref{ch:es_literature} is built on an opinion published in a journal:

\begin{quote}
Limpens, G., \textbf{Rixhon, X.}, Contino, F., \& Jeanmart, H. (2024). ``\emph{EnergyScope Pathway: An open-source model to optimise the energy transition pathways of a regional whole-energy system.}'' In Applied Energy, (Vol. 358). URL: \url{https://doi.org/10.1016/j.apenergy.2023.122501}
\end{quote}

\begin{quote}
\textbf{Rixhon, X.}, Limpens, G., Coppitters, D., Jeanmart, H.,\& Contino, F.(2022). ``\emph{The role of electrofuels under uncertainties for the Belgian energy transition.}'' In Energies (Vol. 14). URL: \url{https://doi.org/10.3390/en14134027}
\end{quote}

\begin{quote}
\textbf{Rixhon, X.}, Limpens, G., Contino, F., \& Jeanmart, H. (2021). ``\emph{Taxonomy of the fuels in a whole-energy system.}'' In Frontiers in Energy Research, Sec. Sustainable Energy Systems, (Vol. 9). URL: \url{https://doi.org/10.3389/fenrg.2021.660073} 
\end{quote}

\begin{quote}
\textbf{Rixhon, X.}, Tonelli, D., Colla, M., Verleysen, K., Limpens, G., Jeanmart, H. ,\& Contino, F.(2022). ``\emph{Integration of non-energy among the end-use demands of bottom-up whole-energy system models.}'' In Frontiers in Energy Research, Sec. Process and Energy Systems Engineering, (Vol. 10). URL: \url{https://doi.org/10.3389/fenrg.2022.904777}
\end{quote}

\begin{quote}
\textbf{Rixhon, X.}, Colla, M., Tonelli, D., Verleysen, K.,  Limpens, G., Jeanmart, H., \& Contino, F.(2021). ``\emph{Comprehensive integration of the non-energy demand within a whole-energy system: Towards a defossilisation of the chemical industry in Belgium.}'' In proceedings of  ECOS 2021 conference (Vol. 34, p. 154).
\end{quote}

\begin{quote}
Limpens, G., Coppitters, D., \textbf{Rixhon, X.}, Contino, F., \& Jeanmart, H. (2020). ``\emph{The impact of uncertainties on the Belgian energy system: application of the Polynomial Chaos Expansion to the EnergyScope model.}'' In proceedings of  ECOS 2020 conference (Vol. 33, p. 711).
\end{quote}

### 2024-01-18 appendices/app_EnergyScope.tex
Drop full implementation of Pathway from the paper

@@
 (i) it should have an hourly resolution spanning a one-year time horizon;
-(ii) it should encompass the entire energy system, including all types of demands (such as heat, electricity, mobility, and non-energy
+(ii) it should encompass the entire energy system, including all types of demands (such as heat, electricity, mobility, and non-energy)
 %\footnote{Non-energy demand is often omitted in models while it represents up to 10\%  of worldwide energy demand and up to 20\% in some countries \cite{rixhon2022integration}.
 \citet{eurostat2019energy} defines the non-energy demand as `\emph{energy products used as raw materials in different sectors, not consumed as a fuel or transformed into another fuel}´.}
-), as well as all resources, conversion processes, and storage technologies;
+, as well as all resources, conversion processes, and storage technologies;
 (iii) it should optimise the system design, accounting for all the options;
@@
 In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
-Based on these reviews, Table \ref{tab:intro_soa} compares models based on all the previous criteria except the computational time (v).
+Based on these reviews, Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis} compare models based on all the previous criteria except the computational time (v).
 Indeed, the latter is hard to compare as models are not apply to the same case study and the information is rarely given.
 The table includes only the models that achieved partially at least four out of the five criteria.
-The authors endeavored to refresh the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.
+We endeavored to refresh the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.
 
@@
 Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
-Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition}.}
+Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition} (part A)}
 \label{tab:intro_soa}
@@
 EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{gray} \checkmark}\footref{foot:freeware} \\
-energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only, while iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & \checkmark \\
+energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only.} & \checkmark & \checkmark \\
 EnergyScope TD & \cite{limpens2019energyscope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:notpathwaystated} & \checkmark \\
-Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{ Only for internal use.} \\
-ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & \checkmark \\
-GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
-\cite{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years.
+Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{Only for internal use.} \\
+%ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & \checkmark \\
+%GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
+\citet{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years.
 However, the time resolution is 16 time slices representing 4 hours per day and one day per season.} & \checkmark \\
-H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \cite{prina2021bottom} classified H2RES as a simulation model on power sector only.
-In their work \cite{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors.
+%H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
+In their work, \citet{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors.
 Their study applied the model to a transition pathway for Croatia.
 In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
-iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
-IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
-OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
-Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \cite{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
-PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\cite{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+%iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+%IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
+%OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
+%Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
+%PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
 Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
 } & \checkmark \\
-RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
-ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+%RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
+%ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+%TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
+\bottomrule
+\end{tabular}}
+\end{minipage}
+\end{table}
+
+\begin{table}[!htbp]
+\caption{Comparison of existing models that partially satisfy at least four of the six criteria (in alphabetical order).
+Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
+Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition} (part B)}
+\label{tab:intro_soa_bis}
+\begin{minipage}{\textwidth}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{lcccccc}
+\toprule
+\textbf{Model} & \textbf{Ref.} & \textbf{Hourly} & \textbf{\parbox{2cm}{\centering Whole-energy}}& \textbf{\parbox{2cm}{\centering Optimis.
+invest.
+\& operation}} & \textbf{Pathway} & \textbf{\parbox{2cm}{\centering Open-source}} \\
+\midrule
+%Calliope & \cite{Atlason2014,Calliope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{Topic is being discussed in the chat of their repository but not yet included in their documentation.} & \checkmark \\
+%COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
+%DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted.}\footnote{\label{foot:industrynotaccounted} Industry not accounted} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{gray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
+%DIETER & \cite{zerrahn2017long} & \checkmark & {\color{gray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
+%E2M2 & \cite{E2M2} & \checkmark & {\color{gray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} individual heating not accounted.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
+%EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark}\footref{foot:freeundersomespecial} \\
+%Ener.
+Trans.
+Model & \cite{ETM} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{The ETM is a simulation model with a simple merit order 'optimisation' for electricity, flex and heat.} & \checkmark & \checkmark \\
+%EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{gray} \checkmark}\footref{foot:freeware} \\
+%energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only, while iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & \checkmark \\
+%EnergyScope TD & \cite{limpens2019energyscope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:notpathwaystated} & \checkmark \\
+%Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{ Only for internal use.} \\
+ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:transportnotsaid_bis} Transport not accounted.}\footnote{\label{foot:industrynotaccounted_bis} Industry not accounted}\footnote{\label{foot:dhnnotaccounted_bis} \gls{DHN} not accounted.}\footnote{\label{foot:decLTHnotaccounted_bis} individual heating not accounted.} & \checkmark & \checkmark & \checkmark \\
+GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
+\citet{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years.
+However, the time resolution is 16 time slices representing 4 hours per day and one day per season.} & \checkmark \\
+H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
+In their work, \citet{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors.
+Their study applied the model to a transition pathway for Croatia.
+In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
+iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & {\color{gray} \checkmark} \footnote{\label{foot:invopt_bis} iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:freeundersomespecial_bis}`\emph{Free under some special conditions}´.} \\
+IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced_bis} Commercially (paid) licensed.} \\
+OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation_bis} Simulation model.} & \checkmark & \checkmark \\
+Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced_bis}  \\
+PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
+} & \checkmark \\
+RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark  & \checkmark & \checkmark \\
+ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial_bis} \\
 TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
@@
 
-From Table \ref{tab:intro_soa}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. 
+From Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. 
 The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models.
@@
 It is noteworthy that \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using 288 time slices, equivalent to a 12-day time resolution, highlighting the sensitivity of results to time resolution.
-Regarding data accessibility, while some publications partially present the dataset used, the overall accessibility of TIMES data remains an area of ongoing inquiry \cite{openmod_times_description}.
+Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data remains an area of ongoing inquiry \cite{openmod_times_description}.
  Calliope is a `\emph{tool that makes it easy to build energy system models}´ at different geographical scale.
-Even if the framework offers the possibility of modelling multi-year systems, the authors didn't find a relevant publication on this topic.
+Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic.
 In fact, the model is typically employed for scenario analysis with a specific focus on the electricity system.
@@
 \cite{prina2019transition}.
-\citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors.
-The time disaggregation can be chosen by the user, for their application they used representative years with three days and two time slice per day.
+\citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors, the time disaggregation being chosen by the user.
+For their application they used representative years with three days and two time slice per day.
 %\cite{bartholdsen2019pathways} performed a similar study for the case of Germany with 16 time slices per representative years and motivated this choice as sufficient as \cite{welsch2014incorporating} demonstrates that it gives consistent results compared to using hourly time resolution over a year.
@@
 Furthermore, PyPSA, a modeling framework recognized for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}.
-Hence, it is worth noting that while Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, the authors have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices.
+Hence, it is worth noting that while Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, we have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices.
 
@@
 
-Hence, it appears that none of the models of Table \ref{tab:intro_soa} fully meet all five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
-This observation is consistent with the findings presented in \cite{prina2019transition} who identified two approaches for optimising the energy transition pathway based on the six criteria.
-The first approach involves running a snapshot model multiple times using an algorithm that optimises the transition path and validates the system's operability.
+Hence, it appears that none of the models of Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis} fully meet all five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
+This observation is consistent with the findings presented by \citet{prina2019transition} who identified two approaches for optimising the energy transition pathway based on the six criteria.
+The first approach involves running a snapshot model multiple times using an algorithm that optimises the transition path and validates the operability of the system.
 The second approach aims to extend a snapshot model to represent the entire transition pathway.
@@
 Thus, the model does not investigate all the options (\ie criteria (iii)).
+
+For the aforementioned reasons, the current work opted for EnergyScope Pathway, an extension of the open-source and documented EnergyScope TD model \cite{limpens2019energyscope} listed in Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis}.
+The latter has an horizon time of one year and does not account for the pathway from an existing energy system to a long-term target.
+The pathway version extends the time horizon to decades and accounts for the pathway transition from an existing energy system to a long term target.
+The computational time is kept low (\ie around a 15 minutes on a personal laptop), mostly due to keeping the linear formulation after extending the snapshot model.
+\citet{limpens2024pathway} provides more detailed insights into the modeling choices made during methodological development.
+In the spirit of the EnergyScope project, the code is fully open-source (under the License Apache 2.0, see repo \cite{PESTD_v1_repo}) with a collaborative documentation \cite{readthedocs_pathway}.
+Compared to existing models, EnergyScope Pathway introduces a rapid computational optimisation tool for exploring diverse transition pathways within an entire energy system while maintaining high temporal precision to accurately capture the integration of intermittent renewables.
+To the best of our knowledge, there are potentially frameworks that could be extended to similar capabilities, but their computational time for similar case study have not been found.
+
+\section{EnergyScope Pathway and its linear formulation} 
+\label{app:ESPathway_full_formulation}
+
+EnergyScope Pathway is the extension of EnergyScope TD \cite{limpens2019energyscope} that follows the snapshot approach \cite{Girones2015}, consisting in optimising the system in a target future year.
+The objective of this section is to present the fundamental variables and constraints of the latter based on which the former was developed.
+There have been formulation choices to be made but they are not discussed in the current manuscript.
+However, the interested reader is invited to refer to the Appendix B of \cite{limpens2024pathway} for further information in this regard.
+
+\subsection{The starting point: a scenario analysis model}
+
+\vspace{0.2cm}\textbf{Overview of the snapshot model}\vspace{-0.3cm}\\
+
+EnergyScope TD \cite{limpens2019energyscope} is a model that optimises both the investment and operating strategy of a '\emph{whole}'-energy system, encompassing electricity, heating, mobility, and non-energy sectors.
+According to \citet{contino2020whole}, a model qualifies as a 'whole-energy' system when it considers all energy sectors, including the non-energy demand such as the production of plastics and other materials using feedstocks that are also considered as energy carriers, with the same level of detail.
+
+The model's hourly resolution over a year makes it well-suited for integrating intermittent renewables.
+Its formulation incorporates typical days and a reconstruction method that captures different time scales from the hour to the season while accounting for the inter-weeks patterns of wind.
+This approach minimally affects the design while significantly reducing computational time \cite{thiran2023validation}.
+The model investigates all the possibilities by optimising the investment decisions and hourly operations over a year, with a computational time of less than a minute on a personal laptop.
+This characteristic was intentionally incorporated into the model design to facilitate uncertainty quantification and other studies that require numerous iterations \cite{rixhon2021role}.
+
+EnergyScope TD has been successfully applied to various national energy systems, including Switzerland \cite{Limpens_role_2019,limpens2019energyscope}, Belgium \cite{Limpens2020}, Italy \cite{borasio2022deep}, and other European countries \cite{dommisse2020modelling}.
+Furthermore, it has been extended to a multi-region energy system model \cite{thiran2023validation}, coupled with other energy models \cite{pavivcevic2022bidirectionnal}, or employed to focus on specific sectors such as the power networks of electricity, gas, and hydrogen \cite{schnidrig2023role}.
+
+\vspace{0.2cm}\textbf{Formulation of the snapshot model}\vspace{-0.3cm}\\
+
+The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost or greenhouse gas emissions of the energy system.
+Typically, the two objectives are integrated by placing a limit on emissions while simultaneously striving to minimize costs.
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{chp_estd_overview.pdf}
+\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimizing model (Optimisation) and results (Energy strategy). The image illustrates what is included (non-exhaustively).}
+\label{fig:estd_overview}
+\end{figure}
+
+\vspace{0.2cm}\textbf{Linear formulation}\vspace{-0.3cm}\\
+
+The following section illustrates, non-exhaustively, the original EnergyScope TD model.
+The objective function, cost and \gls{GHG} formulation will be detailed.
+The rest of the formulation is detailed and available in previous works \cite{limpens2021generating}.
+This work uses the following nomenclature: SETs are in capital letters, \textbf{Variables} are in bold and with first letter capital, and \emph{parameters} are in italic.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+\label{eq:obj_func}%1
+% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
+\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot}} = \sum_{\mathclap{j \in \text{\emph{TECH}}}} \Big(\textbf{$\tau$}(j) \textbf{C\textsubscript{inv}}(j) + \textbf{C\textsubscript{maint}} (j)\Big) + \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{C\textsubscript{op}}(i)&\\
+\label{eq:tau}%2
+ \text{s.t.
+} & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)} - 1} & \forall j \in \text{\emph{TECH}}\\
+ \label{eq:c_inv}%3
+ & \textbf{C\textsubscript{inv}}(j) = c_{\text{\emph{inv}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\
+ \label{eq:c_maint}%4
+ & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\ 
+  \label{eq:c_op}%5
+ & \textbf{C\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t) t_{op} (t)  
+ & \forall i \in \text{\emph{RES}}
+ \end{flalign}
+ \endgroup
+
+The objective, Eq.~(\ref{eq:obj_func}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
+The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau}).
+The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv}).
+The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler.
+The total operation and maintenance costs are calculated in the same way, Eq.~(\ref{eq:c_maint}).
+The total cost of the resources is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op}).
+To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op}) and following equations.
+The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}.
+The full formulation is detailed in \cite{limpens2019energyscope} or in the documentation \cite{readthedocs_estd_v2.2}.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign}
+ \label{eq:GWP_tot}%8
+ \phantom{\hspace{0pt} \min \text{  } }
+ & \textbf{GWP\textsubscript{tot}}  =    \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{GWP\textsubscript{op}} (i) 
+ &\\
+  \label{eq:GWP_op}%7
+ & \textbf{GWP\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t)  t_{op} (t) & \forall i \in \text{\emph{RES}}
+\end{flalign}
+\endgroup
+
+The global annual \gls{GHG} emissions are calculated using a \gls{LCA} approach, i.e.
+taking into account emissions of the resources `\emph{from cradle to use}'.
+It is based on the indicator `\emph{GWP100a-IPCC2013}´ developed by the \acrfull{IPCC} \cite{IPCC2014}.
+For climate change, the natural choice as indicator is the \acrlong{GWP}, expressed in ktCO$_2$-eq./year.
+In Eq.~(\ref{eq:GWP_tot}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}).
+The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Eq.~(\ref{eq:GWP_op}).
+Thus, this version accounts only for operation without accounting for the \gls{GWP} emitted during the construction of the technologies.
+This makes the results comparable with metrics used in the reports by the European Commission and the \acrfull{IEA}.
+
+The above equations (Eqs.
+(\ref{eq:obj_func}) - (\ref{eq:GWP_op})) represent only a part of the formulation and illustrate the syntax that is used.
+Those representing the energy balance, network implementation, sectors representation, etc.
+are not presented in this work but are detailed in the latest version of the model, see \cite{limpens2021generating} and on the documentation \cite{readthedocs_estd_v2.2}.
+
+Finally, energy storage has two dimensions to be optimised: the stored energy quantity (also referred to as 'storage level') and the hourly power flow, encompassing both charging and discharging.
+EnergyScope TD optimises the hourly charge and discharge operations based on the hourly resolution of the typical days.
+In contrast, the optimisation of stored energy is conducted over the entire span of 8760 hours in a year.
+This formulation allows for the effective integration of a wide range of energy storage technologies, spanning short-term solutions like small thermal storage units and daily-use batteries, to longer-term options such as hydro-dam storage for seasonal storage, and even large-scale thermal storage for intra-week patterns.
+A previous study delved into the roles of various storage technologies, considering their sectoral applications and temporal aspects, within the context of the Swiss energy system \cite{Limpens_role_2019}.
+
+\subsection{Extending the model for pathway optimisation}
+\label{sec:path_methodo}
+
+In this section, we delve into the extension of EnergyScope TD from a static yearly snapshot model to a comprehensive pathway model.
+While snapshot models provide insights into the energy system for individual years, they lack the capacity to capture the dynamics inherent in investment strategies throughout a transition period.
+The proposed approach involves segmenting the transition into five-year intervals, during which the energy system is optimized for one specific year.
+This approach results in seven instances of EnergyScope TD -- called representative years -- spanning the 30-year transition period, covering the years from 2020 to 2050.
+To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition.
+Overall, these constraints are integrated into a linear framework, ensuring computational efficiency, with an approximate computational time of 14 minutes on a personal laptop (2.4 GHz Intel Core i5 quad-core).
+
+Figure \ref{fig:meth_path_methodology} illustrates the pathway concept.
+Simplification and choices were necessary to implement linearly the problem while keeping a tractable computational time.
+In this section, we present the formulation retained.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{ES_Pathway.pdf}
+\caption{The pathway methodology relies on 7 representative years (blue boxes) where the model \gls{ESTD} is applied.
+Moreover, the formulation accounts for linking constraints (black boxes) and an initial condition (grey box).
+The overall problem is the pathway model.}
+\label{fig:meth_path_methodology}
+\end{figure}
+
+The proposed formulation relies on representative years, selected every 5 years from 2020 to 2050.  The period between two of them is called `\emph{PHASE}´.
+For each of these 7 representative years, the EnergyScope TD model is run  using the relevant data (such as energy demand, technology costs or \gls{GHG} emissions constraints).
+
+As a consequence, a new dimension `\emph{year}' is added to all \textbf{Variables} and parameters, except the interest rate (\emph{i\textsubscript{rate}}) assumed constant during the transition.
+This new dimension is necessary to represent the changes of technology and resource characteristics over the representative years.
+As an example, the investment cost (\emph{c}\textsubscript{\emph{inv}}) of a solar photovoltaic panels could drastically vary in the next decades (e.g.
+data used ranges between 1220 to 870  [€\textsubscript{2015}/kW] between 2020 and 2035).
+
+\vspace{0.2cm}\textbf{Linking years}\vspace{-0.3cm}\\
+
+At this stage, all years are independent.
+In the following, we introduce new constraints to link representative years.
+The formulation allows to install new capacity (\textbf{F\textsubscript{new}}), remove a capacity that has reached  its lifespan (\textbf{F\textsubscript{old}}) or decommission a technology prematurely (\textbf{F\textsubscript{decom}}).
+These capacity changes occur during a phase, this implies that there is no capacity change during a representative year.
+Figure \ref{fig:path_eg_igcc} illustrates the concept.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{path_e.g._tech.png}
+\caption{Example of how the technologies capacity and associated variables are evolving.
+The example uses a technology with a 20 years lifetime.
+Initially 1~GW of capacity exists (\textbf{F\textsubscript{new}} during phase $2015\_2020$).
+Then another 1~GW is deployed (\textbf{F\textsubscript{new}} during phase $2020\_2025$).
+15 years later, a part of the capacity reaches its lifetime limit and is removed (\textbf{F\textsubscript{old}} phase $2035\_2040$).
+Moreover, during the latter phase, additional capacity is decommissioned  prematurely (\textbf{F\textsubscript{decom}}).
+Finally, the technology reaches its expected lifetime and is fully withdrawn (\textbf{F\textsubscript{old}}).}
+\label{fig:path_eg_igcc}
+\end{figure}
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:F_newBuilt}%5
+&\textbf{F}(y\textsubscript{stop},i) = \textbf{F}(y\textsubscript{start},i)
+ + \textbf{F\textsubscript{new}}(p,i)
+ - \textbf{F\textsubscript{old}}(p,i)
+ - \sum_{\mathclap{p2 \in \text{\emph{PHASE}} \cup \{2015\_2020\}}} \textbf{F\textsubscript{decom}}(p,p2,i)& \notag \nonumber 
+ \end{flalign}
+\begin{flalign} 
+ &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), i \in \text{\emph{TECH}}
+ \end{flalign}
+
+\endgroup
+
+Similarly to a mass balance, Eq.~(\ref{eq:F_newBuilt}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
+at the end of the phase (\emph{y\textsubscript{stop}}), the available capacity is the one used in the next representative year (\textbf{F}(\emph{y\textsubscript{stop}})).
+This capacity is equal to the one available in the previous representative year (\textbf{F}(\emph{y\textsubscript{start}})) plus the new installed capacity (\textbf{F\textsubscript{new}}) minus the capacity that has reached its lifetime (\textbf{F\textsubscript{old}}) minus the early decommissioned capacity (\textbf{F\textsubscript{decom}}).
+One notices that the capacity available for each representative year depends on a year (\emph{y\textsubscript{start}} or \emph{y\textsubscript{stop}}), while the other capacity changes depend on a phase (\emph{p} or \emph{p}2).
+Moreover, the decommissioning term depends on another phase, which is the one when the technology decommissioned has been built.
+As an illustration, Figure \ref{fig:path_eg_igcc} gives an example where 0.5 GW of a capacity built in 2015\_2020 is decommissioned in 2030\_2035 (\textbf{F\textsubscript{decom}}(2030\_2035,2015\_2020,i)).
+
+ \begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+ & \textbf{F\textsubscript{decom}}(p,p2,i) = 0 \hspace{-2cm}&
+\notag \nonumber
+ \end{flalign}
+\begin{flalign} 
+  \label{eq:F_decomNonPhysic1}
+& &  \forall i \in \emph{TECH}, p\in \emph{PHASE}, p2 \in \emph{PHASE} \cup \{2015\_2020\} | \emph{decom\textsubscript{allowed}}(p,p2) = 0 & 
+ \end{flalign}
+  \begin{flalign}
+ \label{eq:Fold_def}%1
+&\textbf{F\textsubscript{old}} (p,i) =  
+\hspace{3mm}\text{if} ( \emph{age} = `\emph{STILL\_IN\_USE}') \text{ then  } 0   &   &\notag\nonumber\\
+&\hspace{2cm}\text{else } \left(\textbf{F\textsubscript{new}}(age,i)  - \sum_{\vspace{5mm}\mathclap{p2 \in \text{\emph{PHASE}}}} \textbf{F\textsubscript{decom}} (p2,age,i)\right)    & \notag\nonumber\\
+&& \hspace{-5cm} \forall p \in PHASE , \forall j \in TECH | age \in AGE(p,j)
+ \end{flalign}
+\endgroup
+
+In linear programming, a solution might be mathematically correct, while not making sense in practice.
+As an example, a technology could be decommissioned before being built ($ \emph{p} < \emph{p\textsubscript{built}}$).
+Eqs.~(\ref{eq:F_decomNonPhysic1}-\ref{eq:Fold_def})  allow preventing these non-sense while keeping the formulation linear.
+Eq.~(\ref{eq:F_decomNonPhysic1}) forces the decommissioned capacity to zero when technology will be built after.
+To do so, a parameter (\emph{decom\textsubscript{allowed}}) is defined \emph{a priori} and is equal to 0 or 1 when decommissioning is not possible or possible, respectively. 
+Eq.~(\ref{eq:Fold_def}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}.
+For each phase, a set (\emph{AGE}) is calculated \emph{a priori}.
+It relates, for a given phase and technology, when the technology should have been built.
+In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built.
+The first part of Eq.~(\ref{eq:Fold_def}) indicates that the technology is still available, and thus no capacity needs to be removed.
+The second part of the equation represents the capacity that reached its expected lifetime minus a part of the capacity that would have been decommissioned.
+As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
+Ones will highlight the use of a `if' in Eq. (\ref{eq:Fold_def}), this formulation is linear as the if is applied to a parameter and not a variable.
+
+ \begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+ \label{eq:Phase2020Design}
+& \textbf{F\textsubscript{new}}(\emph{2015\_2020},i) = \textbf{F}(\emph{YEAR\_2020},i)& \forall i \in \emph{TECH}
+ \end{flalign}
+\endgroup
+
+To initialise the problem in 2020 with the existing design, an additional phase `\emph{2015\_2020}' is created.
+Eq.~(\ref{eq:Phase2020Design}) requires that the capacity used in 2020 is installed in the previous phase.
+
+\vspace{0.2cm}\textbf{Society inertia}\vspace{-0.3cm}\\
+
+To avoid unrealistically fast changes in the system, additional constraints are needed during the phases for the mobility and low temperature heat sectors.
+Without the following constraints, the model would eliminate certain technologies in one phase, such as oil and gas decentralised boilers.
+Even if this result is mathematically and physically correct, (\ie fuels are expensive and investing in more efficient technology is economically and environmentally more profitable), this swap of technology cannot occur in one phase (\ie 5 years).
+Indeed, society inertia to change, available manpower, supply chains and manufacturers limit the change.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:delta_tech_during_phase}
+& \textbf{$\Delta$\textsubscript{change}}(p,i) \geq 
+   \sum_{\mathclap{t \in T }} \left( \textbf{F\textsubscript{t}}(\emph{y\textsubscript{start}},i,t) \right) 
+   -\sum_{\mathclap{t \in T }} \left( \textbf{F\textsubscript{t}}(\emph{y\textsubscript{stop}},i,t) \right)&\notag\nonumber\\
+&&\hspace{-8cm} \forall j \in \emph{TECH}, p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p)
+ \end{flalign}
+ \begin{flalign}
+\label{eq:limit_reno_LTheat}
+&\sum_{\mathclap{\hspace{2cm}i\in TECH(HeatLowT)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
+\leq
+ \emph{lim\textsubscript{LT,ren}} \cdot 
+\big( {eui}(\emph{y\textsubscript{start}},HotWater)  +  \emph{eui}(\emph{y\textsubscript{start}},SpaceHeat)  \big)\hspace{-1cm}&\notag\nonumber\\
+&\hspace{5.5cm} \forall p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p) & \\
+\label{eq:limit_reno_passmob}
+&\sum_{\mathclap{\hspace{2cm}i\in TECH(MobPass)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
+\leq
+ \emph{lim\textsubscript{MobPass}} \cdot \emph{eui}(\emph{y\textsubscript{start}},MobPass) )&\notag\nonumber\\
+&\hspace{5.5cm} \forall p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p) & \\
+\label{eq:limit_reno_freight}
+&\sum_{\mathclap{\hspace{2cm}i\in TECH(MobFreight)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
+\leq
+ \emph{lim\textsubscript{MobFreight}} \cdot \emph{eui}(\emph{y\textsubscript{start}},MobFreight) &\notag\nonumber\\
+&\hspace{5.5cm} \forall p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p) &
+\end{flalign}
+\endgroup
+
+%\hspace{3cm} euc \in EUT\_OF\_CAT(HeatLowT),i\in \emph{TECH\_OF\_EUT}(euc)
+%\hspace{3cm} euc \in EUT\_OF\_CAT(MobPass),i\in \emph{TECH\_OF\_EUT}(euc)
+%\hspace{3cm} euc \in EUT\_OF\_CAT(MobFreight),i\in \emph{TECH\_OF\_EUT}(euc)
+
+Eq.~(\ref{eq:delta_tech_during_phase}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
+%% GL_correction_v2 %% Following line added:
+%  Ici on prend la différence entre ystart et ystop (et non pas le contraire), en effet, on est intéressé par la réticence des personnes à se séparer de leur ancienne technologie, plutot que la réticense à accepter une nouvelle techno.
+%% GL_correction_v2 %% end
+Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Eq.~(\ref{eq:limit_reno_LTheat}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Eq.~(\ref{eq:limit_reno_passmob}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Eq.~(\ref{eq:limit_reno_freight}).
+For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase.
+Consequently, if a technology supplies more than 25\% of the low temperature heat, it would require multiple phases to replace it with a different technology.
+%As an example, setting the maximum change of supplied heat at low temperature to 25\% would limit the change in technology related to heat at low temperature to 25\% during a phase. 
+%In this case, if more than 25\% of the low temperature heat was supplied by a technology, it would take more than one phase to replace it by a different one.
+
+\vspace{0.2cm}\textbf{Cost and emissions of the transition}\vspace{-0.3cm}\\
+
+To optimise the energy system, two key metrics must be adapted: the transition cost and the total \acrfull{GWP}. 
+Concerning the first one, all costs are expressed in €\textsubscript{2015} and an annualisation factor is used to distinguish investments over the transition.
+For the \gls{GWP}, the metric used is based on the contributions of the gases over 100 years.
+It is assumed that the impact of emitting at the beginning or the end of transition are equivalent and thus no annualisation is used.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+\label{eq:obj_func_v2}%1
+% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
+\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
+\label{eq:Capex_v2}
+& \textbf{C\textsubscript{tot,capex}} =
+\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
+\textbf{C\textsubscript{inv,phase}}(p)
+-
+\sum_{\mathclap{i \in \emph{TECH}}} 
+\textbf{C\textsubscript{inv,return}}(i)\\
+  \label{eq:Copex_tot_v2}%5
+& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
++ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}}
+ \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
+\label{eq:path_annu_factor}
+& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+\end{flalign}
+\endgroup
+
+The objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
+%The objective is the minimisation of the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}),
+defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2}).
+The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2}), to which the residual asset investment cost in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
+Thus, the investments account for the installation and dismantlement costs of the technologies.
+The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Eq.~(\ref{eq:Copex_tot_v2}).
+During a phase, the system \gls{OPEX} is
+the product of the annualised phase factor, defined in Eq.~(\ref{eq:path_annu_factor}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
+The annualised phase factor is defined based on an average interest rate during the transition.
+%assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor}).
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:opex_yearly}
+&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{i \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,i) + \sum_{\mathclap{j \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,j)&\forall y\in \emph{YEARS}
+\end{flalign}
+\endgroup
+
+For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Eq.~(\ref{eq:opex_yearly}).
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:PhaseInv}%5
+&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
+\end{flalign}
+\begin{flalign}
+&&\forall p \in \emph{PHASE} | y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)
+\end{flalign}
+\endgroup
+
+The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Eq.~(\ref{eq:PhaseInv}).
+The annualised phase factor  (defined by Eq.~(\ref{eq:path_annu_factor})) is used.
+The specific cost during the phase is defined as  the average between the investment cost for the first and last year of the period.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:salvage}%5
+&\textbf{C\textsubscript{inv,return}}(i) = \hspace{2.5cm}
+\sum_{\mathclap{p \in \emph{PHASE}\cup \{2015\_2020\}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}} 
+\hspace{0.5cm}
+ \tau_{phase}(p)\cdot \left(c_{inv}(y_{start},i)+c_{inv}(y_{stop},i)\right)/2 \cdot
+&\notag\nonumber
+\end{flalign}
+\begin{flalign}
+& 
+\hspace{1.7cm}
+\frac{remaining\_years(i,p)}{lifetime(y\textsubscript{start},i)} \left( \textbf{F\textsubscript{new}}(p,i) - 
+\sum_{\mathclap{p2 \in \emph{PHASE}}} 
+\textbf{F\textsubscript{decom}}(p2,p,i)\right)&
+\forall i \in \emph{TECH}
+\end{flalign}
+\endgroup
+
+A part of the investment will remain after 2050. This residual investment, also called salvage value, can be calculated for each technology. 
+A parameter, calculated \textit{a priori} , gives for each technology and construction phase, the remaining amount of years (\emph{remaining\_years}). 
+As an example, if a PV panel has been built in 2045 and has a 20 years lifetime, the parameter will equal to 15 years. 
+Thus, the salvage value is a fraction of the investment cost of this technology when it has been built. 
+This fraction is the ratio between the number of remaining years and the lifetime of the technology. 
+In the previous example, the residual investment of the PV built is 75\%. 
+Eq.
+(\ref{eq:salvage}) computes, for each technology, the residual value that must be deducted from the total cost.
+The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised.
+The residual value is not applied to technologies that are removed prematurely.
+This differ from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:gwp_tot_transition}
+&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}/2 \left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)&
+\\
+\label{eq:limit_gwp_trans}
+& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
+\end{flalign}
+\endgroup
+
+The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Eq.~(\ref{eq:gwp_tot_transition}).
+The emissions during a phase is estimated as the arithmetic average of the representative years before and after the phase.
+Eq.~(\ref{eq:limit_gwp_trans}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).

### 2024-01-18 appendices/app_case_study.tex
Drop full implementation of Pathway from the paper

@@
 \end{table}
+
+\section{Belgian energy transition pathway towards carbon-neutrality in 2050} 
+\label{app:bel_PF_TD}
+This section presents the results of the deterministic (\ie all parameters at their respective nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon-neutrality in 2050.
+After performing a technical investigation of the pathway by checking the greenhouse gas breakdown by energy sectors, the primary energy mix is analysed.
+To illustrate the sector coupling, a focus is made on the electrification of other sectors.
+Then, the cost implications in terms of investments and operations are discussed.
+
+\subsection{Greenhouse gases and primary energy}
+% GHG reduction
+Figure \ref{fig:pestd_ghg} shows the \acrfull{GHG} per sector. 
+%As a reminder, the total \acrshort{GHG} is capped linearly between its historical value in 2020 (121~MtCO\textsubscript{2,eq}) and 0 in 2050.
+This approach is necessary for the myopic approach, while other approaches, such as imposing a carbon budget over the transition or an end goal (carbon neutrality in 2050) would be more realistic\footnote{The carbon budget has been implemented for the perfect foresight in a previous work \cite{limpens2021generating}.
+In this work, Figure 7.14 shows a Pareto Carbon-budget - transition cost.
+The study shows that no transition would be 7\% less expensive but emitting 75\% more \gls{GHG} while not achieving carbon neutrality.
+Moreover, reducing the carbon budget by a little amount could have a small impact on the cost down to a certain amount.
+As an example, reducing the carbon budget by 13\% will imply an extra cost of 1.3\%.}.
+The system reaches its upper bound (\ie maximum emissions) every year.
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Gwp_per_sector.pdf}
+\caption{Energy sectors have different speed to reduce \gls{GHG} emissions over the transition.
+The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon-neutrality in 2050.}
+\label{fig:pestd_ghg}
+\end{figure}
+
+The defossilisation\footnote{In a more sustainable future, some of the energy carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons (\eg e-methane or e-methanol).
+This is why this paper rather uses ``defossilisation" rather than ``decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.} of the different sectors are not performed at the same rate.
+The non-energy demand of methanol and ammonia are substituted by electrofuels.
+These are the first use of electrofuels as  e-ammonia is the cheapest electrofuel thanks to the high maturity of the Haber-Bosch process.
+The decentralised heat and mobility sectors are also dropping first.
+This is a combination of efficiency and substitution of fossil fuels with electricity.
+Efficiency comes mainly from district heating networks and electrical heat pumps for the heat sector, and public mobility and electric cars for the mobility sector.
+From 2040 onward, the decreases are mainly due to the substitution of the remaining fossil fuels by electrofuels as illustrated in Figure \ref{fig:pestd_primary_energy}.
+
+Figure \ref{fig:pestd_primary_energy} shows the primary energy mix for the different representative years.
+The pathway verifies five trends: (i) reduction of primary energy thanks to energy efficiency; (ii) massive integration of endogenous renewable energies; (iii) importance of electrification; (iv) the usage of gas as the last fossil resource; and (v) the obligation to rely on renewable fuels to achieve carbon neutrality.
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Resources.pdf}
+\caption{Primary energy emitting \gls{GHG} (below Uranium) are reducing linearly with fossil gas remaining until 2045.
+A part of this energy is replaced by renewable ones and starting from 2040, a significant share of electrofuels.
+As end-use demands slightly increase (see Figure \ref{fig:cs_demands}), the drop represents energy efficiency (\ie providing the same services with less primary energy).}
+\label{fig:pestd_primary_energy}
+\end{figure}
+
+The energy supply decreases from 554~TWh/y in 2020 down to 368~TWh/y in 2050 (\ie -34\%) whereas, in the meantime, the demands have increased by 19\%, on average.
+This drop of primary energy consumption reflects the penetration of efficient measures and technologies, such as the previously mentioned public mobility, \gls{DHN} or heat pumps.
+The results in 2050 are aligned with other studies, such as \citet{Devogelaer2013}\footnote{This study was ordered by the National Planning Bureau in 2013.
+Five scenarios are proposed.} and My2050\footnote{The Climate Change Service of the Federal Public Service Health launched an initiative in 2012 entitled `\emph{Low Carbon Belgium by 2050}'.
+This initiative resulted in a report and a calculator in 2013 \cite{Cornet2013}.
+The Belgian calculator has been improved since then into a recent expert version called \textbf{My2050} \cite{My2050}.
+From this study, the results of two scenarios will be used: one based on an optimistic evolution of technologies (Technology), and one focusing on an increased dependence on neighbouring countries (EU integration).} \cite{My2050} which estimates respectively a range of 305-417 TWh/y and 307-364~TWh/y for their central scenarios.
+
+The first fossil energy to phase out is gasoline, which is exclusively used for private cars.
+Indeed, private mobility is partially replaced by public one\footnote{Given the major role played by private cars in the Belgian passenger mobility nowadays (\ie around 80\% \cite{BFP_mob}), public transport (\eg tramways, buses and trains) is assumed to be able to supply only half of it.}; and the cars are switching from gasoline and diesel to electricity.
+Then, diesel and \gls{LFO} are decreasing.
+As diesel is used for trucks and buses mobility, it is harder to phase out compared to gasoline exclusively burned in cars.
+The first drop of \gls{LFO} reflects the switch from oil boilers to other technologies: heat pumps and gas cogeneration mainly.
+Then, it is mainly used for the production of \gls{HVC}, this reflects that \gls{HVC} is a feedstock hard to defossilise.
+Finally, coal is kept mainly for industrial usage because it is a cheap fossil fuel (mainly for industrial usage).
+To phase it out beforehand, a penalty mechanism, such as a carbon tax, would be required, or its strict ban should be put in place.
+The last fossil energy present in the system is fossil gas, used for the production of electricity and heat, through cogeneration mainly.
+Indeed, gas plays a key role to balance the intermittency of solar and wind.
+
+The consumption of uranium declines in 2025, dropping to 2 GW, primarily due to the political framework aimed at phasing out nuclear energy \cite{nuclear_2035}.
+In the initial stages, significant deployment of endogenous energies takes place.
+This includes the utilization of wood, wet biomass, and wind energy, followed by the introduction of solar energy.
+However, solar energy is not fully deployed during this period due to higher integration costs.
+Starting from 2025, the importation of electrofuels begins, although their significant utilisation is observed from 2035 onwards.
+Initially, these fuels are predominantly employed as feedstocks in non-energy sectors.
+From 2040, e-methanol is additionally utilised for the production of High-Value Chemicals (HVCs), e-hydrogen is employed for mobility purposes, and both e-methane and e-ammonia are used for electricity generation through gas \gls{CHP} and ammonia-based \gls{CCGT} plants (see Figure \ref{fig:ELEC_tech_cap}).
+
+In 2020, Belgium has been a net-exporter of electricity, however with the shut-down of nuclear power plants and the increase of electricity consumption, Belgium will become a net importer of electricity.
+These imports reach their maximal allowed capacity by 2035 (i.e.
+30\% of electricity end use).
+This strong dependence on imported electricity illustrates the need for balancing intermittent renewables without relying on fossil fuels.
+
+\subsection{Electricity sector: Capacities and yearly balance}
+\label{subsubsec:elec_sector}
+To better understand the electricity sector, the installed production capacities are given in Figure \ref{fig:ELEC_tech_cap}, while the supply-demand yearly balance is illustrated in Figure \ref{fig:ELEC_layer}.
+
+\begin{figure}[!htbp]
+     \centering
+         \includegraphics[width=\textwidth]{ELEC_tech_cap.pdf}
+         \caption{The electrical production capacity will experience  massive expansion of wind turbines (onshore and then offshore) and a soaring installed capacity of \gls{PV}.
+Ammonia \gls{CCGT} are installed at the end of the transition to provide a flexible capacity as gas \gls{CCGT} are phased out.}
+         \label{fig:ELEC_tech_cap}
+\end{figure}
+
+\begin{figure}[!htbp]
+     \centering
+         \includegraphics[width=\textwidth]{ELEC_layer.pdf}
+         \caption{The electricity supply (positive values) will remain a mix of different technologies where backup is first mainly provided by gas-\gls{CCGT} and then imported electricity, heat and power cogeneration and later ammonia-\gls{CCGT}.
+The electricity demand (negative values) is led by the electricity end-use demand, but the share used to electrify heat (heat pumps), vehicles (cars, trains, trams, ...) and industrial heaters drastically increase.
+This enables a flexible demand that can facilitate the integration of intermittent renewables.}
+         \label{fig:ELEC_layer}
+\end{figure}
+
+As introduced in the primary energy analysis (see Figure \ref{fig:pestd_primary_energy}), renewable capacities soar.
+By 2050, wind and solar technologies deployments are 60~GW of \gls{PV}, 10~GW of onshore wind turbines and 6~GW of offshore wind turbines.
+To compensate the intermittency, the system relies on imported electricity, gas \gls{CCGT}, sector coupling and storage.
+As an illustration, in 2050, 176.8~TWh of electricity transit on the grid which includes 32.4~TWh of electricity imported and 15.4~TWh of electricity from \gls{CCGT}.
+This result is aligned with other studies that estimates different ranges: 180-310 TWh/y \cite{Devogelaer2013}, 126-140~TWh/y \cite{My2050} and in a more recent study
+%\footnote{The EnergyVille consortium, made of belgian universities and research institutes, applied the TIMES-BE model to the case of Belgium to assess the energy transition pathways \cite{PATHS2050}.
+In this study, three scenarios are proposed: electrification, clean molecules and reference (called central in their report).
+They all reach carbon neutrality in 2050.
+The first allows more power production with new nuclear plants and more offshore wind.
+The clean molecules allow the import of synthetic molecules at a lower costs and have a more limited access to cross-border CO\textsubscript{2} storage.}
+using the TIMES-BE model, 185-196~TWh/y \cite{PATHS2050}.
+Higher values from \citet{Devogelaer2013} illustrate an almost exclusively electrified energy system.
+The differences between the study ranges reflect the different assumptions in terms of renewable potentials and availability of nuclear energy.
+A general trend is that Belgium should maximise its use of endogenous renewable resources, which \citet{dubois2023multi} identified as a cheaper option than importing additional renewable energies from abroad.
+Demand management reflects the flexible use of electricity, mainly through heat pumps that uncouple the heat demand and the electricity consumption when combined with thermal storage.
+Gas \gls{CCGT} is also a useful asset to compensate intermittent renewables.
+However, its capacity remains the same as the one installed in 2020.
+These results are verifying an hourly adequacy of the power demand.
+Moreover, in a previous study by \citet{pavivcevic2022bidirectionnal}, the snapshot version of the model has been coupled with Dispa-SET, a dispatch optimisation model.
+Results showed that the backup capacity was underestimated by less than 20\% to respect reserve capacity, mainly due the lack of reserve capacity for grid stability.
+
+From 2025, the electricity mix has a strong renewable share that rises up to 60\% in 2050.
+The remaining 40\% are mainly gas (or ammonia) in \gls{CCGT} and cogeneration and imported electricity.
+From a demand perspective, the electrification first starts with \gls{DHN} heat pumps, then electric cars, then decentralised heat pumps and finally industrial heaters.
+The latter reflects the usage of cheap \gls{PV} production peaks.
+
+\subsection{Costs: Investments and operation}
+
+In the following paragraphs, the results are analysed from an economic perspective to decipher the choices made by the model, as the overall cost of the transition is 1 004~b€\textsubscript{2015} split unequally among the sectors.
+
+Figure \ref{fig:pestd_cumul_inv} illustrates the cumulative investments made throughout the transition, amounting to a total of 377.8 b€\textsubscript{2015}.
+Initially, the infrastructure, transport, and electricity sectors each account for approximately one-third of the investments.
+The investments in infrastructure are primarily driven by the electricity grid and the \acrfull{DHN}, representing a combined investment of 73 b€\textsubscript{2015}.
+The electricity sector's investment is led by power plants, totalling 31.5 b€.
+Notably, the investment costs in the mobility sector are primarily attributed to private cars, constituting 71\% of the total.
+A rough estimation confirms the significant investment in cars, with an average of 500,000 vehicles registered annually in Belgium over the last decade \cite{febiac2021datadigest} and assuming an average cost of 20 k€ per car, the funds allocated to private cars amount to 10 b€ per year.
+This trend in private cars explains why the private mobility sector accounts for half of the investments required to achieve the transition by 2050.
+This finding aligns with other studies, such as \citet{Devogelaer2013}, which estimates cumulative investment expenditures of approximately 600 b€\textsubscript{2005} for the transport sector between 2013 and 2050, which confirms our conservative approach in the estimation.
+
+As a comparison, the investments required to fully deploy the PV and wind potentials from 2020 to 2050 amount to 74.4 b€\textsubscript{2015}, with an additional 22.2 b€\textsubscript{2015} allocated to reinforce the grid.
+The electrification of the heating sectors necessitates investments of 29.2 b€\textsubscript{2015}, including 6.5 b€\textsubscript{2015} for the deployment of the \gls{DHN} infrastructure.
+Storage investments, primarily focused on \gls{DHN} seasonal storage, amount to 3.6 b€\textsubscript{2015}.
+Apart from the investment required to replace all private vehicles (accounting for 44\% of the overall investments), the remaining sectors represent a total of 212 b€\textsubscript{2015}.
+To mitigate the cost of the transition, My2050 suggests deploying a fleet of no more than one million vehicles and implementing a car sharing system, distinct from car-pooling, as an inevitable measure \cite{My2050}.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Investment_transition.pdf}
+\caption{The cumulative investments over the transition is unequally spread between the sectors.
+The energy system in 2020 is imposed to the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
+The investments required during the transition represents 150\% the initial investment and mainly in the same three sectors.
+}
+\label{fig:pestd_cumul_inv}
+\end{figure}
+
+A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technology after 2050.
+Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Eq.
+(\ref{eq:salvage}).
+Out of the 114.4~b€\textsubscript{2015} of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime.
+On the contrary, private mobility has a lower salvage value due to a major drop within the first four years and an average lifetime below 10 years \cite{febiac2021datadigest}.
+% Consequently, out the 240.5~b€\textsubscript{2015}, including the initial investment of 75.7~b€\textsubscript{2015}, only 13.9\% would remain available after 2050.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Investment_return.pdf}
+\caption{By the end of the transition (\ie in 2050), the ratio between the salvage value and its cumulative investment, per sector, is unequal.
+Investments in infrastructures, public mobility, storage and other long-lifetime technologies experience an important salvage value, at the contrary, investments in private mobility will not be recovered as vehicles have a short lifetime.
+All together, these salvage values represent 160.1 b€\textsubscript{2015}, 25\% of the cumulative investment costs in 2050.}
+\label{fig:pestd_inv_return}
+\end{figure}
+
+In addition to investment decisions, the \acrfull{OPEX}, which accounts for resource utilisation and technology maintenance, are significant.
+Figure \ref{fig:pestd_cumul_op} shows the yearly system cost for each sector except the \gls{OPEX} related to resources that are grouped together.
+The latter dominates the \gls{OPEX}, with a significant share of non-renewable resources (\ie 63.6\% in 2020) until 2040, followed by a steep increase in the share of renewable resources (\ie 66.2\% in 2050).
+The substantial reliance on non-renewable resources reflects the prevalent use of fossil fuels in our current energy system.
+The high cost-share of non-renewable fuels underscores the economic challenges of simply substituting fossil fuels with renewables, particularly evident when emphasizing that electrofuels are 2-3 times more expensive.
+Maintenance expenses in the private mobility sector rank second in terms of expenditure.
+On the other hand, maintenance expenses in other sectors are relatively small compared to the aforementioned sectors.
+
+\begin{figure}[!htbp] %@GL: check that text is aligned with new figure.
+\centering
+\includegraphics[width=\textwidth]{Operation_transition.pdf}
+\includegraphics[width=\textwidth]{System_cost.pdf}
+\caption{The yearly system cost shows the shift from non-renewable to renewable resources (mainly electrofuels). Operation cost and maintenance represents almost 50\% of the expenses.}
+\label{fig:pestd_cumul_op}
+\end{figure}
+
+The annualised cost of the energy system in 2020 is estimated to 44.3~b€/y and increases by 5.5~b€/y to reach 49.8~b€/y by 2050.
+\textbf{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y \cite{My2050}, while the other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
+The differences come from the scope of the energy system, as an example \textbf{My2050} also accounts for the agriculture sector.
+These differences highlight the difficulty to compare different studies due to difference of scope and partial availability of used data.
+Overall, comparing with existing studies shows the consistency of the results provided by EnergyScope Pathway.
 

### 2024-01-18 chap_methodo/chap_methodo.tex
Drop full implementation of Pathway from the paper

@@
 \item Check that PCE was appropriate as a method for such a system (ECOS2020)
+\item Work on the $C_inv_return$
 \end{itemize}
@@
 On the contrary, this work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}.
-According to pathway models review, EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is open-source documented model.
+According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is open-source documented model.
 Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation).
-Presenting here only the main constraints of the model and the main input data in Section \ref{chap:case_study}, the reader is invited to refer to the extended paper \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway} for more extensive information.
+From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the model.
+The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from a snapshot approach, EnergyScope TD.
+More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
 
@@
 The pathway's initialisation assumes that all capacities installed in 2020 were built during the pseudo-phase of 2015-2020 (grey box).
-The overall problem is defined as the pathway model.
-Graph adapted from \cite{limpens2024pathway}.}
-\label{fig:meth_path_methodology}
+The overall problem is defined as the pathway model.}
+\label{fig:meth_path_methodology_core}
 \end{figure}
-%
-%We used the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology})---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050.
+
+We used the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core})---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050.
 Even though this lacks realism compared to the myopic approach \cite{babrowski2014reducing,fais2016impact,heuberger2018impact} where the pathway is optimised via a sequence of shorter time windows, the perfect foresight approach allows to optimise the transition with a \ce{CO2}-budget (Section \ref{subsec:cs:CO2-budget}) rather than along a prescribed \ce{CO2}-trajectory.
@@
 For instance, the energy balance is guaranteed at every hour of each of these years.
-%
-%The optimised objective of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as follows: 
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign} 
-%% Objective function + investment 
-%\label{eq:obj_func_v2}%1
-%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
-%\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
-%\label{eq:CAPEX_v2}
-%& \textbf{C\textsubscript{tot,capex}} =
-%\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
-%\textbf{C\textsubscript{inv,phase}}(p)
-%-
-%\sum_{\mathclap{j \in \emph{TECH}}} 
-%\textbf{C\textsubscript{inv,return}}(j)\\
-%  \label{eq:Copex_tot_v2}%5
-%& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
-%+ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}}
-% \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
-%\label{eq:path_annu_factor}
-%& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
-%%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
-%\end{flalign}
-%\endgroup
-%
-%\noindent
-%where $\emph{t\textsubscript{phase}}=5\,\text{years}$ and $\emph{diff\_2015\_year(p)}$ are respectively the duration of a phase between two representative years and the number of years between the middle of a phase and 2015 for a correct annualisation.
+
+The optimised objective of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as follows:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+\label{eq:obj_func_v2}%1
+% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
+\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
+\label{eq:CAPEX_v2}
+& \textbf{C\textsubscript{tot,capex}} =
+\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
+\textbf{C\textsubscript{inv,phase}}(p)
+-
+\sum_{\mathclap{j \in \emph{TECH}}} 
+\textbf{C\textsubscript{inv,return}}(j)\\
+  \label{eq:Copex_tot_v2}%5
+& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
++ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}}
+ \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
+\label{eq:path_annu_factor}
+& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+\end{flalign}
+\endgroup
+
+\noindent
+where $\emph{t\textsubscript{phase}}=5\,\text{years}$ and $\emph{diff\_2015\_year(p)}$ are respectively the duration of a phase between two representative years and the number of years between the middle of a phase and 2015 for a correct annualisation.
 $\textbf{C\textsubscript{inv,return}}$ accounts for the residual value, also called \textit{salvage value}, of the technologies installed during the transition and having not reached the end of their lifetime by 2050.
@@
 The other variables in Eq.\,(\ref{eq:CAPEX_v2}-\ref{eq:Copex_tot_v2}) are detailed here below:
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign} 
-%\label{eq:opex_yearly}
-%&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i)&\forall y\in \emph{YEARS}\\
-%\label{eq:PhaseInv}%5
-%&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
-%\end{flalign}
-%\begin{flalign}
-%&&\forall p \in \emph{PHASE} | y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)
-%\end{flalign}
-%\endgroup
-%
-%\noindent
-%where $\textbf{F\textsubscript{new}}$ are the capacities newly installed. In Eq. (\ref{eq:opex_yearly}-\ref{eq:PhaseInv}), the costs related to each representative year are:
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign} 
-%% Objective function + investment 
-%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
-%\hspace{0pt} 
-% \label{eq:c_inv}%3
-% &\textbf{C\textsubscript{inv}}(y,j) = c_{\text{\emph{inv}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\
-% \label{eq:c_maint}%4
-% &\textbf{C\textsubscript{maint}}(y,j) = c_{\text{\emph{maint}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\ 
-%  \label{eq:c_op}%5
-% &\textbf{C\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t) t_{op} (t)  
-% & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
-% \end{flalign}
-% \endgroup
-%
-%\noindent where the variable $\textbf{F}$ represents the size of the installed capacities (for all technologies $j$) and the variable $\textbf{F\textsubscript{t}}$ is the hourly consumption of the resources; the parameters $c_{\text{\emph{inv}}}$ and $c_{\text{\emph{maint}}}$ are the CAPEX and the OPEX of the technologies, and the parameter $c_{\text{\emph{op}}}$ is the cost of purchasing resources.
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:opex_yearly}
+&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i)&\forall y\in \emph{YEARS}\\
+\label{eq:PhaseInv}%5
+&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
+\end{flalign}
+\begin{flalign}
+&&\forall p \in \emph{PHASE} | y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)
+\end{flalign}
+\endgroup
+
+\noindent
+where $\textbf{F\textsubscript{new}}$ are the capacities newly installed. In Eq. (\ref{eq:opex_yearly}-\ref{eq:PhaseInv}), the costs related to each representative year are:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
+\hspace{0pt} 
+ \label{eq:c_inv}%3
+ &\textbf{C\textsubscript{inv}}(y,j) = c_{\text{\emph{inv}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\
+ \label{eq:c_maint}%4
+ &\textbf{C\textsubscript{maint}}(y,j) = c_{\text{\emph{maint}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\ 
+  \label{eq:c_op}%5
+ &\textbf{C\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t) t_{op} (t)  
+ & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
+ \end{flalign}
+ \endgroup
+
+\noindent where the variable $\textbf{F}$ represents the size of the installed capacities (for all technologies $j$) and the variable $\textbf{F\textsubscript{t}}$ is the hourly consumption of the resources; the parameters $c_{\text{\emph{inv}}}$ and $c_{\text{\emph{maint}}}$ are the CAPEX and the OPEX of the technologies, and the parameter $c_{\text{\emph{op}}}$ is the cost of purchasing resources.
 For the sake of simplicity, as done by \citet{limpens2024pathway}, the sum over the 8760 hours of the year is written as the sum over $t \in T $.
-%
-%Then, as detailed in section \ref{subsec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign} 
-%\label{eq:gwp_tot_transition}
-%&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}\left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)/2 &
-%\\
-%\label{eq:limit_gwp_trans}
-%& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
-%\end{flalign}
-%\endgroup
-%
-%\noindent
-%where the computation of the yearly emissions are based on the \acrfull{GWP} of the resources:
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign}
-%\hspace{0pt}
-% \label{eq:GWP_tot}%8
-% & \textbf{GWP\textsubscript{tot}}(y)  =    \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{GWP\textsubscript{op}}(y,i) 
-% & \forall y \in \text{\emph{YEARS}}\\
-%  \label{eq:GWP_op}%7
-% & \textbf{GWP\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t)  t_{op} (t) & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
-%\end{flalign}
-%\endgroup
-%
-%\noindent
-%where $gwp_{\text{\emph{op}}}$ is the specific emissions (\ie in kt$_{\ce{CO2},\text{eq}}$/GWh) of each resource.
+
+Then, as detailed in section \ref{subsec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:gwp_tot_transition}
+&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}\left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)/2 &
+\\
+\label{eq:limit_gwp_trans}
+& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
+\end{flalign}
+\endgroup
+
+\noindent
+where the computation of the yearly emissions are based on the \acrfull{GWP} of the resources:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign}
+\hspace{0pt}
+ \label{eq:GWP_tot}%8
+ & \textbf{GWP\textsubscript{tot}}(y)  =    \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{GWP\textsubscript{op}}(y,i) 
+ & \forall y \in \text{\emph{YEARS}}\\
+  \label{eq:GWP_op}%7
+ & \textbf{GWP\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t)  t_{op} (t) & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
+\end{flalign}
+\endgroup
+
+\noindent
+where $gwp_{\text{\emph{op}}}$ is the specific emissions (\ie in kt$_{\ce{CO2},\text{eq}}$/GWh) of each resource.
 Based on an approach developed by the Intergovernmental Panel on Climate Change (IPCC) \cite{stocker2014climate}, this work considers the indicator ``GWP100a - IPCC2013'' to compute the emissions related to the use of resources.
@@
 Consequently, it is not included in this work and not accounted for.
-%
-%Besides this constraint on the emissions, the main constraint to link years with each other is the one dictating the installed capacities at the end of each year:
-%
-%\begingroup
-%\belowdisplayskip=2pt
-%\abovedisplayskip=2pt
-%\begin{flalign} 
-%\label{eq:F_newBuilt}%5
-%&\textbf{F}(y\textsubscript{stop},j) = \textbf{F}(y\textsubscript{start},j)
-% + \textbf{F\textsubscript{new}}(p,j)
-% - \textbf{F\textsubscript{old}}(p,j)
-% - \sum_{\mathclap{p2 \in \text{\emph{PHASE}} \cup \{2015\_2020\}}} \textbf{F\textsubscript{decom}}(p,p2,j)& \notag \nonumber 
-% \end{flalign}
-%\begin{flalign} 
-% &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), j \in \text{\emph{TECH}}
-% \end{flalign}
-%\endgroup
-%
-%\noindent
-%where the variables $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$ are the capacities respectively having reached the end of their lifetime and prematurely decommissioned.
+
+Besides this constraint on the emissions, the main constraint to link years with each other is the one dictating the installed capacities at the end of each year:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:F_newBuilt}%5
+&\textbf{F}(y\textsubscript{stop},j) = \textbf{F}(y\textsubscript{start},j)
+ + \textbf{F\textsubscript{new}}(p,j)
+ - \textbf{F\textsubscript{old}}(p,j)
+ - \sum_{\mathclap{p2 \in \text{\emph{PHASE}} \cup \{2015\_2020\}}} \textbf{F\textsubscript{decom}}(p,p2,j)& \notag \nonumber 
+ \end{flalign}
+\begin{flalign} 
+ &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), j \in \text{\emph{TECH}}
+ \end{flalign}
+\endgroup
+
+\noindent
+where the variables $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$ are the capacities respectively having reached the end of their lifetime and prematurely decommissioned.
 Moreover, to account for the society inertia and to prevent unrealistically fast modal share change, constraints limit this change for the sectors of the low-temperature, the passenger mobility and freight mobility demands.

### 2024-01-18 options/symbols_and_cmd.tex
Case study + UQ ranges
Drop publications, first part of methodo + app. Pathway choice
Drop full implementation of Pathway from the paper

@@
 %% C
+\newabbreviation[group={Acronyms}]{CAPEX}{CAPEX}{capital expenditure}
 \newabbreviation[group={Acronyms}]{CCGT}{CCGT}{combined cycle gas turbine}
+\newabbreviation[group={Acronyms}]{CHP}{CHP}{combined heat and power}
 
@@
 
+%% I
+\newabbreviation[group={Acronyms}]{IEA}{IEA}{International Energy Agency}
+\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{intergovernmental panel for climate change}
+
 %% L
+\newabbreviation[group={Acronyms}]{LCA}{LCA}{life cycle assessment}
 \newabbreviation[group={Acronyms}]{LCOE}{LCOE}{levelised cost of energy}
 \newabbreviation[group={Acronyms}]{LFO}{LFO}{light fuel oil}
+\newabbreviation[group={Acronyms}]{LOO}{LOO}{leave-one-out}
 \newabbreviation[group={Acronyms}]{LPG}{LPG}{liquefied petroleum gas}
@@
 \newabbreviation[group={Acronyms}]{NSC}{NSC}{naphtha steam cracker}
+
+%% O
+\newabbreviation[group={Acronyms}]{OPEX}{OPEX}{operational expenditure}
 
@@
 %% B
+\newcommand*{\boldcheckmark}{%
+  \textpdfrender{
+    TextRenderingMode=FillStroke,
+    LineWidth=.8pt, % half of the line width is outside the normal glyph
+  }{\checkmark}%
+}
 

### 2024-01-25 appendices/app_EnergyScope.tex
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom

@@
 \subsection{The starting point: a scenario analysis model}
+\label{app:ESTD}
 
@@
 % Objective function + investment 
-\label{eq:obj_func}%1
+\label{eq:obj_func_app}%1
 % adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
 \hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot}} = \sum_{\mathclap{j \in \text{\emph{TECH}}}} \Big(\textbf{$\tau$}(j) \textbf{C\textsubscript{inv}}(j) + \textbf{C\textsubscript{maint}} (j)\Big) + \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{C\textsubscript{op}}(i)&\\
-\label{eq:tau}%2
+\label{eq:tau_app}%2
  \text{s.t.
 } & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)} - 1} & \forall j \in \text{\emph{TECH}}\\
- \label{eq:c_inv}%3
+ \label{eq:c_inv_app}%3
  & \textbf{C\textsubscript{inv}}(j) = c_{\text{\emph{inv}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\
- \label{eq:c_maint}%4
+ \label{eq:c_maint_app}%4
  & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\ 
-  \label{eq:c_op}%5
+  \label{eq:c_op_app}%5
  & \textbf{C\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t) t_{op} (t)  
@@
 
-The objective, Eq.~(\ref{eq:obj_func}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
-The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau}).
-The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv}).
+The objective, Eq.~(\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
+The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau_app}).
+The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv_app}).
 The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler.
-The total operation and maintenance costs are calculated in the same way, Eq.~(\ref{eq:c_maint}).
-The total cost of the resources is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op}).
-To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op}) and following equations.
+The total operation and maintenance costs are calculated in the same way, Eq.~(\ref{eq:c_maint_app}).
+The total cost of the resources is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
+To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}) and following equations.
 The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}.
@@
 \begin{flalign}
- \label{eq:GWP_tot}%8
+ \label{eq:GWP_tot_app}%8
  \phantom{\hspace{0pt} \min \text{  } }
@@
  &\\
-  \label{eq:GWP_op}%7
+  \label{eq:GWP_op_app}%7
  & \textbf{GWP\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t)  t_{op} (t) & \forall i \in \text{\emph{RES}}
@@
 For climate change, the natural choice as indicator is the \acrlong{GWP}, expressed in ktCO$_2$-eq./year.
-In Eq.~(\ref{eq:GWP_tot}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}).
-The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Eq.~(\ref{eq:GWP_op}).
+In Eq.~(\ref{eq:GWP_tot_app}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}).
+The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Eq.~(\ref{eq:GWP_op_app}).
 Thus, this version accounts only for operation without accounting for the \gls{GWP} emitted during the construction of the technologies.
@@
 The above equations (Eqs.
-(\ref{eq:obj_func}) - (\ref{eq:GWP_op})) represent only a part of the formulation and illustrate the syntax that is used.
+(\ref{eq:obj_func_app}) - (\ref{eq:GWP_op_app})) represent only a part of the formulation and illustrate the syntax that is used.
 Those representing the energy balance, network implementation, sectors representation, etc.
@@
 \begin{flalign} 
-\label{eq:F_newBuilt}%5
+\label{eq:F_newBuilt_app}%5
 &\textbf{F}(y\textsubscript{stop},i) = \textbf{F}(y\textsubscript{start},i)
@@
 
-Similarly to a mass balance, Eq.~(\ref{eq:F_newBuilt}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
+Similarly to a mass balance, Eq.~(\ref{eq:F_newBuilt_app}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
 at the end of the phase (\emph{y\textsubscript{stop}}), the available capacity is the one used in the next representative year (\textbf{F}(\emph{y\textsubscript{stop}})).
@@
 \begin{flalign} 
-  \label{eq:F_decomNonPhysic1}
+  \label{eq:F_decomNonPhysic1_app}
 & &  \forall i \in \emph{TECH}, p\in \emph{PHASE}, p2 \in \emph{PHASE} \cup \{2015\_2020\} | \emph{decom\textsubscript{allowed}}(p,p2) = 0 & 
@@
   \begin{flalign}
- \label{eq:Fold_def}%1
+ \label{eq:Fold_def_app}%1
 &\textbf{F\textsubscript{old}} (p,i) =  
@@
 As an example, a technology could be decommissioned before being built ($ \emph{p} < \emph{p\textsubscript{built}}$).
-Eqs.~(\ref{eq:F_decomNonPhysic1}-\ref{eq:Fold_def})  allow preventing these non-sense while keeping the formulation linear.
-Eq.~(\ref{eq:F_decomNonPhysic1}) forces the decommissioned capacity to zero when technology will be built after.
+Eqs.~(\ref{eq:F_decomNonPhysic1_app}-\ref{eq:Fold_def_app})  allow preventing these non-sense while keeping the formulation linear.
+Eq.~(\ref{eq:F_decomNonPhysic1_app}) forces the decommissioned capacity to zero when technology will be built after.
 To do so, a parameter (\emph{decom\textsubscript{allowed}}) is defined \emph{a priori} and is equal to 0 or 1 when decommissioning is not possible or possible, respectively. 
-Eq.~(\ref{eq:Fold_def}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}.
+Eq.~(\ref{eq:Fold_def_app}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}.
 For each phase, a set (\emph{AGE}) is calculated \emph{a priori}.
@@
 In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built.
-The first part of Eq.~(\ref{eq:Fold_def}) indicates that the technology is still available, and thus no capacity needs to be removed.
+The first part of Eq.~(\ref{eq:Fold_def_app}) indicates that the technology is still available, and thus no capacity needs to be removed.
 The second part of the equation represents the capacity that reached its expected lifetime minus a part of the capacity that would have been decommissioned.
 As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
-Ones will highlight the use of a `if' in Eq. (\ref{eq:Fold_def}), this formulation is linear as the if is applied to a parameter and not a variable.
+Ones will highlight the use of a `if' in Eq. (\ref{eq:Fold_def_app}), this formulation is linear as the if is applied to a parameter and not a variable.
 
@@
 \begin{flalign} 
- \label{eq:Phase2020Design}
+ \label{eq:Phase2020Design_app}
 & \textbf{F\textsubscript{new}}(\emph{2015\_2020},i) = \textbf{F}(\emph{YEAR\_2020},i)& \forall i \in \emph{TECH}
@@
 To initialise the problem in 2020 with the existing design, an additional phase `\emph{2015\_2020}' is created.
-Eq.~(\ref{eq:Phase2020Design}) requires that the capacity used in 2020 is installed in the previous phase.
+Eq.~(\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase.
 
@@
 \begin{flalign} 
-\label{eq:delta_tech_during_phase}
+\label{eq:delta_tech_during_phase_app}
 & \textbf{$\Delta$\textsubscript{change}}(p,i) \geq 
@@
  \begin{flalign}
-\label{eq:limit_reno_LTheat}
+\label{eq:limit_reno_LTheat_app}
 &\sum_{\mathclap{\hspace{2cm}i\in TECH(HeatLowT)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
@@
 &\hspace{5.5cm} \forall p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p) & \\
-\label{eq:limit_reno_passmob}
+\label{eq:limit_reno_passmob_app}
 &\sum_{\mathclap{\hspace{2cm}i\in TECH(MobPass)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
@@
 &\hspace{5.5cm} \forall p \in \emph{PHASE}, \emph{y\textsubscript{start}} \in \emph{Y\_START}(p) & \\
-\label{eq:limit_reno_freight}
+\label{eq:limit_reno_freight_app}
 &\sum_{\mathclap{\hspace{2cm}i\in TECH(MobFreight)}} \textbf{$\Delta$\textsubscript{change}}(p,i) 
@@
 
-Eq.~(\ref{eq:delta_tech_during_phase}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
+Eq.~(\ref{eq:delta_tech_during_phase_app}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
 %% GL_correction_v2 %% Following line added:
@@
 %% GL_correction_v2 %% end
-Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Eq.~(\ref{eq:limit_reno_LTheat}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Eq.~(\ref{eq:limit_reno_passmob}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Eq.~(\ref{eq:limit_reno_freight}).
+Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Eq.~(\ref{eq:limit_reno_LTheat_app}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Eq.~(\ref{eq:limit_reno_passmob_app}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Eq.~(\ref{eq:limit_reno_freight_app}).
 For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase.
@@
 % Objective function + investment 
-\label{eq:obj_func_v2}%1
+\label{eq:obj_func_v2_app}%1
 % adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
 \hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
-\label{eq:Capex_v2}
+\label{eq:Capex_v2_app}
 & \textbf{C\textsubscript{tot,capex}} =
@@
 \textbf{C\textsubscript{inv,return}}(i)\\
-  \label{eq:Copex_tot_v2}%5
+  \label{eq:Copex_tot_v2_app}%5
 & \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
@@
  \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
-\label{eq:path_annu_factor}
+\label{eq:path_annu_factor_app}
 & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
@@
 %The objective is the minimisation of the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}),
-defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2}).
-The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2}), to which the residual asset investment cost in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
+defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2_app}).
+The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2_app}), to which the residual asset investment cost in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
 Thus, the investments account for the installation and dismantlement costs of the technologies.
-The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Eq.~(\ref{eq:Copex_tot_v2}).
+The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Eq.~(\ref{eq:Copex_tot_v2_app}).
 During a phase, the system \gls{OPEX} is
-the product of the annualised phase factor, defined in Eq.~(\ref{eq:path_annu_factor}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
+the product of the annualised phase factor, defined in Eq.~(\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
 The annualised phase factor is defined based on an average interest rate during the transition.
-%assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor}).
+%assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor_app}).
 
@@
 \begin{flalign} 
-\label{eq:opex_yearly}
+\label{eq:opex_yearly_app}
 &\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{i \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,i) + \sum_{\mathclap{j \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,j)&\forall y\in \emph{YEARS}
@@
 
-For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Eq.~(\ref{eq:opex_yearly}).
+For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Eq.~(\ref{eq:opex_yearly_app}).
 
@@
 \begin{flalign} 
-\label{eq:PhaseInv}%5
+\label{eq:PhaseInv_app}%5
 &\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
@@
 
-The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Eq.~(\ref{eq:PhaseInv}).
-The annualised phase factor  (defined by Eq.~(\ref{eq:path_annu_factor})) is used.
+The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Eq.~(\ref{eq:PhaseInv_app}).
+The annualised phase factor  (defined by Eq.~(\ref{eq:path_annu_factor_app})) is used.
 The specific cost during the phase is defined as  the average between the investment cost for the first and last year of the period.
@@
 \begin{flalign} 
-\label{eq:salvage}%5
+\label{eq:salvage_app}%5
 &\textbf{C\textsubscript{inv,return}}(i) = \hspace{2.5cm}
@@
 Eq.
-(\ref{eq:salvage}) computes, for each technology, the residual value that must be deducted from the total cost.
+(\ref{eq:salvage_app}) computes, for each technology, the residual value that must be deducted from the total cost.
 The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised.
@@
 \begin{flalign} 
-\label{eq:gwp_tot_transition}
+\label{eq:gwp_tot_transition_app}
 &\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}/2 \left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)&
 \\
-\label{eq:limit_gwp_trans}
+\label{eq:limit_gwp_trans_app}
 & \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
@@
 
-The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Eq.~(\ref{eq:gwp_tot_transition}).
+The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Eq.~(\ref{eq:gwp_tot_transition_app}).
 The emissions during a phase is estimated as the arithmetic average of the representative years before and after the phase.
-Eq.~(\ref{eq:limit_gwp_trans}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).
+Eq.~(\ref{eq:limit_gwp_trans_app}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).

### 2024-01-25 appendices/app_case_study.tex
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom

@@
 & $c_{\mathrm{op,elec}}$ & Purchase electricity & II & -64.3\% & 179.8\% \\
-& $\bm{c_{\mathrm{op,electrofuels}}}$ & \textbf{Purchase electrofuels} & \textbf{II} & \textbf{-64.3\%} & \textbf{179.8\%} \\
+& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
 & $c_{\mathrm{op,biofuels}}$ & Purchase biofuels & II & -64.3\% & 179.8\% \\
@@
 & $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
-& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
+& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
 \midrule
@@
 & $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
-& $\bm{f_{\mathrm{max,SMR}}}$ & \textbf{Potential capacity \gls{SMR}} & \textbf{-} & \textbf{0} & \textbf{1} \\
+& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
 

### 2024-01-25 chap_atom_mol/chap_atom_mol.tex
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom

\section*{Contributions}
\label{sec:atom_mol:contributions}
\begin{itemize}
\item Pathway UQ
\item Contrast analysis between atom and molecules and local renewables
\item Influence of the parameters on the import of molecules and the installation of SMR
\end{itemize}

\textbf{Add here references and analyses of nuclear and/or molecules as THE solution} \cite{IEA_Nuclear_2022}

The model and the frameworks have been introduced in \cref{sec:methodo} and the case study presented in \cref{sec:case_study}. This section focuses on the results in two ways. First, Section \ref{subsec:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters). Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{subsec:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.

\section{Deterministic impact of integrating \gls{SMR} in 2040}
\label{sec:atom_mol:results_deter} 
In this section, like in the rest of the paper, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW, from 2040 onward. After investigating the deployment of \gls{SMR} through the power sector, the first part of this section focuses on this impact on macro/system-level considerations (\ie overall transition costs, primary energy mix and yearly emissions per sector). The second part will address the impact of \gls{SMR} on each of the other sectors of the system.

\subsection{Power sector}
\label{subsec:atom_mol:results_deter_power_sector}
\Cref{fig:results_deter_tech_cap_elec} shows that \gls{SMR} is deployed as soon as available, \ie 2040, to their maximum capacity, \ie 6\,GW, substituting other flexible power generation units: no ammonia-\gls{CCGT} at the end of the transition and the anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case). To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case. Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.
\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_REF.pdf}
\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_SMR.pdf}
\caption{As soon as available (\ie 2040), \acrfull{SMR} is deployed to their maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie gas and ammonia \gls{CCGT}).}
\label{fig:results_deter_tech_cap_elec}
\end{figure}

When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity, given the 15\% maintenance off-time assumed in this work: 44.6\,TWh. By 2050, it represents 24.6\% of the total electricity production which is less than the current share of conventional nuclear in Belgium, 38.5\%. This resurgence of nuclear electricity occurs at the expense of other, although more efficient technologies: \gls{CCGT} and industrial \gls{CHP}. Besides the unchanged end-use-demand, we observe a slight increase of the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{Elec_Layer_REF.pdf}
\includegraphics[width=0.8\textwidth]{Elec_Layer_SMR.pdf}
\caption{The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
\label{fig:results_deter_layer_elec}
\end{figure}

\subsection{System-level impacts}
\label{subsec:atom_mol:results_deter_overall}
First of all, as far as the objective function (\ie the total transition cost) is concerned, \Cref{fig:results_deter_overall_emissions_sector} shows that the 6\,GW \gls{SMR} installed from 2040 allow reaching a 3.3\% (-36.9\,b€, $\sim$\,6\% of the Belgian GDP) cheaper overall transition. Interestingly, as the model can freely spend the constrained \ce{CO2}-budget over the transition, knowing ahead (\ie perfect foresight) that cheaper and low-emitting \gls{SMR} will be available in the future, cost-savings, that are more important after 2040, also occur before 2040. This is mostly due to the extended use of cheaper \gls{LFO} at the beginning of the transition to produce \gls{HVC}, that is then compensated by the deployment of \gls{SMR} (\Cref{fig:results_deter_overall_emissions_sector}). Then, the capital-intensive investments in \gls{SMR}, mostly recovered by the end of the transition as salvage value, are widely compensated by the smaller resource-related OPEX. This leads, at the end, aggregating the OPEX and the annualised CAPEX, to a system that is yearly 8.8\% less expensive by 2050.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{Cum_total_cost_diff_REF.pdf}
\includegraphics[width=0.8\textwidth]{GWP_per_sector_diff.pdf}
\caption{(Left) Cumulative transition cost difference between the SMR and the REF cases. Including \gls{SMR} ends up in a cheaper overall transition (\ie -3.3\%, -36.9\,b€). (Right) Emissions per energy sector differerence between the SMR and the REF cases. The deployment of \gls{SMR} from 2040 onward allows maintaining longer the use of \acrfull{LFO} for the production of \acrfull{HVC}.}
\label{fig:results_deter_overall_emissions_sector}
\end{figure}

Then, considering the primary energy mix shown in \Cref{fig:results_deter_energy_mix}, three phases in the transition can be identified. Before 2040, thanks to the perfect foresight approach, the model finds it more economical in 2025 to keep on using 33.2\,TWh of \gls{LFO} to produce \gls{HVC} through naphtha/LPG-cracking. In 2040, uranium-driven \gls{SMR} substitutes the electricity originally produced from industrial \gls{CHP} and \gls{CCGT} running respectively on fossil gas and renewable ammonia. Finally, from 2045 onward, the significant drop in the consumption of electrofuels comes from the same industrial \gls{CHP}. This is the illustration of the atom-molecules dilemma where the consumption of local renewables is, on its side, not much affected. In other words, \gls{SMR} competes with importing electrofuels while both support the integration of local solar and wind energies. Then, like the power sector, the SMR case ends up in a less efficient whole-energy system by 2050 at it consumes 47\,TWh (+12.7\%) primary energy more to supply an unchanged \gls{EUD}. Interestingly, in both cases, given the assumptions made on the \gls{GWP} of the resources (\ie $\mathit{gwp}_{\mathrm{op,electrofuels}}=0$\,kt$_{\ce{CO2},\text{eq}}$/GWh and $\mathit{gwp}_{\mathrm{op,uranium}}=0.004$\,kt$_{\ce{CO2},\text{eq}}$/GWh), the constraint on the \ce{CO2}-budget leads to ``carbon-neutrality'' by 2050\footnote{The model being constrained to keep on using all the waste that would keep on being locally produced, the system in 2050 reaches a $\sim 3.5\,Mt_{\ce{CO2},\text{eq}}$/year.}.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{Primary_mix_REF.pdf}
\includegraphics[width=0.8\textwidth]{Primary_mix_SMR.pdf}
\includegraphics[width=0.8\textwidth]{Primary_mix_diff_uranium.pdf}
\caption{(Up) Primary energy mix over the transition for the REF and SMR cases. (Bottom) Difference of the mix between the two cases after aggregation per category. In Uranium, considered as a non-renewable resource in line with \citet{rixhon2021terminology}, is nevertheless dissociated from other non-renewable resources for the sake of clarity. The imported electricity is split between imported non-renewable and renewable resources assuming a linear decrease between the current share of renewables, \ie 37.41\% \cite{eurostat_share_re_elec}, and an assumed 100\%-renewable European electricity mix by 2050.}
\label{fig:results_deter_energy_mix}
\end{figure}

\subsection{Non-power sectors}
\label{subsec:atom_mol:results_deter_others}
\paragraph*{High-temperature heat}
As aforementioned, the main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of overall more efficient heat-and-power co-generation in the benefit of single-output industrial boilers. On the one hand, in the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production" of the 59\,GW solar-PV when fully deployed, in the sunny days. With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity, consequently increasing their load factor and yearly production respectively by 31\% and 48\%. On the other hand, given the 44.6\,TWh of electricity produced by \gls{SMR}, cogeneration units are less relevant and, by 2050, 2.6\,GW industrial gas boilers completely substitute \gls{CHP} to produce 16.6\,TWh (\ie 23\%) of the total production of high-temperature heat.

\paragraph*{Low-temperature heat}
This sector is marginally impacted. In both cases, the major shift of supply from decentralised to centralised productions operates early in the transition, to hit the constraint that \gls{DHN} cannot supply more than 37\% of the \gls{LT}-heat production. Then, from a mix between oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system progressively shifts towards electric \gls{HP} only. Similarly for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economic option.

\paragraph*{Mobility}
The passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} progressively substituting \gls{ICE} cars for the private sector. About the public mobility, trains and tramways supply their \textit{a priori} set maximum share, respectively, 50\% and 30\% complemented by \gls{CNG} buses substituting diesel-driven buses. Similarly, considering the freight transport, technology shifts (\ie from diesel to \gls{FC} trucks) or modal shares (\ie 53\%-47\% split between \gls{NG} and (bio)-diesel boats) are identical between the two cases.

\paragraph*{Non-energy demand}
The supply of ammonia (\ie from Haber-Bosch to direct import of renewable ammonia) and methanol (\ie import of renewable methanol) are unchanged between the two cases. However, as introduced previously, to produce \gls{HVC}, the full substitution of naphtha/LPG-cracking by \gls{MTO} is delayed as the emissions of the former are compensated by the later integration of \gls{SMR}.

\section{Uncertainty quantification on the cost, the atom and the molecules}
\label{sec:atom_mol:results_uq}
``\textit{It is difficult to make predictions, especially about the future.}" (Niels Bohr, foundational contributor to the understanding of atomic structure and quantum theory). Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters. To mitigate the computational burden, we have used the \gls{PCE} approach (\Cref{subsec:meth:UQ}) on the monthly version of EnergyScope Pathway. As detailed by \citet{limpens2023pathway}, this simplified model does not implement daily storage technologies and integrates much more easily intermittent renewables (PVs and wind turbines) in the system as the daily supply-demand intermittency and mismatch are not at stake. Consequently, it requires less flexibility options (\ie electrification of the heating sector or import of electricity from the neighbouring countries) and ends up with a cheaper transition (-3\%). Although, it allows to quantify the full set of 34 uncertain parameters (\ie 1260 samples for an order-2 \gls{PCE}), and avoid the computational burden of the hourly pathway. Nevertheless, comparing the \gls{UQ} on the hourly pathway with a limited set of uncertain parameters and on the monthly model, Appendix \ref{app:UQ_transition_cost} shows that the most impacting parameters are correctly captured by the latter.

After briefly assessing the \gls{GSA} on the objective function of the model, the total transition cost, this section investigates more deeply the atom-molecules dilemma. This time, the Sobol' indices are computed for import of renewable molecules and installed capacities of \gls{SMR}.

\subsection{Total transition cost}
\label{subsec:atom_mol:results_uq_cost}
Exhaustively listed in Appendix \ref{app:UQ_transition_cost}, \Cref{tab:UQ_short} gathers the most impacting parameters\footnote{Per \citet{Turati2017}, parameters are considered as ``impacting'' if their Sobol' index is above the threshold $=1/d$, $d$ being the total number of uncertain parameters after the pre-selection phase. In this case, $d=34$, and, consequently, the threshold is equal to 2.9\%.} on the total transition cost, highlighting the cost of purchasing electrofuels as well as the potentiality to install \gls{SMR} and its CAPEX. The former is the most impacting parameter whereas the two others have much lower influence on the variation of the total transition cost. Given the uncertainty characterisation presented in Section \ref{subsubsec:UQ:UC}, there are 60\% chance that no \gls{SMR} could be installed. In other words, the variation of the parameter $f_{\mathrm{max,SMR}}$ has zero impact on the variation of the total transition cost in 60\% of the samples. Then, in perspective with the local sensitivity analysis of Section \ref{subsec:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward. This represents only 10\% of the samples. Moreover, given its characteristics detailed in \Cref{tab:SMR_features}, mostly its cheap and low-emitting fuel (\ie uranium) and the long lifetime leading to lower annualised CAPEX and higher salvage value, this explains why \gls{SMR} has this lower impact. On the contrary, more expensive renewable electrofuels are always imported, to a smaller or larger extent depending on the sample. For instance, in the REF case (Section \ref{subsubsec:results_deter_overall}), the imported electrofuels represent, by 2050, 152.9\,TWh (\ie 41\% of the primary energy mix) with an average 93€/MWh cost of purchasing and, over the entire transition, a 273\,b€ cumulative OPEX (\ie 25\% of the total transition cost).

\begin{table}[htbp!]
\caption{Total Sobol' indices of the uncertain parameters over the total transition cost. Where the cost of purchasing electrofuels is the top-1 parameter, \gls{SMR}-related parameters have a negligigble impact on this cost.}
\label{tab:UQ_short}
\centering
\begin{tabular}{l c c}
\toprule
\textbf{Parameter}  & \textbf{Ranking} & \textbf{Sobol' index} \\	
\midrule
\textbf{Purchase electrofuels} & 1 & 47.4\% \\
Industry EUD & 2 & 23.5\% \\
Interest rate & 3 & 11.0\% \\
Purchase fossil fuels  & 4 & 6.9\% \\
$\vdots$ & $\vdots$ & $\vdots$\\
\textbf{Potential capacity \gls{SMR}} & 11 & 0.9\% \\
$\vdots$ & $\vdots$ & $\vdots$\\
\textbf{CAPEX \gls{SMR}} & 32 & $<$0.1\% \\
\bottomrule

\end{tabular}
\end{table}

Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter. Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 11\% Sobol' index\footnote{It is important to note here that the model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value). In practice, the interest rate would vary depending on the technology investment risk. This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2019}.}. Finally, similarly to electrofuels, the cost of purchasing fossil fuels is also to consider in the perspective to reduce the uncertainty over the total transition cost. However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their uncertain impact smaller than their renewable alternatives.

\subsection{Atom and molecules}
\label{subsec:atom_mol:results_atom_mol}
The samples used to carry out the \gls{GSA} on the total transition cost, also provide the distribution of other outputs of the model. Among them, \Cref{fig:results_uq_electrofuels} shows the evolution of the import of renewable electrofuels over the transition.

As the general trends are increasing, discrepancies exist between the different energy carriers. E-methane, as the renewable alternative to fossil natural gas, substitutes it, sometimes at a very early stage of the transition, 2025, and to a somehow unrealistically large extent, 173\,TWh, which is more than 13\% more than the total import of electrofuels in the REF case. The necessity to import this molecule is progressive through the transition to supply mostly industrial \gls{CHP} and boilers.

E-hydrogen becomes rapidly the main stream of hydrogen in the system, on top of steam-methane-reforming or electrolysis, to reach a median and a maximum values of, respectively 13.6\,TWh and 40.7\,TWh in 2050. Hydrogen is more frequently used in the mobility sector. Like in the REF case, fuel cell trucks are often the first option but, in some outlying cases, fuel cell cars and buses appear to completely substitute respectively \gls{BEV} cars and \gls{CNG} buses by 2050. Moreover, some samples lead to local production of methanol via the methanolation process, to produce up to 16.3\,TWh of methanol (\ie 30\% of the total supply of methanol of the nominal REF and SMR cases).

Then, the imported e-ammonia, becoming rapidly cost-competitive against its fossil alternative (\Cref{fig:cs_resources_cost}), quickly substitutes it and the Haber-Bosch process. Where the initial purpose of ammonia is to satisfy a relatively limited \acrfull{NED} (\ie 10 $\pm$ 3\,TWh by 2050), the variation of its import is mostly due to the higher or lower need for ammonia-\gls{CCGT} as a flexible option to produce electricity. From 2035, out of the four considered electrofuels, the imported e-ammonia is the one exhibiting the largest uncertainty with, for instance, an interquartile range (IQR)\footnote{The interquartile range is the difference between the third quartile ($Q3$ or $P_{75}$) and the first ($Q1$ or $P_{25}$). It is an indicator of statistical dispersion around the median, $Q2$ or $P_{50}$.} of about 50\,TWh. In some extreme cases, e-ammonia is the most imported molecules, \ie up to 233\,TWh or 63\% of the total primary mix in the REF case in 2050.

Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol or synthetic methanolation exist in some outlying cases. Given its lower \gls{NED} (\ie 1.5$\pm$0.5\,TWh$_{\text{NED,methanol}}$ by 2050), the variation of imported e-methanol is almost exclusively due to its role in the industrial production of \gls{HVC}, \ie 78\% of the total \gls{NED}, through the \acrfull{MTO} process. In some rare samples, methanol is also used to supply the freight transport sector via boats or trucks.

Appendix \ref{app:UQ_electrofuels} gives a more detailed information. On the one hand, it compares the statistics (\ie quartiles and median) with the quantity of imported electrofuels in the REF and SMR cases. On the other hand, this appendix shows the distributions of the different sources of supply and consumption of gas, hydrogen, ammonia and methanol.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{UQ_Electrofuels.pdf}
\includegraphics[height=4cm]{Schematic_boxplot.pdf}
\caption{Distribution of the imported renewable electrofuels over the transition. Starting from no electrofuel in 2020, their respective import rises progressively along the transition (\ie increasing median) at different growth rates and with different ranges of values. Observations being either 1.5 times the interquartile range (IQR) less than the first quartile (Q1) or 1.5 times the interquartile range greater than the third quartile (Q3) are defined as outliers.}
\label{fig:results_uq_electrofuels}
\end{figure}

After investigating the distribution of imports of electrofuels over the transition, this part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe. Figures \ref{fig:results_uq_samples_A} and \ref{fig:results_uq_samples_B} give the trend lines of the key parameters for these imports in 2050\footnote{The authors picked this specific year as it is the one where electrofuels, if imported, are imported in the largest amount, in general, compared to the other years of the transiton.}, as well the installed capacities of \gls{SMR}. Next to the name of a parameter, one can read its Sobol' index versus the output of interest\footnote{For these outputs of interest, different from the total transition cost, the \gls{LOO} error is generally higher than the threshold of \SI{1}{\%} defined in Section \ref{subsubsec:UQ:PCE}. Consequently, the Sobol' indices are less accurate but already allow a fair relative comparison between the different parameters.}. Box plots also point out the distribution of this output for the extreme low or high values of some parameters.

As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane. This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production. Then, considering the smaller-impact parameters, we notice that \gls{SMR} plays a non-negligible role. Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives. This confirms the observation made in Section \ref{subsubsec:results_deter_others}. Highly available local biomass also leads to smaller import of e-methane to supply bio-hydrolysis and produce methane-equivalent gas. Finally, and surprisingly, costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively. In other words, by 2050, more expensive electrofuels induce to more e-methane import and \textit{vice versa} for the fossil fuels. Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}. Subject to the \ce{CO2}-budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity. First running on fossil natural gas, these \gls{CHP} consume more e-methane by 2050. In the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia. This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050. Consequently, there is smaller investment in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers. Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2}-budget allows it. About the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system. If \gls{NG} is more expensive, the system will import less of it. Subsequently, the investments in methane-\gls{CHP} and boilers are more limited. This ends up in smaller need for e-methane by 2050.

In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector. As depicted in \Cref{fig:results_uq_prod_cons} in Appendix \ref{app:UQ_electrofuels}, the utilisation of e-hydrogen is most prevalent in \gls{FC} trucks, followed by \gls{FC} cars and buses to a lesser extent. The adoption of fuel cell engines in trucks contributes, on average, to 63.5\% of the total road freight transport, thereby affecting the level of e-hydrogen imports. Consequently, the smaller is the CAPEX of fuel cell engines, the more the system imports e-hydrogen. Similarly, the cost of purchasing electrofuels influences e-hydrogen imports. Subsequently, the cost of purchasing biofuels emerges as the third most influential parameter. Indeed, biodiesel trucks are the mostly picked alternative to \gls{FC} trucks to provide, on average, 27.6\% of the total. Additionally, \gls{CNG} buses are preferred in public road transport (34.9\%), followed by \gls{FC} buses (11.2\%) competing with biodiesel and hybrid biodiesel buses, accounting for 27.8\% and 26.1\%, respectively. Finally, the last noticeable parameter at stake is the CAPEX of electric vehicles. In competition with \gls{BEV} that stand for 83.4\% on average of the private mobility sector, the cheaper these cars are, the more cost-competitive are these vehicles, and vice versa, versus \gls{FC} cars (\ie 13.7\% of the total passenger mobility, on average).

As already pointed out in Section \ref{subsubsec:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia. As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheap electricity flexibly produced by \gls{SMR} substitutes the \gls{CCGT}. With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 7.3\,TWh, 83.4\% less than in the REF case. Then, with a 9\%-Sobol' index, industrial \gls{EUD} also influences the need for this molecule, due to its \gls{NED}.

The conclusions are more straightforward for the import of e-methanol and the installed capacity of \gls{SMR}. For the former, industrial \gls{EUD} is, by far (\ie 81\% Sobol' index), the key factor. Due to its own \gls{NED} but, above all, since it is the low-emitting alternative picked by the model to supply the significant \gls{NED} of \gls{HVC}, the lower this demand, the lower the need to import e-methanol, and vice versa. For the latter, it is the availability of the technology that drives its installation. Not shown here but all the samples of the \gls{GSA} highlight that \gls{SMR} is installed to its maximum capacity, \ie 6\,GW, as soon as possible. In other words, the only parameter ``Potential capacity \gls{SMR}'' dictates the installation of this technology\footnote{In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from, respectively, 2040, 2045 and 2050.}. Surprisingly, the [-40\%; +44\%]  variation of its CAPEX has a negligible impact, with a Sobol' index of 0.9\%.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
\includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}

\caption{Trend lines of the key parameters on the import of e-methane, e-hydrogen and e-ammonia. Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters. The grey dashed line gives the value of the output of interest in the REF case. Part A}
\label{fig:results_uq_samples_A}
\end{figure}

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
\includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
\caption{Trend lines of the key parameters on the import of e-methanol and the installed capacity of \gls{SMR}, in 2050. Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters. The grey dashed line gives the value of the output of interest in the REF case. Part B}
\label{fig:results_uq_samples_B}
\end{figure}

\subsection{Local renewables}
\label{subsec:atom_mol:results_uq_VRES}
In line with results given in Section \ref{subsec:results_deter}, the \gls{GSA} shows that \gls{SMR} has negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines). \Cref{fig:results_uq_pdf_local_ren} gives the evolution of Sobol' indices for the most impacting parameters on the deployment of \gls{PV} and offshore wind\footnote{As the installed capacities of onshore wind is totally driven by the uncertainty on its maximum potential, $f_{\mathrm{max,windon}}$, it is not represented in the figure.}. The key factor that drives the installed capacities of these two technologies is mostly their respective maximum potential, especially at the end of the transition, much more than their CAPEX. Given its higher \gls{LCOE} (\Cref{fig:LCOE}), \gls{PV} is more impacted in the short-term by the variation on the cost of purchasing electrofuels supplying e-methane (and e-ammonia to a lesser extent) \gls{CCGT}. However, this impact gets negligible by 2050.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.49\textwidth]{Sobol_PV.pdf}
\includegraphics[width=0.49\textwidth]{Sobol_WIND_OFFSHORE.pdf}
\caption{Impacting parameters on the deployment of \gls{PV} and offshore wind over the transition. Progressively, the impact of the uncertainty on the maximum potential increases, unlike the one on the cost of purchasing electrofuels.}
\label{fig:results_uq_pdf_local_ren}
\end{figure}

### 2024-01-25 chap_methodo/chap_methodo.tex
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom

@@
 
-\section{Contributions}
+\section*{Contributions}
 \label{sec:meth:contributions}
 \begin{itemize}
+\item Develop the myopic approach
 \item Apply Stefano's method on the pathway model with a similar approach as Guevara et al.
 \item Check that PCE was appropriate as a method for such a system (ECOS2020)
-\item Work on the $C_inv_return$
+\item Update the $C_inv_return$ based on the initial investigation of \citet{goffauxpathway} and in line with \citet{poncelet2016myopic}.
 \end{itemize}
@@
 
-On the contrary, this work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}.
+This work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}.
 According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is open-source documented model.
 Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation).
-From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the model.
+From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the former approach to further dig into more details about the latter.
 The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from a snapshot approach, EnergyScope TD.
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
+
+\subsection{Perfect foresight: One global optimisation of the transition}
+\label{subsec:meth:PF}
 
@@
 
-We used the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core})---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050.
-Even though this lacks realism compared to the myopic approach \cite{babrowski2014reducing,fais2016impact,heuberger2018impact} where the pathway is optimised via a sequence of shorter time windows, the perfect foresight approach allows to optimise the transition with a \ce{CO2}-budget (Section \ref{subsec:cs:CO2-budget}) rather than along a prescribed \ce{CO2}-trajectory.
-However, the work of \citet{limpens2024pathway} has shown that PF provides similar results for the study case compared to the myopic approach with a prescribed linear decrease of the \ce{CO2}-emissions .
-All the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}, are kept as is with an extra-dimension to relate them to a specific representative year, $y$, of the pathway.
+The whole-energy system model developed in this work originates from the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core}) of EnergyScope---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050 \cite{limpens2024pathway}.
+All the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}, and summarized in Appendix \ref{app:ESTD}, are kept as is with an extra-dimension to relate them to a specific representative year, $y$, of the pathway.
 For instance, the energy balance is guaranteed at every hour of each of these years.
@@
 \abovedisplayskip=2pt
+
 \begin{flalign} 
@@
 
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:salvage}%5
+&\textbf{C\textsubscript{inv,return}}(i) = \hspace{2.5cm}
+\sum_{\mathclap{p \in \emph{PHASE}\cup \{2015\_2020\}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}} 
+\hspace{0.5cm}
+ \tau_{phase}(p)\cdot \left(c_{inv}(y_{start},i)+c_{inv}(y_{stop},i)\right)/2 \cdot
+&\notag\nonumber
+\end{flalign}
+\begin{flalign}
+& 
+\hspace{1.7cm}
+\frac{remaining\_years(i,p)}{lifetime(y\textsubscript{start},i)} \left( \textbf{F\textsubscript{new}}(p,i) - 
+\sum_{\mathclap{p2 \in \emph{PHASE}}} 
+\textbf{F\textsubscript{decom}}(p2,p,i)\right)&
+\forall i \in \emph{TECH}
+\end{flalign}
+\endgroup
+
 \noindent
@@
 This last variable is crucial to avoid penalising heavy (and potentially long-lifetime) investments at the end of the transition as these assets would still be operational beyond 2050.
-The interested reader will find more information about its implementation and the formulation choices related to it in the work of \citet{limpens2024pathway}.
+The interested reader will find more information about the formulation choices related to it in the work of \citet{limpens2024pathway}.
 The other variables in Eq.\,(\ref{eq:CAPEX_v2}-\ref{eq:Copex_tot_v2}) are detailed here below:
@@
 \begin{flalign} 
+\hspace{0pt} 
 \label{eq:opex_yearly}
-&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i)&\forall y\in \emph{YEARS}\\
+&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i) \hspace{2.9cm} \forall y \in \text{\emph{YEARS}}\\
 \label{eq:PhaseInv}%5
@@
 
-Then, as detailed in section \ref{subsec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
+Then, as detailed in section \ref{sec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
 
@@
 
+\subsection{Myopic: Sequential optimisation of the transition with limited foresight}
+\label{subsec:meth:MY}
+
+One of the main methodological contributions of this work regarding the development of the whole-energy system model consists in giving it the possibility to optimise the transition pathway in a myopic approach.After introducing the general concept of it, this section details more the additions brought to the model in terms of implementation.
+
+\vspace{0.1cm}\textbf{General concept of the myopic optimisation}\vspace{-0.3cm}\\
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{MY_schematic.pdf}
+\caption{The myopic approach (in pink) uses several instances of the pathway model (illustrated in Figure \ref{fig:meth_path_methodology}).
+In this example, the pathway instance has a time horizon of 10 years ($N\textsubscript{year,opti}=10$) with a 5 year-overlap ($N\textsubscript{year,overlap}=5$).
+As a comparison the Perfect foresight (in blue) has a time horizon of 30 years.}
+\label{fig:my_schematic}
+\end{figure}
+
+Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has two main advantages: shorter computational time and more realistic representation of the short-sightedness of decision-makers.
+For this reason, several studies are based on this approach \citep{babrowski2014reducing,poncelet2016myopic,nerini2017myopic,heuberger2018impact}.
+\citet{babrowski2014reducing} analysed the benefit of the myopic approach to reduce the computational time.
+\citet{poncelet2016myopic} uses this approach to analyse the expansion planning of the power sector beyond 2050.
+\citet{nerini2017myopic} analysed the impact of the horizon windows and overlapping time.
+Overall these studies decided to choose the myopic approach to analyse the speed of change compared to a perfect foresight approach.
+ % rajout du Réalisme 
+Moreover, the myopic approach allows a sequential optimisation process that opens the doors to decision-making/policy-learning methodologies, like assessing shock events.
+This approach is used by \citet{heuberger2018impact} who assessed the speed of integration of technologies due to these events.
+In their analysis of the overcapacity in European power systems, \citet{moret2020overcapacity} emphasised that such a ``possibility of \textit{recourse}" is very appropriate to address uncertainty gradually unfolding over time.
+Consequently, the development of the myopic approach represents the foundations of the further implementation of the agent-based reinforcement learning framework (see Section \ref{sec:meth:RL}).\\
+
+As illustrated in Figure \ref{fig:my_schematic_2}, after optimising, in design and operation, one time window (\eg from 2020 to 2030), the intermediate system design (\ie the installed capacities) is set as initial conditions for the start of the next time window (\eg from 2025 to 2035) as well as the historical investment decisions (\ie $\textbf{F\textsubscript{new}}$, $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$).
+Consequently, the solution obtained at the end of the first time window (\eg 2030) as well as potential investment decisions between the start of the second time window and this end-year are discarded.
+In other words, they are not taken into account for the optimisation of the second time window.
+This process goes on until the stated end of the transition (\ie 2050, in this case).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{MY_schematic_2.pdf}
+\caption{Sequential optimisation of the transition pathway in the myopic approach: (i) first time-window optimisation, (ii) set-up of the initial conditions of the second time-window, (iii) second time-window optimisation discarding intermediate results}
+\label{fig:my_schematic_2}
+\end{figure}
+
+\vspace{0.1cm}\textbf{Additional sets, parameters and variables}\vspace{-0.3cm}\\
+
+The major add-on from the original EnergyScope Pathway model \cite{limpens2021generating} to the myopic version developed in this thesis, is the possibility to carry out the optimisation on a limited time window, of which the duration is defined by $N\textsubscript{year,opti}$.
+Moreover, to ensure the ``possibility of \textit{recourse}", there is also the possibility of having an overlap between two consecutive time windows.
+The timespan of this overlap is defined by the parameter $N\textsubscript{year,overlap}$.
+The philosophy followed behind the development of the myopic approach was to add another layer on top of the perfect foresight model in order to make it more modular.
+For this reason, the already existing constraints are marginally adapted.
+This way, the newly developed model can easily be used to perform a perfect foresight optimisation by setting the time window to $N\textsubscript{year,opti}=$30 years (\ie between 2020 and 2050) and the overlap between the time windows to $N\textsubscript{year,overlap}=$0.
+Consequently, as defining the actual time window on which the system is optimised as well as the history, \ie what has already been optimised earlier in the transition, are fundamental, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
+
+\begin{table}[htbp]
+\caption[New SETs for myopic pathway formulation.]{New SETs for myopic pathway formulation.} 
+\label{tab:path_my_sets}
+\begin{minipage}{\textwidth}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{l c l}
+\toprule
+\textbf{Set}      & \textbf{Index}	 &	\textbf{Description}\\
+\midrule
+$\text{YEARS}\textsubscript{WND}$ 	&	$y\in$Y	& 	\pbox{20cm}{\vspace{1mm} Representative years of the \\ time window to optimize}\\
+$\text{YEARS}\textsubscript{UP TO}$ &	$y\in$Y	& 	\pbox{20cm}{\vspace{1mm} Representative years including the \\ years already optimised, \ie the history}\\
+$\text{PHASE}\textsubscript{WND}$ &  $\emph{p}\in$P & 	\pbox{20cm}{\vspace{1mm} Phases of the time window to optimize}\\
+$\text{PHASE}\textsubscript{UP TO}$ &  $\emph{p}\in$P & 	\pbox{20cm}{\vspace{1mm} Phases including the phases \\ already optimised, \ie the history}\\
+\bottomrule
+\end{tabular}}
+\end{minipage}
+\end{table}
+
+$\text{YEARS}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{WND}$ substitute $\text{YEARS}$ and $text{PHASE}$ in the constraints defined in the pathway model in Section \ref{subsec:meth:PF}.
+These two sets aim at setting the optimization to a more limited time window.
+Progressing through the transition, $\text{YEARS}\textsubscript{UP TO}$ and $\text{PHASE}\textsubscript{UP TO}$ allow keeping track of the history of the investments (\eg technologies installation, decommissioning or retirement), the consumption of resources, the cumulative amount of emissions, etc.
+
+On top of these four specific sets, some artefacts were also necessary to avoid computational rounding errors.
+Indeed, optimizing the first year of a time window that has already been optimized in the previous time window could lead to rounding errors preventing from the optimization to converge.
+For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimize that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
+This remark stays valid for any time window except the first one of the transition where the year 2020 is optimized even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
+Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Eq.
+\ref{eq:salvage}.\\
+
+\vspace{0.1cm}\textbf{Myopic pathway implementation}\vspace{-0.3cm}\\
+
+Starting this work in 2017, AMPL Optimization Inc.
+has developed a Python \gls{API} called amplpy \cite{amplpy}.
+In a nutshell, this API allows the pre/post-processing of an ampl optimisation problem by accessing its features (\eg constraints,parameters, variables, objective function) from within Python.
+Using this\gls{API}, this updated version of the model interacts with the AMPL problem representing the optimization of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=10cm]{MY_process_code.pdf}
+\caption{Schematic of the iterative optimization of the whole-energy system transition pathway.}
+\label{fig:MY_process_code}
+\end{figure}
+
 \section{Uncertainty quantification}
 \label{sec:meth:UQ}
+In their systematic review, \citet{yue2018review} highlighted that a wide majority of studies addressing the optimisation of energy systems (\ie 75\% out of the 134 reviewed studies) were not investigating the impact of uncertainties.
+However, disregarding these impacts can have drastic consequences on the system design.
+For instance, historical low \gls{NG} prices have led to overcapacity of \gls{CCGT} in Europe \cite{moret2020overcapacity}.
+This is why accounting for uncertainty in \gls{ESOMs} is crucial \cite{mavromatidis2018uncertainty}, especially when it comes to optimise several decades in an inherently uncertain future \cite{peace2008insights}.
+
+This section aims at briefly presenting the methods followed to first characterise these uncertainties, then to quantify their impact on different outputs of interest of the model (\eg amount of molecules imported from abroad, the installed capacity of \gls{SMR} or the total transition cost) and finally, the screening and selection of the parameters to analyse.
 
@@
 \label{subsec:uncert_charac}
-%Presentation here of the uncertainty characterisation from S.
-Moret and adding some parameters specific to the pathway model (\eg $\Delta_{\text{change}}$ linked to change speed) or specific to the main case study (\eg possibility to have nuclear SMR from 2040)
+Characterising precisely the uncertainty---ideally with their respective probability density functions (PDFs)---of the thousands of parameters in the model is daunting if not impossible because of lack of data \cite{marnay2006addressing}.
+Therefore, we used a workaround developed by \citet{Moret2017} that defines relative ranges of variation for different groups of parameters.
+These ranges have been adapted for the Belgian energy system and the pathway formulation.
+Moreover, some ranges have been added to account for new parameters coming from the pathway formulation described in Section \ref{sec:meth:ES} like the society inertia.
+Like other works \cite{li2019renewables,coppitters2021robust}, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+
+%\Cref{tab:UC_short} gives the uncertainty ranges of some key parameters.
+A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_short}.
+As detailed in \Cref{sec:cs:technologies}, the commercial availability of such a technology is uncertain but would not be before 2040.
+Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
+As SMRs are foreseen, if installed, to be around the same locations (\ie Tihange and Doel) as the conventional nuclear power plants and using the same area in kW/ha, the same 6\,GW are assumed to be the maximum capacity for SMRs.
+If it is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about the availability.
+Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
+
+%The exhaustive list of the parameters accounted in this work is presented in Appendix \ref{app:UC_full}.
+
+%\begin{table}[htbp!]
+%\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025.
+$^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
+$^{(b)}$ The nominal value of each parameter is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+$^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}
+%\label{tab:UC_short}
+%\centering
+%\resizebox{\textwidth}{!}{
+%\begin{tabular}{l l l c c c}
+%\toprule
+%\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}$^{(a)}$}  & \multicolumn{2}{c}{\textbf{Relative variation$^{(b)}$}}\\
+%    & & & &	 min 	&	 max \\ 	
+%\midrule		
+%\multirow{2}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+%& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
+%\midrule
+%\multirow{5}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+%& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}$^{(c)}$ & I & -40.0\% & 44.0\% \\
+%\midrule
+%\multirow{1}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
+%& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
+%& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
+%& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
+%\midrule
+%
+%\multirow{2}{*}{\textbf{End-use demand}} & $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
+%& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
+%\midrule
+%
+%\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+%& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
+%& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
+%& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
+%
+%\bottomrule							
+%
+%\end{tabular}}
+%\end{table}
+
+Following the methodology defined by \citet{Moret2017}, uncertainties of types I (investment-type) and II (operation-type, constant uncertainty over time) keep the same range for the whole transition.
+However, parameters with an uncertainty increasing over time, type III, (\ie end-use demands, in this case) will have a wider and wider range over the transition.
+In this work, a +50\% linear increase has been arbitrarily set between the width of the range of such parameters in 2025 and the same ranges in 2050.
+In \Cref{fig:ranges_transition}, this means that for type III uncertainties only, $R_{2050}^+$ is 50\% bigger than $R_{2025}^+$ and $R_{2050}^-$ is 50\% smaller than $R_{2025}^-$.
+For uncertainties of types I and II, the relative variation versus the nominal value remain the same over the transition.
+Inspired by \citet{guevara2022modeling}, the values of the uncertain parameters are set at a fixed relative position from the nominal values for each sampled transition---the values do not zigzag from 2025 to 2050 within the bounds (\Cref{fig:ranges_transition}).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.5\textwidth]{ranges_transition.pdf}
+\caption{$\mu_{2020}$, $\mu_{2025}$, ...,  $\mu_{2050}$ are the nominal values equal to 0 as the uncertain parameters represent a relative increase/decrease of actual parameters of the model.
+$R^+$ and $R^-$ are respectively the upper and lower bounds of the range and $\xi_{2025}$, $\xi_{2030}$, ...,  $\xi_{2050}$ are the values taken by one parameter for a specific sample of the \gls{GSA} for each of the representative years of the transition, always starting from the nominal value in 2020, $\mu_{2020}$.
+The graph has been adapted from \cite{guevara2022modeling}.}
+\label{fig:ranges_transition}
+\end{figure}
+
+Finally, the model accounts for thousands of parameters.
+The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs).
+Therefore, similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped to one single uncertain parameter.
+For instance, the uncertainty on the cost of purchasing renewable electrofuels, $c_{\mathrm{op,electrofuels}}$, identically affects the cost of e-hydrogen, e-methane, e-ammonia and e-methanol.
+Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation of cost of electricity or the electrolyser, that drive the majority of their cost of purchasing \cite{h2coalition}.
+Similarly, the uncertainties impacting the industrial demand, $industry\_EUD$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand.
 
@@
 \label{subsec:screening}
-%Similarly to what we did in the electrofuels+UQ paper (\url{https://www.mdpi.com/1996-1073/14/13/4027}), emphasise that we only consider a limited amount of uncertain parameters to keep a reasonable computation time while capturing the impact of (almost) all the uncertainties.
 
-\section{Reinforcement learning}
+After the initial phase of grouping (Section \ref{subsec:uncert_charac}), a preliminary screening was necessary to identify the key parameters to account for in this \gls{GSA}.
+\citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2}-limits using the snapshot model, \ie EnergyScope TD \cite{limpens2019energyscope}.
+Screening the results of this work, we have discarded some parameters with negligible impact (\eg CAPEX of electrolysers or variation of the freight demand), selected a subset of parameters and added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
+The exhaustive list of these 34 parameters is presented in Appendix \ref{app:UC_full}.
+
+\section{Agent-based reinforcement learning for energy transition support}
 \label{sec:meth:RL}
+
+% See chatGPT
+% 1. Introduction to Reinforcement Learning
+% 2. Problem Formulation
+% 3. Markov Decision Process (MDP)
+% 4. RL Algorithms and Techniques
+% 5. Training Setup
+% 6. Evaluation Metrics
+% 7. Baseline and Comparisons
+% 8. Implementation Details
+% 9. Experimental/Testing Setup
+
+To navigate through the transition and investigate the efficiency of different policies, this work implements the reinforcement learning approach.
+This section aims first at presenting the general concepts of this approach as well as the policy optimisation algorithm.
+Then, the environment, actions, state and reward are detailed.
+
+\subsection{Reinforcement learning fundamentals and algorithm}
+\label{subsec:meth_RL_fund_algo}
+At the initial state, \ie the energy system in 2020, the agent gets an initial observation, $\bm{o}_0$.
+An observation represents a set of the characteristics of the environment accessible to the agent for it to take the next action.
+The state, though, is the exhaustive list of these characteristics.
+Even though an observation is a subset of the state, this work uses these two words interchangeably.
+Then, it takes an action, $\bm{a}_0$, impacting its environment, \ie the energy system limited transition over the first decision window (2020-2030).
+Through this interaction with its environment, the agent is given a reward, $r_1=r\left(\bm{a}_0 | \bm{o}_0 \right)$, and ends up in a new state, \ie the energy system in 2025, characterised by a new observation, $\bm{o}_1$, and so on (\Cref{fig:Schematics_RL}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.5\textwidth]{Schematics_RL.pdf}
+\caption{\acrfull{RL} framework made of the agent interacting with its environment, \ie the energy-system model on a limited decision window of 10 years.}
+\label{fig:Schematics_RL}
+\end{figure}
+
+A learning episode is a succession of such learning steps.
+In the context of the transition pathway between 2020 and 2050, an episode can come to an end for different reasons.
+First, if the actions taken by the agent make the optimisation unfeasible, the episode is prematurely stopped before reaching 2050.
+Similarly, cumulative emissions of the system over the predefined \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) lead to an anticipated end of the episode.
+Finally, the ``natural'' end is the prescribed end of the transition, \ie 2050.
+Consequently, the maximum value of steps for an episode is equal to $N_{ep,max}=5$.
+
+The goal of a \gls{RL} approach is to optimise the mapping between inputs (\ie the observations) and output (\ie the actions), called the policy $\pi\left(\bm{a}_n | \bm{o}_n\right)$.
+To do so, an objective function, $\bm{J}(\pi)$, built on the cumulative rewards collected during each episode.
+Finally, a back-propagation process updates the weights and biases of the \gls{NN} during the learning of the agent.
+Among the wide variety of \gls{RL} algorithms applied in energy systems \cite{perera2021applications}, this work opted for \gls{SAC} \cite{haarnoja2018soft} to train and update the \gls{NN}.
+This algorithm is model-free and off-policy.
+The former characteristic is necessary as the agent is assumed to have no knowledge about the dynamics of the environment, \ie its transition or reward functions.
+The transition function is, given the current state of the environment and the action taken by the agent, a function that outputs a probability to end up to any of the next states.
+The reward function, on its turn, gives a reward given the current state and the taken action.
+In practice, in this model-free approach, the agent estimates the optimal policy directly from experience and without estimating the dynamics of the environment.
+However, model-free methods suffer from two major drawbacks: their sample inefficiency and their brittleness with respect to their hyper-parameters (\eg learning rates, exploration constants) \cite{haarnoja2018soft}.
+The former leads to a too expensive computational burden while the second requires meticulous settings to get good results.
+\gls{SAC} overcomes these two challenges as an off-policy (\ie efficiently re-using past episodes to update the target policy-network) actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function:
+
+\begin{equation}
+    \label{eq:SAC_objective}
+    \bm{J}(\pi) = \underset{\pi}{\mathbb{E}}\left[\underset{n=0}{\overset{N_{ep}}{\sum}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right) - \zeta \log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right) \right]
+\end{equation}
+
+\noindent
+where $\gamma$ is the discount factor and $\zeta$ the temperature parameter.
+The former determines how much importance we want to give to future rewards within an episode.
+The latter balances the trade-off between exploitation of proven actions via the return maximisation, \ie $\sum_{n=0}^{N_{ep}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right)$, and exploration through the entropy\footnote{Where entropy represents the amount of energy in a system not available to produce work in thermodynamics, this term stands for the randomness or stochasticity of the policy in the current context.} term, \ie $\log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right)$.
+This way, \gls{SAC} ensures sample efficiency and low sensitivity to hyper-parameters while improving exploration \cite{haarnoja2017reinforcement} and robustness \cite{ziebart2010modeling}.
+These make \gls{SAC} a state-of-the-art algorithm and one of the most efficient model-free deep RL method nowadays.
+In this work, the authors used the open-source \gls{SAC} package developed by \textsc{Stable-Baselines3} \cite{raffin2021stable} where the policy \gls{NN} is a fully connected multilayer perceptron (MLP) built with \textsc{TensorFlow} \cite{abadi2016tensorflow}.
+For further information on \gls{RL} and the \gls{SAC} algorithm, the interested reader is invited to refer to the works of \citet{sutton2018reinforcement} and \citet{haarnoja2018soft}, respectively.
 

### 2024-01-25 options/symbols_and_cmd.tex
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom

@@
 %% A
+\newabbreviation[group={Acronyms}]{API}{API}{application programming interface}
 
@@
 %% E
+\newabbreviation[group={Acronyms}]{ESOMs}{ESOMs}{energy system optimisation models}
 \newabbreviation[group={Acronyms}]{ESTD}{EnergyScope TD}{EnergyScope Typical Days}
@@
 \newabbreviation[group={Acronyms}]{GHG}{GHG}{greenhouse gases}
+\newabbreviation[group={Acronyms}]{GSA}{GSA}{global sensitivity analysis}
 \newabbreviation[group={Acronyms}]{GWP}{GWP}{global warming potential}
@@
 \newabbreviation[group={Acronyms}]{NG}{NG}{natural gas}
+\newabbreviation[group={Acronyms}]{NN}{NN}{neural network}
 \newabbreviation[group={Acronyms}]{NRE}{NRE}{non-renewable energy}
@@
 \newabbreviation[group={Acronyms}]{RE}{RE}{renewable energy}
+\newabbreviation[group={Acronyms}]{RL}{RL}{reinforcement learning}
 
 %% S
+\newabbreviation[group={Acronyms}]{SAC}{SAC}{Soft Actor Critic}
 \newabbreviation[group={Acronyms}]{SDGs}{SDGs}{Sustainable Development Goals}

### 2024-01-26 appendices/app_results_UQ.tex
\myparagraph + app_UQ_results

\section{Total transition cost}
\label{app:UQ_transition_cost}

\Cref{tab:UQ_full} gives the ranking and total Sobol index over the total transition cost of each of the 34 parameters listed in \Cref{tab:UC_full}. The first column shows these indicators for the \gls{GSA} applied on the monthly pathway model that has some limitations \cite{limpens2023pathway} but has the main advantage to run much faster. The second column gives the same indicators but for an uncertainty quantification carried out on the hourly pathway model and only using the most impacting parameters. Given the similar rankings of the parameters between these two, this comparison shows that the monthly model can be a computationally efficient proxy to quantify the uncertainties of the actual hourly model and point out the key parameters on optimisation-driving objective, the total transition cost.

Besides the top-4 parameters, rankings are slightly different. However, this does not jeopardize the comparative analysis given the similar Sobol' indices. Given their wide range of uncertainty [-64.3\%; 179.8\%] and their significant role to meet the \ce{CO2}-budget, the cost of purchasing electrofuels is the first, by far, impacting parameter. Next, comes, naturally, the industrial \gls{EUD}, representing, at the nominal value, 60\% of the total demands by 2050. The top-3 is completed by the variation of the interest rate, directly impacting the annualisation and the salvage values of the assets. Finally, since the current Belgian whole-energy system deeply relies on fossil resources, and would still do in the near future, the cost of purchasing fossil fuels is part of the impacting parameters. On the contrary, due to the very low annualised, cost, long lifetime leading to a significant salvage value  and a low-emitting fuel, the parameters related to \gls{SMR} barely impact the total transition cost.

\begin{table}[htbp!]
\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the monthly and hourly pathway models. The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties over the hourly model.}
\label{tab:UQ_full}
\centering
\begin{tabular}{l c c| c c}
\toprule
\multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
 & \multicolumn{2}{c|}{Monthly model} 	& \multicolumn{2}{c}{Hourly model} \\ 	
\midrule
\textbf{Purchase electrofuels} & 1 & (47.4\%) & 1 & (44.4\%) \\
Industry EUD & 2 & (23.5\%) & 2 & (26.4\%)  \\
Interest rate & 3 & (11.0\%) &  3 & (13.2\%)  \\
Purchase fossil fuels  & 4 & (6.9\%) & 4 & (6.9\%)   \\
\midrule
Variable OPEX of technologies & 5 & (2.9\%) & 6 & (3.0\%) \\
Purchase biofuels & 6 & (2.6\%) & 5 & (3.0\%) \\
Hourly load factor PV & 7 & (1.9\%) & 9 & (1.3\%) \\
CAPEX electric motor & 8 & (1.9\%) & 7 & (2.5\%) \\
Hourly load factor wind turbines & 9 & (1.3\%) & 8 & (1.4\%) \\
Max capacity PV & 10 & (1.1\%) & 14 & (0.5\%) \\
\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (1.0\%) \\
Available local biomass & 12 & (0.8\%) & 12 & (0.7\%) \\
CAPEX car & 13 & (0.7\%) & 10 & (1.2\%) \\
Passenger mobility EUD & 14 & (0.7\%) & 13 & (0.7\%) \\
\midrule
Modal share change LT-heat & 15 & (0.5\%) & \multicolumn{2}{c}{-} \\
Households EUD & 16 & (0.5\%) & \multicolumn{2}{c}{-} \\
Services EUD & 17 & (0.4\%) & \multicolumn{2}{c}{-} \\
Max capacity onshore wind & 18 & (0.3\%) & \multicolumn{2}{c}{-} \\
Max share of public transport & 19 & (0.3\%) & \multicolumn{2}{c}{-} \\
CAPEX PV & 20 & (0.2\%) & \multicolumn{2}{c}{-} \\
Max capacity offshore wind & 21 & (0.2\%) & \multicolumn{2}{c}{-} \\
Efficiency electric motor & 22 & (0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX fuel cell engine & 23 & (0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX ICE & 24 & (0.1\%) & \multicolumn{2}{c}{-} \\
Efficiency fuel cell engine & 25 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
Modal share change freight mobility & 26 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
Modal share change passenger mobility & 27 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX efficiency measures & 28 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX bus & 29 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX grid reinforcement & 30 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
CAPEX power grid & 31 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
\textbf{CAPEX \gls{SMR}} & 32 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
Available electricity import & 33 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
Purchase electricity & 34 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
\bottomrule

\end{tabular}
\end{table}

% \begin{table}[htbp!]
% \caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the monthly and hourly pathway models. The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties over the hourly model.}
% \label{tab:UQ_full}
% \centering
% \begin{tabular}{l c c| c c}
% \toprule
% \multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
%  & \multicolumn{2}{c|}{Hourly model} 	& \multicolumn{2}{c}{Model model} \\ 	
% \midrule
% Purchase electrofuels & 1 & (44.4\%) & 1 & (47.4\%) \\
% Industry EUD & 2 & (26.4\%) & 2 & (23.5\%)  \\
% Interest rate & 3 & (13.2\%) &  3 & (11.0\%)  \\
% Purchase fossil fuels & 4 & (6.9\%)  & 4 & (6.9\%)  \\
% \midrule
% Purchase biofuels & 5 & (3.0\%) & 6 & (2.6\%)  \\
% Variable OPEX of technologies & 6 & (3.0\%) & 5 & (2.9\%) \\
% CAPEX electric motor & 7 & (2.5\%) & 8 & (1.9\%)  \\
% Hourly load factor wind turbines & 8 & (1.4\%) & 9 & (1.3\%) \\ 
% Hourly load factor PV & 9 & (1.3\%) & 7 & (1.9\%)  \\
% CAPEX car & 10 & (1.2\%) & 13 & (0.7\%) \\
% \textbf{Potential capacity \gls{SMR}} & 11 & (1.0\%) & 11 & (0.9\%)  \\
% Available local biomass & 12 & (0.7\%)  & 12 & (0.8\%) \\
% Passenger mobility EUD & 13 & (0.7\%) & 14 & (0.7\%) \\
% Max capacity PV & 14 & (0.5\%) & 10 & (1.1\%)  \\
% \midrule
% Modal share change LT-heat & \multicolumn{2}{c}{-} & 15 & (0.5\%)  \\
% Households EUD & \multicolumn{2}{c}{-} & 16 & (0.5\%) \\
% Services EUD & \multicolumn{2}{c}{-} & 17 & (0.4\%) \\
% Max capacity onshore wind & \multicolumn{2}{c}{-} & 18 & (0.3\%) \\
% Max share of public transport & \multicolumn{2}{c}{-} & 19 & (0.3\%) \\
% CAPEX PV & \multicolumn{2}{c}{-} & 20 & (0.2\%) \\
% Max capacity offshore wind & \multicolumn{2}{c}{-} & 21 & (0.2\%)\\
% Efficiency electric motor & \multicolumn{2}{c}{-}  & 22 & (0.1\%) \\
% CAPEX fuel cell engine & \multicolumn{2}{c}{-}  & 23 & (0.1\%) \\
% CAPEX ICE & \multicolumn{2}{c}{-} & 24 & (0.1\%)\\
% Efficiency fuel cell engine & \multicolumn{2}{c}{-}  & 25 & ($<$0.1\%) \\
% Modal share change freight mobility & \multicolumn{2}{c}{-}  & 26 & ($<$0.1\%) \\
% Modal share change passenger mobility & \multicolumn{2}{c}{-}  & 27 & ($<$0.1\%)\\
% CAPEX efficiency measures & \multicolumn{2}{c}{-}  & 28 & ($<$0.1\%)\\
% CAPEX bus & \multicolumn{2}{c}{-}  & 29 & ($<$0.1\%)\\
% CAPEX grid reinforcement & \multicolumn{2}{c}{-} & 30 & ($<$0.1\%) \\
% CAPEX power grid & \multicolumn{2}{c}{-} & 31 & ($<$0.1\%)\\
% \textbf{CAPEX \gls{SMR}} & \multicolumn{2}{c}{-} & 32 & ($<$0.1\%) \\
% Available electricity import & \multicolumn{2}{c}{-}  & 33 & ($<$0.1\%) \\
% Purchase electricity  & \multicolumn{2}{c}{-} & 34 & ($<$0.1\%)\\
% \bottomrule

% \end{tabular}
% \end{table}

\newpage
\subsection{Imported renewable electrofuels}
\label{app:UQ_electrofuels}
\begin{table}[htbp!]
\caption{Comparison of the quantities of imported renewable electrofuels, in TWh, between the REF case, the SMR case and the statistical features from the \gls{GSA} (\ie Q1, median and Q3). 2020 is not in the table as, per assumption, no renewable electrofuel is imported for this year. For the sake of clarity, zeroes are replaced by ``-''.}
\label{tab:uq_ref_smr_med}
\begin{minipage}{\linewidth}
\centering
\begin{tabular}{l l | c c c c}
\toprule
\textbf{Year} & \textbf{Case} & \textbf{e-methane} & \textbf{e-hydrogen} & \textbf{e-ammonia} & \textbf{e-methanol}\\	
\toprule							
\multirow{5}{*}{2025}
 & REF & - & - & 10 & 52\\
 & SMR & - & -  & 10 & 29\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 9 & 2\\
 & Median & - & - & 10 & 46\\
 & Q3 & - & - & 11 & 55\\
\toprule
\multirow{5}{*}{2030}
 & REF & - & 1 & 10 & 52\\
 & SMR & - & 1 & 10 & 52\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 9 & 43\\
 & Median & - & - & 10 & 51\\
 & Q3 & - & 1 & 12 & 57\\
\toprule
\multirow{5}{*}{2035}
 & REF & - & 17 & 10 & 53\\
 & SMR & - & 17 & 10 & 53\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 9 & 44\\
 & Median & - & 4 & 11 & 52\\
 & Q3 & - & 16 & 32 & 58\\
\toprule
\multirow{5}{*}{2040}
 & REF & - & 16 & 23 & 54\\
 & SMR & - & 16 & 10 & 54\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 12 & 43\\
 & Median & - & 12 & 36 & 53\\
 & Q3 & 10 & 16 & 68 & 60\\
\toprule
\multirow{5}{*}{2045}
 & REF & 40 & 16 & 42 & 54\\
 & SMR & - & 16 & 11 & 54\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 24 & 43\\
 & Median & - & 13 & 49 & 52\\
 & Q3 & 44 & 17 & 77 & 60\\
\toprule
\multirow{5}{*}{2050}
 & REF & 39 & 16 & 44 & 55\\
 & SMR & 7 & 16 & 11 & 55\\
 \cmidrule{2 - 6}
 & Q1 & - & - & 20 & 44\\
 & Median & 19 & 14 & 44 & 53\\
 & Q3 & 51 & 17 & 72 & 61\\
\bottomrule							
\end{tabular}
\end{minipage}
\end{table}

\Cref{fig:results_uq_prod_cons} gives the distribution of the different routes of supply and consumption of gas like methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.

Given its lower cost of purchasing than its renewable equivalent (\Cref{fig:cs_resources_cost}) and lower \gls{GWP} than other fossil fuels (\ie $\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh versus $\mathit{gwp}_{\mathrm{op,LFO}}=0.31$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), fossil \gls{NG} remains the main source of gas in the system until 2040. Besides bio-hydrolisis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2}-budget for the transition. Its versatility makes gas used by a wide variety of technologies in the different sectors. Initially, in 2020, decentralised gas boilers, \gls{CCGT} and industrial gas boilers represent the biggest consumers of gas with 39\%, 21\% and 16\% of the total consumption, respectively. Progressively, in line with the rest of the system shifting towards more efficiency in the mid-term, industrial \gls{CHP} represent the lion's share, next to other usages in the transport or \gls{LT}-heating sectors.

On the contrary, import of fossil-based hydrogen, largely produced from steam-methane-reforming \cite{spf_economie_h2}, is rarely part of the solution due to the emissions related to the consumption of natural gas. E-hydrogen is the consistent source of hydrogen in the system, next to local production (\ie steam-methane-reforming, electrolysis or ammonia-cracking) in some rare cases where low industrial \gls{EUD} coincides with more abundant electricity from \gls{SMR} or \gls{PV}. In terms of consumption, \gls{FC}-trucks are the more consistent player. \gls{FC}-cars are also at stake but in specific cases where their CAPEX and the CAPEX of electric vehicles are in the bottom and the top of their respective uncertainty range.

Becoming cheaper than its fossil equivalent at early stages of the transition (\ie from 2030 onward), e-ammonia is the exclusive stream of ammonia in the system, except rare cases. Then, on top of its consistent \gls{NED}, the largest consumption of ammonia is \gls{CCGT} as flexible power generation units, to substitute their e-methane equivalents that have a higher \gls{LCOE} (\Cref{fig:LCOE}).

Similarly to ammonia, on top of local production in rare cases, e-methanol is the key source for methanol. Besides its own \gls{NED}, methanol is mostly consumed to produce \gls{HVC} via the \gls{MTO} process instead of naphtha-cracking, in order to respect to \ce{CO2}-budget for the transition.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.49\textwidth]{UQ_Gas_Prod.pdf}
\includegraphics[width=0.49\textwidth]{UQ_Gas_Cons.pdf}
\includegraphics[width=0.49\textwidth]{UQ_H2_Prod.pdf}
\includegraphics[width=0.49\textwidth]{UQ_H2_Cons.pdf}
\includegraphics[width=0.49\textwidth]{UQ_Ammonia_Prod.pdf}
\includegraphics[width=0.49\textwidth]{UQ_Ammonia_Cons.pdf}
\includegraphics[width=0.49\textwidth]{UQ_Methanol_Prod.pdf}
\includegraphics[width=0.49\textwidth]{UQ_Methanol_Cons.pdf}
\caption{Distribution of the different streams of supply (left) and consumption (right) of gas, hydrogen, ammonia and methanol from the \acrfull{GSA}.}
\label{fig:results_uq_prod_cons}
\end{figure}

### 2024-01-26 chap_atom_mol/chap_atom_mol.tex
\myparagraph + app_UQ_results

@@
 
-The model and the frameworks have been introduced in \cref{sec:methodo} and the case study presented in \cref{sec:case_study}.
-This section focuses on the results in two ways.
-First, Section \ref{subsec:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{subsec:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
+After the model and the frameworks have been introduced in Chapter \ref{chap:chap_methodo} and the case study presented in Chapter \ref{chap:case_study}, this one focuses on the atom-molecules dilemma in two ways.
+First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
+Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
 
@@
 \label{sec:atom_mol:results_deter} 
-In this section, like in the rest of the paper, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW, from 2040 onward.
+In this section, like in the rest of the thesis, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW, from 2040 onward.
 After investigating the deployment of \gls{SMR} through the power sector, the first part of this section focuses on this impact on macro/system-level considerations (\ie overall transition costs, primary energy mix and yearly emissions per sector).
@@
 To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case.
-Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.
+Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.\\
+
+\textbf{Compare here with EnergyVille - PATHS2050.}\\
+
 \begin{figure}[htbp!]
@@
 \label{subsec:atom_mol:results_deter_others}
-\paragraph*{High-temperature heat}
+
+\myparagraph{High-temperature heat}\\
+
 As aforementioned, the main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of overall more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
@@
 
-\paragraph*{Low-temperature heat}
+\myparagraph{Low-temperature heat}\\
+
 This sector is marginally impacted.
@@
 
-\paragraph*{Mobility}
+\myparagraph{Mobility}\\
+
 The passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} progressively substituting \gls{ICE} cars for the private sector.
@@
 
-\paragraph*{Non-energy demand}
+\myparagraph{Non-energy demand}\\
+
 The supply of ammonia (\ie from Haber-Bosch to direct import of renewable ammonia) and methanol (\ie import of renewable methanol) are unchanged between the two cases.
@@
 Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters.
-To mitigate the computational burden, we have used the \gls{PCE} approach (\Cref{subsec:meth:UQ}) on the monthly version of EnergyScope Pathway.
-As detailed by \citet{limpens2023pathway}, this simplified model does not implement daily storage technologies and integrates much more easily intermittent renewables (PVs and wind turbines) in the system as the daily supply-demand intermittency and mismatch are not at stake.
+To mitigate the computational burden, we have used the \gls{PCE} approach (Section \ref{subsec:pce}) on the monthly version of EnergyScope Pathway.
+As detailed by \citet{limpens2024pathway}, this simplified model does not implement daily storage technologies and integrates much more easily intermittent renewables (PVs and wind turbines) in the system as the daily supply-demand intermittency and mismatch are not at stake.
 Consequently, it requires less flexibility options (\ie electrification of the heating sector or import of electricity from the neighbouring countries) and ends up with a cheaper transition (-3\%).
@@
 The former is the most impacting parameter whereas the two others have much lower influence on the variation of the total transition cost.
-Given the uncertainty characterisation presented in Section \ref{subsubsec:UQ:UC}, there are 60\% chance that no \gls{SMR} could be installed.
+Given the uncertainty characterisation presented in Section \ref{sec:cs:uncertainty}, there are 60\% chance that no \gls{SMR} could be installed.
 In other words, the variation of the parameter $f_{\mathrm{max,SMR}}$ has zero impact on the variation of the total transition cost in 60\% of the samples.
-Then, in perspective with the local sensitivity analysis of Section \ref{subsec:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
+Then, in perspective with the local sensitivity analysis of Section \ref{sec:atom_mol:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
 This represents only 10\% of the samples.
@@
 On the contrary, more expensive renewable electrofuels are always imported, to a smaller or larger extent depending on the sample.
-For instance, in the REF case (Section \ref{subsubsec:results_deter_overall}), the imported electrofuels represent, by 2050, 152.9\,TWh (\ie 41\% of the primary energy mix) with an average 93€/MWh cost of purchasing and, over the entire transition, a 273\,b€ cumulative OPEX (\ie 25\% of the total transition cost).
+For instance, in the REF case (Section \ref{subsec:atom_mol:results_deter_overall}), the imported electrofuels represent, by 2050, 152.9\,TWh (\ie 41\% of the primary energy mix) with an average 93€/MWh cost of purchasing and, over the entire transition, a 273\,b€ cumulative OPEX (\ie 25\% of the total transition cost).
 
@@
 In practice, the interest rate would vary depending on the technology investment risk.
-This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2019}.}.
+This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.}.
 Finally, similarly to electrofuels, the cost of purchasing fossil fuels is also to consider in the perspective to reduce the uncertainty over the total transition cost.
 However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their uncertain impact smaller than their renewable alternatives.
+
+\Cref{fig:UQ_PDF_total_transition_cost} shows the \gls{PDF} of the total transition cost given the 1260 samples.
+Stretching between 680\,b€ and 1940\,b€, the mean, the median and the nominal value (\ie REF case) are close to each other, respectively 1160\,b€, 1150\,b€ and 1080\,b€.
+Similarly to the analysis carried out by \citet{coppitters2023optimizing}, one observes here that the distribution is right-skewed.
+It could then be qualified as ``fragile" as the top 50\% of the samples cover a bigger range (\ie between 1150 and 1940) than the bottom 50\% of the samples (\ie between 680 and 1150).
+In other words, the bad scenarios, resulting in a total transition cost higher than the median, have a bigger effect on this cost.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_PDF_total_transition_cost.pdf}
+\caption{\acrfull{PDF} of the total transition cost.
+The mean, $\mu=1.16\cdot10^3$\,b€, is slightly higher than the median ($P_{50}=1.15\cdot10^3$\,b€) and the nominal cases cost, $1.08\cdot10^3$\,b€ and $1.04\cdot10^3$\,b€ respectively for the REF and SMR cases.
+Also, with a standard deviation, $\sigma=185$\,b€, a 95\%-confidence interval would be about [0.8; 1.5]$\cdot10^3$\,b€.}
+\label{fig:UQ_PDF_total_transition_cost}
+\end{figure}
 
@@
 Figures \ref{fig:results_uq_samples_A} and \ref{fig:results_uq_samples_B} give the trend lines of the key parameters for these imports in 2050\footnote{The authors picked this specific year as it is the one where electrofuels, if imported, are imported in the largest amount, in general, compared to the other years of the transiton.}, as well the installed capacities of \gls{SMR}.
-Next to the name of a parameter, one can read its Sobol' index versus the output of interest\footnote{For these outputs of interest, different from the total transition cost, the \gls{LOO} error is generally higher than the threshold of \SI{1}{\%} defined in Section \ref{subsubsec:UQ:PCE}.
+Next to the name of a parameter, one can read its Sobol' index versus the output of interest\footnote{For these outputs of interest, different from the total transition cost, the \gls{LOO} error is generally higher than the threshold of \SI{1}{\%} defined in Section \ref{subsec:pce}.
 Consequently, the Sobol' indices are less accurate but already allow a fair relative comparison between the different parameters.}.
@@
 Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives.
-This confirms the observation made in Section \ref{subsubsec:results_deter_others}.
+This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}.
 Highly available local biomass also leads to smaller import of e-methane to supply bio-hydrolysis and produce methane-equivalent gas.
@@
 
-As already pointed out in Section \ref{subsubsec:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia.
+As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia.
 As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheap electricity flexibly produced by \gls{SMR} substitutes the \gls{CCGT}.
@@
 \label{subsec:atom_mol:results_uq_VRES}
-In line with results given in Section \ref{subsec:results_deter}, the \gls{GSA} shows that \gls{SMR} has negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines).
+In line with results given in Section \ref{sec:atom_mol:results_deter}, the \gls{GSA} shows that \gls{SMR} has negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines).
 \Cref{fig:results_uq_pdf_local_ren} gives the evolution of Sobol' indices for the most impacting parameters on the deployment of \gls{PV} and offshore wind\footnote{As the installed capacities of onshore wind is totally driven by the uncertainty on its maximum potential, $f_{\mathrm{max,windon}}$, it is not represented in the figure.}.

### 2024-01-26 chap_case_study/chap_case_study.tex
Drop publications, first part of methodo + app. Pathway choice
Drop full implementation of Pathway from the paper
Methodo (ES, UQ, brief RL) + Resultats mol_vs_atom
\myparagraph + app_UQ_results

@@
 
-\section{Contributions}
+\section*{Contributions}
 \label{sec:cs:contributions}
@@
 
-\textbf{Definition and historical trend}\vspace{-0.3cm}\\
+\myparagraph{Definition and historical trend}\\
 
@@
 
-\textbf{Methodology of quantification}\vspace{-0.3cm}\\
+\myparagraph{Methodology of quantification}\\
 
@@
 \multirow{2}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
-& $\bm{c_{\mathrm{op,electrofuels}}}$ & \textbf{Purchase electrofuels} & \textbf{II} & \textbf{-64.3\%} & \textbf{179.8\%} \\
+& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
 \midrule
@@
 & $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
-& $\bm{c_{\mathrm{inv,nuclear\_SMR}}}$ & \textbf{CAPEX \gls{SMR}}$^{(c)}$ & \textbf{I} & \textbf{-40.0\%} & \textbf{44.0\%} \\
+& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}$^{(c)}$ & I & -40.0\% & 44.0\% \\
 \midrule
@@
 & $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
-& $\bm{f_{\mathrm{max,SMR}}}$ & \textbf{Potential capacity \gls{SMR}} & \textbf{-} & \textbf{0} & \textbf{1} \\
+& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
 

### 2024-01-26 chap_methodo/chap_methodo.tex
\myparagraph + app_UQ_results

@@
 
-One of the main methodological contributions of this work regarding the development of the whole-energy system model consists in giving it the possibility to optimise the transition pathway in a myopic approach.After introducing the general concept of it, this section details more the additions brought to the model in terms of implementation.
+One of the main methodological contributions of this work regarding the development of the whole-energy system model consists in giving it the possibility to optimise the transition pathway in a myopic approach.After introducing the general concept of it, this section details more the additions brought to the model in terms of implementation.\\
 
-\vspace{0.1cm}\textbf{General concept of the myopic optimisation}\vspace{-0.3cm}\\
+\myparagraph{General concept of the myopic optimisation}\\
 
@@
 
-\vspace{0.1cm}\textbf{Additional sets, parameters and variables}\vspace{-0.3cm}\\
+\myparagraph{Additional sets, parameters and variables}\\
 
@@
 
-\vspace{0.1cm}\textbf{Myopic pathway implementation}\vspace{-0.3cm}\\
+\myparagraph{Myopic pathway implementation}\\
 
@@
 
-\vspace{0.1cm}\textbf{Definition}\vspace{-0.3cm}\\
+\myparagraph{Definition}\\
 
@@
 
-where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (\autoref{app:UC_full}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$.
+\noindent where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (\autoref{app:UC_full}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$.
 As uniform distributions are considered, the Legendre polynomials are adopted, as they are the associated family of polynomials that are orthogonal with respect to standard uniform distributions~\cite{Sudret2014}.
@@
 
-\vspace{0.1cm}\textbf{Comparison with a proven method}\vspace{-0.3cm}\\
+\myparagraph{Comparison with a proven method}\\
 
@@
 \end{equation}
-where $M$ is the objective function, $\vec{\xi}\in\omega$, except $\xi_i\leq1-\Delta$ and $\Delta$ is a set multiple of $1/(p-1)\left(\xi_{i,max}-\xi_{i,min}\right)$.
+
+\noindent where $M$ is the objective function, $\vec{\xi}\in\omega$, except $\xi_i\leq1-\Delta$ and $\Delta$ is a set multiple of $1/(p-1)\left(\xi_{i,max}-\xi_{i,min}\right)$.
 As in other studies \cite{Sin2009,Moret2017,Moret2017PhDThesis}, we consider $p$ as even and $\Delta=p/[2(p-1)]\left(\xi_{i,max}-\xi_{i,min}\right)$.\\

### 2024-01-26 options/symbols_and_cmd.tex
\myparagraph + app_UQ_results

@@
 \newabbreviation[group={Acronyms}]{CHP}{CHP}{combined heat and power}
+\newabbreviation[group={Acronyms}]{CNG}{CNG}{compressed natural gas}
 
@@
 %% I
+\newabbreviation[group={Acronyms}]{ICE}{ICE}{internal combustion engine}
 \newabbreviation[group={Acronyms}]{IEA}{IEA}{International Energy Agency}
@@
 \newabbreviation[group={Acronyms}]{LPG}{LPG}{liquefied petroleum gas}
+\newabbreviation[group={Acronyms}]{LT}{LT}{low-temperature}
 
@@
 \newabbreviation[group={Acronyms}]{MTBE}{MTBE}{methyl tert-butyl ether}
+\newabbreviation[group={Acronyms}]{MTO}{MTO}{methanol-to-olefins}
 
@@
 \newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
+\newabbreviation[group={Acronyms}]{PDF}{PDF}{probability density function}
 \newabbreviation[group={Acronyms}]{PV}{PV}{photovoltaic}
@@
 \newabbreviation[group={Acronyms}]{UQ}{UQ}{Uncertainty Quantification}
+
+%% V
+\newabbreviation[group={Acronyms}]{VRES}{VRES}{variable renewable energy sources}
 
@@
 %% P
+\newcommand{\myparagraph}[1]{{\vspace{0cm}\textbf{#1}\vspace{-0.3cm}}}
 

### 2024-02-03 appendices/app_EnergyScope.tex
\myparagraph + app_UQ_results
Caste_study post-reviews by FC

@@
 
-\vspace{0.2cm}\textbf{Overview of the snapshot model}\vspace{-0.3cm}\\
+\myparagraph{Overview of the snapshot model}\\
 
+\noindent
 EnergyScope TD \cite{limpens2019energyscope} is a model that optimises both the investment and operating strategy of a '\emph{whole}'-energy system, encompassing electricity, heating, mobility, and non-energy sectors.
@@
 EnergyScope TD has been successfully applied to various national energy systems, including Switzerland \cite{Limpens_role_2019,limpens2019energyscope}, Belgium \cite{Limpens2020}, Italy \cite{borasio2022deep}, and other European countries \cite{dommisse2020modelling}.
-Furthermore, it has been extended to a multi-region energy system model \cite{thiran2023validation}, coupled with other energy models \cite{pavivcevic2022bidirectionnal}, or employed to focus on specific sectors such as the power networks of electricity, gas, and hydrogen \cite{schnidrig2023role}.
+Furthermore, it has been extended to a multi-region energy system model \cite{thiran2023validation}, coupled with other energy models \cite{pavivcevic2022bidirectionnal}, or employed to focus on specific sectors such as the power networks of electricity, gas, and hydrogen \cite{schnidrig2023role}.\\
 
-\vspace{0.2cm}\textbf{Formulation of the snapshot model}\vspace{-0.3cm}\\
+\myparagraph{Formulation of the snapshot model}\\
 
+\noindent
 The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost or greenhouse gas emissions of the energy system.
@@
 
-\vspace{0.2cm}\textbf{Linear formulation}\vspace{-0.3cm}\\
+\myparagraph{Linear formulation}\\
 
+\noindent
 The following section illustrates, non-exhaustively, the original EnergyScope TD model.
@@
 data used ranges between 1220 to 870  [€\textsubscript{2015}/kW] between 2020 and 2035).
+\\
 
-\vspace{0.2cm}\textbf{Linking years}\vspace{-0.3cm}\\
+\myparagraph{Linking years}\\
 
+\noindent
 At this stage, all years are independent.
@@
 Eq.~(\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase.
+\\
 
-\vspace{0.2cm}\textbf{Society inertia}\vspace{-0.3cm}\\
+\myparagraph{Society inertia}\\
 
+\noindent
 To avoid unrealistically fast changes in the system, additional constraints are needed during the phases for the mobility and low temperature heat sectors.
@@
 For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase.
-Consequently, if a technology supplies more than 25\% of the low temperature heat, it would require multiple phases to replace it with a different technology.
+Consequently, if a technology supplies more than 25\% of the low temperature heat, it would require multiple phases to replace it with a different technology.\\
 %As an example, setting the maximum change of supplied heat at low temperature to 25\% would limit the change in technology related to heat at low temperature to 25\% during a phase. 
@@
 
-\vspace{0.2cm}\textbf{Cost and emissions of the transition}\vspace{-0.3cm}\\
+\myparagraph{Cost and emissions of the transition}\\
 
+\noindent
 To optimise the energy system, two key metrics must be adapted: the transition cost and the total \acrfull{GWP}. 

### 2024-02-03 appendices/app_case_study.tex
Caste_study post-reviews by FC

@@
 
-\section{\ce{CO2}-budget versus linear decrease of emissions}
+\section{Myopic versus perfect foresight pathway optimisation}
+\label{app:my_vs_pf}
+This section aims at digging more into the details of the differences observed between the myopic approaches and the reference case (\ie hourly perfect foresight model).
+
+\begin{table}[htbp]
+\caption{Exhaustive general comparison between the two different foresight approaches: Perfect foresight (PF) and Myopic (MY).
+Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
+\label{tab:PF_MY}
+\begin{minipage}{\textwidth}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{l l r r c}
+\toprule
+& & {\thead{\textbf{PF}}} & {\thead{\textbf{MY}}} & Units \\
+\midrule
+Computational time\footnote{\label{foot:comp_time}These computational times were reached on a 2.4GHz 4-core machine.} & & 830 & \textbf{373} & s \\
+\midrule
+\multirow{4}{*}{Costs in 2050}
+& Total transition\footnote{\label{foot:transition_cost}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & $\simeq$ & b€\textsubscript{2015}\\
+& Cumulative opex & 528 & $\simeq$  & b€\textsubscript{2015}\\
+& Cumulative capex & 636 & $\simeq$ & b€\textsubscript{2015}\\
+& Salvage value & 160 & -1\% & b€\textsubscript{2015}\\
+\midrule
+\multirow{5}{*}{Primary energy mix in 2050} & Total & 368.0 & $\simeq$ & TWh/y\\
+ & e-hydrogen & 15.6 & $\simeq$ & TWh/y\\
+ & e-methane & 41.0 & +6\% & TWh/y\\
+ & e-methanol & 54.8 & $\simeq$ & TWh/y\\
+ & e-ammonia & 40.7 & -10\% & TWh/y\\
+
+ \midrule
+ \multirow{3}{*}{Electrification in 2050\footnote{\label{foot:share_elec}The electrification of the other sectors (\ie centralised heat (100\%-heat pump), private (100\%-\gls{BEV}), public mobility (80\%-train and tramway), freight mobility (25\%-train) and non-energy demand (0\%)) are identical between the three approaches and are, therefore, not presented in the table.}} & System\footnote{\label{foot:system_elec}The electrification of the system is computed as the difference between the total production of electricity and the end-use demand of electricity.} & 63.3 & $\simeq$ & TWh$_{\text{e}}$\\
+ & Industrial heat\footnote{\label{foot:heat_elec}The electrification of the industrial and decentralised heating sectors are expressed in terms of thermal energy (TWh$_{\text{th}}$) provided by electrified processes, respectively industrial resistors and decentralised electric heat pumps.} & 12.3 & -7\% & TWh$_{\text{th}}$\\
+ & Decentralised heat\footref{foot:heat_elec} & 73.9 & $\simeq$ & TWh$_{\text{th}}$\\
+  \midrule
+\multirow{3}{*}{Year of full VRES-deployment} & PV & 2045 & \textbf{2040} & -\\
+ & Wind-offshore & 2030 & \textbf{2025} & -\\
+ & Wind-onshore & 2025 & 2025 & -\\
+\bottomrule
+\end{tabular}
+}
+\end{minipage}
+\end{table}
+
+Similarly to \citet{nerini2017myopic}, Figure \ref{fig:my_pestd_total_cost_cum_diff} shows that myopic optimisation ends up with a slightly more expensive energy transition by 2050 (\ie +3.2~b€$_{2015}$), compared to the perfect foresight, despite the savings done at the early stages.
+Even though this over-cost is negligible compared to the overall cost of the transition (\ie $\sim$1000~b€$_{2015}$), this is explained by the early investments in renewable technologies (\ie PVs and wind turbines) boosted by the significant salvage value retrieved from investing in the consequent reinforcement of the grid.
+
+Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€\textsubscript{2015} in 2030 whereas the overall cumulative investments, so far, are 421.3~b€\textsubscript{2015}.
+The significant lifetime (\eg 80 years) and investment cost (\ie 368M€/GW$_\text{VRES}$ \cite{readthedocs_pathway}) of the power grid, and, on a smaller scale, the district heating network, explain why the myopic optimisation opts for a higher investment in these infrastructures, at early stages.
+Similarly to what \citet{keppo2010short} observed in their studies, these early investments consequently lead to more investments, later in the transition, to renew technologies that have become too old before 2050: during the phase between 2045 and 2050, the myopic approach needs to invest in 9.2GW of PVs that have been installed 25 years before whereas the perfect foresight, by smoothing its investments over the entire transition, has to renew only 2.5GW of PVs.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{MY_cumulative_total_cost_diff.pdf}
+\caption{Cumulative total cost (i.e.
+opex+capex-salvage value) difference between the myopic and perfect foresight (PF) approaches.
+Positive values mean that the myopic approach is higher than perfect foresight.
+Early savings of the myopic vision are overcompensated by late investments further on.}
+\label{fig:my_pestd_total_cost_cum_diff}
+\end{figure}
+
+In 2050, the capacities installed in the different sectors, the design of the system in other words, are very similar between the two approaches.
+The passenger and freight mobility sectors are the same where differences smaller than 1GW are observed in other sectors.
+More interestingly, the myopic optimisation tends to postpone the decommissioning of capacities when the loss of salvage values at the end of an optimisation window would be bigger than the maintenance cost.
+For instance, in the myopic approach, 0.8GW of industrial coal boilers will remain installed in 2045 and 2050 or 3.6GW of naphtha-crackers to produce HVC in 2040, whereas these technologies are not used.
+This is comparable to the ``lock-ins" detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
+
+Highlighted in Figure \ref{fig:my_pestd_res_cat_diff}, the earlier availability of renewable (and intermittent) electricity consequently accelerates the electrification of the other sectors.
+For instance, in 2035, 3.7GW (+75\%) more of industrial electric heaters to produce 5.1TWh/year (+130\%) of additional industrial heat.
+In the low-temperature heat sector, decentralised and centralised electric heat pumps capacities are, respectively, 2.2GW (+19\%) and 1GW (+8\%) higher for each of the representative years between 2030 and 2045, to produce, around 7.8TWh/year (+23\%) and 0.8TWh/year (+1\%), at the expense of other technologies such as gas heat pumps.
+Finally, public trains substitute from 2035 a higher share of the CNG-buses.
+
+In general, due to the formulation of the salvage value (see Eq.
+\ref{eq:salvage}), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition.
+Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}.
+In parallel, in the near term, the system relies on average more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{MY_investment_return_area.pdf}
+\caption{Sum of the respective salvage value of each sector, per end year of each time window.
+All together, these salvage values represent 223.9~b€\textsubscript{2015}, 53\% of the cumulative investment costs in 2030, and 157.7~b€\textsubscript{2015}, 24\% of the cumulative investment costs in 2050.}
+\label{fig:my_pestd_inv_return_area}
+\end{figure}
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{MY_resources_category_diff.pdf}
+\caption{Primary energy resources difference between the myopic and perfect foresight (PF) approaches. Positive values mean that the myopic approach is higher than perfect foresight.}
+\label{fig:my_pestd_res_cat_diff}
+\end{figure}
+
+\section{Assessment of different emissions-trajectories}
+\label{app:CO2_trajectories}
+This section compare results of the optimisation subject to different emissions-trajectories with the REF case where the emissions are constrained to decrease linearly from the level in 2020 to carbon-neutrality in 2050.
+The first comparison is done with the trajectory subject to respect the \ce{CO2}-budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}).
+Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, alleviating the constraint on the emissions-trajectory.
+
+\subsection{\ce{CO2}-budget versus linear decrease of emissions}
 \label{app:CO2_budget}
@@
 \end{figure}
+
+\subsection{Comparison without restriction on \gls{GHG}}
+\label{app:pfmomy_comparison_without_GHG}
+
+The outcomes of a model could be limited when the case study is too restrictive.
+Indeed, ones could argue that the comparison between myopic, monthly and perfect foresight are very similar as the energy system is strongly constrained in terms of \gls{GHG} emissions.
+
+In the following, we perform a similar comparison in the transition pathway without restricting the \gls{GHG} emissions. \\
+
+\myparagraph{Reference case}\\
+
+\noindent
+Figure \ref{fig:EnergyMixPathwayWithoutGHGLimit} illustrates the transition taken by Belgium without restriction on the \gls{GHG} emissions.
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Resources_no_traj.pdf}
+\caption{Primary energy mix of a non-constrained energy transition. In this results, carbon neutrality is not reached in 2050. Some fossil fuels remain used.}
+\label{fig:EnergyMixPathwayWithoutGHGLimit}
+\end{figure}
+
+Similar trends that for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
+However, some changes reflects the cheapest option that the system could utilise to reach a cheaper transition, such as using as little electrofuels as possible.
+In this case, only e-ammonia is used for its end use demand.
+
+Instead of analysing the energy system in details, the following paragraphs will investigate if the comparison findings are consistent on a different case study.\\
+
+\myparagraph{Comparison with Myopic approach}\\
+
+\noindent
+Several key messages of the comparison have been summarised in Table \ref{tab:PF_MY}.
+In the following paragraph we analyse how these conclusions are affected when the constraint on the \gls{GHG}-emissions trajectory is removed.
+
+First, considering the overall transition cost, the myopic approach keeps on making short-term savings, \ie down to -0.2\%, before ending up with a more expensive transition by 2050, \ie +0.2\%.
+Similarly to the case with an imposed \gls{GHG}-emissions trajectory, myopic optimisation invests more, compared to the perfect foresight, at early stages into \gls{VRES} technologies to benefit from the significant salvage values of the related grid infrastructures.
+In 2030, the salvage value of the infrastructures and electricity-generation technologies account for 80.2 and 56.2 b€\textsubscript{2015}, respectively, whereas the overall CAPEX are 404.8 b€\textsubscript{2015} by then.
+
+Then, in the case with a prescribed \gls{GHG}-emissions trajectory, the myopic approach had to invest more by the end of the transition to renew \gls{PV} installed more massively at early stages and that reached the end of their lifetime before the end of the transition.
+In the case without this emissions trajectory, there is less an urgency/need for integrating renewables in the system.
+Consequently, in the latter case, there is not such an extra-investment to make to renew too old renewable assets.
+On top of this, the slower uprise of \gls{VRES} in the case without emissions trajectory leads to a smaller difference of electrification of the other sectors between the perfect foresight and myopic approaches.
+
+Finally, even though the way to get there differs between the perfect foresight and the myopic approaches, the system designs by 2050 are very similar between these two in most of the sectors.
+The main observed difference is in the freight transport where diesel boats are preferred to gas boats.
+This can be looked as a result of the lock-in effect where choices made at early stages, due to the limited foresight, remain in place in the longer term.
+
+In essence, when comparing perfect foresight and myopic approaches, distinctions arise in minor aspects, while the fundamental conclusions of Table \ref{tab:PF_MY} were verified.
+These variances have been elucidated in the preceding enumerated points and can primarily be attributed to the changes in the case study, rather than reflecting limitations inherent to the model or the comparative analysis itself.
 
@@
 \toprule
-\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
-}}  & \multicolumn{2}{c}{\textbf{Relative variation}\footnote{\label{foot:nom_val_uncert}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range_app}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
+}}  & \multicolumn{2}{c}{\textbf{Relative variation}\footnote{\label{foot:nom_val_uncert_app}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
 }}\\
@@
 & $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
-& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
+& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR_app}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
 \midrule

### 2024-02-03 chap_case_study/chap_case_study.tex
Caste_study post-reviews by FC

@@
 As detailed by \citet{limpens2024pathway}, the analysis carried out in this work can be applied to any regional whole-energy system.
-As a densely-populated and highly-industrialised country with limited local renewable potentials (mainly solar and wind), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon-neutrality in 2050 makes it an intricate case study.
+As a densely-populated and highly-industrialised country with limited local renewable potentials (\ie mainly solar and wind representing up to 50\% of the primary mix by 2050), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon-neutrality in 2050 makes it an intricate case study.
 Moreover, this case study and the subsequent analyses can be transferred - to some extent - to other industrialised countries highly dependent on fossil fuels with limited local renewable potentials (\eg the Netherlands or Germany) \cite{dommisse2020modelling}.
@@
 \label{sec:cs:contributions}
-First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heart and mobility), the \gls{NED} is often not considered.
+First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heart and mobility), the \acrfull{NED} is often not considered.
 The latter is defined as ‘\textit{energy products used as raw materials in the different sectors; that is not consumed as a fuel or transformed into another fuel}’ \cite{Eurostat2019}.
@@
 This allows bringing the non-energy sector to a similar level of details as the other sectors.
-Regarding the end-use demands,  keeping the same methodology to define \gls{EUD} in EnergyScope as \citet{Limpens2020}, this work considers updated values given the latest release of the \og EU reference scenario 2020 : energy, transport and GHG emissions: Trends to 2050 \fg by the European Commission \cite{EuropeanCommission2021}.\\
+Keeping the same methodology to define \gls{EUD} in EnergyScope as \citet{Limpens2020}, this work considers updated values given the latest release of the \og EU reference scenario 2020 : energy, transport and GHG emissions: Trends to 2050 \fg by the European Commission \cite{EuropeanCommission2021}.
 
-Second, given the focus of this work on the electrofuels, the case study includes a more explicit representation of e-ammonia and e-methanol (as well as their fossil-based equivalents), on top of e-hydrogen and e-methane, already included in the previous definition of the case study \cite{limpens2021generating}.
+Second, given the focus of this work on the electrofuels, the case study includes a more explicit representation of them: where the previous definition of the case study considered only renewable hydrogen and methane \cite{limpens2021generating}, now, e-ammonia and e-methanol (as well as their fossil-based equivalents), are implemented in the case study.
 As detailed later on, these electrofuels are considered as renewable in the sense that their \gls{GWP} is zero.
-This more explicit representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.\\
+This more explicit representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.
+For instance, considering the case of ammonia, on the top of the import routes, the Haber-Bosch process ,on one hand, is accounted for to supply this molecule.
+On the other hand, ammonia-driven \gls{CCGT} or ammonia-cracking-to-hydrogen are included as ways to consume it.
 
-Third, as nuclear energy could be a real game-changer in the energy transition worldwide \cite{IEA2022nuclear}, and especially in Belgium, this thesis has integrated the potential to install \gls{SMR} from 2040 onward.\\
+Third, as nuclear energy could be a real game-changer in the energy transition worldwide \cite{IEA2022nuclear}, and especially in Belgium, this thesis has integrated the uncertain potential to install \gls{SMR} from 2040 onward.
 
-Fourth, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.\\
+Fourth, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.
+Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.
 
 Finally, to a smaller extent, this work includes updated values for some parameters compared to the work of \citet{limpens2021generating}.
-The main change concerns the cost and performance of private mobility vehicles, which is specifically a key components in the Belgian energy transition.
-The previous version of the case study was excessively favouring fuel cell car versus \gls{BEV}.
-Comparing with other works \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
+The main change concerns the cost and performance of private mobility vehicles, which is specifically a key components in the European \cite{biresselioglu2018electric} and Belgian \cite{BFP_mob} energy transitions.
+Based on the work of \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
+Besides their lower efficiencies (\ie about 50\% less efficient), \gls{FC} car had the lion share of the private mobility, compared to \gls{BEV} given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter, in previous works \gls{limpens2021generating,rixhon2021role}.
+In these, the more limited potential to import electricity from abroad and produce it locally via \gls{VRES} forced the model to rather electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.
+To align with other similar works on the modelling of whole-energy system \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
 Regarding \gls{BEV}, while the CAPEX has been kept unchanged, the efficiency and the battery capacity, \ie the range, have been increased.
@@
 End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative years of the transition, in order to account for their daily or seasonal variability \cite{Limpens2020,limpens2021generating}.
-In this work, the yearly end-use demands (EUD) for all sectors are calculated from the rather slightly increasing forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}).
+In this work, the yearly end-use demands (EUD) for all sectors are calculated from the forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}).
 
@@
 \label{subsec:cs:NED}
-Where previously published works of \citet{rixhon2021comprehensive,rixhon2022integration} investigated more extensively the integration of the \gls{NED} in the case study of Belgium, this section summarises the rationale of doing so as well as the methodology used to quantify this demand.\\
+The \gls{NED} currently represents around 20\% of the final energy consumption in Belgium \cite{FPSEconomy2021}.
+This section summarises the rationale of adding a higher level of details to the\gls{NED} compared to what was done in the previous version of the case study \cite{limpens2021generating}.
+Then, it explains the methodology used to quantify this demand.\\
 
@@
 
-The \gls{NED} can be split into four main categories of final molecules \cite{IEA2018_petrochemicals}: (i) \gls{HVC} (worldwide production of $\sim$365Mt/year); (ii) ammonia ($\sim$185Mt/year); (iii) methanol ($\sim$100Mt/year) and (iv) the other products.
+\noindent
+The \gls{NED} can be split into four main categories of final molecules \cite{IEA2018_petrochemicals}: (i) \gls{HVC} (worldwide production of $\sim$365Mt/year, equivalent to $\sim$4770TWh/year); (ii) ammonia ($\sim$185Mt/year, equivalent to $\sim$964TWh/year); (iii) methanol ($\sim$100Mt/year, equivalent to $\sim$540TWh/year) and (iv) the other products.
 \Gls{HVC} gather the light olefins (e.g.
@@
 Ammonia is  mainly used for the production of fertilizers ($\sim$80\% of global ammonia consumption).
-Its production is dominated by \gls{NG} via steam reforming to produce hydrogen, used as feedstock in the Haber-Bosch process.
+Its production is dominated by fossil gas via steam methane reforming to produce hydrogen, used as feedstock in the Haber-Bosch process.
 Methanol is mainly converted to formaldehyde (resin) but also used for the production of other chemicals (e.g.
 solvents and gasoline-blends).
-Currently, its synthesis, like ammonia, is mainly relying on natural gas via steam reforming.
+Currently, its synthesis, like ammonia, is mainly relying on natural gas via steam methane reforming.
 Finally, the other products gather all chemicals not mentioned in the other categories such as bitumen, lubricants and other heavy products from oil refineries \cite{daioglou2014energy}.\\
 
-The \gls{NED} currently represents around 20\% of the final energy consumption in Belgium \cite{FPSEconomy2021}.
-Over the recent history, there has been a relatively constant evolution of three main categories of the final consumption for non-energy use in Belgium, \cite{statbel_NED_2019}: (i) naphtha and \gls{LPG} (between 59\% and 67\% of the total final consumption, around 59.4\,TWh in 2019), (ii) \gls{NG} (between 9\% and 14\%, 11.8\,TWh in 2019), and (iii) others (\ie bitumen, coal tar and other oil products) (between 21\% and 28\%).
+Over the recent history, there has been a relatively constant evolution of three main categories of the final consumption for non-energy use in Belgium, \cite{statbel_NED_2019}: (i) naphtha and \gls{LPG} (between 59\% and 67\% of the total final consumption, around 59.4\,TWh in 2019), (ii) fossil gas (between 9\% and 14\%, 11.8\,TWh in 2019), and (iii) others (\ie bitumen, coal tar and other oil products) (between 21\% and 28\%).
 Naphtha and \gls{LPG} are consumed in a naphtha cracker, which results in ethylene and propylene, what will be considered as \gls{HVC} in the rest of this work.
-Similarly, \gls{NG}, as non-energy carrier, is used in steam methane reformer to produce the required hydrogen to the synthesis process of ammonia.
+Similarly, fossil gas, as non-energy carrier, is used in steam methane reformer to produce the required hydrogen to the synthesis process of ammonia.
 The small shares of bitumen and coal tar are respectively used for roadworks and to produce synthetic gas through gasification.
-Finally \og other oil products\fg take into account, indistinguishably, tar and sulphur as well as by-products of the refineries (\eg \gls{BTX}).
+Finally \og other oil products'' take into account, indistinguishably, tar and sulphur as well as by-products of the refineries (\eg \gls{BTX}).
 About methanol, there is currently no production plant in Belgium even if the country plays a role in trading this commodity between its neighbouring countries and consumes part of what it imports.\\
@@
 
+\noindent
 The non-energy demand studied in this analysis focuses on the chemical industry (more than 90\% of the non-energy use in Belgium) and, similarly to other studies \cite{IEA2018_petrochemicals, daioglou2014energy}, is split between the three aforementioned main groups of products (\ie \gls{HVC}, ammonia and methanol).
 Before describing these three demands, this study excludes bitumen, coal tar and \og other oil products\fg.
-The first two represent marginal shares of the current non-energy use in such a way that they should not affect the big trends provided by this study.
+The first two represent marginal shares of the current non-energy use (\ie 4\% and 5\% respectively) in such a way that they should not affect the conclusions provided by this study.
 As described previously, the latest are mostly by-products from refineries that the system uses because they are available.
-However, in a perspective of defossilisation, since the future of fossil-based refineries is unclear, they have not be implemented in this study nor their by-products.\\
+However, in a perspective of defossilisation, since the future of fossil-based refineries is unclear, they have not been implemented in this study nor their by-products.\\
 
 Regarding \gls{HVC}, the future of their production is highly uncertain.
-One of the reasons is new regulations and strategies promoting recycling and limitation of single-use plastics \cite{EU_plastics}.
+Because of the new regulations and strategies promoting recycling and limitation of single-use plastics \cite{EU_plastics}.
 Besides this uncertainty, Belgium stays a major exporter as approximately 2/3 of plastic raw materials produced locally are exported abroad \cite{agoria_plastics}.
 Even if a significant part of \gls{HVC} produced in Belgium is not locally consumed, this demand has been set based on the assumption that Belgium will keep its industrial activity in this sector.
-This assumption does not consider the net imports of \gls{HVC}, unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
+In other words, this assumption does not deduce the part of local production  of \gls{HVC} being then exported (and not consumed locally), unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
 Therefore, the actual demand of \gls{HVC} is inferred from the consumption of naphtha and \gls{LPG} as non-energy use as well as energy-carrier in the chemical and petrochemical industries \cite{statbel_NED_2019}.
@@
 The ammonia sector in Belgium is quite different: the country locally produces and imports ammonia much more than it exports it.
-Thanks to a database from the United Nations \cite{UN_statistics} and the National Bank of Belgium, it has been identified that, over the last ten years, Belgium has imported, exported and locally produced, on average, respectively, 1010\,kt, 105\,kt and 990\,kt of ammonia.
+Thanks to a database from the United Nations \cite{UN_statistics} and the National Bank of Belgium, it has been identified that, over the last ten years, Belgium has imported 1010\,kt of ammonia, exported 105\,kt and locally produced 990\,kt on average.
 Therefore, on top of the local production, the net import (\ie import minus export) is also included in this non-energy demand.
@@
 To define the actual non-energy demand of methanol, only a 51\%-share of this net import is kept since, according to the Methanol Institute and \gls{MMSA}, this share is used for formaldehyde production in Belgium \cite{MMSA51}.
-The rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
+Currently, the rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
 This methodology gives a current non-energy demand of methanol of 269\,kt.\\
 
-Finally, after converting these masses of products into energy content (\ie LHV: \gls{HVC} - 47\,MJ/kg, ammonia - 18.8\,MJ/kg and methanol - 19.9\,MJ/kg), this work assumed constant shares between the three categories within the \gls{NED} and the same growing rate as presented by \citet{EuropeanCommission2016}.
+Eventually, after converting these masses of products into energy content (\ie LHV: \gls{HVC} - 47\,MJ/kg, ammonia - 18.8\,MJ/kg and methanol - 19.9\,MJ/kg), two final assumptions are made: (i) the current shares of each of the three commodities (\ie 77.9\%, 19.2\% and 2.9\% for \gls{HVC}, ammonia and methanol, respectively) are supposed to remain unchanged over the transition; (ii) the overall \gls{NED}, in absolute terms, follows the growing rate presented by \citet{EuropeanCommission2016}, \ie around 7\% between 2020 and 2050.
 
@@
 \label{subsec:cs:EUD_forecast}
-Although, given a significant and unsubstantiated discrepancy in the non-energy use forecasts compared to their previous report (\ie +80\% over the 2020-2030 time window), the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}.
-Looking at \Cref{fig:cs_demands}, between 2020 and 2050, one observes a noteworthy increase of the electricity (+40\%), passenger (+45\%) and freight mobility (+35\%) demands.
-The rise of the non-energy demand is more limited, \ie +6\%, whereas the heating demands is forecast to decrease: -11\% and -3\% respectively for the low and high-temperature heat demands.
+In their latest report \cite{EuropeanCommission2021}, the European Commission forecasts a significant and abrupt increase of the \gls{NED} compared to their previous report \cite{EuropeanCommission2016}, \ie +80\% over the 2020-2030 time window.
+Given this discrepancy that is unsubstantiated and specific to the case of Belgium, the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}.
+Between 2020 and 2050, one observes a noteworthy increase of the electricity (+40\%), passenger (+45\%) and freight mobility (+35\%) demands (see \Cref{fig:cs_demands}).
+The rise of the non-energy demand is more limited, \ie +6\%, whereas the heating demands is forecast to decrease: -11\% for the low-temperature heat demand and -3\% for the high-temperature heat demand.
+This is explained by a better insulation of buildings and an improved efficiency of industrial processes.
 Regarding the center graph of \Cref{fig:cs_demands}, it is the aggregation of the same data as in the left graph but per category, rather than per sector, with the non-energy demand being associated with the industry.
@@
 The sharp increase from 2020 to 2025 is due the COVID-crisis that has significantly reduced these demands in 2020.
-As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for parts of electricity and low-temperature heating demands \cite{Limpens2020}.
+As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for the fluctuating parts of electricity and low-temperature heating demands \cite{Limpens2020}.
 A daily time series is used for the passenger mobility and applied similarly to every typical days.
-Finally, for the other demands, the yearly demand is distributed uniformly over the different hours of the year.
+Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and constant share of electricity and low-temperature heat demands), the yearly demand is distributed uniformly over the different hours of the year.
 
@@
 \centering
-\includegraphics[width=0.32\textwidth]{EUD_sec.pdf}
-\includegraphics[width=0.32\textwidth]{EUD_cat.pdf}
-\includegraphics[width=0.32\textwidth]{EUD_mob.pdf}
+\includegraphics[width=0.49\textwidth]{EUD_sec.pdf}
+\includegraphics[width=0.49\textwidth]{EUD_cat.pdf}
+\includegraphics[width=0.49\textwidth]{EUD_mob.pdf}
 \caption{EnergyScope splits the whole-energy system end-use demands (EUD) into two sets: (non)-energy and transport-related.
@@
 In the center graph, the non-energy demand has been fully associated with the industrial demand.
-As detailed previously, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh.
-Graphs have been adapted from \cite{limpens2024pathway}.}
+As detailed previously, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh.}
 \label{fig:cs_demands}
@@
 availability, as detailed by \citet{limpens2024pathway}.
-\Cref{fig:cs_resources_cost} depicts the evolution the respective costs of purchasing.
+First, the evolution of the respective costs of purchasing is presented (see \Cref{fig:cs_resources_cost}).
 Regarding ``renewable electrofuels'', these are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis\cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
@@
 The limitation in terms of availability can be direct or indirect.
-On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use.
-On the other hand, wind, solar, hydro and uranium are limited by the technical potentials respectively, of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and nuclear power plants (6\,GW).
-As \gls{SMR} are foreseen, if installed, to be around the same locations (\ie Thiange and Doel) as the conventional nuclear power plants and using the same area in kW/ha, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}.
-This is even without considering the potential limit due to the local availability, in terms of volume and flow rate, of enough water that would be a more socially-accepted solution than cooling towers exhausting a dense plume.
-Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050).
+On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are arbitrarily limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use.
+On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and nuclear power plants (6\,GW), respectively.
+In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}.
+Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050) \cite{limpens2021generating}.
 In the current work, the electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``sustainable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
@@
 
-\includegraphics[width=0.32\textwidth]{Res_ren.pdf}
-\includegraphics[width=0.32\textwidth]{Res_foss.pdf}
-\includegraphics[width=0.32\textwidth]{Res_others.pdf}
+\includegraphics[width=0.49\textwidth]{Res_ren.pdf}
+\includegraphics[width=0.49\textwidth]{Res_foss.pdf}
+\includegraphics[width=0.49\textwidth]{Res_others.pdf}
 \caption{Cost of purchasing the different resources.
@@
 
-\section{Conversion technologies: SMR, \gls{NED}-related and others}
+\section{Conversion technologies}
 \label{sec:cs:technologies}
 As the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, for instance, technologies are implemented to convert these resources into the end-use demands.
-Besides their CAPEX, OPEX and lifetime defined in Section \ref{subsec:meth:ES_Pathway}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) exhibit their own charge/discharge losses.
-Eventually, there are also infrastructure technologies like the grid or the \gls{DHN} that allow to account for the investment necessary, respectively, to integrate more intermittent renewables in the power sector and the expand the use of centralised heating systems.
-The exhaustive list of these technologies have already been presented in previous works \cite{limpens2021generating}.\\
+Besides their CAPEX, OPEX and lifetime defined in Section \ref{sec:meth:ES}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) exhibit their own charge/discharge losses.
+There are also infrastructure technologies.
+They encompass, for instance, the power grid, the \gls{DHN} or technologies to produce intermediate energy carriers (\eg wood pyrolysis, biomethanolation or steam methane reforming to produce hydrogen).
+Not digging into too much details about the exhaustive list of these technologies presented in previous works \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} then the technologies to supply the \glsxtrfull{NED}.\\
 
-A specific attention is to put on the implementation of \glsxtrfull{SMR} whereas the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
+\subsection{Small modular reactor}
+\label{subsec:cs:SMR_tech}
+
+A specific attention is to put on the implementation of \gls{SMR} whereas the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
 Similarly to the analysis of \citet{PATHS2050}, a Belgian consortium for energy research, and in line with the Belgian Nuclear Research Centre (SCK-CEN) \cite{SCK-CEN_SMR}, \gls{SMR} are implemented with the features listed in \Cref{tab:SMR_features}.
@@
 Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity independently at any hour of each representative year.
-Here, we simplify SMRs as only producing electricity and assume that the after-heat is lost to the atmosphere anyway.
+Here, we simplify SMRs as only producing electricity and disregard the heat produced by the nuclear reaction.
+This is considered as lost to the atmosphere.
 
@@
 \caption{Nominal features of the SMRs in EnergyScope.
-\gls{SMR} exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year.
-$^{(a)}$ This annual availability accounts for yearly maintenance where the reactors might not operate or, at least, not at their maximum capacity.
-$^{(b)}$ 2040 is the soonest year at which \gls{SMR} could be available, optimistically assuming industrial prototypes being completed by 2035 and 5 additional years for their commercial installation.}
+\gls{SMR} exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year.}
 \label{tab:SMR_features}
+\begin{minipage}{\linewidth}
 \centering
-\begin{tabular}{l c c|c}
+\begin{tabular}{l c c}
 \toprule
-\multirow{2}{*}{\textbf{Feature}} & \multirow{2}{*}{\textbf{Value}} & \multirow{2}{*}{\textbf{Unit}} & \textbf{Similarity with}\\
- & & & \textbf{conventional nuclear}\\
+\textbf{Feature} & \textbf{Value} & \textbf{Unit}\\
 \midrule
-CAPEX & 4850 & €/kW & \checkmark\\
-Annual OPEX & 103 & €/kW/year & \checkmark\\
-Lifetime & 60 & year & \checkmark\\
-Efficiency & 40\% & -& \checkmark\\
-Maximum capacity & 6 & GW & \checkmark\\
-Annual availability & 85\%$^{(a)}$ & -& \checkmark\\
-\midrule
-Operational year & 2040$^{(b)}$ & - & \xmark\\
-Flexibility & Full & - & \xmark\\
+CAPEX & 4850 & €/kW \\
+Annual OPEX & 103 & €/kW/year \\
+Lifetime & 60 & year \\
+Efficiency & 40\% & -\\
+Maximum capacity & 6 & GW \\
+Annual availability & 85\%\footnote{\label{foot:avail_SMR}This annual availability accounts for yearly maintenance where the reactors might not operate or, at least, not at their maximum capacity.
+} & -\\
+Operational year & 2040\footnote{\label{foot:op_year_SMR}2040 is the soonest year at which \gls{SMR} could be available, optimistically assuming industrial prototypes being completed by 2035 and 5 additional years for their commercial installation.} & - \\
+Flexibility & Full & - \\
 \bottomrule
@@
 \end{tabular}
+\end{minipage}
 \end{table}
 
-For the sake of comparison, \Cref{fig:LCOE} gives the \gls{LCOE} of the principal technologies to produce electricity, based on the computation used by \citet{limpens2021generating}.
-Compared to the other flexible generation units, \gls{SMR} is significantly more cost-effective.
-In addition, we see that \gls{CCGT} supplied by e-ammonia outcompetes its e-methane equivalent, unlike their respective fossil-based equivalent.
+%\begin{table}[htbp!]
+%\caption{Nominal features of the SMRs in EnergyScope.
+\gls{SMR} exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year.}
+%\label{tab:SMR_features}
+%\centering
+%\begin{tabular}{l c c|c}
+%\toprule
+%\multirow{2}{*}{\textbf{Feature}} & \multirow{2}{*}{\textbf{Value}} & \multirow{2}{*}{\textbf{Unit}} & \textbf{Similarity with}\\
+% & & & \textbf{conventional nuclear}\\
+%\midrule
+%CAPEX & 4850 & €/kW & \checkmark\\
+%Annual OPEX & 103 & €/kW/year & \checkmark\\
+%Lifetime & 60 & year & \checkmark\\
+%Efficiency & 40\% & -& \checkmark\\
+%Maximum capacity & 6 & GW & \checkmark\\
+%Annual availability & 85\%\footnote{\label{foot:avail_SMR}This annual availability accounts for yearly maintenance where the reactors might not operate or, at least, not at their maximum capacity.
+} & -& \checkmark\\
+%\midrule
+%Operational year & 2040\footnote{\label{foot:op_year_SMR}2040 is the soonest year at which \gls{SMR} could be available, optimistically assuming industrial prototypes being completed by 2035 and 5 additional years for their commercial installation.} & - & \xmark\\
+%Flexibility & Full & - & \xmark\\
+%\bottomrule							
+%
+%\end{tabular}
+%\end{table}
+
+For the sake of comparison, the \gls{LCOE} of the principal technologies to produce electricity, based on the computation used by \citet{limpens2021generating}, are detailed (see \Cref{fig:LCOE}).
+Not including here the cost of integrating a technology in the system (\eg reinforcement of the grid and storage capacities for \gls{VRES}), the \gls{LCOE} aims at aggregating and normalizing the CAPEX and OPEX of technologies providing a common commodity, \ie electricity.
+Compared to the other flexible generation units (\ie \gls{CCGT}), \gls{SMR} is significantly more cost-effective.
+Besides being about six times more capital-intensive in €/kW, the investment is amortized over a longer expected lifetime (\ie 60 years versus 25 for \gls{CCGT}).
+Moreover, the cost of purchasing uranium driving \gls{SMR} is expected to remain stable and low whereas the expected increase of the cost of purchasing fossil fuels dominates the \gls{LCOE} of \gls{CCGT}.
+In addition, one sees that \gls{CCGT} supplied by e-ammonia outcompetes its e-methane equivalent, unlike their respective fossil-based equivalent.
+This due to the fact that e-ammonia, not requiring carbon capture, is expected to be more cost-effective to produce versus e-methane \cite{h2coalition}.
+On the contrary, fossil-based ammonia, mostly relying on steam methane reforming, requires additional steps in the production process compared to fossil gas, as introduced in Section \ref{subsec:cs:NED}.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{LCOE.pdf}
+\includegraphics[width=0.49\textwidth]{LCOE_line_2.pdf}
 \caption{Levelised cost of energy (LCOE) for the main technologies in the power sector.
-Where \gls{SMR} is cheaper than the other flexible options, CCGT running on e-ammonia is, a priori, cheaper than its e-methane alternative.}
+Gray and black curves are related to technologies runnning on renewable and fossil resources, respectively.
+Where \gls{SMR} is cheaper than the other flexible options, \gls{CCGT} running on e-ammonia is, a priori, cheaper than its e-methane alternative.}
 \label{fig:LCOE}
@@
 
-\autoref{fig:NED_tech} illustrates the different paths to produce the final molecules of the NED integrated within the model.
-Similarly to \cite{tsiropoulos2018emerging}, naphtha, here considered as \gls{LFO}, resulting from refinery operation is modeled as an imported commodity.
+\subsection{Technologies supplying the non-energy demand}
+\label{subsec:cs:NED_tech}
+
+Different paths are implemented to produce the final molecules of the NED (see \Cref{fig:NED_tech}).
+Similarly to \citet{tsiropoulos2018emerging}, naphtha, here considered as \gls{LFO}, resulting from refinery operation is modeled as an imported commodity.
 Presented here for the specific year of 2035, all data and related references can be found in \cite{GIT_NED}.
@@
 For ammonia and methanol, the molecules can either be produced locally from other resources or directly imported (with distinction between non-renewable and renewable molecules).
-Some other technologies included in the model (not represented here for the sake of clarity), are able to turn some resources presented here into others (\eg \gls{NG} to hydrogen, woody biomass to hydrogen or to synthetic natural gas, more information in \cite{Limpens2020}).
 
@@
 
+\myparagraph{The impact of integrating the \gls{NED} in the Belgian whole-energy system}\\
+
+\noindent
+Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium, using EnergyScope TD (see Appendix \ref{app:ESTD}).
+This snapshot model, \ie optimising the system in a target future year (\ie 2050 in this case) considering a green field approach, investigated the defossilisation of the system.
+To analyse the whole-energy system at different ``climate targets'', the model forced the total emissions to decrease by reducing their upper limit while optimising the total cost.
+In practice, 10\% steps of \gls{GWP} reduction were made from the ``reference scenario - 100\%''.
+This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon-neutrality).\\
+
+First and foremost, when the \gls{NED} is implemented with this higher level of details, we highlighted that woody biomass was ``cannibalised'' to produce methanol, instead of high-temperature heat, in the cost-optimum situation.
+At more ambitious ``climate targets'', e-methanol rises as the keystone to defossilise the \gls{NED} sector as \gls{MTO} becomes the favoured option to produce \gls{HVC}, representing the major share of the \gls{NED}.
+Then, including the \gls{NED} affects the selection of the technologies for the satisfaction of the heat and the electricity demands.
+The additional high-temperature demand required by the \gls{NED} forces the system to invest in more efficient technologies like \gls{CHP} instead of \gls{CCGT}.
+This additional heat demand mostly supplies naphtha-cracking substituted by \gls{MTO} at more ambitious ``climate targets'' to produce \gls{HVC}.
+To a lesser extent, to respect the emissions-caps, integrating the \gls{NED} leads to a higher integration of solar-\gls{PV} to support the electrification of the low-temperature heat sector.\\
+
+For further details on these analyses, the interested reader is invited to refer to aforementioned for further published works.
+
 \section{Uncertainty ranges}
 \label{sec:cs:uncertainty}
+As detailed in Section \ref{sec:meth:UQ}, accounting for uncertainty in \gls{ESOMs} is crucial \cite{mavromatidis2018uncertainty}, especially when it comes to optimise several decades in an inherently uncertain future.
+The fundamental step in this ambition is to characterise these uncertainties.
+In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined range of uncertainties for the model parameters.
 \Cref{tab:UC_short} gives the uncertainty ranges of some key parameters.
@@
 Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
-As SMRs are foreseen, if installed, to be around the same locations (\ie Tihange and Doel) as the conventional nuclear power plants and using the same area in kW/ha, the same 6\,GW are assumed to be the maximum capacity for SMRs.
-If it is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% of chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+}.
 Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about the availability.
@@
 \begin{table}[htbp!]
-\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025.
-$^{(a)}$ Per \cite{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
-$^{(b)}$ The nominal value of each parameter is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
-$^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}
+\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025. }
 \label{tab:UC_short}
+\begin{minipage}{\linewidth}
 \centering
@@
 \toprule
-\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}$^{(a)}$}  & \multicolumn{2}{c}{\textbf{Relative variation$^{(b)}$}}\\
+\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
+}}  & \multicolumn{2}{c}{\textbf{Relative variation\footnote{\label{foot:nom_val_uncert}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+}}}\\
     & & & &	 min 	&	 max \\ 	
@@
 & $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
-& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}$^{(c)}$ & I & -40.0\% & 44.0\% \\
+& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}& I & -40.0\% & 44.0\% \\
 \midrule
@@
 In this work, the effect of greenhouse gases is cumulative over time and a constraint is set on the overall emissions of the transition---a \ce{CO2}-budget for the transition.
-The arbitrarily chosen way of attributing emissions-budget to Belgium is usually called \og grandfathering\fg.
-Even though this approach has his pros and cons out of discussion within the scope of this work, it consists in \og maintaining that prior emissions increase future emission entitlements\fg \cite{knight2013grandfathering}.
+This approach is in line with the works defining safe operating spaces within the nine different global planetary boundaries (\ie (i) novel entities, (ii) stratospheric ozone depletion, (iii) atmospheric aerosol loading, (iv) ocean acidification, (v) biogeochemical flows, (vi) fresh water change, (vii) land system change, (viii) biosphere integrity, and, (ix) climate change) \cite{richardson2023earth,steffen2015planetary,rockstrom2009safe}.
+This ``systemic framework for addressing global anthropogenic impacts on Earth system'' gives quantitative recommendations about the \ce{CO2} concentration, among others, to maintain ``the stability and resilience of Earth system as a whole'' \cite{richardson2023earth}.
+In their review, \citet{ryberg2020downscaling} identified three main sharing principle categories when considering these safe spaces: \ie \textit{utilitarian}, \textit{egalitarian} and \textit{acquired rights} principles.In a nutshell, the former, mostly applied to the scale of industry sector \cite{ryberg2018bring,brejnrod2017absolute}, aims at maximising the sum of welfare.
+The second shares the so-called ``budget'' equally among the total population, allocating the same share to each individual \cite{hoff2017bringing,o2018good}.
+Finally, in \textit{acquired rights} principles, also called ``grandfathering'', the sharing is based on \og maintaining that prior emissions increase future emission entitlements\fg  \cite{knight2013grandfathering}.In this thesis, we have chosen the latter principle to allocate the \ce{CO2}-budget to the Belgian transition.
 This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.

### 2024-02-03 chap_methodo/chap_methodo.tex
Caste_study post-reviews by FC

@@
 
+Assessing the robustness of a whole-energy system transition pathway calls a variety of methodological tools.
+First and foremost, such an extensive system needs to be represented, \ie modelled, to further be optimized.
+This model requires some characteristics to capture the peculiarities of this system (\eg intermittency of \gls{VRES}, sector coupling) while keeping an affordable computational time and an open-documentation.
+Then, as looking into the future (\ie up to 2050 in this work) comes with its lot of uncertainties, these have to be assessed carefully in terms of characterisation and quantification.
+The former aims at defining the range over which parameters of the model vary.
+The latter allows assessing the impact that such uncertainties can have on the output of the model.
+Finally, meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
+Finally, given the limited foresight in the uncertain future that has to deal with long-term objective, \ie limiting the cumulative \gls{GHG} emissions to a certain budget by 2050, this work encompasses the optimisation of the policy,\ie set of actions to take along the transition with a specific methodology to assess its robustness.
+
+Detailing the different tools needed to answer the research questions, this chapter starts with the presentation of the whole-energy system optimization model, EnergyScope Pathway and its myopic formulation.
+Then, the focus is put on the uncertainties, their characterisation as well as quantification.
+Finally, an agent-based \acrfull{RL} approach is detailed to address the sequential decision-making process in the uncertain transition with limited vision in the future.
+The robustness of these policies is assessed via the use of the \gls{PCA}.
+
 \section*{Contributions}
 \label{sec:meth:contributions}
-\begin{itemize}
-\item Develop the myopic approach
-\item Apply Stefano's method on the pathway model with a similar approach as Guevara et al.
-\item Check that PCE was appropriate as a method for such a system (ECOS2020)
-\item Update the $C_inv_return$ based on the initial investigation of \citet{goffauxpathway} and in line with \citet{poncelet2016myopic}.
-\end{itemize}
+The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimize the behaviour of an artificial agent interacting with its environment, \ie the transition of the energy system.
+Given the uncertainties and potential shocks of the future, this approach allows the agent to play the transition in a sequential, \ie myopic, way and optimize the choice and the timing of its actions.
+This optimization is done through multiple tries-and-errors where the agent repeats the transition with a new set of uncertainties.
+
+Then, to support this step-by-step transition with limited foresight in the future, we have extended the EnergyScope Pathway model \cite{limpens2021generating}.
+Originally developed to optimize the transition in one global optimisation up to 2050, \ie perfect foresight, part of this thesis consisted in making this model able to optimize the same transition but with sequential more limited time windows, \ie myopic approach.
+
+The third principal methodological added-value is the use of \acrfull{PCA} to assess the robustness of a policy.
+Given the uncertainties and the timespan of the transition, this approach allows highlighting the main \og directions'' of variation of the system design (\ie the installed capacities).
+After this step of identification, strategies and policies can be projected on these directions to see how robust they are to the overall transition uncertainties.
+
+Finally, more minor methodological developments are part of this thesis.
+Following an approach similar to \citet{guevara2022modeling}, we have extended the ranges of uncertainty developed by \citet{Moret2017} to the pathway optimisation.
+After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model subject to different emission-constraints \citet{rixhon2021role} as well as the pathway model.
+Eventually, starting from the initial investigation of \citet{goffauxpathway}, this work has converged to the most relevant formulation of the salvage value, \ie \textbf{C\textsubscript{inv,return}} in Eq.
+\ref{eq:salvage}, for the model EnergyScope.
 
 \section*{Other authors' main contribution statement}
-On top of the main contributions of this thesis that are aforementioned, three main authors are to be mentioned for having brought a significant part of the methodological work.
+Novelty does not stand in the reinvention of the wheel.
+This thesis, instead, finds its fundamentals in great tools previously developed by other authors.
+As developers of the building blocks of the main contributions of this thesis, three main authors are to be mentioned for having brought a significant part of the methodological work.
 Based on Stefano Moret's monthly whole-energy system model (\ie EnergyScope) \cite{moret2016strategic}, Gauthier Limpens has developed the hourly version of the snapshot model (\ie EnergyScope TD) \cite{limpens2019energyscope}, as well as the perfect foresight pathway model \cite{limpens2024pathway}, to which I personally contributed too.
@@
 This work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}.
-According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is open-source documented model.
+According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is an open-source documented model.
 Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation).
@@
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
+
+\subsection{Typical days to break the curse of dimensionality}
+\label{subsec:meth:TD}
+\textbf{Add here some details about TDs}
 
@@
 
+\noindent
 Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has two main advantages: shorter computational time and more realistic representation of the short-sightedness of decision-makers.
@@
 
+\noindent
 The major add-on from the original EnergyScope Pathway model \cite{limpens2021generating} to the myopic version developed in this thesis, is the possibility to carry out the optimisation on a limited time window, of which the duration is defined by $N\textsubscript{year,opti}$.
@@
 
+\noindent
 Starting this work in 2017, AMPL Optimization Inc.
@@
 \end{figure}
+
+\myparagraph{Impact of myopic formulation on the system}\\
+
+\noindent
+As expected, the computational time is reduced drastically (\ie by 55\%) while the design remains similar.
+Given the continuous change of the input parameters over the considered time frame, the perfect foresight and myopic approaches results are very similar, like in \cite{krey2006vergleich}: less than 1\% cost difference over the transition, similar system designs by 2050 and slight shifts in time in terms of adoption of technologies.
+
+The main difference lies in the transition itself and especially in the earlier deployment of PVs and offshore wind turbines.
+These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset.
+This is mostly due to the chosen formulation of the salvage value, Eq.
+\ref{eq:salvage}.
+Since the objective function is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at early stages of the transition.
+A more detailed comparison between the myopic and perfect foresight approaches is available in Appendix \ref{app:my_vs_pf}.
 
@@
 
+\noindent
 The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie results) in the EnergyScope Pathway model ($M$).
@@
 
+\noindent
 Besides being an in-house used method, an early step of this thesis consisted in assessing \gls{PCE} with similar approach used in the literature \cite{limpens2020impact}. \\
@@
 
+The transition towards carbon-neutrality of a whole-energy system (i.e.
+including all streams of energy carriers and demands) is uncertain.
+Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), i.e.
+perfect foresight, policy makers rather go through multiple rolling-horizon short-term decisions, i.e.
+myopic approach.
+Yet, these decisions can have long-term impacts, 20 to 50 years.
+Meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
 To navigate through the transition and investigate the efficiency of different policies, this work implements the reinforcement learning approach.
@@
 \label{sec:meth:PCA}
+
+Born in 1901 with Pearson

### 2024-02-06 appendices/app_EnergyScope.tex
Appendices A, B and C post-reviews by FC

@@
 Increasing the technical resolution of energy system models often comes at the expense of a shorter simulation horizon, and vice versa.
-For instance, day-ahead grid operation models prioritise accurate grid resolution and capacity reserves for uncertainties, but they may not incorporate long-term market trends.
+For instance, day-ahead grid operation models prioritise accurate grid resolution and capacity reserves in case of foreseeable deviations, but they may not incorporate long-term market trends.
 Different model classes cater to various needs, with decreasing technical resolution.
@@
 
-In accordance with the previous classification, models aimed at aiding decision-makers in the energy transition primarily fall under the categories of planning and scenario analysis, with a relatively lower technical resolution.
+In accordance with the previous classification, models aimed at aiding decision-makers in the energy transition primarily fall under the categories of planning and scenario analysis, with a lower technical resolution than the other classes of model (see Figure \ref{fig:energy_models_classification}).
 Nonetheless, ensuring technical accuracy is of paramount importance to ensure the effective functionality of future energy systems.
 Hence, these models should meet the following requirements as a minimum:
-(i) assessment of intermittent renewable energy integration;
-(ii) accounting for all energy flows in different sectors, including the measurement of greenhouse gas emissions in the energy sector;
-(iii) exploration of all available options;
-(iv) consideration of investments throughout the transition process; and
-(v) ensuring a reasonable computation time for analysing different trajectories.
+(i) assessment of intermittent renewable energy integration thanks to an \textbf{hourly resolution} spanning a one-year time horizon;
+(ii) accounting for the \textbf{whole-energy system} as including all energy (\ie heat, electricity, mobility) and non-energy flows in different sectors, accounting for their respective greenhouse gas emissions, as well as all resources, conversion processes, and storage technologies;
+(iii) exploration of all available options through the \textbf{optimisation of investments and operations};
+(iv) consideration of long-term investments throughout the \textbf{transition pathway} process (\ie 30 years up to 2050, in our case); and
+(v) ensuring a reasonable \textbf{computational time} (\ie less than one hour on a personal laptop 15 minutes) for analysing different trajectories.
 Additionally, to enhance result reproducibility and user understanding, it is advantageous for such models to:
-(vi) maintain transparency and preferably be open-source, accompanied by collaborative documentation.
+(vi) maintain transparency and preferably be \textbf{open-source}, with accessible data and accompanied by collaborative documentation.
 
-These requirements can be transposed into criteria that a model should match:
-(i) it should have an hourly resolution spanning a one-year time horizon;
-(ii) it should encompass the entire energy system, including all types of demands (such as heat, electricity, mobility, and non-energy)
-%\footnote{Non-energy demand is often omitted in models while it represents up to 10\%  of worldwide energy demand and up to 20\% in some countries \cite{rixhon2022integration}.
+%These requirements can be transposed into criteria that a model should match:
+%(i) it should have an hourly resolution spanning a one-year time horizon;
+%(ii) it should encompass the entire energy system, including all types of demands (such as heat, electricity, mobility, and non-energy)
+%%\footnote{Non-energy demand is often omitted in models while it represents up to 10\%  of worldwide energy demand and up to 20\% in some countries \cite{rixhon2022integration}.
 \citet{eurostat2019energy} defines the non-energy demand as `\emph{energy products used as raw materials in different sectors, not consumed as a fuel or transformed into another fuel}´.}
-, as well as all resources, conversion processes, and storage technologies;
-(iii) it should optimise the system design, accounting for all the options;
-(iv) it should have a long-term investment horizon, spanning several decades;
-(v) its computational time should be reasonable, typically less than one hour on a personal laptop;
-(vi) it should be open-source, with accessible data and comprehensive documentation.
+%, as well as all resources, conversion processes, and storage technologies;
+%(iii) it should optimise the system design, accounting for all the options;
+%(iv) it should have a long-term investment horizon, spanning several decades;
+%(v) its computational time should be reasonable, typically less than one hour on a personal laptop;
+%(vi) it should be open-source, with accessible data and comprehensive documentation.
+
 These requirements are commonly found in reviews of energy system models.
-In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria: (i-iv) and (vi), along with others such as the number of users and market equilibrium.
+In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the number of users and market equilibrium.
 In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv).
@@
 In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
-Based on these reviews, Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis} compare models based on all the previous criteria except the computational time (v).
-Indeed, the latter is hard to compare as models are not apply to the same case study and the information is rarely given.
+Based on these reviews, Table \ref{tab:intro_soa} compares models based on all the previous criteria except the computational time (v).
+Indeed, the latter is hard to compare as models are not applied to the same case study and the information is rarely given.
 The table includes only the models that achieved partially at least four out of the five criteria.
@@
 \begin{table}[!htbp]
-\caption{Comparison of existing models that partially satisfy at least four of the six criteria (in alphabetical order).
+\caption{Comparison of existing models that partially satisfy at least four of the five criteria (in alphabetical order).
 Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
-Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition} (part A)}
+Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition}}
 \label{tab:intro_soa}
@@
 Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{Only for internal use.} \\
-%ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & \checkmark \\
-%GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
+ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & \checkmark \\
+GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark
+%\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
 \citet{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years.
-However, the time resolution is 16 time slices representing 4 hours per day and one day per season.} & \checkmark \\
-%H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
+However, the time resolution is 16 time slices representing 4 hours per day and one day per season.}
+ & \checkmark \\
+H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}
+%\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
 In their work, \citet{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors.
 Their study applied the model to a transition pathway for Croatia.
-In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
-%iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
-%IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
-%OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
-%Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
-%PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.}
+&  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\ 
+iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
+OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
+Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
+PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
 Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
 } & \checkmark \\
-%RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
-%ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
-%TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
+RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
+ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
 \bottomrule
@@
 
-\begin{table}[!htbp]
-\caption{Comparison of existing models that partially satisfy at least four of the six criteria (in alphabetical order).
+%\begin{table}[!htbp]
+%\caption{Comparison of existing models that partially satisfy at least four of the five criteria (in alphabetical order).
 Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
 Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition} (part B)}
-\label{tab:intro_soa_bis}
-\begin{minipage}{\textwidth}
-\centering
-\resizebox{\textwidth}{!}{
-\begin{tabular}{lcccccc}
-\toprule
-\textbf{Model} & \textbf{Ref.} & \textbf{Hourly} & \textbf{\parbox{2cm}{\centering Whole-energy}}& \textbf{\parbox{2cm}{\centering Optimis.
+%\label{tab:intro_soa_bis}
+%\begin{minipage}{\textwidth}
+%\centering
+%\resizebox{\textwidth}{!}{
+%\begin{tabular}{lcccccc}
+%\toprule
+%\textbf{Model} & \textbf{Ref.} & \textbf{Hourly} & \textbf{\parbox{2cm}{\centering Whole-energy}}& \textbf{\parbox{2cm}{\centering Optimis.
 invest.
 \& operation}} & \textbf{Pathway} & \textbf{\parbox{2cm}{\centering Open-source}} \\
-\midrule
+%\midrule
 %Calliope & \cite{Atlason2014,Calliope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{Topic is being discussed in the chat of their repository but not yet included in their documentation.} & \checkmark \\
@@
 %Enertile & \cite{enertile} & \checkmark & \checkmark \footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{ Only for internal use.} \\
-ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:transportnotsaid_bis} Transport not accounted.}\footnote{\label{foot:industrynotaccounted_bis} Industry not accounted}\footnote{\label{foot:dhnnotaccounted_bis} \gls{DHN} not accounted.}\footnote{\label{foot:decLTHnotaccounted_bis} individual heating not accounted.} & \checkmark & \checkmark & \checkmark \\
-GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
+%ESO-XEL & \cite{heuberger2017electricity} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:transportnotsaid_bis} Transport not accounted.}\footnote{\label{foot:industrynotaccounted_bis} Industry not accounted}\footnote{\label{foot:dhnnotaccounted_bis} \gls{DHN} not accounted.}\footnote{\label{foot:decLTHnotaccounted_bis} individual heating not accounted.} & \checkmark & \checkmark & \checkmark \\
+%GENeSYS-MOD & \cite{loffler2017designing} & \checkmark & \checkmark & \checkmark &  \checkmark\footnote{\citet{loffler2017designing} applied a pathway transition, but the time resolution was increased to 12h and it uses 3 typical days over a year.
 \citet{bartholdsen2019pathways} performed a multi-regional pathway (16 nodes) for the case of Germany from 2020 to 2050 with a time step of 5 years.
 However, the time resolution is 16 time slices representing 4 hours per day and one day per season.} & \checkmark \\
-H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
+%H2RES & \cite{herc2021energy} & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:h2res} In their review in 2021, \citet{prina2021bottom} classified H2RES as a simulation model on power sector only.
 In their work, \citet{herc2021energy} presented a new version of H2RES claiming to optimise the power system and partially represent other sectors.
@@
 In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
-iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & {\color{gray} \checkmark} \footnote{\label{foot:invopt_bis} iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:freeundersomespecial_bis}`\emph{Free under some special conditions}´.} \\
-IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced_bis} Commercially (paid) licensed.} \\
-OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation_bis} Simulation model.} & \checkmark & \checkmark \\
-Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced_bis}  \\
-PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+%iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & {\color{gray} \checkmark} \footnote{\label{foot:invopt_bis} iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:freeundersomespecial_bis}`\emph{Free under some special conditions}´.} \\
+%IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced_bis} Commercially (paid) licensed.} \\
+%OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation_bis} Simulation model.} & \checkmark & \checkmark \\
+%Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced_bis}  \\
+%PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
 Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
 } & \checkmark \\
-RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark  & \checkmark & \checkmark \\
-ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial_bis} \\
-TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
-\bottomrule
-\end{tabular}}
-\end{minipage}
-\end{table}
+%RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark  & \checkmark & \checkmark \\
+%ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial_bis} \\
+%TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
+%\bottomrule
+%\end{tabular}}
+%\end{minipage}
+%\end{table}
 
-From Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. 
+From Table \ref{tab:intro_soa}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. \\
+
+\myparagraph{TIMES}\\
+
+\noindent
 The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models.
@@
 This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute in their research endeavors.
-Notably, TIMES was reported commercial in 2010 \cite{Connolly2010}.
+Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}.
 A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}.
@@
 In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}.
-It is noteworthy that \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using 288 time slices, equivalent to a 12-day time resolution, highlighting the sensitivity of results to time resolution.
+It is noteworthy that \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using 12 typical days with an hourly resolution, highlighting the sensitivity of results to time resolution.
 Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data remains an area of ongoing inquiry \cite{openmod_times_description}.
- Calliope is a `\emph{tool that makes it easy to build energy system models}´ at different geographical scale.
+\\
+
+\myparagraph{Calliope}\\
+
+\noindent
+Calliope is a ``\emph{tool that makes it easy to build energy system models}'' at different geographical scale.
 Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic.
-In fact, the model is typically employed for scenario analysis with a specific focus on the electricity system.
+In fact, the model is typically employed for scenario analysis.
 Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}.
@@
 Moreover, the model has been used with decades of weather data.
-However, its application has been limited to assessing the impact of inter-year variability in wind and PV on the results, rather than evaluating a transition pathway \cite{pfenninger2017dealing}.
+However, its application has been limited to assessing the impact of inter-year variability in wind and PV on the results,
+rather than evaluating a transition pathway \cite{pfenninger2017dealing}. \\
+
+\myparagraph{GENeSYS-MOD}\\
+
+\noindent
 Similarly GENeSYS-MOD presents some limitations.
-This model is an application of the open-source energy modelling system (OSeMOSYS), itself represented as a model with a poor time discretisation and a heavy computational burden according to
-\cite{prina2019transition}.
+This model is an application of the open-source energy modelling system (OSeMOSYS), itself represented as a model with a poor time discretisation and a heavy computational burden according to \cite{prina2019transition}.
 \citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors, the time disaggregation being chosen by the user.
 For their application they used representative years with three days and two time slice per day.
+\\
 %\cite{bartholdsen2019pathways} performed a similar study for the case of Germany with 16 time slices per representative years and motivated this choice as sufficient as \cite{welsch2014incorporating} demonstrates that it gives consistent results compared to using hourly time resolution over a year.
@@
 %PyPSA, one of the most promising model thanks to its strong and active community, has also been used to perform a myopic transition \cite{pedersen2022long}.
+
+\myparagraph{PyPSA}\\
+
+\noindent
 Among the open-source models with an active community, PyPSA is one of the best-performing, with a large and active community, development at the state of the art, worldwide applications, and usage not only limited to academia.
@@
 Furthermore, PyPSA, a modeling framework recognized for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}.
+\\
+
 Hence, it is worth noting that while Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, we have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices.
-
-% SOA
-
-Hence, it appears that none of the models of Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis} fully meet all five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
+Hence, it appears that none of the models of Table \ref{tab:intro_soa} fully meets all five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
 This observation is consistent with the findings presented by \citet{prina2019transition} who identified two approaches for optimising the energy transition pathway based on the six criteria.
@@
 
-For the aforementioned reasons, the current work opted for EnergyScope Pathway, an extension of the open-source and documented EnergyScope TD model \cite{limpens2019energyscope} listed in Tables \ref{tab:intro_soa} and \ref{tab:intro_soa_bis}.
+For the aforementioned reasons, the current work opted for EnergyScope Pathway, an extension of the open-source and documented EnergyScope TD model \cite{limpens2019energyscope} listed in Table \ref{tab:intro_soa}.
 The latter has an horizon time of one year and does not account for the pathway from an existing energy system to a long-term target.
@@
 
-EnergyScope Pathway is the extension of EnergyScope TD \cite{limpens2019energyscope} that follows the snapshot approach \cite{Girones2015}, consisting in optimising the system in a target future year.
+EnergyScope Pathway is the extension of EnergyScope TD \cite{limpens2019energyscope} that follows the snapshot approach \cite{Girones2015}.
 The objective of this section is to present the fundamental variables and constraints of the latter based on which the former was developed.
@@
 \label{app:ESTD}
+
+\myparagraph{Typical days to break the curse of dimensionality}\\
+
+\noindent
+In the field of bottom-up energy system modelling\footnote{As detailed by \citet{prina2020classification}, bottom-up models offer a detailed analysis of components and interconnections within different energy sectors, allowing for a techno-economic comparison of technologies and assessment of alternatives for achieving energy targets and reducing greenhouse gas emissions.
+On the contrary, top-down modes, mostly used by economists and administrations, integrate a simplified representation of the energy system as interacting with the other macro-economic sectors.}, one of the biggest challenges is the time resolution \cite{prina2020classification}.
+With the rise of \gls{VRES}, being able to integrate them and capture their interactions with the rest of the energy system requires an hourly time resolution while optimising a whole year (\ie 8760 hours), if not a whole transition (\ie several decades).
+This long-term target combined with a fine time resolution usually leads to the so-called ``curse of dimensionality'' \cite{kuo2005lifting}.
+As an example, running EnergyScope TD over each of the 8760 hours to optimise a single target year takes more than 19h \cite{limpens2019energyscope}.
+
+To break this curse, EnergyScope TD, like other models \cite{Gabrielli2018,Despres2017,Nahmmacher2014}, relies on a subset representative days called typical days (TDs).
+This more limited number of days, \ie 12 in the rest of this thesis, cluster the days of the year that have similar time series of demands (\ie varying electricity and heat demands) and weather data (\ie sun, onshore and offshore wind).
+This way, each day of the year is associated to one of these typical days (see Figure \ref{fig:Typical_days}).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.9\textwidth]{Typical_days.pdf}
+\caption{Association of each day of the year (grays) to one of the 12 typical days (TDs) (black dots). Graph adapted from \citet{limpens2019energyscope}.}
+\label{fig:Typical_days}
+\end{figure}
+
+Finally, to properly capture the inter days dynamics, EnergyScope TD uses the ``Coupling typical days'' method from \citet{Gabrielli2018}.
+Among others, this allows representing the dynamics and the seasonality of storage capacities.
+This method as well as the clustering approach selected in our case, \ie k-medoids \cite{Kaufman2009,Park2009}, extensively detailed and compared to other methods, in the work of \citet{limpens2019energyscope}.
+\\
 
@@
 Its formulation incorporates typical days and a reconstruction method that captures different time scales from the hour to the season while accounting for the inter-weeks patterns of wind.
-This approach minimally affects the design while significantly reducing computational time \cite{thiran2023validation}.
 The model investigates all the possibilities by optimising the investment decisions and hourly operations over a year, with a computational time of less than a minute on a personal laptop.
@@
 \noindent
-The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost or greenhouse gas emissions of the energy system.
+The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost and greenhouse gas emissions of the energy system.
 Typically, the two objectives are integrated by placing a limit on emissions while simultaneously striving to minimize costs.
@@
 \noindent
-The following section illustrates, non-exhaustively, the original EnergyScope TD model.
+The following section illustrates the main constraints of the original EnergyScope TD model.
 The objective function, cost and \gls{GHG} formulation will be detailed.
@@
  \text{s.t.
-} & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)} - 1} & \forall j \in \text{\emph{TECH}}\\
+} & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{\left(\left(i_{\text{\emph{rate}}}+1\right)^{\emph{lifetime}(j)}\right) - 1} & \forall j \in \text{\emph{TECH}}
+% \label{eq:c_inv_app}%3
+% & \textbf{C\textsubscript{inv}}(j) = c_{\text{\emph{inv}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\
+% \label{eq:c_maint_app}%4
+% & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\ 
+%  \label{eq:c_op_app}%5
+% & \textbf{C\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t) t_{op} (t)  
+% & \forall i \in \text{\emph{RES}}
+ \end{flalign}
+ \endgroup
+
+The objective, Eq.~(\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
+The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau_app}).
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+%\label{eq:obj_func_app}%1
+%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
+%\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot}} = \sum_{\mathclap{j \in \text{\emph{TECH}}}} \Big(\textbf{$\tau$}(j) \textbf{C\textsubscript{inv}}(j) + \textbf{C\textsubscript{maint}} (j)\Big) + \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{C\textsubscript{op}}(i)&\\
+%\label{eq:tau_app}%2
+% \text{s.t.
+} & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{\left(\left(i_{\text{\emph{rate}}}+1\right)^{\emph{lifetime}(j)}\right) - 1} & \forall j \in \text{\emph{TECH}}\\
  \label{eq:c_inv_app}%3
@@
  \label{eq:c_maint_app}%4
- & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\ 
+ & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}
+%  \label{eq:c_op_app}%5
+% & \textbf{C\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t) t_{op} (t)  
+% & \forall i \in \text{\emph{RES}}
+ \end{flalign}
+ \endgroup
+
+The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv_app}).
+The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler.
+The total operation and maintenance costs (\textbf{C\textsubscript{maint}}) are calculated in the same way, Eq.~(\ref{eq:c_maint_app}).
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+%\label{eq:obj_func_app}%1
+%% adding 25pt space, otherwise flalign with two "&" would flush to the extreme left
+%\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot}} = \sum_{\mathclap{j \in \text{\emph{TECH}}}} \Big(\textbf{$\tau$}(j) \textbf{C\textsubscript{inv}}(j) + \textbf{C\textsubscript{maint}} (j)\Big) + \sum_{\mathclap{i \in \text{\emph{RES}}}} \textbf{C\textsubscript{op}}(i)&\\
+%\label{eq:tau_app}%2
+% \text{s.t.
+} & \textbf{$\tau$}(j) =  \frac{i_{\text{\emph{rate}}}(i_{\text{\emph{rate}}}+1)^{\emph{lifetime}(j)}}{\left(\left(i_{\text{\emph{rate}}}+1\right)^{\emph{lifetime}(j)}\right) - 1} & \forall j \in \text{\emph{TECH}}\\
+% \label{eq:c_inv_app}%3
+% & \textbf{C\textsubscript{inv}}(j) = c_{\text{\emph{inv}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\
+% \label{eq:c_maint_app}%4
+% & \textbf{C\textsubscript{maint}}(j) = c_{\text{\emph{maint}}}(j) \textbf{F}(j) & \forall j \in \text{\emph{TECH}}\\ 
   \label{eq:c_op_app}%5
@@
 
-The objective, Eq.~(\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
-The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau_app}).
-The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv_app}).
-The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler.
-The total operation and maintenance costs are calculated in the same way, Eq.~(\ref{eq:c_maint_app}).
-The total cost of the resources is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
+The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
 To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}) and following equations.
@@
   \label{eq:GWP_op_app}%7
- & \textbf{GWP\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t)  t_{op} (t) & \forall i \in \text{\emph{RES}}
+ & \textbf{GWP\textsubscript{op}}(i) = \sum_{\mathclap{t \in T }} \emph{gwp}_{\text{\emph{op}}}(i) \textbf{F\textsubscript{t}}(i,t)  t_{op} (t) & \forall i \in \text{\emph{RES}}
 \end{flalign}
@@
 While snapshot models provide insights into the energy system for individual years, they lack the capacity to capture the dynamics inherent in investment strategies throughout a transition period.
-The proposed approach involves segmenting the transition into five-year intervals, during which the energy system is optimized for one specific year.
+The proposed approach involves segmenting the transition into five-year intervals.
 This approach results in seven instances of EnergyScope TD -- called representative years -- spanning the 30-year transition period, covering the years from 2020 to 2050.
-To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition.
+To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition (see Figure \ref{fig:meth_path_methodology}).
 Overall, these constraints are integrated into a linear framework, ensuring computational efficiency, with an approximate computational time of 14 minutes on a personal laptop (2.4 GHz Intel Core i5 quad-core).
-
-Figure \ref{fig:meth_path_methodology} illustrates the pathway concept.
 Simplification and choices were necessary to implement linearly the problem while keeping a tractable computational time.

### 2024-02-06 appendices/app_case_study.tex
Appendices A, B and C post-reviews by FC

@@
 
-The defossilisation\footnote{In a more sustainable future, some of the energy carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons (\eg e-methane or e-methanol).
-This is why this paper rather uses ``defossilisation" rather than ``decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.} of the different sectors are not performed at the same rate.
+The defossilisation of the different sectors are not performed at the same rate.
 The non-energy demand of methanol and ammonia are substituted by electrofuels.
@@
 The annualised cost of the energy system in 2020 is estimated to 44.3~b€/y and increases by 5.5~b€/y to reach 49.8~b€/y by 2050.
-\textbf{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y \cite{My2050}, while the other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
-The differences come from the scope of the energy system, as an example \textbf{My2050} also accounts for the agriculture sector.
+The work of \citet{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y, while the other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
+The differences come from the scope of the energy system: as an example \citet{My2050} also account for the agriculture sector.
 These differences highlight the difficulty to compare different studies due to difference of scope and partial availability of used data.

### 2024-02-06 chap_case_study/chap_case_study.tex
Appendices A, B and C post-reviews by FC

@@
 Currently, its synthesis, like ammonia, is mainly relying on natural gas via steam methane reforming.
-Finally, the other products gather all chemicals not mentioned in the other categories such as bitumen, lubricants and other heavy products from oil refineries \cite{daioglou2014energy}.\\
+Finally, the other products gather all chemicals not mentioned in the other categories such as bitumen, lubricants and other heavy products from oil refineries \cite{daioglou2014energy}.
 
@@
 As described previously, the latest are mostly by-products from refineries that the system uses because they are available.
-However, in a perspective of defossilisation, since the future of fossil-based refineries is unclear, they have not been implemented in this study nor their by-products.\\
+However, in a perspective of defossilisation, since the future of fossil-based refineries is unclear, they have not been implemented in this study nor their by-products.
 
@@
 Then, given the respective efficiencies (1.83t$_{\text{naphtha}}$/t$_{\text{HVC}}$ and 1.67t$_{\text{LPG}}$/t$_{\text{HVC}}$) \cite{IEA2018_petrochemicals}, the current demand of \gls{HVC} is estimated equal to 3069\,kt, without making distinctions between the different chemicals (\ie ethylene, propylene and \gls{BTX}).
-\\
 
@@
 Therefore, on top of the local production, the net import (\ie import minus export) is also included in this non-energy demand.
-This gives a current demand of 1895\,kt of ammonia.\\
+This gives a current demand of 1895\,kt of ammonia.
 
@@
 Currently, the rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
-This methodology gives a current non-energy demand of methanol of 269\,kt.\\
+This methodology gives a current non-energy demand of methanol of 269\,kt.
 
@@
 As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for the fluctuating parts of electricity and low-temperature heating demands \cite{Limpens2020}.
-A daily time series is used for the passenger mobility and applied similarly to every typical days.
+A daily time series is used for the passenger mobility and applied similarly to every days of the year.
 Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and constant share of electricity and low-temperature heat demands), the yearly demand is distributed uniformly over the different hours of the year.
@@
 \label{sec:cs:resources}
-To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$, their global warming potential, $\mathit{gwp}_{\mathrm{op}}$, as well as their
+To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$, their global warming potential, $\emph{gwp}_{\mathrm{op}}$, as well as their
 availability, as detailed by \citet{limpens2024pathway}.
@@
 As the amount to import from each of these locations is hard to forecast, the current work considers the average cost between the different locations.
-Besides these, every other resource has its specific \gls{GWP} like coal ($\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), natural gas ($\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh) or the fossil-based molecules equivalent to the electrofuels (\eg $\mathit{gwp}_{\mathrm{op,ammonia}}=0.46$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,methanol}}=0.41$\,kt$_{\ce{CO2},\text{eq}}$/GWh).
+Besides these, every other resource has its specific \gls{GWP} like coal ($\emph{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), natural gas ($\emph{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh) or the fossil-based molecules equivalent to the electrofuels (\eg $\emph{gwp}_{\mathrm{op,ammonia}}=0.46$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\emph{gwp}_{\mathrm{op,methanol}}=0.41$\,kt$_{\ce{CO2},\text{eq}}$/GWh).
 
@@
 They encompass, for instance, the power grid, the \gls{DHN} or technologies to produce intermediate energy carriers (\eg wood pyrolysis, biomethanolation or steam methane reforming to produce hydrogen).
-Not digging into too much details about the exhaustive list of these technologies presented in previous works \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} then the technologies to supply the \glsxtrfull{NED}.\\
+Not digging into too much details about the exhaustive list of these technologies presented in previous works \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} then the technologies to supply the \glsxtrfull{NED}.
 
@@
 In practice, 10\% steps of \gls{GWP} reduction were made from the ``reference scenario - 100\%''.
-This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon-neutrality).\\
+This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon-neutrality).
 
@@
 This additional heat demand mostly supplies naphtha-cracking substituted by \gls{MTO} at more ambitious ``climate targets'' to produce \gls{HVC}.
-To a lesser extent, to respect the emissions-caps, integrating the \gls{NED} leads to a higher integration of solar-\gls{PV} to support the electrification of the low-temperature heat sector.\\
+To a lesser extent, to respect the emissions-caps, integrating the \gls{NED} leads to a higher integration of solar-\gls{PV} to support the electrification of the low-temperature heat sector.
 
@@
 Others only set the objective as the carbon-neutrality by 2050.
-For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of \og unicorn technologies\fg or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.\\
+For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of \og unicorn technologies\fg or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.
 
@@
 This is another sign of the urgency to act to mitigate climate change as this 30-year budget represents only 10 years of the current emissions.
-\\
 

### 2024-02-06 chap_intro_ccl/intro.tex
Appendices A, B and C post-reviews by FC

@@
 As the question of policymakers is not only what to do but how to do it, we need to address the optimisation of policies. This would introduce the RL part
+
+\textbf{In a more sustainable future, some of the energy carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons (\eg e-methane or e-methanol).
+This is why this paper rather uses ``defossilisation" rather than ``decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.}

### 2024-02-06 chap_methodo/chap_methodo.tex
Appendices A, B and C post-reviews by FC

@@
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
-
-\subsection{Typical days to break the curse of dimensionality}
-\label{subsec:meth:TD}
-\textbf{Add here some details about TDs}
 
@@
 In their analysis of the overcapacity in European power systems, \citet{moret2020overcapacity} emphasised that such a ``possibility of \textit{recourse}" is very appropriate to address uncertainty gradually unfolding over time.
-Consequently, the development of the myopic approach represents the foundations of the further implementation of the agent-based reinforcement learning framework (see Section \ref{sec:meth:RL}).\\
+Consequently, the development of the myopic approach represents the foundations of the further implementation of the agent-based reinforcement learning framework (see Section \ref{sec:meth:RL}).
 
@@
 \noindent
-Besides being an in-house used method, an early step of this thesis consisted in assessing \gls{PCE} with similar approach used in the literature \cite{limpens2020impact}. \\
+Besides being an in-house used method, an early step of this thesis consisted in assessing \gls{PCE} with similar approach used in the literature \cite{limpens2020impact}.
 
@@
 
-to rank parameters among each other. In (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.\\
+to rank parameters among each other. In (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.
 

### 2024-02-12 chap_RL/chap_RL.tex
First version of PCA method

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

\section{Definition of the actions, states and rewards}
\label{sec:act_states_rew}
After pointing out that the environment is basically the EnergyScope myopic model at the different steps of the transition, define the actions, the states as well as the shape of the reward.

\subsection{Actions}
\subsection{States}
\subsection{Reward}

\section{Results - As in Smart Energy Systems Conference}

%\subsection{Training}
%Here present results RL-oriented (actions, states, reward)
%
%Then, present results energy system-oriented (installed capacities, costs, generation)

%\subsection{Testing}
%See how the different policies saved successively during the training behave when facing to new samples. This way, we can pick an optimal policy (the one giving the maximum average (or another metric) reward). 
%
%We then compare the results of RL with the perfect foresight-TD deterministic and the perfect foresight-monthly with uncertainties (by setting the actions of the agent as variables in these two optimisations). The comparison would go over different aspects:
%\begin{itemize}
%\item Over-cost
%\item Over-change, a bit like what Paolo defined as design error/change (in terms in consumed resources, installed capacities,...)
%\end{itemize}

### 2024-02-12 chap_RobPol/chap_RobPol.tex
First version of PCA method

%!TEX root = ../thesis_main.tex
%!TEX encoding = UTF-8 Unicode

The purpose of this chapter is to display results regarding PCA as presented orally to you in December 2023

### 2024-02-12 chap_atom_mol/chap_atom_mol.tex
Caste_study post-reviews by FC
First version of PCA method

@@
+\textbf{Add here references and analyses of nuclear and/or molecules as THE solution} \cite{IEA_Nuclear_2022}
+
 \section*{Contributions}
@@
 \end{itemize}
-
-\textbf{Add here references and analyses of nuclear and/or molecules as THE solution} \cite{IEA_Nuclear_2022}
 
@@
 
-\textbf{Compare here with EnergyVille - PATHS2050.}\\
+{\color{red} \textbf{Compare here with EnergyVille - PATHS2050 and other works from Seb's report.}}\\
 
@@
 With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity, consequently increasing their load factor and yearly production respectively by 31\% and 48\%.
-On the other hand, given the 44.6\,TWh of electricity produced by \gls{SMR}, cogeneration units are less relevant and, by 2050, 2.6\,GW industrial gas boilers completely substitute \gls{CHP} to produce 16.6\,TWh (\ie 23\%) of the total production of high-temperature heat.
+On the other hand, given the 44.6\,TWh of electricity produced by \gls{SMR}, cogeneration units are less relevant and, by 2050, 2.6\,GW industrial gas boilers completely substitute \gls{CHP} to produce 16.6\,TWh (\ie 23\%) of the total production of high-temperature heat.\\
 
@@
 Then, from a mix between oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system progressively shifts towards electric \gls{HP} only.
-Similarly for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economic option.
+Similarly for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economic option.\\
 
@@
 About the public mobility, trains and tramways supply their \textit{a priori} set maximum share, respectively, 50\% and 30\% complemented by \gls{CNG} buses substituting diesel-driven buses.
-Similarly, considering the freight transport, technology shifts (\ie from diesel to \gls{FC} trucks) or modal shares (\ie 53\%-47\% split between \gls{NG} and (bio)-diesel boats) are identical between the two cases.
+Similarly, considering the freight transport, technology shifts (\ie from diesel to \gls{FC} trucks) or modal shares (\ie 53\%-47\% split between \gls{NG} and (bio)-diesel boats) are identical between the two cases.\\
 

### 2024-02-12 chap_methodo/chap_methodo.tex
First version of PCA method

@@
 % See chatGPT
-% 1. Introduction to Reinforcement Learning
-% 2. Problem Formulation
+% 1. Introduction to Reinforcement Learning --> Methodo
+% 2. Problem Formulation --> Methodo
 % 3. Markov Decision Process (MDP)
-% 4. RL Algorithms and Techniques
+% 4. RL Algorithms and Techniques --> Methodo
 % 5. Training Setup
@@
 
-\subsection{Reinforcement learning fundamentals and algorithm}
-\label{subsec:meth_RL_fund_algo}
+\subsection{Reinforcement learning fundamentals and application to energy systems}
+\label{subsec:meth_RL_fundamentals}
+
+\subsection{Problem formulation and algorithm}
+\label{subsec:meth_RL_algo}
 At the initial state, \ie the energy system in 2020, the agent gets an initial observation, $\bm{o}_0$.
@@
 Then, it takes an action, $\bm{a}_0$, impacting its environment, \ie the energy system limited transition over the first decision window (2020-2030).
-Through this interaction with its environment, the agent is given a reward, $r_1=r\left(\bm{a}_0 | \bm{o}_0 \right)$, and ends up in a new state, \ie the energy system in 2025, characterised by a new observation, $\bm{o}_1$, and so on (\Cref{fig:Schematics_RL}).
+Through this interaction with its environment, the agent is given a reward, $r_1=r\left(\bm{a}_0 | \bm{o}_0 \right)$, and ends up in a new state, \ie the energy system in 2025, characterised by a new observation, $\bm{o}_1$, and so on (see Figure \ref{fig:Schematics_RL}).
 
@@
 \includegraphics[width=0.5\textwidth]{Schematics_RL.pdf}
-\caption{\acrfull{RL} framework made of the agent interacting with its environment, \ie the energy-system model on a limited decision window of 10 years.}
+\caption{\Acrfull{RL} framework made of the agent interacting with its environment, \ie the energy-system model on a limited decision window of 10 years.}
 \label{fig:Schematics_RL}
@@
 
-\section{Principal Components Analysis}
+\section{Robustness assessment via PCA}
 \label{sec:meth:PCA}
+When optimizing a transition pathway of a whole-energy system, including its uncertainties, capturing the most variable changes of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
+For the case study detailed in Chapter \ref{chap:case_study}, this consists of \ie 7 representative years of the transition (\ie from 2020 to 2050), 113 possible technologies subject to uncertain parameters.
+To tackle this challenge, we have developed a methodology based on the \acrfull{PCA}.
+This methodology provides two main outputs.
+First and foremost, using the model runs necessary to quantify the impact of the uncertain parameters on the total cost of the transition (see Section \ref{sec:meth:UQ}), it gives a metric on which to assess the robustness of energy transition policies resulting from different approaches.
+This metric gives more insight that, for instance, the variation of the total transition cost that encompasses too many aspects (\ie design and operation strategies, variation along the transition) in one single value.
+Second, these ``directions of variation'' can highlight key modal shifts or highly varying design strategies over the transition.
+After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to give these ``directions of variation'' and to assess the robustness of policies.
 
-Born in 1901 with Pearson
+\subsection{Principal Component Analysis: General concept}
+\label{subsec:meth:PCA:PCA}
+Born in the early 20th century \cite{pearson1901on,hotelling1933analysis}, the \acrfull{PCA} finds its fundamentals from the \gls{SVD}.
+The latter is a generalization, to an arbitrary (\ie not especially square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
+The core concept of principal component analysis (PCA) involves simplifying a dataset with numerous interconnected variables by reducing its dimensionality.
+The aim is to preserve as much variability within the data as feasible.
+This is accomplished by transforming the data, $\mathbf{x}$, into a new set of variables called principal components (PCs), $\mathbf{z}$.
+These components are uncorrelated and arranged in such a way that the first ones retain the majority of the variability found in all of the original variables.
+On the other hand, the final principal components (PCs) pinpoint directions where there is minimal variation, indicating nearly constant linear relationships among the original variables \cite{jolliffe2002principal}.
+
+Easier to represent in two dimensions, let us consider a vector $\mathbf{x}$ composed of the variables $x_1$ and $x_2$, $p=2$, and 25 realisations of them (See Figure \ref{fig:PCA_intro} (left)).
+The first \gls{PC}, $z_1$, is a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $x$ with maximum variance:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign}
+\hspace{0pt}
+  \label{eq:PCA_z1}%7
+ z_1=\mathbf{\alpha_1^{T}x}=\alpha_{11}x_1+\alpha_{12}x_2+...+\alpha_{1p}x_p=\sum_{j=1}^{p} \alpha_{1j}x_j
+\end{flalign}
+\endgroup
+
+\noindent
+where $^T$ means the transpose vector.
+Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and maximizing the variance.
+These transformations lead to the construction of the PCs (see Figure \ref{fig:PCA_intro} (right)).
+In more general cases, one can write $z_k=\mathbf{\alpha_k^{T}x}$ as the $k^{\text{th}}$ PC.
+There can be up to $p$ PCs even though, usually, most of the variance of the original data can be captured by $m$ PCs where $m\ll p$.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.4\textwidth]{PCA_intro_original.pdf}
+\hspace{1.5cm}
+\includegraphics[width=0.4\textwidth]{PCA_intro_PCA.pdf}
+\caption{Original observations of the dataset (left) and their projection with respect to their PCs (right).
+The variation of the realisations is more significant in the direction of $x_2$ than $x_1$.
+Once projected with respect to their PCs, the variation is even more significant in the direction of $z_1$ than in either of the original variables, as it captures most of their variance.
+Graph adapted from \cite{jolliffe2002principal}.}
+\label{fig:PCA_intro}
+\end{figure}
+
+The PCs are computed based on the covariance matrix of $\mathbf{x}$, $\mathbf{\Sigma}$.
+The diagonal of this matrix gives the variance of the $i^{\text{th}}$ variable whereas the other elements give the covariance between the $i^{\text{th}}$ and the $j^{\text{th}}$ variables where $i\neq j$.
+Out of this matrix, $\mathbf{\alpha_k}$ is the eigenvector of $\mathbf{\Sigma}$ corresponding to its $k^{\text{th}}$ highest eigenvalue $\lambda_k$.
+Finally, as the $\mathbf{\alpha_k}$ for all $k=1,2,\cdots, p$ are normalized,  \ie $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
+In other words, $\alpha_{ki}$, \ie the components of$ \mathbf{\alpha_k}$, give the weight of the $i^{\text{th}}$ original variable, $x_i$, in the $k^{\text{th}}$ PC, \ie $z_k$, where $\lambda_k$ gives the variance of the original data captured by $z_k$.
+In other words, a high absolute value of $\alpha_{ki}$ means that $x_i$ has a significant impact in the direction given by the $k^{\text{th}}$ PC \cite{zdybal2022advancing}.
+The interested reader is invited to refer to the work of \citet{jolliffe2002principal} for further mathematical demonstrations, information and examples.
+
+\subsection{Principal components of the transition}
+\label{subsec:meth:PCA:transition}
+As introduced, the objective is to define the main technological drivers of the variation through the transition to 2050 subject to uncertainties.
+To do so, before calculating the PCs of the transition, three preliminary steps are necessary: (i) data selection, (ii) data scaling and, (iii)outliers management.
+After this preprocessing, PCs can be computed for each of the year of the transition then aggregated to give a bigger picture over the whole transition.\\
+
+\myparagraph{Data selection}\\
+
+\noindent
+Prior to any data processing, the choice of the original dataset is fundamental.
+In the will to characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Eq.
+\ref{eq:F_newBuilt}).
+Even though these represent only a part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Eq.
+\ref{eq:c_inv}) and, more indirectly, the resources to use.
+In other words, it captures the technological landscape of the transition.
+Having defined the type of variable to consider, we need to assemble a relevant dataset.
+This is given by the runs to quantify the impact of the uncertain parameters on the total cost of transition required by the method described in Section \ref{sec:meth:UQ}.
+Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ states for the sample number of the uncertainty quantification method.
+For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=240$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
+Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}).
+In other words, we focus here only on the years 2025, 2030, 2035, 2040, 2045 and 2050.
+This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}).\\
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{PCA_data.pdf}
+\caption{Original raw data considered in the \acrfull{PCA} of the variation of the design strategy through the transition, $\textbf{x}(y,j,s)$.}
+\label{fig:PCA_data}
+\end{figure}
+
+\myparagraph{Data scaling}\\
+
+\noindent
+Preprocessing the dataset before employing a method to reduce dimensionality, like \gls{PCA}, can greatly affect the structure of the simplified representation and the characteristics of the features extracted from the dataset \cite{parente2013principal,peerenboom2015dimension}.
+Scaling the original raw data via normalization, \ie reducing data to $[0, 1]$ interval has a double purposes: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
+Consequently, the first part of this data preprocessing consists in scaling the installed capacities versus their respective sector and representative year (see Eq.
+\ref{eq:PCA_scaling_0}).
+The sectors, as defined in EnergyScope, are the electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{NED}, storage and infrastructures.
+For instance, the installed capacity of \gls{PV} panels in the year $y$ of the sample $s$ is scaled by the maximum installed capacity in the electricity sector in the year $y$ among all the samples.
+
+\begin{equation}
+ \label{eq:PCA_scaling_0}
+\textbf{x}^{*}(y,j,s)=\frac{\textbf{x}(y,j,s)}{\mathrm{max}_{sec,y}\left(\textbf{x}(y,j,s)\right)}
+ \quad \forall y \in \text{\emph{YEARS}}, sec \in \text{\emph{SECTORS}}
+\end{equation}
+
+Then, to give ``directions/metrics'' representative to the size of each sector within the energy system, we added another weight based on the relative share of commodity produced by each sector (see Figure \ref{fig:Share_EUD_sectors}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Share_EUD_sectors.pdf}
+\caption{Multiplying factor for each of the end-use sectors. Over the transition, sectors like electricity or methanol become more important due to sector coupling, \eg e-mobility or \gls{MTO}.}
+\label{fig:Share_EUD_sectors}
+\end{figure}
+
+To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal value (see Chapter \ref{chap:atom_mol}).
+The end-use-demands as well as the commodity produced for the sector coupling are based on the results of this deterministic REF case.
+For instance, the share of the electricity sector accounts for its \gls{EUD} and the electricity produced to supply other sectors (\eg heat, mobility).
+Finally, to compare apples with apples, we converted the \gls{EUD} in the mobility sectors, \ie passenger and freight, into the \gls{FEC} they require in the REF case.
+This gives the second weighing factor to scale data, on top of the one of Eq.
+\ref{eq:PCA_scaling_0} (see Eq.
+\ref{eq:PCA_scaling_1}).
+
+\begin{equation}
+ \label{eq:PCA_scaling_1}
+\textbf{x}^{**}(y,j,s)=\textbf{x}^{*}(y,j,s)\cdot \text{share}_{\text{EUD}}(y,sec)
+ \quad \forall y \in \text{\emph{YEARS}}, sec \in \text{\emph{SECTORS}}
+\end{equation}
+
+One would notice that this second scaling factor omits the infrastructure and storage technologies.
+In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact.
+Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies (\eg \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}).
+\\
+
+\myparagraph{Outliers management}\\
+
+\begin{wrapfigure}{r}{3cm}
+\centering
+\captionsetup{justification=centering}
+\includegraphics[width=2.8cm]{Schematic_boxplot.pdf}
+\caption{}
+\label{fig:Schematic_boxplot_methodo}
+\end{wrapfigure}
+
+\noindent
+Handling outliers is one of the biggest challenges in data science \cite{aguinis2013best}.
+These are defined as data points differing significantly from the rest of the data set.
+Being extreme values, outliers influence the overall dataset variance and, consequently, rotate the PCs directions towards them \cite{stanimirova2007dealing}.
+In the context of \gls{PCA}, outliers could be defined as ``model fit outliers'' as their presence influences the fit of the model.
+There are several techniques to detect/define and handle the outliers \cite{aguinis2013best}.
+In this work, detection is performed via the box plot technique, as outliers are identified as those points lying beyond the plot’s whiskers, or fences.
+These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}).
+Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the rest of the dataset for this specific technology and year across all the samples.
+Then, one option to handle these outliers would be to remove the whole data point for which one technology presents an outlying installed capacity.
+However, it would lead to reducing the entire dataset (\ie several thousands of points) to too few data points (\ie several tens) limiting the applicability of the \gls{PCA} approach.
+Consequently, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
+In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively.\\
+
+\myparagraph{Principal components of each representative year}\\
+
+\noindent
+Now that data are selected and preprocessed, principal components are first computed for each representative year separately, using the Python package \textsc{PCA} from \textsc{sklearn.decomposition}.
+As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.} in our case study) which is intractable.
+Moreover, the first PCs keep track of most of the variance of the system whereas the last ones present a smaller interest.
+Choosing the appropriate threshold involves a trade-off.
+Retaining too few principal components may result in loss of important information, while retaining too many may lead to over-fitting or computational inefficiency.
+The good practice followed in this work is to compute, for each of the representative years of the transition (except 2020),  the PCs that capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
+At the end of this step, we have a list of $m$ PCs, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and
+
+\begin{equation}
+\label{eq:PC_year}
+\sum_{i=0}^m\text{var}\left(\text{PC}_{y,i}\right)\geq 90\% \sum_{i=0}^p\text{var}\left(\text{PC}_{y,i}\right) \quad \forall y \in \text{\emph{YEARS}}
+\end{equation}
+
+\noindent
+where $m$ is presumably different for each representative year and $p$ is the total number of variables, hence the maximum number of PCs.
+For the entire transition, it gives a total of $M$ $\text{PC}_{y,i}$.\\
+
+\myparagraph{Principal components of the transition}\\
+
+\noindent
+The final step consists in defining metrics on the whole transition based on the $\text{PC}_{y,i}$ computed for each representative year, separately.
+To do so, all the $\text{PC}_{y,i}$ from every year are sorted together in a descending order based on their respective variance.
+Then, starting with the one with the highest absolute variance, all the other $\text{PC}_{y,i}$ similar to it are clustered together.
+The similarity between two PCs is defined according to the cosine similarity approach.
+Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest.
+Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is greater or equal to 90\%:
+
+\begin{equation}
+\label{eq:cos_similarity}
+S_C (A,B):= \cos(\theta) = {\mathbf{A} \cdot \mathbf{B} \over \|\mathbf{A}\| \|\mathbf{B}\|} = \frac{ \sum\limits_{i=1}^{n}{A_i  B_i} }{ \sqrt{\sum\limits_{i=1}^{n}{A_i^2}} \cdot \sqrt{\sum\limits_{i=1}^{n}{B_i^2}} } \geq 90\%
+\end{equation}
+
+\noindent
+where $A$ and $B$ represent two different PCs.
+The components of these similar PCs are then averaged to form the first PC of the transition, $\text{PC}_{\text{transition},1}$.
+Then, the process repeats with the $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$.
+This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to 90\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$ generated at the previous step:
+
+\begin{equation}
+\label{eq:PC_transition}
+\sum_{i=0}^N\text{var}\left(\text{PC}_{y,i}\right)\geq 90\% \sum_{i=0}^M\text{var}\left(\text{PC}_{y,i}\right)
+\end{equation}
+
+\myparagraph{Robustness assessment of roadmaps}\\
+
+\noindent
+Similarly to the work of \citet{moret2020overcapacity}, roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimizations (see Chapter \ref{chap:chap_RobPol}).
+The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps.
+In conclusion, a roadmap would be defined as more robust than another one if the projection of its myopic runs on the different $\text{PC}_{\text{transition}}$ span on a more narrow range.

### 2024-02-12 options/symbols_and_cmd.tex
Caste_study post-reviews by FC
First version of PCA method

@@
 \newabbreviation[group={Acronyms}]{HP}{HP}{heat pump}
+\newabbreviation[group={Acronyms}]{HT}{HT}{high-temperature}
 \newabbreviation[group={Acronyms}]{HVC}{HVC}{high value chemicals}
@@
 \newabbreviation[group={Acronyms}]{IPCC}{IPCC}{intergovernmental panel for climate change}
+\newabbreviation[group={Acronyms}]{IQR}{IQR}{interquatile range}
 
@@
 \newabbreviation[group={Acronyms}]{NED}{NED}{non-energy demand}
-\newabbreviation[group={Acronyms}]{NG}{NG}{natural gas}
+\newabbreviation[group={Acronyms}]{NG}{NG}{fossil gas}
 \newabbreviation[group={Acronyms}]{NN}{NN}{neural network}
@@
 %% P
+\newabbreviation[group={Acronyms}]{PC}{PC}{principal component}
+\newabbreviation[group={Acronyms}]{PCA}{PCA}{Principal Component Analysis}
 \newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
@@
 \newabbreviation[group={Acronyms}]{SMR}{SMR}{small modular reactor}
+\newabbreviation[group={Acronyms}]{SVD}{SVD}{singular value decomposition}
 

### 2024-02-12 Figures.py

@@ -1,753 +1,0 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Jan 10 14:47:49 2024
-
-@author: xrixhon
-"""
-
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-"""
-Created on Thu Jul 20 10:52:57 2023
-
-@author: xrixhon
-"""
-
-import rheia.POST_PROCESS.post_process as rheia_pp
-import rheia.UQ.uncertainty_quantification as rheia_uq
-import matplotlib.pyplot as plt
-import plotly.express as px
-import plotly.graph_objects as go
-
-import pandas as pd
-import plotly.io as pio
-import numpy as np
-import os, sys
-from pathlib import Path
-import pickle as pkl
-import hashlib
-
-pio.templates.default = 'simple_white'
-pio.kaleido.scope.mathjax = None
-
-pio.renderers.default = 'browser'
-
-dir_to_save = '/Users/xrixhon/Documents/PhD_Thesis/Papers/5 - Nuke_vs_Mol/Images/'
-
-def dict_color_full():
-    categories = ['Sectors','Electricity','Heat_low_T','Heat_high_T','Mobility','Freight','Ammonia',
-               'Methanol','HVC','Conversion','Storage','Storage','Storage_daily','Resources',
-               'Infrastructure','Years_1','Years_2','Phases']
-    color_dict_full = dict.fromkeys(categories)
-
-    for c in categories:
-        color_dict_full.update(dict_color(c))
-    
-    return color_dict_full
-
-def dict_color(category):
-    color_dict = {}
-    
-    if category == 'Electricity':
-        color_dict = {"NUCLEAR":"deeppink", "NUCLEAR_SMR": "deeppink", "CCGT":"darkorange", "CCGT_AMMONIA":"slateblue", "COAL_US" : "black", "COAL_IGCC" : "dimgray", "PV" : "yellow", "WIND_ONSHORE" : "lawngreen", "WIND_OFFSHORE" : "green", "HYDRO_RIVER" : "blue", "GEOTHERMAL" : "firebrick", "ELECTRICITY" : "dodgerblue"}
-    elif category == 'Heat_low_T':
[…]

### 2024-02-16 appendices/app_case_study.tex
Atom_molecules: first part of intro

@@
 To compensate the intermittency, the system relies on imported electricity, gas \gls{CCGT}, sector coupling and storage.
+This is in line with the work of \citet{Devogelaer2020} that ends up with about 80~GW of total installed power generation capacity by 2050, with less \gls{PV} (\ie 39~GW) and more wind capacities (\ie 25~GW of which offshore takes 8.3~GW).
 As an illustration, in 2050, 176.8~TWh of electricity transit on the grid which includes 32.4~TWh of electricity imported and 15.4~TWh of electricity from \gls{CCGT}.
-This result is aligned with other studies that estimates different ranges: 180-310 TWh/y \cite{Devogelaer2013}, 126-140~TWh/y \cite{My2050} and in a more recent study
+This result is aligned with other studies that estimates different ranges: 180-310 TWh/y \cite{Devogelaer2013}, about 250~TWh/y \cite{Devogelaer2020}, 126-140~TWh/y \cite{My2050} and in a more recent study
 %\footnote{The EnergyVille consortium, made of belgian universities and research institutes, applied the TIMES-BE model to the case of Belgium to assess the energy transition pathways \cite{PATHS2050}.
@@
 
-Figure \ref{fig:pestd_cumul_inv} illustrates the cumulative investments made throughout the transition, amounting to a total of 377.8 b€\textsubscript{2015}.
+Figure \ref{fig:pestd_cumul_inv} illustrates the cumulative investments made throughout the transition, amounting to a total of 377.8~b€\textsubscript{2015}.
+This makes 12.6~b€\textsubscript{2015} on average per year, which represents 2.1\% of the current Belgian \gls{GDP}.
+This in the range of other studies on climate-neutral scenarios concluding additional investment needs of the order of 2 to 3\% of the \gls{GDP} on a global scale \cite{IEA2021,IRENA2021} or 2.7 to 3.8\% at the European level \cite{Widuto2023}.
+
 Initially, the infrastructure, transport, and electricity sectors each account for approximately one-third of the investments.
@@
 To mitigate the cost of the transition, My2050 suggests deploying a fleet of no more than one million vehicles and implementing a car sharing system, distinct from car-pooling, as an inevitable measure \cite{My2050}.
+
+%% INCREASE OF CAPEX in terms of GDP
+%At international level, work on climate-neutral scenarios indicates additional investment needs of the order of 2 to 3% of world GDP on average per year over the period between now and 2050, with levels of USD 4,800 billion according to IEA (2021) and IRENEA (2021) and USD 5,800 billion according to BNEF (2022).
+At European level, modelling work by the European Commission (EC, 2022) suggests additional investment of €360 billion on average per year, or in the region of 1.5 to 2% of GDP.
+National studies in the UK (CCC, 2020) and France (Pisani-Ferry, 2022) indicate similar levels, also between 1.5 and 2% of GDP.
+At Belgian level, McKinsey (2023) calculates an average annual additional requirement equivalent to 2 to 3% of GDP.
 

### 2024-02-16 chap_atom_mol/chap_atom_mol.tex
Atom_molecules: first part of intro

@@
-\textbf{Add here references and analyses of nuclear and/or molecules as THE solution} \cite{IEA_Nuclear_2022}
+On top of scenarios with more profound behavioural changes, variety of technological pathways are often investigated to meet the ambitions of climate change mitigation.
+For instance, in their work, \citet{My2050} assessed different scenarios for a climate neutral Belgium by 2050.
+Depending on the scenarios, the emphasis is put on a higher electrification of the whole-energy system, a higher consumption of hydrogen or more complex moleclules (\ie electrofuels) or a bigger reliance on biomass and the related \gls{BECCS}.
+Besides these, ``unicorn'' technologies are also investigated \cite{heuberger2018impact}.
+These are technological solutions that have not (yet) reached a high enough maturity, \ie TRL 3 to 7, or facing non-techno-economic hurdles (\ie social acceptance) to be currently deployed on a large scale.
+Among others, nuclear energy, potentially in the form of \acrfull{SMR}, finds an interest in the literature \cite{IEA_Nuclear_2022,PATHS2050} as well as in actual current investments, in Belgium for instance \cite{SMRlesoir}.
+
+In a society with an increasing electricity demand, notably due to the electrification of sectors like mobility, low-temperature heat or industry, and a deeper integration of \gls{VRES}, there seems to be a case for nuclear energy to produce a reliable low-\ce{CO2}emission electricity \cite{IAEA2008}, especially given the willingness to phase out of imported Russian fossil fuels anchored in the European REPowerEU Plan \cite{REPowerEU}.
+Unlike fossil natural gas that is a ``flow-based'' resource, uranium is ``stock-based'', which favours the security of electricity supply, in case of a conflict like the invasion of Ukraine, for instance.
+
+``Country specific energy studies are needed as a prerequisite to the decision of following the nuclear route'' \cite{IAEA2008}.
+Even though, in a country like Belgium, reaching the goal of energy transition will not be a ``winner-takes-all'' situation but rather a combination of solutions, implemented simultaneously \cite{Limpens2020}, this chapter focuses on this atom-molecules dilemma in two ways.
+First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
+Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
+
+\textbf{Disclaimer}: Relying on ``local'' nuclear energy for some (or on importing electrofuels from abroad for others) is highly controversial.
+On top of purely techno-economic aspects, non-exhaustively mentioned beforehand, there are other ethical, societal or even political considerations to account for when addressing this question.
+Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} in the Belgian energy landscape, on a strictly techno-economical point of view.
+This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in the Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
 
@@
 \item Pathway UQ
-\item Contrast analysis between atom and molecules and local renewables
 \item Influence of the parameters on the import of molecules and the installation of SMR
+
+\item Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad.
+
 \end{itemize}
-
-After the model and the frameworks have been introduced in Chapter \ref{chap:chap_methodo} and the case study presented in Chapter \ref{chap:case_study}, this one focuses on the atom-molecules dilemma in two ways.
-First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
 
@@
 
-{\color{red} \textbf{Compare here with EnergyVille - PATHS2050 and other works from Seb's report.}}\\
+{\color{red} \textbf{Compare here with EnergyVille - PATHS2050.}}\\
 
@@
 Not shown here but all the samples of the \gls{GSA} highlight that \gls{SMR} is installed to its maximum capacity, \ie 6\,GW, as soon as possible.
-In other words, the only parameter ``Potential capacity \gls{SMR}'' dictates the installation of this technology\footnote{In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from, respectively, 2040, 2045 and 2050.}.
+In other words, the only parameter ``Potential capacity \gls{SMR}'' dictates the installation of this technology\footnote{In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from 2040, 2045 and 2050, respectively.}.
 Surprisingly, the [-40\%; +44\%]  variation of its CAPEX has a negligible impact, with a Sobol' index of 0.9\%.

### 2024-02-16 chap_case_study/chap_case_study.tex
Atom_molecules: first part of intro

@@
 Based on the work of \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
-Besides their lower efficiencies (\ie about 50\% less efficient), \gls{FC} car had the lion share of the private mobility, compared to \gls{BEV} given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter, in previous works \gls{limpens2021generating,rixhon2021role}.
+Besides their lower efficiencies (\ie about 50\% less efficient), \gls{FC} car had the lion share of the private mobility, compared to \gls{BEV} given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter, in previous works \cite{limpens2021generating,rixhon2021role}.
 In these, the more limited potential to import electricity from abroad and produce it locally via \gls{VRES} forced the model to rather electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.

### 2024-02-16 chap_intro_ccl/conclusion.tex
Atom_molecules: first part of intro

@@
 List the future works to build upon the thesis
+
+\textbf{Word about sufficiency} oui mais c'est sans doute celle qui est le plus enviable car la moins risqué à tenter d'atteindre, les solutions mirages technologiques, si on y croit trop on met tout la dessus et si ca foire, on est encore plus dans la merde pcq la direction est mauvaise, la solution sobriété, meme si jamais atteint les efforts pr l'atteindre ne seront pas contre productif

### 2024-02-16 chap_intro_ccl/intro.tex
Atom_molecules: first part of intro

@@
+It has been proven that the climate change (among other environmental challenges) is (mostly) due to the concentration of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
+This concentration resulting from the cumulative emissions (and subtractions), usually expressed in kt$_{\ce{CO2},\text{eq}}$, over time could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
+
+\begin{equation}
+\label{eq:equality_GHG}
+\mathrm{GHG} =  \frac{\mathrm{GHG}}{\text{Primary energy}} \times \frac{\text{Primary energy}}{\mathrm{EUD}}\times \frac{\mathrm{EUD}}{\text{Population}} \times \text{Population}\\
+ \end{equation}
+
+\noindent
+where the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third could stand as the energy intensity per capita.
+Such an identity, mathematically-correct though, is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its global/encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
+However, Eq.
+\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition.
+Besides the question of the total population and its growth \cite{dodson2020population,scovronick2017impact}, these three levers of actions are: renewables, efficiency and sufficiency aiming at reducing the first, the second and the third terms on the right-hand side of Eq.
+\ref{eq:equality_GHG}, respectively.
+The latter, explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, is defined by \citet{lage2023citizens} as ``a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people''.
+Altough this finds a growing interest in the scientific community \cite{o2018good}, it requires, maybe more than the two other levers, interdisciplinarity \cite{schmidt2015interdisciplinary}, \ie the combination of multiple academic disciplines like sociology, psychology or politics, that are out of the scope of my expertise, and, consequently, this thesis.
+However, the work developed in the present manuscript aims at providing support to such interdisciplinary projects to assess sufficiency policies.
+More within the grasp of the engineering world, this thesis rather focuses on the first two terms of Eq.
+\ref{eq:equality_GHG}, \ie renewables and efficiency.
+This aligns with the current European policies binding the Member States of the European Union.
+For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind, an \textbf{increase in energy efficiency} and significantly \textbf{higher shares of energy from renewable sources} in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
+
+It aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
+
 %!TEX root = ../thesis_main.tex
@@
 
-Our society In their 2030 Agenda, United Nations have worked on identifying 17 \gls{SDGs} as a plan of action for society (or people), environment (or planet) and economy (or prosperity) \cite{un_sdgs}.
+In their 2030 Agenda, United Nations have worked on identifying 17 \gls{SDGs} as a plan of action for society (or people), environment (or planet) and economy (or prosperity) \cite{un_sdgs}. 
 %

### 2024-02-16 chap_methodo/chap_methodo.tex
Atom_molecules: first part of intro

@@
 % See chatGPT
-% 1. Introduction to Reinforcement Learning --> Methodo
-% 2. Problem Formulation --> Methodo
-% 3. Markov Decision Process (MDP)
-% 4. RL Algorithms and Techniques --> Methodo
+% 1. Introduction to Reinforcement Learning --> Methodo --> DONE
+% 2. Problem Formulation --> Methodo --> DONE
+% 3. Markov Decision Process (MDP) --> Methodo --> DONE
+% 4. RL Algorithms and Techniques --> Methodo --> DONE
 % 5. Training Setup
 % 6. Evaluation Metrics
-% 7. Baseline and Comparisons
+% 7. Baseline and Comparisons --> Results
 % 8. Implementation Details
@@
 \label{subsec:meth_RL_fundamentals}
+
+\begin{wrapfigure}{r}{3cm}
+\centering
+\captionsetup{justification=centering}
+\includegraphics[width=2.8cm]{RL_Fundamentals.pdf}
+\caption{}
+\label{fig:RL_Fundamentals}
+\end{wrapfigure}
+
+\Gls{RL} is a subfield of machine learning focused on training an agent to make sequential decisions by interacting with an environment to achieve specific goals.
+Unlike supervised learning, where data is labeled, and unsupervised learning, where patterns are inferred from unlabeled data, reinforcement learning deals with learning from interaction, typically through trial and error.
+This way, \gls{RL} is considered as active learning \cite{cao2020reinforcement}.
+Starting from an initial state, the agent takes an action that impacts its environment.
+The latter feeds back the agent with a reward and the new state (see Figure \ref{fig:RL_Fundamentals}).
+This goes on until reaching the end of the episode.
+When the episode is done, the agent starts again from an initial state, takes an action and so on.
+The agent learns to optimize its policy by maximizing a notion of cumulative reward over time.
+This policy refers to the strategy or mapping from states to actions that the agent employs to make decisions.
+Essentially, it defines the behavior of the agent in the environment.
+The ultimate goal of the agent is often to find an optimal policy, which maximizes the expected cumulative reward over time.
+All these concepts and interactions between the agent and its environment are formalized as a \gls{MDP} \cite{sutton2018reinforcement}, represented by the tuple $<s,a,T,r,\pi,\gamma>$.
+$s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions.
+$a\in A$ is the action among the action space, $A$, and $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$.$r$ is the reward received by the action when taking the action $a$ from a state $s$, $R(s,a)$.
+Finally, $\pi$ is the policy as already described and $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
+During the learning/optimization process, the agent acts according to the exploitation-exploration trade-off.
+In the former case, the action $a$ is directly given the mapping provided by the current policy $\pi$,  depending on the state $s$.
+In the latter, the action is randomly picked within the action space.
+For further information, the interested reader is invited to refer to work of \citet{sutton2018reinforcement} or the course given by \citet{davidsilver_RL_online} available online.
+
+Due to the increasing complexity of the systems and the integration of uncertainties, the last decades have seen the emergence of publications where \gls{RL} is applied to energy systems \cite{perera2021applications}.
+In their respective reviews, \citet{cao2020reinforcement} and \citet{perera2021applications} highlighted groups of problems addressed with \gls{RL} in the research field of energy systems: building energy management system (BEMS), optimization of dispatch and operational control closely linked with the energy market and the optimal power flow problem in the grid, micro-grid management, electro-mobility or even demand-side management or optimal control of energy system devices like maximum power point tracking (MPPT) of wind turbines and \gls{PV} panels.
+The major novelty of this thesis in the use of \gls{RL} stands then in its application to a new kind of energy system problem: the optimization of the transition pathway of a whole-energy system.
+In this sense, the objective is to optimize and provide a policy to support this transition subject to uncertainties.
 
@@
 In the context of the transition pathway between 2020 and 2050, an episode can come to an end for different reasons.
-First, if the actions taken by the agent make the optimisation unfeasible, the episode is prematurely stopped before reaching 2050.
+First, if the actions taken by the agent make the optimisation infeasible, the episode is prematurely stopped before reaching 2050.
 Similarly, cumulative emissions of the system over the predefined \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) lead to an anticipated end of the episode.
@@
 
-Then, to give ``directions/metrics'' representative to the size of each sector within the energy system, we added another weight based on the relative share of commodity produced by each sector (see Figure \ref{fig:Share_EUD_sectors}).
+Then, to give ``directions/metrics'' representative to the size of each sector within the energy system, we added another weight based on the relative share of commodity produced by each sector.
+For the case study of Belgium detailed in Chapter \ref{chap:case_study}, this gives a higher weight for electricity and low-temperature heat sectors (see Figure \ref{fig:Share_EUD_sectors}).
 
@@
 \includegraphics[width=0.8\textwidth]{Share_EUD_sectors.pdf}
-\caption{Multiplying factor for each of the end-use sectors. Over the transition, sectors like electricity or methanol become more important due to sector coupling, \eg e-mobility or \gls{MTO}.}
+\caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition.
+These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimized through the perfect foresight approach.
+Over the transition, sectors like electricity or methanol become more important due to sector coupling, \eg e-mobility or \gls{MTO}.}
 \label{fig:Share_EUD_sectors}

### 2024-02-16 options/symbols_and_cmd.tex
Atom_molecules: first part of intro

@@
 %% B
+\newabbreviation[group={Acronyms}]{BECCS}{BECCS}{bioenergy with carbon capture and storage}
 \newabbreviation[group={Acronyms}]{BEV}{BEV}{battery electric vehicle}
@@
 \newabbreviation[group={Acronyms}]{CCGT}{CCGT}{combined cycle gas turbine}
+\newabbreviation[group={Acronyms}]{CCS}{CCS}{carbon capture and storage}
 \newabbreviation[group={Acronyms}]{CHP}{CHP}{combined heat and power}
@@
 %% G
+\newabbreviation[group={Acronyms}]{GDP}{GDP}{gross domestic product}
 \newabbreviation[group={Acronyms}]{GHG}{GHG}{greenhouse gases}
@@
 %% M
+\newabbreviation[group={Acronyms}]{MDP}{MDP}{Markov decision process}
 \newabbreviation[group={Acronyms}]{MMSA}{MMSA}{Methanol Market Services Asia}

### 2024-02-19 appendices/app_taxonomy.tex
Merge taxonomy paper + Implement Hervé's comments on case_study

\section{Introduction}
\label{sec:intro}
Since stone age, humans use fuels, defined as any energy carriers intended for energy conversion \cite{FAO_biofuel, ISO16559}. First evidence of the use of domesticated fire was established in 790 000 B.C. \cite{alperson2010acheulian}. Thus, biomass has been the first fuel used by the human kind for security, cooking and heating. Nowadays, most of the used energy sources are fossil fuels. In 2019, oil, coal and gas represented respectively 31\%, 25\% and 23\% of the global primary energy consumption \cite{ourworld_2019}. Despite their great advantage, high energy density, these fuels have a major drawback: their combustion releases huge quantities of carbon dioxide (35 Gt of \ce{CO2} in 2019) mainly responsible for climate change \cite{iea2020world}. The biggest challenge of the energy transition is to secure the energy supply while reducing the greenhouse gas emissions. In practice, this means finding alternatives to fossil fuels.

First and foremost, in the context of the energy transition, fuels will keep playing a major role in the global energy system \cite{Ahlgren2012}. Even if electricity gains shares through the electrification of the energy demand, it will not entirely displace fuels for three main reasons: storage, infrastructure compatibility and cross-sectorial links. Because of their intermittency and space disparity, a deeper integration of variable renewable energy sources (VRES) requires storage and transport in order to supply the energy demand at the right time and in the right place \cite{Brouwer2016,Evans2012,Gallo2016,hall2008,Rosa2017}. Where typical container of batteries are limited in terms of storage capacity (up to 10~MWh) and present significant costs and self-discharge losses, the energy conversion into fuels provides a more affordable solution for higher storage capacity (from 100~GWh) and longer storage time scales (months to year) \cite{Rosa2017}. Due to the economic inertia and their infrastructure legacy \cite{Ahlgren2012}, fuels remain the most appropriate solution for sectors requiring high energy density (\eg heavy-duty transport, shipping, aviation or chemical industry) \cite{Albrecht2020,Decker2019,Goede2018,Pearson2012,Rosa2017,Stancin2020,Trieb2018,Zeman2008}. 
\citet{contino2020whole} point out that the energy transition is an interdisciplinary effort and not solely about the power sector. The latter represents only a fifth of the global energy consumption \cite{iea_renewable}. Also, \citet{Goede2018} has shown in 2018 that \ce{CO2} emissions of the Netherlands are equally apportioned between the different types of end-use-demands (\ie power, heat, mobility and non-energy). This highlights the necessity to consider every energy sector rather than focusing all efforts on the power system and even more, to shift towards a multi-vector interconnected energy system. In this cross-sectorial approach and with the perspective of increasing shares of VRES, fuels are promising energy carriers in order to maximise the overall system efficiency \cite{mathiesen2015, Stancin2020}.

Given the growing diversity of pathways for transforming the renewable energy sources into fuels, clear classification and terminology are necessary \cite{bailera2017}. As predicted by \citet{ridjan2016}, there is now a need to support the right fuel technology development, via the use of a more comprehensive and quantitative terminology (\eg specifying the share of biomass in the energy balance of the production of a biofuel). The objective of this terminology is to avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions. To a further extent, such an harmonisation may help ``to facilitate comparison between national fuel market and enhance trading" \cite{FAO_biofuel}. It also aims at promoting and increasing the use of biomass as energy carriers and electricity transformed into fuel. On a general public perspective, this terminology aims at simplifying and clarifying the addressed terms and allows to have a more critical mind. In \Cref{sec:lit_review}, we present the terminology currently used in the literature. On top of this review of the scientific literature, this work also includes review of terminology defined by the International Organization for Standardization (ISO) and other official institutions (\eg the Food and Agriculture Organisation of the United Nations (FAO) or the International Energy Agency (IEA)) and presents the motivation for a new one, more quantitative and measurable. The latter is introduced in \Cref{sec:definition}. Finally, \Cref{sec:discussion} gives the room for a discussion and the actual recommendations.

\section{Terminology used in the literature}
\label{sec:lit_review}
\citet{ridjan2016} performed a first literature review between 2006 and 2014, looking in Scopus for the three keywords: (i) ``synthetic fuels", (ii) ``electrofuels" and (iii)``alternative fuels". Based on this extensive review work, they proposed the following definitions: (i) \textit{Synthetic fuels} were defined as x-To-Liquid (xTL) processed fuels, within the scope of ``Fischer-Tropsch fuels that are produced by gasification of either coal, natural gas or biomass". Comparatively, similar definitions are given in \cite{eia2006} and \cite{iea2007}, except that the latter excludes biomass from the potential feedstocks.
(ii) \textit{Electrofuels} were considered  as a storage capacity of electricity into chemical bonds through so called coal-, biomass- and emission(\ce{CO2})-to-electrofuel (xTE) processes. 
(iii) In accordance with \cite{european2013proposal}, they presented \textit{alternative fuels} as any fuel used as a substitution for fossil oil sources in the energy supply with no specific restrictions regarding the feedstock.

More recently, \citet{Stancin2020} has shown that scientific research about ``\textit{alternative fuels}" has experienced a significant impulse since the 2000s. Therefore, in the continuity of \cite{ridjan2016}, we performed a similar literature review in order to update the terminology used by the scientific community. Hence, the current work extends the previous review to cover publications from 2015 to 2020. The same methodology was followed, through a manual identification, and the same search terms were looked for in Scopus: the first search screened the terms of interest (\ie synthetic fuels or electrofuels) complemented by narrowing-down terms (\eg “alternative fuels”, “hydrocarbon”, “biomass”, or “ammonia”). This initial search resulted in 251 documents. Then, after screening the abstract (and, if necessary, the articles themselves), 75 relevant documents were selected as they used the searched terminology and gave an actual definition as well as identified the production process. The non-relevant documents usually contained information that fall out of the scope of the current work, \eg information on specific catalysts. Out of the relevant documents, the main outcomes concern the relative majority of ``synthetic fuels", the emergence of ``electrofuels" and the scattered diversity of other terms.

Where terms like ``synthetic gas", ``electrofuels" or ``synthetic natural gas" account for 19\%, 13\% and 8\% of the additional relevant literature, respectively, ``synthetic fuels" is used in 35\% of the case to cover fuels produced from a wide variety of primary resources (renewable or not) and through various conversion processes. To pick only a few examples, its usage ranged from liquid hydrocarbons produced from synthesis gas through the Fischer-Tropsch process \cite{haarlemmer2014,Trieb2018} to ammonia from the Haber-Bosch process \cite{bargiacchi2019} or biomass gasification or pyrolysis \cite{monaco2018, rao2015}. As a consequence of an increasing share of VRES in the energy system, ``electrofuels" are gaining more popularity: where this term was used in 4\% of the relevant literature between 2006 and 2014 \cite{ridjan2016}, it is now present in 13\% of the additional relevant literature reviewed for this work. These fuels are considered as being produced mainly based on the electricity provided by VRES \cite{brynolf2018}. Besides water electrolysis, this electricity also supplies processes like \ce{CO2} hydrogenation \cite{brynolf2018,Decker2019,Pearson2012} or co-electrolysis of \ce{CO2} and \ce{H2O} \cite{hanggi2019,larsson2015}. Aside the two aforementioned keywords, the literature makes use of a scattered diversity of terms that aim at precising the features of the considered fuel: for instance, ``solar fuel", ``power gas" or ``green synthetic fuel". However, out of context, words like ``green fuel", ``advanced fuel" or ``solar fuel" may be misunderstood because of their multiple interpretations \cite{ridjan2016}.

Consequently, in the context of the energy transition where new use of primary resources and conversion technologies are booming, we encompass in this work all the resources and conversion processes one might consider in the production of fuels. The objective of this taxonomy is to fit with the long-term perspective of this transition (\eg carbon-neutrality, defossilisation) while being adapted for the current context.  We do not only focus on carbon-based fuels nor on specific end-use sectors (\eg transport), contrary to what is mostly seen in the literature. We propose a comprehensive and harmonised taxonomy, without having the ambition to apply strict classification methodologies (\eg questionnaires, text-mining of corpora of the domain) \cite{ncbi_fossilfuel}.  %This aims to set the key distinguishing features of fuels to be able to clearly picture the technological implications behind their labels.

\section{Classification and definition of the fuels}
\label{sec:definition}
%To meet climate objectives while ensuring energy supply, any energy carrier or conversion technologies of an energy system aims at being sustainable. Thereon, European Commission defines ``sustainable development as a development that meets the needs of current generations without compromising the ability of future generations to meet theirs" \cite{eu_sustainable}.
Given the difficulty to quantify and circumscribe the sustainability of a fuel, we decided here to suggest classification and definition of the fuels regardless of this feature. As described below, we opted for a focus on the energy balance of a fuel with regard to its chemical building blocks and the origin of the energy to supply the conversion and production process, without having the ambition to quantify the sustainability of the different fuels.

An essential distinction has to be pointed out between renewable and non-renewable fuels. In accordance with \cite{eu2003directive}, \textit{renewable fuels are fuels produced from renewable energy sources}. Renewable energy sources are non-fossil sources that are naturally replenished on a human timescale (wind, solar, geothermal, wave, tidal, hydro-power, biomass) \cite{ellabban2014}. A fuel can be defined as ``renewable" only when it is based on renewable sources. To be ``sustainable", a ``renewable fuel" should not increase the concentration of \ce{CO2} in the atmosphere \cite{Dincer2020}, among other things \cite{eu_sustainable}.
%and components as well as the energy necessary to its production come from renewable sources.
On the contrary, non-renewable fuels are those that are not based on renewable sources. In practice, these cluster fuels produced from fossil sources (\eg coal, natural gas, oil). 
%There is still a long way towards a fully renewable world. In 2018, fossil-based fuels accounted for 81.2\% of the world primary energy supply \cite{iea_fossil}. 
Finally, the nuclear-based fuels are considered separately. Even though nuclear power plants do not emit \ce{CO2} to produce electricity or heat, uranium is not a renewable resource.

%While the resources are simple to simplify, the pathways can be used by different resources and thus harder to retrace.

Given these definitions and looking at the primary resources, the distinction is clear between renewable and non-renewable sources. However, the production and end-use fuels are harder to group as they can use both renewable and non-renewable resources. In other words, the same end-use fuel can be produced from different resources by different processes, which explains the ambiguity in the definition of these fuels. As illustrated on \Cref{fig:synfuels_paths}, hydrogen is a good example as it can be produced from fossil or renewable resources and by different processes: for instance, (i) steam reforming of fossil natural gas (75\% of the current global production of hydrogen today \cite{iea2019hydrogen}); (ii) partial oxidation of heavy hydrocarbons; (iii) coal gasification; (iv) electrolysis with electricity generated from fossil fuels; (v) electrolysis with electricity generated from renewable sources; (vi) thermochemical conversion of ligno-cellulosic biomass; (vii) steam reforming of biogas or even (viii) thermochemical water splitting using heat from concentrated solar power or nuclear wastes \cite{acar2014}.

Therefore, besides the aforementioned renewable/non-renewable distinction, we propose to categorise fuels in three groups whose definitions are broad enough to capture the diversity of the fuels while maintaining a simplified representation: (i) \textit{Biofuels} as the absolute majority of the energy balance of their production consists of biomass; (ii) \textit{Electrofuels}, similarly to biofuels, but with electricity instead of biomass. \textit{A priori}, it is complicated to give an exact number to quantify this ``majority" of the energy balance but we propose a solution in \Cref{sec:discussion} to remove this ambiguity. (iii) \textit{Synthetic fuels} when the end-use fuel results from the upgrade of another fuel to improve physical and/or chemical characteristics (\eg to increase the volumetric energy density). \Cref{fig:synfuels_paths} illustrates this taxonomy by representing the physical, not lexical, link between different kinds of concepts (\eg resource, conversion and fuel).

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.9\textwidth]{taxonomy.pdf}
    \caption{Overview of the main pathways to produce fuels. The boxes in the ``conversion" category illustrates the terminology to use in the different cases: first, a fuel can be defined as \emph{bio} or \emph{electro} if the absolute majority of the energy balance of its production consists of biomass or electricity, respectively. Then, a fuel can be defined as \emph{synthetic} if it has been obtained through a synthesis process from another fuel, regardless of the feedstock. The box "Fuels" gives some examples to illustrate the terminology used and applied to very common fuels.}
    \label{fig:synfuels_paths}
\end{figure}

% \textit{Fossil} pathways account, nowadays, for most of the fuels. They are mainly refined from fossil hydrocarbons.  Thus, we define ``\textit{fossil} fuels" as: \textit{fuel based on fossil resources.}

\textit{Biofuels} are those based on biomass as their chemical building blocks and energy supply for their own conversion process (\Cref{fig:synfuels_paths}). Biomass is mainly harvested in a solid form and has a lower energy density than fossil hydrocarbons. Thus, the ``bio" processes aim at concentrating the energy and converting it into a fuel more convenient to use (\eg pellet from woody wastes). %liquid or gaseous fuels.
Similarly to the \citet{OEERE_biofuel}, in the literature, biofuels mainly target liquid or gaseous transport fuels (\eg biodiesel and bioethanol) and imply a sustainable production \cite{eu_biofuel, pottering2009directive}. The definition given by official institutions (\eg the FAO or the European Commission) is aligned with the ISO 16559:2014: ``biofuel is a solid, liquid or gaseous fuel produced directly or indirectly from biomass" \cite{eu_biofuels, FAO_biofuel, ISO16559}. To add a quantitative and measurable aspect in the terminology, we propose the following definition: \textit{biofuels are fuels produced from biomass as the major component of their energy balance}. Even if our terminology lacks the subtlety brought by terms like ``woodfuels", ``agrofuels", or ``municipal by-products", that allow to highlight the source of biomass at stake, provided by the FAO \cite{FAO_biofuel}, this terminology aims at harmonising and desambiguating the meaning of the general term ``biofuel".

%According to the European Commission, biofuels are ``liquid or gaseous transport fuels such as biodiesel and bioethanol which are made from biomass. For biofuels to reduce greenhouse gas emissions without adversely affecting the environment or social sustainability, they must be produced in a sustainable way" \cite{eu_biofuel}. This is also stated in the directive of the European Parliament: ``(10) Biofuel production should be sustainable" \cite{pottering2009directive}. However, this definition is too restrictive as it does not include solid biofuels (\eg bio-char that could replace coke in steel production \cite{mousa2016biomass}) nor biofuels used for other purposes than transport (\ie heat and power). 
%Definition of biomass : ``\textit{the biodegradable fraction of products, waste and residues from agriculture (including vegetal and animal substances), forestry and related industries, as well as the biodegradable fraction of industrial and municipal waste}'' \cite{eu2001directive}.

%For example, the pyrolysis, amongst others, produces coal-chars which is an illustration of a higher energy density fuel produced based on thermo-chemical conversion, or biomethanation which transforms solid digestable biomass into biogas.

\textit{Electrofuels} are, according to \citet{JRC_alternative}, fuels whose energy content comes from renewable electricity. On top of that, \citet{goldmann2018} even add that they should be carbon-neutral and serve as a storage capacity of electricity. To be more comprehensive, we define \textit{electrofuels as fuels produced from electricity}, as the resource used to produce this electricity may vary. This electricity is transformed into hydrogen through electrolysis. This hydrogen can be upgraded in more complex fuels when combined with other molecules (\eg carbon mono/dioxide or nitrogen). Electrofuels group all the fuels where the absolute majority of the energy contained into the fuel comes from electricity. In this sense, it is important to point out that ``electrofuels" do not especially imply ``renewable fuels" (\Cref{fig:synfuels_paths}). For instance, the electricity needed to produce \ce{H2} through an electrolyser can be generated from renewable or non-renewable sources \cite{bhandari2014}, even if the most sustainable way to produce electrofuels remains through the electricity produced in excess from VRES.

\textit{Synthetic fuels are fuels which have been obtained through a synthesis process from another fuel}. It encompasses upgrading processes, such as methane produced through the Sabatier process or longer hydrocarbons based on syngas through the Fischer-Tropsch process. We extend the definition of the European Commission and Energy Information Agency who consider synthetic fuels as ``any liquid fuel obtained from coal, natural gas or biomass" \cite{eia2006, eu_synthetic}. By extending this definition and focusing on the synthesis operation, similarly to \citet{speight2020synthetic}, we are able to include emerging fuels like ammonia from the Haber-Bosch process. Contrary to biofuels and electrofuels, the resources used for producing synthetic fuels are not strictly identified. For instance, electrolysis-produced \ce{H2} can be upgraded with synthesis gas from the gasification of biomass to form methanol \cite{mignard2008}. Another example could be the production of ammonia via the synthesis of \ce{H2} from methane-reforming. Consequently, defining a fuel only as ``synthetic" gives a reduced insight about the fuel, especially about its ``renewability". Therefore, we recommend to avoid using ``synthetic fuel". In the former example, one would then rather define the produced methanol as a ``bio-electrofuel" given that its main sources are electricity and biomass. 
% To summarise, we define \textit{synthetic as a generic, and therefore less specific, term that gathers all the aforementioned ones (\ie fossil, bio or electro). Therefore, we recommend to use ``synthetic'' for fuels produced from another one and only when the origin of the resources is not directly identifiable.}
%These fuels cannot be included in the other groups as the used fuel may be produced from different pathways. 
%To summarise, we define synthetic fuel as \textit{a fuel produced from another one; and for which fossil, bio or electro cannot be used.}
%To summarise, we define synthetic fuels as \textit{a fuel produced from another one, where the energy origin used for the fuel cannot be clearly specified. This term is necessary as a synthetic fuel can be produced from a same process but with different resources.}

Finally, besides these recommended terms in the context of the energy transition, others are mainly misleading: ``alternative" or ``non-conventional" fuels. These terms are defined in opposition to another kind of fuels, mostly fossil. However, alternatives of today will be conventions of tomorrow. This means that depending on the ``conventional" of the context, these terms will represent different fuels. Because of this implied ambiguity, we recommend not to use these words as their interpretation is case and time sensitive.

\section{Discussion}
\label{sec:discussion}
% Representing a simple picture with well discerned paths is challenging, as most of the processes can be hybridised, such as gasification with renewable hydrogen injection \textbf{Source}, or electrolysis which can either use fossil or renewable electricity.
In the urging necessity for an energy transition and the persisting need for fuels, technologies to convert primary resources into end-use fuels are rapidly multiplying. In this context, we suggest comprehensive terminology and classification to characterise the fuels. Given the limited insight provided by the widely-used term ``synthetic fuels", we recommend to use more specific terms like ``biofuels", ``electrofuels", or even hybridised terms (\eg ``bio-electrofuels"). These allow to highlight the contribution of biomass and electricity in the energy balance of the fuel.

Giving an exact share of biomass (or electricity) above which a fuel can be defined as biofuel (or electrofuel) would open an endless debate. Therefore, to emphasise the effort to step away from fossil fuels, we suggest to specify the contribution of biomass (or electricity) within the energy balance of the production of the fuel looking at both the feedstock and the energy to supply the conversion or production process. This would give, for instance, ``bio(25) methanol" if 25\% of the energy balance to produce the methanol comes from biomass (\eg biomass gasification). Another example could be ``electro(40) ammonia" if 40\% of the energy balance comes from electricity (\eg water electrolysis, air-captured nitrogen and power supply of the Haber-Bosch process). As electricity can be produced from a wide variety of sources (\Cref{fig:synfuels_paths}), we recommend to give insights on the composition of this electricity (\eg from wind, solar or the grid mix).

To fully embrace these new terminology and classification, industries and academics will have to switch of perspective as studies on energy balance of biofuels or electrofuels too often consider the energy distribution over the different production steps or the feedstock only rather than over all the resources at stake \cite{OConnell2019,slade2013micro}.

% terms like ``renewable fuels", ``electrofuels", ``biofuels" or ``synthetic fuels" are more and more popular in the literature and used for a vast variety of fuels. These terms imply specific use of primary resources or process. Therefore, they shall be used accurately in order to drive consistent policies and discussions between academia and industries.

% Through a comprehensive literature review, we first identified the top-used term (\ie ``synthetic fuel"), the emergence of new ones like ``electrofuel" and the wide variety of other terms. Then, on the purpose for setting a comprehensive terminology and classification, we pointed out different categories of fuels.

% First of all, the definition of ``renewable fuels" was clearly reminded as fuels produced from non-fossil sources that are naturally replenished on a human timescale. As we aim at a pathway towards sustainability where fuels will keep playing a significant role, the renewable fuels shall gain more and more shares in the global energy mix. Therefore, we recommend to use ``renewable fuel" when it is not produced from fossil sources. Although, we also point out that ``renewable" does not directly mean ``sustainable" as this term implies many more aspects than ``Affordable and Clean Energy" \cite{eu_sustainable}, especially when we consider biofuels (\eg use of lands, indirect impacts on environment).

% Then, besides this first distinction between renewable and non-renewable, mostly fossil fuels, we identified three terms to describe them: ``biofuels", ``electrofuels" and ``synthetic fuels". On one hand, the first two really focus on the resource and the energy balance to produce the fuel. Fuels are defined as ``bio" or ``electro" if a majority of their energy balance (\ie feedstock and energy supply for the conversion processes) consists respectively of biomass or electricity. On the other hand, the last one targets the synthesis process from which a fuel can be produced based on another (or other) one(s). As observed in the literature, this includes many different fuels, from many different processes. Therefore, we encourage to use specific terms like ``biofuels" or ``electrofuels", even hybridised term (\eg ``bio-electrofuels") to give more insight to the reader about the technical implications behind the concerned fuel.

% Finally, we recommended not to use case and time sensitive terms like ``alternative" or ``non-conventional" fuels, given their ambiguous dependency on the alternatives or conventions which these fuels are opposed to.

% For the purpose of being concise and comprehensive, it is necessary to cluster the fuels in wide aggregates. The objective was to clarify the terminology and propose a classification, mainly based on the used resources. Consequently, there is not a straightforward label corresponding to one fuel. One will have to consider the resources and different process to label the end-use fuel.

% As aforementioned, the innovation is running in the energy sector. Therefore, even if we aim at setting a comprehensive and time insensitive terminology and classification, we do not exclude that, in the coming years, new fuels might emerge and would not fit in our framework.

% As the objective was to clarify the terminology while keeping a simple classification, some processes have been aggregated or not mentioned. As a trade-off between overview and technical insights, technologies were aggregated when their process and inputs were similar, such as thermo-chemical processes, that gather gasification, together with slow or fast pyrolysis. Some technologies are under investigation, such as hydrogen production based on high temperature water cracking from concentrated solar \cite{steinfeld2005solar} or nuclear \cite{fujiwara2008hydrogen} energy.

### 2024-02-19 chap_atom_mol/chap_atom_mol.tex
Merge taxonomy paper + Implement Hervé's comments on case_study

@@
 On top of purely techno-economic aspects, non-exhaustively mentioned beforehand, there are other ethical, societal or even political considerations to account for when addressing this question.
-Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} in the Belgian energy landscape, on a strictly techno-economical point of view.
+Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} in the Belgian energy landscape, on a strictly techno-economical point of view with a cost-based optimisation.
 This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in the Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
@@
 Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.\\
-
-{\color{red} \textbf{Compare here with EnergyVille - PATHS2050.}}\\
 
@@
 \end{figure}
+
+\section{Discussion}
+\label{sec:atom_mol:discuss}
+
+{\color{red} \textbf{Compare here with EnergyVille - PATHS2050.}}\\
+
+{\color{red} \textbf{Discuss the question of the availability SVB \& CL}}\\

### 2024-02-19 chap_case_study/chap_case_study.tex
Merge taxonomy paper + Implement Hervé's comments on case_study

--- 2024-02-06 chap_case_study/chap_case_study.tex
@@
 Moreover, this case study and the subsequent analyses can be transferred - to some extent - to other industrialised countries highly dependent on fossil fuels with limited local renewable potentials (\eg the Netherlands or Germany) \cite{dommisse2020modelling}.
-This chapter presents the different demands to satisfy, with a particular focus on the non-energy demand, as well as the resources available and the conversion technologies to supply those.
+This chapter presents the different demands to satisfy, with a particular focus on the non-energy demand, as well as the resources available and the conversion technologies to supply these demands.
 For a comprehensive understanding and detailed descriptions of the technologies, please refer to the documentation \cite{readthedocs_pathway}.
@@
 \label{sec:cs:contributions}
-First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heart and mobility), the \acrfull{NED} is often not considered.
+First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heat and mobility), the \acrfull{NED} is often not considered.
 The latter is defined as ‘\textit{energy products used as raw materials in the different sectors; that is not consumed as a fuel or transformed into another fuel}’ \cite{Eurostat2019}.
@@
 To minimise the total cost of the system, the model simply selected the cheapest between the two resources (\ie natural gas).
-This work goes one step further and accounts for the \gls{NED} as a demand of three commodities (\ie \gls{HVC}, ammonia and methanol) as well as with associated production technologies.
+This work goes one step further and accounts for the \gls{NED} as a demand for three commodities (\ie \gls{HVC}, ammonia and methanol) as well as the associated production technologies.
 This allows bringing the non-energy sector to a similar level of details as the other sectors.
@@
 
-Second, given the focus of this work on the electrofuels, the case study includes a more explicit representation of them: where the previous definition of the case study considered only renewable hydrogen and methane \cite{limpens2021generating}, now, e-ammonia and e-methanol (as well as their fossil-based equivalents), are implemented in the case study.
+Second, given the focus of this work on the electrofuels, the case study includes a more detailed representation of them: where the previous definition of the case study considered only renewable hydrogen and methane \cite{limpens2021generating}, now, e-ammonia and e-methanol (as well as their fossil-based equivalents), are implemented in the case study.
 As detailed later on, these electrofuels are considered as renewable in the sense that their \gls{GWP} is zero.
-This more explicit representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.
-For instance, considering the case of ammonia, on the top of the import routes, the Haber-Bosch process ,on one hand, is accounted for to supply this molecule.
+This more detailed representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.
+For instance, considering the case of ammonia, on the top of the import routes, on the one hand, the Haber-Bosch process is accounted for to supply this molecule.
 On the other hand, ammonia-driven \gls{CCGT} or ammonia-cracking-to-hydrogen are included as ways to consume it.
 
-Third, as nuclear energy could be a real game-changer in the energy transition worldwide \cite{IEA2022nuclear}, and especially in Belgium, this thesis has integrated the uncertain potential to install \gls{SMR} from 2040 onward.
-
-Fourth, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.
+Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.
 Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.
 
-Finally, to a smaller extent, this work includes updated values for some parameters compared to the work of \citet{limpens2021generating}.
-The main change concerns the cost and performance of private mobility vehicles, which is specifically a key components in the European \cite{biresselioglu2018electric} and Belgian \cite{BFP_mob} energy transitions.
-Based on the work of \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
-Besides their lower efficiencies (\ie about 50\% less efficient), \gls{FC} car had the lion share of the private mobility, compared to \gls{BEV} given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter, in previous works \gls{limpens2021generating,rixhon2021role}.
+Fourth, as nuclear energy could be a real game-changer in the energy transition worldwide \cite{IEA2022nuclear}, and especially in Belgium, this thesis has integrated the uncertain decision to install \gls{SMR} from 2040 onward.
+
+Finally, this work includes updated values for some parameters compared to the work of \citet{limpens2021generating}.
+The main change concerns the cost and performance of private mobility vehicles, which is a key components in the European \cite{biresselioglu2018electric} and Belgian \cite{BFP_mob} energy transitions.
+Based on the work of the \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
+In previous works \cite{limpens2021generating,rixhon2021role}, despite their lower efficiencies (\ie about 50\% less efficient), \gls{FC} cars had the lion share of the private mobility compared to battery electric vehicles(BEV) given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter.
 In these, the more limited potential to import electricity from abroad and produce it locally via \gls{VRES} forced the model to rather electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.
-To align with other similar works on the modelling of whole-energy system \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
+To align with other similar works on the modelling of whole-energy systems \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
 Regarding \gls{BEV}, while the CAPEX has been kept unchanged, the efficiency and the battery capacity, \ie the range, have been increased.
@@
 \label{sec:cs:demand}
-End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative years of the transition, in order to account for their daily or seasonal variability \cite{Limpens2020,limpens2021generating}.
+End-use demands, exogenously imposed as inputs to the model, are characterised by yearly quantities to satisfy and are also distributed over the different hours of each representative year of the transition, to account for their daily or seasonal variability \cite{Limpens2020,limpens2021generating}.
 In this work, the yearly end-use demands (EUD) for all sectors are calculated from the forecast proposed by the European Commission for Belgium (Appendix 2 in report \cite{EuropeanCommission2021}).
@@
 The \gls{NED} currently represents around 20\% of the final energy consumption in Belgium \cite{FPSEconomy2021}.
-This section summarises the rationale of adding a higher level of details to the\gls{NED} compared to what was done in the previous version of the case study \cite{limpens2021generating}.
+This section summarises the rationale of adding a higher level of details to the \gls{NED} compared to what was done in the previous version of the case study \cite{limpens2021generating}.
 Then, it explains the methodology used to quantify this demand.\\
@@
 
-Over the recent history, there has been a relatively constant evolution of three main categories of the final consumption for non-energy use in Belgium, \cite{statbel_NED_2019}: (i) naphtha and \gls{LPG} (between 59\% and 67\% of the total final consumption, around 59.4\,TWh in 2019), (ii) fossil gas (between 9\% and 14\%, 11.8\,TWh in 2019), and (iii) others (\ie bitumen, coal tar and other oil products) (between 21\% and 28\%).
+Over the recent history, there has been a relatively constant share of three main categories of the final consumption for non-energy use in Belgium \cite{statbel_NED_2019} (see Figure \ref{fig:cs_NED_hist}): (i) naphtha and \gls{LPG} (between 59\% and 67\% of the total final consumption, around 59.4\,TWh in 2019), (ii) fossil gas (between 9\% and 14\%, 11.8\,TWh in 2019), and (iii) others (\ie bitumen, coal tar and other oil products) (between 21\% and 28\%).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{NED_hist.pdf}
+\caption{Historical data for the non-energy use in Belgium \cite{statbel_NED_2019}.
+``Other oil products" account for tar and sulphur.
+This group also includes by-products of the oil-refining industry like aromatics (\eg \acrfull{BTX}) and olefins (\eg propylene).
+Graph adapted from \cite{rixhon2021comprehensive}.}
+\label{fig:cs_NED_hist}
+\end{figure}
+
 Naphtha and \gls{LPG} are consumed in a naphtha cracker, which results in ethylene and propylene, what will be considered as \gls{HVC} in the rest of this work.
-Similarly, fossil gas, as non-energy carrier, is used in steam methane reformer to produce the required hydrogen to the synthesis process of ammonia.
-The small shares of bitumen and coal tar are respectively used for roadworks and to produce synthetic gas through gasification.
+Similarly, fossil gas, as non-energy carrier, is used in steam methane reformers to produce the required hydrogen for the synthesis of ammonia.
+The small shares of bitumen and coal tar are used for roadworks and to produce synthetic gas through gasification, respectively.
 Finally \og other oil products'' take into account, indistinguishably, tar and sulphur as well as by-products of the refineries (\eg \gls{BTX}).
@@
 \noindent
-The non-energy demand studied in this analysis focuses on the chemical industry (more than 90\% of the non-energy use in Belgium) and, similarly to other studies \cite{IEA2018_petrochemicals, daioglou2014energy}, is split between the three aforementioned main groups of products (\ie \gls{HVC}, ammonia and methanol).
+The non-energy demand studied in this work focuses on the chemical industry (more than 90\% of the non-energy use in Belgium) and, similarly to other studies \cite{IEA2018_petrochemicals, daioglou2014energy}, is split between the three aforementioned main groups of products (\ie \gls{HVC}, ammonia and methanol).
 Before describing these three demands, this study excludes bitumen, coal tar and \og other oil products\fg.
-The first two represent marginal shares of the current non-energy use (\ie 4\% and 5\% respectively) in such a way that they should not affect the conclusions provided by this study.
+The first two represent marginal shares of the current non-energy use (\ie 4\% and 5\% respectively) in such a way that they should not affect the conclusions.
 As described previously, the latest are mostly by-products from refineries that the system uses because they are available.
@@
 Even if a significant part of \gls{HVC} produced in Belgium is not locally consumed, this demand has been set based on the assumption that Belgium will keep its industrial activity in this sector.
-In other words, this assumption does not deduce the part of local production  of \gls{HVC} being then exported (and not consumed locally), unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
+In other words, we do not deduce the part of local production of \gls{HVC} being then exported (and not consumed locally), unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
 Therefore, the actual demand of \gls{HVC} is inferred from the consumption of naphtha and \gls{LPG} as non-energy use as well as energy-carrier in the chemical and petrochemical industries \cite{statbel_NED_2019}.
-This assumption is based on the fact that, in the conversion processes to produce \gls{HVC} from naphtha or \gls{LPG}, these fuels also serve as energy-carrier to supply the process itself.
-Then, given the respective efficiencies (1.83t$_{\text{naphtha}}$/t$_{\text{HVC}}$ and 1.67t$_{\text{LPG}}$/t$_{\text{HVC}}$) \cite{IEA2018_petrochemicals}, the current demand of \gls{HVC} is estimated equal to 3069\,kt, without making distinctions between the different chemicals (\ie ethylene, propylene and \gls{BTX}).
+This assumption is based on the fact that, in the conversion processes to produce \gls{HVC} from naphtha or \gls{LPG}, these fuels also serve as energy-carrier to supply the processes themselves.
+Then, given the respective efficiencies (1.83t$_{\text{naphtha}}$/t$_{\text{HVC}}$ and 1.67t$_{\text{LPG}}$/t$_{\text{HVC}}$) \cite{IEA2018_petrochemicals}, the current yearly demand of \gls{HVC} is estimated equal to 3069\,kt, without making distinctions between the different chemicals (\ie ethylene, propylene and \gls{BTX}).
 
 The ammonia sector in Belgium is quite different: the country locally produces and imports ammonia much more than it exports it.
-Thanks to a database from the United Nations \cite{UN_statistics} and the National Bank of Belgium, it has been identified that, over the last ten years, Belgium has imported 1010\,kt of ammonia, exported 105\,kt and locally produced 990\,kt on average.
+Thanks to a database from the United Nations \cite{UN_statistics} and the National Bank of Belgium, it has been identified that, over the last ten years, Belgium, on average, has imported 1010\,kt of ammonia, exported 105\,kt and locally produced 990\,kt per year.
 Therefore, on top of the local production, the net import (\ie import minus export) is also included in this non-energy demand.
-This gives a current demand of 1895\,kt of ammonia.
+This gives a current yearly demand of 1895\,kt of ammonia.
 
@@
 Currently, the rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
-This methodology gives a current non-energy demand of methanol of 269\,kt.
+This methodology gives a current yearly non-energy demand of methanol of 269\,kt.
 
-Eventually, after converting these masses of products into energy content (\ie LHV: \gls{HVC} - 47\,MJ/kg, ammonia - 18.8\,MJ/kg and methanol - 19.9\,MJ/kg), two final assumptions are made: (i) the current shares of each of the three commodities (\ie 77.9\%, 19.2\% and 2.9\% for \gls{HVC}, ammonia and methanol, respectively) are supposed to remain unchanged over the transition; (ii) the overall \gls{NED}, in absolute terms, follows the growing rate presented by \citet{EuropeanCommission2016}, \ie around 7\% between 2020 and 2050.
+Eventually, after converting these masses of products into energy contents (\ie LHV: \gls{HVC} - 47\,MJ/kg, ammonia - 18.8\,MJ/kg and methanol - 19.9\,MJ/kg), we assume that the current shares of each of the three commodities (\ie 77.9\%, 19.2\% and 2.9\% for \gls{HVC}, ammonia and methanol, respectively) are supposed to remain unchanged over the transition.
 
@@
 \label{subsec:cs:EUD_forecast}
-In their latest report \cite{EuropeanCommission2021}, the European Commission forecasts a significant and abrupt increase of the \gls{NED} compared to their previous report \cite{EuropeanCommission2016}, \ie +80\% over the 2020-2030 time window.
+In its latest report \cite{EuropeanCommission2021}, the European Commission forecasts a significant and abrupt increase of the \gls{NED} compared to their previous report \cite{EuropeanCommission2016}, \ie +80\% over the 2020-2030 time window.
 Given this discrepancy that is unsubstantiated and specific to the case of Belgium, the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}.
@@
 This is explained by a better insulation of buildings and an improved efficiency of industrial processes.
-Regarding the center graph of \Cref{fig:cs_demands}, it is the aggregation of the same data as in the left graph but per category, rather than per sector, with the non-energy demand being associated with the industry.
-This illustrates how industrialised is Belgium, compared to households and services, and, consequently, highly energy-intensive.
-The right graph of \Cref{fig:cs_demands} gives the passenger and the freight mobility.
-The sharp increase from 2020 to 2025 is due the COVID-crisis that has significantly reduced these demands in 2020.
+Regarding the top-right graph of \Cref{fig:cs_demands}, it is the aggregation of the same data as in the top-left graph but per economic sector, rather than per energy sector, with the non-energy demand being associated with the industry.
+This illustrates how industrialised Belgium is, compared to households and services, and, consequently, highly energy-intensive.
+The bottom graph of \Cref{fig:cs_demands} gives the passenger and the freight mobility.
+The sharp increase from 2020 to 2025 is due the COVID-crisis that led to significantly reduced demands in 2020.
 As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for the fluctuating parts of electricity and low-temperature heating demands \cite{Limpens2020}.
 A daily time series is used for the passenger mobility and applied similarly to every days of the year.
-Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and constant share of electricity and low-temperature heat demands), the yearly demand is distributed uniformly over the different hours of the year.
+Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and the share of electricity and low-temperature heat demands that are considered as constant over the year), the yearly demand is distributed uniformly over the different hours of the year.
 
@@
 \caption{EnergyScope splits the whole-energy system end-use demands (EUD) into two sets: (non)-energy and transport-related.
-This figure presents the nominal values of each of these demands.
-In the center graph, the non-energy demand has been fully associated with the industrial demand.
-As detailed previously, the non-energy demand is expressed in tons of physical products (\ie \glsxtrfull{HVC}, ammonia and methanol) and then translated into their respective energy equivalent, in TWh.}
+This figure presents the yearly values of each of these demands.
+In the top-right graph, the non-energy demand has been fully associated with the industrial demand.
+This highlights the significant level of industrialisation of Belgium compared to the other sectors.}
 \label{fig:cs_demands}
@@
 First, the evolution of the respective costs of purchasing is presented (see \Cref{fig:cs_resources_cost}).
-Regarding ``renewable electrofuels'', these are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis\cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
-Then, besides their cost, the resources are either limited or unlimited in terms of availability and either renewable or not.
+Regarding ``renewable electrofuels'', the costs are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis \cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
+Then, besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
 The limitation in terms of availability can be direct or indirect.
-On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are arbitrarily limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use.
-On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and nuclear power plants (6\,GW), respectively.
+On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are directly limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use.
+The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh.
+On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
 In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}.
@@
 In practice, it means that their \gls{GWP} is assumed to be zero in the model.
-Regarding specifically these electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco).
+Regarding these electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco).
 As the amount to import from each of these locations is hard to forecast, the current work considers the average cost between the different locations.
@@
 \caption{Cost of purchasing the different resources.
-Besides the free local renewables (\ie sun, wind and hydro) limited by technical potentials, EnergyScope accounts for renewable energy carriers and their respective fossil counterparts (left and center graphs).
+Besides the free local renewables (\ie sun, wind and hydro) limited by technical potentials, EnergyScope accounts for renewable energy carriers and their respective fossil counterparts (top-left and top-right graphs).
 These fuels can be imported from abroad without limitation on their availability.
@@
 As the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, for instance, technologies are implemented to convert these resources into the end-use demands.
-Besides their CAPEX, OPEX and lifetime defined in Section \ref{sec:meth:ES}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) exhibit their own charge/discharge losses.
+Besides their CAPEX, OPEX and lifetime defined in Section \ref{sec:meth:ES}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) have charge/discharge losses.
 There are also infrastructure technologies.
 They encompass, for instance, the power grid, the \gls{DHN} or technologies to produce intermediate energy carriers (\eg wood pyrolysis, biomethanolation or steam methane reforming to produce hydrogen).
-Not digging into too much details about the exhaustive list of these technologies presented in previous works \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} then the technologies to supply the \glsxtrfull{NED}.
+Not digging into too much details about the exhaustive list of these technologies presented in a previous work \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} and the technologies to supply the \glsxtrfull{NED}.
 
@@
 
-A specific attention is to put on the implementation of \gls{SMR} whereas the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
+A specific attention is put on the implementation of \gls{SMR} as the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
 Similarly to the analysis of \citet{PATHS2050}, a Belgian consortium for energy research, and in line with the Belgian Nuclear Research Centre (SCK-CEN) \cite{SCK-CEN_SMR}, \gls{SMR} are implemented with the features listed in \Cref{tab:SMR_features}.
-Where most of the features are similar to conventional nuclear power plants, it differs from these on two main points: their potential year start, 2040, and their flexibility.
-Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity independently at any hour of each representative year.
+Where most of the features are similar to conventional nuclear power plants, they differ from these on two main points: their potential year start, 2040, and their flexibility.
+Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity at any hour of each representative year.
 Here, we simplify SMRs as only producing electricity and disregard the heat produced by the nuclear reaction.
@@
 Not including here the cost of integrating a technology in the system (\eg reinforcement of the grid and storage capacities for \gls{VRES}), the \gls{LCOE} aims at aggregating and normalizing the CAPEX and OPEX of technologies providing a common commodity, \ie electricity.
-Compared to the other flexible generation units (\ie \gls{CCGT}), \gls{SMR} is significantly more cost-effective.
-Besides being about six times more capital-intensive in €/kW, the investment is amortized over a longer expected lifetime (\ie 60 years versus 25 for \gls{CCGT}).
+Compared to the other flexible generation units, \gls{SMR} is significantly more cost-effective.
+Besides being about six times more capital-intensive in €/kW, the investment is amortized over a longer expected lifetime (\ie 60 years).
 Moreover, the cost of purchasing uranium driving \gls{SMR} is expected to remain stable and low whereas the expected increase of the cost of purchasing fossil fuels dominates the \gls{LCOE} of \gls{CCGT}.
@@
 Gray and black curves are related to technologies runnning on renewable and fossil resources, respectively.
-Where \gls{SMR} is cheaper than the other flexible options, \gls{CCGT} running on e-ammonia is, a priori, cheaper than its e-methane alternative.}
+\gls{SMR} is cheaper than the other flexible options.
+\gls{CCGT} running on e-ammonia is cheaper than its e-methane alternative.}
 \label{fig:LCOE}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{NED_tech.pdf}
+\includegraphics[width=\textwidth]{NED_tech.pdf}
 \caption{Schematic view of the different resources able to produce \gls{HVC}, ammonia and methanol with their related conversion technologies (including energy efficiency and their CAPEX - in €/kW of final molecules).
 Values stand for 2035.
-Graph from \cite{rixhon2021comprehensive}}
+Graph from \cite{rixhon2021comprehensive}.}
 \label{fig:NED_tech}
 \end{figure}
+%
+%\myparagraph{The impact of integrating the \gls{NED} in the Belgian whole-energy system}\\
 
-\myparagraph{The impact of integrating the \gls{NED} in the Belgian whole-energy system}\\
-
-\noindent
+%\noindent
 Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium, using EnergyScope TD (see Appendix \ref{app:ESTD}).
@@
 
-For further details on these analyses, the interested reader is invited to refer to aforementioned for further published works.
+For further details on these analyses, the interested reader is invited to refer to previously published works \cite{rixhon2021comprehensive,rixhon2022integration}.
 
@@
 
-The exhaustive list of the parameters accounted in this work is presented in Appendix \ref{app:UC_full}.
+The exhaustive list of the parameters accounted for in this work is presented in Appendix \ref{app:UC_full}.
 
@@
 \multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
-}}  & \multicolumn{2}{c}{\textbf{Relative variation\footnote{\label{foot:nom_val_uncert}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+}}  & \multicolumn{2}{c}{\textbf{Relative variation\footnote{\label{foot:nom_val_uncert}The nominal values for each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
 }}}\\
@@
 \label{sec:cs:CO2-budget}
-In most of the studies carried out on the pathway optimisation of a whole-energy system, a \ce{CO2}-trajectory is \textit{a priori} set to reach carbon-neutrality by 2050.
+In most studies carried out on the pathway optimisation of a whole-energy system, a \ce{CO2}-trajectory is \textit{a priori} set to reach carbon-neutrality by 2050.
 \citet{nerini2017myopic} used the emission trajectory indicated by the UK's Committee on Climate Change in their analysis of the impact of limited foresight to achieve the target of 80\% reduction of \gls{GHG} by 2050 in the United Kingdom.
-In their assessment of the impacts of economy-wide emissions policies in the water-energy-land nexus, \citet{licandeo2023assessing} analysed different \ce{CO2}-trajectories considering more or less severe water scarcity for the US.
+In their assessment of the impacts of economy-wide emission policies in the water-energy-land nexus, \citet{licandeo2023assessing} analysed different \ce{CO2}-trajectories considering more or less severe water scarcity for the US.
 \citet{poncelet2016myopic} with LUSYM (Leuven University SYstem Model) and \citet{PATHS2050} with TIMES-BE also set decreasing emission trajectories in their analysis of respectively the Belgian power sector and whole-energy system.
@@
 This ``systemic framework for addressing global anthropogenic impacts on Earth system'' gives quantitative recommendations about the \ce{CO2} concentration, among others, to maintain ``the stability and resilience of Earth system as a whole'' \cite{richardson2023earth}.
-In their review, \citet{ryberg2020downscaling} identified three main sharing principle categories when considering these safe spaces: \ie \textit{utilitarian}, \textit{egalitarian} and \textit{acquired rights} principles.In a nutshell, the former, mostly applied to the scale of industry sector \cite{ryberg2018bring,brejnrod2017absolute}, aims at maximising the sum of welfare.
+In their review, \citet{ryberg2020downscaling} identified three main sharing principle categories when considering these safe spaces: \ie \textit{utilitarian}, \textit{egalitarian} and \textit{acquired rights} principles.
+In a nutshell, the former, mostly applied to the industry sector \cite{ryberg2018bring,brejnrod2017absolute}, aims at maximising the sum of welfare.
 The second shares the so-called ``budget'' equally among the total population, allocating the same share to each individual \cite{hoff2017bringing,o2018good}.
-Finally, in \textit{acquired rights} principles, also called ``grandfathering'', the sharing is based on \og maintaining that prior emissions increase future emission entitlements\fg  \cite{knight2013grandfathering}.In this thesis, we have chosen the latter principle to allocate the \ce{CO2}-budget to the Belgian transition.
-This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.
+Finally, in \textit{acquired rights} principles, also called ``grandfathering'', the sharing is based on \og maintaining that prior emissions increase future emission entitlements\fg  \cite{knight2013grandfathering}.
+In this thesis, we have chosen the latter principle to allocate the \ce{CO2}-budget to the Belgian transition.
+This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world energy-related emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.
 Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in Eq.\,(\ref{eq:limit_gwp_trans}).
 This is another sign of the urgency to act to mitigate climate change as this 30-year budget represents only 10 years of the current emissions.
+In this approach, the residual emissions in 2050 are not explicitly constrained down to zero.
+This carbon-neutrality is rather implicit given the ambitious \ce{CO2}-budget.
 
@@
 Appendix \ref{app:CO2_budget} compares the emissions trajectory between the REF case and a case (without \gls{SMR}) where the linear decrease is imposed between 2020 and carbon-neutrality in 2050.
+
+\section*{Conclusion}
+\label{sec:cs:conclusion}
+Mainly based on data collected during previous works \cite{Moret2017PhDThesis,limpens2021generating}, this thesis has developed the case study along different axes.
+First, the non-energy demand, that currently represents about 20\% of the final energy consumed in Belgium, has been brought to a level of details similar to the other energy sectors of the system.
+Previously considered as a fixed demand supplied by fossil gas, it has been detailed to account for final-use products (\ie \acrfull{HVC}, ammonia and methanol) and their respective production routes.
+Second, electrofuels are now more detailed in the model as well as their different supply and consumption routes.
+Third, instead of defining \textit{a priori} the \ce{CO2}-trajectory through the transition, this work sets a limit on the overall \ce{CO2}-budget.
+Then, based on the methodology of \citet{Moret2017}, uncertainty ranges are defined for the 5-year step transition from 2020 to 2050.
+Finally, on top of adding the possibility to install small modular reactors (SMR), this thesis accounts for updated values for the \gls{EUD} as well as adjusted assumptions regarding the two key drivers of the transition in the private mobility, \ie battery electric vehicles (BEV) and \acrfull{FC} cars.

### 2024-02-20 appendices/app_taxonomy.tex
First part of intro: transition, VRES, electrofuels, ESOMs

@@
 Even if electricity gains shares through the electrification of the energy demand, it will not entirely displace fuels for three main reasons: storage, infrastructure compatibility and cross-sectorial links.
-Because of their intermittency and space disparity, a deeper integration of variable renewable energy sources (VRES) requires storage and transport in order to supply the energy demand at the right time and in the right place \cite{Brouwer2016,Evans2012,Gallo2016,hall2008,Rosa2017}.
+Because of their intermittency and space disparity, a deeper integration of \gls{VRES} requires storage and transport in order to supply the energy demand at the right time and in the right place \cite{Brouwer2016,Evans2012,Gallo2016,hall2008,Rosa2017}.
 Where typical container of batteries are limited in terms of storage capacity (up to 10~MWh) and present significant costs and self-discharge losses, the energy conversion into fuels provides a more affordable solution for higher storage capacity (from 100~GWh) and longer storage time scales (months to year) \cite{Rosa2017}.
@@
 This highlights the necessity to consider every energy sector rather than focusing all efforts on the power system and even more, to shift towards a multi-vector interconnected energy system.
-In this cross-sectorial approach and with the perspective of increasing shares of VRES, fuels are promising energy carriers in order to maximise the overall system efficiency \cite{mathiesen2015, Stancin2020}.
+In this cross-sectorial approach and with the perspective of increasing shares of \gls{VRES}, fuels are promising energy carriers in order to maximise the overall system efficiency \cite{mathiesen2015, Stancin2020}.
 
@@
 To pick only a few examples, its usage ranged from liquid hydrocarbons produced from synthesis gas through the Fischer-Tropsch process \cite{haarlemmer2014,Trieb2018} to ammonia from the Haber-Bosch process \cite{bargiacchi2019} or biomass gasification or pyrolysis \cite{monaco2018, rao2015}.
-As a consequence of an increasing share of VRES in the energy system, ``electrofuels" are gaining more popularity: where this term was used in 4\% of the relevant literature between 2006 and 2014 \cite{ridjan2016}, it is now present in 13\% of the additional relevant literature reviewed for this work.
-These fuels are considered as being produced mainly based on the electricity provided by VRES \cite{brynolf2018}.
+As a consequence of an increasing share of \gls{VRES} in the energy system, ``electrofuels" are gaining more popularity: where this term was used in 4\% of the relevant literature between 2006 and 2014 \cite{ridjan2016}, it is now present in 13\% of the additional relevant literature reviewed for this work.
+These fuels are considered as being produced mainly based on the electricity provided by \gls{VRES} \cite{brynolf2018}.
 Besides water electrolysis, this electricity also supplies processes like \ce{CO2} hydrogenation \cite{brynolf2018,Decker2019,Pearson2012} or co-electrolysis of \ce{CO2} and \ce{H2O} \cite{hanggi2019,larsson2015}.
@@
 In this sense, it is important to point out that ``electrofuels" do not especially imply ``renewable fuels" (\Cref{fig:synfuels_paths}).
-For instance, the electricity needed to produce \ce{H2} through an electrolyser can be generated from renewable or non-renewable sources \cite{bhandari2014}, even if the most sustainable way to produce electrofuels remains through the electricity produced in excess from VRES.
+For instance, the electricity needed to produce \ce{H2} through an electrolyser can be generated from renewable or non-renewable sources \cite{bhandari2014}, even if the most sustainable way to produce electrofuels remains through the electricity produced in excess from \gls{VRES}.
 

### 2024-02-20 chap_intro_ccl/intro.tex
Merge taxonomy paper + Implement Hervé's comments on case_study
First part of intro: transition, VRES, electrofuels, ESOMs

@@
 
+Consequently, to ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
+Besides behavioral changes, an overall reshape of the energy system is necessary in terms of both primary energy sources, \ie more renewables, and technologies used to convert these resources into the \gls{EUD} (\ie the energy service required by the the final consumer), \ie more efficiency \cite{iea2020world}.
+The former corresponds to a whole ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}) where energy carriers called, in the literature, ``\emph{biofuels}'', ``\emph{electrofuels}'', ``\emph{synthetic fuels}'', ``\emph{renewable fuels}'' or even ``\emph{sustainable fuels}'', will more and more play a crucial role.
+To avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions, we have suggested a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
+In the rest of this thesis, the electro - and bio - fuels are considered as renewable and with no \gls{GWP}.
+
+\begin{figure}[ht!]
+\centering
+\includegraphics[width=0.9\textwidth]{IEA_WEO_2019.png}
+\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario \cite{iea2020world}.}
+\label{fig:intro:IEA_WEO_2019}
+\end{figure}
+
+In the general perspective to decrease the \gls{GWP} of the primary energy mix, \gls{VRES} like wind and solar, have already emerged as the keystone to defossilise the energy system.
+However, their intermittency and space disparity could hold back their vaster integration in the future.
+To address this issue, due to some limitations (\eg range, power, costs) of electricity-focused solutions like \gls{DC} lines, the transport and long-term storage of the renewable electricity produced in excess should be optimised.
+
+This challenge can be tackled by \textit{electrofuels} \cite{rozzi2020}.
+These fuels represent energy carriers where electricity has the major share in the energy balance of the fuel.
+In practice, this electricity is mainly converted into hydrogen (\ie electrolysis) and then potentially upgraded into more complex fuels (\eg methane, methanol or ammonia).
+Even if the share of electricity increases in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during (and after) the energy transition \cite{Ahlgren2012}.
+They offer three main advantages: infrastructure compatibility, storage and capacity to link sectors (\ie from electricity to mobility, heat, or industry).
+Development on electrofuels aims at getting them more and more compatible with existing and mature technologies \cite{Ahlgren2012}.
+An example is carbon-free ammonia-hydrogen blends burned in spark ignition engines \cite{lhuillier2020experimental} or \gls{CHP} applications \cite{pochet202022}.
+With a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
+Besides direct electrification of other sectors (\eg electrical heat pumps, battery electric vehicles), \citet{brown2018synergies} showed that converting power to hydrogen and methane was advantageous at high shares of renewables, in their optimisation of the European whole-energy system.
+Electrofuels have the ability to couple energy and non-energy sectors \cite{Stancin2020}.
+For instance, electricity produced in excess from \gls{VRES} can be converted in ammonia through the Haber-Bosch process and subsequently transformed into fertiliser - coupling the power and industry sectors \cite{verleysen2020can}.
+Gas networks present much more storage potential than electrical network (\eg 50 times more in Germany and 300 times more in France) \cite{Rosa2017}.
+Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}.) Besides storing energy, in their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
+Moreover, some applications (\eg marine, aviation and heavy-duty transport) will be hard to electrify and keep on requiring high-density energy carriers \cite{horvath2018techno, brynolf2018}.
+These carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons in a renewable world.
+This is why this paper rather uses "defossilisation" rather than "decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.9\textwidth]{Storage_electrofuels.pdf}
+\caption{Energy carriers and technologies to store electricity. Electrofuels are an economical solution for high capacity and long-term storage of energy. Graph adapted from \cite{ISPT2017}.}
+\label{fig:intro:Storage_electrofuels}
+\end{figure}
+
+To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a multi-sector and whole-energy system \cite{Contino2020}.
+To reach this goal, an energy system optimisation model (ESOM) can define the design of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
+In this research field, \citet{yue2018review} highlighted that most of ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
+However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system, such as a country.
+Given the lifetime of the conversion technologies, such strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
+Besides the uncertainty on the model structure (not addressed in this work), this long-term and large-system optimisation motivates the need to account for \gls{UQ} and consider it as a major challenge of such models \cite{pfenninger2014energy}.
+This challenge, along with a large number (\ie more than a hundred) of uncertain parameters and limited information of their distribution, leads to the "curse of dimensionality" \cite{kuo2005lifting}.
+
 It aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
@@
 
-General introduction to the need of energy transition\\
-
-How electrofuels will be part of the solution.
-Here include part of the terminology paper \url{https://www.frontiersin.org/articles/10.3389/fenrg.2021.660073/full} to clearly set the context and make sure the readers understand what's behind electrofuels and emphasize at the end that in our case, we consider electrofuels as the ones from renewable sources (with a 0-gwp).\\
-
 From this need, develop a thread like I did in my FRIA application, we need whole-energy system model to give more insights to policymakers\\

### 2024-02-22 appendices/app_EnergyScope.tex
Atom_molecules: first part of intro
Review of Appendix on EnergyScope by Hervé

@@
-\section{EnergyScope Pathway: The right model to choose} 
+\section{EnergyScope Pathway: The right model} 
 \label{app:ESPathway_choice}
 
+``\textit{Only when single-model results are contextualized by the model's position in the larger ensemble, the reader would be able to have a complete and correct interpretation of the output}" \cite{dekker2023identifying}.
 Energy system models of varying complexity are valuable tools for guiding policymakers and projecting future trends.
-These models enable the exploration of different energy scenarios and the assessment of their consequences based on the underlying assumptions.
+These models enable the exploration of different energy scenarios and the assessment of their consequences.
 Specifically, techno-economic models play a crucial role in identifying technically feasible pathways for the energy transition while considering the associated economic costs.
@@
 These categories can be broken down into subcategories.
-This paper focuses on the system planning and scenario analysis models.
+This work focuses on the system planning and scenario analysis models.
 Inspired from \cite{palmintier2013incorporating}.
@@
 In accordance with the previous classification, models aimed at aiding decision-makers in the energy transition primarily fall under the categories of planning and scenario analysis, with a lower technical resolution than the other classes of model (see Figure \ref{fig:energy_models_classification}).
-Nonetheless, ensuring technical accuracy is of paramount importance to ensure the effective functionality of future energy systems.
+Nonetheless, ensuring technical accuracy is of paramount importance to ensure the effective performance of future energy systems.
 Hence, these models should meet the following requirements as a minimum:
 (i) assessment of intermittent renewable energy integration thanks to an \textbf{hourly resolution} spanning a one-year time horizon;
-(ii) accounting for the \textbf{whole-energy system} as including all energy (\ie heat, electricity, mobility) and non-energy flows in different sectors, accounting for their respective greenhouse gas emissions, as well as all resources, conversion processes, and storage technologies;
+(ii) accounting for the \textbf{whole-energy system} by including all energy (\ie heat, electricity and mobility) and non-energy flows in different sectors, accounting for their respective greenhouse gas emissions, as well as all resources, conversion processes, and storage technologies;
 (iii) exploration of all available options through the \textbf{optimisation of investments and operations};
 (iv) consideration of long-term investments throughout the \textbf{transition pathway} process (\ie 30 years up to 2050, in our case); and
-(v) ensuring a reasonable \textbf{computational time} (\ie less than one hour on a personal laptop 15 minutes) for analysing different trajectories.
+(v) ensuring a reasonable \textbf{computational time} (\ie less than one hour on a personal laptop) for analysing different trajectories.
 Additionally, to enhance result reproducibility and user understanding, it is advantageous for such models to:
@@
 
-These requirements are commonly found in reviews of energy system models.
-In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the number of users and market equilibrium.
+These requirements are commonly found in reviews on energy system models.
+In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the ``popularity'' of the models via the number of downloads/sales or the integration of economic market equilibrium.
 In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv).
@@
 In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
-Based on these reviews, Table \ref{tab:intro_soa} compares models based on all the previous criteria except the computational time (v).
+Based on these reviews, models are compared based on all the previous criteria except the computational time (v) (see Table \ref{tab:intro_soa}).
 Indeed, the latter is hard to compare as models are not applied to the same case study and the information is rarely given.
 The table includes only the models that achieved partially at least four out of the five criteria.
-We endeavored to refresh the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.
+We endeavored to update the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.
 
@@
 \caption{Comparison of existing models that partially satisfy at least four of the five criteria (in alphabetical order).
-Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
+Legend: \checkmark criterion satisfied; {\color{lightgray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
 Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition}}
@@
 Calliope & \cite{Atlason2014,Calliope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{Topic is being discussed in the chat of their repository but not yet included in their documentation.} & \checkmark \\
-COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
-DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted.}\footnote{\label{foot:industrynotaccounted} Industry not accounted} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{gray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
-DIETER & \cite{zerrahn2017long} & \checkmark & {\color{gray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
-E2M2 & \cite{E2M2} & \checkmark & {\color{gray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} individual heating not accounted.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
-EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark}\footref{foot:freeundersomespecial} \\
+COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{lightgray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
+DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{lightgray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted for.}\footnote{\label{foot:industrynotaccounted} Industry not accounted for.} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{lightgray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
+DIETER & \cite{zerrahn2017long} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted for.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
+E2M2 & \cite{E2M2} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} Individual heating not accounted for.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
+EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{lightgray} \checkmark}\footref{foot:freeundersomespecial} \\
 Ener.
 Trans.
-Model & \cite{ETM} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{The ETM is a simulation model with a simple merit order 'optimisation' for electricity, flex and heat.} & \checkmark & \checkmark \\
-EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{gray} \checkmark}\footref{foot:freeware} \\
-energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only.} & \checkmark & \checkmark \\
+Model & \cite{ETM} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{The ETM is a simulation model with a simple merit order 'optimisation' for electricity, flexibility and heat.} & \checkmark & \checkmark \\
+EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{lightgray} \checkmark}\footref{foot:freeware} \\
+energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only.} & \checkmark & \checkmark \\
 EnergyScope TD & \cite{limpens2019energyscope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:notpathwaystated} & \checkmark \\
@@
 In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.}
-&  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\ 
-iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{gray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
-IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
+&  {\color{lightgray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\ 
+iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{lightgray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:freeundersomespecial} \\
+IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{lightgray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:paidlicenced} \\
 OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:simulation} & \checkmark & \checkmark \\
-Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
-PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{lightgray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced}  \\
+PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
 Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
 } & \checkmark \\
-RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
-ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial} \\
+RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{lightgray} \checkmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark  & \checkmark & \checkmark \\
+ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{lightgray} \checkmark}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footref{foot:freeundersomespecial} \\
 TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
@@
 %\caption{Comparison of existing models that partially satisfy at least four of the five criteria (in alphabetical order).
-Legend: \checkmark criterion satisfied; {\color{gray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
+Legend: \checkmark criterion satisfied; {\color{lightgray} \checkmark} criterion partially satisfied; {\color{gray} \xmark} criterion not satisfied.
 Data from \cite{Connolly2010,prina2020classification,chang2021trends,prina2019transition} (part B)}
@@
 %Calliope & \cite{Atlason2014,Calliope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{Topic is being discussed in the chat of their repository but not yet included in their documentation.} & \checkmark \\
-%COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
-%DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{gray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted.}\footnote{\label{foot:industrynotaccounted} Industry not accounted} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{gray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
-%DIETER & \cite{zerrahn2017long} & \checkmark & {\color{gray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
-%E2M2 & \cite{E2M2} & \checkmark & {\color{gray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} individual heating not accounted.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
-%EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{gray} \checkmark}\footref{foot:freeundersomespecial} \\
+%COMPOSE & \cite{COMPOSE} & \checkmark & \checkmark & \checkmark & \checkmark & {\color{lightgray} \checkmark}\footnote{\label{foot:freeundersomespecial}`\emph{Free under some special conditions}´.} \\
+%DER-CAM & \cite{popovic2017mixed,dercam_website} & \checkmark & {\color{lightgray} \checkmark}\footnote{\label{foot:transportnotsaid} Transport not accounted.}\footnote{\label{foot:industrynotaccounted} Industry not accounted} & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:notpathway} Not specified but time horizon is 1 year.} & {\color{lightgray} \checkmark} \footnote{\label{foot:freeware} Freeware.} \\
+%DIETER & \cite{zerrahn2017long} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:industrynotaccounted}\footnote{\label{foot:dhnnotaccounted} \gls{DHN} not accounted.} & \checkmark & {\color{gray} \xmark}\footref{foot:notpathway} & \checkmark \\
+%E2M2 & \cite{E2M2} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footnote{\label{foot:decLTHnotaccounted} individual heating not accounted.} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced} Commercially (paid) licensed.} \\
+%EMPIRE & \cite{backe2022empire} & \checkmark & {\color{gray} \xmark} \footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & \checkmark & \checkmark & {\color{lightgray} \checkmark}\footref{foot:freeundersomespecial} \\
 %Ener.
@@
 Model & \cite{ETM} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{The ETM is a simulation model with a simple merit order 'optimisation' for electricity, flex and heat.} & \checkmark & \checkmark \\
-%EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{gray} \checkmark}\footref{foot:freeware} \\
-%energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only, while iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & \checkmark \\
+%EnergyPLAN & \cite{lundenergyplan} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation} Simulation model.} & {\color{gray} \xmark} \footnote{\label{foot:notpathwaystated} Yearly horizon without pathway.} & {\color{lightgray} \checkmark}\footref{foot:freeware} \\
+%energyRt & \cite{EnergyRT} & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footnote{\label{foot:invopt} EnergyRT optimises investments only, while iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & \checkmark \\
 %EnergyScope TD & \cite{limpens2019energyscope} & \checkmark & \checkmark & \checkmark & {\color{gray} \xmark} \footref{foot:notpathwaystated} & \checkmark \\
@@
 Their study applied the model to a transition pathway for Croatia.
-In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{gray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
-%iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & {\color{gray} \checkmark} \footnote{\label{foot:invopt_bis} iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & {\color{gray} \checkmark} \footnote{\label{foot:freeundersomespecial_bis}`\emph{Free under some special conditions}´.} \\
-%IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced_bis} Commercially (paid) licensed.} \\
+In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.} &  {\color{lightgray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\
+%iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & {\color{lightgray} \checkmark} \footnote{\label{foot:invopt_bis} iHOGA conducts optimisation and simulation without specifying timing or scope.} & \checkmark & {\color{lightgray} \checkmark} \footnote{\label{foot:freeundersomespecial_bis}`\emph{Free under some special conditions}´.} \\
+%IMAKUS & \cite{kuhn2012iteratives} & \checkmark & {\color{lightgray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \xmark}\footnote{\label{foot:paidlicenced_bis} Commercially (paid) licensed.} \\
 %OpenDSS & \cite{opendss} & \checkmark & \checkmark & {\color{gray} \xmark} \footnote{\label{foot:simulation_bis} Simulation model.} & \checkmark & \checkmark \\
-%Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{gray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced_bis}  \\
-%PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{gray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
+%Plexos & \cite{energyexemplarplexos9} & \checkmark & {\color{lightgray} \checkmark}\footnote{Does not account for all sectors but allow to implement them according to \citet{waucquez2023validation}.} & \checkmark & \checkmark & {\color{gray} \xmark}\footref{foot:paidlicenced_bis}  \\
+%PyPSA & \cite{brown2017pypsa,PyPsa} & \checkmark & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footnote{\citet{pedersen2022long} applied PyPSA to a whole energy system split in 37 nodes.
 Using a myopic approach, the model optimises the energy transition with a 3-hours resolution).
 } & \checkmark \\
-%RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{gray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark  & \checkmark & \checkmark \\
-%ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{gray} \checkmark}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark & \checkmark & {\color{gray} \checkmark} \footref{foot:freeundersomespecial_bis} \\
+%RamsesR & \cite{energistyrelsen2023ramses} & \checkmark & {\color{lightgray} \checkmark}\footref{foot:transportnotsaid_bis}\footref{foot:industrynotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark  & \checkmark & \checkmark \\
+%ReEDS & \cite{short2011regional} & {\color{gray} \xmark}\footnote{ Seasonal time slice.} & {\color{lightgray} \checkmark}\footref{foot:industrynotaccounted_bis}\footref{foot:dhnnotaccounted_bis}\footref{foot:decLTHnotaccounted_bis} & \checkmark & \checkmark & {\color{lightgray} \checkmark} \footref{foot:freeundersomespecial_bis} \\
 %TIMES & \cite{loulou2005documentation} & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\footnote{Model is now open-source with limited access to data \cite{openmod_times_description}.} \\
@@
 
-\myparagraph{TIMES}\\
-
-\noindent
-The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models.
-It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration.
-This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute in their research endeavors.
-Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}.
-A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}.
-Recent developments by the IEA-ETSAP have resulted in a version that is compatible with open-source solver CBC.
-In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}.
-It is noteworthy that \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using 12 typical days with an hourly resolution, highlighting the sensitivity of results to time resolution.
-Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data remains an area of ongoing inquiry \cite{openmod_times_description}.
-\\
-
 \myparagraph{Calliope}\\
@@
 Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic.
-In fact, the model is typically employed for scenario analysis.
+In fact, the model is typically employed for snapshot analysis, \ie optimization of a target future year.
 Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}.
@@
 This model is an application of the open-source energy modelling system (OSeMOSYS), itself represented as a model with a poor time discretisation and a heavy computational burden according to \cite{prina2019transition}.
-\citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors, the time disaggregation being chosen by the user.
+\citet{loffler2017designing} applied the model to the world by splitting it into 10 regions and most of the energy demand sectors, leaving to the user the choice of the time resolution.
 For their application they used representative years with three days and two time slice per day.
@@
 Among the open-source models with an active community, PyPSA is one of the best-performing, with a large and active community, development at the state of the art, worldwide applications, and usage not only limited to academia.
-A study conducted by \citet{bartholdsen2019pathways} centered on Germany employed a representation comprising 16 time slices for representative years.
+A study conducted by \citet{bartholdsen2019pathways} centered on Germany employed a representation comprising 16 time slices per representative year.
 This choice was substantiated by the work of \citet{welsch2014incorporating}, which demonstrated that this level of temporal granularity yields consistent results in comparison to hourly time resolution over a year.
-However, it is noteworthy that the utilization of a limited number of time slices may simplify the optimization of storage technologies, especially those designed for inter-month energy storage.
+However, it is noteworthy that the utilization of a limited number of time slices may oversimplify the optimization of storage technologies, especially those designed for inter-month energy storage.
 This simplification can be viewed as a pragmatic approach to reduce the computational burden while over-simplifying the challenge of accurately integrating intermittent renewable energy sources.
@@
 
-Hence, it is worth noting that while Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, we have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices.
-Hence, it appears that none of the models of Table \ref{tab:intro_soa} fully meets all five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
+\myparagraph{TIMES}\\
+
+\noindent
+The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models.
+It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration.
+This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute.
+Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}.
+A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}.
+Recent developments by the IEA-ETSAP have resulted in a version that is compatible with the open-source solver CBC.
+In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}.
+To highlight the sensitivity of results to time resolution, \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using up to 12 typical days with an hourly resolution.
+Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data is not ensured \cite{openmod_times_description}.
+\\
+
+While Calliope, OSeMOSYS, PyPSA and TIMES frameworks have the potential to be used for evaluating a transition pathway, we have not come across any publication that explicitly demonstrates their application to such cases with an hourly time resolution over significant time slices to accurately capture the seasonality within each representative year.
+Hence, it appears that none of the models of Table \ref{tab:intro_soa} fully meets the five criteria outlined in the table, topped with the additional consideration of acceptable computational time.
 This observation is consistent with the findings presented by \citet{prina2019transition} who identified two approaches for optimising the energy transition pathway based on the six criteria.
@@
 For the aforementioned reasons, the current work opted for EnergyScope Pathway, an extension of the open-source and documented EnergyScope TD model \cite{limpens2019energyscope} listed in Table \ref{tab:intro_soa}.
-The latter has an horizon time of one year and does not account for the pathway from an existing energy system to a long-term target.
+The latter has a time horizon of one year and does not account for the pathway from an existing energy system to a long-term target.
 The pathway version extends the time horizon to decades and accounts for the pathway transition from an existing energy system to a long term target.
-The computational time is kept low (\ie around a 15 minutes on a personal laptop), mostly due to keeping the linear formulation after extending the snapshot model.
+The computational time is kept low (\ie around 15 minutes on a personal laptop), mostly due to keeping the linear formulation after extending the snapshot model.
 \citet{limpens2024pathway} provides more detailed insights into the modeling choices made during methodological development.
@@
 Compared to existing models, EnergyScope Pathway introduces a rapid computational optimisation tool for exploring diverse transition pathways within an entire energy system while maintaining high temporal precision to accurately capture the integration of intermittent renewables.
-To the best of our knowledge, there are potentially frameworks that could be extended to similar capabilities, but their computational time for similar case study have not been found.
+To the best of our knowledge, there are potentially other frameworks that could be extended to similar capabilities, but their computational times for similar case studies have not been found.
 
@@
 The objective of this section is to present the fundamental variables and constraints of the latter based on which the former was developed.
-There have been formulation choices to be made but they are not discussed in the current manuscript.
-However, the interested reader is invited to refer to the Appendix B of \cite{limpens2024pathway} for further information in this regard.
+Formulation choices have been made but they are not discussed here.
+The interested reader is invited to refer to Appendix B of \cite{limpens2024pathway} for further information in this regard.
 
@@
 
-To break this curse, EnergyScope TD, like other models \cite{Gabrielli2018,Despres2017,Nahmmacher2014}, relies on a subset representative days called typical days (TDs).
-This more limited number of days, \ie 12 in the rest of this thesis, cluster the days of the year that have similar time series of demands (\ie varying electricity and heat demands) and weather data (\ie sun, onshore and offshore wind).
+To break this curse, EnergyScope TD, like other models \cite{Gabrielli2018,Despres2017,Nahmmacher2014}, relies on a subset of representative days called typical days (TDs).
+This more limited number of days, \ie 12 in the rest of this thesis, clusters the days of the year that have similar time series of demands (\ie varying electricity and heat demands) and weather data (\ie sun, onshore and offshore wind).
 This way, each day of the year is associated to one of these typical days (see Figure \ref{fig:Typical_days}).
@@
 \includegraphics[width=0.9\textwidth]{Typical_days.pdf}
-\caption{Association of each day of the year (grays) to one of the 12 typical days (TDs) (black dots). Graph adapted from \citet{limpens2019energyscope}.}
+\caption{Association of each day of the year (gray dots) to one of the 12 typical days (TDs) (black dots). Graph adapted from \citet{limpens2019energyscope}.}
 \label{fig:Typical_days}
@@
 Among others, this allows representing the dynamics and the seasonality of storage capacities.
-This method as well as the clustering approach selected in our case, \ie k-medoids \cite{Kaufman2009,Park2009}, extensively detailed and compared to other methods, in the work of \citet{limpens2019energyscope}.
+This method as well as the clustering approach selected in our case, \ie k-medoids \cite{Kaufman2009,Park2009}, are extensively detailed and compared to other methods, in the work of \citet{limpens2019energyscope}.
 \\
@@
 The model's hourly resolution over a year makes it well-suited for integrating intermittent renewables.
-Its formulation incorporates typical days and a reconstruction method that captures different time scales from the hour to the season while accounting for the inter-weeks patterns of wind.
-The model investigates all the possibilities by optimising the investment decisions and hourly operations over a year, with a computational time of less than a minute on a personal laptop.
+Its formulation incorporates a reconstruction method that captures different time scales from the hour to the season while accounting for the inter-weeks patterns of wind.
+The model optimises the investment decisions and hourly operations over a year, with a computational time of less than a minute on a personal laptop.
 This characteristic was intentionally incorporated into the model design to facilitate uncertainty quantification and other studies that require numerous iterations \cite{rixhon2021role}.
@@
 EnergyScope TD has been successfully applied to various national energy systems, including Switzerland \cite{Limpens_role_2019,limpens2019energyscope}, Belgium \cite{Limpens2020}, Italy \cite{borasio2022deep}, and other European countries \cite{dommisse2020modelling}.
-Furthermore, it has been extended to a multi-region energy system model \cite{thiran2023validation}, coupled with other energy models \cite{pavivcevic2022bidirectionnal}, or employed to focus on specific sectors such as the power networks of electricity, gas, and hydrogen \cite{schnidrig2023role}.\\
+Furthermore, it has been extended to a multi-region energy system model \cite{thiran2023validation}, coupled with other energy models \cite{pavivcevic2022bidirectionnal}, or employed to focus on specific sectors such as the networks of electricity, gas, and hydrogen \cite{schnidrig2023role}.\\
 
@@
 The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost and greenhouse gas emissions of the energy system.
-Typically, the two objectives are integrated by placing a limit on emissions while simultaneously striving to minimize costs.
+Typically, the two objectives are integrated by placing a limit on emissions while simultaneously minimizing the costs.
 
@@
 \includegraphics[width=\textwidth]{chp_estd_overview.pdf}
-\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimizing model (Optimisation) and results (Energy strategy). The image illustrates what is included (non-exhaustively).}
+\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimizing model (Optimisation) and results (Energy strategy). }
 \label{fig:estd_overview}
@@
 \noindent
-The following section illustrates the main constraints of the original EnergyScope TD model.
-The objective function, cost and \gls{GHG} formulation will be detailed.
-The rest of the formulation is detailed and available in previous works \cite{limpens2021generating}.
-This work uses the following nomenclature: SETs are in capital letters, \textbf{Variables} are in bold and with first letter capital, and \emph{parameters} are in italic.
+The following section illustrates the formulation of the original EnergyScope TD model.
+The objective function, cost and \gls{GHG} formulation are detailed.
+The rest of the formulation is detailed and available in a previous work \cite{limpens2021generating}.
+This work uses the following nomenclature: SETs are in capital letters, \textbf{Variables} are in bold and with the first letter in upper case, and \emph{parameters} are in italic.
 
@@
 
-The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over different periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
-To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}) and following equations.
+The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over the different time-periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
+To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}).
 The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}.
@@
 Those representing the energy balance, network implementation, sectors representation, etc.
-are not presented in this work but are detailed in the latest version of the model, see \cite{limpens2021generating} and on the documentation \cite{readthedocs_estd_v2.2}.
+are not presented in this work but are detailed in the latest version of the model, see \cite{limpens2021generating} and in the documentation \cite{readthedocs_estd_v2.2}.
 
-Finally, energy storage has two dimensions to be optimised: the stored energy quantity (also referred to as 'storage level') and the hourly power flow, encompassing both charging and discharging.
-EnergyScope TD optimises the hourly charge and discharge operations based on the hourly resolution of the typical days.
-In contrast, the optimisation of stored energy is conducted over the entire span of 8760 hours in a year.
+Finally, energy storage has two dimensions to be optimised: (i) the hourly power flow, encompassing both charging and discharging and, (ii) the stored energy quantity (also referred to as 'storage level').
+EnergyScope TD optimises the former based on the hourly resolution of the typical days and the latter over the entire span of 8760 hours in a year.
 This formulation allows for the effective integration of a wide range of energy storage technologies, spanning short-term solutions like small thermal storage units and daily-use batteries, to longer-term options such as hydro-dam storage for seasonal storage, and even large-scale thermal storage for intra-week patterns.
-A previous study delved into the roles of various storage technologies, considering their sectoral applications and temporal aspects, within the context of the Swiss energy system \cite{Limpens_role_2019}.
+A previous study investigated the roles of various storage technologies, considering their sectoral applications and temporal aspects, within the context of the Swiss energy system \cite{Limpens_role_2019}.
 
@@
 Simplification and choices were necessary to implement linearly the problem while keeping a tractable computational time.
-In this section, we present the formulation retained.
+In this section, the retained formulation is presented.
 
@@
 This new dimension is necessary to represent the changes of technology and resource characteristics over the representative years.
-As an example, the investment cost (\emph{c}\textsubscript{\emph{inv}}) of a solar photovoltaic panels could drastically vary in the next decades (e.g.
+As an example, the investment cost (\emph{c}\textsubscript{\emph{inv}}) of solar photovoltaic panels could drastically vary in the next decades (e.g.
 data used ranges between 1220 to 870  [€\textsubscript{2015}/kW] between 2020 and 2035).
@@
 \includegraphics[width=\textwidth]{path_e.g._tech.png}
-\caption{Example of how the technologies capacity and associated variables are evolving.
+\caption{Example of how the technology capacities and associated variables are evolving.
 The example uses a technology with a 20 years lifetime.
@@
 &\textbf{F\textsubscript{old}} (p,i) =  
-\hspace{3mm}\text{if} ( \emph{age} = `\emph{STILL\_IN\_USE}') \text{ then  } 0   &   &\notag\nonumber\\
+\hspace{3mm}\text{if} ( \emph{age} = \text{`\emph{STILL\_IN\_USE}'}) \text{ then} 0   &   &\notag\nonumber\\
 &\hspace{2cm}\text{else } \left(\textbf{F\textsubscript{new}}(age,i)  - \sum_{\vspace{5mm}\mathclap{p2 \in \text{\emph{PHASE}}}} \textbf{F\textsubscript{decom}} (p2,age,i)\right)    & \notag\nonumber\\
@@
 For each phase, a set (\emph{AGE}) is calculated \emph{a priori}.
-It relates, for a given phase and technology, when the technology should have been built.
+It relates, for a given phase and technology, when the technology was built.
 In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built.
@@
 As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
-Ones will highlight the use of a `if' in Eq. (\ref{eq:Fold_def_app}), this formulation is linear as the if is applied to a parameter and not a variable.
+The `if' in Eq. (\ref{eq:Fold_def_app}) is linear as it is applied to a parameter and not a variable.
 
@@
 
-The objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
+As an extension of Eq. \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
 %The objective is the minimisation of the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}),
 defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2_app}).
-The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2_app}), to which the residual asset investment cost in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
+The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2_app}), to which the residual asset value in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
 Thus, the investments account for the installation and dismantlement costs of the technologies.
@@
 Eq.
-(\ref{eq:salvage_app}) computes, for each technology, the residual value that must be deducted from the total cost.
+(\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost.
 The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised.

### 2024-02-22 chap_atom_mol/chap_atom_mol.tex
Review of Appendix on EnergyScope by Hervé

--- 2024-02-16 chap_atom_mol/chap_atom_mol.tex
@@
 First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
+Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
 
-\textbf{Disclaimer}: Relying on ``local'' nuclear energy for some (or on importing electrofuels from abroad for others) is highly controversial.
-On top of purely techno-economic aspects, non-exhaustively mentioned beforehand, there are other ethical, societal or even political considerations to account for when addressing this question.
-Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} in the Belgian energy landscape, on a strictly techno-economical point of view.
+\textbf{Disclaimer}: Relying on ``local'' nuclear energy for some is highly controversial.
+On top of purely techno-economic aspects, non-exhaustively mentioned beforehand, there are other ethical, societal or even political considerations to account for when addressing this question \cite{kempf2022}.
+In the same sense, on top of ethical or geopolitical aspects, one could question the availability of the imported electrofuels, assumed to be unlimited in this work (see Chapter \ref{chap:case_study}).
+In their work, \citet{lefebvre2022electrofuel} investigated this topic for the case of Belgium considering lower and upper bounds in terms of availability based on, on the one hand, the already signed agreements with the exporting countries and, on the other hand, their maximal technological potential of \gls{VRES}, respectively.
+They highlighted that the lower bound, \ie currently the most reliable information, resulted in a total availability of electrofuels, at least, one order of magnitude lower than the needs provided by the cost-optimum carbon-neutral Belgium in 2050.
+Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} as well as the need of importing electrofuels in the Belgian energy landscape, on a strictly techno-economical point of view with a cost-based optimisation.
 This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in the Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
@@
 \label{sec:atom_mol:contributions}
-\begin{itemize}
-\item Pathway UQ
-\item Influence of the parameters on the import of molecules and the installation of SMR
+In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimize a transition pathway with a small number of time stages, \ie two or three.
+Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with tens of uncertain parameters (see Appendix \ref{app:UC_full}).
 
-\item Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad.
+Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad, in a country where the local potential of \gls{VRES} is limited compared to its \gls{EUD}.
 
-\end{itemize}
+On top of identifying the most impactful uncertain parameters on the objective function, \ie the total transition cost, this chapter aims also at highlighting the ones that mostly drive other variables resulting from the optimization process: the installation of \gls{SMR} and the import of each of the electrofuels by the end of the transition, \ie 2050.
 
@@
 To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case.
-Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.\\
-
-{\color{red} \textbf{Compare here with EnergyVille - PATHS2050.}}\\
+Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.
+In their work, \citet{PATHS2050} also showed that \gls{SMR} first substitutes ``e-fuels/hydrogen'' turbines before reducing the need for \gls{PV}.
+However, in their ``Central'' scenario where no \gls{SMR} is installed by 2050, they rely on 96.6\,GW of \gls{PV}, 63\% more than the 59.2\,GW potential considered in our work and about 15 times more than the current installed capacity, \ie 6.5\,GW.
+\\
 
@@
 \end{figure}
+
+\section{Discussion}
+\label{sec:atom_mol:discuss}
+Given its lower \gls{LCOE}, \gls{SMR} is installed as soon as available.
+It directly substitutes other flexible power generation units (\ie CCGT) and provides, by 2050, 44.6\,TWh, 25\% of the total electricity production.
+Consequently, this reduces the need to produce electricity via \gls{CHP} and allows increasing by 48\% the high-temperature heat produced by electric heaters.
+Besides these two sectors, the others (\ie low-temperature heat, mobility and non-energy demand) are marginally or not impacted by the integration of \gls{SMR}.
+
+Betting on \gls{SMR} means letting the emissions go up at early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are respectively 3.3\% and 8.8\% cheaper than in the REF case.
+
+Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), \acrfull{GSA} highlights the biggest impact of the cost of purchasing electrofuels on the variation of the total transition cost, around 45\%.
+On the contrary, parameters directly related to \gls{SMR} (\ie its availability and its CAPEX) have limited impact, below 1\%.
+This \gls{GSA} also points out the key drivers for the import of renewable electrofuels and the installation of \gls{SMR} by 2050.
+Besides the cost of purchasing electrofuels (\ie the lower this cost, the bigger the imports), it shows that available \gls{SMR} would have mainly a direct impact on e-ammonia, by substituting the ammonia \gls{CCGT} that is the biggest consumer of this molecule.
+Indirectly, this parameter will also reduce the import of e-methane by reducing the need for gas \gls{CHP} and gas boilers.
+About e-hydrogen and e-methanol, their imports are impacted by the other technologies in competition in the transport sector and the industrial demand, respectively.
+
+In conclusion, this work puts under the spotlight the ``competition'' between \gls{SMR} and imported electrofuels while both of them support to the integration of local \gls{VRES}.

### 2024-02-22 chap_case_study/chap_case_study.tex
Review of Appendix on EnergyScope by Hervé

@@
 CAPEX & 4850 & €/kW \\
-Annual OPEX & 103 & €/kW/year \\
+Annual OPEX & 103\footnote{\label{foot:c_maint_SMR}This value is in line with \citet{smr_canada} that evaluate annual opex of 0.015\$/kWh, which makes 112\$/kW/year assuming a 85\% full-load availability, and \citet{PATHS2050} that accounts for 83.3€/kW/year} & €/kW/year \\
 Lifetime & 60 & year \\
@@
 For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of \og unicorn technologies\fg or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.
+In their ``near-term to net zero'' (NT2NZ) approach to estimate \ce{CO2} prices, \citet{kaufman2020near} emphasises the need to select, \textit{a priori}, an emissions pathway to the net-zero target.
+Some authors suggest limiting near-term technological disruptions and, consequently, the initial rate of emission reductions \cite{wigley1996economic} where others, to avoid technology lock-in and to benefit from early investments in the transition, encourage a sharper decrease of the emissions at an earlier stage \cite{vogt2018starting}.
 

### 2024-03-06 appendices/app_EnergyScope.tex
First parts of intro and RL chapters

@@
 In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the ``popularity'' of the models via the number of downloads/sales or the integration of economic market equilibrium.
+Eight years later, \citet{lopion2018review} enriched the review of \citet{Connolly2010} using similar critieria and including models developed in the 2010s.
+In 2014, \citet{pfenninger2014energy} pointed out the current paradigms and challenges to face as well as the emerging approaches to address them in the 21st century energy systems modeling community.
+Besides the behavioral and social factors, they also highlighted the challenges related to mutli-sectorial systems, time and space resolution or the open-accessibility of data and models and their ability to account for uncertainties.
 In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv).

### 2024-03-06 appendices/app_case_study.tex
First parts of intro and RL chapters

@@
 This is a combination of efficiency and substitution of fossil fuels with electricity.
+In particular, the private mobility drops to near-zero emissions at early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift.
+First, while accounting for the society inertia in this sector (see Eq.
+\ref{eq:limit_reno_passmob_app}), battery electric vehicles (BEV) substitute \gls{ICE} cars.
+This reduces the overall emissions of the private vehicles as electric motors are more efficient than \gls{ICE} and they are driven by a less-emitting resource (\ie electricity versus diesel and gasoline).
+On top of this, the model allows a modal shift from the private to the public mobility.
+From the current share of the passenger mobility provided by private vehicles (\ie 80\%), the model allows this share to go down to 50\%.
 Efficiency comes mainly from district heating networks and electrical heat pumps for the heat sector, and public mobility and electric cars for the mobility sector.

### 2024-03-06 chap_RL/chap_RL.tex
Review of Appendix on EnergyScope by Hervé
First parts of intro and RL chapters

@@
-%!TEX root = ../thesis_main.tex
-%!TEX encoding = UTF-8 Unicode
+\vspace{-0.2cm}
+\begin{flushright}
+\emph{``For the things we have to learn before we can do them, we learn by doing them.''}\\
+Aristotle, in \textit{The Nicomachean Ethics}, IV$^\text{th}$ century BC
+\end{flushright}
+\vspace{0.4cm}
+
+Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}.
+For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track with the 1.5°C (if not, 2°C) increase of global temperature by the end of the century.
+On the contrary, using their top-down model DICE, \citet{nordhaus2014question} state that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.
+This is even more valid when models assess a myopic transition pathway subject, with limited foresight through the future with progressively unveiled uncertainties.
+
+To address this issue, several approaches have been used.
+Among them,  multi-stage stochastic programming is often put forward as a promising method.
+Stochastic programming formulates the problem as a mathematical program with probabilistic constraints or objective functions.
+These models explicitly consider the uncertainty by incorporating probability distributions for the uncertain parameters.
+The goal is to find an optimal decision that minimizes/maximizes the expected value of the objective function while satisfying the probabilistic constraints, modelled as a scenario tree.
+At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}.
+Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the 5-year steps transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
+The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
+However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10,and relies on probability distribution that are often difficult to define properly.
+Increasing the number of these uncertainties in stochastic programming usually leads to a computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
+Based on the approach of \citet{bertsimas2004price}, and similarly to \citet{Moret2017PhDThesis}, \citet{nicolas2021robust} rather opted for the robust optimisation of the global pathway up to 2200 given different temperature deviation targets, \ie 2 or 3°C by 2200 via the use of uncertainty budget, $\Gamma$, in the TIAM-World model \cite{loulou2005documentation}.
+Considering 9 climate parameters and their respective lower and upper bounds, the idea behind the uncertainty budget stems from the improbability of all parameters simultaneously reaching one of their two extreme values.
+
+In the exploration of the myopic transition pathway under uncertainties, we decided to investigate the \gls{RL} approach to benefit from its policy optimisation mechanism.
+Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
+\gls{RL} exhibits two main advantages: its effectiveness to handle uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimize the policy \cite{perera2021applications}.Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.
+Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2}-budget, are very limited.
+Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.
+
+\section*{Contributions}
+\label{sec:meth:contributions}
+Applying the \gls{RL} approach to the optimization of the myopic transition pathway of a whole-energy system presents several novelties.
+First of all, as introduced in Section \ref{subsec:meth_RL_fundamentals}, when applied to energy systems, \gls{RL} is more dedicated either to smaller scale systems (\eg \gls{BEMS}, vehicles and energy devices) or to sector-specific, often the power sector, problems (\eg dispatch problems, energy markets and grid) \cite{perera2021applications}.
+In our case, the sector-coupling, the long-term goal at the end of a multiple-steps transition and the number of uncertain parameters make this application new for \gls{RL}.
+
+Then, applying to this optimisation environment, \ie EnergyScope myopic Pathway,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework.
+In this agent, while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.
+
+Finally, comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight, if not myopic, optimisation without learning process, highlights the added-value brought by the optimised policy.
 
 \section{Definition of the actions, states and rewards}
-\label{sec:act_states_rew}
-After pointing out that the environment is basically the EnergyScope myopic model at the different steps of the transition, define the actions, the states as well as the shape of the reward.
+\label{sec:RL:act_states_rew}
+As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}).
+In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment.
+Then, the window 2020-2030 is optimised via EnergyScope.
+Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state he took it.
+Based on these two pieces of information, \ie the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised.
+This goes on until eventually reaching 2050.
+The main purpose of this section is to define the shape of the reward as well as the sets of actions and states.\\
 
-\subsection{Actions}
-\subsection{States}
-\subsection{Reward}
+\myparagraph{Reward}\\
+
+Properly defined the reward fed by the environment to the agent is crucial in \gls{RL} for several reasons.If the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
+Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
+On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.
+It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.
+
+\myparagraph{Actions}\\
+
+Defining the levers of action, the core of the policy, to support the transition of a country-size whole-energy system is challenging, especially when accounting for political and socio-technical aspects \cite{castrejon2020making}.
+In our work, focusing only on the techno-economic aspect, we assume that the actions taken by the agent are directly implemented and impacting the environment, without ``misfire''.
+In other words, considering only the techno-economic lens, there is no moderation nor contest towards the agent's actions, as the objective is to assess how far and when within the transition to push the different levers of action.
+Given the overall objective of the agent to succeed the transition, \ie respecting the \ce{CO2}-budget by 2050, we have defined the actions in this sense.
+The first action, $\mathrm{act}_{\mathrm{gwp}} \in [0,1]$, aims at limiting the emissions at the representative year ending the concerned time window, $\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})$, between the level of emissions in 2020, \ie $\textbf{GWP\textsubscript{tot}}(2020)=123\,\text{Mt}_{\ce{CO2},\text{eq}}$, and carbon-neutrality:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:RL:act_gwp}
+&\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})\leq \mathrm{act}_{\mathrm{gwp}} \cdot \textbf{GWP\textsubscript{tot}}(2020) &
+\end{flalign}
+\endgroup
+
+Out the total \gls{GHG} emissions in Belgium in 2020,oil (\ie so-called \gls{LFO} in the model), on the one hand and, on the other hand, fossil gas, account for roughly 40\% and 31\%, respectively.
+Then, even though its use in 2020 is much more limited compared to the two formers, \ie 28\,TWh of solid fossil fuels (\ie so-called COAL in the model) versus 159 and 142\,TWh for oil and fossil gas, respectively, coal is a cheap, about 0.017€/GWh, and highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
+For these reasons, three additional actions to support the strict limitation of overall emissions of the first action: limiting the consumption of these three fossil resources up to the level of consumption in 2020, $\textbf{Cons\textsubscript{fossil gas}}(2020)$, $\textbf{Cons\textsubscript{LFO}}(2020)$ and $\textbf{Cons\textsubscript{coal}}(2020)$,  over the entire concerned time window, except the first one as this year is the initial condition of the time window and cannot be optimised any more:
+
+\begingroup
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+\label{eq:RL:act_NG}
+&\textbf{Cons\textsubscript{fossil gas}}(y)\leq \mathrm{act}\textsubscript{fossil gas} \cdot \textbf{Cons\textsubscript{fossil gas}}(2020) & \forall y \in \text{time window}\\
+\label{eq:RL:act_LFO}
+&\textbf{Cons\textsubscript{LFO}}(y)\leq \mathrm{act}\textsubscript{LFO} \cdot \textbf{Cons\textsubscript{LFO}}(2020) & \forall y \in \text{time window}\\
+\label{eq:RL:act_COAL}
+&\textbf{Cons\textsubscript{coal}}(y)\leq \mathrm{act}\textsubscript{coal} \cdot \textbf{Cons\textsubscript{coal}}(2020) & \forall y \in \text{time window}
+\end{flalign}
+\endgroup
+
+\noindent
+where $\mathrm{act}\textsubscript{fossil gas}$, $\mathrm{act}\textsubscript{LFO}$ and $\mathrm{act}\textsubscript{coal}$ can take values between 0 and 1.
+These complete the action space of the agent, $A\in \mathbb{R}^4_{[0,1]}$.\\
+
+\myparagraph{States}\\
 
@@
 %\end{itemize}
+
+\textbf{Confirm here with what is said in \cite{luderer2018residual}: 
+Importantly, our results also show that near-term policy stringency is an important driver of cumulative Res-FFI-CO2 in climate change mitigation scenarios.
+If strengthening of NDCs fails, Res-FFI-CO2 will be even higher, not only because of additional near-term emissions, but also due to a decrease of economic mitigation potentials in the longer term caused by further carbon lock-in.
+Delaying the strengthening of mitigation action will increase the world’s dependence on CDR for holding warming to well below 2°C, and is likely to push the 1.5°C target out of reach for this century.}

### 2024-03-06 chap_RobPol/chap_RobPol.tex
First parts of intro and RL chapters

@@
 
-The purpose of this chapter is to display results regarding PCA as presented orally to you in December 2023
+Assessing the robustness of a policy driving the transition pathway a whole-energy system is complex, especially due to the curse of dimensionality.
+This curse comes from the number of variables of the system (\eg the installed capacity of technologies), the multiple-year approach specific to the pathway optimisation (\ie versus the snapshot approach) or the number of uncertain parameters.
+On top of this, the sector coupling interconnecting the installed capacities and the used resources among the different (non-)energy sectors
+
+Present here the different definition of robustness and, among them, : 
+Policymakers can choose a
+strategy among various options: (1) plan for the worst-case scenario
+(i.e. resistance), (2) regardless of the conditions in the future, select a
+policy mix that allow the system to recover rapidly (i.e. resilience), (3)
+seek a strategy that will be able to perform reasonably well in almost all
+plausible futures (i.e. static robustness), and (4) in case of a change in
+conditions, be prepared to change the policy mix (i.e. adaptive ro-bustness) \cite{walker2012deep}.
+
+According to \citet{castrejon2020making}, ``a policy mix is considered robust, if the system of interest performs satisfactorily under a broad range of plausible futures.''\cite{walker2012deep}.
+In their work, they mostly focus on the electricity sector, variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economy models (like EnergyScope) that usually assume rational choice within an overall cost minimization.
+
+``Robust policy mixes aim to answer, ‘knowing the fact that the future is uncertain, what available policy instruments are likely to perform well in multiple plausible futures?’ [12].
+'' \cite{castrejon2020making}.
+
+\citet{herman2014beyond} defined robustness as ``… the fraction of sampled states of the world in which a solution satisfies all performance requirements''
+
+\section*{Contributions}
+\label{sec:RobPol:contributions}
+The main contributions of this chapter is the application of the methodology proposed in Section \ref{sec:meth:PCA} to the case study of the Belgian energy transition.
+First, we develop the different steps that lead to the principal components of the transition.
+We analyse these big trends of variation and highlight the fact that these variations stand for the entire pathway, a group of consecutive representative years or rather on a tipping-year.
+Then, and most importantly, we assess the robustness of different technological roadmaps by projecting their resulting myopic pathway against these directions of variation.
+The application of \gls{PCA} to provide a new metric for robustness applied to the case of Belgium is the added-value of this chapter.
+
+\section{Definition of the principal components of the transition}
+\label{sec:RobPol:PC_transition}
+As detailed in Section \ref{subsec:meth:PCA:transition}, we have decided to define the directions of variation, \ie the robustness metrics, based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol.
+These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.
+To extract these

### 2024-03-06 chap_atom_mol/chap_atom_mol.tex
First parts of intro and RL chapters

@@
 This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
+In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the models finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
 Consequently, there is smaller investment in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers.
@@
 
-Betting on \gls{SMR} means letting the emissions go up at early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are respectively 3.3\% and 8.8\% cheaper than in the REF case.
-
 Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), \acrfull{GSA} highlights the biggest impact of the cost of purchasing electrofuels on the variation of the total transition cost, around 45\%.
@@
 In conclusion, this work puts under the spotlight the ``competition'' between \gls{SMR} and imported electrofuels while both of them support to the integration of local \gls{VRES}.
+Betting on \gls{SMR} means letting the emissions go up at early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are 3.3\% and 8.8\% cheaper than in the REF case, respectively.
+However, given the need to import molecules at earlier stages of the transition, with or without \gls{SMR}, it seems reasonable to keep on investing in the transport and the infrastructure to support these imports.
+This would allow covering ourselves against the risk of eventually not having \gls{SMR} available by the end of the first half of this century.
+In the EnergyScope Pathway model, assuming the same technology mix as in the SMR case up to 2040 but without installing \gls{SMR} for the rest of the transition, \ie between 2040 and 2050, would require drastic (and quick) changes to still meet the \ce{CO2}-budget: around 44\,b€ of additional cumulative OPEX in purchasing electrofuels, mostly e-ammonia (+32\,TWh by 2050) and e-methane (+31\,TWh) to supply \gls{CCGT} and industrial \gls{CHP}.
+As a more concrete example, Fluxys, the manager of the gas network in Belgium, already presents significant investments by 2032 (\ie 1.3\,b€) to support this transition in the near future \cite{Fluxys_2023}.
+
+Chapters \ref{chap:chap_RL} and \ref{chap:chap_RobPol} put in practice the methodology detailed in Chapter \ref{chap:chap_methodo} to come up with a policy to get a system robust to this possibility of finally not having access to \gls{SMR}, as well as other uncertainties, especially when considering a myopic progression through the transition.

### 2024-03-06 chap_intro_ccl/intro.tex
First parts of intro and RL chapters

@@
-It has been proven that the climate change (among other environmental challenges) is (mostly) due to the concentration of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
-This concentration resulting from the cumulative emissions (and subtractions), usually expressed in kt$_{\ce{CO2},\text{eq}}$, over time could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
+It has been proven that the climate change, among other environmental challenges, is (mostly) due to the concentration of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
+Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
+Among these emissions, globally, about 75\% are directly related to the whole-energy system \cite{ourworldindata_CO2_world}.
+The \gls{GHG} emissions usually expressed in kt$_{\ce{CO2},\text{eq}}$, could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
 
@@
 However, Eq.
-\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition.
-Besides the question of the total population and its growth \cite{dodson2020population,scovronick2017impact}, these three levers of actions are: renewables, efficiency and sufficiency aiming at reducing the first, the second and the third terms on the right-hand side of Eq.
+\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition, leaving aside the question of the total population and its growth \cite{dodson2020population,scovronick2017impact}.
+These three levers of actions are: renewables, efficiency and sufficiency aiming at reducing the first, the second and the third terms on the right-hand side of Eq.
 \ref{eq:equality_GHG}, respectively.
@@
 This aligns with the current European policies binding the Member States of the European Union.
-For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind, an \textbf{increase in energy efficiency} and significantly \textbf{higher shares of energy from renewable sources} in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
+For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind\footnote{This directly relates to sufficiency as it encompasses social justice in parallel with a minimization of the energy use.}, an \textbf{increase in energy efficiency} and significantly \textbf{higher shares of energy from renewable sources} in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
 
 Consequently, to ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
-Besides behavioral changes, an overall reshape of the energy system is necessary in terms of both primary energy sources, \ie more renewables, and technologies used to convert these resources into the \gls{EUD} (\ie the energy service required by the the final consumer), \ie more efficiency \cite{iea2020world}.
+Besides behavioral changes, an overall reshape of the energy system is necessary in terms of both primary energy sources, \ie more renewables, and technologies used to convert these resources into the \gls{EUD} (\ie the energy service required by the the final consumer), \ie more efficiency \cite{iea2020world,luderer2018residual}.
 The former corresponds to a whole ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}) where energy carriers called, in the literature, ``\emph{biofuels}'', ``\emph{electrofuels}'', ``\emph{synthetic fuels}'', ``\emph{renewable fuels}'' or even ``\emph{sustainable fuels}'', will more and more play a crucial role.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{IEA_WEO_2019.png}
+\includegraphics[width=\textwidth]{IEA_WEO_2019.png}
 \caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario \cite{iea2020world}.}
@@
 It aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
+Instead of aiming at answering the question ``What could possibly happen in the future?'', let's rather address ``What could or should we do to make the future possible?'' ``Our task is not to foresee the future, but to enable it'' by Saint-Exupéry in Citadelle, 1948.
 

### 2024-03-06 chap_methodo/chap_methodo.tex
First parts of intro and RL chapters

@@
-%\begin{flushright}
-%%%\emph{``For the things we have to learn before we can do them, we learn by doing them.''}\\%Each wrong scenario you have analysed, is a wrong decision you won't made.''}\\
-%%%Aristote
-%\emph{All models are wrong, but some are useful.}\\
-%George P. Box
-%\end{flushright}
-
+\vspace{-0.2cm}
+\begin{flushright}
+\emph{``Technique aussi brûlante que les derniers bilans du GIEC.''}\\
+ Primero (ft. Romeo), in \textit{Deux deux}, 2022
+\end{flushright}
+\vspace{0.4cm}
 %\medskip
@@
 To navigate through the transition and investigate the efficiency of different policies, this work implements the reinforcement learning approach.
-This section aims first at presenting the general concepts of this approach as well as the policy optimisation algorithm.
-Then, the environment, actions, state and reward are detailed.
+This section aims at presenting the general concepts of this approach.
+Then, its application to the myopic optimisation of a whole-energy system is introduced as well as the policy optimisation algorithm.
 
@@
 $s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions.
-$a\in A$ is the action among the action space, $A$, and $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$.$r$ is the reward received by the action when taking the action $a$ from a state $s$, $R(s,a)$.
+$a\in A$ is the action among the action space, $A$, and $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$.$r$ is the reward received by the agent when taking the action $a$ from a state $s$, $R(s,a)$.
 Finally, $\pi$ is the policy as already described and $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{Schematics_RL.pdf}
+\includegraphics[width=0.8\textwidth]{Schematics_RL.pdf}
 \caption{\Acrfull{RL} framework made of the agent interacting with its environment, \ie the energy-system model on a limited decision window of 10 years.}
@@
 For further information on \gls{RL} and the \gls{SAC} algorithm, the interested reader is invited to refer to the works of \citet{sutton2018reinforcement} and \citet{haarnoja2018soft}, respectively.
+
+\textbf{Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity}
 
@@
 These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimized through the perfect foresight approach.
-Over the transition, sectors like electricity or methanol become more important due to sector coupling, \eg e-mobility or \gls{MTO}.}
+Over the transition, sectors like electricity (\ie from 22\% in 2025 to 33\% in 2050) or ammonia (\ie from 2\% to 8\%) become more important due to sector coupling, \eg e-mobility or ammonia-\gls{CCGT}.}
 \label{fig:Share_EUD_sectors}
@@
 In conclusion, a roadmap would be defined as more robust than another one if the projection of its myopic runs on the different $\text{PC}_{\text{transition}}$ span on a more narrow range.
+The bounds of this ``range of projection'' are computed as the mean, $\mu$, of the projected data $\pm$ a 95\% confidence level, CL, on the margin of error, MOE:
+$$
+\text{range of projection} = [\mu-\mathrm{MOE}; \mu + \mathrm{MOE}] 
+$$
+
+\noindent
+where the margin of error, MOE,  is computed thanks to standard error of the mean, SEM, and assuming a Student's, distribution, t,  of the $N$ projected data:
+$$
+\text{MOE} = \mathrm{SEM}\cdot \mathrm{PPF}_{\mathrm{t}}\left((1+\mathrm{CL})/2,N-1\right)
+$$
+
+\noindent
+where $\mathrm{PPF}_{\mathrm{t}}$ is the percent point function (\ie inverse of cumulative distribution function) of the Student's distribution.

### 2024-03-06 options/symbols_and_cmd.tex
Merge taxonomy paper + Implement Hervé's comments on case_study
First part of intro: transition, VRES, electrofuels, ESOMs
First parts of intro and RL chapters

@@
 \newabbreviation[group={Acronyms}]{BECCS}{BECCS}{bioenergy with carbon capture and storage}
+\newabbreviation[group={Acronyms}]{BEMS}{BEMS}{building energy management system}
 \newabbreviation[group={Acronyms}]{BEV}{BEV}{battery electric vehicle}
@@
 %% D
+\newabbreviation[group={Acronyms}]{DC}{DC}{direct current}
 \newabbreviation[group={Acronyms}]{DHN}{DHN}{district heating network}
@@
 %% I
+\newabbreviation[group={Acronyms}]{IAMs}{IAMs}{integrated assessment models}
 \newabbreviation[group={Acronyms}]{ICE}{ICE}{internal combustion engine}
 \newabbreviation[group={Acronyms}]{IEA}{IEA}{International Energy Agency}
-\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{intergovernmental panel for climate change}
+\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{intergovernmental panel on climate change}
 \newabbreviation[group={Acronyms}]{IQR}{IQR}{interquatile range}
@@
 \newabbreviation[group={Acronyms}]{PC}{PC}{principal component}
+\newabbreviation[group={Acronyms}]{PCs}{PCs}{principal components}
 \newabbreviation[group={Acronyms}]{PCA}{PCA}{Principal Component Analysis}

### 2024-03-07 chap_RL/chap_RL.tex
1st section of RL chapter

@@
 
-\myparagraph{Reward}\\
-
-Properly defined the reward fed by the environment to the agent is crucial in \gls{RL} for several reasons.If the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
-Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
-On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.
-It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.
-
 \myparagraph{Actions}\\
@@
 
+\myparagraph{Reward}\\
+
+Properly defined the reward fed by the environment to the agent is crucial in \gls{RL} for several reasons.If the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
+Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
+On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.
+It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.
+
+Through its maximisation of the expected return (see Section \ref{subsec:meth_RL_algo}), a \gls{RL}-agent is as sensitive to positive reward, \ie the carrot, as negative reward, \ie the stick.
+When the former encourages desired behaviours, the latter can be seen as a penalty or a punishment and discourages the undesirable behaviours \cite{sutton2018reinforcement}.
+In our case, we have decided to combine these two approaches (see Figure \ref{fig:Reward}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.7\textwidth]{Reward.pdf}
+\caption{Reward, $R$, function scheme.
+Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded.
+If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues.
+Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
+\label{fig:Reward}
+\end{figure}
+
+First of all, taking a set of actions at a certain state might lead to an infeasible optimisation problem.
+In other words, as actions have a direct impact on some constraints of the problem, they might limit too much the feasible domain to the point where no solution can be found.
+For instance, the extreme case of aiming at carbon-neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve.
+In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300.
+If EnergyScope is able to provide a solution to the time window to optimise and the end of the transition, \ie 2050, is not reached, a test on the cumulative emissions so far.
+On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent.
+This penalisation is proportional to the difference between the \ce{CO2}-budget and the actual cumulative emissions.
+On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded.
+Eventually, if reaching 2050, we decided to tweak the reward function in capping the share of the cumulative emissions and integrating the transition cost.
+Given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
+Moreover, to make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans.
+cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€.
+This value comes from the mean of the total transition costs obtained through the \gls{GSA} performed on the perfect foresight transition pathway optimisation (see Section \ref{subsec:atom_mol:results_uq_cost}).
+In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost.
+The values of these weights are the results of a trial and error.
+This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.\\
+
+\newpage
 \myparagraph{States}\\
+
+Besides the reward, states are the other piece of information provided by the environment to the agent.
+In \gls{RL}, the purpose of states are to represent the current situation or configuration of the environment in which the agent operates.
+The primary function of states in RL is to provide the necessary context for the agent to choose appropriate actions based on its current observations and goals \cite{sutton2018reinforcement}.
+The challenge in the definition of the states is to provide enough information but not too much to avoid overwhelming the agent with non-informative features.
+
+Consequently, after another process of trial and error, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window.
+The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050.
+Therefore, the cumulative emissions emitted so far in the current step of the transition is the first dimension of the states.
+Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment.
+Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified by in Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency.
+The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy.
+Even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis.
+Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio.
 
@@
 Delaying the strengthening of mitigation action will increase the world’s dependence on CDR for holding warming to well below 2°C, and is likely to push the 1.5°C target out of reach for this century.}
+
+\section{Discussion}
+\label{sec:RL:discussion}
+
+Potentially interesting to implement reward shaping to accelerate learning or guide the agent towards achieving the desired behaviour more efficiently.
+In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode.
+However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy.
+Reward shaping addresses this issue by providing additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights.
+These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective.
+However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).

### 2024-03-10 appendices/app_results_UQ.tex
Caste_study post-reviews by FC
Edit since UQ_TD_full has been completed

@@
 \Cref{tab:UQ_full} gives the ranking and total Sobol index over the total transition cost of each of the 34 parameters listed in \Cref{tab:UC_full}.
-The first column shows these indicators for the \gls{GSA} applied on the monthly pathway model that has some limitations \cite{limpens2023pathway} but has the main advantage to run much faster.
-The second column gives the same indicators but for an uncertainty quantification carried out on the hourly pathway model and only using the most impacting parameters.
+The first column shows these indicators for the \gls{GSA} applied on the hourly pathway model.
+For information, the second column gives the same indicators but for an uncertainty quantification carried out on the monthly pathway model that has some limitations \cite{limpens2024pathway} but has the main advantage to run much faster.
 Given the similar rankings of the parameters between these two, this comparison shows that the monthly model can be a computationally efficient proxy to quantify the uncertainties of the actual hourly model and point out the key parameters on optimisation-driving objective, the total transition cost.
@@
 Besides the top-4 parameters, rankings are slightly different.
+The main difference in terms of ranking relates to the import of electricity from abroad, \ie its cost of purchasing and its availability.
+Indeed, as observed by \citet{limpens2024pathway}, the monthly model does not require this import given the easier integration of monthly-averaged local \gls{VRES}.
 However, this does not jeopardize the comparative analysis given the similar Sobol' indices.
@@
 \centering
-\begin{tabular}{l c c| c c}
+\begin{tabular}{l c c|c c}
 \toprule
 \multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
- & \multicolumn{2}{c|}{Monthly model} 	& \multicolumn{2}{c}{Hourly model} \\ 	
+& \multicolumn{2}{c|}{Hourly model} & \multicolumn{2}{c}{Monthly model}\\ 	
 \midrule
-\textbf{Purchase electrofuels} & 1 & (47.4\%) & 1 & (44.4\%) \\
-Industry EUD & 2 & (23.5\%) & 2 & (26.4\%)  \\
-Interest rate & 3 & (11.0\%) &  3 & (13.2\%)  \\
-Purchase fossil fuels  & 4 & (6.9\%) & 4 & (6.9\%)   \\
+\textbf{Purchase electrofuels} & 1 & (46.8\%) & 1 & (47.4\%) \\
+Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) \\
+Interest rate & 3 & (12.0\%) & 3 & (11.0\%) \\
+Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) \\
 \midrule
-Variable OPEX of technologies & 5 & (2.9\%) & 6 & (3.0\%) \\
-Purchase biofuels & 6 & (2.6\%) & 5 & (3.0\%) \\
-Hourly load factor PV & 7 & (1.9\%) & 9 & (1.3\%) \\
-CAPEX electric motor & 8 & (1.9\%) & 7 & (2.5\%) \\
-Hourly load factor wind turbines & 9 & (1.3\%) & 8 & (1.4\%) \\
-Max capacity PV & 10 & (1.1\%) & 14 & (0.5\%) \\
-\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (1.0\%) \\
-Available local biomass & 12 & (0.8\%) & 12 & (0.7\%) \\
-CAPEX car & 13 & (0.7\%) & 10 & (1.2\%) \\
-Passenger mobility EUD & 14 & (0.7\%) & 13 & (0.7\%) \\
-\midrule
-Modal share change LT-heat & 15 & (0.5\%) & \multicolumn{2}{c}{-} \\
-Households EUD & 16 & (0.5\%) & \multicolumn{2}{c}{-} \\
-Services EUD & 17 & (0.4\%) & \multicolumn{2}{c}{-} \\
-Max capacity onshore wind & 18 & (0.3\%) & \multicolumn{2}{c}{-} \\
-Max share of public transport & 19 & (0.3\%) & \multicolumn{2}{c}{-} \\
-CAPEX PV & 20 & (0.2\%) & \multicolumn{2}{c}{-} \\
-Max capacity offshore wind & 21 & (0.2\%) & \multicolumn{2}{c}{-} \\
-Efficiency electric motor & 22 & (0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX fuel cell engine & 23 & (0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX ICE & 24 & (0.1\%) & \multicolumn{2}{c}{-} \\
-Efficiency fuel cell engine & 25 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-Modal share change freight mobility & 26 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-Modal share change passenger mobility & 27 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX efficiency measures & 28 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX bus & 29 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX grid reinforcement & 30 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-CAPEX power grid & 31 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-\textbf{CAPEX \gls{SMR}} & 32 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-Available electricity import & 33 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
-Purchase electricity & 34 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+Variable OPEX of technologies & 5 & (3.1\%) & 5 & (2.9\%)\\
+Purchase biofuels & 6 & (2.6\%) & 6 & (2.6\%) \\
+CAPEX electric motor & 7 & (2.1\%) & 8 & (1.9\%) \\
+Purchase electricity & 8 & (1.5\%) & 34 & ($<$0.1\%) \\
+Hourly load factor wind turbines & 9 & (1.1\%) & 9 & (1.3\%)\\
+Hourly load factor PV & 10 & (1.1\%) & 7 & (1.9\%) \\
+\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (0.9\%)\\
+CAPEX car & 12 & (0.8\%) & 13 & (0.7\%) \\
+Available local biomass & 13 & (0.7\%) & 12 & (0.8\%) \\
+Passenger mobility EUD & 14 & (0.7\%) & 14 & (0.7\%) \\
+Modal share change LT-heat & 15 & (0.5\%) & 15 & (0.5\%)\\
+Max capacity PV & 16 & (0.5\%) & 10 & (1.1\%)\\
+Households EUD & 17 & (0.5\%) & 16 & (0.5\%) \\
+Services EUD & 18 & (0.5\%) &  17 & (0.4\%) \\
+Max share of public transport & 19 & (0.3\%) & 19 & (0.3\%) \\
+Max capacity onshore wind & 20 & (0.3\%) & 18 & (0.3\%) \\
+CAPEX PV & 21 & (0.2\%) & 20 & (0.2\%) \\
+Efficiency electric motor & 22 & (0.2\%) & 22 & (0.1\%)\\
+Max capacity offshore wind & 23 & (0.2\%) & 21 & (0.2\%) \\
+Available electricity import & 24 & (0.1\%) & 33 & ($<$0.1\%) \\
+CAPEX ICE & 25 & (0.1\%) & 24 & (0.1\%)\\
+CAPEX fuel cell engine & 26 & (0.1\%) & 23 & (0.1\%) \\
+Efficiency fuel cell engine & 27 & (0.1\%) & 25 & ($<$0.1\%) \\
+Modal share change freight mobility & 28 & (0.1\%) & 26 & ($<$0.1\%) \\
+Modal share change passenger mobility & 29 & ($<$0.1\%) & 27 & ($<$0.1\%) \\
+CAPEX grid reinforcement & 30 & ($<$0.1\%) & 30 & ($<$0.1\%) \\
+CAPEX efficiency measures & 31 & ($<$0.1\%) & 28 & ($<$0.1\%) \\
+CAPEX bus & 32 &($<$0.1\%)  & 29 & ($<$0.1\%) \\
+\textbf{CAPEX \gls{SMR}} & 33 & ($<$0.1\%)  & 32 & ($<$0.1\%) \\
+CAPEX power grid & 34 & ($<$0.1\%) & 31 & ($<$0.1\%) \\
 \bottomrule
@@
 
-% \begin{table}[htbp!]
-% \caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the monthly and hourly pathway models.
+%\begin{table}[htbp!]
+%\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the monthly and hourly pathway models.
 The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties over the hourly model.}
-% \label{tab:UQ_full}
-% \centering
-% \begin{tabular}{l c c| c c}
-% \toprule
-% \multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
-%  & \multicolumn{2}{c|}{Hourly model} 	& \multicolumn{2}{c}{Model model} \\ 	
-% \midrule
-% Purchase electrofuels & 1 & (44.4\%) & 1 & (47.4\%) \\
-% Industry EUD & 2 & (26.4\%) & 2 & (23.5\%)  \\
-% Interest rate & 3 & (13.2\%) &  3 & (11.0\%)  \\
-% Purchase fossil fuels & 4 & (6.9\%)  & 4 & (6.9\%)  \\
-% \midrule
-% Purchase biofuels & 5 & (3.0\%) & 6 & (2.6\%)  \\
-% Variable OPEX of technologies & 6 & (3.0\%) & 5 & (2.9\%) \\
-% CAPEX electric motor & 7 & (2.5\%) & 8 & (1.9\%)  \\
-% Hourly load factor wind turbines & 8 & (1.4\%) & 9 & (1.3\%) \\ 
-% Hourly load factor PV & 9 & (1.3\%) & 7 & (1.9\%)  \\
-% CAPEX car & 10 & (1.2\%) & 13 & (0.7\%) \\
-% \textbf{Potential capacity \gls{SMR}} & 11 & (1.0\%) & 11 & (0.9\%)  \\
-% Available local biomass & 12 & (0.7\%)  & 12 & (0.8\%) \\
-% Passenger mobility EUD & 13 & (0.7\%) & 14 & (0.7\%) \\
-% Max capacity PV & 14 & (0.5\%) & 10 & (1.1\%)  \\
-% \midrule
-% Modal share change LT-heat & \multicolumn{2}{c}{-} & 15 & (0.5\%)  \\
-% Households EUD & \multicolumn{2}{c}{-} & 16 & (0.5\%) \\
-% Services EUD & \multicolumn{2}{c}{-} & 17 & (0.4\%) \\
-% Max capacity onshore wind & \multicolumn{2}{c}{-} & 18 & (0.3\%) \\
-% Max share of public transport & \multicolumn{2}{c}{-} & 19 & (0.3\%) \\
-% CAPEX PV & \multicolumn{2}{c}{-} & 20 & (0.2\%) \\
-% Max capacity offshore wind & \multicolumn{2}{c}{-} & 21 & (0.2\%)\\
-% Efficiency electric motor & \multicolumn{2}{c}{-}  & 22 & (0.1\%) \\
-% CAPEX fuel cell engine & \multicolumn{2}{c}{-}  & 23 & (0.1\%) \\
-% CAPEX ICE & \multicolumn{2}{c}{-} & 24 & (0.1\%)\\
-% Efficiency fuel cell engine & \multicolumn{2}{c}{-}  & 25 & ($<$0.1\%) \\
-% Modal share change freight mobility & \multicolumn{2}{c}{-}  & 26 & ($<$0.1\%) \\
-% Modal share change passenger mobility & \multicolumn{2}{c}{-}  & 27 & ($<$0.1\%)\\
-% CAPEX efficiency measures & \multicolumn{2}{c}{-}  & 28 & ($<$0.1\%)\\
-% CAPEX bus & \multicolumn{2}{c}{-}  & 29 & ($<$0.1\%)\\
-% CAPEX grid reinforcement & \multicolumn{2}{c}{-} & 30 & ($<$0.1\%) \\
-% CAPEX power grid & \multicolumn{2}{c}{-} & 31 & ($<$0.1\%)\\
-% \textbf{CAPEX \gls{SMR}} & \multicolumn{2}{c}{-} & 32 & ($<$0.1\%) \\
-% Available electricity import & \multicolumn{2}{c}{-}  & 33 & ($<$0.1\%) \\
-% Purchase electricity  & \multicolumn{2}{c}{-} & 34 & ($<$0.1\%)\\
-% \bottomrule
-
-% \end{tabular}
-% \end{table}
+%\label{tab:UQ_full}
+%\centering
+%\begin{tabular}{l c c|c c| c c}
+%\toprule
+%\multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{6}{c}{\textbf{Ranking (Sobol' index)}}\\
+%& \multicolumn{2}{c}{Hourly model} & \multicolumn{2}{c|}{Monthly model} 	& \multicolumn{2}{c}{Hourly model} \\ 	
+%\midrule
+%\textbf{Purchase electrofuels} & 1 & (46.8\%) & 1 & (47.4\%) & 1 & (44.4\%) \\
+%Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) & 2 & (26.4\%)  \\
+%Interest rate & 3 & (12.0\%) & 3 & (11.0\%) &  3 & (13.2\%)  \\
+%Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) & 4 & (6.9\%)   \\
+%\midrule
+%Variable OPEX of technologies & 5 & (3.1\%) & 5 & (2.9\%) & 6 & (3.0\%) \\
+%Purchase biofuels & 6 & (2.6\%) & 6 & (2.6\%) & 5 & (3.0\%) \\
+%CAPEX electric motor & 7 & (2.1\%) & 8 & (1.9\%) & 7 & (2.5\%) \\
+%Purchase electricity & 8 & (1.5\%) & 34 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%Hourly load factor wind turbines & 9 & (1.1\%) & 9 & (1.3\%) & 8 & (1.4\%) \\
+%Hourly load factor PV & 10 & (1.1\%) & 7 & (1.9\%) & 9 & (1.3\%) \\
+%\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (0.9\%) & 11 & (1.0\%) \\
+%CAPEX car & 12 & (0.8\%) & 13 & (0.7\%) & 10 & (1.2\%) \\
+%Available local biomass & 13 & (0.7\%) & 12 & (0.8\%) & 12 & (0.7\%) \\
+%Passenger mobility EUD & 14 & (0.7\%) & 14 & (0.7\%) & 13 & (0.7\%) \\
+%Modal share change LT-heat & 15 & (0.5\%) & 15 & (0.5\%) & \multicolumn{2}{c}{-} \\
+%Max capacity PV & 16 & (0.5\%) & 10 & (1.1\%) & 14 & (0.5\%) \\
+%Households EUD & 17 & (0.5\%) & 16 & (0.5\%) & \multicolumn{2}{c}{-} \\
+%Services EUD & 18 & (0.5\%) &  17 & (0.4\%) & \multicolumn{2}{c}{-} \\
+%Max share of public transport & 19 & (0.3\%) & 19 & (0.3\%) & \multicolumn{2}{c}{-} \\
+%Max capacity onshore wind & 20 & (0.3\%) & 18 & (0.3\%) & \multicolumn{2}{c}{-} \\
+%CAPEX PV & 21 & (0.2\%) & 20 & (0.2\%) & \multicolumn{2}{c}{-} \\
+%Efficiency electric motor & 22 & (0.2\%) & 22 & (0.1\%) & \multicolumn{2}{c}{-} \\
+%Max capacity offshore wind & 23 & (0.2\%) & 21 & (0.2\%) & \multicolumn{2}{c}{-} \\
+%Available electricity import & 24 & (0.1\%) & 33 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX ICE & 25 & (0.1\%) & 24 & (0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX fuel cell engine & 26 & (0.1\%) & 23 & (0.1\%) & \multicolumn{2}{c}{-} \\
+%Efficiency fuel cell engine & 27 & (0.1\%) & 25 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%Modal share change freight mobility & 28 & (0.1\%) & 26 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%Modal share change passenger mobility & 29 & ($<$0.1\%) & 27 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX grid reinforcement & 30 & ($<$0.1\%) & 30 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX efficiency measures & 31 & ($<$0.1\%) & 28 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX bus & 32 &($<$0.1\%)  & 29 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%\textbf{CAPEX \gls{SMR}} & 33 & ($<$0.1\%)  & 32 & ($<$0.1\%) & \multicolumn{2}{c}{-} \\
+%CAPEX power grid & 34 & ($<$0.1\%) & 31 &  & \multicolumn{2}{c}{-} \\
+%\bottomrule							
+%
+%\end{tabular}
+%\end{table}
 
 \newpage
-\subsection{Imported renewable electrofuels}
+\section{Imported renewable electrofuels}
 \label{app:UQ_electrofuels}
@@
  & Q1 & - & - & 9 & 2\\
- & Median & - & - & 10 & 46\\
+ & Median & - & - & 10 & 47\\
  & Q3 & - & - & 11 & 55\\
@@
  & Q1 & - & - & 9 & 43\\
- & Median & - & - & 10 & 51\\
+ & Median & - & - & 10 & 50\\
  & Q3 & - & 1 & 12 & 57\\
@@
  \cmidrule{2 - 6}
- & Q1 & - & - & 9 & 44\\
- & Median & - & 4 & 11 & 52\\
- & Q3 & - & 16 & 32 & 58\\
+ & Q1 & - & - & 9 & 42\\
+ & Median & - & 5 & 11 & 51\\
+ & Q3 & - & 16 & 33 & 58\\
 \toprule
@@
  \cmidrule{2 - 6}
- & Q1 & - & - & 12 & 43\\
- & Median & - & 12 & 36 & 53\\
- & Q3 & 10 & 16 & 68 & 60\\
+ & Q1 & - & - & 10 & 41\\
+ & Median & - & 12 & 26 & 51\\
+ & Q3 & 8 & 16 & 68 & 60\\
 \toprule
@@
  \cmidrule{2 - 6}
- & Q1 & - & - & 24 & 43\\
- & Median & - & 13 & 49 & 52\\
- & Q3 & 44 & 17 & 77 & 60\\
+ & Q1 & - & - & 12 & 42\\
+ & Median & - & 12 & 37 & 52\\
+ & Q3 & 35 & 17 & 71 & 60\\
 \toprule
@@
  \cmidrule{2 - 6}
- & Q1 & - & - & 20 & 44\\
- & Median & 19 & 14 & 44 & 53\\
- & Q3 & 51 & 17 & 72 & 61\\
+ & Q1 & - & - & 11 & 43\\
+ & Median & 1 & 13 & 31 & 53\\
+ & Q3 & 36 & 17 & 66 & 61\\
 \bottomrule							

### 2024-03-10 chap_RL/chap_RL.tex
Edit since UQ_TD_full has been completed

@@
 
+\noindent
 Defining the levers of action, the core of the policy, to support the transition of a country-size whole-energy system is challenging, especially when accounting for political and socio-technical aspects \cite{castrejon2020making}.
@@
 
+\noindent
 Properly defined the reward fed by the environment to the agent is crucial in \gls{RL} for several reasons.If the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
@@
 
-\newpage
 \myparagraph{States}\\
 
+\noindent
 Besides the reward, states are the other piece of information provided by the environment to the agent.
@@
 
-\section{Results - As in Smart Energy Systems Conference}
+\section{Learning process and convergence}
+\label{sec:RL:learning}
+
+Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the training of the \gls{NN}.
+For this, numerous episodes are played through the myopic optimization of the transition pathway of Belgium.
+At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
+Moreover, a sample of values is drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
+This sample will remain valid for the following time windows.
+In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).
+Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward.
+This goes on until the end of the episode.
+Similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+
+\begin{itemize}
+\item Convergence of reward or success rate
+\item Actions to take, which ones are binding
+\item Show the state space: share of \gls{VRES} and efficiency
+\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
+\end{itemize}
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
+\caption{Binding versus non-binding constraints.
+In \gls{LP} where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$).
+The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution.
+These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
+\label{fig:Reward}
+\end{figure}
+
+\section{Testing and comparison with references}
+\label{sec:RL:testing}
 

### 2024-03-10 chap_atom_mol/chap_atom_mol.tex
Edit since UQ_TD_full has been completed

@@
 
+Beyond the direct impact that \gls{SMR} has on the power sector, it is worth assessing the other sectors given the sector-coupling effect related to the whole-energy system optimisation \cite{contino2020whole}.
+
 \myparagraph{High-temperature heat}\\
@@
 \midrule
-\textbf{Purchase electrofuels} & 1 & 47.4\% \\
-Industry EUD & 2 & 23.5\% \\
-Interest rate & 3 & 11.0\% \\
-Purchase fossil fuels  & 4 & 6.9\% \\
+\textbf{Purchase electrofuels} & 1 & 46.8\% \\
+Industry EUD & 2 & 23.2\% \\
+Interest rate & 3 & 12.0\% \\
+Purchase fossil fuels  & 4 & 5.7\% \\
 $\vdots$ & $\vdots$ & $\vdots$\\
@@
 $\vdots$ & $\vdots$ & $\vdots$\\
-\textbf{CAPEX \gls{SMR}} & 32 & $<$0.1\% \\
+\textbf{CAPEX \gls{SMR}} & 33 & $<$0.1\% \\
 \bottomrule
@@
 Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
-Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 11\% Sobol' index\footnote{It is important to note here that the model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
+Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index\footnote{It is important to note here that the model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
 In practice, the interest rate would vary depending on the technology investment risk.
@@
 \Cref{fig:UQ_PDF_total_transition_cost} shows the \gls{PDF} of the total transition cost given the 1260 samples.
-Stretching between 680\,b€ and 1940\,b€, the mean, the median and the nominal value (\ie REF case) are close to each other, respectively 1160\,b€, 1150\,b€ and 1080\,b€.
+Stretching between 660\,b€ and 2050\,b€, the mean, the median and the nominal value (\ie REF case) are close to each other, respectively 1180\,b€, 1160\,b€ and 1080\,b€.
 Similarly to the analysis carried out by \citet{coppitters2023optimizing}, one observes here that the distribution is right-skewed.
-It could then be qualified as ``fragile" as the top 50\% of the samples cover a bigger range (\ie between 1150 and 1940) than the bottom 50\% of the samples (\ie between 680 and 1150).
+It could then be qualified as ``fragile" as the top 50\% of the samples cover a bigger range (\ie between 1160 and 2050) than the bottom 50\% of the samples (\ie between 660 and 1160).
 In other words, the bad scenarios, resulting in a total transition cost higher than the median, have a bigger effect on this cost.
@@
 \caption{\acrfull{PDF} of the total transition cost.
-The mean, $\mu=1.16\cdot10^3$\,b€, is slightly higher than the median ($P_{50}=1.15\cdot10^3$\,b€) and the nominal cases cost, $1.08\cdot10^3$\,b€ and $1.04\cdot10^3$\,b€ respectively for the REF and SMR cases.
-Also, with a standard deviation, $\sigma=185$\,b€, a 95\%-confidence interval would be about [0.8; 1.5]$\cdot10^3$\,b€.}
+The mean, $\mu=1.18\cdot10^3$\,b€, is slightly higher than the median ($P_{50}=1.16\cdot10^3$\,b€) and the nominal cases cost, $1.08\cdot10^3$\,b€ and $1.04\cdot10^3$\,b€ respectively for the REF and SMR cases.
+Also, with a standard deviation, $\sigma=197$\,b€, a 95\%-confidence interval would be about [0.8; 1.6]$\cdot10^3$\,b€.}
 \label{fig:UQ_PDF_total_transition_cost}
@@
 As the general trends are increasing, discrepancies exist between the different energy carriers.
-E-methane, as the renewable alternative to fossil natural gas, substitutes it, sometimes at a very early stage of the transition, 2025, and to a somehow unrealistically large extent, 173\,TWh, which is more than 13\% more than the total import of electrofuels in the REF case.
+E-methane, as the renewable alternative to fossil natural gas, substitutes it, sometimes at a very early stage of the transition, 2025, and to a somehow unrealistically large extent, 163\,TWh, which is more than 6\% more than the total import of electrofuels in the REF case.
 The necessity to import this molecule is progressive through the transition to supply mostly industrial \gls{CHP} and boilers.
 
-E-hydrogen becomes rapidly the main stream of hydrogen in the system, on top of steam-methane-reforming or electrolysis, to reach a median and a maximum values of, respectively 13.6\,TWh and 40.7\,TWh in 2050.
+E-hydrogen becomes rapidly the main stream of hydrogen in the system, on top of steam-methane-reforming or electrolysis, to reach a median and a maximum values of, respectively 13.0\,TWh and 42.1\,TWh in 2050.
 Hydrogen is more frequently used in the mobility sector.
 Like in the REF case, fuel cell trucks are often the first option but, in some outlying cases, fuel cell cars and buses appear to completely substitute respectively \gls{BEV} cars and \gls{CNG} buses by 2050.
-Moreover, some samples lead to local production of methanol via the methanolation process, to produce up to 16.3\,TWh of methanol (\ie 30\% of the total supply of methanol of the nominal REF and SMR cases).
+Moreover, some samples lead to local production of methanol via the methanolation process, to produce up to 17.8\,TWh of methanol (\ie 33\% of the total supply of methanol of the nominal REF and SMR cases).
 
@@
 It is an indicator of statistical dispersion around the median, $Q2$ or $P_{50}$.} of about 50\,TWh.
-In some extreme cases, e-ammonia is the most imported molecules, \ie up to 233\,TWh or 63\% of the total primary mix in the REF case in 2050.
+In some extreme cases, e-ammonia is the most imported molecules, \ie up to 216\,TWh or 58\% of the total primary mix in the REF case in 2050.
 

### 2024-03-10 chap_intro_ccl/conclusion.tex
Merge taxonomy paper + Implement Hervé's comments on case_study
Edit since UQ_TD_full has been completed

--- 2023-12-21 chap_intro_ccl/conclusion.tex
@@
 List the future works to build upon the thesis
+
+\begin{itemize}
+\item \textbf{Word about sufficiency} oui mais c'est sans doute celle qui est le plus enviable car la moins risqué à tenter d'atteindre, les solutions mirages technologiques, si on y croit trop on met tout la dessus et si ca foire, on est encore plus dans la merde pcq la direction est mauvaise, la solution sobriété, meme si jamais atteint les efforts pr l'atteindre ne seront pas contre productif
+\item \textbf{Word about availability of electrfuels} Voir mémoire Ced et Simon
+\item Extract a roadmap representative of a multiple-run UQ or RL process.
+\end{itemize}

### 2024-03-13 chap_atom_mol/chap_atom_mol.tex
Implementation of FC's comments on chapter methodo

@@
 Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters.
-To mitigate the computational burden, we have used the \gls{PCE} approach (Section \ref{subsec:pce}) on the monthly version of EnergyScope Pathway.
-As detailed by \citet{limpens2024pathway}, this simplified model does not implement daily storage technologies and integrates much more easily intermittent renewables (PVs and wind turbines) in the system as the daily supply-demand intermittency and mismatch are not at stake.
-Consequently, it requires less flexibility options (\ie electrification of the heating sector or import of electricity from the neighbouring countries) and ends up with a cheaper transition (-3\%).
-Although, it allows to quantify the full set of 34 uncertain parameters (\ie 1260 samples for an order-2 \gls{PCE}), and avoid the computational burden of the hourly pathway.
-Nevertheless, comparing the \gls{UQ} on the hourly pathway with a limited set of uncertain parameters and on the monthly model, Appendix \ref{app:UQ_transition_cost} shows that the most impacting parameters are correctly captured by the latter.
-
 After briefly assessing the \gls{GSA} on the objective function of the model, the total transition cost, this section investigates more deeply the atom-molecules dilemma.

### 2024-03-13 appendices/app_EnergyScope.tex
Implementation of FC comments on chapter methodo

@@
 As an extension of Eq. \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
-%The objective is the minimisation of the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}),
 defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2_app}).
@@
 \hspace{1.7cm}
-\frac{remaining\_years(i,p)}{lifetime(y\textsubscript{start},i)} \left( \textbf{F\textsubscript{new}}(p,i) - 
+\frac{remaining\_years(i,p)}{\emph{lifetime}(y\textsubscript{start},i)} \left( \textbf{F\textsubscript{new}}(p,i) - 
 \sum_{\mathclap{p2 \in \emph{PHASE}}} 

### 2024-03-13 chap_methodo/chap_methodo.tex
Edit since UQ_TD_full has been completed
Implementation of FC comments on chapter methodo

@@
 First and foremost, such an extensive system needs to be represented, \ie modelled, to further be optimized.
-This model requires some characteristics to capture the peculiarities of this system (\eg intermittency of \gls{VRES}, sector coupling) while keeping an affordable computational time and an open-documentation.
+This model requires some characteristics to capture the peculiarities of this system such as the intermittency of \gls{VRES}, the coupling between the different energy and non-energy sectors and a pathway vision to pave the way from where we are to where we want to go.
 Then, as looking into the future (\ie up to 2050 in this work) comes with its lot of uncertainties, these have to be assessed carefully in terms of characterisation and quantification.
@@
 Finally, meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
-Finally, given the limited foresight in the uncertain future that has to deal with long-term objective, \ie limiting the cumulative \gls{GHG} emissions to a certain budget by 2050, this work encompasses the optimisation of the policy,\ie set of actions to take along the transition with a specific methodology to assess its robustness.
+This work encompasses the optimisation of the policy,\ie set of actions to take along the transition with a specific methodology to assess its robustness.
 
 Detailing the different tools needed to answer the research questions, this chapter starts with the presentation of the whole-energy system optimization model, EnergyScope Pathway and its myopic formulation.
-Then, the focus is put on the uncertainties, their characterisation as well as quantification.
+Then, it focuses on the uncertainties, their characterisation as well as their quantification.
 Finally, an agent-based \acrfull{RL} approach is detailed to address the sequential decision-making process in the uncertain transition with limited vision in the future.
@@
 \label{sec:meth:contributions}
-The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimize the behaviour of an artificial agent interacting with its environment, \ie the transition of the energy system.
+The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimize the behaviour of an artificial agent interacting with its environment, \ie the the whole-energy system through its transition.
 Given the uncertainties and potential shocks of the future, this approach allows the agent to play the transition in a sequential, \ie myopic, way and optimize the choice and the timing of its actions.
-This optimization is done through multiple tries-and-errors where the agent repeats the transition with a new set of uncertainties.
+This optimization is done through trial and error where the agent repeats the transition with a new set of uncertainties.
 
@@
 Originally developed to optimize the transition in one global optimisation up to 2050, \ie perfect foresight, part of this thesis consisted in making this model able to optimize the same transition but with sequential more limited time windows, \ie myopic approach.
+Besides its shorter computational time, this approach is more representative an actual decision process with limited foresight in the future \cite{babrowski2014reducing}.
 
-The third principal methodological added-value is the use of \acrfull{PCA} to assess the robustness of a policy.
+The third principal methodological contribution is the use of \acrfull{PCA} to assess the robustness of a policy, \ie how much the optimal pathway is affected by the uncertainties.
 Given the uncertainties and the timespan of the transition, this approach allows highlighting the main \og directions'' of variation of the system design (\ie the installed capacities).
@@
 After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model subject to different emission-constraints \citet{rixhon2021role} as well as the pathway model.
-Eventually, starting from the initial investigation of \citet{goffauxpathway}, this work has converged to the most relevant formulation of the salvage value, \ie \textbf{C\textsubscript{inv,return}} in Eq.
-\ref{eq:salvage}, for the model EnergyScope.
+Eventually, starting from the initial investigation of \citet{goffauxpathway}, this work has converged to the most relevant formulation of the salvage value for the model EnergyScope.
+This aims at considering the residual value of assets that would still be in place after the end of the optimisation.
+It avoids penalising capital-intensive and long-lasting asset.In this formulation, the capacities that have been anticipatively decommissioned are removed from the total installed capacities.
+Therefore,  Therefore, this penalises decisions that would lead to investments that are later decommissioned before having reached the end of their lifetime.
 
@@
 From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the former approach to further dig into more details about the latter.
-The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from a snapshot approach, EnergyScope TD.
+The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from, EnergyScope TD, the a snapshot model, optimising a target future year with a greenfield approach \cite{geidl2006greenfield}.
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
@@
 The whole-energy system model developed in this work originates from the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core}) of EnergyScope---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050 \cite{limpens2024pathway}.
-All the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}, and summarized in Appendix \ref{app:ESTD}, are kept as is with an extra-dimension to relate them to a specific representative year, $y$, of the pathway.
-For instance, the energy balance is guaranteed at every hour of each of these years.
+Each representative year is represented via the the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
+Then, to draw a consistent pathway between these years, additional constraints aim at linking them, \eg limiting the modal shifts within some energy sectors.
 
-The optimised objective of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as follows:
+The objective function of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as the sum of the total \gls{CAPEX}, \textbf{C\textsubscript{tot,capex}}, and the \gls{OPEX}, \textbf{C\textsubscript{tot,opex}} (see Eq.
+\ref{eq:obj_func_v2}).
+
+\begingroup
+\vspace{-0.3cm}
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\begin{flalign} 
+% Objective function + investment 
+\label{eq:obj_func_v2}%1
+% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
+\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&%\\
+\end{flalign}
+\endgroup
+
+%\label{eq:Capex_v2_app}
+%& \textbf{C\textsubscript{tot,capex}} =
+%\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
+%\textbf{C\textsubscript{inv,phase}}(p)
+%-
+%\sum_{\mathclap{i \in \emph{TECH}}} 
+%\textbf{C\textsubscript{inv,return}}(i)\\
+%  \label{eq:Copex_tot_v2_app}%5
+%& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
+%+ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}}
+% \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
+%\label{eq:path_annu_factor_app}
+%& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+
+The total \gls{CAPEX} is the difference between the total investments done during the transition, $\textbf{C\textsubscript{inv,phase}}$, and the residual value of the assets that would still be in place after the end of the transition, $\textbf{C\textsubscript{inv,return}}$ (see Eq.
+\ref{eq:CAPEX_v2}).
+
+\begingroup
+\begin{flalign} 
+\belowdisplayskip=2pt
+\abovedisplayskip=2pt
+\label{eq:CAPEX_v2}
+\hspace{0pt}&\textbf{C\textsubscript{tot,capex}} =
+\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
+\textbf{C\textsubscript{inv,phase}}(p)
+- \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{inv,return}}(j),&
+\end{flalign}
+\endgroup
+
+\noindent
+where $\textbf{C\textsubscript{inv,phase}}$ is given as the sum over all the  newly installed technologies at the phase $p$, $\textbf{F\textsubscript{new}}(p)$, multiplied by the average investment costs, $\emph{c\textsubscript{inv}}$,  of the year starting and ending the corresponding phase, $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$ respectively (see Eq.
+\ref{eq:PhaseInv}).
 
@@
 \abovedisplayskip=2pt
-
 \begin{flalign} 
-% Objective function + investment 
-\label{eq:obj_func_v2}%1
-% adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
-\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&\\
-\label{eq:CAPEX_v2}
-& \textbf{C\textsubscript{tot,capex}} =
-\sum_{\mathclap{p \in \text{\emph{PHASE}}\cup \{2015\_2020\}}} 
-\textbf{C\textsubscript{inv,phase}}(p)
--
-\sum_{\mathclap{j \in \emph{TECH}}} 
-\textbf{C\textsubscript{inv,return}}(j)\\
-  \label{eq:Copex_tot_v2}%5
-& \textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
-+ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)}}
- \Big(\textbf{C\textsubscript{opex}}(y\textsubscript{start}) + \textbf{C\textsubscript{opex}}(y\textsubscript{stop}) \Big)/2&\\
-\label{eq:path_annu_factor}
-& \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
-%% GL_correction_v2 %% & \tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}} &
+\hspace{0pt} 
+\label{eq:PhaseInv}%5
+&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \emph{c\textsubscript{inv}}(p,j)&\forall p \in \emph{PHASE},
 \end{flalign}
 \endgroup
+
+\noindent where $\tau\textsubscript{\emph{phase}}$ is the annualised phase factor and $\emph{c\textsubscript{inv}}(p,j)$ is the arithmetic average of the investment cost of the technology $j$ at the beginning and the end of the phase, $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j)$ and $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)$.
+Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed (see Eq.
+\ref{eq:salvage})
 
@@
 \label{eq:salvage}%5
-&\textbf{C\textsubscript{inv,return}}(i) = \hspace{2.5cm}
-\sum_{\mathclap{p \in \emph{PHASE}\cup \{2015\_2020\}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}} 
+&\textbf{C\textsubscript{inv,return}}(j) = \sum_{\mathclap{p \in \emph{PHASE}\cup \{2015\_2020\}}} 
 \hspace{0.5cm}
- \tau_{phase}(p)\cdot \left(c_{inv}(y_{start},i)+c_{inv}(y_{stop},i)\right)/2 \cdot
+ \tau_{phase}(p)\cdot \emph{c\textsubscript{inv}}(p,j) \cdot
 &\notag\nonumber
@@
 \hspace{1.7cm}
-\frac{remaining\_years(i,p)}{lifetime(y\textsubscript{start},i)} \left( \textbf{F\textsubscript{new}}(p,i) - 
+\frac{remaining\_years(j,p)}{\emph{lifetime}(y\textsubscript{start},j)} \left( \textbf{F\textsubscript{new}}(p,j) - 
 \sum_{\mathclap{p2 \in \emph{PHASE}}} 
-\textbf{F\textsubscript{decom}}(p2,p,i)\right)&
-\forall i \in \emph{TECH}
+\textbf{F\textsubscript{decom}}(p2,p,j)\right)&
+\forall j \in \emph{TECH}
+\end{flalign}
+\endgroup
+
+About the total \gls{OPEX} of the transition, $\textbf{C\textsubscript{tot,opex}}$, on top of the initial costs in 2020, we assume that the \gls{OPEX} of a phase is equal to the average operational costs, $ \textbf{C\textsubscript{opex}}$,  of $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$, multiplied by the duration of a phase $\emph{t\textsubscript{phase}}$ equal to 5 years in our case (see Eq.
+\ref{eq:Copex_tot_v2}).
+
+\begingroup
+\begin{flalign} 
+  \label{eq:Copex_tot_v2}%5
+&\textbf{C\textsubscript{tot,opex}} =  \textbf{C\textsubscript{opex}}(2020)
++ \emph{t\textsubscript{phase}}\cdot \tau\textsubscript{\emph{phase}}(p) \cdot \sum_{\mathclap{p \in \emph{PHASE}}} 
+\textbf{C\textsubscript{opex}}(p),&
 \end{flalign}
@@
 \noindent
-where $\emph{t\textsubscript{phase}}=5\,\text{years}$ and $\emph{diff\_2015\_year(p)}$ are respectively the duration of a phase between two representative years and the number of years between the middle of a phase and 2015 for a correct annualisation.
-$\textbf{C\textsubscript{inv,return}}$ accounts for the residual value, also called \textit{salvage value}, of the technologies installed during the transition and having not reached the end of their lifetime by 2050.
-This last variable is crucial to avoid penalising heavy (and potentially long-lifetime) investments at the end of the transition as these assets would still be operational beyond 2050.
-The interested reader will find more information about the formulation choices related to it in the work of \citet{limpens2024pathway}.
-The other variables in Eq.\,(\ref{eq:CAPEX_v2}-\ref{eq:Copex_tot_v2}) are detailed here below:
+where $\textbf{C\textsubscript{opex}}(p)$ is the arithmetic average of the operational cost at the beginning and the end of the phase $p$,  $\textbf{C\textsubscript{opex}}(y\textsubscript{start})$ and $\textbf{C\textsubscript{opex}}(y\textsubscript{stop})$.
+The operational cost of a year, $y$, $\textbf{C\textsubscript{opex}} (y)$, is the sum of the costs related to maintenance and operation of technologies, $ \textbf{C\textsubscript{maint}}$, and the consumption of resources, $\textbf{C\textsubscript{op}}$ (see Eq.
+\ref{eq:opex_yearly}).
 
@@
 \label{eq:opex_yearly}
-&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i) \hspace{2.9cm} \forall y \in \text{\emph{YEARS}}\\
-\label{eq:PhaseInv}%5
-&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \left(\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j) + \emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)\right)/2&\notag\nonumber
-\end{flalign}
-\begin{flalign}
-&&\forall p \in \emph{PHASE} | y\textsubscript{start}\in \emph{P\_START}(p),y\textsubscript{stop}\in \emph{P\_STOP}(p)
+&\textbf{C\textsubscript{opex}} (y) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{C\textsubscript{maint}}(y,j) + \sum_{\mathclap{i \in \emph{RES}}} \textbf{C\textsubscript{op}}(y,i) & \forall y \in \text{\emph{YEARS}},
 \end{flalign}
@@
 \noindent
-where $\textbf{F\textsubscript{new}}$ are the capacities newly installed. In Eq. (\ref{eq:opex_yearly}-\ref{eq:PhaseInv}), the costs related to each representative year are:
+where the costs related to each representative year are:
 
@@
 \hspace{0pt} 
- \label{eq:c_inv}%3
- &\textbf{C\textsubscript{inv}}(y,j) = c_{\text{\emph{inv}}}(y,j) \textbf{F}(y,j) & \forall y \in \text{\emph{YEARS}}, \forall j \in \text{\emph{TECH}}\\
  \label{eq:c_maint}%4
@@
  &\textbf{C\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} c_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t) t_{op} (t)  
- & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
+ & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}},
  \end{flalign}
@@
 
-\noindent where the variable $\textbf{F}$ represents the size of the installed capacities (for all technologies $j$) and the variable $\textbf{F\textsubscript{t}}$ is the hourly consumption of the resources; the parameters $c_{\text{\emph{inv}}}$ and $c_{\text{\emph{maint}}}$ are the CAPEX and the OPEX of the technologies, and the parameter $c_{\text{\emph{op}}}$ is the cost of purchasing resources.
+\noindent where the variable $\textbf{F}$ represents the size of the installed capacities (for all technologies $j$) and the variable $\textbf{F\textsubscript{t}}$ is the hourly consumption of the resources; the parameter $c_{\text{\emph{maint}}}$ is the OPEX of the technologies, and the parameter $c_{\text{\emph{op}}}$ is the cost of purchasing resources.
 For the sake of simplicity, as done by \citet{limpens2024pathway}, the sum over the 8760 hours of the year is written as the sum over $t \in T $.
 
-Then, as detailed in section \ref{sec:cs:CO2-budget}, the \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$, is computed and constrained as follows:
+he \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Eq.
+\ref{eq:gwp_tot_transition}).
+Similarly to initial operational costs to account for the system in place in 2020 (see Eq.
+\ref{eq:Copex_tot_v2}), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
+Then, as detailed in Section \ref{sec:cs:CO2-budget}, these cumulative emissions is constrained by a budget (see Eq.
+\ref{eq:limit_gwp_trans}).
 
@@
 \label{eq:gwp_tot_transition}
-&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}|y\textsubscript{start}\in \emph{Y\_START}(p),y\textsubscript{stop}\in \emph{Y\_STOP}(p)}}\left(\textbf{GWP\textsubscript{tot}}(y\textsubscript{start}) +\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop}) \right)/2 &
+&\textbf{GWP\textsubscript{tot,trans}}= \textbf{GWP\textsubscript{tot}}(2020) + \emph{t\textsubscript{phase}}\sum_{\mathclap{p \in \emph{PHASE}}}\textbf{GWP\textsubscript{tot}}(p) &
 \\
 \label{eq:limit_gwp_trans}
-& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}}&
+& \textbf{GWP\textsubscript{tot,trans}} \leq \emph{gwp\textsubscript{lim,trans}},&
 \end{flalign}
@@
 \noindent
-where the computation of the yearly emissions are based on the \acrfull{GWP} of the resources:
+where $\textbf{GWP\textsubscript{tot}}(p)$ is the arithmetic average of the yearly emissions at the beginning and the end of the phase $p$,  $\textbf{GWP\textsubscript{tot}}(y\textsubscript{start})$ and $\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop})$.
+The computation of these yearly emissions are based on the \acrfull{GWP} of the resources:
 
@@
   \label{eq:GWP_op}%7
- & \textbf{GWP\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t)  t_{op} (t) & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}}
+ & \textbf{GWP\textsubscript{op}}(y,i) = \sum_{\mathclap{t \in T }} gwp_{\text{\emph{op}}}(y,i) \textbf{F\textsubscript{t}}(y,i,t)  t_{op} (t) & \forall y \in \text{\emph{YEARS}}, \forall i \in \text{\emph{RES}},
 \end{flalign}
@@
 Moreover, to account for the society inertia and to prevent unrealistically fast modal share change, constraints limit this change for the sectors of the low-temperature, the passenger mobility and freight mobility demands.
-The parameters $\Delta_{\mathrm{change,LT\_heat}}$, $\Delta_{\mathrm{change,pass}}$ and $\Delta_{\mathrm{change,freight}}$ respectively limit their respective modal share change up to 33\%, 50\% and 50\% per phase of 5 years.
+The interested reader will find more information about the formulation choices related to it in the work of \citet{limpens2024pathway}.
 
@@
 \citet{babrowski2014reducing} analysed the benefit of the myopic approach to reduce the computational time.
-\citet{poncelet2016myopic} uses this approach to analyse the expansion planning of the power sector beyond 2050.
+\citet{poncelet2016myopic} uses this approach to analyse the expansion planning of the power sector beyond 2050 to assess the realism of the decision making process brought up by the myopic implementation.
 \citet{nerini2017myopic} analysed the impact of the horizon windows and overlapping time.
@@
 In their analysis of the overcapacity in European power systems, \citet{moret2020overcapacity} emphasised that such a ``possibility of \textit{recourse}" is very appropriate to address uncertainty gradually unfolding over time.
-Consequently, the development of the myopic approach represents the foundations of the further implementation of the agent-based reinforcement learning framework (see Section \ref{sec:meth:RL}).
+Consequently, the development of the myopic approach with an overlap between two successive time windows has been implemented.
+This sequential optimisation framework also represents the foundations of the further implementation of the agent-based reinforcement learning framework (see Section \ref{sec:meth:RL}).
 
@@
 Consequently, the solution obtained at the end of the first time window (\eg 2030) as well as potential investment decisions between the start of the second time window and this end-year are discarded.
-In other words, they are not taken into account for the optimisation of the second time window.
+In other words, they are not taken into account for the optimisation of the second time window since the new final year is further into the future.
 This process goes on until the stated end of the transition (\ie 2050, in this case).
@@
 The major add-on from the original EnergyScope Pathway model \cite{limpens2021generating} to the myopic version developed in this thesis, is the possibility to carry out the optimisation on a limited time window, of which the duration is defined by $N\textsubscript{year,opti}$.
-Moreover, to ensure the ``possibility of \textit{recourse}", there is also the possibility of having an overlap between two consecutive time windows.
+Moreover, there is also the possibility of having an overlap between two consecutive time windows.
 The timespan of this overlap is defined by the parameter $N\textsubscript{year,overlap}$.
@@
 This way, the newly developed model can easily be used to perform a perfect foresight optimisation by setting the time window to $N\textsubscript{year,opti}=$30 years (\ie between 2020 and 2050) and the overlap between the time windows to $N\textsubscript{year,overlap}=$0.
-Consequently, as defining the actual time window on which the system is optimised as well as the history, \ie what has already been optimised earlier in the transition, are fundamental, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
+Consequently, as it is fundamental to define, on the one hand, the actual time window on which the system is optimised, and on the other hand, the history, \ie what has already been optimised earlier in the transition, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
 
@@
 \label{tab:path_my_sets}
-\begin{minipage}{\textwidth}
+%\begin{minipage}{\textwidth}
 \centering
-\resizebox{\textwidth}{!}{
+%\resizebox{\textwidth}{!}{
 \begin{tabular}{l c l}
@@
 \bottomrule
-\end{tabular}}
-\end{minipage}
+\end{tabular}%}
+%\end{minipage}
 \end{table}
 
-$\text{YEARS}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{WND}$ substitute $\text{YEARS}$ and $text{PHASE}$ in the constraints defined in the pathway model in Section \ref{subsec:meth:PF}.
+$\text{YEARS}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{WND}$ substitute $\text{YEARS}$ and $\text{PHASE}$ in the constraints defined in the pathway model in Section \ref{subsec:meth:PF}.
 These two sets aim at setting the optimization to a more limited time window.
@@
 On top of these four specific sets, some artefacts were also necessary to avoid computational rounding errors.
-Indeed, optimizing the first year of a time window that has already been optimized in the previous time window could lead to rounding errors preventing from the optimization to converge.
+For instance, the first year of a time window is the result of the optimisation of the previous one.
+Therefore, optimising again this first year could lead to rounding errors preventing from the optimization to converge.
 For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimize that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
@@
 \noindent
-As expected, the computational time is reduced drastically (\ie by 55\%) while the design remains similar.
+In line with the work of \citet{babrowski2014reducing}, the computational time is reduced drastically (\ie by 55\%).
+On top of this, we observed that the resulting design, \ie the technological mix, remains similar.
 Given the continuous change of the input parameters over the considered time frame, the perfect foresight and myopic approaches results are very similar, like in \cite{krey2006vergleich}: less than 1\% cost difference over the transition, similar system designs by 2050 and slight shifts in time in terms of adoption of technologies.
 
-The main difference lies in the transition itself and especially in the earlier deployment of PVs and offshore wind turbines.
+The main difference lies in the myopic transition itself and especially in the earlier deployment of PVs and offshore wind turbines.
 These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset.
-This is mostly due to the chosen formulation of the salvage value, Eq.
-\ref{eq:salvage}.
-Since the objective function is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at early stages of the transition.
+This is mostly due to impact of the salvage value, Eq.
+\ref{eq:salvage}, in the objective function.
+Since this is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at early stages of the transition.
 A more detailed comparison between the myopic and perfect foresight approaches is available in Appendix \ref{app:my_vs_pf}.
@@
 Moreover, some ranges have been added to account for new parameters coming from the pathway formulation described in Section \ref{sec:meth:ES} like the society inertia.
-Like other works \cite{li2019renewables,coppitters2021robust}, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+Like other works \cite{li2019renewables,coppitters2021robust} and given the scarcity of information, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+Alternatives like PERT or Gaussian distributions could also have been considered \cite{coppittersthesis}.
 
@@
 
-Following the methodology defined by \citet{Moret2017}, uncertainties of types I (investment-type) and II (operation-type, constant uncertainty over time) keep the same range for the whole transition.
-However, parameters with an uncertainty increasing over time, type III, (\ie end-use demands, in this case) will have a wider and wider range over the transition.
-In this work, a +50\% linear increase has been arbitrarily set between the width of the range of such parameters in 2025 and the same ranges in 2050.
+Following the methodology defined by \citet{Moret2017}, uncertainties of types I (investment-type) and II (operation-type, constant uncertainty over time) keep the same range width for the whole transition.
+In other words, unlike type III parameters, this width is not expanding (nor narrowing) for the different representative years of the transition.
+However, parameters with an uncertainty increasing over time, type III, (\ie end-use demands, in this case) will have a wider and wider range over the transition (see Figure \ref{fig:ranges_transition}).
+In this work, a +50\% linear increase has been set between the width of the range of such parameters in 2025 and the same ranges in 2050.
+This choice leads to an industrial \gls{EUD} that could be, in 2050, -30.8\% compared to its nominal value.
+This potential drop compared to the reference is in line with the work of \citet{My2050}.
+In their work, the total energy demand in the industry sector in 2050 could be between -19\% to -50\% of the reference value, depending on the scenario.
 In \Cref{fig:ranges_transition}, this means that for type III uncertainties only, $R_{2050}^+$ is 50\% bigger than $R_{2025}^+$ and $R_{2050}^-$ is 50\% smaller than $R_{2025}^-$.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{ranges_transition.pdf}
-\caption{$\mu_{2020}$, $\mu_{2025}$, ...,  $\mu_{2050}$ are the nominal values equal to 0 as the uncertain parameters represent a relative increase/decrease of actual parameters of the model.
+\includegraphics[width=0.7\textwidth]{ranges_transition.pdf}
+\caption{Expansion of the width of uncertainty range for type III parameters.
+$\mu_{2020}$, $\mu_{2025}$, ...,  $\mu_{2050}$ are the nominal values equal to 0 as the uncertain parameters represent a relative increase/decrease of actual parameters of the model.
 $R^+$ and $R^-$ are respectively the upper and lower bounds of the range and $\xi_{2025}$, $\xi_{2030}$, ...,  $\xi_{2050}$ are the values taken by one parameter for a specific sample of the \gls{GSA} for each of the representative years of the transition, always starting from the nominal value in 2020, $\mu_{2020}$.
@@
 Finally, the model accounts for thousands of parameters.
-The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs).
-Therefore, similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped to one single uncertain parameter.
+The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs\footnote{As detailed in Section \ref{subsec:pce}, the number of runs required for the \gls{GSA} is proportional to the factorial of the number of uncertain parameters.
+As second order \gls{PCE} is the minimum to ensure accuracy of the surrogate model, considering thousands of independent uncertain parameters would lead to millions of runs, if no more.}).
+Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped to one single uncertain parameter.
+On top of mitigating the computational burden, this aims at grouping parameters that are closely linked with each other.
 For instance, the uncertainty on the cost of purchasing renewable electrofuels, $c_{\mathrm{op,electrofuels}}$, identically affects the cost of e-hydrogen, e-methane, e-ammonia and e-methanol.
 Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation of cost of electricity or the electrolyser, that drive the majority of their cost of purchasing \cite{h2coalition}.
-Similarly, the uncertainties impacting the industrial demand, $industry\_EUD$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand.
+Similarly, the uncertainties impacting the industrial demand, $industry\_\emph{EUD}$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand.
 
@@
 
-We used \gls{PCE}, an approach for surrogate-assisted \gls{UQ}, to propagate uncertainties in input parameters through the system model.
+To avoid the avoid the computational burden of well-known method like Monte-Carlo analysis \cite{yue2018review}, we used \gls{PCE} to carry out a \gls{GSA}.
+\gls{PCE} is an approach for surrogate-assisted \gls{UQ} that propagates uncertainties in input parameters through the system model.
 This allowed us to assess statistical moments on the quantity of interest and determine Sobol' indices~\cite{coppitters2020robust}.
@@
 \noindent
-The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie results) in the EnergyScope Pathway model ($M$).
+The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie the value of the objective function, see Eq.
+\ref{eq:obj_func_v2}) in the EnergyScope Pathway model ($M$).
 This representation is constructed as a truncated series of multivariate orthonormal polynomials $\bm{\Psi}$, weighted by coefficients $u$:
@@
 \noindent where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (\autoref{app:UC_full}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$.
+The coefficients ($u_0, u_1, \dots, u_{P+1}$) are quantified using a regression method applied to orthonormal polynomials~\cite{Sudret2014}.
 As uniform distributions are considered, the Legendre polynomials are adopted, as they are the associated family of polynomials that are orthogonal with respect to standard uniform distributions~\cite{Sudret2014}.
@@
 
-The coefficients ($u_0, u_1, \dots, u_{P+1}$) are quantified using a regression method applied to orthonormal polynomials~\cite{Sudret2014}.
 To ensure a well-posed least-square minimisation, it is recommended to have a number of training samples at least twice the number of coefficients~\cite{Sudret2014}.
@@
 Starting with $p=1$, a PCE is constructed and the \gls{LOO} error is evaluated.
-If the \gls{LOO} error is below a specified threshold, the corresponding polynomial order is considered sufficient for generating an accurate PCE.
+If the \gls{LOO} error is below a specified threshold ($\sim 1\%$), the corresponding polynomial order is considered sufficient for generating an accurate PCE.
 However, if the error exceeds the threshold, the order is increased, and additional samples are generated following the rule of Eq.\,(\ref{eq:pce:nterms}).
@@
 \begin{equation}
-  EE_{i}=\frac{M(\xi_1,\xi_2,...,\xi_i+\Delta,...,\xi_d)-M(\vec{\xi})}{\Delta}
+  EE_{i}=\frac{M(\xi_1,\xi_2,...,\xi_i+\Delta,...,\xi_d)-M(\vec{\xi})}{\Delta},
 \end{equation}
@@
 \citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2}-limits using the snapshot model, \ie EnergyScope TD \cite{limpens2019energyscope}.
-Screening the results of this work, we have discarded some parameters with negligible impact (\eg CAPEX of electrolysers or variation of the freight demand), selected a subset of parameters and added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
+Screening the results of this work, we have discarded some parameters with negligible impact\footnote{Per \citet{Turati2017}, parameters are considered as ``negligible'' if their Sobol' index is below the threshold $=1/d$, $d$ being the total number of uncertain parameters} (\eg CAPEX of electrolysers or variation of the freight demand), selected a subset of parameters and added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
 The exhaustive list of these 34 parameters is presented in Appendix \ref{app:UC_full}.
@@
 % 7. Baseline and Comparisons --> Results
-% 8. Implementation Details
+% 8. Implementation Details --> Chap_RL --> DONE
 % 9. Experimental/Testing Setup
@@
 including all streams of energy carriers and demands) is uncertain.
-Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), i.e.
-perfect foresight, policy makers rather go through multiple rolling-horizon short-term decisions, i.e.
-myopic approach.
+Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), policy makers rather go through multiple rolling-horizon short-term decisions.
 Yet, these decisions can have long-term impacts, 20 to 50 years.
+This long-term future is intrinsically uncertain and could be the place for potential sudden unexpected events.
 Meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
@@
 
-\begin{wrapfigure}{r}{3cm}
-\centering
-\captionsetup{justification=centering}
-\includegraphics[width=2.8cm]{RL_Fundamentals.pdf}
-\caption{}
-\label{fig:RL_Fundamentals}
-\end{wrapfigure}
-
-\Gls{RL} is a subfield of machine learning focused on training an agent to make sequential decisions by interacting with an environment to achieve specific goals.
-Unlike supervised learning, where data is labeled, and unsupervised learning, where patterns are inferred from unlabeled data, reinforcement learning deals with learning from interaction, typically through trial and error.
+\Gls{RL} is a subfield of machine learning focused on training an agent to make sequential decisions by interacting with an environment to achieve specific goals (see Figure \ref{fig:RL_Fundamentals}).
+Unlike supervised learning, where data is labelled, and unsupervised learning, where patterns are inferred from unlabelled data, reinforcement learning deals with learning from interaction, typically through trial and error.
 This way, \gls{RL} is considered as active learning \cite{cao2020reinforcement}.
@@
 When the episode is done, the agent starts again from an initial state, takes an action and so on.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.3\textwidth]{RL_Fundamentals.pdf}
+\caption{General concept of \acrfull{RL} as the interactions between the agent and its environment.
+The agent takes some action that has an impact on the environment which feeds back the agent with a reward and the new state.
+The objective of the agent is to optimize its policy, \ie mapping between the state it is at and the action to take, by maximizing its cumulative reward.}
+\label{fig:RL_Fundamentals}
+\end{figure}
+
 The agent learns to optimize its policy by maximizing a notion of cumulative reward over time.
@@
 All these concepts and interactions between the agent and its environment are formalized as a \gls{MDP} \cite{sutton2018reinforcement}, represented by the tuple $<s,a,T,r,\pi,\gamma>$.
-$s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions.
-$a\in A$ is the action among the action space, $A$, and $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$.$r$ is the reward received by the agent when taking the action $a$ from a state $s$, $R(s,a)$.
-Finally, $\pi$ is the policy as already described and $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
+The Markov property of such a decision process states that a decision is made only based on this tuple and not on the history/path that has led to it.
+In this tuple, $s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions; $a\in A$ is the action among the action space, $A$; $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$; $r$ is the reward received by the agent when taking the action $a$ from state $s$, $R(s,a)$; $\pi$ is the policy telling the action to take depending on the current state and; $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
 During the learning/optimization process, the agent acts according to the exploitation-exploration trade-off.
-In the former case, the action $a$ is directly given the mapping provided by the current policy $\pi$,  depending on the state $s$.
-In the latter, the action is randomly picked within the action space.
+In the exploitation, the action $a$ is directly given by the mapping provided by the current policy $\pi$,  depending on the state $s$.
+In the exploration, the action is randomly picked within the action space.
 For further information, the interested reader is invited to refer to work of \citet{sutton2018reinforcement} or the course given by \citet{davidsilver_RL_online} available online.
 
-Due to the increasing complexity of the systems and the integration of uncertainties, the last decades have seen the emergence of publications where \gls{RL} is applied to energy systems \cite{perera2021applications}.
+Due to the increasing complexity of the systems and the integration of uncertainties, the last decades have seen the emergence of publications where \gls{RL} is applied to energy systems \cite{cao2020reinforcement,perera2021applications}.
 In their respective reviews, \citet{cao2020reinforcement} and \citet{perera2021applications} highlighted groups of problems addressed with \gls{RL} in the research field of energy systems: building energy management system (BEMS), optimization of dispatch and operational control closely linked with the energy market and the optimal power flow problem in the grid, micro-grid management, electro-mobility or even demand-side management or optimal control of energy system devices like maximum power point tracking (MPPT) of wind turbines and \gls{PV} panels.
-The major novelty of this thesis in the use of \gls{RL} stands then in its application to a new kind of energy system problem: the optimization of the transition pathway of a whole-energy system.
+The major novelty of this thesis is the application of \gls{RL} to a new kind of energy system problem: the optimization of the transition pathway of a whole-energy system.
 In this sense, the objective is to optimize and provide a policy to support this transition subject to uncertainties.
@@
 
+Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of \gls{RL} with \gls{DNN}, called \gls{DRL}.
+Among others, one of the main drawbacks of traditional \gls{RL} algorithms, \ie without the use of \gls{NN},  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter \ref{chap:chap_RL}).
+By approximating the state-action function with its parameters (\ie weights and biases), \gls{DNN} can address this difficulty.
+
+Given the assumed absence of knowledge of the agent about the dynamics of the environment, \ie its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.
+In practice, in a model-free approach, the agent estimates the optimal policy directly from experience and without estimating the dynamics of the environment.
+However, model-free methods suffer from two major drawbacks: their sample inefficiency and their sensitivity with respect to their hyper-parameters (\eg learning rates, exploration constants) \cite{haarnoja2018soft}.
+The former leads to a too expensive computational burden while the second requires meticulous settings to get good results.
+To overcome these two challenges, we needed to chose between an ``on-policy'' or ``off-policy'' algorithm.
+In a nutshell, in on-policy learning, the agent learns the value function or policy based on the data it generates by following its current policy whereas, in off-policy, the agent can learn from data collected by any policy, not just the one it is currently following, which provides greater flexibility and potential for reusing data.
+This makes off-policy algorithms more data efficient and ensuring better exploration by reusing past experiences or even following random exploration \cite{haarnoja2018soft}.
+
 The goal of a \gls{RL} approach is to optimise the mapping between inputs (\ie the observations) and output (\ie the actions), called the policy $\pi\left(\bm{a}_n | \bm{o}_n\right)$.
-To do so, an objective function, $\bm{J}(\pi)$, built on the cumulative rewards collected during each episode.
+To do so, an objective function, $\bm{J}(\pi)$, is built on the cumulative rewards collected during each episode.
 Finally, a back-propagation process updates the weights and biases of the \gls{NN} during the learning of the agent.
 Among the wide variety of \gls{RL} algorithms applied in energy systems \cite{perera2021applications}, this work opted for \gls{SAC} \cite{haarnoja2018soft} to train and update the \gls{NN}.
-This algorithm is model-free and off-policy.
-The former characteristic is necessary as the agent is assumed to have no knowledge about the dynamics of the environment, \ie its transition or reward functions.
-The transition function is, given the current state of the environment and the action taken by the agent, a function that outputs a probability to end up to any of the next states.
-The reward function, on its turn, gives a reward given the current state and the taken action.
-In practice, in this model-free approach, the agent estimates the optimal policy directly from experience and without estimating the dynamics of the environment.
-However, model-free methods suffer from two major drawbacks: their sample inefficiency and their brittleness with respect to their hyper-parameters (\eg learning rates, exploration constants) \cite{haarnoja2018soft}.
-The former leads to a too expensive computational burden while the second requires meticulous settings to get good results.
-\gls{SAC} overcomes these two challenges as an off-policy (\ie efficiently re-using past episodes to update the target policy-network) actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function:
+Like other actor-critic-based algorithms, \gls{SAC} works with two \gls{NN} in parallel: the actor learning the control policy and the critic judging the actor (see Figure \ref{fig:Actor-critic}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.5\textwidth]{Actor-critic.pdf}
+\caption{General concept of actor-critic-based algorithms.
+The two \gls{NN} are trained against each other for the actor to improve the control policy and for the critic to provide a better judgement of the actor's action via the temporal-difference (TD) error.
+Graph adapted from \cite{cao2020reinforcement}.}
+\label{fig:Actor-critic}
+\end{figure}
+
+\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented\footnote{The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Eq.
+\ref{eq:SAC_objective}.} objective function (see Eq.
+\ref{eq:SAC_objective}).
+Where entropy represents the amount of energy in a system not available to produce work in thermodynamics, this term, also called Shannon entropy in the \gls{RL} context, stands for the randomness or stochasticity of the policy.
 
@@
     \label{eq:SAC_objective}
-    \bm{J}(\pi) = \underset{\pi}{\mathbb{E}}\left[\underset{n=0}{\overset{N_{ep}}{\sum}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right) - \zeta \log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right) \right]
+    \bm{J}(\pi) = \underset{\pi}{\mathbb{E}}\left[\underset{n=0}{\overset{N_{ep}}{\sum}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right) - \zeta \log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right) \right],
 \end{equation}
@@
 where $\gamma$ is the discount factor and $\zeta$ the temperature parameter.
-The former determines how much importance we want to give to future rewards within an episode.
-The latter balances the trade-off between exploitation of proven actions via the return maximisation, \ie $\sum_{n=0}^{N_{ep}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right)$, and exploration through the entropy\footnote{Where entropy represents the amount of energy in a system not available to produce work in thermodynamics, this term stands for the randomness or stochasticity of the policy in the current context.} term, \ie $\log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right)$.
-This way, \gls{SAC} ensures sample efficiency and low sensitivity to hyper-parameters while improving exploration \cite{haarnoja2017reinforcement} and robustness \cite{ziebart2010modeling}.
-These make \gls{SAC} a state-of-the-art algorithm and one of the most efficient model-free deep RL method nowadays.
-In this work, the authors used the open-source \gls{SAC} package developed by \textsc{Stable-Baselines3} \cite{raffin2021stable} where the policy \gls{NN} is a fully connected multilayer perceptron (MLP) built with \textsc{TensorFlow} \cite{abadi2016tensorflow}.
+$\gamma$ determines how much importance we want to give to future rewards within an episode.
+$\zeta$ balances the trade-off between exploitation of proven actions via the return maximisation, \ie $\sum_{n=0}^{N_{ep}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right)$, and exploration through the entropy term, \ie $\log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right)$.
+This way, \gls{SAC} ensures sample efficiency while improving exploration \cite{haarnoja2017reinforcement} and robustness \cite{ziebart2010modeling}.
+In their work, \citet{haarnoja2017reinforcement} showed a lower sensitivity of \gls{SAC} to hyper-parameters.
+These make \gls{SAC} a state-of-the-art algorithm and one of the most efficient model-free deep RL method nowadays \cite{haarnoja2017reinforcement}.
+In this thesis, we used the open-source \gls{SAC} package developed by \textsc{Stable-Baselines3} \cite{raffin2021stable} where the policy \gls{NN} is a fully connected multilayer perceptron (MLP) built with \textsc{TensorFlow} \cite{abadi2016tensorflow}.
 For further information on \gls{RL} and the \gls{SAC} algorithm, the interested reader is invited to refer to the works of \citet{sutton2018reinforcement} and \citet{haarnoja2018soft}, respectively.
-
-\textbf{Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity}
 
@@
 To tackle this challenge, we have developed a methodology based on the \acrfull{PCA}.
+The philosophy behind this approach is to identify key-combinations of design variables giving relevant dimensions to compare different systems, beyond their sole objective function, \ie total transition cost, but without having to compare each design variable individually.
 This methodology provides two main outputs.
@@
 Born in the early 20th century \cite{pearson1901on,hotelling1933analysis}, the \acrfull{PCA} finds its fundamentals from the \gls{SVD}.
-The latter is a generalization, to an arbitrary (\ie not especially square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
+\gls{SVD} is a generalization, to an arbitrary (\ie not especially square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
 The core concept of principal component analysis (PCA) involves simplifying a dataset with numerous interconnected variables by reducing its dimensionality.
 The aim is to preserve as much variability within the data as feasible.
-This is accomplished by transforming the data, $\mathbf{x}$, into a new set of variables called principal components (PCs), $\mathbf{z}$.
+This is accomplished by transforming the $p$-dimension data, $\mathbf{x}$, into a new set of variables called principal components (PCs), $\mathbf{z}$.
 These components are uncorrelated and arranged in such a way that the first ones retain the majority of the variability found in all of the original variables.
@@
 
+The PCs are computed based on the covariance matrix of $\mathbf{x}$, $\mathbf{\Sigma}$ where the diagonal of this matrix gives the variance of the $i^{\text{th}}$ variable and the other elements give the covariance between the $i^{\text{th}}$ and the $j^{\text{th}}$ variables where $i\neq j$.
+Out of this matrix, $\mathbf{\alpha_k}$ is the eigenvector of $\mathbf{\Sigma}$ corresponding to its $k^{\text{th}}$ highest eigenvalue $\lambda_k$.
+One crucial aspect of these eigenvectors is their normalization, \ie $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$ \cite{jolliffe2002principal}.
+This normalization has several objectives.
+Among them, this ensures orthogonality of the PCs ensuring that they represent independent directions in the original feature space.
+Then, normalizing the eigenvectors ensures that the magnitude of each eigenvector represents the importance or variance explained by its corresponding principal component.
+This makes it easier to interpret the relative importance of each principal component in explaining the variability of the data.
+Finally, this ensures a fair comparison between the original features.
+Without normalization, variables with larger scales would dominate the principal components, potentially skewing the results and leading to misinterpretation of the principal components.
+In other words, given this normalization, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
+Moreover, this means that $\alpha_{ki}$, \ie the component of $ \mathbf{\alpha_k}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
+This PC captures $\lambda_k$ variance of the original data.
+In other words, a high absolute value of $\alpha_{ki}$ means that $x_i$ has a significant impact in the direction given by the $k^{\text{th}}$ PC \cite{zdybal2022advancing}.
+
 Easier to represent in two dimensions, let us consider a vector $\mathbf{x}$ composed of the variables $x_1$ and $x_2$, $p=2$, and 25 realisations of them (See Figure \ref{fig:PCA_intro} (left)).
-The first \gls{PC}, $z_1$, is a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $x$ with maximum variance:
+These realisations of the original variables allow building the covariance matrix.
+Then, the first \gls{PC}, $z_1$, is a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $\mathbf{x}$ with maximum variance:
 
@@
   \label{eq:PCA_z1}%7
- z_1=\mathbf{\alpha_1^{T}x}=\alpha_{11}x_1+\alpha_{12}x_2+...+\alpha_{1p}x_p=\sum_{j=1}^{p} \alpha_{1j}x_j
+ z_1=\sum_{j=1}^{p} \alpha_{1j}x_j=\mathbf{\alpha_1^{T}x}=\alpha_{11}x_1+\alpha_{12}x_2,
 \end{flalign}
@@
 \noindent
-where $^T$ means the transpose vector.
+where $\mathbf{^T}$ means the transpose vector.
 Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and maximizing the variance.
-These transformations lead to the construction of the PCs (see Figure \ref{fig:PCA_intro} (right)).
+These linear transformations can be seen as projection of the original data on the principal direction, \ie PCs (see Figure \ref{fig:PCA_intro} (right)).
 In more general cases, one can write $z_k=\mathbf{\alpha_k^{T}x}$ as the $k^{\text{th}}$ PC.
 There can be up to $p$ PCs even though, usually, most of the variance of the original data can be captured by $m$ PCs where $m\ll p$.
+The interested reader is invited to refer to the work of \citet{jolliffe2002principal} for further mathematical demonstrations, information and examples.
 
@@
 
-The PCs are computed based on the covariance matrix of $\mathbf{x}$, $\mathbf{\Sigma}$.
-The diagonal of this matrix gives the variance of the $i^{\text{th}}$ variable whereas the other elements give the covariance between the $i^{\text{th}}$ and the $j^{\text{th}}$ variables where $i\neq j$.
-Out of this matrix, $\mathbf{\alpha_k}$ is the eigenvector of $\mathbf{\Sigma}$ corresponding to its $k^{\text{th}}$ highest eigenvalue $\lambda_k$.
-Finally, as the $\mathbf{\alpha_k}$ for all $k=1,2,\cdots, p$ are normalized,  \ie $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
-In other words, $\alpha_{ki}$, \ie the components of$ \mathbf{\alpha_k}$, give the weight of the $i^{\text{th}}$ original variable, $x_i$, in the $k^{\text{th}}$ PC, \ie $z_k$, where $\lambda_k$ gives the variance of the original data captured by $z_k$.
-In other words, a high absolute value of $\alpha_{ki}$ means that $x_i$ has a significant impact in the direction given by the $k^{\text{th}}$ PC \cite{zdybal2022advancing}.
-The interested reader is invited to refer to the work of \citet{jolliffe2002principal} for further mathematical demonstrations, information and examples.
-
 \subsection{Principal components of the transition}
@@
 As introduced, the objective is to define the main technological drivers of the variation through the transition to 2050 subject to uncertainties.
-To do so, before calculating the PCs of the transition, three preliminary steps are necessary: (i) data selection, (ii) data scaling and, (iii)outliers management.
+To do so, before calculating the PCs of the transition, three preliminary steps are necessary: (i) selection of the right data, (ii) data scaling and, (iii) outliers management.
+Like any other dimension-reduction process, \gls{PCA} has to be supplied with relevant data to reach the stated objective.
+Like the normalization of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude.
+Finally, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
 After this preprocessing, PCs can be computed for each of the year of the transition then aggregated to give a bigger picture over the whole transition.\\
@@
 \noindent
-Prior to any data processing, the choice of the original dataset is fundamental.
-In the will to characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Eq.
+To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Eq.
 \ref{eq:F_newBuilt}).
-Even though these represent only a part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Eq.
+Even though these represent only the design part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Eq.
 \ref{eq:c_inv}) and, more indirectly, the resources to use.
@@
 This is given by the runs to quantify the impact of the uncertain parameters on the total cost of transition required by the method described in Section \ref{sec:meth:UQ}.
-Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ states for the sample number of the uncertainty quantification method.
+Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ stands for the sample number of the uncertainty quantification method.
 For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=240$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
@@
 \includegraphics[width=0.8\textwidth]{PCA_data.pdf}
-\caption{Original raw data considered in the \acrfull{PCA} of the variation of the design strategy through the transition, $\textbf{x}(y,j,s)$.}
+\caption{Original raw data considered in the \acrfull{PCA} of the variation of the design strategy through the transition, $\textbf{x}(y,j,s)$, accounting for the $p$ possible technologies to install.}
 \label{fig:PCA_data}
@@
 Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the rest of the dataset for this specific technology and year across all the samples.
-Then, one option to handle these outliers would be to remove the whole data point for which one technology presents an outlying installed capacity.
-However, it would lead to reducing the entire dataset (\ie several thousands of points) to too few data points (\ie several tens) limiting the applicability of the \gls{PCA} approach.
-Consequently, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
-In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively.\\
+There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
+Since all the data points correspond to a result provided by the optimization, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
+In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively.
+In practice, this modification narrows the variation range of features presenting outliers and, consequently, reduces their weight in the different PCs.\\
 
@@
 Choosing the appropriate threshold involves a trade-off.
-Retaining too few principal components may result in loss of important information, while retaining too many may lead to over-fitting or computational inefficiency.
-The good practice followed in this work is to compute, for each of the representative years of the transition (except 2020),  the PCs that capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
+Retaining too few principal components may result in loss of important information, while retaining too many may lead to unclear analysis.
+In this work is to compute, for each of the representative years of the transition (except 2020),  the PCs capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
 At the end of this step, we have a list of $m$ PCs, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and
@@
 \label{eq:PC_year}
-\sum_{i=0}^m\text{var}\left(\text{PC}_{y,i}\right)\geq 90\% \sum_{i=0}^p\text{var}\left(\text{PC}_{y,i}\right) \quad \forall y \in \text{\emph{YEARS}}
+\sum_{i=0}^m\text{var}\left(\text{PC}_{y,i}\right)\geq 90\% \sum_{i=0}^p\text{var}\left(\text{PC}_{y,i}\right) \quad \forall y \in \text{\emph{YEARS}},
 \end{equation}
@@
 Then, starting with the one with the highest absolute variance, all the other $\text{PC}_{y,i}$ similar to it are clustered together.
-The similarity between two PCs is defined according to the cosine similarity approach.
+The similarity between two PCs is defined according to the cosine similarity approach, especially appropriate in high-dimensional positive spaces \cite{xia2015learning}.
 Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest.
@@
 \label{eq:cos_similarity}
-S_C (A,B):= \cos(\theta) = {\mathbf{A} \cdot \mathbf{B} \over \|\mathbf{A}\| \|\mathbf{B}\|} = \frac{ \sum\limits_{i=1}^{n}{A_i  B_i} }{ \sqrt{\sum\limits_{i=1}^{n}{A_i^2}} \cdot \sqrt{\sum\limits_{i=1}^{n}{B_i^2}} } \geq 90\%
+S_C (A,B):= \cos(\theta) = {\mathbf{A} \cdot \mathbf{B} \over \|\mathbf{A}\| \|\mathbf{B}\|} = \frac{ \sum\limits_{i=1}^{n}{A_i  B_i} }{ \sqrt{\sum\limits_{i=1}^{n}{A_i^2}} \cdot \sqrt{\sum\limits_{i=1}^{n}{B_i^2}} } \geq 90\%,
 \end{equation}
@@
 The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps.
-In conclusion, a roadmap would be defined as more robust than another one if the projection of its myopic runs on the different $\text{PC}_{\text{transition}}$ span on a more narrow range.
+In conclusion, a roadmap would be defined as more robust than another one if the projection of its myopic runs on the different $\text{PC}_{\text{transition}}$ spans on a more narrow range (see Figure \ref{fig:Projection_PC_transition}).
 The bounds of this ``range of projection'' are computed as the mean, $\mu$, of the projected data $\pm$ a 95\% confidence level, CL, on the margin of error, MOE:
 $$
-\text{range of projection} = [\mu-\mathrm{MOE}; \mu + \mathrm{MOE}] 
+\text{range of projection} = [\mu-\mathrm{MOE}; \mu + \mathrm{MOE}],
 $$
@@
 $$
-\text{MOE} = \mathrm{SEM}\cdot \mathrm{PPF}_{\mathrm{t}}\left((1+\mathrm{CL})/2,N-1\right)
+\text{MOE} = \mathrm{SEM}\cdot \mathrm{PPF}_{\mathrm{t}}\left((1+\mathrm{CL})/2,N-1\right),
 $$
@@
 where $\mathrm{PPF}_{\mathrm{t}}$ is the percent point function (\ie inverse of cumulative distribution function) of the Student's distribution.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Projection_PC_transition.pdf}
+\caption{Projection on the first PC of the transition, $\text{PC}_{\text{transition},1}$, of the different myopic runs under uncertainties based on different roadmaps.
+Given that the distribution resulting from roadmap 1 spans over a more narrow range of this PC of the transition, we conclude that this roadmap is more robust than roadmap 2 according to the direction of variation described by $\text{PC}_{\text{transition},1}$.}
+\label{fig:Projection_PC_transition}
+\end{figure}

### 2024-03-13 options/symbols_and_cmd.tex
1st section of RL chapter
Edit since UQ_TD_full has been completed
Implementation of FC comments on chapter methodo

@@
 \newabbreviation[group={Acronyms}]{DHN}{DHN}{district heating network}
+\newabbreviation[group={Acronyms}]{DNN}{DNN}{deep neural network}
+\newabbreviation[group={Acronyms}]{DRL}{DRL}{deep reinforcement learning}
 
@@
 \newabbreviation[group={Acronyms}]{FC}{FC}{fuel cell}
+\newabbreviation[group={Acronyms}]{FEC}{FEC}{final energy consumed}
 
@@
 \newabbreviation[group={Acronyms}]{LOO}{LOO}{leave-one-out}
+\newabbreviation[group={Acronyms}]{LP}{LP}{linear programming}
 \newabbreviation[group={Acronyms}]{LPG}{LPG}{liquefied petroleum gas}
@@
 %% U
-\newabbreviation[group={Acronyms}]{UQ}{UQ}{Uncertainty Quantification}
+\newabbreviation[group={Acronyms}]{UQ}{UQ}{uncertainty quantification}
 

### 2024-03-20 chap_methodo/chap_methodo.tex
Figures of Tech_Cap for UQ + First part of RobPol

@@
 
-Easier to represent in two dimensions, let us consider a vector $\mathbf{x}$ composed of the variables $x_1$ and $x_2$, $p=2$, and 25 realisations of them (See Figure \ref{fig:PCA_intro} (left)).
-These realisations of the original variables allow building the covariance matrix.
-Then, the first \gls{PC}, $z_1$, is a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $\mathbf{x}$ with maximum variance:
+Easier to represent in two dimensions, let us consider a vector $\mathbf{x}$ composed of the variables $x_1$ and $x_2$, $p=2$, and 25 realisations of them (See Figure \ref{fig:PCA_intro_original}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.4\textwidth]{PCA_intro_original.pdf}
+\caption{Original observations of a two-dimension dataset, $x_1$ and $x_2$.}
+\label{fig:PCA_intro_original}
+\end{figure}
+
+From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximize their spread, \ie variance (see \ref{fig:PCA_intro_proj}).
+Another way to understand this is to look at the other side of the same coin.
+Indeed, \gls{PCA} also looks for properties that allow reconstructing the original features as accurately as possible.
+In practice, it aims at minimizing the total reconstruction error that is the average squared distance between the original observations and their respective projection.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.4\textwidth]{PCA_intro_proj_1.pdf}
+\hspace{1.5cm}
+\includegraphics[width=0.4\textwidth]{PCA_intro_proj_2.pdf}
+\caption{Two different projections (red dots) of the original observations (grey dots).
+Projection on the left captures a bigger variance than the projection on the right.
+At the same time, the average squared distance between the original observations and their respective projection is smaller for the projection on the left.
+Consequently, the left projection is closer to be the first PC than to the right projection.}
+\label{fig:PCA_intro_proj}
+\end{figure}
+
+More mathematically, the first \gls{PC}, $z_1$, is then a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $\mathbf{x}$ with maximum variance:
 
@@
 Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and maximizing the variance.
-These linear transformations can be seen as projection of the original data on the principal direction, \ie PCs (see Figure \ref{fig:PCA_intro} (right)).
+These linear transformations can be seen as projection of the original data on the principal direction, \ie PCs (see Figure \ref{fig:PCA_intro_PCA}).
 In more general cases, one can write $z_k=\mathbf{\alpha_k^{T}x}$ as the $k^{\text{th}}$ PC.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{PCA_intro_original.pdf}
-\hspace{1.5cm}
 \includegraphics[width=0.4\textwidth]{PCA_intro_PCA.pdf}
-\caption{Original observations of the dataset (left) and their projection with respect to their PCs (right).
-The variation of the realisations is more significant in the direction of $x_2$ than $x_1$.
+\caption{Projection of the original observations with respect to their PCs.
+The variation of the realisations was more significant in the direction of $x_2$ than $x_1$.
 Once projected with respect to their PCs, the variation is even more significant in the direction of $z_1$ than in either of the original variables, as it captures most of their variance.
-Graph adapted from \cite{jolliffe2002principal}.}
-\label{fig:PCA_intro}
+}
+\label{fig:PCA_intro_PCA}
 \end{figure}
@@
 Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ stands for the sample number of the uncertainty quantification method.
-For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=240$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
+For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
 Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}).
 In other words, we focus here only on the years 2025, 2030, 2035, 2040, 2045 and 2050.
-This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}).\\
+This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}).
+Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.\\
 
@@
 Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest.
-Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is greater or equal to 90\%:
+Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Eq.
+\ref{eq:cos_similarity}).
+It is important to consider to second specific case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors.
+For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$.
+The second vector being the opposite of the first, they are considered as equivalent in the regard of \gls{PCA}.
 
@@
 \label{eq:cos_similarity}
-S_C (A,B):= \cos(\theta) = {\mathbf{A} \cdot \mathbf{B} \over \|\mathbf{A}\| \|\mathbf{B}\|} = \frac{ \sum\limits_{i=1}^{n}{A_i  B_i} }{ \sqrt{\sum\limits_{i=1}^{n}{A_i^2}} \cdot \sqrt{\sum\limits_{i=1}^{n}{B_i^2}} } \geq 90\%,
+S_C (A,B):= \cos(\theta) = {\mathbf{A} \cdot \mathbf{B} \over \|\mathbf{A}\| \|\mathbf{B}\|} = \frac{ \sum\limits_{i=1}^{n}{A_i  B_i} }{ \sqrt{\sum\limits_{i=1}^{n}{A_i^2}} \cdot \sqrt{\sum\limits_{i=1}^{n}{B_i^2}} } \geq 90\%\text{ or } \leq -90\%,
 \end{equation}
@@
 Then, the process repeats with the $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$.
-This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to 90\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$ generated at the previous step:
+This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to $\alpha$\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$, \ie the ``total transition variance'', generated at the previous step:
 
@@
 \label{eq:PC_transition}
-\sum_{i=0}^N\text{var}\left(\text{PC}_{y,i}\right)\geq 90\% \sum_{i=0}^M\text{var}\left(\text{PC}_{y,i}\right)
-\end{equation}
+\sum_{i=0}^N\text{var}\left(\text{PC}_{y,i}\right)\geq \alpha\% \sum_{i=0}^M\text{var}\left(\text{PC}_{y,i}\right)
+\end{equation}
+
+In our case, we have decided to set the value of $\alpha$ to 80\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{y}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).
+For instance, capturing up to 90\% of the the total transition variance would mean almost doubling the number of $\text{PC}_{y}$ to consider and, consequently, the number of directions of variation to assess.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.5\textwidth]{PC_y_share_tot_var.pdf}
+\caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{y}$.
+Passing a certain threshold, here 80\%, the number of $\text{PC}_{y}$ to consider increase significantly compared to the share of transition variance that they capture.}
+\label{fig:PC_y_share_tot_var}
+\end{figure}
 

### 2024-03-20 chap_RL/chap_RL.tex
Implementation of FC comments on chapter methodo
Figures of Tech_Cap for UQ + First part of RobPol

@@
 At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
-Moreover, a sample of values is drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
+Moreover, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
 This sample will remain valid for the following time windows.
@@
 This goes on until the end of the episode.
-Similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
 
@@
 \label{sec:RL:testing}
+
+How does it react to potential shocks (e.g. no fuels from 2040) compared to classic myopic?
+
+\citet{herman2014beyond} defined robustness as ``… the fraction of sampled states of the world in which a solution satisfies all performance requirements''
 
@@
 However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).
+
+\textbf{Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity}

### 2024-03-20 chap_RobPol/chap_RobPol.tex
Figures of Tech_Cap for UQ + First part of RobPol

--- 2024-02-12 chap_RobPol/chap_RobPol.tex
@@
 %!TEX encoding = UTF-8 Unicode
+\vspace{-0.2cm}
+\begin{flushright}
+\emph{``The more data we have, the more likely we are to drown in it.''}\\
+Nassim Nicholas Taleb, in \textit{Fooled by Randomness: The Hidden Role of Chance in Life and in the Markets}, 2008
+\end{flushright}
+\vspace{0.4cm}
 
-The purpose of this chapter is to display results regarding PCA as presented orally to you in December 2023
+Assessing the robustness of a roadmap driving the transition pathway of a whole-energy system is complex, especially due to the curse of dimensionality.
+This curse comes from the number of variables of the system (\eg the installed capacity of technologies), the multiple-year approach specific to the pathway optimisation (\ie versus the snapshot approach) or the number of uncertain parameters.
+On top of this, the sector coupling interconnecting the installed capacities and the used resources among the different (non-)energy sectors can make harder the understanding of big trends of such a system.
+To navigate through this load of uncertain and interconnected data, it is necessary to assess the robustness of pathway roadmaps.
+
+To deal with such uncertainties, decision-makers have several options: (i) resistance; (ii) resilience; (iii) static robustness; and (iv) adaptive robustness \cite{walker2012deep}.
+Where resistance consists in planning for the worst-case scenario, resilience aims at a fast recovery whatever the conditions in the future.
+Finally, in static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions.
+Where the adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}, the objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.
+\citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}.
+In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economy model, like EnergyScope, that usually assume rational choice within an overall cost minimization.
+However, in the case of the transition pathway of a whole-eenrgy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''.
+Between the sole total transition cost and the entire set of installed technologies that give too few and too much information, respectively, the performance metric here is defined through the \gls{PCA} approach.
+Then, when comes the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps.
+In other terms, one roadmap will not be robust or not in itself but rather more or less robust than another one.
+
+\section*{Contributions}
+\label{sec:RobPol:contributions}
+The main contributions of this chapter is the application of the methodology proposed in Section \ref{sec:meth:PCA} to the case study of the Belgian energy transition.
+First, we develop the different steps that lead to the principal components of the transition.
+We analyse these big trends of variation and highlight the fact that these variations stand for the entire pathway, a group of consecutive representative years or rather on a tipping-year.
+Then, and most importantly, we assess the robustness of different technological roadmaps by projecting their resulting myopic pathway against these directions of variation.
+The application of \gls{PCA} to provide a new metric for robustness applied to the case of Belgium is the added-value of this chapter.
+
+\section{Definition of the principal components of the transition}
+\label{sec:RobPol:PC_transition}
+As detailed in Section \ref{subsec:meth:PCA:transition}, we have decided to define the directions of variation, \ie the robustness metrics, based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol.
+These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.
+As introduced in Section \ref{subsec:meth:PCA:transition}, the data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}).
+This gave 1260 different transitions resulting, for each of them, from the pathway optimisation subject to a sample of uncertain parameters (see Section \ref{subsec:uncert_charac}).
+Appendix \ref{app:UQ_tech_cap} gives the exhaustive distributions of the installed capacities among the different end-use sectors from the \gls{GSA}.
+
+\subsection{Principal components of each representative year}
+\label{subsec:RobPol:PC_year}
+After the pre-preprocessing of the raw data (\ie data scaling and outliers management, see Section \ref{subsec:meth:PCA:transition}), the \gls{PCs} of each representative year of the transition, except 2020 as the initialisation year, can be computed.

### 2024-03-21 chap_RobPol/chap_RobPol.tex
PC_y of RobPol chapter

@@
 After the pre-preprocessing of the raw data (\ie data scaling and outliers management, see Section \ref{subsec:meth:PCA:transition}), the \gls{PCs} of each representative year of the transition, except 2020 as the initialisation year, can be computed.
+
+First of all, before investigating the \gls{PCs}, it is worth looking at the total variance of each representative year (see Table \ref{tab:design_variance}).
+Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition.
+In other words, the further goes the transition, the more limited are the degrees of freedom to respect the \ce{CO2}-budget.
+
+\begin{table}[htbp]
+\caption{Whole-system design variance of the different representative years and their comparison with 2025.} 
+\label{tab:design_variance}
+\centering
+\begin{tabular}{c c c}
+\toprule
+\textbf{Year}      & \textbf{Design variance} [$10^{-3}$]	 &	\textbf{vs. 2025}\\
+\midrule
+2025 	&	$10.4$	& - 	\\
+2030 	&	$12.1$	& +15\%	\\
+2035 	&	$9.7$	& -7\% 	\\
+2040 	&	$6.1$	& -42\% 	\\
+2045 	&	$5.1$	& -51\% 	\\
+2050 	&	$4.8$	& -54\% 	\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Then, keeping the \gls{PCs} capturing at least 90\% of the total variance of each representative year, this gives between four, in 2035, and seven, in 2050, \gls{PCs} depending on the year (see Figure \ref{fig:PC_over_the_years}), and a total of 34 \gls{PCs}.
+At later stages of the transition, the increasing number of required \gls{PCs}, in line with their smaller share of explained variance, is another indication that the variance of the system design is more spread over a wider range of technologies and with a more limited amplitude.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{PC_over_the_years.pdf}
+\caption{\gls{PCs} capturing at least 90\% of the total variance of their respective representative year of the transition. }
+\label{fig:PC_over_the_years}
+\end{figure}
+
+Finally, we consider the respective contribution of the different technologies in the different $\text{PC}_{y}$, \ie their corresponding component in the different eigenvectors.
+Highlighting the top-5 technologies for $\text{PC}_{y,1}$, $\text{PC}_{y,2}$ and $\text{PC}_{y,3}$, we observe general trends over the whole transition as well as tipping year where there is a clear trade-off between several technologies (see Figure \ref{fig:Top5_PC_year}).
+As pointed out in Section \ref{subsec:meth:PCA:transition}, \gls{PCA} does not make any distinction between a vector of variation and its opposite.
+This is why $\text{PC}_{2025,1}$ and $\text{PC}_{2035,1}$ are actually very similar even though there are mostly on the opposite sides of the 0-axis.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Top5_PC1.pdf}
+\includegraphics[width=0.8\textwidth]{Top5_PC2.pdf}
+\includegraphics[width=0.8\textwidth]{Top5_PC3.pdf}
+\caption{Contribution of the top-5 technologies to the first three \gls{PCs} over the different representative years.
+Where the variation of \gls{PV} panels supported by the electrification of the \gls{HT} heat via industrial electric resistors is the main varying drivers over the whole transition, there are tipping years like 2025 for the production of \gls{HVC} (\ie naphtha-cracker to \gls{MTO} or 2025-2030 for the production of \gls{LT} heat (\ie from oil and gas boilers to electric \gls{HP}.}
+\label{fig:Top5_PC_year}
+\end{figure}
+
+Even though the following observations could be made by analysing the distribution of installed capacities (see Appendix \ref{app:UQ_tech_cap}) or the covariance matrices, the \gls{PCA} decomposition offers a more visual and summarising representation of the main trends of variation.
+Due to their intermittency, integrating more \gls{PV} panels leads to the installation of other technologies to benefit from free and renewable electricity when it exceeds the electrical \gls{EUD}.
+Therefore, we observe that the variation of installed \gls{PV} is directly linked with the variation of installed industrial electrical resistors and, to a smaller extent, of decentralised and \gls{DHN} electrical \gls{HP}.
+These variations cover the whole transition as the main varying factors given their significant contributions to the $\text{PC}_{1}$.
+Where this was an example for correlated technologies, we can also identify some key modal shifts where one technology is either substituted by others or in balance with another one.
+First, about the \gls{LT} heat sector, the early stages of the transition, \ie 2025-2030, sees the shift from mainly decentralised oil and gas boilers towards decentralised and \gls{DHN} electrical \gls{HP}.
+Later in the transition, there seems to be a tight competition between (bio)diesel and \gls{FC} trucks that drive the design variance to a smaller extent as they mostly appear in the second and third \gls{PCs} of the representative years.
+Not capturing a significant share of variance, there are other modal shifts (\eg \gls{BEV} substituting diesel and gasoline cars) that are not visible through the \gls{PCs}.
+Besides these modal shifts spread over several representative years, 2025 is the tipping year concerning the shift from naphtha-cracker and methanol-to-olefins to supply \gls{HVC}.
+Finally, there are also technologies contributing to \gls{PCs} because they are the main producing assets of their respective end-use sector and the demand varies significantly, \eg biomass-to-methanol.
+
+\subsection{Principal components of the transition}
+\label{subsec:RobPol:PC_transition}
+Based on the \gls{PCs} of each representative year, $\text{PC}_{y}$, the \gls{PCs} of the transition, \ie the metrics to assess the robustness of roadmaps, can be computed.
+Before aggregating and averaging similar $\text{PC}_{y}$, it is necessary to rank them to ensure capturing most of the transition variance in the subsequent $\text{PC}_{\text{transition}}$.
+
+Based on the design variance captured by each $\text{PC}_{y}$ in their respective representative year, we can rank them (see Table \ref{tab:ranking_PCs}).
+Summing all of these variances over the different years results in a ``pseudo'' total variance of the transition.
+To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 80\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).
+This results in keeping 11 $\text{PC}_{y}$: the first PC of each representative year, the second of 2025, 2030, 2035 and 2040 and the third PC of 2025.
+
+\begin{table}[htbp!]
+\caption{Ranking of \gls{PCs} per design variance captured in their respective representative year and cumulative share of the captured total variance of the transition}
+\label{tab:ranking_PCs}
+\centering
+\begin{tabular}{c c c c c}
+\toprule
+\multirow{2}{*}{\textbf{Ranking}} & \multirow{2}{*}{\textbf{Year}}  & \multirow{2}{*}{\textbf{PC}} & \multirow{2}{*}{\textbf{Design variance} [$10^{-4}$]} & \textbf{Cumulative share of} \\	
+ &   &  &  & \textbf{total variance} [\%] \\	
+ \midrule	
+1 & 2030 & $\text{PC}_1$ & 61.1 & 13.9 \\
+2 & 2025 & $\text{PC}_1$ & 55.7 & 26.5 \\
+3 & 2035 & $\text{PC}_1$ & 52.7 & 38.4 \\
+4 & 2030 & $\text{PC}_2$ & 38.4 & 47.2 \\
+5 & 2040 & $\text{PC}_1$ & 33.4 & 54.7 \\
+6 & 2045 & $\text{PC}_1$ & 26.5 & 60.7 \\
+7 & 2050 & $\text{PC}_1$ & 22.2 & 65.8 \\
+8 & 2035 & $\text{PC}_2$ & 20.9 & 70.5 \\
+9 & 2025 & $\text{PC}_2$ & 19.1 & 74.8 \\
+10 & 2025 & $\text{PC}_3$ & 15.0 & 78.2 \\
+11 & 2040 & $\text{PC}_2$ & 8.7 & \textbf{80.2} \\
+\midrule
+$\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$\\
+34 & 2050 & $\text{PC}_7$ & 1.6 & 100 \\
+\bottomrule
+
+\end{tabular}
+\end{table}
+
+Given the similarity between $\text{PC}_y$, some $\text{PC}_{\text{transition}}$ result from the aggregation and averaging of the components of several $\text{PC}_y$.

### 2024-03-27 appendices/app_case_study.tex
Integration of HJ comments on appendices + RobPol start part 2

@@
 \label{app:bel_2020}
-The Belgian whole-energy system of 2020 was largely based (88.6\% of the primary energy mix) on \og conventional fuels\fg (\ie oil and oil products (38.2\%), natural gas (29.5\%), uranium (16.3\%) and solid fossil fuels (4.6\%) while the rest mainly accounts for 26.7\,TWh of lignocellulosic and wet biomass, 12.8\,TWh of wind and 5.1\,TWh of solar \cite{spf_economy_2022}.
+The Belgian whole-energy system of 2020 was largely based (88.6\% of the primary energy mix) on ``conventional fuels'' (\ie oil and oil products (38.2\%), natural gas (29.5\%), uranium (16.3\%) and solid fossil fuels (4.6\%) while the rest mainly accounts for 26.7\,TWh of lignocellulosic and wet biomass, 12.8\,TWh of wind and 5.1\,TWh of solar \cite{spf_economy_2022}.
 Given the data available in the literature (mostly for the power sector) and, when not available, following the assumptions made by \citet{Limpens2020}, \Cref{tab:Belgium_2020} gives the major technologies used in 2020 to supply the different demands of \Cref{fig:cs_demands}.
@@
 \multirow{2}{*}{Private mobility\footnote{\label{foot:privatemob_80}The private mobility accounts for 80\% of the passengers mobility.}}
- & Diesel car & 49\% & 93.5\,Mpass.-km/h\\
+ & Diesel car & 48\% & 93.5\,Mpass.-km/h\\
  & Gasoline car & 49\% & 94.7\,Mpass.-km/h\\
@@
 \label{app:bel_PF_TD}
-This section presents the results of the deterministic (\ie all parameters at their respective nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon-neutrality in 2050.
+This section presents the results of the deterministic (\ie all parameters at their nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon-neutrality in 2050.
 After performing a technical investigation of the pathway by checking the greenhouse gas breakdown by energy sectors, the primary energy mix is analysed.
-To illustrate the sector coupling, a focus is made on the electrification of other sectors.
+To illustrate the sector coupling, a focus is made on the electrification of the sectors.
 Then, the cost implications in terms of investments and operations are discussed.
@@
 \includegraphics[width=\textwidth]{Gwp_per_sector.pdf}
-\caption{Energy sectors have different speed to reduce \gls{GHG} emissions over the transition.
+\caption{Energy sectors have different rate to reduce \gls{GHG} emissions over the transition.
 The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon-neutrality in 2050.}
@@
 Figure \ref{fig:pestd_primary_energy} shows the primary energy mix for the different representative years.
-The pathway verifies five trends: (i) reduction of primary energy thanks to energy efficiency; (ii) massive integration of endogenous renewable energies; (iii) importance of electrification; (iv) the usage of gas as the last fossil resource; and (v) the obligation to rely on renewable fuels to achieve carbon neutrality.
+The pathway verifies five trends: (i) reduction of primary energy thanks to energy efficiency; (ii) massive integration of endogenous renewable energies; (iii) importance of electrification; (iv) the use of gas as the last fossil resource; and (v) the obligation to rely on renewable fuels to achieve carbon neutrality.
 
@@
 \includegraphics[width=\textwidth]{Resources.pdf}
-\caption{Primary energy emitting \gls{GHG} (below Uranium) are reducing linearly with fossil gas remaining until 2045.
+\caption{Primary energy emitting \gls{GHG} (below Uranium) are decreasing linearly with fossil gas remaining until 2045.
 A part of this energy is replaced by renewable ones and starting from 2040, a significant share of electrofuels.
@@
          \caption{The electricity supply (positive values) will remain a mix of different technologies where backup is first mainly provided by gas-\gls{CCGT} and then imported electricity, heat and power cogeneration and later ammonia-\gls{CCGT}.
-The electricity demand (negative values) is led by the electricity end-use demand, but the share used to electrify heat (heat pumps), vehicles (cars, trains, trams, ...) and industrial heaters drastically increase.
+The electricity demand (negative values) is led by the electricity end-use demand, but the share used to electrify heat (heat pumps), vehicles (cars, trains, trams...) and industrial heaters drastically increases.
 This enables a flexible demand that can facilitate the integration of intermittent renewables.}
@@
 
-In the following paragraphs, the results are analysed from an economic perspective to decipher the choices made by the model, as the overall cost of the transition is 1 004~b€\textsubscript{2015} split unequally among the sectors.
+In the following paragraphs, the results are analysed from a cost perspective to decipher the choices made by the model, as the overall cost of the transition is 1004~b€\textsubscript{2015} split unequally among the sectors.
 
@@
 \includegraphics[width=\textwidth]{Investment_transition.pdf}
-\caption{The cumulative investments over the transition is unequally spread between the sectors.
-The energy system in 2020 is imposed to the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
+\caption{The cumulative investments over the transition are unequally spread between the sectors.
+The energy system in 2020 is imposed as the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
 The investments required during the transition represents 150\% the initial investment and mainly in the same three sectors.
@@
 
-A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technology after 2050.
+A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technologies after 2050.
 Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Eq.
@@
 \includegraphics[width=\textwidth]{Investment_return.pdf}
-\caption{By the end of the transition (\ie in 2050), the ratio between the salvage value and its cumulative investment, per sector, is unequal.
-Investments in infrastructures, public mobility, storage and other long-lifetime technologies experience an important salvage value, at the contrary, investments in private mobility will not be recovered as vehicles have a short lifetime.
+\caption{By the end of the transition (\ie in 2050), the ratio between the salvage value in a sector and its cumulative investment is unequal.
+Investments in infrastructures, public mobility, storage and other long-lifetime technologies experience an important salvage value on the contrary, investments in private mobility will not be recovered as vehicles have a short lifetime.
 All together, these salvage values represent 160.1 b€\textsubscript{2015}, 25\% of the cumulative investment costs in 2050.}
@@
 The annualised cost of the energy system in 2020 is estimated to 44.3~b€/y and increases by 5.5~b€/y to reach 49.8~b€/y by 2050.
-The work of \citet{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y, while the other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
+The work of \citet{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y, while other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
 The differences come from the scope of the energy system: as an example \citet{My2050} also account for the agriculture sector.
@@
 \begin{table}[htbp]
-\caption{Exhaustive general comparison between the two different foresight approaches: Perfect foresight (PF) and Myopic (MY).
+\caption{Comparison between the two different foresight approaches: Perfect foresight (PF) and Myopic (MY).
 Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
@@
 opex+capex-salvage value) difference between the myopic and perfect foresight (PF) approaches.
-Positive values mean that the myopic approach is higher than perfect foresight.
+Positive values mean that the myopic approach is more expensive than perfect foresight.
 Early savings of the myopic vision are overcompensated by late investments further on.}
@@
 
-In 2050, the capacities installed in the different sectors, the design of the system in other words, are very similar between the two approaches.
-The passenger and freight mobility sectors are the same where differences smaller than 1GW are observed in other sectors.
+In 2050, the capacities installed in the different sectors, in other words the design of the system, are very similar between the two approaches.
+The passenger and freight mobility sectors are the same and differences smaller than 1GW are observed in other sectors.
 More interestingly, the myopic optimisation tends to postpone the decommissioning of capacities when the loss of salvage values at the end of an optimisation window would be bigger than the maintenance cost.
 For instance, in the myopic approach, 0.8GW of industrial coal boilers will remain installed in 2045 and 2050 or 3.6GW of naphtha-crackers to produce HVC in 2040, whereas these technologies are not used.
-This is comparable to the ``lock-ins" detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
+This could be compare to the ``lock-ins" detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
+However, in our case, keeping in place unused technologies is due to the formulation of the salvage value.
+Indeed, to maximise the total transition cost, the model prefers recovering part of the salvage value of these assets rather than prematurely decommissioning them.
 
-Highlighted in Figure \ref{fig:my_pestd_res_cat_diff}, the earlier availability of renewable (and intermittent) electricity consequently accelerates the electrification of the other sectors.
-For instance, in 2035, 3.7GW (+75\%) more of industrial electric heaters to produce 5.1TWh/year (+130\%) of additional industrial heat.
+Highlighted in Figure \ref{fig:my_pestd_res_cat_diff}, the earlier availability of renewable (and intermittent) electricity accelerates the electrification of the other sectors.
+For instance, in 2035, 3.7GW (+75\%) more of industrial electric heaters produce 5.1TWh/year (+130\%) of additional industrial heat.
 In the low-temperature heat sector, decentralised and centralised electric heat pumps capacities are, respectively, 2.2GW (+19\%) and 1GW (+8\%) higher for each of the representative years between 2030 and 2045, to produce, around 7.8TWh/year (+23\%) and 0.8TWh/year (+1\%), at the expense of other technologies such as gas heat pumps.
@@
 Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}.
-In parallel, in the near term, the system relies on average more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.
+In parallel, in the near term, the system relies on average slightly more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.
 
@@
 \includegraphics[width=\textwidth]{MY_investment_return_area.pdf}
-\caption{Sum of the respective salvage value of each sector, per end year of each time window.
+\caption{Sum of the respective salvage values of each sector, per end year of each time window.
 All together, these salvage values represent 223.9~b€\textsubscript{2015}, 53\% of the cumulative investment costs in 2030, and 157.7~b€\textsubscript{2015}, 24\% of the cumulative investment costs in 2050.}
@@
 \label{app:CO2_trajectories}
-This section compare results of the optimisation subject to different emissions-trajectories with the REF case where the emissions are constrained to decrease linearly from the level in 2020 to carbon-neutrality in 2050.
+This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon-neutrality in 2050.
 The first comparison is done with the trajectory subject to respect the \ce{CO2}-budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}).
-Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, alleviating the constraint on the emissions-trajectory.
+Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions-trajectory.
 
@@
 
-\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed for each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead.
+\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead.
 Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050.
 The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat.
-In the former, this is linked to the extended use of oil products through naphtha-cracking.
-The latter is produced by industrial coal boilers for longer, until 2040.
+In the former, this is linked to the earlier phase-out of oil products through naphtha-cracking.
+In the latter, it is the use of coal in industrial coal boilers that is phased-out.
 Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{GWP_per_sector_REF.pdf}
-\includegraphics[width=0.49\textwidth]{GWP_per_sector_lin.pdf}
+\includegraphics[width=\textwidth]{GWP_per_sector_lin.pdf}
+\includegraphics[width=\textwidth]{GWP_per_sector_REF.pdf}
 \caption{Respecting the \ce{CO2}-budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
@@
 The outcomes of a model could be limited when the case study is too restrictive.
-Indeed, ones could argue that the comparison between myopic, monthly and perfect foresight are very similar as the energy system is strongly constrained in terms of \gls{GHG} emissions.
+Indeed, one could argue that the comparison between myopic, monthly and perfect foresight are very similar as the energy system is strongly constrained in terms of \gls{GHG} emissions.
 
-In the following, we perform a similar comparison in the transition pathway without restricting the \gls{GHG} emissions. \\
+In the following, we perform a similar comparison of the transition pathway without restricting the \gls{GHG} emissions. \\
 
@@
 Similar trends that for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
-However, some changes reflects the cheapest option that the system could utilise to reach a cheaper transition, such as using as little electrofuels as possible.
+However, some changes reflects the cheapest option that the system could utilise such as using as little electrofuels as possible.
 In this case, only e-ammonia is used for its end use demand.
@@
 Several key messages of the comparison have been summarised in Table \ref{tab:PF_MY}.
-In the following paragraph we analyse how these conclusions are affected when the constraint on the \gls{GHG}-emissions trajectory is removed.
+In the following paragraphs, we analyse how these conclusions are affected when the constraint on the \gls{GHG}-emissions trajectory is removed.
 
-First, considering the overall transition cost, the myopic approach keeps on making short-term savings, \ie down to -0.2\%, before ending up with a more expensive transition by 2050, \ie +0.2\%.
+First, considering the overall transition cost, the myopic approach keeps on making limited short-term savings, \ie down to -0.2\%, before ending up with a slightly more expensive transition by 2050, \ie +0.2\%.
 Similarly to the case with an imposed \gls{GHG}-emissions trajectory, myopic optimisation invests more, compared to the perfect foresight, at early stages into \gls{VRES} technologies to benefit from the significant salvage values of the related grid infrastructures.
@@
 
-Then, in the case with a prescribed \gls{GHG}-emissions trajectory, the myopic approach had to invest more by the end of the transition to renew \gls{PV} installed more massively at early stages and that reached the end of their lifetime before the end of the transition.
-In the case without this emissions trajectory, there is less an urgency/need for integrating renewables in the system.
+Then, in the case with a prescribed \gls{GHG}-emissions trajectory, the myopic approach invests more by the end of the transition to renew \gls{PV} installed more massively at early stages and that reached the end of their lifetime before the end of the transition.
+In the case without this emissions trajectory, there is less urgency/need for integrating renewables in the system.
 Consequently, in the latter case, there is not such an extra-investment to make to renew too old renewable assets.
@@
 
-\section{Uncertainty characterisation for the 5-year steps transition} 
-\label{app:UC_full}
-Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.
+%\section{Uncertainty characterisation for the 5-year steps transition} 
+%\label{app:UC_full}
+%Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.
 Refer to \cite{Moret2017, Moret2017PhDThesis} for the methodology and sources.
@@
 
-\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits.
+%\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits.
 Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}.
@@
 
-This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
+%This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
 For the specific case of \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ will influence the maximum capacity (\ie 6~GW) to install to translate somehow the readiness of this technology.
@@
 $^{(c)}$ This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.
-
-\begin{table}[htbp!]
-\caption{Application of the uncertainty characterization method to the EnergyScope Pathway model for the year 2025. }
-\label{tab:UC_full}
-\begin{minipage}{\linewidth}
-\centering
-\resizebox{\textwidth}{!}{
-\begin{tabular}{l l l c c c}
-\toprule
-\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range_app}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
+%
+%\begin{table}[htbp!]
+%\caption{Application of the uncertainty characterization method of \citet{Moret2017PhDThesis} to the EnergyScope Pathway model for the year 2025. }
+%\label{tab:UC_full}
+%\begin{minipage}{\linewidth}
+%\centering
+%\resizebox{\textwidth}{!}{
+%\begin{tabular}{l l l c c c}
+%\toprule
+%\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range_app}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
 }}  & \multicolumn{2}{c}{\textbf{Relative variation}\footnote{\label{foot:nom_val_uncert_app}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
 }}\\
-    & & & &	 min 	&	 max \\ 	
-\midrule		
-\multirow{4}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
-& $c_{\mathrm{op,elec}}$ & Purchase electricity & II & -64.3\% & 179.8\% \\
-& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
-& $c_{\mathrm{op,biofuels}}$ & Purchase biofuels & II & -64.3\% & 179.8\% \\
-\midrule
-\multirow{9}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
-& $c_{\mathrm{inv,bus}}$ & CAPEX bus & I & -21.6\% & 25.0\% \\
-& $c_{\mathrm{inv,ic\_prop}}$ & CAPEX ICE & I & -21.6\% & 25.0\% \\
-& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
-& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
-& $c_{\mathrm{inv,efficiency}}$ & CAPEX efficiency measures & I & -39.3\%  & 39.3\% \\
-& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
-& $c_{\mathrm{inv,grid}}$ & CAPEX power grid & I & -39.3\% & 39.3\% \\
-& $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
-& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR_app}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
-\midrule
-\textbf{Maintenance cost} & $c_{\mathrm{maint,var}}$ & Variable OPEX of technologies & I & -48.2\% & 35.7\% \\
-\midrule
-\multirow{2}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
-& $\eta_{\mathrm{fc\_prop}}$ & Consumption fuel cell vehicles & I & -28.7\% & 28.7\% \\
-\midrule
-\multirow{3}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
-& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
-& $f_{\mathrm{max,windoff}}$ & Max capacity offshore wind & I & -24.1\% & 24.1\% \\
-\midrule
-\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
-& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
-\midrule
-\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
-& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
-\midrule
-
-\multirow{4}{*}{\textbf{End-use demand}} &$HH\_EUD$ & Households EUD & III & -13.8\% & 11.2\% \\
-& $services\_EUD$ & Services EUD & III & -14.3\% & 11\% \\
-& $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
-& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
-\midrule
-
-\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
-& $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\
-& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
-& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
-& $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
-& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
-
-\bottomrule
-
-\end{tabular}}
-\end{minipage}
-\end{table}
+%    & & & &	 min 	&	 max \\ 	
+%\midrule		
+%\multirow{4}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+%& $c_{\mathrm{op,elec}}$ & Purchase electricity & II & -64.3\% & 179.8\% \\
+%& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
+%& $c_{\mathrm{op,biofuels}}$ & Purchase biofuels & II & -64.3\% & 179.8\% \\
+%\midrule
+%\multirow{9}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+%& $c_{\mathrm{inv,bus}}$ & CAPEX bus & I & -21.6\% & 25.0\% \\
+%& $c_{\mathrm{inv,ic\_prop}}$ & CAPEX ICE & I & -21.6\% & 25.0\% \\
+%& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,efficiency}}$ & CAPEX efficiency measures & I & -39.3\%  & 39.3\% \\
+%& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,grid}}$ & CAPEX power grid & I & -39.3\% & 39.3\% \\
+%& $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
+%& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR_app}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
+%\midrule
+%\textbf{Maintenance cost} & $c_{\mathrm{maint,var}}$ & Variable OPEX of technologies & I & -48.2\% & 35.7\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+%& $\eta_{\mathrm{fc\_prop}}$ & Consumption fuel cell vehicles & I & -28.7\% & 28.7\% \\
+%\midrule
+%\multirow{3}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
+%& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
+%& $f_{\mathrm{max,windoff}}$ & Max capacity offshore wind & I & -24.1\% & 24.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
+%& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
+%& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
+%\midrule
+%
+%\multirow{4}{*}{\textbf{End-use demand}} &$HH\_EUD$ & Households EUD & III & -13.8\% & 11.2\% \\
+%& $services\_EUD$ & Services EUD & III & -14.3\% & 11\% \\
+%& $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
+%& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
+%\midrule
+%
+%\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+%& $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\
+%& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
+%& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
+%& $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
+%& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
+%
+%\bottomrule							
+%
+%\end{tabular}}
+%\end{minipage}
+%\end{table}

### 2024-03-27 appendices/app_results_UQ.tex
Figures of Tech_Cap for UQ + First part of RobPol
Integration of HJ comments on appendices + RobPol start part 2

@@
 \begin{table}[htbp!]
-\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the monthly and hourly pathway models.
-The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties over the hourly model.}
+\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the hourly and monthly pathway models.
+The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties.}
 \label{tab:UQ_full}
+\begin{minipage}{\textwidth}
 \centering
+\resizebox{0.8\textwidth}{!}{
 \begin{tabular}{l c c|c c}
@@
 
-\end{tabular}
+\end{tabular}}
+\end{minipage}
 \end{table}
@@
 \label{tab:uq_ref_smr_med}
-\begin{minipage}{\linewidth}
+\begin{minipage}{\textwidth}
 \centering
+\resizebox{0.8\textwidth}{!}{
 \begin{tabular}{l l | c c c c}
@@
 \bottomrule							
-\end{tabular}
+\end{tabular}}
 \end{minipage}
@@
 
-\Cref{fig:results_uq_prod_cons} gives the distribution of the different routes of supply and consumption of gas like methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.
+Figures \ref{fig:results_uq_prod_cons_Gas}, \ref{fig:results_uq_prod_cons_H2}, \ref{fig:results_uq_prod_cons_Ammonia} and \ref{fig:results_uq_prod_cons_Methanol} give the distribution of the different routes of supply and consumption methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.
 
 Given its lower cost of purchasing than its renewable equivalent (\Cref{fig:cs_resources_cost}) and lower \gls{GWP} than other fossil fuels (\ie $\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh versus $\mathit{gwp}_{\mathrm{op,LFO}}=0.31$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), fossil \gls{NG} remains the main source of gas in the system until 2040.
-Besides bio-hydrolisis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2}-budget for the transition.
-Its versatility makes gas used by a wide variety of technologies in the different sectors.
+Besides bio-hydrolysis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2}-budget for the transition.
+As a versatile energy carrier, gas is used by a wide variety of technologies in the different sectors.
 Initially, in 2020, decentralised gas boilers, \gls{CCGT} and industrial gas boilers represent the biggest consumers of gas with 39\%, 21\% and 16\% of the total consumption, respectively.
@@
 In terms of consumption, \gls{FC}-trucks are the more consistent player.
-\gls{FC}-cars are also at stake but in specific cases where their CAPEX and the CAPEX of electric vehicles are in the bottom and the top of their respective uncertainty range.
+\gls{FC}-cars are also at stake but in specific cases where their CAPEX and the CAPEX of electric vehicles are in the bottom and the top of their respective uncertainty ranges.
 
 Becoming cheaper than its fossil equivalent at early stages of the transition (\ie from 2030 onward), e-ammonia is the exclusive stream of ammonia in the system, except rare cases.
-Then, on top of its consistent \gls{NED}, the largest consumption of ammonia is \gls{CCGT} as flexible power generation units, to substitute their e-methane equivalents that have a higher \gls{LCOE} (\Cref{fig:LCOE}).
+Then, on top of its consistent \gls{NED}, the largest consumption of ammonia is \gls{CCGT} as flexible power generation units, to substitute their e-methane equivalents that have a higher \gls{LCOE} (see \Cref{fig:LCOE}).
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{UQ_Gas_Prod.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_Gas_Cons.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_H2_Prod.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_H2_Cons.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_Ammonia_Prod.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_Ammonia_Cons.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_Methanol_Prod.pdf}
-\includegraphics[width=0.49\textwidth]{UQ_Methanol_Cons.pdf}
-\caption{Distribution of the different streams of supply (left) and consumption (right) of gas, hydrogen, ammonia and methanol from the \acrfull{GSA}.}
-\label{fig:results_uq_prod_cons}
+\includegraphics[width=\textwidth]{UQ_Gas_Prod.pdf}
+\includegraphics[width=\textwidth]{UQ_Gas_Cons.pdf}
+\caption{Distribution of the different streams of supply (top) and consumption (bottom) of methane from the \acrfull{GSA}.}
+\label{fig:results_uq_prod_cons_Gas}
 \end{figure}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{UQ_H2_Prod.pdf}
+\includegraphics[width=\textwidth]{UQ_H2_Cons.pdf}
+\caption{Distribution of the different streams of supply (top) and consumption (bottom) of hydrogen from the \acrfull{GSA}.}
+\label{fig:results_uq_prod_cons_H2}
+\end{figure}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{UQ_Ammonia_Prod.pdf}
+\includegraphics[width=\textwidth]{UQ_Ammonia_Cons.pdf}
+\caption{Distribution of the different streams of supply (top) and consumption (bottom) of ammonia from the \acrfull{GSA}.}
+\label{fig:results_uq_prod_cons_Ammonia}
+\end{figure}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{UQ_Methanol_Prod.pdf}
+\includegraphics[width=\textwidth]{UQ_Methanol_Cons.pdf}
+\caption{Distribution of the different streams of supply (top) and consumption (bottom) of methanol from the \acrfull{GSA}.}
+\label{fig:results_uq_prod_cons_Methanol}
+\end{figure}
+
+\newpage
+\section{Installed capacities}
+\label{app:UQ_tech_cap}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{ELECTRICITY_Tech.pdf}
+\includegraphics[width=\textwidth]{HT_HEAT_Tech.pdf}
+\includegraphics[width=\textwidth]{LT_HEAT_Tech.pdf}
+\caption{Distribution of the installed capacities among the end-use sectors of electricity, \gls{HT} and \gls{LT} heat from the \acrfull{GSA}.}
+\label{fig:results_uq_tech_cap}
+\end{figure}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{AMMONIA_Tech.pdf}
+\includegraphics[width=\textwidth]{METHANOL_Tech.pdf}
+\includegraphics[width=\textwidth]{HVC_Tech.pdf}
+\caption{Distribution of the installed capacities among the end-use sectors of ammonia, methanol and \gls{HVC} from the \acrfull{GSA}.}
+\label{fig:results_uq_tech_cap}
+\end{figure}
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{PASS_MOB_Tech.pdf}
+\includegraphics[width=\textwidth]{FREIGHT_MOB_Tech.pdf}
+\caption{Distribution of the installed capacities among the end-use sectors of passenger and freight mobility from the \acrfull{GSA}.}
+\label{fig:results_uq_tech_cap}
+\end{figure}

### 2024-03-27 chap_RobPol/chap_RobPol.tex
Integration of HJ comments on appendices + RobPol start part 2

@@
 To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 80\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).
-This results in keeping 11 $\text{PC}_{y}$: the first PC of each representative year, the second of 2025, 2030, 2035 and 2040 and the third PC of 2025.
+This results in keeping 14 $\text{PC}_{y}$: the first and second \gls{PCs} of each representative year and the third PC of 2025 and 2035.
+Even though the 80\%-threshold is reached with the first 11 \gls{PCs},  $\text{PC}_{2045,2}$, $\text{PC}_{2050,2}$ and $\text{PC}_{2035,3}$ are accounted for as they are similar to $\text{PC}_{2040,2}$.
+Consequently, aiming at capturing at least 80\% of the total transition variance,
 
@@
 11 & 2040 & $\text{PC}_2$ & 8.7 & \textbf{80.2} \\
-\midrule
+12 & 2045 & $\text{PC}_2$ & 7.9 & 82.0 \\
+13 & 2050 & $\text{PC}_2$ & 7.7 & 83.7 \\
+14 & 2035 & $\text{PC}_3$ & 7.3 & 85.4 \\
 $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$\\
@@
 
-Given the similarity between $\text{PC}_y$, some $\text{PC}_{\text{transition}}$ result from the aggregation and averaging of the components of several $\text{PC}_y$.
+Given the similarity between $\text{PC}_y$ (see Figure \ref{fig:Top5_PC_year}), some $\text{PC}_{\text{transition}}$ result from the aggregation and averaging of the components of several $\text{PC}_y$ (see Table \ref{tab:PC_transition_aggregation}).
+This aggregation step has a double objective: limiting the dimension of the robustness metrics and avoiding ineffective redundancy in terms of $\text{PC}_{\text{transition}}$ where several of them would otherwise point towards similar directions of variation.
+
+\begin{table}[htbp!]
+\caption{Aggregation of $\text{PC}_y$ to construct the $\text{PC}_{\text{transition}}$ and share of the captured total variance of the transition by each $\text{PC}_{\text{transition}}$.}
+\label{tab:PC_transition_aggregation}
+\centering
+\begin{tabular}{c c c}
+\toprule
+ \multirow{2}{*}{$\mathbf{\textbf{PC}_{\textbf{transition}}}$} & \multirow{2}{*}{$\mathbf{\textbf{PC}_{\textbf{y}}}$} & \textbf{Share of total} \\	
+ &  & \textbf{variance} [\%] \\	
+ \midrule
+\multirow{6}{*}{$\text{PC}_{\text{transition,1}}$} & $\text{PC}_{\text{2025,1}}$ & \multirow{6}{*}{57.0}\\
+& $\text{PC}_{\text{2030,1}}$ & \\
+& $\text{PC}_{\text{2035,1}}$ & \\
+& $\text{PC}_{\text{2040,1}}$ & \\
+& $\text{PC}_{\text{2045,1}}$ & \\
+& $\text{PC}_{\text{2050,1}}$ & \\
+ \midrule
+\multirow{2}{*}{$\text{PC}_{\text{transition,2}}$} & $\text{PC}_{\text{2030,2}}$ & \multirow{2}{*}{13.4}\\
+& $\text{PC}_{\text{2035,2}}$ & \\
+\midrule
+\multirow{4}{*}{$\text{PC}_{\text{transition,3}}$} & $\text{PC}_{\text{2040,2}}$ & \multirow{4}{*}{7.2}\\
+& $\text{PC}_{\text{2045,2}}$ & \\
+& $\text{PC}_{\text{2050,2}}$ & \\
+& $\text{PC}_{\text{2035,3}}$ & \\
+ \midrule
+$\text{PC}_{\text{transition,3}}$ & $\text{PC}_{\text{2025,2}}$ & 4.3\\
+ \midrule
+$\text{PC}_{\text{transition,4}}$ & $\text{PC}_{\text{2025,3}}$ & 3.4\\
+\bottomrule
+
+\end{tabular}
+\end{table}
+
+Averaging the components of similar $\text{PC}_y$ allows constructing the $\text{PC}_{\text{transition}}$ (see Figure \ref{fig:Top5_PC_transition}).
+These directions of variation form the robustness matrix on which it is possible to project the results of transition roadmaps tested under uncertainties and myopic pathway optimisation.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Top5_PCtransition.pdf}
+\caption{Contribution of the top-5 technologies to the different $\text{PC}_{\text{transition}}$. }
+\label{fig:Top5_PC_transition}
+\end{figure}
+
+\section{Robustness assessment of pathway roadmaps}
+\label{sec:RobPol:Rob_Assessment}
+Now that the performance metric is defined, we can assess the robustness of different roadmaps.
+These roadmaps are defined as the technological mix given by the deterministic optimisation of the perfect foresight pathway under certain conditions.
+The two first roadmaps are the \textbf{REF} and \textbf{SMR} cases (see Chapter \ref{chap:atom_mol}).
+All the uncertain parameters are considered at their nominal value in these two cases.
+The only difference consists in allowing \gls{SMR} from 2040 onward in the \textbf{SMR} case.
+On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''.
+Then, we assess the robustness of these three roadmaps via a similar ``two-step rolling horizon approach'' as in the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the potential additional investments needed through myopic pathway optimisation under uncertainties.
+
+\subsection{The robust roadmap}
+\label{subsec:RobPol:Rob_roadmap}
+Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow here the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}.
+In their works, the authors consider a factor $\Gamma_{\text{obj}}\in [0,d]$ that represents a ``protection parameter'' of the objective function where $d$ is the total number of uncertain parameters (see Section \ref{subsec:pce}).
+If $\Gamma_{\text{obj}}= 0$, where all the uncertain parameters are at their nominal value, we obtain the deterministic solution of the \textbf{REF} case.
+If $\Gamma_{\text{obj}}= d$, this means considering the ``fully robust'' solution where all the uncertain parameters are their worst value, as in \citet{soyster1973convex}.
+Between these two extreme cases, the uncertain parameters to account for at their worst values follow the ranking given by the \gls{GSA} (see Chapter \ref{chap:atom_mol}).
+In practice, $\Gamma_{\text{obj}}= 1$ means considering the cost of purchasing electrofuels at its worst value.
+$\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value in the lot, and so on.
+In the present analysis, we consider the ``robust'' roadmap by setting $\Gamma_{\text{obj}}= 6$.
+Consequently, costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies are considered at their worst values, \ie the upper bound of their respective range.
+In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{LT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
+The following paragraphs investigate the differences in the different sectors between the ROB and the REF cases, similarly to the comparison carried out with the SMR case in Chapter \ref{chap:atom_mol}.\\
+
+\myparagraph{Power sector}\\
+
+Given the more expensive imported energy carriers, the model opts for an earlier electrification of the system and more efficiency.
+Consequently, the power sector is one of the most impacted ones (see Figure \ref{fig:rob_pol_ROB_elec}).
+At the earlier stages of the transition,  the model opts for the full deployment of local \gls{VRES} as soon as possible and importing more electricity (\ie +8.7~TWh, +52\%) from abroad in 2025.
+This earlier and bigger integration of \gls{VRES} is supported by the higher electrification of \gls{HT} and, to a smaller extent the \gls{LT}, heating sector as well as in the mobility sectors.
+It is mostly the industrial electric heaters that absorb the abundant and intermittent electricity produced by \gls{PV} panels and wind turbines.
+In the mobility sectors, \gls{BEV} substitutes gasoline cars from 2025 onward and electric trucks make the transition from diesel to \gls{FC} trucks.
+On the supply side, besides the local \gls{VRES} and the direct import of electricity from abroad, \gls{CHP} units, mostly industrial, are favoured given their higher efficiency at the expense of ammonia-\gls{CCGT}.
+In the ROB case, by 2050, these \gls{CHP} units produce 43.3~TWh, \ie 23\% of the total electricity production, versus 21.9~TWh in the REF case.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_ROB.pdf}
+\includegraphics[width=0.8\textwidth]{Elec_Layer_ROB.pdf}
+\caption{Installed capacities and layer balance of the power sector in the ROB case.
+Compared to the REF case, local \gls{VRES} are deployed as soon as possible supported by an earlier electrification of the heat and transport sectors.
+\gls{CHP} units also represents a higher share of the electricity supply given their higher efficiency at the expense of other flexible production units like ammonia-\gls{CCGT}.}
+\label{fig:rob_pol_ROB_elec}
+\end{figure}
+
+\myparagraph{Non-energy sector}\\
+
+Technologically speaking, by 2050, the ROB case accounts for an additional 1.2~GW of \gls{MTO} to supply the extra 10.3~TWh of \gls{HVC}.
+Besides this, the extra 2.5~TWh and 0.4~TWh of e-ammonia and e-methanol are supplied by supplied by the import of their respective energy carrier.

### 2024-03-27 chap_atom_mol/chap_atom_mol.tex
Integration of HJ comments on appendices + RobPol start part 2

@@
+%!TEX root = ../thesis_main.tex
+%!TEX encoding = UTF-8 Unicode
+\vspace{-0.2cm}
+\begin{flushright}
+\emph{``It is difficult to make predictions, especially about the future.''}\\
+Niels Bohr, in \textit{Bulletin of the Atomic Scientists}, 1971
+\end{flushright}
+\vspace{0.4cm}
+
 On top of scenarios with more profound behavioural changes, variety of technological pathways are often investigated to meet the ambitions of climate change mitigation.
@@
 
-Beyond the direct impact that \gls{SMR} has on the power sector, it is worth assessing the other sectors given the sector-coupling effect related to the whole-energy system optimisation \cite{contino2020whole}.
+Beyond the direct impact that \gls{SMR} has on the power sector, it is worth assessing the other sectors given the sector-coupling effect related to the whole-energy system optimisation \cite{contino2020whole}.\\
 
@@
 As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheap electricity flexibly produced by \gls{SMR} substitutes the \gls{CCGT}.
-With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 7.3\,TWh, 83.4\% less than in the REF case.
-Then, with a 9\%-Sobol' index, industrial \gls{EUD} also influences the need for this molecule, due to its \gls{NED}.
+With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 2.0\,TWh, 95.4\% less than in the REF case.
+Then, with a 12\%-Sobol' index, the cost of purchasing electricity from abroad, considered as renewable by 2050 and, therefore, a direct competitor to e-ammonia \gls{CCGT}, also affects the need of this molecule, especially when this cost is low.
 
 The conclusions are more straightforward for the import of e-methanol and the installed capacity of \gls{SMR}.
-For the former, industrial \gls{EUD} is, by far (\ie 81\% Sobol' index), the key factor.
+For the former, industrial \gls{EUD} is, by far (\ie 79\% Sobol' index), the key factor.
 Due to its own \gls{NED} but, above all, since it is the low-emitting alternative picked by the model to supply the significant \gls{NED} of \gls{HVC}, the lower this demand, the lower the need to import e-methanol, and vice versa.

### 2024-03-27 chap_case_study/chap_case_study.tex
PC_y of RobPol chapter
Integration of HJ comments on appendices + RobPol start part 2

@@
-%\begin{flushright}
-%%%\emph{``For the things we have to learn before we can do them, we learn by doing them.''}\\%Each wrong scenario you have analysed, is a wrong decision you won't made.''}\\
-%%%Aristote
-%\emph{All models are wrong, but some are useful.}\\
-%George P. Box
-%\end{flushright}
-
+\vspace{-0.2cm}
+\begin{flushright}
+\emph{``Ils vont le faire!''}\\
+Philippe Albert, during Belgium-Japan FIFA World Cup round of 16, 2018
+\end{flushright}
+\vspace{0.4cm}
 %\medskip
@@
 The fundamental step in this ambition is to characterise these uncertainties.
+
 In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined range of uncertainties for the model parameters.
 \Cref{tab:UC_short} gives the uncertainty ranges of some key parameters.
+As the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III, based on the work of \citet{Moret2017PhDThesis}.
+For type III uncertainties (\ie uncertainty ranges increasing with time), a 50\% increase has been set arbitrarily between the ranges for 2025 and these same ranges for 2050.
+In other words, for these specific uncertainties, the ranges for 2050 are 50\% larger than for 2025.
 Like other works \cite{li2019renewables,coppitters2021robust}, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+
+\citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits.
+Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}.
+The uncertainty characterisation gives the uncertainty ranges per parameter or group of parameters (category).
+
+This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
 A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_short}.
@@
 }.
-Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about the availability.
+Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability.
 Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
 
-The exhaustive list of the parameters accounted for in this work is presented in Appendix \ref{app:UC_full}.
+Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.
 
 \begin{table}[htbp!]
-\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025. }
-\label{tab:UC_short}
+\caption{Application of the uncertainty characterization method of \citet{Moret2017PhDThesis} to the EnergyScope Pathway model for the year 2025. }
+\label{tab:UC_full}
 \begin{minipage}{\linewidth}
@@
 \toprule
-\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
-}}  & \multicolumn{2}{c}{\textbf{Relative variation\footnote{\label{foot:nom_val_uncert}The nominal values for each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
-}}}\\
+\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range_app}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg .
+}}  & \multicolumn{2}{c}{\textbf{Relative variation}\footnote{\label{foot:nom_val_uncert_app}The nominal values of each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+}}\\
     & & & &	 min 	&	 max \\ 	
 \midrule		
-\multirow{2}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+\multirow{4}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+& $c_{\mathrm{op,elec}}$ & Purchase electricity & II & -64.3\% & 179.8\% \\
 & $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
+& $c_{\mathrm{op,biofuels}}$ & Purchase biofuels & II & -64.3\% & 179.8\% \\
 \midrule
-\multirow{5}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+\multirow{9}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+& $c_{\mathrm{inv,bus}}$ & CAPEX bus & I & -21.6\% & 25.0\% \\
+& $c_{\mathrm{inv,ic\_prop}}$ & CAPEX ICE & I & -21.6\% & 25.0\% \\
 & $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
 & $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
+& $c_{\mathrm{inv,efficiency}}$ & CAPEX efficiency measures & I & -39.3\%  & 39.3\% \\
 & $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
-& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}& I & -40.0\% & 44.0\% \\
+& $c_{\mathrm{inv,grid}}$ & CAPEX power grid & I & -39.3\% & 39.3\% \\
+& $c_{\mathrm{inv,grid\_enforce}}$ & CAPEX grid reinforcement & I & -39.3\% & 39.3\% \\
+& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR_app}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.} & I & -40.0\% & 44.0\% \\
 \midrule
-\multirow{1}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+\textbf{Maintenance cost} & $c_{\mathrm{maint,var}}$ & Variable OPEX of technologies & I & -48.2\% & 35.7\% \\
 \midrule
-\multirow{2}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
+\multirow{2}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+& $\eta_{\mathrm{fc\_prop}}$ & Consumption fuel cell vehicles & I & -28.7\% & 28.7\% \\
+\midrule
+\multirow{3}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
 & $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
+& $f_{\mathrm{max,windoff}}$ & Max capacity offshore wind & I & -24.1\% & 24.1\% \\
 \midrule
@@
 
-\multirow{2}{*}{\textbf{End-use demand}} & $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
+\multirow{4}{*}{\textbf{End-use demand}} &$HH\_EUD$ & Households EUD & III & -13.8\% & 11.2\% \\
+& $services\_EUD$ & Services EUD & III & -14.3\% & 11\% \\
+& $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
 & $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
@@
 
-\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+& $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\
 & $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
 & $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
+& $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
 & $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
@@
 \end{table}
+
+%\begin{table}[htbp!]
+%\caption{Illustration of the uncertainty characterisation for different parameters for the year 2025. }
+%\label{tab:UC_short}
+%\begin{minipage}{\linewidth}
+%\centering
+%\resizebox{\textwidth}{!}{
+%\begin{tabular}{l l l c c c}
+%\toprule
+%\multirow{2}{*}{\textbf{Category}} & \multirow{2}{*}{\textbf{Parameter}} & \multirow{2}{*}{\textbf{Meaning}} & \multirow{2}{*}{\textbf{Type}\footnote{\label{foot:type_uncert_range}Per \citet{Moret2017PhDThesis}, \og I: investment-type, II: operation-type (constant uncertainty over time), III: operation-type (uncertainty increasing over time)\fg.
+}}  & \multicolumn{2}{c}{\textbf{Relative variation\footnote{\label{foot:nom_val_uncert}The nominal values for each of the parameters is 0, meaning no variation compared to the nominal values of the impacted parameter in the model.
+}}}\\
+%    & & & &	 min 	&	 max \\ 	
+%\midrule		
+%\multirow{2}{*}{\textbf{Cost of purchasing}} & $c_{\mathrm{op,fossil}}$ & Purchase fossil fuels & II & -64.3\% & 179.8\% \\
+%& $c_{\mathrm{op,electrofuels}}$ & Purchase electrofuels & II & -64.3\% & 179.8\% \\
+%\midrule
+%\multirow{5}{*}{\textbf{Investment cost}} &$c_{\mathrm{inv,car}}$ & CAPEX car  & I & -21.6\% & 25.0\% \\
+%& $c_{\mathrm{inv,e\_prop}}$ & CAPEX electric motor & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,fc\_prop}}$ & CAPEX fuel cell engine & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,PV}}$ & CAPEX PV & I & -39.6\% & 39.6\% \\
+%& $c_{\mathrm{inv,nuclear\_SMR}}$ & CAPEX \gls{SMR}\footnote{\label{foot:range_SMR}This range has been inferred from the local sensitivity analysis performed by \citet{PATHS2050}.}& I & -40.0\% & 44.0\% \\
+%\midrule
+%\multirow{1}{*}{\textbf{Consumption}} &$\eta_{\mathrm{e\_prop}}$ & Consumption electric vehicles & I & -28.7\% & 28.7\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Potential installed capacity}} &$f_{\mathrm{max,PV}}$ & Max capacity PV & I & -24.1\% & 24.1\% \\
+%& $f_{\mathrm{max,windon}}$ & Max capacity onshore wind & I & -24.1\% & 24.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Hourly load factor}} & $c_{\mathrm{p,t,PV}}$ & Hourly load factor PV & II & -22.1\% & 22.1\% \\
+%& $c_{\mathrm{p,t,winds}}$ & Hourly load factor wind turbines & II & -22.1\% & 22.1\% \\
+%\midrule
+%\multirow{2}{*}{\textbf{Resource availability}} & $avail_{\mathrm{elec}}$ & Available electricity import & I & -32.1\% & 32.1\% \\
+%& $avail_{\mathrm{biomass}}$ & Available local biomass & I & -32.1\% & 32.1\% \\
+%\midrule
+%
+%\multirow{2}{*}{\textbf{End-use demand}} & $pass\_EUD$ & Passenger mobility EUD & III & -7.5\% & 7.5\% \\
+%& $industry\_EUD$ & Industry EUD & III & -20.5\% & 16.0\% \\
+%\midrule
+%
+%\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+%& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\
+%& $\Delta_{\mathrm{change,pass}}$ & Modal share change passenger mobility & - & -30\% & 30\% \\
+%& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
+%
+%\bottomrule							
+%
+%\end{tabular}}
+%\end{minipage}
+%\end{table}
 

### 2024-03-27 chap_intro_ccl/conclusion.tex
Integration of HJ comments on appendices + RobPol start part 2

@@
 I took here the same sections as in Gauthier's thesis
-\section*{Thesis contribution}
-\addcontentsline{toc}{section}{Thesis contribution}
+\section*{Thesis contributions}
+\addcontentsline{toc}{section}{Thesis contributions}
 Insist here on the methodological added value of the thesis
+
+\section*{Limitations}
+\addcontentsline{toc}{section}{Limitations}
+\begin{itemize}
+\item Due to the formulation of the salvage value, some technologies remain in place whereas they are not used anymore (see Appendix B.3, e.g.
+coal boilers and naphtha-cracker).
+It's better for the system to keep unused assets to recover part of their salvage value than prematurely decommissioning them.
+\end{itemize}
 

### 2024-03-29 chap_RobPol/chap_RobPol.tex
RobPol - Description case ROB

@@
 On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''.
-Then, we assess the robustness of these three roadmaps via a similar ``two-step rolling horizon approach'' as in the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the potential additional investments needed through myopic pathway optimisation under uncertainties.
+Then, we assess the robustness of these three roadmaps via a similar ``three-step rolling horizon approach'' as in the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.
 
@@
 In the ROB case, by 2050, these \gls{CHP} units produce 43.3~TWh, \ie 23\% of the total electricity production, versus 21.9~TWh in the REF case.
+These observations are in line with the work of  \citet{moret2020overcapacity} focusing on the problem of overcapacity in the European power sector.
+The authors highlighted that the robust strategy diversified the supply sources of electricity between \gls{VRES}, import of electricity and more efficient technologies like \gls{CHP} and \gls{HP}.
 
@@
 
-\myparagraph{Non-energy sector}\\
+\myparagraph{Heating sectors}\\
+
+By 2050, the additional 23.9\% of industrial \gls{EUD} directly impacts the technological mix to supply the \gls{HT} heat, \ie +13.6~TWh as direct \gls{HT} heat \gls{EUD} on top of the extra 3.4~TWh consumed by the \gls{MTO} process.
+As aforementioned, at earlier stages of the transition, up to 20.9~GW of industrial electric heaters supply the additional demand and support the integration of solar \gls{PV}.
+However, in the mid-term, 27\% of these heaters are prematurely decommissioned where 24\% are not renewed after reaching the end of their lifetime.
+Instead, up to +3.5~GW of industrial \gls{CHP} units are installed to efficiently use e-methane that is considered more expensive in the ROB case to supply electricity and 52\% of the \gls{HT} heat demand.
+About the \gls{LT} heat sector, the principal impact is the bigger installation of \gls{DHN} and decentralised electric \gls{HP}, up to +2.1~GW and +0.8~GW respectively.
+\\
+
+\myparagraph{Mobility sectors}\\
+
+In the transport of freight, the ROB roadmap transition from diesel to \gls{FC} trucks with the use of electric trucks from 2025 to 2035.
+Similarly to the industrial electric heaters, it is a way to integrate to early installation of solar \gls{PV} and absorb part of its intermittent electricity.
+Freight transport via trains and boats are not different from the REF roadmap.
+About the passenger mobility, compared to the REF case, the shift from \gls{ICE} cars to \gls{BEV} occurs earlier in the transition, in line with the increased production of intermittent electricity from \gls{VRES}.
+\\
+
+\myparagraph{Non-energy}\\
 
 Technologically speaking, by 2050, the ROB case accounts for an additional 1.2~GW of \gls{MTO} to supply the extra 10.3~TWh of \gls{HVC}.
-Besides this, the extra 2.5~TWh and 0.4~TWh of e-ammonia and e-methanol are supplied by supplied by the import of their respective energy carrier.
+Besides this, the extra 2.5~TWh and 0.4~TWh of e-ammonia and e-methanol are supplied by supplied by the import of their respective energy carrier.\\
+
+In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}.More expensive imported energy carriers lead to, on the one hand, an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) supported by a higher electrification of the heating and transport sectors.
+On the other hand, it increases the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
+This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}.
+The increased interest rate and variable OPEX of technologies have a more negligible impact on the roadmap strategy as they identically affect the entire set of technologies.
+
+\subsection{Projection on robustness metric}
+\label{subsec:RobPol:Projection}
+After defining the robustness metric and describing the roadmaps to assess (\ie REF, SMR and ROB), the final step consists in testing these roadmaps under uncertainties in a myopic pathway optimisation.
+This aims at assessing the adaptation of the roadmaps as uncertainty gradually unfolds over time \cite{moret2020overcapacity}, which brings more realism than perfect foresight \cite{poncelet2016myopic}.
+The three roadmaps are tested 500 times fixing the installed capacities $f_{\mathrm{max}}\geq\textbf{F}\geq\textbf{F}^*$ for all the end-use technologies, and taking values of the 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}) following the sampling technique detailed in Section \ref{subsec:pce}, \ie Sobol' sequence.
+However, uncertain parameters limit the potential installed capacity, $f_{\mathrm{max}}$, of \gls{PV} panels, onshore and offshore wind turbines and \gls{SMR}.
+For these, in case the installed capacity prescribed by the roadmap is higher than the maximum potential affected by the value of the uncertain parameter, the actual installed capacity is set to this new limit.
+In reality, this would be similar to revise to a smaller extent the expected roadmap due to external events (\eg smaller public acceptance \cite{zoellner2008public,sam2014small}, lower \gls{TRL} or longer installation time than expected)
+
+\section{Discussion}
+\label{sec:RobPol:Discussion}

### 2024-03-29 chap_case_study/chap_case_study.tex
RobPol - Description case ROB

@@
 & $f_{\mathrm{max,windoff}}$ & Max capacity offshore wind & I & -24.1\% & 24.1\% \\
+& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
 \midrule
@@
 & $\Delta_{\mathrm{change,LT\_heat}}$ & Modal share change LT-heat & - & -30\% & 30\% \\
-& $f_{\mathrm{max,SMR}}$ & Potential capacity \gls{SMR} & - & 0 & 1 \\
 

### 2024-03-29 chap_methodo/chap_methodo.tex
RobPol - Description case ROB

@@
 \noindent
-Similarly to the work of \citet{moret2020overcapacity}, roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimizations (see Chapter \ref{chap:chap_RobPol}).
-The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps.
+Similarly to the work of \citet{moret2020overcapacity}, roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimizations.
+The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps (see Figure \ref{fig:Methodo_Robustness}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Methodo_Robustness.pdf}
+\caption{Overview of the three-step methodology to assess the robustness of roadmaps. Schematic adapted from \cite{moret2020overcapacity}.}
+\label{fig:Methodo_Robustness}
+\end{figure}
+
 In conclusion, a roadmap would be defined as more robust than another one if the projection of its myopic runs on the different $\text{PC}_{\text{transition}}$ spans on a more narrow range (see Figure \ref{fig:Projection_PC_transition}).

### 2024-04-03 chap_intro_ccl/intro.tex
Intro sent for review to Diederik and Arnaud

@@
 This aligns with the current European policies binding the Member States of the European Union.
-For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind\footnote{This directly relates to sufficiency as it encompasses social justice in parallel with a minimization of the energy use.}, an \textbf{increase in energy efficiency} and significantly \textbf{higher shares of energy from renewable sources} in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
+For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind\footnote{This directly relates to sufficiency as it encompasses social justice in parallel with a minimization of the energy use.}, an increase in energy efficiency and significantly higher shares of energy from renewable sources in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
 
-Consequently, to ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
+\section*{Context and motivations}
+
+To ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
 Besides behavioral changes, an overall reshape of the energy system is necessary in terms of both primary energy sources, \ie more renewables, and technologies used to convert these resources into the \gls{EUD} (\ie the energy service required by the the final consumer), \ie more efficiency \cite{iea2020world,luderer2018residual}.
@@
 To address this issue, due to some limitations (\eg range, power, costs) of electricity-focused solutions like \gls{DC} lines, the transport and long-term storage of the renewable electricity produced in excess should be optimised.
-
 This challenge can be tackled by \textit{electrofuels} \cite{rozzi2020}.
@@
 With a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
-Besides direct electrification of other sectors (\eg electrical heat pumps, battery electric vehicles), \citet{brown2018synergies} showed that converting power to hydrogen and methane was advantageous at high shares of renewables, in their optimisation of the European whole-energy system.
+Besides direct electrification of other sectors (\eg electrical heat pumps, \gls{BEV}), \citet{brown2018synergies} showed that converting power to hydrogen and methane was advantageous at high shares of renewables, in their optimisation of the European whole-energy system.
 Electrofuels have the ability to couple energy and non-energy sectors \cite{Stancin2020}.
-For instance, electricity produced in excess from \gls{VRES} can be converted in ammonia through the Haber-Bosch process and subsequently transformed into fertiliser - coupling the power and industry sectors \cite{verleysen2020can}.
+For instance, electricity produced in excess from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser - coupling the power and industry sectors \cite{verleysen2020can}.
 Gas networks present much more storage potential than electrical network (\eg 50 times more in Germany and 300 times more in France) \cite{Rosa2017}.
-Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}.) Besides storing energy, in their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
+Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}).
+In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
 Moreover, some applications (\eg marine, aviation and heavy-duty transport) will be hard to electrify and keep on requiring high-density energy carriers \cite{horvath2018techno, brynolf2018}.
-These carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons in a renewable world.
-This is why this paper rather uses "defossilisation" rather than "decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
+These carriers, currently produced from fossil resources, will still consist of hydrocarbons in a renewable world.
+This is why this thesis rather uses "defossilisation" rather than "decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
 
@@
 To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a multi-sector and whole-energy system \cite{Contino2020}.
-To reach this goal, an energy system optimisation model (ESOM) can define the design of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
-In this research field, \citet{yue2018review} highlighted that most of ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
+To reach this goal, an energy system optimisation model (ESOM) can optimise the design and the operation of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
+
+In this research field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most of ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
 However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system, such as a country.
@@
 Besides the uncertainty on the model structure (not addressed in this work), this long-term and large-system optimisation motivates the need to account for \gls{UQ} and consider it as a major challenge of such models \cite{pfenninger2014energy}.
-This challenge, along with a large number (\ie more than a hundred) of uncertain parameters and limited information of their distribution, leads to the "curse of dimensionality" \cite{kuo2005lifting}.
+This challenge, along with a large number (\ie more than a hundred) of uncertain parameters and limited information of their distribution, leads to the "curse of dimensionality" \cite{kuo2005lifting}, where the computational burden rapidly explodes with the number of considered uncertain parameters.
 
-It aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
-Instead of aiming at answering the question ``What could possibly happen in the future?'', let's rather address ``What could or should we do to make the future possible?'' ``Our task is not to foresee the future, but to enable it'' by Saint-Exupéry in Citadelle, 1948.
+On top of dealing with these uncertainties, the reality of decision-makers is a limited foresight into the future \cite{poncelet2016myopic}.
+In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'', \ie 2050 (and beyond), and make the planning decisions once and for all accordingly.
+On the contrary, they uncover the realisation of these uncertainties step-by-step (\ie in a ``myopic'' way), and progressively act on them to, hopefully, meet the set target to mitigate the climate change.
+In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there was a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
+On top of the ``what to do?'', this approach aims at helping the policymakers to answer the  question ``how to do it?''.
 
-%!TEX root = ../thesis_main.tex
-%!TEX encoding = UTF-8 Unicode
+Finally, when assessing the robustness of transition pathways provided by \gls{ESOMs}, the literature shows a variety of techniques, \eg Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
+However, the robustness is commonly applied to assess the sensibility of the solution regarding the objective function to optimise (\ie the total cost).
+Given the complexity of whole-energy system models (\ie multiple sectors and multiple energy carriers) along with the time-scale of the transition and the lot of uncertainties, the total cost does not provide information on the sensibility of the design strategy, \ie the investment decisions.
+Where \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
+However, their work focused on the power sector only and for a target future year, \ie 2035.
+The literature was missing an approach to tackle these two challenges: to go beyond the total transition cost encompassing the details of a solution into a single value while assessing the whole-energy system over its whole transition.
 
-In their 2030 Agenda, United Nations have worked on identifying 17 \gls{SDGs} as a plan of action for society (or people), environment (or planet) and economy (or prosperity) \cite{un_sdgs}. 
-%
-%As Fatih
-%
-%On top of the depletion of (economically and environmentally) easy to harvest conventional fossil resources, the most urgent motivation for energy transition is the growing threat of climate change.
-This change jeopardizes most of the \gls{SDGs} listed by the United Nations .
-Besides their environmental pillars (\ie 13-climate action, 14-life below water and 15-life on land)\cite{vinuesa2020role}, succeeding such a transition would contribute in reaching others of these goals like affordable
+\section*{Objectives and tasks}
+``\textit{Our task is not to foresee the future, but to enable it.}'' Saint-Exupéry in Citadelle, 1948.
+In that sense, this thesis aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
+Rather than trying, in vain, to answer the question ``What could possibly happen in the future?'', this work rather addresses the ``What could or should we do to make the future possible?''.
+Given these general context and motivations, the research questions are described as follows:
+\begin{itemize}
+\item What is the role of electrofuels in the transition pathway of a whole-energy system subject to uncertainties, limited foresight into the future and a \ce{CO2}-budget?
+What are the key uncertainties driving their import?
+\item How to explore the multiple pathway possibilities through the optimisation of a policy, \ie sequence of actions to support a transition?
+\item How to assess the robustness of a pathway roadmap defining the design strategy over the transition?
+\end{itemize}
 
-From this need, develop a thread like I did in my FRIA application, we need whole-energy system model to give more insights to policymakers\\
+To answer these questions, several tasks have been carried out on the whole-energy system model accounting for the uncertainties, the method to explore the myopic pathway of such a system and the approach to assess the robustness of a pathway roadmap.
+All this work has been applied to the case of Belgium, a densely populated country with a limited amount of renewable energies, which represents roughly one third of the forecast energy demand \cite{Limpens2020}.
+This makes it a challenging case to go from a highly fossil-dominated system in 2020 (\ie 73\% of the primary energy mix \cite{spf_economy_2022}) to carbon-neutrality by 2050.
 
-Listing some examples of situations where uncertainties have not been considered and ended up in over-cost/waste of time/waste of..., highlight the need as well to consider uncertainties when we want to advise policymakers.\\
+The model developed and used in the context of this thesis is EnergyScope Pathway \cite{limpens2024pathway}.
+First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for the pathway transition from an existing system to a long term target, \ie 2050.
+Compared to other models, EnergyScope Pathway introduces a rapid (\ie around 15~min on a personal laptop for the optimisation of a 30-year transition) computational optimisation tool for exploring diverse transition pathways within an entire energy system while maintaining temporal precision (\ie hourly time-resolution) to accurately capture the integration of \gls{VRES}.
+Based on this model originally implemented in a perfect foresight approach, \ie one overall optimisation for the whole pathway, we have developed the myopic method.
+In this, the whole time horizon (\ie 30 years) is optimised through a sequence of 10-year long time-windows having a 5-year overlap between each other.
+To address the question about the role of electrofuels, we have detailed further on their implementation in the model to account for four main ones: hydrogen, methane, ammonia and methanol.
+Moreover, given their current and (expected) future role into the sector of the \gls{NED}, we have implemented this sector with a similar level of detail as the other sectors of the system, \ie electricity, heat and transport.
 
-As the question of policymakers is not only what to do but how to do it, we need to address the optimisation of policies. This would introduce the RL part
+To account for uncertainties, we have used, and adapted to our case study, the work of \citet{Moret2017PhDThesis} and \citet{coppittersthesis} on the uncertainty characterisation and quantification, respecitvely.
+In his thesis, \citet{Moret2017PhDThesis} developed a framework to obtain uncertainty range for a variety of parameters like cost of purchasing and availability of resources or investment cost and efficiency of technologies.
+These ranges have then been sampled and propagated through the EnegyScope Pathway via the RHEIA framework developed by \citet{coppittersthesis}.
+Using the surrogate-modeling approach called \gls{PCE}, this framework allows identifying the uncertain parameters with the biggest impact on the variation of total transition cost or other outputs of interest like the imported amount of electrofuels.
 
-\textbf{In a more sustainable future, some of the energy carriers, currently produced mostly from fossil resources, will still consist of hydrocarbons (\eg e-methane or e-methanol).
-This is why this paper rather uses ``defossilisation" rather than ``decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.}
+To explore the different pathway trajectories in this myopic optimisation process, we have applied the \gls{RL} method where an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie sequence of actions to take to support the transition.
+Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
+Although, these actions are taken every five years, they impact the system, for the next ten years - their time window.
+The intermediate solutions obtained in the middle of the time window are used as a new starting point for the agent that makes a new series of decisions for the next ten years, etc.
+Repeating the whole transition with different sequences of actions-states allow the agent to come up with an optimised policy towards sustainability, considering the variation of the parameters of its environment.
+
+Finally, in the objective to assess the sensibility to uncertainties of different design transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
+This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a frame.
+Roadmaps resulting from perfect foresight optimisation have been tested in a myopic and uncertain pathways.
+The results of these myopic runs have then been ``projected'' on the aforementioned frame to be able to compare the robustness of different roadmaps between each other.
+
+\section*{Outline}
+This thesis is composed of five chapters to provide answers to the different research questions.
+Chapter \ref{chap:chap_methodo} brings more details about the different methodological aspects of this work.
+It starts with the main constraints, parameters and variables of the whole-energy system optimisation model, EnergyScope Pathway.
+Then, it gives information on the \gls{UQ} approach and the way it has been adapted to the case of Belgium and its transition pathway.
+Finally, general fundamentals and more case-specific considerations are brought up about the \gls{RL} and \gls{PCA}-based robustness approaches.
+
+Chapter \ref{chap:case_study} presents the case study of this work, \ie Belgium and its energy transition.
+Without exhaustively detailing the data \cite{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the date related \gls{SMR} as an option to produce nuclear-based electricity in the future.
+
+Chapter \ref{chap:atom_mol} is the first of the three chapters presenting results.
+In this one, we detail the results of the \gls{GSA} carried out on the Belgian energy transition under the lens of the atom-molecules dilemma.
+On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear capacities by 2040 onward and the driving parameters on the import of electrofuels.
+
+In Chapter \ref{chap:chap_RL}, the rules of the ``\gls{RL} game'' are detailed.
+In other words, we define the action and state spaces as well as the reward function driving the behaviour of the agent in its quest to optimise its policy.
+Then, we analyse the results of the learning phase before testing under uncertainties the learned policy versus more ``classic'' myopic optimisation, \ie without the support of such a policy.
+
+After detailing the robustness metric based on the \gls{PCA} approach, Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting of deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
+The first one, the reference case, considers nominal values for all the uncertain parameters.
+The SMR case is the one introduced in Chapter \ref{chap:atom_mol} where we allow the model to install \gls{SMR} from 2040 onward.
+Eventually, the ROB case accounts for the highest values of parameters for those having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, interest rate and variable \gls{OPEX} of technologies).
+
+Finally, we draw general conclusions and suggest potential perspectives for future works in terms of uses and further developments of the methodological tools.

### 2024-04-04 appendices/app_case_study.tex
Intro post-review DC

@@
 
+\section{Monthly versus hourly pathway optimisation}
+\label{app:mo_vs_td}
+
+This section compares the perfect foresight optimisation based on a monthly time resolution with the reference case (hourly time resolution).
+The main advantage of the monthly approach is its computational time, \ie a reduction of 99.8\% compared to the hourly resolution (see Table \ref{tab:TD_MO}).
+This tractability represents a significant advantage when several thousands of runs are necessary, \eg for uncertainty quantification as it has been performed in \cite{rixhon2021role}.
+
+\begin{table}[htbp]
+\caption{Comparison between the two different time-resolutions: Hourly (PF) and Monthly (MO).
+Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
+\label{tab:TD_MO}
+\begin{minipage}{\textwidth}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{l l r r c}
+\toprule
+& & {\thead{\textbf{PF}}} & {\thead{\textbf{MO}}} & Units \\
+\midrule
+Computational time\footnote{\label{foot:comp_time_mo}These computational times were reached on a 2.4GHz 4-core machine.} & & 830 & \textbf{2} & s \\
+\midrule
+\multirow{4}{*}{Costs in 2050}
+& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\textsubscript{2015}\\
+& Cumulative opex & 528 & -3\%  & b€\textsubscript{2015}\\
+& Cumulative capex & 636 & -2\% & b€\textsubscript{2015}\\
+& Salvage value & 160 & -5\% & b€\textsubscript{2015}\\
+\midrule
+\multirow{5}{*}{Primary energy mix in 2050} & Total & 368.0 & +2\% & TWh/y\\
+ & e-hydrogen & 15.6 & +2\% & TWh/y\\
+ & e-methane & 41.0 & \textbf{+51\%} & TWh/y\\
+ & e-methanol & 54.8 & $\simeq$ & TWh/y\\
+ & e-ammonia & 40.7 & \textbf{+55\%} & TWh/y\\
+
+ \midrule
+ \multirow{3}{*}{Electrification in 2050\footnote{\label{foot:share_elec_mo}The electrification of the other sectors (\ie centralised heat (100\%-heat pump), private (100\%-\gls{BEV}), public mobility (80\%-train and tramway), freight mobility (25\%-train) and non-energy demand (0\%)) are identical between the three approaches and are, therefore, not presented in the table.}} & System\footnote{\label{foot:system_elec_mo}The electrification of the system is computed as the difference between the total production of electricity and the end-use demand of electricity.} & 63.3 & \textbf{-27\%} & TWh$_{\text{e}}$\\
+ & Industrial heat\footnote{\label{foot:heat_elec_mo}The electrification of the industrial and decentralised heating sectors are expressed in terms of thermal energy (TWh$_{\text{th}}$) provided by electrified processes, respectively industrial resistors and decentralised electric heat pumps.} & 12.3 & \textbf{-98\%} & TWh$_{\text{th}}$\\
+ & Decentralised heat\footref{foot:heat_elec_mo} & 73.9 & \textbf{-13\%} & TWh$_{\text{th}}$\\
+  \midrule
+\multirow{3}{*}{Year of full VRES-deployment} & PV & 2045 & \textbf{2040} & -\\
+ & Wind-offshore & 2030 & \textbf{2025} & -\\
+ & Wind-onshore & 2025 & 2025 & -\\
+\bottomrule
+\end{tabular}
+}
+\end{minipage}
+\end{table}
+
+However, averaging time series of end-use demands and renewable productions brings some discrepancies.
+First and foremost, \gls{VRES} loose their intrinsic power intermittency \cite{haydt2011relevance}.
+This has a series of consequences.
+Overall, this overestimates the uptake and emergence of solar and wind power generation and, consequently, overestimate their share in the primary electricity mix, especially at intermediate stages of the transition \cite{haydt2011relevance,Poncelet2016} (see Figure \ref{fig:MO_Share_VRES_elec_mix}).
+In line with these studies, the more there is \gls{VRES} in the primary electricity mix, the higher is the error\footnote{Like \citet{Poncelet2016}, this error is defined as:$\sum_{i} \frac{|\text{supply share}_i^{MO}-\text{supply share}_i^{PF}|}{2}$, where index i runs over all technologies and import of electricity and the supply shares are expressed as a percentage.} with lower time resolution.
+In 2025 and 2050, where the share of \gls{VRES} in the electricity primary mix is 46\% and 66\%, respectively, this error goes from 10\% to 18\%.
+
+ \begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{MO_Share_VRES_elec_mix.pdf}
+\caption{Share of \gls{VRES} (\ie solar and wind) in the electricity primary mix for the monthly and hourly time resolutions.}
+\label{fig:MO_Share_VRES_elec_mix}
+\end{figure}
+
+When considering the system overall, this averaged time-variability requires a more limited sector-coupling in the monthly model to absorb the intermittency of renewables.
+Industrial and decentralised heat demands are less electrified (see \autoref{tab:TD_MO}).
+This leads to 30\% electrification of the overall system versus 36\% for the hourly approach.
+This smaller production of electricity is done to the detriment of direct import of electricity, completely removed from the primary mix, as assumed to keep on having a higher global warming potential than conventional assets (\eg \gls{CCGT} running on gas and, later on, renewable gas and ammonia).
+Then, the generation mix in heating sectors or non-energy demands is less diversified, leading the sharper and later switch of resources and technologies.
+For instance, lignocellulosic biomass is ``cannibalized" by industrial boilers to the detriment of biomass-to-methanol or \gls{HVC}.
+Besides this, in general, we observe an overestimation of conventional technologies (\eg running on gas or oil) generation, either by installing more capacities or having a higher load factor for similar capacities.
+In line with \citet{Poncelet2016}, this leads to an overall underestimation of the operational (and investment) costs (see \autoref{tab:TD_MO}).
+
+Finally, in the mobility sectors, there is no difference between low and high time resolutions.
+This is due to the fact that these demands are assumed to stay the same over the different hours of the year (\ie freight mobility) or have a favorite technology to supply them (\eg \gls{BEV} for private mobility).
+
 %\section{Uncertainty characterisation for the 5-year steps transition} 

### 2024-04-04 appendices/app_results_UQ.tex
Intro post-review DC

@@
 \begin{table}[htbp!]
-\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the hourly and monthly pathway models.
+\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the hourly and monthly (see Appendix \ref{app:mo_vs_td}) pathway models.
 The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties.}
@@
 
-Figures \ref{fig:results_uq_prod_cons_Gas}, \ref{fig:results_uq_prod_cons_H2}, \ref{fig:results_uq_prod_cons_Ammonia} and \ref{fig:results_uq_prod_cons_Methanol} give the distribution of the different routes of supply and consumption methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.
+Figures \ref{fig:results_uq_prod_cons_Gas}, \ref{fig:results_uq_prod_cons_H2}, \ref{fig:results_uq_prod_cons_Ammonia} and \ref{fig:results_uq_prod_cons_Methanol} give the distribution of the different routes of supply and consumption of methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.
 

### 2024-04-04 chap_RL/chap_RL.tex
Intro post-review DC

@@
 At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
-Moreover, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
+Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
 This sample will remain valid for the following time windows.
@@
 For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+
+To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase, also called ``training'', of the agent has been done on the monthly model.
+Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
 

### 2024-04-04 chap_atom_mol/chap_atom_mol.tex
Intro post-review DC

@@
 First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
+Second, accounting for uncertainties as presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
 

### 2024-04-04 chap_intro_ccl/intro.tex
Intro post-review DC

@@
-It has been proven that the climate change, among other environmental challenges, is (mostly) due to the concentration of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
+Climate change, among other environmental challenges, is (mostly) due to the concentration of anthropogenic \acrfull{GHG} in the environment \cite{IPCC_CO2_budget}.
 Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
 Among these emissions, globally, about 75\% are directly related to the whole-energy system \cite{ourworldindata_CO2_world}.
-The \gls{GHG} emissions usually expressed in kt$_{\ce{CO2},\text{eq}}$, could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
+The \gls{GHG} emissions, usually expressed in kt$_{\ce{CO2},\text{eq}}$, could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
 
@@
 \label{eq:equality_GHG}
-\mathrm{GHG} =  \frac{\mathrm{GHG}}{\text{Primary energy}} \times \frac{\text{Primary energy}}{\mathrm{EUD}}\times \frac{\mathrm{EUD}}{\text{Population}} \times \text{Population}\\
+\mathrm{GHG} =  \frac{\mathrm{GHG}}{\text{Primary energy}} \times \frac{\text{Primary energy}}{\mathrm{EUD}}\times \frac{\mathrm{EUD}}{\text{Population}} \times \text{Population,}\\
  \end{equation}
@@
 \noindent
-where the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third could stand as the energy intensity per capita.
-Such an identity, mathematically-correct though, is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its global/encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
+where the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third stands as the energy intensity per capita.
+Such an identity, mathematically-correct though, is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
 However, Eq.
-\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition, leaving aside the question of the total population and its growth \cite{dodson2020population,scovronick2017impact}.
+\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition, considering that the population will continue to grow \cite{dodson2020population,scovronick2017impact}.
 These three levers of actions are: renewables, efficiency and sufficiency aiming at reducing the first, the second and the third terms on the right-hand side of Eq.
@@
 To ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
-Besides behavioral changes, an overall reshape of the energy system is necessary in terms of both primary energy sources, \ie more renewables, and technologies used to convert these resources into the \gls{EUD} (\ie the energy service required by the the final consumer), \ie more efficiency \cite{iea2020world,luderer2018residual}.
+Besides behavioral changes, an overall reshape of the energy system is necessary in terms of primary energy sources and technologies used to convert more efficiently these resources into the \gls{EUD} (\ie the energy service required by the the final consumer) \cite{iea2020world,luderer2018residual}.
 The former corresponds to a whole ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}) where energy carriers called, in the literature, ``\emph{biofuels}'', ``\emph{electrofuels}'', ``\emph{synthetic fuels}'', ``\emph{renewable fuels}'' or even ``\emph{sustainable fuels}'', will more and more play a crucial role.
-To avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions, we have suggested a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
+Even if the share of electricity increases in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during (and after) the energy transition \cite{Ahlgren2012}.
+To avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions, we have developed a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
 In the rest of this thesis, the electro - and bio - fuels are considered as renewable and with no \gls{GWP}.
@@
 
-In the general perspective to decrease the \gls{GWP} of the primary energy mix, \gls{VRES} like wind and solar, have already emerged as the keystone to defossilise the energy system.
+In the general perspective to decrease the \gls{GWP} of the primary energy mix, \gls{VRES}, like wind and solar, have already emerged as the keystone to defossilise\footnote{Hydrocarbons, currently produced from fossil resources, will still be composed of carbon in a renewable world.
+This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.} the energy system.
 However, their intermittency and space disparity could hold back their vaster integration in the future.
@@
 In practice, this electricity is mainly converted into hydrogen (\ie electrolysis) and then potentially upgraded into more complex fuels (\eg methane, methanol or ammonia).
-Even if the share of electricity increases in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during (and after) the energy transition \cite{Ahlgren2012}.
 They offer three main advantages: infrastructure compatibility, storage and capacity to link sectors (\ie from electricity to mobility, heat, or industry).
@@
 An example is carbon-free ammonia-hydrogen blends burned in spark ignition engines \cite{lhuillier2020experimental} or \gls{CHP} applications \cite{pochet202022}.
-With a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
+Moreover, some applications (\eg marine, aviation and heavy-duty transport) will be hard to electrify and keep on requiring high-density energy carriers \cite{horvath2018techno, brynolf2018}.
+Then, about storage, gas networks present much more storage potential than electrical network (\eg 50 times more in Germany and 300 times more in France) \cite{Rosa2017}.
+Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}).
+In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
+Finally, with a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
 Besides direct electrification of other sectors (\eg electrical heat pumps, \gls{BEV}), \citet{brown2018synergies} showed that converting power to hydrogen and methane was advantageous at high shares of renewables, in their optimisation of the European whole-energy system.
@@
 For instance, electricity produced in excess from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser - coupling the power and industry sectors \cite{verleysen2020can}.
-Gas networks present much more storage potential than electrical network (\eg 50 times more in Germany and 300 times more in France) \cite{Rosa2017}.
-Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}).
-In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
-Moreover, some applications (\eg marine, aviation and heavy-duty transport) will be hard to electrify and keep on requiring high-density energy carriers \cite{horvath2018techno, brynolf2018}.
-These carriers, currently produced from fossil resources, will still consist of hydrocarbons in a renewable world.
-This is why this thesis rather uses "defossilisation" rather than "decarbonisation" as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
 
@@
 \includegraphics[width=0.9\textwidth]{Storage_electrofuels.pdf}
-\caption{Energy carriers and technologies to store electricity. Electrofuels are an economical solution for high capacity and long-term storage of energy. Graph adapted from \cite{ISPT2017}.}
+\caption{Energy carriers and technologies to store electricity. Electrofuels provide a solution for high capacity and long-term storage of energy. Graph adapted from \cite{ISPT2017}.}
 \label{fig:intro:Storage_electrofuels}
@@
 
-To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a multi-sector and whole-energy system \cite{Contino2020}.
+To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a multi-sector and whole-energy system \cite{contino2020whole}.
 To reach this goal, an energy system optimisation model (ESOM) can optimise the design and the operation of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
 
-In this research field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most of ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
+In this research field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
 However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system, such as a country.
 Given the lifetime of the conversion technologies, such strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
-Besides the uncertainty on the model structure (not addressed in this work), this long-term and large-system optimisation motivates the need to account for \gls{UQ} and consider it as a major challenge of such models \cite{pfenninger2014energy}.
-This challenge, along with a large number (\ie more than a hundred) of uncertain parameters and limited information of their distribution, leads to the "curse of dimensionality" \cite{kuo2005lifting}, where the computational burden rapidly explodes with the number of considered uncertain parameters.
+This long-term and large-system optimisation motivates the need to account for \gls{UQ}, which is considered a major challenge of such models in the literature \cite{pfenninger2014energy}.
+This challenge, along with a large number (\ie more than a hundred) of uncertain parameters, leads to the "curse of dimensionality" \cite{kuo2005lifting}, where the computational burden rapidly explodes with the number of considered uncertain parameters.
 
@@
 In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there was a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
-On top of the ``what to do?'', this approach aims at helping the policymakers to answer the  question ``how to do it?''.
+On top of the ``what to do?'', this \gls{RL}-based framework aims at helping the policymakers to answer the  question ``how to do it?''.
 
-Finally, when assessing the robustness of transition pathways provided by \gls{ESOMs}, the literature shows a variety of techniques, \eg Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
-However, the robustness is commonly applied to assess the sensibility of the solution regarding the objective function to optimise (\ie the total cost).
-Given the complexity of whole-energy system models (\ie multiple sectors and multiple energy carriers) along with the time-scale of the transition and the lot of uncertainties, the total cost does not provide information on the sensibility of the design strategy, \ie the investment decisions.
-Where \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
+Finally, when assessing the robustness of transition pathways provided by ESOMs, the literature shows a variety of techniques, \eg Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
+However, the robustness is commonly applied to assess the sensibility of the solution regarding the objective function (\ie the total cost).
+Given the complexity of whole-energy system models (\ie multiple sectors and multiple energy carriers) along with the time-scale of the transition and the large number of uncertainties, the total cost does not provide information on the sensibility of the design strategy, \ie the investment decisions.
+For instance, the optimal total cost might need huge investments cost that are not possible in the beginning, and it is more preferable to balance between \gls{CAPEX} and \gls{OPEX}.
+\citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
 However, their work focused on the power sector only and for a target future year, \ie 2035.
@@
 First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for the pathway transition from an existing system to a long term target, \ie 2050.
-Compared to other models, EnergyScope Pathway introduces a rapid (\ie around 15~min on a personal laptop for the optimisation of a 30-year transition) computational optimisation tool for exploring diverse transition pathways within an entire energy system while maintaining temporal precision (\ie hourly time-resolution) to accurately capture the integration of \gls{VRES}.
 Based on this model originally implemented in a perfect foresight approach, \ie one overall optimisation for the whole pathway, we have developed the myopic method.
@@
 Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
-Although, these actions are taken every five years, they impact the system, for the next ten years - their time window.
+Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
 The intermediate solutions obtained in the middle of the time window are used as a new starting point for the agent that makes a new series of decisions for the next ten years, etc.
@@
 Chapter \ref{chap:case_study} presents the case study of this work, \ie Belgium and its energy transition.
-Without exhaustively detailing the data \cite{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the date related \gls{SMR} as an option to produce nuclear-based electricity in the future.
+Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the date related \gls{SMR} as an option to produce nuclear-based electricity in the future.
 
@@
 In this one, we detail the results of the \gls{GSA} carried out on the Belgian energy transition under the lens of the atom-molecules dilemma.
-On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear capacities by 2040 onward and the driving parameters on the import of electrofuels.
+On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear (``atom'') capacities by 2040 onward and the driving parameters on the import of electrofuels (``molecules'').
 
@@
 
-After detailing the robustness metric based on the \gls{PCA} approach, Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting of deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
+Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting of deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
 The first one, the reference case, considers nominal values for all the uncertain parameters.

### 2024-04-05 appendices/app_EnergyScope.tex
RobPol - Description case ROB
Intro post-review AR

@@
 
-The objective, Eq.~(\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
-The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau_app}).
+The objective, Equation (\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
+The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Equation (\ref{eq:tau_app}).
 
@@
 
-The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv_app}).
+The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Equation (\ref{eq:c_inv_app}).
 The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler.
-The total operation and maintenance costs (\textbf{C\textsubscript{maint}}) are calculated in the same way, Eq.~(\ref{eq:c_maint_app}).
+The total operation and maintenance costs (\textbf{C\textsubscript{maint}}) are calculated in the same way, Equation (\ref{eq:c_maint_app}).
 
@@
 
-The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over the different time-periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}).
-To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}).
+The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over the different time-periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Equation (\ref{eq:c_op_app}).
+To simplify the reading, we write the sum over typical days as $t\in T$ such as in Equation (\ref{eq:c_op_app}).
 The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}.
@@
 For climate change, the natural choice as indicator is the \acrlong{GWP}, expressed in ktCO$_2$-eq./year.
-In Eq.~(\ref{eq:GWP_tot_app}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}).
-The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Eq.~(\ref{eq:GWP_op_app}).
+In Equation (\ref{eq:GWP_tot_app}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}).
+The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Equation (\ref{eq:GWP_op_app}).
 Thus, this version accounts only for operation without accounting for the \gls{GWP} emitted during the construction of the technologies.
@@
 
-The above equations (Eqs.
-(\ref{eq:obj_func_app}) - (\ref{eq:GWP_op_app})) represent only a part of the formulation and illustrate the syntax that is used.
+The above equations (Equations (\ref{eq:obj_func_app}) - (\ref{eq:GWP_op_app})) represent only a part of the formulation and illustrate the syntax that is used.
 Those representing the energy balance, network implementation, sectors representation, etc.
@@
 \centering
-\includegraphics[width=\textwidth]{path_e.g._tech.png}
+\includegraphics[width=\textwidth]{path_e.g._tech.pdf}
 \caption{Example of how the technology capacities and associated variables are evolving.
@@
 
-Similarly to a mass balance, Eq.~(\ref{eq:F_newBuilt_app}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
+Similarly to a mass balance, Equation (\ref{eq:F_newBuilt_app}) is the technology capacity balance.
+The constraint forces the installation or withdrawing of capacities between two representative years:
 at the end of the phase (\emph{y\textsubscript{stop}}), the available capacity is the one used in the next representative year (\textbf{F}(\emph{y\textsubscript{stop}})).
@@
 As an example, a technology could be decommissioned before being built ($ \emph{p} < \emph{p\textsubscript{built}}$).
-Eqs.~(\ref{eq:F_decomNonPhysic1_app}-\ref{eq:Fold_def_app})  allow preventing these non-sense while keeping the formulation linear.
-Eq.~(\ref{eq:F_decomNonPhysic1_app}) forces the decommissioned capacity to zero when technology will be built after.
+Equations (\ref{eq:F_decomNonPhysic1_app}-\ref{eq:Fold_def_app})  allow preventing these non-sense while keeping the formulation linear.
+Equation (\ref{eq:F_decomNonPhysic1_app}) forces the decommissioned capacity to zero when technology will be built after.
 To do so, a parameter (\emph{decom\textsubscript{allowed}}) is defined \emph{a priori} and is equal to 0 or 1 when decommissioning is not possible or possible, respectively. 
-Eq.~(\ref{eq:Fold_def_app}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}.
+Equation (\ref{eq:Fold_def_app}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}.
 For each phase, a set (\emph{AGE}) is calculated \emph{a priori}.
@@
 In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built.
-The first part of Eq.~(\ref{eq:Fold_def_app}) indicates that the technology is still available, and thus no capacity needs to be removed.
+The first part of Equation (\ref{eq:Fold_def_app}) indicates that the technology is still available, and thus no capacity needs to be removed.
 The second part of the equation represents the capacity that reached its expected lifetime minus a part of the capacity that would have been decommissioned.
 As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
-The `if' in Eq. (\ref{eq:Fold_def_app}) is linear as it is applied to a parameter and not a variable.
+The `if' in Equation (\ref{eq:Fold_def_app}) is linear as it is applied to a parameter and not a variable.
 
@@
 To initialise the problem in 2020 with the existing design, an additional phase `\emph{2015\_2020}' is created.
-Eq.~(\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase.
+Equation (\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase.
 \\
@@
 
-Eq.~(\ref{eq:delta_tech_during_phase_app}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
+Equation (\ref{eq:delta_tech_during_phase_app}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
 %% GL_correction_v2 %% Following line added:
@@
 %% GL_correction_v2 %% end
-Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Eq.~(\ref{eq:limit_reno_LTheat_app}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Eq.~(\ref{eq:limit_reno_passmob_app}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Eq.~(\ref{eq:limit_reno_freight_app}).
+Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Equation (\ref{eq:limit_reno_LTheat_app}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Equation (\ref{eq:limit_reno_passmob_app}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Equation (\ref{eq:limit_reno_freight_app}).
 For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase.
@@
 
-As an extension of Eq. \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
-defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2_app}).
-The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2_app}), to which the residual asset value in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
+As an extension of Equation \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
+defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Equation (\ref{eq:obj_func_v2_app}).
+The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Equation (\ref{eq:Capex_v2_app}), to which the residual asset value in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}).
 Thus, the investments account for the installation and dismantlement costs of the technologies.
-The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Eq.~(\ref{eq:Copex_tot_v2_app}).
+The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Equation (\ref{eq:Copex_tot_v2_app}).
 During a phase, the system \gls{OPEX} is
-the product of the annualised phase factor, defined in Eq.~(\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
+the product of the annualised phase factor, defined in Equation (\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
 The annualised phase factor is defined based on an average interest rate during the transition.
@@
 
-For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Eq.~(\ref{eq:opex_yearly_app}).
+For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Equation (\ref{eq:opex_yearly_app}).
 
@@
 
-The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Eq.~(\ref{eq:PhaseInv_app}).
-The annualised phase factor  (defined by Eq.~(\ref{eq:path_annu_factor_app})) is used.
+The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Equation (\ref{eq:PhaseInv_app}).
+The annualised phase factor  (defined by Equation (\ref{eq:path_annu_factor_app})) is used.
 The specific cost during the phase is defined as  the average between the investment cost for the first and last year of the period.
@@
 In the previous example, the residual investment of the PV built is 75\%. 
-Eq.
-(\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost.
+Equation (\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost.
 The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised.
@@
 
-The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Eq.~(\ref{eq:gwp_tot_transition_app}).
+The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Equation (\ref{eq:gwp_tot_transition_app}).
 The emissions during a phase is estimated as the arithmetic average of the representative years before and after the phase.
-Eq.~(\ref{eq:limit_gwp_trans_app}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).
+Equation (\ref{eq:limit_gwp_trans_app}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).

### 2024-04-05 appendices/app_case_study.tex
Intro post-review AR

@@
 In particular, the private mobility drops to near-zero emissions at early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift.
-First, while accounting for the society inertia in this sector (see Eq.
-\ref{eq:limit_reno_passmob_app}), battery electric vehicles (BEV) substitute \gls{ICE} cars.
+First, while accounting for the society inertia in this sector (see Equation (\ref{eq:limit_reno_passmob_app})), battery electric vehicles (BEV) substitute \gls{ICE} cars.
 This reduces the overall emissions of the private vehicles as electric motors are more efficient than \gls{ICE} and they are driven by a less-emitting resource (\ie electricity versus diesel and gasoline).
@@
 A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technologies after 2050.
-Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Eq.
-(\ref{eq:salvage}).
+Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Equation (\ref{eq:salvage}).
 Out of the 114.4~b€\textsubscript{2015} of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime.
@@
 
-In general, due to the formulation of the salvage value (see Eq.
-\ref{eq:salvage}), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition.
+In general, due to the formulation of the salvage value (see Equation (\ref{eq:salvage})), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition.
 Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}.

### 2024-04-05 chap_RL/chap_RL.tex
Intro post-review AR

@@
 \includegraphics[width=0.7\textwidth]{Reward.pdf}
-\caption{Reward, $R$, function scheme.
+\caption{Reward function, $R$.
 Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded.
@@
 
-Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the training of the \gls{NN}.
-For this, numerous episodes are played through the myopic optimization of the transition pathway of Belgium.
+Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the learning of the \gls{NN}, also called ``training''.
+For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium.
 At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
@@
 
-To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase, also called ``training'', of the agent has been done on the monthly model.
+To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model.
 Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
+
+The learning phase has been split by batch of 500 steps.
+For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
+This makes sure replay buffer is full enough for useful updates.
+At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved.
+This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
+\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
+\caption{Mean reward and success rate of the different learning batches.
+The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view.
+The success rate also confirms this progression and convergence.}
+\label{fig:reward_success}
+\end{figure}
 
@@
 These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
-\label{fig:Reward}
+\label{fig:Binding_constr}
 \end{figure}

### 2024-04-05 chap_atom_mol/chap_atom_mol.tex
Intro post-review AR

@@
 \label{sec:atom_mol:results_uq}
-``\textit{It is difficult to make predictions, especially about the future.}" (Niels Bohr, foundational contributor to the understanding of atomic structure and quantum theory).
-Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters.
+``\textit{It is difficult to make predictions, especially about the future.}"\footnote{Niels Bohr, foundational contributor to the understanding of atomic structure and quantum theory} Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters.
 After briefly assessing the \gls{GSA} on the objective function of the model, the total transition cost, this section investigates more deeply the atom-molecules dilemma.

### 2024-04-05 chap_case_study/chap_case_study.tex
Intro post-review AR

@@
 This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world energy-related emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.
-Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in Eq.\,(\ref{eq:limit_gwp_trans}).
+Therefore, in this work, a limit has been put on $\emph{gwp\textsubscript{lim,trans}}=1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ in Equation\,(\ref{eq:limit_gwp_trans}).
 This is another sign of the urgency to act to mitigate climate change as this 30-year budget represents only 10 years of the current emissions.

### 2024-04-05 chap_intro_ccl/conclusion.tex
Intro post-review AR

@@
 \end{itemize}
+
+Quand je parle de ce que j'ai fait avec RL : Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
+Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
+The intermediate solutions obtained in the middle of the time window are used as a new starting point for the agent that makes a new series of decisions for the next ten years, etc.
+Repeating the whole transition with different sequences of actions-states allow the agent to come up with an optimised policy towards sustainability, considering the variation of the parameters of its environment.

### 2024-04-05 chap_intro_ccl/intro.tex
Intro post-review AR

@@
-Climate change, among other environmental challenges, is (mostly) due to the concentration of anthropogenic \acrfull{GHG} in the environment \cite{IPCC_CO2_budget}.
+Climate change, among other environmental changes, is mostly due to the concentration of anthropogenic \acrfull{GHG} in the environment \cite{IPCC_CO2_budget}.
 Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
-Among these emissions, globally, about 75\% are directly related to the whole-energy system \cite{ourworldindata_CO2_world}.
-The \gls{GHG} emissions, usually expressed in kt$_{\ce{CO2},\text{eq}}$, could be developed as an adapted, \ie less economy-oriented, version of the original Kaya identity \cite{kaya1997environment}:
+Energy consumption, mainly from fossil fuels, is the main contributor: it is responsible for 75\% of \gls{GHG} emissions, the remaining 25\% originating from agriculture and industrial processes \cite{ourworldindata_CO2_world}.
+With the perspective of the energy system, the Kaya identity \cite{kaya1997environment} can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
+\gls{EUD} is the energy service required by the the final consumer.
+In Equation (\ref{eq:equality_GHG}), the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
 
@@
 \label{eq:equality_GHG}
-\mathrm{GHG} =  \frac{\mathrm{GHG}}{\text{Primary energy}} \times \frac{\text{Primary energy}}{\mathrm{EUD}}\times \frac{\mathrm{EUD}}{\text{Population}} \times \text{Population,}\\
+\mathrm{GHG} =  \frac{\mathrm{GHG}}{\text{Primary energy}} \times \frac{\text{Primary energy}}{\mathrm{EUD}}\times \frac{\mathrm{EUD}}{\text{Population}} \times \text{Population}\\
  \end{equation}
 
+Such an identity is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
+Considering the expected growing trend of the Population, Equation (\ref{eq:equality_GHG}) highlights three levers of action that should be activated to reduce the \gls{GHG} emissions and favour the transition \cite{dodson2020population,scovronick2017impact}: renewables, efficiency and sufficiency.
+
+Focusing on the levers that are within grasp of engineering studies, this thesis addresses the two ``technical'' levers of the energy transition identified in Equation (\ref{eq:equality_GHG}): renewables and efficiency.
+This focus is aligned with the current European policies binding the Member States of the European Union.
+For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, states that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind, an increase in energy efficiency and significantly higher shares of energy from renewable sources in an integrated energy system''.% (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
+
+Despite the growing echo it finds in the scientific community \cite{o2018good}, sufficiency is lacking in the RED III.
+Explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, it is defined as ``\emph{a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people}'' \cite{lage2023citizens}.
+Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving sociology, psychology and political science, as it is outside the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
+One of the objectives of this manuscript is to provide support to such interdisciplinary projects in the assessment of sufficiency policy.
+\\
+
 \noindent
-where the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third stands as the energy intensity per capita.
-Such an identity, mathematically-correct though, is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
-However, Eq.
-\ref{eq:equality_GHG} has the merit to highlight three levers of action that should be activated to reduce the \gls{GHG} emissions and, consequently, favour the transition, considering that the population will continue to grow \cite{dodson2020population,scovronick2017impact}.
-These three levers of actions are: renewables, efficiency and sufficiency aiming at reducing the first, the second and the third terms on the right-hand side of Eq.
-\ref{eq:equality_GHG}, respectively.
-The latter, explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, is defined by \citet{lage2023citizens} as ``a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people''.
-Altough this finds a growing interest in the scientific community \cite{o2018good}, it requires, maybe more than the two other levers, interdisciplinarity \cite{schmidt2015interdisciplinary}, \ie the combination of multiple academic disciplines like sociology, psychology or politics, that are out of the scope of my expertise, and, consequently, this thesis.
-However, the work developed in the present manuscript aims at providing support to such interdisciplinary projects to assess sufficiency policies.
-More within the grasp of the engineering world, this thesis rather focuses on the first two terms of Eq.
-\ref{eq:equality_GHG}, \ie renewables and efficiency.
-This aligns with the current European policies binding the Member States of the European Union.
-For instance, the Renewable Energy Directive (RED) III, published in October 2023 \cite{REDIII}, highlights that ``the Union’s climate neutrality objective (by 2050) requires a just energy transition which leaves no territory or citizen behind\footnote{This directly relates to sufficiency as it encompasses social justice in parallel with a minimization of the energy use.}, an increase in energy efficiency and significantly higher shares of energy from renewable sources in an integrated energy system'' (\ie 42.5\% of the Union's gross final consumption of energy by 2030).
-
-\section*{Context and motivations}
-
 To ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
-Besides behavioral changes, an overall reshape of the energy system is necessary in terms of primary energy sources and technologies used to convert more efficiently these resources into the \gls{EUD} (\ie the energy service required by the the final consumer) \cite{iea2020world,luderer2018residual}.
-The former corresponds to a whole ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}) where energy carriers called, in the literature, ``\emph{biofuels}'', ``\emph{electrofuels}'', ``\emph{synthetic fuels}'', ``\emph{renewable fuels}'' or even ``\emph{sustainable fuels}'', will more and more play a crucial role.
-Even if the share of electricity increases in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during (and after) the energy transition \cite{Ahlgren2012}.
-To avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions, we have developed a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
-In the rest of this thesis, the electro - and bio - fuels are considered as renewable and with no \gls{GWP}.
+Similarly to the RED III, the transformations identified by \citet{iea2020world} focus on technical aspects, disregarding the sufficiency.
+The technical transformations that can be brought to the energy system concern both the primary energy sources and the technologies used to convert more efficiently these resources into the \gls{EUD} \cite{iea2020world,luderer2018residual}.
 
@@
 \includegraphics[width=\textwidth]{IEA_WEO_2019.png}
-\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario \cite{iea2020world}.}
+\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario of \citet{iea2020world}.}
 \label{fig:intro:IEA_WEO_2019}
@@
 
-In the general perspective to decrease the \gls{GWP} of the primary energy mix, \gls{VRES}, like wind and solar, have already emerged as the keystone to defossilise\footnote{Hydrocarbons, currently produced from fossil resources, will still be composed of carbon in a renewable world.
-This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.} the energy system.
-However, their intermittency and space disparity could hold back their vaster integration in the future.
-To address this issue, due to some limitations (\eg range, power, costs) of electricity-focused solutions like \gls{DC} lines, the transport and long-term storage of the renewable electricity produced in excess should be optimised.
-This challenge can be tackled by \textit{electrofuels} \cite{rozzi2020}.
+Changing primary energy sources corresponds to integrating more renewables.
+\gls{VRES} like wind and solar are the keystone to defossilise the energy system but they face challenges related to their intermittency and space disparity.
+To address this issue,  limitations like range, power or  costs make \gls{DC} lines or batteries not the most relevant solutions.
+The transport and long-term storage of the renewable electricity produced in excess should be optimised.
+This is why, to reach the objective of sustainable development, more renewables is associated with ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}).
+
+The fuel switch consists in replacing traditional fossil fuels with new energy carriers.
+Among them, \textit{electrofuels} are promising solutions to tackle the challenges related to renewable electricity \cite{rozzi2020}.
 These fuels represent energy carriers where electricity has the major share in the energy balance of the fuel.
-In practice, this electricity is mainly converted into hydrogen (\ie electrolysis) and then potentially upgraded into more complex fuels (\eg methane, methanol or ammonia).
-They offer three main advantages: infrastructure compatibility, storage and capacity to link sectors (\ie from electricity to mobility, heat, or industry).
+In practice, this electricity is mainly converted into hydrogen through electrolysis and then potentially upgraded into more complex fuels such as methane, methanol or ammonia.
+
+Electrofuels offer three main advantages: infrastructure compatibility, storage and capacity to link sectors (\ie from electricity to mobility, heat, or industry).
 Development on electrofuels aims at getting them more and more compatible with existing and mature technologies \cite{Ahlgren2012}.
-An example is carbon-free ammonia-hydrogen blends burned in spark ignition engines \cite{lhuillier2020experimental} or \gls{CHP} applications \cite{pochet202022}.
-Moreover, some applications (\eg marine, aviation and heavy-duty transport) will be hard to electrify and keep on requiring high-density energy carriers \cite{horvath2018techno, brynolf2018}.
-Then, about storage, gas networks present much more storage potential than electrical network (\eg 50 times more in Germany and 300 times more in France) \cite{Rosa2017}.
-Where batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses, electrofuels are an economical solution for high capacity (from 100 GWh) and long-term (\ie from months to years) storage of energy \cite{child2018role, dias2020energy} (see Figure \ref{fig:intro:Storage_electrofuels}).
-In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of the ambient \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
-Finally, with a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
-Besides direct electrification of other sectors (\eg electrical heat pumps, \gls{BEV}), \citet{brown2018synergies} showed that converting power to hydrogen and methane was advantageous at high shares of renewables, in their optimisation of the European whole-energy system.
-Electrofuels have the ability to couple energy and non-energy sectors \cite{Stancin2020}.
-For instance, electricity produced in excess from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser - coupling the power and industry sectors \cite{verleysen2020can}.
+An example is ammonia-hydrogen blends burned in spark ignition engines \cite{lhuillier2020experimental} or \gls{CHP} applications \cite{pochet202022}.
+Mobility applications such as marine, aviation and heavy-duty transport are hard to electrify and require high-density energy carriers \cite{horvath2018techno, brynolf2018}.
+
+Concerning storage,  batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses.
+On the contrary, electrofuels are promising solutions for high capacity (from 100 GWh) and long-term storage of energy \cite{child2018role, dias2020energy} (Figure \ref{fig:intro:Storage_electrofuels}).
+In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
+gas networks present much more storage potential than electrical network: 50 times more in Germany and 300 times more in France \cite{Rosa2017}.
 
@@
 
-To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a multi-sector and whole-energy system \cite{contino2020whole}.
+Finally, with a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
+Besides direct electrification of other sectors with electrical heat pumps, Battery Electric Vehicles (BEV),  converting power to hydrogen and methane is advantageous at high shares of renewables for the optimisation of the European energy system \cite{brown2018synergies}.
+Electrofuels have also the ability to couple energy and non-energy sectors \cite{Stancin2020}.
+For instance, electricity from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser, coupling the power and agro-industrial sectors \cite{verleysen2020can}.
+
+Even if the share of electricity will increase in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during, and after, the energy transition \cite{Ahlgren2012}.
+In other words, hydrocarbons, currently produced from fossil resources, will still be composed of carbon in a renewable world.
+This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
+
+Besides electrofuels, ``\emph{biofuels}'', ``\emph{synthetic fuels}'', ``\emph{renewable fuels}'' or even ``\emph{sustainable fuels}'' will play a key role in the transition.
+To avoid the confusion between all these ``alternative'' fuels and thereby reduce misunderstanding in political or academic discussions, we have developed a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
+In the rest of this thesis, the electro --- and bio --- fuels are considered as renewable with no associated \gls{GWP}.
+
+To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a whole-energy system which is multi-sector and account for multiple energy carriers \cite{contino2020whole}.
 To reach this goal, an energy system optimisation model (ESOM) can optimise the design and the operation of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
 
-In this research field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most ESOMs use a deterministic approach (\ie 75\% out of the 134 reviewed ESOM studies).
-However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system, such as a country.
-Given the lifetime of the conversion technologies, such strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
+In the field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most ESOMs use a deterministic approach (75\% of the 134 reviewed studies).
+However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system such as a country.
+Given the lifetime of energy conversion technologies, such strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
 This long-term and large-system optimisation motivates the need to account for \gls{UQ}, which is considered a major challenge of such models in the literature \cite{pfenninger2014energy}.
-This challenge, along with a large number (\ie more than a hundred) of uncertain parameters, leads to the "curse of dimensionality" \cite{kuo2005lifting}, where the computational burden rapidly explodes with the number of considered uncertain parameters.
+%This challenge, along with a large number of uncertain parameters (more than a hundred), leads to the "curse of dimensionality" \cite{kuo2005lifting}, where the computational burden rapidly explodes with the number of considered uncertain parameters.
 
 On top of dealing with these uncertainties, the reality of decision-makers is a limited foresight into the future \cite{poncelet2016myopic}.
-In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'', \ie 2050 (and beyond), and make the planning decisions once and for all accordingly.
-On the contrary, they uncover the realisation of these uncertainties step-by-step (\ie in a ``myopic'' way), and progressively act on them to, hopefully, meet the set target to mitigate the climate change.
+In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'' in 2050 and make the planning decisions once and for all accordingly.
+On the contrary, they uncover the realisation of these uncertainties step-by-step, in a ``myopic'' way, and progressively act on them to, hopefully, meet the set target to mitigate the climate change.
 In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there was a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
@@
 
-Finally, when assessing the robustness of transition pathways provided by ESOMs, the literature shows a variety of techniques, \eg Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
-However, the robustness is commonly applied to assess the sensibility of the solution regarding the objective function (\ie the total cost).
-Given the complexity of whole-energy system models (\ie multiple sectors and multiple energy carriers) along with the time-scale of the transition and the large number of uncertainties, the total cost does not provide information on the sensibility of the design strategy, \ie the investment decisions.
-For instance, the optimal total cost might need huge investments cost that are not possible in the beginning, and it is more preferable to balance between \gls{CAPEX} and \gls{OPEX}.
+The robustness of transition pathways provided by ESOMs is an essential question due to their sensitivity to uncertain factors.
+A variety of techniques are used to assess this robustness including Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
+However, the robustness is commonly assessed to evaluate the sensibility of the total cost (objective function) to input parameters.
+Given the complexity of whole-energy system models, the time scale of the transition and the large number of uncertainties, only considering the total cost does not give information on the sensibility of the design strategy, \ie the investment decisions.
+For instance, total cost might consist in excessive upfront investments cost and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable depsite a slightly higher total cost.
 \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
-However, their work focused on the power sector only and for a target future year, \ie 2035.
-The literature was missing an approach to tackle these two challenges: to go beyond the total transition cost encompassing the details of a solution into a single value while assessing the whole-energy system over its whole transition.
+However, their work focused on the power sector only and for a target future year (2035).
+Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost that and provide more details on the sensibility of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
 
-\section*{Objectives and tasks}
-``\textit{Our task is not to foresee the future, but to enable it.}'' Saint-Exupéry in Citadelle, 1948.
-In that sense, this thesis aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
+\noindent
+``\textit{Our task is not to foresee the future, but to enable it}''\footnote{Saint-Exupéry in Citadelle, 1948}.
+This thesis aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
 Rather than trying, in vain, to answer the question ``What could possibly happen in the future?'', this work rather addresses the ``What could or should we do to make the future possible?''.
-Given these general context and motivations, the research questions are described as follows:
+In that spirit, the research questions are described as follows:
 \begin{itemize}
-\item What is the role of electrofuels in the transition pathway of a whole-energy system subject to uncertainties, limited foresight into the future and a \ce{CO2}-budget?
+\item What can be the role of electrofuels in the transition pathway of a whole-energy system subject to uncertainties, limited foresight into the future and a \ce{CO2}-budget?
 What are the key uncertainties driving their import?
@@
 To answer these questions, several tasks have been carried out on the whole-energy system model accounting for the uncertainties, the method to explore the myopic pathway of such a system and the approach to assess the robustness of a pathway roadmap.
-All this work has been applied to the case of Belgium, a densely populated country with a limited amount of renewable energies, which represents roughly one third of the forecast energy demand \cite{Limpens2020}.
-This makes it a challenging case to go from a highly fossil-dominated system in 2020 (\ie 73\% of the primary energy mix \cite{spf_economy_2022}) to carbon-neutrality by 2050.
 
@@
 First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for the pathway transition from an existing system to a long term target, \ie 2050.
-Based on this model originally implemented in a perfect foresight approach, \ie one overall optimisation for the whole pathway, we have developed the myopic method.
-In this, the whole time horizon (\ie 30 years) is optimised through a sequence of 10-year long time-windows having a 5-year overlap between each other.
-To address the question about the role of electrofuels, we have detailed further on their implementation in the model to account for four main ones: hydrogen, methane, ammonia and methanol.
-Moreover, given their current and (expected) future role into the sector of the \gls{NED}, we have implemented this sector with a similar level of detail as the other sectors of the system, \ie electricity, heat and transport.
+Based on this model originally implemented in a perfect foresight approach, we have developed the myopic method in which the whole time horizon (30 years) is optimised through a sequence of 10-year time-windows with 5-year overlaps.
 
-To account for uncertainties, we have used, and adapted to our case study, the work of \citet{Moret2017PhDThesis} and \citet{coppittersthesis} on the uncertainty characterisation and quantification, respecitvely.
-In his thesis, \citet{Moret2017PhDThesis} developed a framework to obtain uncertainty range for a variety of parameters like cost of purchasing and availability of resources or investment cost and efficiency of technologies.
+To address the question of the role of electrofuels, they have been integrated in the pre-existing model, focusing on four molecules: hydrogen, methane, ammonia and methanol.
+Moreover, given their current and (expected) future role into the sector of the \gls{NED}, we have added to the model this sector with a similar level of detail as the other sectors of the system, \ie electricity, heat and transport.
+
+To account for uncertainties, uncertainty characterisation and quantification have been applied by adapting the works of \citet{Moret2017PhDThesis} and \citet{coppittersthesis}, respecitvely.
+In his thesis, \citet{Moret2017PhDThesis} developed a framework to obtain uncertainty ranges for a variety of parameters such as cost of purchasing and availability of resources or investment cost and efficiency of technologies.
 These ranges have then been sampled and propagated through the EnegyScope Pathway via the RHEIA framework developed by \citet{coppittersthesis}.
-Using the surrogate-modeling approach called \gls{PCE}, this framework allows identifying the uncertain parameters with the biggest impact on the variation of total transition cost or other outputs of interest like the imported amount of electrofuels.
+Using the surrogate modeling approach called \gls{PCE}, this framework allows identifying the uncertain parameters with the biggest impact on the variation of total transition cost or other outputs of interest such as the imported amount of electrofuels.
 
-To explore the different pathway trajectories in this myopic optimisation process, we have applied the \gls{RL} method where an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie sequence of actions to take to support the transition.
-Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
-Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
-The intermediate solutions obtained in the middle of the time window are used as a new starting point for the agent that makes a new series of decisions for the next ten years, etc.
-Repeating the whole transition with different sequences of actions-states allow the agent to come up with an optimised policy towards sustainability, considering the variation of the parameters of its environment.
+To explore the different pathway trajectories in this myopic optimisation process, we have applied the \gls{RL} method.
+In this branch of machine learning, an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie sequence of actions to take to support the transition.
 
@@
 
-\section*{Outline}
+All this work has been applied to the case of Belgium, a densely populated country with a limited amount of renewable energies, which represents roughly one third of the forecast energy demand \cite{Limpens2020}.
+This makes it a challenging case to go from a highly fossil-dominated system in 2020 (73\% of the primary energy mix \cite{spf_economy_2022}) to carbon-neutrality by 2050.\\
+
+\noindent
 This thesis is composed of five chapters to provide answers to the different research questions.
-Chapter \ref{chap:chap_methodo} brings more details about the different methodological aspects of this work.
+Chapter \ref{chap:chap_methodo} details about the tools and methods used in this work.
 It starts with the main constraints, parameters and variables of the whole-energy system optimisation model, EnergyScope Pathway.
@@
 
-Chapter \ref{chap:case_study} presents the case study of this work, \ie Belgium and its energy transition.
+Chapter \ref{chap:case_study} presents the case study of this work: Belgium and its energy transition.
 Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the date related \gls{SMR} as an option to produce nuclear-based electricity in the future.
 
-Chapter \ref{chap:atom_mol} is the first of the three chapters presenting results.
-In this one, we detail the results of the \gls{GSA} carried out on the Belgian energy transition under the lens of the atom-molecules dilemma.
-On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear (``atom'') capacities by 2040 onward and the driving parameters on the import of electrofuels (``molecules'').
+Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} carried out on the Belgian energy transition under.
+On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear (``atom'') capacities by 2040 onward and the driving parameters on the import of electrofuels (``molecules''): the atom-molecules dilemma.
 

### 2024-04-05 chap_methodo/chap_methodo.tex
Intro post-review AR

@@
 
-The objective function of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as the sum of the total \gls{CAPEX}, \textbf{C\textsubscript{tot,capex}}, and the \gls{OPEX}, \textbf{C\textsubscript{tot,opex}} (see Eq.
-\ref{eq:obj_func_v2}).
+The objective function of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as the sum of the total \gls{CAPEX}, \textbf{C\textsubscript{tot,capex}}, and the \gls{OPEX}, \textbf{C\textsubscript{tot,opex}} (see Equation (\ref{eq:obj_func_v2})).
 
@@
 
-The total \gls{CAPEX} is the difference between the total investments done during the transition, $\textbf{C\textsubscript{inv,phase}}$, and the residual value of the assets that would still be in place after the end of the transition, $\textbf{C\textsubscript{inv,return}}$ (see Eq.
-\ref{eq:CAPEX_v2}).
+The total \gls{CAPEX} is the difference between the total investments done during the transition, $\textbf{C\textsubscript{inv,phase}}$, and the residual value of the assets that would still be in place after the end of the transition, $\textbf{C\textsubscript{inv,return}}$ (see Equation  (\ref{eq:CAPEX_v2})).
 
@@
 \noindent
-where $\textbf{C\textsubscript{inv,phase}}$ is given as the sum over all the  newly installed technologies at the phase $p$, $\textbf{F\textsubscript{new}}(p)$, multiplied by the average investment costs, $\emph{c\textsubscript{inv}}$,  of the year starting and ending the corresponding phase, $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$ respectively (see Eq.
-\ref{eq:PhaseInv}).
+where $\textbf{C\textsubscript{inv,phase}}$ is given as the sum over all the  newly installed technologies at the phase $p$, $\textbf{F\textsubscript{new}}(p)$, multiplied by the average investment costs, $\emph{c\textsubscript{inv}}$,  of the year starting and ending the corresponding phase, $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$ respectively (see Equation (\ref{eq:PhaseInv})).
 
@@
 \noindent where $\tau\textsubscript{\emph{phase}}$ is the annualised phase factor and $\emph{c\textsubscript{inv}}(p,j)$ is the arithmetic average of the investment cost of the technology $j$ at the beginning and the end of the phase, $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j)$ and $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)$.
-Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed (see Eq.
-\ref{eq:salvage})
+Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed (see Equation (\ref{eq:salvage}))
 
@@
 
-About the total \gls{OPEX} of the transition, $\textbf{C\textsubscript{tot,opex}}$, on top of the initial costs in 2020, we assume that the \gls{OPEX} of a phase is equal to the average operational costs, $ \textbf{C\textsubscript{opex}}$,  of $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$, multiplied by the duration of a phase $\emph{t\textsubscript{phase}}$ equal to 5 years in our case (see Eq.
-\ref{eq:Copex_tot_v2}).
+About the total \gls{OPEX} of the transition, $\textbf{C\textsubscript{tot,opex}}$, on top of the initial costs in 2020, we assume that the \gls{OPEX} of a phase is equal to the average operational costs, $ \textbf{C\textsubscript{opex}}$,  of $\emph{y\textsubscript{start}}$ and $\emph{y\textsubscript{stop}}$, multiplied by the duration of a phase $\emph{t\textsubscript{phase}}$ equal to 5 years in our case (see Equation (\ref{eq:Copex_tot_v2})).
 
@@
 where $\textbf{C\textsubscript{opex}}(p)$ is the arithmetic average of the operational cost at the beginning and the end of the phase $p$,  $\textbf{C\textsubscript{opex}}(y\textsubscript{start})$ and $\textbf{C\textsubscript{opex}}(y\textsubscript{stop})$.
-The operational cost of a year, $y$, $\textbf{C\textsubscript{opex}} (y)$, is the sum of the costs related to maintenance and operation of technologies, $ \textbf{C\textsubscript{maint}}$, and the consumption of resources, $\textbf{C\textsubscript{op}}$ (see Eq.
-\ref{eq:opex_yearly}).
+The operational cost of a year, $y$, $\textbf{C\textsubscript{opex}} (y)$, is the sum of the costs related to maintenance and operation of technologies, $ \textbf{C\textsubscript{maint}}$, and the consumption of resources, $\textbf{C\textsubscript{op}}$ (see Equation (\ref{eq:opex_yearly})).
 
@@
 
-he \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Eq.
-\ref{eq:gwp_tot_transition}).
-Similarly to initial operational costs to account for the system in place in 2020 (see Eq.
-\ref{eq:Copex_tot_v2}), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
-Then, as detailed in Section \ref{sec:cs:CO2-budget}, these cumulative emissions is constrained by a budget (see Eq.
-\ref{eq:limit_gwp_trans}).
+he \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Equation (\ref{eq:gwp_tot_transition})).
+Similarly to initial operational costs to account for the system in place in 2020 (see Equation (\ref{eq:Copex_tot_v2})), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
+Then, as detailed in Section \ref{sec:cs:CO2-budget}, these cumulative emissions is constrained by a budget (see Equation (\ref{eq:limit_gwp_trans})).
 
@@
 This remark stays valid for any time window except the first one of the transition where the year 2020 is optimized even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
-Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Eq.
-\ref{eq:salvage}.\\
+Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Equation (\ref{eq:salvage}).\\
 
@@
 These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset.
-This is mostly due to impact of the salvage value, Eq.
-\ref{eq:salvage}, in the objective function.
+This is mostly due to impact of the salvage value, Equation (\ref{eq:salvage}), in the objective function.
 Since this is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at early stages of the transition.
@@
 \noindent
-The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie the value of the objective function, see Eq.
-\ref{eq:obj_func_v2}) in the EnergyScope Pathway model ($M$).
+The PCE model ($\hat{M}$) is a representation of the relationship between the input parameters and the output variable of interest (\ie the value of the objective function, see Equation (\ref{eq:obj_func_v2})) in the EnergyScope Pathway model ($M$).
 This representation is constructed as a truncated series of multivariate orthonormal polynomials $\bm{\Psi}$, weighted by coefficients $u$:
@@
 If the \gls{LOO} error is below a specified threshold ($\sim 1\%$), the corresponding polynomial order is considered sufficient for generating an accurate PCE.
-However, if the error exceeds the threshold, the order is increased, and additional samples are generated following the rule of Eq.\,(\ref{eq:pce:nterms}).
-
-For the specific study of this work, a polynomial order of 2 is necessary (with 1260 training samples as per Eq.\,(\ref{eq:pce:nterms})) to achieve a \gls{LOO} error below \SI{1}{\%} for the total transition cost.
+However, if the error exceeds the threshold, the order is increased, and additional samples are generated following the rule of Equation\,(\ref{eq:pce:nterms}).
+
+For the specific study of this work, a polynomial order of 2 is necessary (with 1260 training samples as per Equation\,(\ref{eq:pce:nterms})) to achieve a \gls{LOO} error below \SI{1}{\%} for the total transition cost.
 
@@
 
-to rank parameters among each other. In (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.
+to rank parameters among each other. In Equation (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.
 
@@
 To overcome these two challenges, we needed to chose between an ``on-policy'' or ``off-policy'' algorithm.
-In a nutshell, in on-policy learning, the agent learns the value function or policy based on the data it generates by following its current policy whereas, in off-policy, the agent can learn from data collected by any policy, not just the one it is currently following, which provides greater flexibility and potential for reusing data.
+In a nutshell, in on-policy learning, the agent learns the value function or policy based on the data it generates by following its current policy whereas, in off-policy, the agent can learn from data collected by any policy, not just the one it is currently following, which provides greater flexibility and potential for reusing data stored in the so-called replay buffer.
 This makes off-policy algorithms more data efficient and ensuring better exploration by reusing past experiences or even following random exploration \cite{haarnoja2018soft}.
@@
 
-\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented\footnote{The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Eq.
-\ref{eq:SAC_objective}.} objective function (see Eq.
-\ref{eq:SAC_objective}).
+\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented\footnote{The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Equation (\ref{eq:SAC_objective}).} objective function (see Equation (\ref{eq:SAC_objective})).
 Where entropy represents the amount of energy in a system not available to produce work in thermodynamics, this term, also called Shannon entropy in the \gls{RL} context, stands for the randomness or stochasticity of the policy.
@@
 \noindent
-To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Eq.
-\ref{eq:F_newBuilt}).
-Even though these represent only the design part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Eq.
-\ref{eq:c_inv}) and, more indirectly, the resources to use.
+To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})).
+Even though these represent only the design part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:c_inv})) and, more indirectly, the resources to use.
 In other words, it captures the technological landscape of the transition.
@@
 Scaling the original raw data via normalization, \ie reducing data to $[0, 1]$ interval has a double purposes: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
-Consequently, the first part of this data preprocessing consists in scaling the installed capacities versus their respective sector and representative year (see Eq.
-\ref{eq:PCA_scaling_0}).
+Consequently, the first part of this data preprocessing consists in scaling the installed capacities versus their respective sector and representative year (see Equation (\ref{eq:PCA_scaling_0})).
 The sectors, as defined in EnergyScope, are the electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{NED}, storage and infrastructures.
@@
 Finally, to compare apples with apples, we converted the \gls{EUD} in the mobility sectors, \ie passenger and freight, into the \gls{FEC} they require in the REF case.
-This gives the second weighing factor to scale data, on top of the one of Eq.
-\ref{eq:PCA_scaling_0} (see Eq.
-\ref{eq:PCA_scaling_1}).
+This gives the second weighing factor to scale data, on top of the one of Equation (\ref{eq:PCA_scaling_0}) (see Equation (\ref{eq:PCA_scaling_1})).
 
@@
 Indeed, as detailed in Section \ref{subsec:meth:PCA:PCA}, a PC represents a vector for which the components are related to each variable of interest.
-Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Eq.
-\ref{eq:cos_similarity}).
+Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Equation (\ref{eq:cos_similarity})).
 It is important to consider to second specific case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors.

### 2024-04-05 options/symbols_and_cmd.tex
RobPol - Description case ROB
Intro post-review AR

--- 2024-02-16 options/symbols_and_cmd.tex
@@
 %% A
-\newabbreviation[group={Acronyms}]{API}{API}{application programming interface}
+\newabbreviation[group={Acronyms}]{API}{API}{Application Programming Interface}
 
 %% B
-\newabbreviation[group={Acronyms}]{BECCS}{BECCS}{bioenergy with carbon capture and storage}
-\newabbreviation[group={Acronyms}]{BEV}{BEV}{battery electric vehicle}
-\newabbreviation[group={Acronyms}]{BTX}{BTX}{benzene, toluene and xylene}
+\newabbreviation[group={Acronyms}]{BECCS}{BECCS}{Bioenergy with Carbon Capture and Storage}
+\newabbreviation[group={Acronyms}]{BEMS}{BEMS}{Building Energy Management System}
+\newabbreviation[group={Acronyms}]{BEV}{BEV}{Battery Electric Vehicle}
+\newabbreviation[group={Acronyms}]{BTX}{BTX}{Benzene, Toluene and Xylene}
 
 %% C
-\newabbreviation[group={Acronyms}]{CAPEX}{CAPEX}{capital expenditure}
-\newabbreviation[group={Acronyms}]{CCGT}{CCGT}{combined cycle gas turbine}
-\newabbreviation[group={Acronyms}]{CCS}{CCS}{carbon capture and storage}
-\newabbreviation[group={Acronyms}]{CHP}{CHP}{combined heat and power}
-\newabbreviation[group={Acronyms}]{CNG}{CNG}{compressed natural gas}
+\newabbreviation[group={Acronyms}]{CAPEX}{CAPEX}{Capital Expenditure}
+\newabbreviation[group={Acronyms}]{CCGT}{CCGT}{Combined Cycle Gas Turbine}
+\newabbreviation[group={Acronyms}]{CCS}{CCS}{Carbon Capture and Storage}
+\newabbreviation[group={Acronyms}]{CHP}{CHP}{Combined Heat and Power}
+\newabbreviation[group={Acronyms}]{CNG}{CNG}{Compressed Natural Gas}
 
 %% D
-\newabbreviation[group={Acronyms}]{DHN}{DHN}{district heating network}
+\newabbreviation[group={Acronyms}]{DC}{DC}{Direct Current}
+\newabbreviation[group={Acronyms}]{DHN}{DHN}{District Heating Network}
+\newabbreviation[group={Acronyms}]{DNN}{DNN}{Deep Neural Network}
+\newabbreviation[group={Acronyms}]{DRL}{DRL}{Deep Reinforcement Learning}
 
 %% E
-\newabbreviation[group={Acronyms}]{ESOMs}{ESOMs}{energy system optimisation models}
-\newabbreviation[group={Acronyms}]{ESTD}{EnergyScope TD}{EnergyScope Typical Days}
-\newabbreviation[group={Acronyms}]{EUD}{EUD}{end-use demand}
+\newabbreviation[group={Acronyms}]{ESOMs}{ESOMs}{Energy System Optimisation Models}
+\newabbreviation[group={Acronyms}]{EUD}{EUD}{End-Use Demand}
 
 %% F
-\newabbreviation[group={Acronyms}]{FC}{FC}{fuel cell}
+\newabbreviation[group={Acronyms}]{FC}{FC}{Fuel Cell}
+\newabbreviation[group={Acronyms}]{FEC}{FEC}{Final Energy Consumed}
 
 %% G
-\newabbreviation[group={Acronyms}]{GDP}{GDP}{gross domestic product}
-\newabbreviation[group={Acronyms}]{GHG}{GHG}{greenhouse gases}
-\newabbreviation[group={Acronyms}]{GSA}{GSA}{global sensitivity analysis}
-\newabbreviation[group={Acronyms}]{GWP}{GWP}{global warming potential}
+\newabbreviation[group={Acronyms}]{GDP}{GDP}{Gross Domestic Product}
+\newabbreviation[group={Acronyms}]{GHG}{GHG}{Greenhouse Gases}
+\newabbreviation[group={Acronyms}]{GSA}{GSA}{Global Sensitivity Analysis}
+\newabbreviation[group={Acronyms}]{GWP}{GWP}{Global Warming Potential}
 
 %% H
-\newabbreviation[group={Acronyms}]{HP}{HP}{heat pump}
-\newabbreviation[group={Acronyms}]{HT}{HT}{high-temperature}
-\newabbreviation[group={Acronyms}]{HVC}{HVC}{high value chemicals}
+\newabbreviation[group={Acronyms}]{HP}{HP}{Heat Pump}
+\newabbreviation[group={Acronyms}]{HT}{HT}{High-Temperature}
+\newabbreviation[group={Acronyms}]{HVC}{HVC}{High-Value Chemicals}
 
 %% I
-\newabbreviation[group={Acronyms}]{ICE}{ICE}{internal combustion engine}
+\newabbreviation[group={Acronyms}]{IAMs}{IAMs}{Integrated Assessment Models}
+\newabbreviation[group={Acronyms}]{ICE}{ICE}{Internal Combustion Engine}
 \newabbreviation[group={Acronyms}]{IEA}{IEA}{International Energy Agency}
-\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{intergovernmental panel for climate change}
-\newabbreviation[group={Acronyms}]{IQR}{IQR}{interquatile range}
+\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{Intergovernmental Panel on Climate Change}
+\newabbreviation[group={Acronyms}]{IQR}{IQR}{Interquatile Range}
 
 %% L
-\newabbreviation[group={Acronyms}]{LCA}{LCA}{life cycle assessment}
-\newabbreviation[group={Acronyms}]{LCOE}{LCOE}{levelised cost of energy}
-\newabbreviation[group={Acronyms}]{LFO}{LFO}{light fuel oil}
-\newabbreviation[group={Acronyms}]{LOO}{LOO}{leave-one-out}
-\newabbreviation[group={Acronyms}]{LPG}{LPG}{liquefied petroleum gas}
-\newabbreviation[group={Acronyms}]{LT}{LT}{low-temperature}
+\newabbreviation[group={Acronyms}]{LCA}{LCA}{Life Cycle Assessment}
+\newabbreviation[group={Acronyms}]{LCOE}{LCOE}{Levelised Cost of Energy}
+\newabbreviation[group={Acronyms}]{LFO}{LFO}{Light Fuel Oil}
+\newabbreviation[group={Acronyms}]{LOO}{LOO}{Leave-One-Out}
+\newabbreviation[group={Acronyms}]{LP}{LP}{Linear Programming}
+\newabbreviation[group={Acronyms}]{LPG}{LPG}{Liquefied Petroleum Gas}
+\newabbreviation[group={Acronyms}]{LT}{LT}{Low-Temperature}
 
 %% M
-\newabbreviation[group={Acronyms}]{MDP}{MDP}{Markov decision process}
+\newabbreviation[group={Acronyms}]{MDP}{MDP}{Markov Decision Process}
 \newabbreviation[group={Acronyms}]{MMSA}{MMSA}{Methanol Market Services Asia}
-\newabbreviation[group={Acronyms}]{MTBE}{MTBE}{methyl tert-butyl ether}
-\newabbreviation[group={Acronyms}]{MTO}{MTO}{methanol-to-olefins}
+\newabbreviation[group={Acronyms}]{MTBE}{MTBE}{Methyl Tert-butyl Ether}
+\newabbreviation[group={Acronyms}]{MTO}{MTO}{Methanol-to-olefins}
 
 %% N
-\newabbreviation[group={Acronyms}]{NED}{NED}{non-energy demand}
-\newabbreviation[group={Acronyms}]{NG}{NG}{fossil gas}
-\newabbreviation[group={Acronyms}]{NN}{NN}{neural network}
-\newabbreviation[group={Acronyms}]{NRE}{NRE}{non-renewable energy}
-\newabbreviation[group={Acronyms}]{NSC}{NSC}{naphtha steam cracker}
+\newabbreviation[group={Acronyms}]{NED}{NED}{Non-energy Demand}
+\newabbreviation[group={Acronyms}]{NG}{NG}{Fossil Gas}
+\newabbreviation[group={Acronyms}]{NN}{NN}{Neural Network}
+\newabbreviation[group={Acronyms}]{NRE}{NRE}{Non-renewable Energy}
+\newabbreviation[group={Acronyms}]{NSC}{NSC}{Naphtha Steam Cracker}
 
 %% O
-\newabbreviation[group={Acronyms}]{OPEX}{OPEX}{operational expenditure}
+\newabbreviation[group={Acronyms}]{OPEX}{OPEX}{Operational Expenditure}
 
 %% P
-\newabbreviation[group={Acronyms}]{PC}{PC}{principal component}
+\newabbreviation[group={Acronyms}]{PC}{PC}{Principal Component}
+\newabbreviation[group={Acronyms}]{PCs}{PCs}{Principal Components}
 \newabbreviation[group={Acronyms}]{PCA}{PCA}{Principal Component Analysis}
 \newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
-\newabbreviation[group={Acronyms}]{PDF}{PDF}{probability density function}
-\newabbreviation[group={Acronyms}]{PV}{PV}{photovoltaic}
+\newabbreviation[group={Acronyms}]{PDF}{PDF}{Probability Density Function}
+\newabbreviation[group={Acronyms}]{PV}{PV}{Photovoltaic}
 
 %% R
-\newabbreviation[group={Acronyms}]{RE}{RE}{renewable energy}
-\newabbreviation[group={Acronyms}]{RL}{RL}{reinforcement learning}
+\newabbreviation[group={Acronyms}]{RE}{RE}{Renewable Energy}
+\newabbreviation[group={Acronyms}]{RL}{RL}{Reinforcement Learning}
 
@@
 \newabbreviation[group={Acronyms}]{SAC}{SAC}{Soft Actor Critic}
-\newabbreviation[group={Acronyms}]{SDGs}{SDGs}{Sustainable Development Goals}
-\newabbreviation[group={Acronyms}]{SMR}{SMR}{small modular reactor}
-\newabbreviation[group={Acronyms}]{SVD}{SVD}{singular value decomposition}
+\newabbreviation[group={Acronyms}]{SMR}{SMR}{Small Modular Reactor}
+\newabbreviation[group={Acronyms}]{SVD}{SVD}{Singular Value Decomposition}
+
+%% S
+\newabbreviation[group={Acronyms}]{TRL}{TRL}{Technology Readiness Level}
 
@@
 %% V
-\newabbreviation[group={Acronyms}]{VRES}{VRES}{variable renewable energy sources}
+\newabbreviation[group={Acronyms}]{VRES}{VRES}{Variable Renewable Energy Sources}
 

### 2024-04-08 chap_RL/chap_RL.tex
Chap RL - Reward and success

@@
 
-\section{Learning process and convergence}
+\section{Convergence and learning process}
 \label{sec:RL:learning}
@@
 To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model.
-Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
+Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
+\\
 
-The learning phase has been split by batch of 500 steps.
+\myparagraph{Reward and success}\\
+
+The learning phase has been split into batches of 500 steps.
 For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
@@
 This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).
+The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal.
+Right hand side of Figure \ref{fig:reward_success} shows the success rate as the share of transitions meeting the \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) over the total number of attempted transitions, \ie episodes.
+Even though this success rate is not what drives the agent's optimisation, it shows that the shape of the reward (see Figure \ref{fig:Reward}) leads towards more and more successes.
 
@@
 The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view.
-The success rate also confirms this progression and convergence.}
+The evolution of the success rate also shows the shape of the reward aims at more and more successful transitions.}
 \label{fig:reward_success}
@@
 
+Over the entire learning process, 23\% of the episodes have been successful versus 73\% of failures, the rest being infeasible problems where the actions taken by the agent are too constraining for the model (see Section \ref{sec:RL:act_states_rew}).
+Besides these infeasible cases, there is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}).
+This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive.
+Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive.
+This is another indication that the shape of the reward is appropriate in this exploration of transition pathways.
+
+Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year''.
+Beyond this point, the chances to succeed the transition are 47\%.
+In other words, mid-term actions are necessary to hope succeeding the transition.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.49\textwidth]{Reward_status.pdf}
+\includegraphics[width=0.49\textwidth]{Reward_status_2.pdf}
+\caption{Reward distribution between successes and failures.
+Right hand side details at the end of which time-window the failure occurred.
+The ``tipping year'' is 2040 as failing the transition by 2040 represents 58\% of all the failures.
+Beyond this point, succeeding the transition represents 47\% of the cases.}
+\label{fig:reward_status}
+\end{figure}
+
+\myparagraph{States: Renewables and efficiency}\\
+
 \begin{itemize}
-\item Convergence of reward or success rate
 \item Actions to take, which ones are binding

### 2024-04-10 appendices/app_EnergyScope.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

@@
 As an example, the investment cost (\emph{c}\textsubscript{\emph{inv}}) of solar photovoltaic panels could drastically vary in the next decades (e.g.
-data used ranges between 1220 to 870  [€\textsubscript{2015}/kW] between 2020 and 2035).
+data used ranges between 1220 to 870  [€/kW] between 2020 and 2035).
 \\
@@
 To optimise the energy system, two key metrics must be adapted: the transition cost and the total \acrfull{GWP}. 
-Concerning the first one, all costs are expressed in €\textsubscript{2015} and an annualisation factor is used to distinguish investments over the transition.
+Concerning the first one, all costs are expressed in €\textsubscript{2020} and an annualisation factor is used to distinguish investments over the transition.
 For the \gls{GWP}, the metric used is based on the contributions of the gases over 100 years.

### 2024-04-10 appendices/app_case_study.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

@@
 
-In the following paragraphs, the results are analysed from a cost perspective to decipher the choices made by the model, as the overall cost of the transition is 1004~b€\textsubscript{2015} split unequally among the sectors.
+In the following paragraphs, the results are analysed from a cost perspective to decipher the choices made by the model, as the overall cost of the transition is 1004~b€ split unequally among the sectors.
 
-Figure \ref{fig:pestd_cumul_inv} illustrates the cumulative investments made throughout the transition, amounting to a total of 377.8~b€\textsubscript{2015}.
-This makes 12.6~b€\textsubscript{2015} on average per year, which represents 2.1\% of the current Belgian \gls{GDP}.
+Figure \ref{fig:pestd_cumul_inv} illustrates the cumulative investments made throughout the transition, amounting to a total of 377.8~b€.
+This makes 12.6~b€ on average per year, which represents 2.1\% of the current Belgian \gls{GDP}.
 This in the range of other studies on climate-neutral scenarios concluding additional investment needs of the order of 2 to 3\% of the \gls{GDP} on a global scale \cite{IEA2021,IRENA2021} or 2.7 to 3.8\% at the European level \cite{Widuto2023}.
@@
 Initially, the infrastructure, transport, and electricity sectors each account for approximately one-third of the investments.
-The investments in infrastructure are primarily driven by the electricity grid and the \acrfull{DHN}, representing a combined investment of 73 b€\textsubscript{2015}.
+The investments in infrastructure are primarily driven by the electricity grid and the \acrfull{DHN}, representing a combined investment of 73 b€.
 The electricity sector's investment is led by power plants, totalling 31.5 b€.
@@
 
-As a comparison, the investments required to fully deploy the PV and wind potentials from 2020 to 2050 amount to 74.4 b€\textsubscript{2015}, with an additional 22.2 b€\textsubscript{2015} allocated to reinforce the grid.
-The electrification of the heating sectors necessitates investments of 29.2 b€\textsubscript{2015}, including 6.5 b€\textsubscript{2015} for the deployment of the \gls{DHN} infrastructure.
-Storage investments, primarily focused on \gls{DHN} seasonal storage, amount to 3.6 b€\textsubscript{2015}.
-Apart from the investment required to replace all private vehicles (accounting for 44\% of the overall investments), the remaining sectors represent a total of 212 b€\textsubscript{2015}.
+As a comparison, the investments required to fully deploy the PV and wind potentials from 2020 to 2050 amount to 74.4 b€, with an additional 22.2 b€ allocated to reinforce the grid.
+The electrification of the heating sectors necessitates investments of 29.2 b€, including 6.5 b€ for the deployment of the \gls{DHN} infrastructure.
+Storage investments, primarily focused on \gls{DHN} seasonal storage, amount to 3.6 b€.
+Apart from the investment required to replace all private vehicles (accounting for 44\% of the overall investments), the remaining sectors represent a total of 212 b€.
 To mitigate the cost of the transition, My2050 suggests deploying a fleet of no more than one million vehicles and implementing a car sharing system, distinct from car-pooling, as an inevitable measure \cite{My2050}.
@@
 Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Equation (\ref{eq:salvage}).
-Out of the 114.4~b€\textsubscript{2015} of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime.
+Out of the 114.4~b€ of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime.
 On the contrary, private mobility has a lower salvage value due to a major drop within the first four years and an average lifetime below 10 years \cite{febiac2021datadigest}.
-% Consequently, out the 240.5~b€\textsubscript{2015}, including the initial investment of 75.7~b€\textsubscript{2015}, only 13.9\% would remain available after 2050.
+% Consequently, out the 240.5~b€, including the initial investment of 75.7~b€, only 13.9\% would remain available after 2050.
 
@@
 Investments in infrastructures, public mobility, storage and other long-lifetime technologies experience an important salvage value on the contrary, investments in private mobility will not be recovered as vehicles have a short lifetime.
-All together, these salvage values represent 160.1 b€\textsubscript{2015}, 25\% of the cumulative investment costs in 2050.}
+All together, these salvage values represent 160.1 b€, 25\% of the cumulative investment costs in 2050.}
 \label{fig:pestd_inv_return}
@@
 \multirow{4}{*}{Costs in 2050}
-& Total transition\footnote{\label{foot:transition_cost}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & $\simeq$ & b€\textsubscript{2015}\\
-& Cumulative opex & 528 & $\simeq$  & b€\textsubscript{2015}\\
-& Cumulative capex & 636 & $\simeq$ & b€\textsubscript{2015}\\
-& Salvage value & 160 & -1\% & b€\textsubscript{2015}\\
+& Total transition\footnote{\label{foot:transition_cost}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & $\simeq$ & b€\\
+& Cumulative opex & 528 & $\simeq$  & b€\\
+& Cumulative capex & 636 & $\simeq$ & b€\\
+& Salvage value & 160 & -1\% & b€\\
 \midrule
@@
 
-Similarly to \citet{nerini2017myopic}, Figure \ref{fig:my_pestd_total_cost_cum_diff} shows that myopic optimisation ends up with a slightly more expensive energy transition by 2050 (\ie +3.2~b€$_{2015}$), compared to the perfect foresight, despite the savings done at the early stages.
-Even though this over-cost is negligible compared to the overall cost of the transition (\ie $\sim$1000~b€$_{2015}$), this is explained by the early investments in renewable technologies (\ie PVs and wind turbines) boosted by the significant salvage value retrieved from investing in the consequent reinforcement of the grid.
+Similarly to \citet{nerini2017myopic}, Figure \ref{fig:my_pestd_total_cost_cum_diff} shows that myopic optimisation ends up with a slightly more expensive energy transition by 2050 (\ie +3.2~b€), compared to the perfect foresight, despite the savings done at the early stages.
+Even though this over-cost is negligible compared to the overall cost of the transition (\ie $\sim$1000~b€), this is explained by the early investments in renewable technologies (\ie PVs and wind turbines) boosted by the significant salvage value retrieved from investing in the consequent reinforcement of the grid.
 
-Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€\textsubscript{2015} in 2030 whereas the overall cumulative investments, so far, are 421.3~b€\textsubscript{2015}.
+Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€ in 2030 whereas the overall cumulative investments, so far, are 421.3~b€.
 The significant lifetime (\eg 80 years) and investment cost (\ie 368M€/GW$_\text{VRES}$ \cite{readthedocs_pathway}) of the power grid, and, on a smaller scale, the district heating network, explain why the myopic optimisation opts for a higher investment in these infrastructures, at early stages.
@@
 \caption{Sum of the respective salvage values of each sector, per end year of each time window.
-All together, these salvage values represent 223.9~b€\textsubscript{2015}, 53\% of the cumulative investment costs in 2030, and 157.7~b€\textsubscript{2015}, 24\% of the cumulative investment costs in 2050.}
+All together, these salvage values represent 223.9~b€, 53\% of the cumulative investment costs in 2030, and 157.7~b€, 24\% of the cumulative investment costs in 2050.}
 \label{fig:my_pestd_inv_return_area}
@@
 Similarly to the case with an imposed \gls{GHG}-emissions trajectory, myopic optimisation invests more, compared to the perfect foresight, at early stages into \gls{VRES} technologies to benefit from the significant salvage values of the related grid infrastructures.
-In 2030, the salvage value of the infrastructures and electricity-generation technologies account for 80.2 and 56.2 b€\textsubscript{2015}, respectively, whereas the overall CAPEX are 404.8 b€\textsubscript{2015} by then.
+In 2030, the salvage value of the infrastructures and electricity-generation technologies account for 80.2 and 56.2 b€, respectively, whereas the overall CAPEX are 404.8 b€ by then.
 
@@
 \multirow{4}{*}{Costs in 2050}
-& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\textsubscript{2015}\\
-& Cumulative opex & 528 & -3\%  & b€\textsubscript{2015}\\
-& Cumulative capex & 636 & -2\% & b€\textsubscript{2015}\\
-& Salvage value & 160 & -5\% & b€\textsubscript{2015}\\
+& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\\
+& Cumulative opex & 528 & -3\%  & b€\\
+& Cumulative capex & 636 & -2\% & b€\\
+& Salvage value & 160 & -5\% & b€\\
 \midrule

### 2024-04-10 chap_RL/chap_RL.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

@@
 
+During the learning process, the algorithm explores numerous transition pathways (2439 successful transitions out of 10,565 attempts).
+Each pathway provides valuable insight into the best course of actions --- the primary goal of reinforcement learning.
+As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken.
+Yet, the exploration during this learning process is not exhaustive.
+The trends provided below are therefore not proven.
+The randomness of the process and the number of explored transition pathways still give us high confidence.
+
 Over the entire learning process, 23\% of the episodes have been successful versus 73\% of failures, the rest being infeasible problems where the actions taken by the agent are too constraining for the model (see Section \ref{sec:RL:act_states_rew}).
@@
 
-Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year''.
-Beyond this point, the chances to succeed the transition are 47\%.
+Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
+Beyond this point, through this learning process, the chances to succeed the transition are 47\%.
 In other words, mid-term actions are necessary to hope succeeding the transition.
@@
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 58\% of all the failures.
-Beyond this point, succeeding the transition represents 47\% of the cases.}
+Beyond this point, through this learning process, succeeding the transition represents 47\% of the cases.}
 \label{fig:reward_status}
@@
 \myparagraph{States: Renewables and efficiency}\\
+
+In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{RE_in_mix.pdf}
+\caption{BLABLABLA}
+\label{fig:RE_in_mix}
+\end{figure}
 

### 2024-04-10 chap_atom_mol/chap_atom_mol.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

--- 2024-04-04 chap_atom_mol/chap_atom_mol.tex
@@
 Besides these, ``unicorn'' technologies are also investigated \cite{heuberger2018impact}.
-These are technological solutions that have not (yet) reached a high enough maturity, \ie TRL 3 to 7, or facing non-techno-economic hurdles (\ie social acceptance) to be currently deployed on a large scale.
-Among others, nuclear energy, potentially in the form of \acrfull{SMR}, finds an interest in the literature \cite{IEA_Nuclear_2022,PATHS2050} as well as in actual current investments, in Belgium for instance \cite{SMRlesoir}.
+These technologies have not (yet) reached a high enough maturity, \ie Technology Readiness Level (TRL) 3 to 7, or they face hurdles such as social acceptance.
+Therefore, they are currently not deployed on a large scale.
+Among others, nuclear energy finds an interest in the literature \cite{IEA_Nuclear_2022,PATHS2050}.
+In the form of \acrfull{SMR}, nuclear energy is also discussed in actual current investments (in Belgium for instance \cite{SMRlesoir}).
 
-In a society with an increasing electricity demand, notably due to the electrification of sectors like mobility, low-temperature heat or industry, and a deeper integration of \gls{VRES}, there seems to be a case for nuclear energy to produce a reliable low-\ce{CO2}emission electricity \cite{IAEA2008}, especially given the willingness to phase out of imported Russian fossil fuels anchored in the European REPowerEU Plan \cite{REPowerEU}.
-Unlike fossil natural gas that is a ``flow-based'' resource, uranium is ``stock-based'', which favours the security of electricity supply, in case of a conflict like the invasion of Ukraine, for instance.
+In the future, deeper integration of \gls{VRES} will come with the bigger electrification of sectors like mobility, low-temperature and industry \cite{IEA2023electrification}.
+In such a context, there is a case for nuclear energy to produce a reliable low-\ce{CO2}emission electricity \cite{IAEA2008}.
+This is also in line with the willingness to phase out of imported Russian fossil fuels anchored in the European REPowerEU Plan \cite{REPowerEU}.
+Unlike the consumption of fossil gas that represents about 80\% of the \gls{LCOE} from \gls{CCGT}, the price of uranium accounts for 25-30\% of \gls{LCOE} from nuclear power plants.
+This favours nuclear energy against price volatility and for the security of electricity supply, in case of a conflict like the invasion of Ukraine, for instance.
 
-``Country specific energy studies are needed as a prerequisite to the decision of following the nuclear route'' \cite{IAEA2008}.
-Even though, in a country like Belgium, reaching the goal of energy transition will not be a ``winner-takes-all'' situation but rather a combination of solutions, implemented simultaneously \cite{Limpens2020}, this chapter focuses on this atom-molecules dilemma in two ways.
-First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-Second, accounting for uncertainties as presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
+In a country like Belgium, reaching the goal of energy transition will not be a ``winner-takes-all'' situation but rather a combination of solutions \cite{Limpens2020,limpens2021generating}.
+However, this chapter focuses on this atom-molecules dilemma.
+Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad, in a country where the local potential of \gls{VRES} is limited compared to its \gls{EUD}.
 
-\textbf{Disclaimer}: Relying on ``local'' nuclear energy for some is highly controversial.
+Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
+
+In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimize a transition pathway with a small number of time stages, \ie two or three.
+Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Appendix \ref{app:UC_full}).
+Accounting for uncertainties presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}, by the end of the transition, \ie 2050.
+
+Relying on nuclear energy for some is highly controversial.
 On top of purely techno-economic aspects, non-exhaustively mentioned beforehand, there are other ethical, societal or even political considerations to account for when addressing this question \cite{kempf2022}.
@@
 In their work, \citet{lefebvre2022electrofuel} investigated this topic for the case of Belgium considering lower and upper bounds in terms of availability based on, on the one hand, the already signed agreements with the exporting countries and, on the other hand, their maximal technological potential of \gls{VRES}, respectively.
-They highlighted that the lower bound, \ie currently the most reliable information, resulted in a total availability of electrofuels, at least, one order of magnitude lower than the needs provided by the cost-optimum carbon-neutral Belgium in 2050.
+Their lower bound resulted in a total availability of electrofuels that is one order of magnitude lower than the needs provided by the cost-optimum carbon-neutral Belgium in 2050.
 Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} as well as the need of importing electrofuels in the Belgian energy landscape, on a strictly techno-economical point of view with a cost-based optimisation.
@@
 
-\section*{Contributions}
-\label{sec:atom_mol:contributions}
-In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimize a transition pathway with a small number of time stages, \ie two or three.
-Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with tens of uncertain parameters (see Appendix \ref{app:UC_full}).
-
-Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad, in a country where the local potential of \gls{VRES} is limited compared to its \gls{EUD}.
-
-On top of identifying the most impactful uncertain parameters on the objective function, \ie the total transition cost, this chapter aims also at highlighting the ones that mostly drive other variables resulting from the optimization process: the installation of \gls{SMR} and the import of each of the electrofuels by the end of the transition, \ie 2050.
-
 \section{Deterministic impact of integrating \gls{SMR} in 2040}
 \label{sec:atom_mol:results_deter} 
-In this section, like in the rest of the thesis, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW, from 2040 onward.
-After investigating the deployment of \gls{SMR} through the power sector, the first part of this section focuses on this impact on macro/system-level considerations (\ie overall transition costs, primary energy mix and yearly emissions per sector).
+In this section, like in the rest of the thesis, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW (see Chapter \ref{chap:case_study}), from 2040 onward.
+After investigating the deployment of \gls{SMR} through the power sector, the first part of this section focuses on this impact on the whole-system level considerations (\ie overall transition costs, primary energy mix and yearly emissions per sector).
 The second part will address the impact of \gls{SMR} on each of the other sectors of the system.
@@
 \label{subsec:atom_mol:results_deter_power_sector}
-\Cref{fig:results_deter_tech_cap_elec} shows that \gls{SMR} is deployed as soon as available, \ie 2040, to their maximum capacity, \ie 6\,GW, substituting other flexible power generation units: no ammonia-\gls{CCGT} at the end of the transition and the anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case).
+\gls{SMR} is deployed as soon as available, \ie 2040, to their maximum capacity, 6\,GW, substituting other flexible power generation units (see \Cref{fig:results_deter_tech_cap_elec}).
+There is no ammonia-\gls{CCGT} at the end of the transition and an anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case).
 To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case.
@@
 
-When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity, given the 15\% maintenance off-time assumed in this work: 44.6\,TWh.
-By 2050, it represents 24.6\% of the total electricity production which is less than the current share of conventional nuclear in Belgium, 38.5\%.
-This resurgence of nuclear electricity occurs at the expense of other, although more efficient technologies: \gls{CCGT} and industrial \gls{CHP}.
-Besides the unchanged end-use-demand, we observe a slight increase of the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.
+When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity.
+Given the 15\% maintenance off-time assumed in this work, this represents 44.6\,TWh by 2050.
+In comparison, in 2020, conventional nuclear power plants produced 34.4\,TWh Belgium.
+This resurgence of nuclear electricity occurs at the expense of other, although more efficient, technologies: \gls{CCGT} and industrial \gls{CHP}.
+Besides the unchanged end-use-demand compared with the REF case, we observe a slight increase of the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.
 
@@
 \label{subsec:atom_mol:results_deter_overall}
-First of all, as far as the objective function (\ie the total transition cost) is concerned, \Cref{fig:results_deter_overall_emissions_sector} shows that the 6\,GW \gls{SMR} installed from 2040 allow reaching a 3.3\% (-36.9\,b€, $\sim$\,6\% of the Belgian GDP) cheaper overall transition.
-Interestingly, as the model can freely spend the constrained \ce{CO2}-budget over the transition, knowing ahead (\ie perfect foresight) that cheaper and low-emitting \gls{SMR} will be available in the future, cost-savings, that are more important after 2040, also occur before 2040.
-This is mostly due to the extended use of cheaper \gls{LFO} at the beginning of the transition to produce \gls{HVC}, that is then compensated by the deployment of \gls{SMR} (\Cref{fig:results_deter_overall_emissions_sector}).
+First of all, the 6\,GW \gls{SMR} installed from 2040 allow reaching a 3.3\% (-36.9\,b€) cheaper overall transition (\Cref{fig:results_deter_overall_emissions_sector}).
+Over the 30-year transition, this represents an annual cost saving equal to 0.2\% of the Belgian GDP.
+Thanks to the perfect foresight approach, the model knows ahead that cheaper and low-emitting \gls{SMR} will be available in the future.
+As the model can freely spend the constrained \ce{CO2}-budget over the transition, cost-savings also occur before 2040.
+These early-stage cost-savings are equally shared between the extended use of cheaper \gls{LFO} to produce \gls{HVC} and the delayed deployment of \gls{PV}.
 Then, the capital-intensive investments in \gls{SMR}, mostly recovered by the end of the transition as salvage value, are widely compensated by the smaller resource-related OPEX.
-This leads, at the end, aggregating the OPEX and the annualised CAPEX, to a system that is yearly 8.8\% less expensive by 2050.
+This leads, at the end, aggregating the OPEX and the annualised CAPEX, to a system that is yearly 4.4\,b€ (8.8\%) cheaper by 2050 than in the REF case (Figure \ref{fig:results_deter_overall_emissions_sector}).
 
@@
 \includegraphics[width=0.8\textwidth]{Cum_total_cost_diff_REF.pdf}
+\includegraphics[width=0.8\textwidth]{System_cost_DIFF.pdf}
+\caption{(Top) Cumulative transition cost and (down) annual system cost differences between the SMR and the REF cases.
+Including \gls{SMR} ends up in a cheaper overall transition (-36.9\,b€) and cheaper whole-energy system by 2050 (-4.4\,b€).}
+\label{fig:results_deter_overall_emissions_sector}
+\end{figure}
+
+Extending the use of naphtha-cracker and, to a lesser extent the postponing the deployment of PV, lead to higher \ce{CO2}-emissions at the beginning of the transition that are then compensated by the deployment of \gls{SMR} (\Cref{fig:GWP_per_sector_diff}).
+
+\begin{figure}[htbp!]
+\centering
 \includegraphics[width=0.8\textwidth]{GWP_per_sector_diff.pdf}
-\caption{(Left) Cumulative transition cost difference between the SMR and the REF cases.
-Including \gls{SMR} ends up in a cheaper overall transition (\ie -3.3\%, -36.9\,b€).
-(Right) Emissions per energy sector differerence between the SMR and the REF cases.
+\caption{Emissions per energy sector differerence between the SMR and the REF cases.
 The deployment of \gls{SMR} from 2040 onward allows maintaining longer the use of \acrfull{LFO} for the production of \acrfull{HVC}.}
-\label{fig:results_deter_overall_emissions_sector}
+\label{fig:GWP_per_sector_diff}
 \end{figure}
@@
 \includegraphics[width=0.8\textwidth]{Primary_mix_diff_uranium.pdf}
-\caption{(Up) Primary energy mix over the transition for the REF and SMR cases.
-(Bottom) Difference of the mix between the two cases after aggregation per category.
-In Uranium, considered as a non-renewable resource in line with \citet{rixhon2021terminology}, is nevertheless dissociated from other non-renewable resources for the sake of clarity.
-The imported electricity is split between imported non-renewable and renewable resources assuming a linear decrease between the current share of renewables, \ie 37.41\% \cite{eurostat_share_re_elec}, and an assumed 100\%-renewable European electricity mix by 2050.}
+\caption{(Top) Primary energy mix over the transition for the REF and SMR cases.
+(Down) Difference of the mix between the two cases after aggregation per category.
+In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
+It is nevertheless dissociated from other non-renewable resources for the sake of clarity.
+The imported electricity is split between imported non-renewable and renewable resources between the current share of renewables, \ie 37.41\% \cite{eurostat_share_re_elec}, and an assumed 100\%-renewable European electricity mix by 2050.}
 \label{fig:results_deter_energy_mix}
@@
 
-Beyond the direct impact that \gls{SMR} has on the power sector, it is worth assessing the other sectors given the sector-coupling effect related to the whole-energy system optimisation \cite{contino2020whole}.\\
+Beyond the direct impact that \gls{SMR} has on the power sector, it also affects the other sectors.
+This is due to the sector-coupling typical of a whole-energy system optimisation \cite{contino2020whole}.\\
 
@@
 
-As aforementioned, the main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of overall more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
-On the one hand, in the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production" of the 59\,GW solar-PV when fully deployed, in the sunny days.
-With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity, consequently increasing their load factor and yearly production respectively by 31\% and 48\%.
-On the other hand, given the 44.6\,TWh of electricity produced by \gls{SMR}, cogeneration units are less relevant and, by 2050, 2.6\,GW industrial gas boilers completely substitute \gls{CHP} to produce 16.6\,TWh (\ie 23\%) of the total production of high-temperature heat.\\
+The main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
+In the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production" of the 59\,GW solar-PV when fully deployed, in the sunny days.
+With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity.
+This increases the load factor (31\%) and yearly production (48\%) of these industrial electric heaters.
+Then, given the 44.6\,TWh of electricity produced by \gls{SMR}, cogeneration units are less relevant and, by 2050, 2.6\,GW industrial gas boilers completely substitute \gls{CHP} to produce 16.6\,TWh (\ie 23\%) of the total production of high-temperature heat.\\
 
@@
 In both cases, the major shift of supply from decentralised to centralised productions operates early in the transition, to hit the constraint that \gls{DHN} cannot supply more than 37\% of the \gls{LT}-heat production.
-Then, from a mix between oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system progressively shifts towards electric \gls{HP} only.
+Then, from a mix between oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system shifts towards electric \gls{HP} only by 2035.
 Similarly for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economic option.\\
@@
 
-The passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} progressively substituting \gls{ICE} cars for the private sector.
+The passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} substituting \gls{ICE} cars for the private sector by 2030.
 About the public mobility, trains and tramways supply their \textit{a priori} set maximum share, respectively, 50\% and 30\% complemented by \gls{CNG} buses substituting diesel-driven buses.
@@
 \label{sec:atom_mol:results_uq}
-``\textit{It is difficult to make predictions, especially about the future.}" (Niels Bohr, foundational contributor to the understanding of atomic structure and quantum theory).
-Besides the deep understanding of the deterministic results, it is important to challenge these conclusions out-of-sample, accounting for the uncertainty of the parameters.
-After briefly assessing the \gls{GSA} on the objective function of the model, the total transition cost, this section investigates more deeply the atom-molecules dilemma.
-This time, the Sobol' indices are computed for import of renewable molecules and installed capacities of \gls{SMR}.
+Besides the detailed understanding of the deterministic results, it is important to challenge these conclusions accounting for the uncertainty of the parameters \cite{guevara2020machine}.
+After briefly assessing the \acrfull{GSA} on the total transition cost (Section \ref{subsec:atom_mol:results_uq_cost}), this section investigates more deeply the atom-molecules dilemma (Section \ref{subsec:atom_mol:results_atom_mol}).
+The Sobol' indices are computed for import of renewable molecules and installed capacities of \gls{SMR}.
 
@@
 \label{subsec:atom_mol:results_uq_cost}
-Exhaustively listed in Appendix \ref{app:UQ_transition_cost}, \Cref{tab:UQ_short} gathers the most impacting parameters\footnote{Per \citet{Turati2017}, parameters are considered as ``impacting'' if their Sobol' index is above the threshold $=1/d$, $d$ being the total number of uncertain parameters after the pre-selection phase.
-In this case, $d=34$, and, consequently, the threshold is equal to 2.9\%.} on the total transition cost, highlighting the cost of purchasing electrofuels as well as the potentiality to install \gls{SMR} and its CAPEX.
-The former is the most impacting parameter whereas the two others have much lower influence on the variation of the total transition cost.
+Exhaustively listed in Appendix \ref{app:UQ_transition_cost}, \Cref{tab:UQ_short} gathers the most impacting parameters on the total transition cost.
+Per \citet{Turati2017}, parameters are considered as ``impacting'' if their Sobol' index is above the threshold $=1/d$, $d$ being the total number of uncertain parameters after the pre-selection phase.
+In this case, $d=34$, and, consequently, the threshold is equal to 2.9\%.
+\Cref{tab:UQ_short} shows that the cost of purchasing electrofuels is the most impacting parameter.
+On the contrary, the potentiality to install \gls{SMR} and its CAPEX have much lower influence on the variation of the total transition cost.
 Given the uncertainty characterisation presented in Section \ref{sec:cs:uncertainty}, there are 60\% chance that no \gls{SMR} could be installed.
 In other words, the variation of the parameter $f_{\mathrm{max,SMR}}$ has zero impact on the variation of the total transition cost in 60\% of the samples.
-Then, in perspective with the local sensitivity analysis of Section \ref{sec:atom_mol:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
+Then, in perspective with the scenario analysis of Section \ref{sec:atom_mol:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
 This represents only 10\% of the samples.
-Moreover, given its characteristics detailed in \Cref{tab:SMR_features}, mostly its cheap and low-emitting fuel (\ie uranium) and the long lifetime leading to lower annualised CAPEX and higher salvage value, this explains why \gls{SMR} has this lower impact.
+Moreover, when it is possible, \gls{SMR} is always installed thanks to its characteristics: cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.
+However, given its limited deployment, up to 6\,GW, \gls{SMR}-related parameters have a lower impact on the variation of the total transition cost.
 On the contrary, more expensive renewable electrofuels are always imported, to a smaller or larger extent depending on the sample.
 For instance, in the REF case (Section \ref{subsec:atom_mol:results_deter_overall}), the imported electrofuels represent, by 2050, 152.9\,TWh (\ie 41\% of the primary energy mix) with an average 93€/MWh cost of purchasing and, over the entire transition, a 273\,b€ cumulative OPEX (\ie 25\% of the total transition cost).
+
+Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
+
+Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index.
+The model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
+In practice, the interest rate would vary depending on the technology investment risk.
+This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.
+
+Finally, the cost of purchasing fossil fuels is also a key parameter in terms of variation of the total transition cost.
+However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their impact smaller than their renewable alternatives.
 
@@
 
-Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
-Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index\footnote{It is important to note here that the model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
-In practice, the interest rate would vary depending on the technology investment risk.
-This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.}.
-Finally, similarly to electrofuels, the cost of purchasing fossil fuels is also to consider in the perspective to reduce the uncertainty over the total transition cost.
-However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their uncertain impact smaller than their renewable alternatives.
-
 \Cref{fig:UQ_PDF_total_transition_cost} shows the \gls{PDF} of the total transition cost given the 1260 samples.
@@
 \label{subsec:atom_mol:results_atom_mol}
-The samples used to carry out the \gls{GSA} on the total transition cost, also provide the distribution of other outputs of the model.
-Among them, \Cref{fig:results_uq_electrofuels} shows the evolution of the import of renewable electrofuels over the transition.
+The samples used to carry out the \gls{GSA} on the total transition cost, also provide the distribution of other outputs of the model like the import of renewable electrofuels over the transition.
+As the general trends are increasing, discrepancies exist between the different energy carriers (see \Cref{fig:results_uq_electrofuels}).
+%
+%\begin{figure}[htbp!]
+%\centering
+%\includegraphics[width=0.7\textwidth]{UQ_Electrofuels.pdf}
+%\includegraphics[height=4cm]{Schematic_boxplot.pdf}
+%\caption{Distribution of the imported renewable electrofuels over the transition.
+Starting from no electrofuel in 2020, their respective import rises progressively along the transition (\ie increasing median) at different growth rates and with different ranges of values.
+Observations being either 1.5 times the interquartile range (IQR) less than the first quartile (Q1) or 1.5 times the interquartile range greater than the third quartile (Q3) are defined as outliers.}
+%\label{fig:results_uq_electrofuels}
+%\end{figure}
 
-As the general trends are increasing, discrepancies exist between the different energy carriers.
-E-methane, as the renewable alternative to fossil natural gas, substitutes it, sometimes at a very early stage of the transition, 2025, and to a somehow unrealistically large extent, 163\,TWh, which is more than 6\% more than the total import of electrofuels in the REF case.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.7\textwidth]{UQ_Electrofuels_no_outlier.pdf}
+\includegraphics[height=3.8cm]{Schematic_boxplot_no_outlier.pdf}
+\caption{Distribution of the imported renewable electrofuels over the transition.
+Starting from no electrofuel in 2020, their respective import rises progressively along the transition (\ie increasing median) at different growth rates and with different ranges of values.
+}
+\label{fig:results_uq_electrofuels}
+\end{figure}
+
+E-methane, as the renewable alternative to fossil methane, substitutes it, sometimes at a very early stage of the transition, 2025, and large extent, 163\,TWh, which is more than 6\% more than the total import of electrofuels in the REF case.
 The necessity to import this molecule is progressive through the transition to supply mostly industrial \gls{CHP} and boilers.
 
-E-hydrogen becomes rapidly the main stream of hydrogen in the system, on top of steam-methane-reforming or electrolysis, to reach a median and a maximum values of, respectively 13.0\,TWh and 42.1\,TWh in 2050.
+E-hydrogen becomes rapidly the main stream of hydrogen in the system, to reach a median and a maximum values of, respectively 13.0\,TWh and 42.1\,TWh in 2050.
 Hydrogen is more frequently used in the mobility sector.
@@
 
-Then, the imported e-ammonia, becoming rapidly cost-competitive against its fossil alternative (\Cref{fig:cs_resources_cost}), quickly substitutes it and the Haber-Bosch process.
+Then, the imported e-ammonia rapidly becomes cost-competitive against its fossil alternative (see \Cref{fig:cs_resources_cost}).
+At early stage of the transition, e-ammonia substitutes fossil ammonia and the Haber-Bosch process.
 Where the initial purpose of ammonia is to satisfy a relatively limited \acrfull{NED} (\ie 10 $\pm$ 3\,TWh by 2050), the variation of its import is mostly due to the higher or lower need for ammonia-\gls{CCGT} as a flexible option to produce electricity.
@@
 It is an indicator of statistical dispersion around the median, $Q2$ or $P_{50}$.} of about 50\,TWh.
-In some extreme cases, e-ammonia is the most imported molecules, \ie up to 216\,TWh or 58\% of the total primary mix in the REF case in 2050.
+In some extreme cases, e-ammonia is the most imported molecules, \ie up to 216\,TWh or 58\% of the total primary mix.
 
-Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol or synthetic methanolation exist in some outlying cases.
-Given its lower \gls{NED} (\ie 1.5$\pm$0.5\,TWh$_{\text{NED,methanol}}$ by 2050), the variation of imported e-methanol is almost exclusively due to its role in the industrial production of \gls{HVC}, \ie 78\% of the total \gls{NED}, through the \acrfull{MTO} process.
-In some rare samples, methanol is also used to supply the freight transport sector via boats or trucks.
+Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand in methanol.
+Its \gls{NED} represents, on average, 3\% of the total consumption of methnaol.
+The variation of imported e-methanol is due to its role in the industrial production of \gls{HVC} through the \acrfull{MTO} process as it represents 95\% of the total consumption.
+For the remaining 2\%, methanol is also used to supply the freight transport sector via boats or trucks.
+More detailed statistics including the different sources of supply and consumption of gas, hydrogen, ammonia and methanol are provided in Appendix \ref{app:UQ_electrofuels}.\\
 
-Appendix \ref{app:UQ_electrofuels} gives a more detailed information.
-On the one hand, it compares the statistics (\ie quartiles and median) with the quantity of imported electrofuels in the REF and SMR cases.
-On the other hand, this appendix shows the distributions of the different sources of supply and consumption of gas, hydrogen, ammonia and methanol.
+This part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe.
+The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050 (see Figures \ref{fig:results_uq_samples_A} and \ref{fig:results_uq_samples_B}).
+We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transiton.Next to the name of a parameter, one can read its Sobol' index versus the output of interest.
+For these outputs of interest, different from the total transition cost, the \gls{LOO} error is around 20\%.
+Even though this is higher than the threshold of \SI{1}{\%}, the Sobol' indices allow ranking the most impacting parameters (see Section \ref{subsec:pce}).
+Box plots also point out the distribution of this output for the extreme low or high values of some parameters.
 
@@
 \centering
-\includegraphics[width=0.7\textwidth]{UQ_Electrofuels.pdf}
-\includegraphics[height=4cm]{Schematic_boxplot.pdf}
-\caption{Distribution of the imported renewable electrofuels over the transition.
-Starting from no electrofuel in 2020, their respective import rises progressively along the transition (\ie increasing median) at different growth rates and with different ranges of values.
-Observations being either 1.5 times the interquartile range (IQR) less than the first quartile (Q1) or 1.5 times the interquartile range greater than the third quartile (Q3) are defined as outliers.}
-\label{fig:results_uq_electrofuels}
+\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
+\includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
+\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
+
+\caption{Trend lines of the key parameters on the import of e-methane, e-hydrogen and e-ammonia.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_A}
 \end{figure}
 
-After investigating the distribution of imports of electrofuels over the transition, this part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe.
-Figures \ref{fig:results_uq_samples_A} and \ref{fig:results_uq_samples_B} give the trend lines of the key parameters for these imports in 2050\footnote{The authors picked this specific year as it is the one where electrofuels, if imported, are imported in the largest amount, in general, compared to the other years of the transiton.}, as well the installed capacities of \gls{SMR}.
-Next to the name of a parameter, one can read its Sobol' index versus the output of interest\footnote{For these outputs of interest, different from the total transition cost, the \gls{LOO} error is generally higher than the threshold of \SI{1}{\%} defined in Section \ref{subsec:pce}.
-Consequently, the Sobol' indices are less accurate but already allow a fair relative comparison between the different parameters.}.
-Box plots also point out the distribution of this output for the extreme low or high values of some parameters.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
+\includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
+\caption{Trend lines of the key parameters on the import of e-methanol and the installed capacity of \gls{SMR}, in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_B}
+\end{figure}
 
@@
 Finally, and surprisingly, costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively.
-In other words, by 2050, more expensive electrofuels induce to more e-methane import and \textit{vice versa} for the fossil fuels.
+In other words, by 2050, more expensive electrofuels result in more e-methane import.
+On the contrary, cheaper electrofuels result in higher import of fossil methane.
 Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}.
 Subject to the \ce{CO2}-budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity.
-First running on fossil natural gas, these \gls{CHP} consume more e-methane by 2050.
-In the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
+First running on fossil gas, these \gls{CHP} consume more e-methane by 2050.
+On the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
 This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
-In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the models finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
+In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
 Consequently, there is smaller investment in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers.
@@
 In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector.
-As depicted in \Cref{fig:results_uq_prod_cons} in Appendix \ref{app:UQ_electrofuels}, the utilisation of e-hydrogen is most prevalent in \gls{FC} trucks, followed by \gls{FC} cars and buses to a lesser extent.
+The utilisation of e-hydrogen is most prevalent in \gls{FC} trucks, followed by \gls{FC} cars and buses to a lesser extent (Appendix \ref{app:UQ_electrofuels}).
 The adoption of fuel cell engines in trucks contributes, on average, to 63.5\% of the total road freight transport, thereby affecting the level of e-hydrogen imports.
@@
 As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia.
-As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheap electricity flexibly produced by \gls{SMR} substitutes the \gls{CCGT}.
+As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheaper electricity produced by \gls{SMR} (40 versus 151 €/MWh$_{\text{elec}}$) substitutes these \gls{CCGT}.
 With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 2.0\,TWh, 95.4\% less than in the REF case.
@@
 Not shown here but all the samples of the \gls{GSA} highlight that \gls{SMR} is installed to its maximum capacity, \ie 6\,GW, as soon as possible.
-In other words, the only parameter ``Potential capacity \gls{SMR}'' dictates the installation of this technology\footnote{In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from 2040, 2045 and 2050, respectively.}.
-Surprisingly, the [-40\%; +44\%]  variation of its CAPEX has a negligible impact, with a Sobol' index of 0.9\%.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
-
-\caption{Trend lines of the key parameters on the import of e-methane, e-hydrogen and e-ammonia.
-Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
-The grey dashed line gives the value of the output of interest in the REF case.
-Part A}
-\label{fig:results_uq_samples_A}
-\end{figure}
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-methanol and the installed capacity of \gls{SMR}, in 2050.
-Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
-The grey dashed line gives the value of the output of interest in the REF case.
-Part B}
-\label{fig:results_uq_samples_B}
-\end{figure}
+In other words, the only parameter ``Potential capacity \gls{SMR}'' dictates the installation of this technology.
+In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from 2040, 2045 and 2050, respectively.
+Large range (-40\%; +44\%) of its CAPEX has a negligible impact on the installed capacity, with a Sobol' index of 0.9\%.
+This is explained by the long lifetime of the technology (60 years) and the salvage value recovered by 2050.
 
@@
 In line with results given in Section \ref{sec:atom_mol:results_deter}, the \gls{GSA} shows that \gls{SMR} has negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines).
-\Cref{fig:results_uq_pdf_local_ren} gives the evolution of Sobol' indices for the most impacting parameters on the deployment of \gls{PV} and offshore wind\footnote{As the installed capacities of onshore wind is totally driven by the uncertainty on its maximum potential, $f_{\mathrm{max,windon}}$, it is not represented in the figure.}.
+Whereas the installed capacities of onshore wind is totally driven by the uncertainty on its maximum potential, Sobol' indices change for the most impacting parameters on the deployment of \gls{PV} and offshore wind (see Figure \ref{fig:results_uq_pdf_local_ren}).
 The key factor that drives the installed capacities of these two technologies is mostly their respective maximum potential, especially at the end of the transition, much more than their CAPEX.
@@
 \label{sec:atom_mol:discuss}
-Given its lower \gls{LCOE}, \gls{SMR} is installed as soon as available.
+Given its lower \gls{LCOE} (40\,€/MWh$_{\text{elec}}$) and the important share of the investment recovered via its salvage value (between 79\% and 96\%), \gls{SMR} is installed as soon as available.
 It directly substitutes other flexible power generation units (\ie CCGT) and provides, by 2050, 44.6\,TWh, 25\% of the total electricity production.
@@
 
-Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), \acrfull{GSA} highlights the biggest impact of the cost of purchasing electrofuels on the variation of the total transition cost, around 45\%.
+Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the biggest impact of the cost of purchasing electrofuels on the variation of the total transition cost, around 45\%.
 On the contrary, parameters directly related to \gls{SMR} (\ie its availability and its CAPEX) have limited impact, below 1\%.
@@
 
-In conclusion, this work puts under the spotlight the ``competition'' between \gls{SMR} and imported electrofuels while both of them support to the integration of local \gls{VRES}.
+In conclusion, this work puts under the spotlight the ``competition'' between \gls{SMR} and imported electrofuels while both of them support the integration of local \gls{VRES}.
 Betting on \gls{SMR} means letting the emissions go up at early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are 3.3\% and 8.8\% cheaper than in the REF case, respectively.
@@
 This would allow covering ourselves against the risk of eventually not having \gls{SMR} available by the end of the first half of this century.
+The perfect foresight approach does not model this risk as the possibility to install or not \gls{SMR} is known in advance.
 In the EnergyScope Pathway model, assuming the same technology mix as in the SMR case up to 2040 but without installing \gls{SMR} for the rest of the transition, \ie between 2040 and 2050, would require drastic (and quick) changes to still meet the \ce{CO2}-budget: around 44\,b€ of additional cumulative OPEX in purchasing electrofuels, mostly e-ammonia (+32\,TWh by 2050) and e-methane (+31\,TWh) to supply \gls{CCGT} and industrial \gls{CHP}.

### 2024-04-10 chap_case_study/chap_case_study.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

@@
 availability, as detailed by \citet{limpens2024pathway}.
+
 First, the evolution of the respective costs of purchasing is presented (see \Cref{fig:cs_resources_cost}).
 Regarding ``renewable electrofuels'', the costs are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis \cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
+
 Then, besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
@@
 Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050) \cite{limpens2021generating}.
-In the current work, the electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``sustainable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
+
+In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
 In practice, it means that their \gls{GWP} is assumed to be zero in the model.
+Compared to equivalent electrofuels that would be locally produced, this assumption does not account account for the emissions related to their transport of imported electrofuels from abroad to Belgium.
+However, the share of these emissions is limited (between 5 and 10\%) compared to the whole cycle of production and use of imported electrofuels \cite{coppitters2024towards}.
+Therefore, the conclusions drawn in this thesis are not affected by this assumption.
+
 Regarding these electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco).

### 2024-04-10 chap_methodo/chap_methodo.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR

@@
 \noindent
-Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has two main advantages: shorter computational time and more realistic representation of the short-sightedness of decision-makers.
+Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has three main advantages: shorter computational time, more realistic representation of the short-sightedness of decision-makers and the ability to capture shocks and unexpected events \cite{mccollum2020energy}.
 For this reason, several studies are based on this approach \citep{babrowski2014reducing,poncelet2016myopic,nerini2017myopic,heuberger2018impact}.
@@
 
+\myparagraph{Qualitative ranking based on Sobol' indices}\\
+
+When a \gls{LOO} error of 1\% cannot be achieved, Sobol' indices can still provide a qualitative ranking to identify the most impacting parameters on the output of interest (see Table \ref{tab:methodo:Sobol_ranking_order_1_2}).
+A first-order \gls{PCE} to assess the uncertainties on the total transition cost has a \gls{LOO} error of 16\%.
+However, it has the main advantage to require significantly fewer runs than more accurate second-order \gls{PCE} (70 versus 1260) while computing similar Sobol' indices for the most impacting parameters, and consequently giving similar ranking.
+
+\begin{table}[htbp!]
+\caption{Total Sobol' indices of the uncertain parameters over the total transition cost via \gls{PCE} of orders 2 and 1.
+The similar rankings (and indices) show the validity of using the faster (even though less accurate) lower order \gls{PCE} to rank the most impacting uncertain parameters.}
+\label{tab:methodo:Sobol_ranking_order_1_2}
+\begin{minipage}{\textwidth}
+\centering
+\resizebox{0.8\textwidth}{!}{
+\begin{tabular}{l c c|c c}
+\toprule
+\multirow{3}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
+& \multicolumn{2}{c|}{PCE order 2} & \multicolumn{2}{c}{PCE order 1}\\
+& \multicolumn{2}{c|}{LOO error 1\%} & \multicolumn{2}{c}{LOO error 16\%}\\
+\midrule
+Purchase electrofuels & 1 & (46.8\%) & 1 & (43.5\%) \\
+Industry EUD & 2 & (23.2\%) & 2 & (18.3\%) \\
+Interest rate & 3 & (12.0\%) & 3 & (12.1\%) \\
+Purchase fossil fuels & 4 & (5.7\%) & 5 & (3.5\%) \\
+Variable OPEX of technologies & 5 & (3.1\%) & 6 & (2.4\%)\\
+Purchase biofuels & 6 & (2.6\%) & 4 & (4.1\%) \\
+\bottomrule							
+\end{tabular}}
+\end{minipage}
+\end{table}
+
 \myparagraph{Comparison with a proven method}\\
@@
 After the initial phase of grouping (Section \ref{subsec:uncert_charac}), a preliminary screening was necessary to identify the key parameters to account for in this \gls{GSA}.
-\citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2}-limits using the snapshot model, \ie EnergyScope TD \cite{limpens2019energyscope}.
-Screening the results of this work, we have discarded some parameters with negligible impact\footnote{Per \citet{Turati2017}, parameters are considered as ``negligible'' if their Sobol' index is below the threshold $=1/d$, $d$ being the total number of uncertain parameters} (\eg CAPEX of electrolysers or variation of the freight demand), selected a subset of parameters and added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
+\citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2}-limits using the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
+Screening the results of this work, we have discarded some parameters with negligible impact such as CAPEX of electrolysers or variation of the freight demand.
+Parameters are considered as ``negligible'' if their Sobol' index is below the threshold $=1/d$, $d$ being the total number of uncertain parameters \cite{Turati2017}.
+On top of this selection, we have added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
 The exhaustive list of these 34 parameters is presented in Appendix \ref{app:UC_full}.

### 2024-04-11 chap_RL/chap_RL.tex
Atom_molecules chap. fully reviewed by FC and integrated
Final touch for UQ_samples figures

@@
 \centering
-\includegraphics[width=0.7\textwidth]{Reward.pdf}
+\includegraphics[width=0.65\textwidth]{Reward.pdf}
 \caption{Reward function, $R$.
@@
 \includegraphics[width=0.8\textwidth]{RE_in_mix.pdf}
-\caption{BLABLABLA}
+\caption{Exploration of the state space over the learning process with a focus on the share of renewable energy carriers in the primary energy mix.
+Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
+Below a near-term threshold (about 60\%), the chances of success are limited, \ie no-go zones.}
 \label{fig:RE_in_mix}
 \end{figure}
+
+\myparagraph{Actions}\\
 
@@
 \centering
+\includegraphics[width=0.8\textwidth]{Actions_learning.pdf}
+\caption{Over the whole learning process, actions taken by the agent.
+Besides the wide exploration of the action space, it does not seem to be any clear set of actions to take to support successful transitions.}
+\label{fig:Actions_learning}
+\end{figure}
+
+\begin{figure}[!htbp]
+\centering
 \includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
@@
 \label{fig:Binding_constr}
+\end{figure}
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Binding_learning.pdf}
+\caption{Out of the successful transitions, binding and not binding actions taken by the agent.
+Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
+Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.
+Limiting coal consumption is always constraining, unlike \gls{LFO} that is ``naturally'' substituted by EnergyScope Pathway in the near-term.}
+\label{fig:Binding_learning}
 \end{figure}

### 2024-04-11 chap_atom_mol/chap_atom_mol.tex
Atom_molecules chap. fully reviewed by FC and integrated
Final touch for UQ_samples figures

@@
 
-Then, the imported e-ammonia rapidly becomes cost-competitive against its fossil alternative (see \Cref{fig:cs_resources_cost}).
+Then, the imported e-ammonia rapidly becomes cost-competitive against its fossil alternative (see Chapter \ref{chap:case_study}).
 At early stage of the transition, e-ammonia substitutes fossil ammonia and the Haber-Bosch process.
@@
 This part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe.
-The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050 (see Figures \ref{fig:results_uq_samples_A} and \ref{fig:results_uq_samples_B}).
+The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050.
 We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transiton.Next to the name of a parameter, one can read its Sobol' index versus the output of interest.
@@
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
-
-\caption{Trend lines of the key parameters on the import of e-methane, e-hydrogen and e-ammonia.
-Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
-The grey dashed line gives the value of the output of interest in the REF case.
-}
-\label{fig:results_uq_samples_A}
-\end{figure}
+For the import of e-methanol, industrial \gls{EUD} is, by far (\ie $\sim$80\% Sobol' index), the key factor (Figure \ref{fig:results_uq_samples_methanol}).
+Due to its own \gls{NED} but, above all, since it is the low-emitting alternative picked by the model to supply the significant \gls{NED} of \gls{HVC}, the lower this demand, the lower the need to import e-methanol, and vice versa.
 
@@
 \includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
-\includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-methanol and the installed capacity of \gls{SMR}, in 2050.
+\caption{Trend lines of the key parameters on the import of e-methanol in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 }
-\label{fig:results_uq_samples_B}
+\label{fig:results_uq_samples_methanol}
 \end{figure}
 
-As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane.
+As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}).
 This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production.
@@
 
-In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
+\caption{Trend lines of the key parameters on the import of e-methane in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_gas}
+\end{figure}
+
+In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector (Figure \ref{fig:results_uq_samples_H2}).
 The utilisation of e-hydrogen is most prevalent in \gls{FC} trucks, followed by \gls{FC} cars and buses to a lesser extent (Appendix \ref{app:UQ_electrofuels}).
@@
 
-As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
+\caption{Trend lines of the key parameters on the import of e-hydrogen in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_H2}
+\end{figure}
+
+As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia (Figure \ref{fig:results_uq_samples_ammonia}).
 As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheaper electricity produced by \gls{SMR} (40 versus 151 €/MWh$_{\text{elec}}$) substitutes these \gls{CCGT}.
@@
 
-The conclusions are more straightforward for the import of e-methanol and the installed capacity of \gls{SMR}.
-For the former, industrial \gls{EUD} is, by far (\ie 79\% Sobol' index), the key factor.
-Due to its own \gls{NED} but, above all, since it is the low-emitting alternative picked by the model to supply the significant \gls{NED} of \gls{HVC}, the lower this demand, the lower the need to import e-methanol, and vice versa.
-For the latter, it is the availability of the technology that drives its installation.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
+\caption{Trend lines of the key parameters on the import of e-ammonia in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_ammonia}
+\end{figure}
+
+For the installed capacity of \gls{SMR}, it is the availability of the technology that drives its installation (Figure \ref{fig:results_uq_samples_SMR}).
 Not shown here but all the samples of the \gls{GSA} highlight that \gls{SMR} is installed to its maximum capacity, \ie 6\,GW, as soon as possible.
@@
 This is explained by the long lifetime of the technology (60 years) and the salvage value recovered by 2050.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
+\caption{Trend lines of the key parameters on the installed capacity of \gls{SMR} in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_SMR}
+\end{figure}
 

### 2024-04-11 chap_RL/chap_RL.tex
Keep going on chapter RL

@@
 
-\begin{itemize}
-\item Actions to take, which ones are binding
-\item Show the state space: share of \gls{VRES} and efficiency
-\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
-\end{itemize}
+After investigating the intermediate milestones to meet the \ce{CO2}-budget by 2050, this section details the actions the agent has taken during the learning process (see Figure \ref{fig:Actions_learning}).
+Rows represent the beginning of the time window at which the set of actions is taken.
+Similarly to the state space, we observe a wide exploration of the action space.
+The more the agent was able to progress through transition, without exceeding the \ce{CO2}-budget, the bigger is the share of successes compared to failures.
+Besides this observation, no specific range of values for the different actions at the different timing seems to lead to more successes.
+In other words, there is no clear set of actions that would support more effectively the transition.
 
@@
 \end{figure}
+
+To identify the actions that have an actual impact on the environment, we looked if they were binding or not.
+In a \gls{LP} problem, constraints represent straight lines in the domain of variables.
+When this problem is bounded and feasible, these lines are the edges of a convex polygon, the domain of feasibility (see Figure \ref{fig:Binding_constr}).
+The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function.
+Besides being within the domain of feasibility, it is proven that this optimal solution, when unique \footnote{There are cases where the objective function has the same optimal value along an entire edge.
+In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}.
+Among This vertex is t
 
@@
 \end{figure}
+
+\begin{itemize}
+\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
+\end{itemize}
 

### 2024-04-12 chap_RL/chap_RL.tex
Graphs chap RL

@@
 
+Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of system \cite{perera2021applications}.
+On top of it, this thesis investigate the learning episodes themselves to explore the field of possibilities to succeed the transition.
+
 Then, applying to this optimisation environment, \ie EnergyScope myopic Pathway,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework.
@@
 
-\section{Definition of the actions, states and rewards}
+\section{Definition of the actions, reward and states}
 \label{sec:RL:act_states_rew}
@@
 Out the total \gls{GHG} emissions in Belgium in 2020,oil (\ie so-called \gls{LFO} in the model), on the one hand and, on the other hand, fossil gas, account for roughly 40\% and 31\%, respectively.
-Then, even though its use in 2020 is much more limited compared to the two formers, \ie 28\,TWh of solid fossil fuels (\ie so-called COAL in the model) versus 159 and 142\,TWh for oil and fossil gas, respectively, coal is a cheap, about 0.017€/GWh, and highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
-For these reasons, three additional actions to support the strict limitation of overall emissions of the first action: limiting the consumption of these three fossil resources up to the level of consumption in 2020, $\textbf{Cons\textsubscript{fossil gas}}(2020)$, $\textbf{Cons\textsubscript{LFO}}(2020)$ and $\textbf{Cons\textsubscript{coal}}(2020)$,  over the entire concerned time window, except the first one as this year is the initial condition of the time window and cannot be optimised any more:
+Then, even though its use in 2020 is much more limited compared to the two formers, \ie 28\,TWh of solid fossil fuels (\ie so-called COAL in the model) versus 159 and 142\,TWh for oil and fossil gas, respectively, coal is a cheap, 17€/MWh, and highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
+For these reasons, three additional actions support the strict limitation of overall emissions of the first action: limiting the consumption of these three fossil resources up to the level of consumption in 2020, $\textbf{Cons\textsubscript{fossil gas}}(2020)$, $\textbf{Cons\textsubscript{LFO}}(2020)$ and $\textbf{Cons\textsubscript{coal}}(2020)$,  over the entire concerned time window, except the first one as this year is the initial condition of the time window and cannot be optimised any more:
 
@@
 The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy.
-Even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis.
+Electricity imported from abroad is not considered in the set of renewable energy carriers even if it can be assumed to be fully renewable by 2050.
+Finally, even though energy efficiency is usually defined as the ratio between the \gls{FEC} and the primary energy mix, we decided to define this efficiency with a focus on the \gls{EUD}, like in the rest of this thesis.
 Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio.
@@
 
-During the learning process, the algorithm explores numerous transition pathways (2439 successful transitions out of 10,565 attempts).
+During the learning process, the algorithm explores numerous transition pathways: 2037 successful transitions out of 10,751 attempts.
 Each pathway provides valuable insight into the best course of actions --- the primary goal of reinforcement learning.
@@
 
-Over the entire learning process, 23\% of the episodes have been successful versus 73\% of failures, the rest being infeasible problems where the actions taken by the agent are too constraining for the model (see Section \ref{sec:RL:act_states_rew}).
-Besides these infeasible cases, there is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}).
+There is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}).
 This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive.
@@
 Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
-Beyond this point, through this learning process, the chances to succeed the transition are 47\%.
+Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
 In other words, mid-term actions are necessary to hope succeeding the transition.
@@
 Right hand side details at the end of which time-window the failure occurred.
-The ``tipping year'' is 2040 as failing the transition by 2040 represents 58\% of all the failures.
-Beyond this point, through this learning process, succeeding the transition represents 47\% of the cases.}
+The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
+Beyond this point, through this learning process, succeeding the transition represents 38\% of the cases.}
 \label{fig:reward_status}
@@
 
-To identify the actions that have an actual impact on the environment, we looked if they were binding or not.
-In a \gls{LP} problem, constraints represent straight lines in the domain of variables.
-When this problem is bounded and feasible, these lines are the edges of a convex polygon, the domain of feasibility (see Figure \ref{fig:Binding_constr}).
+To identify the actions that have an actual impact on the environment, we can check if they were binding or not.
+In a \gls{LP} problem, constraints represent hyperplanes in the domain of variables.
+In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}).
+When the problem is bounded and feasible, these lines are the edges of a convex polygon, the domain of feasibility.
 The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function.
@@
 In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}.
-Among This vertex is t
+The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal.
+In other words, binding constraints, when tightened, aggravate the objective value function.
+If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.
 
@@
 \end{figure}
+
+After filtering out failures of the learning episodes and keeping only the successful transitions, only a limited set of the actions are binding and have an actual impact on the result of the optimisation in EnergyScope Pathway (see Figure \ref{fig:Binding_learning}).
+This allows identifying key actions to support the myopic transition.Limiting the \gls{GWP} in the near-term is a key-factor for success.
+However, this action has an effective impact on the environment only when it forces the system to be close to carbon-neutrality.
+The range over which limiting the use of fossil gas binds the optimisation is wider.
+Compared to other non-renewable fuels, this is due to the longer use of this energy carrier favoured by its low \gls{GWP} (the second after uranium) and its versatility (applications in the electricity, heat and mobility sectors).
+
+When it comes to limiting the use of \gls{LFO} and coal, the conclusions are more straightforward.
+At the beginning of the transition, most of the 159TWh of \gls{LFO} are consumed by naphtha-cracker (46\%) and decentralised boilers (45\%).
+The remaining 10\% are consumed by industrial boilers.
+Even though \gls{LFO} represents  30\% of the primary energy mix in 2020, the cost-optimum removes it of the mix without requiring the action of the agent.
+Naphtha-crackers, decentralised and industrial boilers get substituted by \gls{MTO}, decentralised \gls{HP} and industrial resistors and \gls{CHP}, respectively.
+This ``non-bindness'' of limiting \gls{LFO} is an indication that this action could be removed from the agent's levers of action without impacting the optimisation of its policy.
+
+On the contrary, limiting coal is always binding.
+Before all, this is due because coal is a cheap resource (17€/MWh).
+In other words, the cost-driven environment will favour it.
+Then, as the maximum amount of coal (28\,TWh) is much smaller than fossil gas and \gls{LFO}, high value of $\mathrm{act}\textsubscript{coal}$ still represents small consumption of coal.
 
@@
 The comparison would go over different aspects:
-%\begin{itemize}
-%\item Over-cost
-%\item Over-change, a bit like what Paolo defined as design error/change (in terms in consumed resources, installed capacities,...)
-%\end{itemize}
 

### 2024-04-15 chap_RL/chap_RL.tex
RL chapter: State exploration

@@
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
-Beyond this point, through this learning process, succeeding the transition represents 38\% of the cases.}
+Beyond this point, through this learning process, succeeding the transition represents 38\% of the episodes.}
 \label{fig:reward_status}
@@
 
-\myparagraph{States: Renewables and efficiency}\\
+\myparagraph{States}\\
 
-In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
+The first two dimensions of the state space are the cumulative emissions and costs.
+They drive the value of the reward and, consequently, the optimisation of the agent's policy.
+Per definition, the threshold of 1.2\,Gt\ce{CO2} splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
+In the successful transitions, the median cumulative emissions, $\text{P}_{50}$,are about 0.9\,Gt\ce{CO2}.
+This is due to the combination of efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
+Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt\ce{CO2}, 12\% higher than the \ce{CO2}-budget.
+As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
+Where 98\% of the failures were below the \ce{CO2}-budget in 2035, only 37\% passed this threshold in 2040.
+This reminds the importance of near-term actions to hope for a successful transition.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{RE_in_mix.pdf}
-\caption{Exploration of the state space over the learning process with a focus on the share of renewable energy carriers in the primary energy mix.
+\includegraphics[width=0.8\textwidth]{Cum_gwp_cost.pdf}
+\caption{Exploration of the state space over the learning process: cumulative emissions and costs.
+Successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
+}
+\label{fig:Cum_gwp_cost}
+\end{figure}
+
+Given the shape of the reward (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.
+The skewness of the cumulative emissions and costs in 2050 are indications of this shape of reward (see Table \ref{tab:skewness_gwp_cost}).
+When succeeding the transitions, the cumulative emissions and costs have a negative and positive skewness, respectively.
+On the contrary, when it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).
+
+\begin{table}[htbp!]
+\caption{Skewness of cumulative emissions and costs in 2050.
+Cumulative emissions are skewed to the left and to the right for the successes and failures, respectively.
+The skewness of the cumulative costs for successful transitions are compared to failures.On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
+\label{tab:skewness_gwp_cost}
+\centering
+\begin{tabular}{l c c}
+\toprule
+\textbf{Status of episode}  & \textbf{Skewness of cumulative} & \textbf{Skewness of cumulative} \\
+\textbf{in 2050}  & \textbf{emissions} & \textbf{costs} \\	
+\midrule
+Success & -0.52 & 0.50 \\
+Failure & 0.61 & 0.24 \\
+
+\bottomrule
+
+\end{tabular}
+\end{table}
+
+The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
+From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition.
+For the successful occurrences to exceed failures, this share increases to 54\% in 2025.
+Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}.
+In 2050, the success-failure threshold has been observed at 82\%.
+In the REF case of Chapter \ref{chap:atom_mol}, the share of renewables reached 86\% by 2050.
+However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another threshold at lower share of renewables in the mix.
+This area corresponds to the possibility to install \gls{SMR}.
+As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold to 60\% as in the SMR case of Chapter \ref{chap:atom_mol}.
+Besides these milestones to respect the \ce{CO2}-budget of the transition, one can also look at the other side of the thresholds.
+Below the near-term threshold of 60\%, this the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.
+
+The efficiency, as defined in Section \ref{sec:RL:act_states_rew}, gives less valuable information to direct towards successful transition.
+Through the transition, besides the share of success increasing over the failures, these two scenarios indistinguishably spread over the whole range.
+Similarly than the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{RE_in_mix_Efficiency.pdf}
+\caption{Exploration of the state space over the learning process: share of renewable energy carriers in the primary energy mix and efficiency.
 Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
-Below a near-term threshold (about 60\%), the chances of success are limited, \ie no-go zones.}
-\label{fig:RE_in_mix}
+Below a near-term threshold (about 60\%), the chances of success are limited, \ie no-go zones.
+Efficiency is a less valuable information for the agent to succeed transitions as failures and successes indistinguishably spread over the whole range.}
+\label{fig:RE_in_mix_Efficiency}
 \end{figure}
@@
 \includegraphics[width=0.8\textwidth]{Actions_learning.pdf}
-\caption{Over the whole learning process, actions taken by the agent.
-Besides the wide exploration of the action space, it does not seem to be any clear set of actions to take to support successful transitions.}
+\caption{Besides the wide exploration of the action space, successes and failures indistinguishably spread over the whole ranges of actions.
+In other words, it does not seem to be any clear set of actions to support successful transitions.}
 \label{fig:Actions_learning}
@@
 \includegraphics[width=0.8\textwidth]{Binding_learning.pdf}
-\caption{Out of the successful transitions, binding and not binding actions taken by the agent.
-Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
+\caption{Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
 Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.

### 2024-04-15 chap_RobPol/chap_RobPol.tex
RobPol chapter: comparison of projections

@@
  \midrule
-$\text{PC}_{\text{transition,3}}$ & $\text{PC}_{\text{2025,2}}$ & 4.3\\
+$\text{PC}_{\text{transition,4}}$ & $\text{PC}_{\text{2025,2}}$ & 4.3\\
  \midrule
@@
 
-\subsection{The robust roadmap}
+\subsection{A robust roadmap?}
 \label{subsec:RobPol:Rob_roadmap}
@@
 
-In the transport of freight, the ROB roadmap transition from diesel to \gls{FC} trucks with the use of electric trucks from 2025 to 2035.
-Similarly to the industrial electric heaters, it is a way to integrate to early installation of solar \gls{PV} and absorb part of its intermittent electricity.
+In the transport of freight, the ROB roadmap transitions from diesel to \gls{FC} trucks with the temporal use of electric trucks from 2025 to 2035.
+Similarly to the industrial electric heaters, it is a way to integrate the early installation of solar \gls{PV} and absorb part of its intermittent electricity.
 Freight transport via trains and boats are not different from the REF roadmap.
@@
 Technologically speaking, by 2050, the ROB case accounts for an additional 1.2~GW of \gls{MTO} to supply the extra 10.3~TWh of \gls{HVC}.
-Besides this, the extra 2.5~TWh and 0.4~TWh of e-ammonia and e-methanol are supplied by supplied by the import of their respective energy carrier.\\
+Besides this, the extra 2.5~TWh and 0.4~TWh of e-ammonia and e-methanol are supplied by the import of their respective energy carrier.\\
 
@@
 This aims at assessing the adaptation of the roadmaps as uncertainty gradually unfolds over time \cite{moret2020overcapacity}, which brings more realism than perfect foresight \cite{poncelet2016myopic}.
-The three roadmaps are tested 500 times fixing the installed capacities $f_{\mathrm{max}}\geq\textbf{F}\geq\textbf{F}^*$ for all the end-use technologies, and taking values of the 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}) following the sampling technique detailed in Section \ref{subsec:pce}, \ie Sobol' sequence.
+The three roadmaps are tested 500 times fixing the installed capacities $f_{\mathrm{max}}\geq\textbf{F}\geq\textbf{F}^*$ for all the end-use technologies, and taking values of the 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}) following the Sobol' sequence.
 However, uncertain parameters limit the potential installed capacity, $f_{\mathrm{max}}$, of \gls{PV} panels, onshore and offshore wind turbines and \gls{SMR}.
 For these, in case the installed capacity prescribed by the roadmap is higher than the maximum potential affected by the value of the uncertain parameter, the actual installed capacity is set to this new limit.
-In reality, this would be similar to revise to a smaller extent the expected roadmap due to external events (\eg smaller public acceptance \cite{zoellner2008public,sam2014small}, lower \gls{TRL} or longer installation time than expected)
+In reality, this would be similar to revise to a smaller extent the expected roadmap due to external events (\eg smaller public acceptance \cite{zoellner2008public,sam2014small}, lower \gls{TRL} or longer installation time than expected).
+
+No tested roadmap is more robust than the others on all the \gls{PCs} of transition (Table \ref{tab:projection}).
+Overall, the roadmaps follow the same trend over the different PCs of transition: they span over the most narrow range for PC 3 and over the widest range for PC 2.
+In practice, it means that roadmaps resulting from perfect foresight optimisation, when tested through a myopic process, would be more robust towards the variations driven by the freight transport sector (\ie \gls{FC} and diesel trucks and diesel boat) and less robust towards the variation driven by the decentralised heating sector (\ie decentralised \gls{HP} and wood and gas boilers) and \gls{PV}.
+
+\begin{table}[htbp!]
+\caption{95\% confidence range of projected data from roadmaps tested in myopic optimisation [$\cdot 10^{-3}$].
+Values in brackets give the relative difference with the REF roadmap.
+There is no ``best overall'' roadmap.}
+\label{tab:projection}
+\centering
+\begin{tabular}{c| c| c c}
+\toprule
+\multirow{2}{*}{\textbf{PC}} & \multicolumn{3}{c}{\textbf{95\% confidence range}}\\
+& REF & SMR & ROB \\
+\midrule
+1 & 4.2 & 4.2 (=) & 3.7 (\textbf{-12\%}) \\
+2 & 5.7 & 5.6 (-2\%) & 6.1 (\textbf{+8\%})\\
+3 & 2.2 & 2.3 (+3\%) & 1.9 (\textbf{-17\%}) \\
+4 & 4.5 & 4.4 (-2\%) & 4.5 (+1\%) \\
+5 & 2.7 & 3.1 (\textbf{+15\%}) & 2.6 (\textbf{-4\%})\\
+
+\bottomrule
+
+\end{tabular}
+\end{table}
+
+The SMR roadmap spans over similar ranges as the REF case.
+The most significant difference concerns PC 5.
+This PC is driven by the variations in the \gls{HVC} sector.
+As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}.
+Consequently, in the near-term, the SMR roadmap presents more variation in terms of installed capacity of \gls{MTO}.
+
+The ROB roadmap is more robust than the REF case on PCs 1, 3 and 5, by 12\%, 17\% and 4\% respectively.
+About PC 1, it is driven by the variation of installed capacities of \gls{PV}.
+In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap more robust towards these variations.
+Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
+Finally, as the ROB roadmap accounts for additional capacities to supply the potential additional demand of \gls{HVC}, it is more robust towards the variations in this sector.
+
+On the contrary, the ROB roadmap is less when considering PC 2, driven by the variations in the \gls{LT} and \gls{HT} heat sectors.
+The biggest impact comes from the variations of industrial resistors.
+In the ROB case, even though they are more installed at earlier stages of the transition to support the integration of \gls{PV}, a part of them are either prematurely decommissioned or not renewed.
+This makes this roadmap more sensitive to additional industrial heaters needed in the myopic optimisation at the latest stages of the transition.
 

### 2024-04-16 chap_RobPol/chap_RobPol.tex
RobPol chapter: submission for review

@@
 
-\subsection{Principal components of the transition}
+\subsection{Principal components of the perfect foresight REF transitions}
 \label{subsec:RobPol:PC_transition}
@@
 
+\newpage
 \section{Robustness assessment of pathway roadmaps}
@@
 Consequently, costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies are considered at their worst values, \ie the upper bound of their respective range.
-In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{LT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
+In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{LT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
+
+Besides the additional operational costs due to more expensive energy carriers, this case presents cumulative investments over the transition that are 2.6\% (17\,b€) higher than the REF case.
 The following paragraphs investigate the differences in the different sectors between the ROB and the REF cases, similarly to the comparison carried out with the SMR case in Chapter \ref{chap:atom_mol}.\\
@@
 
-\section{Discussion}
-\label{sec:RobPol:Discussion}
+\subsection{Principal components of myopic transitions}
+\label{subsec:RobPol:PC_MY_transition}
+When computing the \gls{PCs} of the different myopic transitions, the ROB case has a variance that is 10\% smaller than the one of the REF case.
+Moreover, we observe that the variations of \gls{PV}, industrial resistors and \gls{LT}-heat technologies still have a major contribution on the variation of the system roadmaps (see Figure \ref{fig:Top5_PCtransition_cases}).
+More surprising is the contribution of the variation of the technology to produce methanol from methane.
+Where methanol is preferably imported from abroad (or produced from biomass) in the perfect foresight approach, the myopic optimisation opts for this technology in the near-term (2025) to compete with the imports as the main supplier of this energy carrier.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.49\textwidth]{Top5_PCtransition_REF_MY.pdf}
+\includegraphics[width=0.49\textwidth]{Top5_PCtransition_SMR_MY.pdf}
+\end{figure}
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.49\textwidth]{Top5_PCtransition_ROB_MY.pdf}
+\caption{Contribution of the top-5 technologies to the \gls{PCs} of myopic transitions for the REF, SMR and ROB cases.
+The variation of \gls{PV}, industrial electric resistors and \gls{LT}-heat technologies still have a major contribution of the variation in the system roadmaps.
+At the early stages of the myopic transitions, the installation methane-to-methanol is also subject to significant variations.}
+\label{fig:Top5_PCtransition_cases}
+\end{figure}
+
+\section{Conclusions}
+\label{sec:RobPol:Conclusions}
+In this chapter, we applied the \gls{PCA}-based approach detailed in Chapter \ref{chap:chap_methodo} to assess the robustness of different transition roadmaps for the case of Belgium.
+First, starting from the principal components of each representative year, we have computed the PCs of the transition to serve as robustness metric.
+This showed that the variation of installed capacities of \gls{PV}, industrial electric resistors and \gls{LT}-heat technologies drove the major share (75\%) of the variation of the transition design strategy.
+
+Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB.
+Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}.
+From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation the additional capacities needed in myopic optimisations subject to uncertainties.
+This showed that counting on the installation of \gls{SMR} by 2040 was not providing more or less robustness than the reference case.
+However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario.
+Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} (12\%) and technologies in the transport of freight (17\%).
+Overall, this \gls{PCA}-based approach allows evaluating the variation of multiple-year transition pathway of a whole-energy system and assessing its robustness.
+Doing so, it brings more information than when considering only the variation of the objective function.

### 2024-04-16 chap_atom_mol/chap_atom_mol.tex
RobPol chapter: submission for review

@@
 
-\section{Discussion}
-\label{sec:atom_mol:discuss}
+\section{Conclusions}
+\label{sec:atom_mol:conclusions}
 Given its lower \gls{LCOE} (40\,€/MWh$_{\text{elec}}$) and the important share of the investment recovered via its salvage value (between 79\% and 96\%), \gls{SMR} is installed as soon as available.

### 2024-04-16 chap_case_study/chap_case_study.tex
RobPol chapter: submission for review

@@
 
-\section*{Conclusion}
-\label{sec:cs:conclusion}
+\section{Conclusions}
+\label{sec:cs:conclusions}
 Mainly based on data collected during previous works \cite{Moret2017PhDThesis,limpens2021generating}, this thesis has developed the case study along different axes.

### 2024-04-18 chap_RL/chap_RL.tex
RobPol chapter: submission for review
Figures RL comparaison

@@
 Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive.
-This is another indication that the shape of the reward is appropriate in this exploration of transition pathways.
+This is another indication that the shape of the reward is appropriate in this exploration of successful transition pathways.
 
@@
 Cumulative emissions are skewed to the left and to the right for the successes and failures, respectively.
-The skewness of the cumulative costs for successful transitions are compared to failures.On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
+The skewness of the cumulative costs for successful transitions is higher compared to failures.On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
 \label{tab:skewness_gwp_cost}
@@
 
-\begin{itemize}
-\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
-\end{itemize}
-
-\section{Testing and comparison with references}
+\section{Comparison with references and testing}
 \label{sec:RL:testing}
@@
 
+\begin{itemize}
+\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
+\end{itemize}
 %\subsection{Training}
@@
 
-\section{Discussion}
-\label{sec:RL:discussion}
+\section{Conclusions}
+\label{sec:RL:conclusions}
 

### 2024-04-18 chap_RobPol/chap_RobPol.tex
Figures RL comparaison

@@
 \label{subsec:RobPol:Projection}
-After defining the robustness metric and describing the roadmaps to assess (\ie REF, SMR and ROB), the final step consists in testing these roadmaps under uncertainties in a myopic pathway optimisation.
+After defining the robustness metric and describing the roadmaps to assess (\ie REF, SMR and ROB), the final step consists in testing these roadmaps under uncertainties in myopic pathway optimisations.
 This aims at assessing the adaptation of the roadmaps as uncertainty gradually unfolds over time \cite{moret2020overcapacity}, which brings more realism than perfect foresight \cite{poncelet2016myopic}.
 The three roadmaps are tested 500 times fixing the installed capacities $f_{\mathrm{max}}\geq\textbf{F}\geq\textbf{F}^*$ for all the end-use technologies, and taking values of the 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}) following the Sobol' sequence.
-However, uncertain parameters limit the potential installed capacity, $f_{\mathrm{max}}$, of \gls{PV} panels, onshore and offshore wind turbines and \gls{SMR}.
+However,  some uncertain parameters limit the potential installed capacity, $f_{\mathrm{max}}$, of \gls{PV} panels, onshore and offshore wind turbines and \gls{SMR}.
 For these, in case the installed capacity prescribed by the roadmap is higher than the maximum potential affected by the value of the uncertain parameter, the actual installed capacity is set to this new limit.
@@
 No tested roadmap is more robust than the others on all the \gls{PCs} of transition (Table \ref{tab:projection}).
-Overall, the roadmaps follow the same trend over the different PCs of transition: they span over the most narrow range for PC 3 and over the widest range for PC 2.
-In practice, it means that roadmaps resulting from perfect foresight optimisation, when tested through a myopic process, would be more robust towards the variations driven by the freight transport sector (\ie \gls{FC} and diesel trucks and diesel boat) and less robust towards the variation driven by the decentralised heating sector (\ie decentralised \gls{HP} and wood and gas boilers) and \gls{PV}.
+The SMR roadmap spans over similar ranges as the REF case.
+The most significant difference concerns PC 5.
+This PC is driven by the variations in the \gls{HVC} sector.
+As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}.
+As naphtha-cracker is the more economical option in the near-term of a myopic transition, this smoother modal shift leads to less variations of installed capacities.
+This makes the SMR roadmap more robust than the REF case.
+
+% Overall, the roadmaps follow the same trend over the different PCs of transition: they span over the most narrow range for PC 3 and over the widest range for PC 1.
+In practice, it means that roadmaps resulting from perfect foresight optimisation, when tested through a myopic process, would be more robust towards the variations driven by the freight transport sector (\ie \gls{FC} and diesel trucks and diesel boat) and less robust towards the variation driven by \gls{PV} and the electrification of the heating sectors.
 
 \begin{table}[htbp!]
-\caption{95\% confidence range of projected data from roadmaps tested in myopic optimisation [$\cdot 10^{-3}$].
-Values in brackets give the relative difference with the REF roadmap.
-There is no ``best overall'' roadmap.}
+\caption{Comparison of the robustness with the REF case where robustness is defined by the width of the 95\% confidence range of the additional installed capacities projected on the PC.
+Differences below 5\% are not shown ($\simeq$).}
 \label{tab:projection}
 \centering
-\begin{tabular}{c| c| c c}
+\begin{tabular}{c| c c}
 \toprule
-\multirow{2}{*}{\textbf{PC}} & \multicolumn{3}{c}{\textbf{95\% confidence range}}\\
-& REF & SMR & ROB \\
+\multirow{3}{*}{\textbf{PC}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
+& SMR & ROB \\
 \midrule
-1 & 4.2 & 4.2 (=) & 3.7 (\textbf{-12\%}) \\
-2 & 5.7 & 5.6 (-2\%) & 6.1 (\textbf{+8\%})\\
-3 & 2.2 & 2.3 (+3\%) & 1.9 (\textbf{-17\%}) \\
-4 & 4.5 & 4.4 (-2\%) & 4.5 (+1\%) \\
-5 & 2.7 & 3.1 (\textbf{+15\%}) & 2.6 (\textbf{-4\%})\\
+1 & $\simeq$ & +49\% \\
+2 &  $\simeq$ & +37\%\\
+3 &  $\simeq$ & +33\%\\
+4 & $\simeq$ & $\simeq$ \\
+5 &  +34\% & -21\%\\
 
@@
 
-The SMR roadmap spans over similar ranges as the REF case.
-The most significant difference concerns PC 5.
-This PC is driven by the variations in the \gls{HVC} sector.
-As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}.
-Consequently, in the near-term, the SMR roadmap presents more variation in terms of installed capacity of \gls{MTO}.
-
-The ROB roadmap is more robust than the REF case on PCs 1, 3 and 5, by 12\%, 17\% and 4\% respectively.
-About PC 1, it is driven by the variation of installed capacities of \gls{PV}.
-In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap more robust towards these variations.
+The ROB roadmap is more robust than the REF case all PCs but PC 4.
+About PC 1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors.
+In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations.
+For the same reasons, the ROB roadmap is 37\% more robust for PC 2.
+The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers.
+In myopic transitions though, this technology keeps on being installed in the mid-term.
 Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
-Finally, as the ROB roadmap accounts for additional capacities to supply the potential additional demand of \gls{HVC}, it is more robust towards the variations in this sector.
-
-On the contrary, the ROB roadmap is less when considering PC 2, driven by the variations in the \gls{LT} and \gls{HT} heat sectors.
-The biggest impact comes from the variations of industrial resistors.
-In the ROB case, even though they are more installed at earlier stages of the transition to support the integration of \gls{PV}, a part of them are either prematurely decommissioned or not renewed.
-This makes this roadmap more sensitive to additional industrial heaters needed in the myopic optimisation at the latest stages of the transition.
+Finally, the ROB roadmap is 21\% less robust for PC 5.
+Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller.
+As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.
 
@@
 From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation the additional capacities needed in myopic optimisations subject to uncertainties.
-This showed that counting on the installation of \gls{SMR} by 2040 was not providing more or less robustness than the reference case.
+This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation.
 However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario.
-Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} (12\%) and technologies in the transport of freight (17\%).
+Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
+
 Overall, this \gls{PCA}-based approach allows evaluating the variation of multiple-year transition pathway of a whole-energy system and assessing its robustness.

### 2024-04-19 chap_RL/chap_RL.tex
RL chapter: submission for review

@@
 Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of system \cite{perera2021applications}.
-On top of it, this thesis investigate the learning episodes themselves to explore the field of possibilities to succeed the transition.
+This thesis rather investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.
 
@@
 
-Out the total \gls{GHG} emissions in Belgium in 2020,oil (\ie so-called \gls{LFO} in the model), on the one hand and, on the other hand, fossil gas, account for roughly 40\% and 31\%, respectively.
+Out of the total \gls{GHG} emissions in Belgium in 2020,oil (\ie so-called \gls{LFO} in the model), on the one hand and, on the other hand, fossil gas, account for roughly 40\% and 31\%, respectively.
 Then, even though its use in 2020 is much more limited compared to the two formers, \ie 28\,TWh of solid fossil fuels (\ie so-called COAL in the model) versus 159 and 142\,TWh for oil and fossil gas, respectively, coal is a cheap, 17€/MWh, and highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
@@
 Failure & 0.61 & 0.24 \\
-
 \bottomrule
@@
 
-The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
+Finally, we observe that the majority of the successful transitions are cheaper than the reference transition cost, 1.1\,b€.
+Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is more increased than the one of electrofuels (see Table \ref{tab:param_RL}).
+Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases.
+On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and interest rate.
+These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2}-budget.
+
+\begin{table}[htbp!]
+\caption{Uncertain parameters impacting the most the total transition cost, their Sobol' index and, for the successful transitions, their mean of their values between 0 and 100\%, $\mu$, and their skewness, $\gamma$.
+On top of being supported by the agent's actions, successful transitions occur when the cost of purchasing fossil fuels is more increased than the one of electrofuels.}
+\label{tab:param_RL}
+\centering
+\begin{tabular}{l c c c}
+\toprule
+\textbf{Parameter}  & \textbf{$\mu$} & \textbf{$\gamma$}  \\	
+\midrule
+Purchase electrofuels & 50.4\% & 0.004  \\
+Industry EUD & 49.8\% & 0.026 \\
+Interest rate & 48.4\% & 0.089\\
+Purchase fossil fuels  & 55.0\% & -0.068\\
+\bottomrule
+
+\end{tabular}
+\end{table}
+
+Besides the cumulative emissions and costs, the agent also observes the share of renewable energy carriers in the primary mix and the efficiency of the system The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
 From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition.
@@
 The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function.
-Besides being within the domain of feasibility, it is proven that this optimal solution, when unique \footnote{There are cases where the objective function has the same optimal value along an entire edge.
+Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge.
 In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}.
@@
 
-\section{Comparison with references and testing}
+\section{Comparison with references}
 \label{sec:RL:testing}
+After investigating the exploration of transition pathways on the monthly model, this section compares these results with two references: the \gls{RL}-agent trained on the hourly model and the perfect foresight under uncertainties from the \gls{UQ} analysis presented in Chapter \ref{chap:atom_mol}.
+The first reference has been built by applying the same rules (see Section \ref{sec:RL:act_states_rew}) but on the hourly model over a more limited number of batches, 43 versus 80, leading to 928 versus 2037 successful transitions for the results presented in Section \ref{sec:RL:learning}.
+First, the purpose is to assess the capacity of the monthly model to be a good proxy of the hourly myopic transitions and, then,  compare the explored pathways with the perfect foresight under uncertainties.
 
-How does it react to potential shocks (e.g. no fuels from 2040) compared to classic myopic?
+Considering the actions taken by the agent and their binding effect on the environment,  we see similar trends as in the monthly model (see Figure \ref{fig:Binding_learning_TD}).
+Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation.
+Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon-neutrality becomes more crucial at the end of the transition.
 
-\citet{herman2014beyond} defined robustness as ``… the fraction of sampled states of the world in which a solution satisfies all performance requirements''
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Binding_learning_TD.pdf}
+\caption{When the agent learns on the hourly model, the actions are binding in the similar timing and similar ranges as when learning on the monthly model. }
+\label{fig:Binding_learning_TD}
+\end{figure}
 
-\begin{itemize}
-\item Assess the severity of learning on Monthly model, versus Hourly model. Compare the policies: one versus the other versus one followed by the other
-\end{itemize}
-%\subsection{Training}
-%Here present results RL-oriented (actions, states, reward)
-%
-%Then, present results energy system-oriented (installed capacities, costs, generation)
+\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2}-budget (see Figure \ref{fig:Gwp_pathway}).
+However, driven first by this \ce{CO2}-budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}).
+This comes from the agent's actions that limit the emissions and/or the consumption of fossil resources at the early stages.
+Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition.
+This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 9 or more Mt\ce{CO2} compared to 4 for the perfect foresight approach.
+These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% on average by 2050.
+The difference in the early drop of emissions between the monthly and hourly models comes from the easier integration of local \gls{VRES} as the hourly intermittency is averaged over a month (see Appendix \ref{app:mo_vs_td}).
+Besides this sharper decrease of emissions early in the transition, the monthly myopic pathways fit with the trend provided by \gls{RL} optimisation on the hourly model.
+Finally, the long-term vision of the perfect foresight approach results in smoother reduction of the emissions to end up with less emissions by 2050.
 
-%\subsection{Testing}
-%See how the different policies saved successively during the training behave when facing to new samples.
-This way, we can pick an optimal policy (the one giving the maximum average (or another metric) reward).
-%
-%We then compare the results of RL with the perfect foresight-TD deterministic and the perfect foresight-monthly with uncertainties (by setting the actions of the agent as variables in these two optimisations).
-The comparison would go over different aspects:
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Gwp_pathway.pdf}
+\caption{Comparison of \ce{CO2}-emissions pathways from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\label{fig:Gwp_pathway}
+\end{figure}
 
-\textbf{Confirm here with what is said in \cite{luderer2018residual}: 
-Importantly, our results also show that near-term policy stringency is an important driver of cumulative Res-FFI-CO2 in climate change mitigation scenarios.
-If strengthening of NDCs fails, Res-FFI-CO2 will be even higher, not only because of additional near-term emissions, but also due to a decrease of economic mitigation potentials in the longer term caused by further carbon lock-in.
-Delaying the strengthening of mitigation action will increase the world’s dependence on CDR for holding warming to well below 2°C, and is likely to push the 1.5°C target out of reach for this century.}
+The transition pathways of the annual system cost also show the advantage of the \gls{RL}-based exploration (see Figure \ref{fig:System_cost_pathway}).
+Via its actions and its cost-based reward, the agent can reach systems that are cheaper than any other solution obtained by the perfect foresight approach.
+Even more than for the emissions, the monthly model is a good proxy to assess the annual system cost of the pathway on the hourly model.
+The perfect foresight approach gives a wider variability in its cost since this method always found a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{System_cost_pathway.pdf}
+\caption{Comparison of annual system costs pathways from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\label{fig:System_cost_pathway}
+\end{figure}
+
+The analysis of the cumulative costs shows that the \gls{OPEX} is the main difference between myopic and perfect foresight transitions (see Figure \ref{fig:Opex_Capex_Salvage_comp}).
+Successful myopic transitions have a lower \gls{OPEX} thanks to the agent's actions and importing more electrofuels, especially e-ammonia that is more imported in the myopic optimisation to supply \gls{CCGT} in case \gls{SMR} is not available.
+
+We observe also similar distributions between the monthly and hourly myopic results that confirm the good approximation provided by the monthly approach.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.325\textwidth]{Opex_2050_comp.pdf}
+\includegraphics[width=0.325\textwidth]{Capex_2050_comp.pdf}
+\includegraphics[width=0.325\textwidth]{Salvage_2050_comp.pdf}
+\caption{Comparison of cumulative OPEX (top-left), CAPEX (top-right) and salvage value (bottom) in 2050 from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\label{fig:Opex_Capex_Salvage_comp}
+\end{figure}
 
@@
 \label{sec:RL:conclusions}
+In the literature, two options are investigated to explore transition pathways of a whole-energy system.
+On the one hand, in perfect foresight optimisation, a full knowledge of the parameters is assumed over the entire time horizon.
+On the other hand, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon, to add more realism \cite{poncelet2016myopic}.
+To meet objective of climate change mitigation, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
+However, since the effect of \ce{CO2}-emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself.
+Consequently, we have applied the \acrfull{RL} approach on the environment of the Belgian myopic pathway optimisation under uncertainties.
 
-Potentially interesting to implement reward shaping to accelerate learning or guide the agent towards achieving the desired behaviour more efficiently.
-In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode.
-However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy.
-Reward shaping addresses this issue by providing additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights.
-These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective.
-However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).
+In \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
+The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt\ce{CO2} equivalent to 10 years at the current level of emissions.
+A secondary focus is also put on aiming at minimum transition cost.
+To know its progress through the transition, the agent had the information of the cumulative emissions and costs as well the share of renewable energy carriers in the primary energy mix and the system overall efficiency.
 
-\textbf{Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity}
+This \gls{RL}-based exploration pointed out that short-term actions were needed to hope succeeding such a transition, as also demonstrated by \citet{luderer2018residual}.
+Where \gls{LFO} becomes less cost-competitive in the near-future, limiting the use of coal should be done at any cost.
+Then, fossil gas should be replaced by e-methane in the mid-term while limiting the overall emissions becomes more effective by the end of the transition.
+The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances to succeed the transition.
+Below 54\% of renewables in the mix in the near-future, these chances become much more limited, \ie the no-go zone.
+
+Finally, we have compared the results coming from this \gls{RL}-based myopic optimisation with monthly time-resolution with the hourly perfect foresight approach.
+These myopic optimisations provided pathways respecting the \ce{CO2}-budget that were more drastic in cutting emissions in the short-term than the perfect.
+Supported by the agent's actions, we could reach cheaper energy system than what the perfect foresight could give.
+With the advantage of the smaller computational burden, the monthly model myopic model was also found to be a good proxy with its equivalent based on hourly time-resolution.
+
+In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2}-budget target.
+It also highlighted the actions to take to effectively support a whole-energy system transition.

### 2024-04-19 chap_intro_ccl/conclusion.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR
Graphs chap RL
RL chapter: submission for review

@@
 It's better for the system to keep unused assets to recover part of their salvage value than prematurely decommissioning them.
+\item Progressive scale-up of \gls{VRES} instead of full potential possible from 2025.
 \end{itemize}
@@
 \begin{itemize}
+\item Electrofuels is the unicorn solution. What about carbon removal
 \item \textbf{Word about sufficiency} oui mais c'est sans doute celle qui est le plus enviable car la moins risqué à tenter d'atteindre, les solutions mirages technologiques, si on y croit trop on met tout la dessus et si ca foire, on est encore plus dans la merde pcq la direction est mauvaise, la solution sobriété, meme si jamais atteint les efforts pr l'atteindre ne seront pas contre productif
@@
 \item Extract a roadmap representative of a multiple-run UQ or RL process.
+\item In this work, we only consider emissions related to the use of energy carriers and not the construction --> LCA
+\item Multi-objective optimisation
+\item Potentially interesting to implement reward shaping to accelerate learning or guide the agent towards achieving the desired behaviour more efficiently.
+In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode.
+However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy.
+Reward shaping addresses this issue by providing additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights.
+These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective.
+However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).
+
+\item Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity.
+Assess the severity of learning on Monthly model, versus Hourly model.
+Compare the policies: one versus the other versus one followed by the other
 \end{itemize}

### 2024-04-20 chap_RL/chap_RL.tex
RL chapter: analysis about import of electrofuels

@@
 Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is more increased than the one of electrofuels (see Table \ref{tab:param_RL}).
+In other words, to have higher chances to succeed a myopic transition, the key factor is to reduce the uncertainty on the cost of purchasing electrofuels.
 Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases.
@@
 
-Considering the actions taken by the agent and their binding effect on the environment,  we see similar trends as in the monthly model (see Figure \ref{fig:Binding_learning_TD}).
+Considering the actions taken by the agent and their binding effect on the environment with a hourly resolution,  we see similar trends as in the monthly model (see Figure \ref{fig:Binding_learning_TD}).
 Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation.
@@
 The analysis of the cumulative costs shows that the \gls{OPEX} is the main difference between myopic and perfect foresight transitions (see Figure \ref{fig:Opex_Capex_Salvage_comp}).
-Successful myopic transitions have a lower \gls{OPEX} thanks to the agent's actions and importing more electrofuels, especially e-ammonia that is more imported in the myopic optimisation to supply \gls{CCGT} in case \gls{SMR} is not available.
-
-We observe also similar distributions between the monthly and hourly myopic results that confirm the good approximation provided by the monthly approach.
+Supported by the agent's actions, successful myopic transitions have a lower \gls{OPEX} than the perfect foresight ones.
 
@@
 \label{fig:Opex_Capex_Salvage_comp}
+\end{figure}
+
+The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}.
+The assessment of the primary energy mix by the 2050 highlights where the differences of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
+In the majority of the cases, e-ammonia is twice more imported in the myopic transitions.
+This is even more the case for the myopic transitions carried out on the hourly model.
+Being cheaper than e-methane (see Chapter \ref{chap:case_study}), e-ammonia brings flexibility in the production of electricity via \gls{CCGT} (see Chapter \ref{chap:atom_mol}).
+Besides the slightly favourable economical conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge in the future, and, among others, the availability of \gls{SMR}.
+This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt\ce{CO2} (see Figure \ref{fig:Cum_gwp_cost}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Mix_2050_comp.pdf}
+\caption{Comparison of the primary energy mix in 2050 from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.
+The biggest difference is about e-ammonia to supply \gls{CCGT}}
+\label{fig:Mix_2050_comp}
 \end{figure}
@@
 
+With the advantage of the smaller computational burden, the monthly model myopic model was found to be a good proxy with its equivalent based on hourly time-resolution.
+
 Finally, we have compared the results coming from this \gls{RL}-based myopic optimisation with monthly time-resolution with the hourly perfect foresight approach.
@@
 Supported by the agent's actions, we could reach cheaper energy system than what the perfect foresight could give.
-With the advantage of the smaller computational burden, the monthly model myopic model was also found to be a good proxy with its equivalent based on hourly time-resolution.
+This comparative analysis also pointed out that investing more into renewable electrofuels was the option when knowledge about the future is limited.
+To do so, it requires reducing the uncertainties of their cost of purchasing.
 

### 2024-04-20 chap_RobPol/chap_RobPol.tex
RL chapter: analysis about import of electrofuels

@@
 \label{subsec:RobPol:PC_year}
-After the pre-preprocessing of the raw data (\ie data scaling and outliers management, see Section \ref{subsec:meth:PCA:transition}), the \gls{PCs} of each representative year of the transition, except 2020 as the initialisation year, can be computed.
+Before computing the \gls{PCs} of each representative year of the transition, we pre-preprocessed the raw data though scaling and managing the outliers detailed in Section \ref{subsec:meth:PCA:transition}.
+The second step discarded decentralised \gls{HP} running on methane.
+Without managing the outliers, this technology has a contribution in the \gls{PCs} of transition for several reasons.
+First of all, the \gls{LT} heating sector represents between 22\% and 27\% of the \gls{EUD} in the case of Belgium.
+Being part of this major sector, the scaling of decentralised gas \gls{HP} is higher than technologies in other sectors.
+Then, given the \gls{LP} implementation of EnergyScope Pathway, the gas heat pumps completely substitute electric heat pumps in specific conditions (less electricity and cheaper methane).
+However, theses cases represent less than 10\% of the 1260 runs.
+For these reasons, we have decided that managing the outliers was needed to assess the main directions of variation of the transition.
 

### 2024-04-20 chap_intro_ccl/conclusion.tex
RL chapter: analysis about import of electrofuels

@@
 %!TEX encoding = UTF-8 Unicode
+
+The general objective of this thesis has been to investigate the impact of uncertainties on the optimisation of the transition pathway of a whole-energy system that accounts for multiple energy carriers and multiple sectors, \ie electricity, high and low temperature heat, passenger and freight transport and non-energy demand.
+
 I took here the same sections as in Gauthier's thesis

### 2024-04-20 chap_methodo/chap_methodo.tex
RL chapter: State exploration
RL chapter: analysis about import of electrofuels

@@
 \begin{table}[htbp!]
-\caption{Total Sobol' indices of the uncertain parameters over the total transition cost via \gls{PCE} of orders 2 and 1.
+\caption{Total Sobol' indices of the uncertain parameters over the total transition cost via \gls{PCE} of orders 2 and 1 having a \gls{LOO} error of 1\% and 16\%, respectively.
 The similar rankings (and indices) show the validity of using the faster (even though less accurate) lower order \gls{PCE} to rank the most impacting uncertain parameters.}
@@
 \label{sec:meth:PCA}
-When optimizing a transition pathway of a whole-energy system, including its uncertainties, capturing the most variable changes of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
-For the case study detailed in Chapter \ref{chap:case_study}, this consists of \ie 7 representative years of the transition (\ie from 2020 to 2050), 113 possible technologies subject to uncertain parameters.
+When optimizing a transition pathway of a whole-energy system, including its uncertainties, capturing the variance of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
+For the case study detailed in Chapter \ref{chap:case_study}, this consists of 113 possibles technologies over 7 representative years.
 To tackle this challenge, we have developed a methodology based on the \acrfull{PCA}.
-The philosophy behind this approach is to identify key-combinations of design variables giving relevant dimensions to compare different systems, beyond their sole objective function, \ie total transition cost, but without having to compare each design variable individually.
+The philosophy behind this approach is to identify key-combinations of design variables capturing most of the system variance, beyond their sole objective function, \ie total transition cost, but without having to compare each design variable individually.
 This methodology provides two main outputs.
-First and foremost, using the model runs necessary to quantify the impact of the uncertain parameters on the total cost of the transition (see Section \ref{sec:meth:UQ}), it gives a metric on which to assess the robustness of energy transition policies resulting from different approaches.
-This metric gives more insight that, for instance, the variation of the total transition cost that encompasses too many aspects (\ie design and operation strategies, variation along the transition) in one single value.
+First and foremost, using the model runs necessary to quantify the impact of the uncertain parameters on the total cost of the transition (see Section \ref{sec:meth:UQ}), it gives a metric on which to assess the robustness of different energy transition policies.
+Where the total transition cost is a single value representing the entire system and analysing individually all technologies would be overwhelming, this metric gives the happy medium.
 Second, these ``directions of variation'' can highlight key modal shifts or highly varying design strategies over the transition.
-After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to give these ``directions of variation'' and to assess the robustness of policies.
+After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of policies.
 
@@
 Born in the early 20th century \cite{pearson1901on,hotelling1933analysis}, the \acrfull{PCA} finds its fundamentals from the \gls{SVD}.
-\gls{SVD} is a generalization, to an arbitrary (\ie not especially square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
-The core concept of principal component analysis (PCA) involves simplifying a dataset with numerous interconnected variables by reducing its dimensionality.
-The aim is to preserve as much variability within the data as feasible.
+\gls{SVD} is a generalization, to an arbitrary (\ie not necessary square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
+The core concept of principal component analysis (PCA) involves simplifying the original dataset with numerous interconnected variables by reducing its dimensionality while preserving as much variability within the data.
 This is accomplished by transforming the $p$-dimension data, $\mathbf{x}$, into a new set of variables called principal components (PCs), $\mathbf{z}$.
-These components are uncorrelated and arranged in such a way that the first ones retain the majority of the variability found in all of the original variables.
+These components are uncorrelated and arranged in such a way that the first ones retain the majority of the variability found in the original dataset.
 On the other hand, the final principal components (PCs) pinpoint directions where there is minimal variation, indicating nearly constant linear relationships among the original variables \cite{jolliffe2002principal}.
 
-The PCs are computed based on the covariance matrix of $\mathbf{x}$, $\mathbf{\Sigma}$ where the diagonal of this matrix gives the variance of the $i^{\text{th}}$ variable and the other elements give the covariance between the $i^{\text{th}}$ and the $j^{\text{th}}$ variables where $i\neq j$.
+The PCs are computed based on the covariance matrix of $\mathbf{x}$, $\mathbf{\Sigma}$, where the diagonal of this matrix gives the variance of the $i^{\text{th}}$ variable and the other elements give the covariance between the $i^{\text{th}}$ and the $j^{\text{th}}$ variables where $i\neq j$.
 Out of this matrix, $\mathbf{\alpha_k}$ is the eigenvector of $\mathbf{\Sigma}$ corresponding to its $k^{\text{th}}$ highest eigenvalue $\lambda_k$.
-One crucial aspect of these eigenvectors is their normalization, \ie $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$ \cite{jolliffe2002principal}.
-This normalization has several objectives.
-Among them, this ensures orthogonality of the PCs ensuring that they represent independent directions in the original feature space.
-Then, normalizing the eigenvectors ensures that the magnitude of each eigenvector represents the importance or variance explained by its corresponding principal component.
+These eigenvectors are orthogonal to each other so that they represent independent directions in the original feature space.
+These eigenvectors are also normalized such that $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$ \cite{jolliffe2002principal}.
 This makes it easier to interpret the relative importance of each principal component in explaining the variability of the data.
@@
 
-More mathematically, the first \gls{PC}, $z_1$, is then a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $\mathbf{x}$ with maximum variance:
+Mathematically, the first \gls{PC}, $z_1$, is a linear function, $\mathbf{\alpha_1^{T}x}$,  of the different variables of $\mathbf{x}$ with maximum variance:
 
@@
 where $\mathbf{^T}$ means the transpose vector.
-Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and maximizing the variance.
+Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and capturing the remaining variance.
 These linear transformations can be seen as projection of the original data on the principal direction, \ie PCs (see Figure \ref{fig:PCA_intro_PCA}).

### 2024-04-21 chap_RL/chap_RL.tex
Conclusion: First half
Work (How?) + Results (What?)

@@
 Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
-In other words, mid-term actions are necessary to hope succeeding the transition.
+In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
 
@@
 They drive the value of the reward and, consequently, the optimisation of the agent's policy.
-Per definition, the threshold of 1.2\,Gt\ce{CO2} splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
-In the successful transitions, the median cumulative emissions, $\text{P}_{50}$,are about 0.9\,Gt\ce{CO2}.
+Per definition, the threshold of 1.2\,Gt$_{\ce{CO2},\text{eq}}$ splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
+In the successful transitions, the median cumulative emissions, $\text{P}_{50}$,are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
 This is due to the combination of efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
-Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt\ce{CO2}, 12\% higher than the \ce{CO2}-budget.
+Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$, 12\% higher than the \ce{CO2}-budget.
 As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
 Where 98\% of the failures were below the \ce{CO2}-budget in 2035, only 37\% passed this threshold in 2040.
-This reminds the importance of near-term actions to hope for a successful transition.
+This reminds the importance of near-term, 2025-2030, actions to hope for a successful transition.
 
@@
 Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition.
-This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 9 or more Mt\ce{CO2} compared to 4 for the perfect foresight approach.
+This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 9 or more Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
 These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% on average by 2050.
@@
 Besides the slightly favourable economical conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge in the future, and, among others, the availability of \gls{SMR}.
-This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt\ce{CO2} (see Figure \ref{fig:Cum_gwp_cost}).
+This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt$_{\ce{CO2},\text{eq}}$ (see Figure \ref{fig:Cum_gwp_cost}).
 
@@
 In \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
-The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt\ce{CO2} equivalent to 10 years at the current level of emissions.
+The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt$_{\ce{CO2},\text{eq}}$ equivalent to 10 years at the current level of emissions.
 A secondary focus is also put on aiming at minimum transition cost.

### 2024-04-21 chap_RobPol/chap_RobPol.tex
Conclusion: First half
Work (How?) + Results (What?)

@@
  \midrule
-$\text{PC}_{\text{transition,4}}$ & $\text{PC}_{\text{2025,3}}$ & 3.4\\
+$\text{PC}_{\text{transition,5}}$ & $\text{PC}_{\text{2025,3}}$ & 3.4\\
 \bottomrule
@@
 However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario.
-Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
+Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
 

### 2024-04-21 chap_intro_ccl/conclusion.tex
Conclusion: First half
Work (How?) + Results (What?)

--- 2024-04-19 chap_intro_ccl/conclusion.tex
@@
 %!TEX encoding = UTF-8 Unicode
-I took here the same sections as in Gauthier's thesis
-\section*{Thesis contributions}
-\addcontentsline{toc}{section}{Thesis contributions}
-Insist here on the methodological added value of the thesis
 
-\section*{Limitations}
-\addcontentsline{toc}{section}{Limitations}
+% WHAT WE DID (HOW?)
+The general objective of this thesis has been to investigate the impact of uncertainties and foresight on the transition pathway of a whole-energy system (\ie multi-energy carriers and multi-sectors) subject to \ce{CO2}-budget without a prescribed trajectory to reduce the emissions.
+We have pursued this objective on the case study of Belgium with a focus on the role of renewable electrofuels (\ie e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely-populated and highly industrialised country with limited local renewable potentials.
+To have 66\% chance of limiting warming to 1.5°C \cite{IPCC_CO2_budget} and using the ``grandfathering'' sharing principle of the global budget \cite{knight2013grandfathering}, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$ which represents 10 years of the current level of emissions.
+
+To optimise such a system, we have developed EnergyScope Pathway that minimise the total transition cost of the system from 2020 to 2050 and implemented the myopic approach.
+This approach considers a sequence of 10-year time windows with an overlap of 5 years between them rather than one global optimisation over the 30 years of the transition.
+This allows assessing the realism of the decision making process where knowledge about the future is progressively unfold rather than assumed complete from the beginning of the transition, the perfect foresight (PF) approach.
+
+The future of a national whole-energy system is subject to a plethora of uncertainties such as the cost of purchasing resources, the \acrfull{CAPEX} of technologies or the level of demands.
+In this thesis, we have used the uncertainty characterisation approach developed by \citet{moret2016strategic} to provide the range of values for the uncertain parameters.
+To propagate and quantify the impact of these uncertainties on the outputs of the model, we have used the \acrfull{PCE} method implemented in the Coppitters' RHEIA framework \cite{coppittersthesis}.
+
+To explore the myopic pathways respecting a \ce{CO2}-budget equivalent to 10 years of current emissions, without \ce{CO2}-trajectory, we have used the \acrfull{RL} approach.
+In this framework, an agent took actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
+Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
+Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
+Repeating the whole transition with different sequences of actions-states allow the agent to come up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
+
+Finally, we have assessed the robustness of technological roadmaps tested in myopic transitions.
+To do so, we used the approach of \acrfull{PCA} to identify the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.
+Projecting the results of myopic transitions on these principal components of the transition pointed out which roadmap was more robust and less likely to require additional investments along the transition.\\
+
+% THE RESULTS WE GOT (WHAT?)
+Given the ambitious \ce{CO2}-budget, the presumably high levels of demands and the limited potential of local renewables, electrofuels found to be the third pillar of the Belgian transition after opting for more efficient technologies and fully deploying local solar and wind energies.
+The uncertainty on the cost of purchasing these energy carriers drives by far (46.8\%) the variation of the total transition cost, before the industrial end-use demands (23.2\%), the interest rate (12.0\%) and the cost of purchasing fossil fuels (5.7\%).
+E-methanol is the key-molecule to import in the mid-term to defossilise the non-energy demand and substitute naphtha-crackers to produce on average 43\,TWh \acrfull{HVC}.
+The uncertainty on this demand is the most impacting parameter (80\%) on the variation of importing e-methanol.
+In the longer-term, importing e-ammonia becomes crucial to supply its own non-energy demand ($\sim$ 10\,TWh) and, before all, \acrfull{CCGT} to provide flexibility in the production of electricity.
+For this reason, after the variation of its cost of purchasing (50\%), the quantity of imported e-ammonia varies depending on the installation of nuclear \acrfull{SMR} (20\%) that substitutes \gls{CCGT}.
+To a smaller extent, e-methane and e-hydrogen help defossilising the industrial sectors and the heavy-duty road transport, respectively.
+E-methane imports is mostly impacted by industrial demands (45\%).
+The CAPEX of fuel cell vehicles (25\%) and its cost of purchasing (20\%) are the key parameters on the variation of the e-hydrogen import.
+
+Through the \gls{RL}-based exploration of myopic transitions, we pointed out that near-term actions were mandatory to hope succeeding this ambitious transition.
+Beyond 2030, if the right actions have not been taken, the cumulative emissions are likely to overshoot the \ce{CO2}-budget.
+Besides limiting the use of coal at any cost, limiting the consumption of fossil gas was needed at the early stages.
+Once the fuel switch, from fossil fuels to their renewable equivalents, has been operated, limiting the overall emissions is the most effective action to stick to the direction towards sustainability.
+The system overall efficiency provides less informative insight than the share of renewable energy carriers in the primary energy mix.
+The latter shows intermediate milestones to reach, 62\%-share by 2030, to have higher chance to succeed the transition by 2050.
+Below this threshold, these chances are much more limited, \ie no-go zones.
+Finally, due to limited foresight into the future and its uncertainties, successful myopic transitions rely more on importing renewable electrofuels than PF approach.
+This is even more the case for e-ammonia that is in general twice more imported by 2050 in myopic than PF transitions.
+
+Finally, in the case of Belgium, the integration of solar \gls{PV} supported by the electrification of the industrial and decentralised heating sectors drive 57\% of the variance of the technological mix through the transition.
+This affects the entire transition.
+On the contrary, other key contributions to this variance are more focused on tipping years.
+For instance, the low-temperature heating sectors shift from oil and gas decentralised boilers to electric \acrfull{HP} and \acrfull{DHN} from 2025 to 2035.
+In the non-energy sector, the shift between naphtha-crackers and \acrfull{MTO} takes place in 2025.
+Then, testing roadmaps resulting from perfect foresight optimisations in myopic conditions show that relying on \gls{SMR} by 2040 onward was not providing more or less robustness in terms of additional capacities to install.However, despite representing additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \acrfull{CHP} provide significantly better level of robustness (between 33\% and 49\%).
+
+% WHAT TO DO WITH THAT (SO WHAT?)
+\textbf{Results outcome}
 \begin{itemize}
-\item Due to the formulation of the salvage value, some technologies remain in place whereas they are not used anymore (see Appendix B.3, e.g.
-coal boilers and naphtha-cracker).
-It's better for the system to keep unused assets to recover part of their salvage value than prematurely decommissioning them.
-\item Progressive scale-up of \gls{VRES} instead of full potential possible from 2025.
+\item Developing local \gls{VRES} needs to be done at any cost
+\item Future nuclear, in the form of \gls{SMR} for instance, would come in competition with electrofuels.
+Betting on \gls{SMR} in the longer-term would post-pone the effort to reduce the \gls{GHG} emissions and would require important (and fast) changes in the case this technology would eventually not be available.
+A more robust strategy, especially with the limited foresight we have in the future is to invest in electrofuels to hope succeeding this ambitious transition.
 \end{itemize}
 
-\section*{Application outcomes}
-\addcontentsline{toc}{section}{Application outcomes}
-What this new methodology has brought over when applied to the case of Belgium
+\textbf{Methodology outcome}
+\begin{itemize}
+\item Results could be applied to other countries with a high demand and low renewable potentials (e.g. Netherlands and Germany)
+\item Assess the robustness of other technological roadmaps
+\item Application to other whole-energy system
+\end{itemize}
 
-\section*{Recommendations and guidance}
-\addcontentsline{toc}{section}{Recommendations and guidance}
-What to do then for policymakers, how to use the tool
-
-\section*{Perspectives}
-\addcontentsline{toc}{section}{Perspectives}
-List the future works to build upon the thesis
-
+% FUTURE PERSPECTIVES (WHAT'S NEXT?)
 \begin{itemize}
@@
 Compare the policies: one versus the other versus one followed by the other
+
+\item Progressive scale-up of \gls{VRES} instead of full potential possible from 2025.
 \end{itemize}
-
-Quand je parle de ce que j'ai fait avec RL : Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
-Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
-The intermediate solutions obtained in the middle of the time window are used as a new starting point for the agent that makes a new series of decisions for the next ten years, etc.
-Repeating the whole transition with different sequences of actions-states allow the agent to come up with an optimised policy towards sustainability, considering the variation of the parameters of its environment.

### 2024-04-21 chap_intro_ccl/conclusion.tex
Conclusion: Done

@@
 In the non-energy sector, the shift between naphtha-crackers and \acrfull{MTO} takes place in 2025.
-Then, testing roadmaps resulting from perfect foresight optimisations in myopic conditions show that relying on \gls{SMR} by 2040 onward was not providing more or less robustness in terms of additional capacities to install.However, despite representing additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \acrfull{CHP} provide significantly better level of robustness (between 33\% and 49\%).
+Then, testing roadmaps resulting from perfect foresight optimisations in myopic conditions show that relying on \gls{SMR} by 2040 onward was not providing more or less robustness in terms of additional capacities to install.However, despite representing additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \acrfull{CHP} provide significantly better level of robustness (between 33\% and 49\%).\\
 
 % WHAT TO DO WITH THAT (SO WHAT?)
-\textbf{Results outcome}
-\begin{itemize}
-\item Developing local \gls{VRES} needs to be done at any cost
-\item Future nuclear, in the form of \gls{SMR} for instance, would come in competition with electrofuels.
-Betting on \gls{SMR} in the longer-term would post-pone the effort to reduce the \gls{GHG} emissions and would require important (and fast) changes in the case this technology would eventually not be available.
-A more robust strategy, especially with the limited foresight we have in the future is to invest in electrofuels to hope succeeding this ambitious transition.
-\end{itemize}
+In the context of the energy transition, this thesis aimed at providing decision-makers with new methods and informed guidances accounting for the intrinsic uncertainties of the future.
+Even though the following conclusions have been drawn from studies on Belgium, the trends can be transposed to other countries with a high demand and low renewable potentials such as the Netherlands or Germany.
 
-\textbf{Methodology outcome}
-\begin{itemize}
-\item Results could be applied to other countries with a high demand and low renewable potentials (e.g. Netherlands and Germany)
-\item Assess the robustness of other technological roadmaps
-\item Application to other whole-energy system
-\end{itemize}
+Besides a wider electrification of the system and the integration of more efficient technologies, a robust policy consists in investing in electrofuels.
+Chemically equivalent to their fossil-based equivalents, they can benefit from the current infrastructures and provide flexibility to the mismatch in space and time between the production of renewable electricity and its consumption.
+Already stated via an analysis on the snapshot model by \citet{rixhon2021role}, this work confirms through the optimisation of the whole transition pathway that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
+On the road to a robust whole-energy system in Belgium, the policy makers, the industries, and academia should invest in projects to produce and use these fuels or developing the exchange networks with neighbouring countries.
+This would provide significant reduction in the variation of the total transition cost and robustness towards uncertainties progressively unfolding with the future.
 
-% FUTURE PERSPECTIVES (WHAT'S NEXT?)
-\begin{itemize}
-\item Electrofuels is the unicorn solution. What about carbon removal
-\item \textbf{Word about sufficiency} oui mais c'est sans doute celle qui est le plus enviable car la moins risqué à tenter d'atteindre, les solutions mirages technologiques, si on y croit trop on met tout la dessus et si ca foire, on est encore plus dans la merde pcq la direction est mauvaise, la solution sobriété, meme si jamais atteint les efforts pr l'atteindre ne seront pas contre productif
-\item \textbf{Word about availability of electrfuels} Voir mémoire Ced et Simon
-\item Extract a roadmap representative of a multiple-run UQ or RL process.
-\item In this work, we only consider emissions related to the use of energy carriers and not the construction --> LCA
-\item Multi-objective optimisation
-\item Potentially interesting to implement reward shaping to accelerate learning or guide the agent towards achieving the desired behaviour more efficiently.
+About the use of the electrofuels, the two molecules imported in largest quantity, e-methanol and e-ammonia, would aim at defossilising the non-energy and providing flexibility in the production of electricity, respectively.
+The main use of e-hydrogen is in the freight transport via trucks and e-methane would supply the industrial sectors through \gls{CHP} and boilers.
+
+Also brought up as another ``unicorn'' solution to the problem of the energy transition, \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but would still support the deployment of local \acrfull{VRES} such as solar and wind.
+This confirmed that investing as soon as possible in local renewables is one pillar of the transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2}-budget of the transition.
+
+On a methodological point of view, this thesis developed novel approaches to explore myopic pathway transitions, highlight intermediate milestones and no-go zones towards sustainability and assess the robustness of transition technological roadmaps.
+Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them on their specific case study.
+\\
+
+As the research in the field of the energy transition is a never-ending story, we conclude with perspectives that could benefit from the methods and the results developed in this thesis.
+Future studies could focus on different scenario analyses, the development of EnergyScope Pathway or the approaches applied to it such as \gls{RL}.
+
+The Belgian transition pathways explored in this thesis rely massively on the early deployment of local \gls{VRES} and import of electrofuels.
+In some of these pathways, in a 5 to 10 year time window, the current capacities of \gls{PV} and wind turbines are multiplied by 10 and 6, respectively.
+On their side, the amount of imported electrofuels could be 10 to 20 times more than the potential that Belgium could benefit from exporting countries, given the already signed agreements \cite{lefebvre2022electrofuel}.
+As the objective of this work was to explore as much as possible transition pathways, we have decided not to put further constraints on these energy carriers.
+One future work would be to constrain the potential of local renewables to a more progressive deployment or the availability of the electrofuels.
+This could provide different pathways where electrofuels would be produced more locally using carbon-capture technologies.
+This could also show the limit of the Belgian energy to meet such an ambitious \ce{CO2}-budget in its energy transition.
+
+To assess more accurately the boundary conditions of the Belgium in its transition pathway, a future work could be to merge the EnergyScope Pathway with EnergyScope Multi-Cells \cite{thiran2023validation}.
+Rather than model a single country and the rest of the world as the exterior, the Multi-Cells optimises multiple interconnected regions.
+At a European scale, Belgium could benefit from energy exchanges with countries where the renewable potential is more important such as France or Spain.
+
+Rather than limiting the availability of renewable resources, one could question the level of demands and address the question of the sufficiency.
+For a growing part of the scientific community \cite{o2018good}, sufficiency is seen as a no-regret lever of action to mitigate climate change.
+A study could consist in analysing the level of demands, within their respective sector, we should reach to meet the \gls{CO2}-budget and limit the installation of solar PV and wind turbines or the import of electrofuels.
+
+On the methodological side, several perspectives could be considered about the whole-energy system model or the implementation of \gls{RL}.
+
+In this thesis, we have constrained the cumulative \acrfull{GHG} emissions of the transition based on the use of the resources.
+However, in a system reducing drastically its emissions, indirect emissions such as the ones related to the construction and dismantling of assets, would have a much higher share of the total emissions.
+Depending on the level of emission reduction, these indirect emissions could be one to three times higher than the direct emissions \cite{blanco2020life}.
+To accurately integrate these emissions in the model and account for them in the \ce{CO2}-budget, \gls{LCA} can compute these indirect emissions.
+On top of this, \gls{LCA} could provide data to assess other impact indicators besides climate change \cite{astudillo2018integrating}.
+Based on this multi-criteria approach, another thread of future work would be to implement multi-objective optimisation \cite{dubois2023multi}.
+In this method, objectives other than the sole total transition cost such as the ones identified as planetary boundaries \cite{richardson2023earth} would be minimised.
+
+Regarding the \gls{RL} approach, we identify two threads of future development: reward shaping and multi-fidelity.
+Rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}.
+Reward shaping aims at accelerating the learning or guiding the agent towards achieving the desired behaviour more efficiently.
 In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode.
 However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy.
-Reward shaping addresses this issue by providing additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights.
+Reward shaping addresses this issue.
 These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective.
-However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task (see Section \ref{sec:RL:act_states_rew}).
+However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task.
 
-\item Talk here about potential improvement of the approach like the Multi-Fidelity Reinforcement Learning by \citet{cutler2014reinforcement}, rather than just unidirectional transfer from low to high-fidelity.
-Assess the severity of learning on Monthly model, versus Hourly model.
-Compare the policies: one versus the other versus one followed by the other
-
-\item Progressive scale-up of \gls{VRES} instead of full potential possible from 2025.
-\end{itemize}
+Given the similarities between the policy learned on the monthly (coarse) and the hourly (fine) model, another way to improve the efficiency of the \gls{RL} process would be to implement transfer knowledge \cite{mann2013directed}, or even multi-fidelity \cite{cutler2014reinforcement}.
+The first step, transfer learning, would consist in starting the learning process on the monthly and faster model, refine the action space, and continue learning on the more accurate hourly model.
+More than this unidirectional transfer of information, \citet{cutler2014reinforcement} developed the Mutli-Fidelity Reinforcement Learning (MFRL) framework that defines rules for when the agent should move up to a higher fidelity environment, as well as moving down in fidelity before over-exploring in a more expensive optimisation.
+This would result in an agent that can, on the one hand, use information from the monthly model to limit the exploration in the hourly model, and, on the other hand, exploit the data of the hourly model to update the policy learned on the monthly one.

### 2024-04-21 chap_intro_ccl/conclusion.tex
Intro: Implementation of OP's comments

@@
 % WHAT TO DO WITH THAT (SO WHAT?)
-In the context of the energy transition, this thesis aimed at providing decision-makers with new methods and informed guidances accounting for the intrinsic uncertainties of the future.
+In the context of the energy transition, this thesis aimed at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
 Even though the following conclusions have been drawn from studies on Belgium, the trends can be transposed to other countries with a high demand and low renewable potentials such as the Netherlands or Germany.
@@
 Rather than limiting the availability of renewable resources, one could question the level of demands and address the question of the sufficiency.
-For a growing part of the scientific community \cite{o2018good}, sufficiency is seen as a no-regret lever of action to mitigate climate change.
+For a growing part of the scientific community \cite{o2018good}, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
 A study could consist in analysing the level of demands, within their respective sector, we should reach to meet the \gls{CO2}-budget and limit the installation of solar PV and wind turbines or the import of electrofuels.

### 2024-04-21 chap_intro_ccl/intro.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR
Intro: Implementation of OP's comments

@@
-Climate change, among other environmental changes, is mostly due to the concentration of anthropogenic \acrfull{GHG} in the environment \cite{IPCC_CO2_budget}.
-Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
+Anthropogenic \gls{GHG} emissions must be reduced to mitigate environmental damages \cite{IPCC_CO2_budget}.
 Energy consumption, mainly from fossil fuels, is the main contributor: it is responsible for 75\% of \gls{GHG} emissions, the remaining 25\% originating from agriculture and industrial processes \cite{ourworldindata_CO2_world}.
-With the perspective of the energy system, the Kaya identity \cite{kaya1997environment} can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
+The Kaya identity is an economic formulation stating that the total emission level of the \gls{GHG} can be expressed as the product of four factors: human population, \gls{GDP} per capita, energy intensity (per unit of GDP), and carbon intensity (emissions per unit of energy consumed) \cite{kaya1997environment}.From an energy point of view, this identity can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
 \gls{EUD} is the energy service required by the the final consumer.
@@
 
-Such an identity is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not translate properly the heterogeneity of the situation \cite{IPCC2000}.
-Considering the expected growing trend of the Population, Equation (\ref{eq:equality_GHG}) highlights three levers of action that should be activated to reduce the \gls{GHG} emissions and favour the transition \cite{dodson2020population,scovronick2017impact}: renewables, efficiency and sufficiency.
+Such an identity is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not appropriately mirror the heterogeneity of the situation \cite{IPCC2000}.
+Considering the expected growing trend of the Population, Equation (\ref{eq:equality_GHG}) highlights three levers of action that should be activated to reduce the \gls{GHG} emissions and favour the energy transition \cite{dodson2020population,scovronick2017impact}: renewables, efficiency and sufficiency.
 
@@
 Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving sociology, psychology and political science, as it is outside the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
-One of the objectives of this manuscript is to provide support to such interdisciplinary projects in the assessment of sufficiency policy.
+One of the objectives of this manuscript is to support interdisciplinary projects in the assessment of sufficiency policy.
 \\
@@
 \noindent
-To ensure the energy supply of an assumably more and more demanding society in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
+To ensure the energy supply of a growing and more demanding population in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
 Similarly to the RED III, the transformations identified by \citet{iea2020world} focus on technical aspects, disregarding the sufficiency.
@@
 Changing primary energy sources corresponds to integrating more renewables.
-\gls{VRES} like wind and solar are the keystone to defossilise the energy system but they face challenges related to their intermittency and space disparity.
+\gls{VRES} like wind and solar are the keystones to defossilise the energy system but they face challenges related to their intermittency and space disparity.
 To address this issue,  limitations like range, power or  costs make \gls{DC} lines or batteries not the most relevant solutions.
@@
 In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
-gas networks present much more storage potential than electrical network: 50 times more in Germany and 300 times more in France \cite{Rosa2017}.
+gas networks present much more storage potential than electrical networks: 50 times more in Germany and 300 times more in France \cite{Rosa2017}.
 
@@
 In other words, hydrocarbons, currently produced from fossil resources, will still be composed of carbon in a renewable world.
-This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in a carbon-neutral energy transition \cite{mertens2020carbon}.
+This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in the energy transition that aims at carbon-neutrality where there is no additional, if no removal, of \gls{GHG} emissions in the atmosphere \cite{mertens2020carbon}.
 
@@
 To avoid the confusion between all these ``alternative'' fuels and thereby reduce misunderstanding in political or academic discussions, we have developed a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
-In the rest of this thesis, the electro --- and bio --- fuels are considered as renewable with no associated \gls{GWP}.
+In the rest of this thesis, the electro --- and bio --- fuels, imported from the exterior of the considered energy system, are considered as renewable with no associated \gls{GWP}.
 
@@
 However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system such as a country.
-Given the lifetime of energy conversion technologies, such strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
+Given the lifetime of energy conversion technologies, such a strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
 This long-term and large-system optimisation motivates the need to account for \gls{UQ}, which is considered a major challenge of such models in the literature \cite{pfenninger2014energy}.
@@
 
-On top of dealing with these uncertainties, the reality of decision-makers is a limited foresight into the future \cite{poncelet2016myopic}.
-In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'' in 2050 and make the planning decisions once and for all accordingly.
-On the contrary, they uncover the realisation of these uncertainties step-by-step, in a ``myopic'' way, and progressively act on them to, hopefully, meet the set target to mitigate the climate change.
+On top of dealing with these uncertainties, the reality of decision-makers includes a limited foresight into the future \cite{poncelet2016myopic}.
+In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'' in 2050 and accordingly make the planning decisions once and for all.
+On the contrary, they uncover the realisation of these uncertainties step-by-step, in a ``myopic'' way, and progressively act on them to, hopefully, meet the set target to reduce the anthropogenic \gls{GHG}.
 In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there was a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
@@
 Given the complexity of whole-energy system models, the time scale of the transition and the large number of uncertainties, only considering the total cost does not give information on the sensibility of the design strategy, \ie the investment decisions.
-For instance, total cost might consist in excessive upfront investments cost and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable depsite a slightly higher total cost.
+For instance, total cost might consist in excessive upfront investments and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable depsite a slightly higher total cost.
 \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
 However, their work focused on the power sector only and for a target future year (2035).
-Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost that and provide more details on the sensibility of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
+Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost and provide more details on the sensibility of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
 
@@
 ``\textit{Our task is not to foresee the future, but to enable it}''\footnote{Saint-Exupéry in Citadelle, 1948}.
-This thesis aims at providing decision-makers with new methods and informed policies accounting for the intrinsic uncertainties of the future.
+This thesis aims at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
 Rather than trying, in vain, to answer the question ``What could possibly happen in the future?'', this work rather addresses the ``What could or should we do to make the future possible?''.
@@
 \begin{itemize}
-\item What can be the role of electrofuels in the transition pathway of a whole-energy system subject to uncertainties, limited foresight into the future and a \ce{CO2}-budget?
+\item What can be the role of electrofuels in the transition pathways of a whole-energy system considering uncertainties,  a limited foresight, and a given \ce{CO2}-budget?
 What are the key uncertainties driving their import?
-\item How to explore the multiple pathway possibilities through the optimisation of a policy, \ie sequence of actions to support a transition?
-\item How to assess the robustness of a pathway roadmap defining the design strategy over the transition?
+\item How to explore the multiple possible pathways through the optimisation of a sequence of actions to support an energy transition?
+\item How to assess the robustness of a transition roadmap whilst developing strategy and designing possible pathways?
 \end{itemize}
@@
 The model developed and used in the context of this thesis is EnergyScope Pathway \cite{limpens2024pathway}.
-First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for the pathway transition from an existing system to a long term target, \ie 2050.
-Based on this model originally implemented in a perfect foresight approach, we have developed the myopic method in which the whole time horizon (30 years) is optimised through a sequence of 10-year time-windows with 5-year overlaps.
+First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for transition pathways from an existing system to a long-term target, \ie 2050.
+Based on this model originally implemented for a perfect foresight approach, we have developed the myopic method in which the whole time horizon (30 years) is optimised through a sequence of 10-year time-windows with 5-year overlaps.
 
 To address the question of the role of electrofuels, they have been integrated in the pre-existing model, focusing on four molecules: hydrogen, methane, ammonia and methanol.
-Moreover, given their current and (expected) future role into the sector of the \gls{NED}, we have added to the model this sector with a similar level of detail as the other sectors of the system, \ie electricity, heat and transport.
+Moreover, given their current and (expected) future role in the sector of the \gls{NED}, we have added this sector to the model with a similar level of detail as for the other sectors of the system, \ie electricity, heat and transport.
 
@@
 
-To explore the different pathway trajectories in this myopic optimisation process, we have applied the \gls{RL} method.
+To explore the different pathway trajectories in this myopic optimisation process, we have applied the \acrfull{RL} method.
 In this branch of machine learning, an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie sequence of actions to take to support the transition.
 
-Finally, in the objective to assess the sensibility to uncertainties of different design transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
+Finally, in the objective to assess the sensibility to uncertainties of different transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
 This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a frame.
@@
 This thesis is composed of five chapters to provide answers to the different research questions.
-Chapter \ref{chap:chap_methodo} details about the tools and methods used in this work.
-It starts with the main constraints, parameters and variables of the whole-energy system optimisation model, EnergyScope Pathway.
-Then, it gives information on the \gls{UQ} approach and the way it has been adapted to the case of Belgium and its transition pathway.
+Chapter \ref{chap:chap_methodo} details the tools and methods we used.
+It starts with the main constraints, parameters and variables of EnergyScope Pathway, the whole-energy system optimisation model.
+Then, it explains the \acrfull{UQ} approach and the way it has been adapted to the case of Belgium and its transition pathway.
 Finally, general fundamentals and more case-specific considerations are brought up about the \gls{RL} and \gls{PCA}-based robustness approaches.
@@
 Chapter \ref{chap:case_study} presents the case study of this work: Belgium and its energy transition.
-Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the date related \gls{SMR} as an option to produce nuclear-based electricity in the future.
+Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the data-related \gls{SMR} as an option to produce nuclear-based electricity in the future.
 
-Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} carried out on the Belgian energy transition under.
-On top of the impact of uncertainties on the total transition cost and the system design in general, we target the impact of having new nuclear (``atom'') capacities by 2040 onward and the driving parameters on the import of electrofuels (``molecules''): the atom-molecules dilemma.
+Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} carried out on the Belgian energy transition under uncertainties.
+On top of the impact of uncertainties on the total transition cost and the system design in general, we analyse the impact of having new nuclear (``atom'') capacities by 2040 onward and the driving parameters on the import of electrofuels (``molecules''): the so-called atom-molecules dilemma.
 
-In Chapter \ref{chap:chap_RL}, the rules of the ``\gls{RL} game'' are detailed.
+In Chapter \ref{chap:chap_RL}, we detail the rules of the ``\gls{RL} game''.
 In other words, we define the action and state spaces as well as the reward function driving the behaviour of the agent in its quest to optimise its policy.
-Then, we analyse the results of the learning phase before testing under uncertainties the learned policy versus more ``classic'' myopic optimisation, \ie without the support of such a policy.
+Then, we analyse the results of the learning phase before testing --- under uncertainties --- the learned policy versus more ``classic'' myopic optimisation, \ie without the support of such a policy.
 
-Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting of deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
+Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting from deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
 The first one, the reference case, considers nominal values for all the uncertain parameters.
 The SMR case is the one introduced in Chapter \ref{chap:atom_mol} where we allow the model to install \gls{SMR} from 2040 onward.
-Eventually, the ROB case accounts for the highest values of parameters for those having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, interest rate and variable \gls{OPEX} of technologies).
+Eventually, the ROB case accounts for the highest values of parameters having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, interest rate and variable \gls{OPEX} of technologies).
 
-Finally, we draw general conclusions and suggest potential perspectives for future works in terms of uses and further developments of the methodological tools.
+Finally, we draw the main conclusions of this work: near-term actions are needed to respect ambitious \ce{CO2}-budget and investing into electrofuels provide robustness to uncertainties and the limited foresight into the future.
+Perspectives are suggested for future research on uses and further developments of the methodological tools.

### 2024-04-22 chap_intro_ccl/conclusion.tex
Conclusion: Reviews by Jabibi and Thomas

@@
 % WHAT WE DID (HOW?)
-The general objective of this thesis has been to investigate the impact of uncertainties and foresight on the transition pathway of a whole-energy system (\ie multi-energy carriers and multi-sectors) subject to \ce{CO2}-budget without a prescribed trajectory to reduce the emissions.
-We have pursued this objective on the case study of Belgium with a focus on the role of renewable electrofuels (\ie e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely-populated and highly industrialised country with limited local renewable potentials.
-To have 66\% chance of limiting warming to 1.5°C \cite{IPCC_CO2_budget} and using the ``grandfathering'' sharing principle of the global budget \cite{knight2013grandfathering}, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$ which represents 10 years of the current level of emissions.
+The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2}-budget without prescribed trajectory for the reduction of the emissions.
+We have pursued this objective on the case study of Belgium with a focus on the role that renewable electrofuels such as e-ammonia, e-methane, e-methanol and e-hydrogen, could play in this densely-populated and highly industrialised country with limited local renewable potentials.
+To have 66\% chance of limiting warming to 1.5°C \cite{IPCC_CO2_budget} and using the ``grandfathering'' sharing principle of the global budget \cite{knight2013grandfathering}, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
+This budget represents 10 years of the current level of emissions.
 
-To optimise such a system, we have developed EnergyScope Pathway that minimise the total transition cost of the system from 2020 to 2050 and implemented the myopic approach.
-This approach considers a sequence of 10-year time windows with an overlap of 5 years between them rather than one global optimisation over the 30 years of the transition.
-This allows assessing the realism of the decision making process where knowledge about the future is progressively unfold rather than assumed complete from the beginning of the transition, the perfect foresight (PF) approach.
+To optimise such a system, we have developed EnergyScope Pathway that minimises the total transition cost of the system from 2020 to 2050 and implemented the myopic approach.
+This approach considers a sequence of 10-year time windows with an overlap of 5 years rather than one global optimisation over the 30 years of the transition.
+This allows assessing the realism of the decision making process where knowledge about the future is progressively unfold rather than assumed to be fully known from the beginning of the transition as in the perfect foresight (PF) approach.
 
-The future of a national whole-energy system is subject to a plethora of uncertainties such as the cost of purchasing resources, the \acrfull{CAPEX} of technologies or the level of demands.
-In this thesis, we have used the uncertainty characterisation approach developed by \citet{moret2016strategic} to provide the range of values for the uncertain parameters.
-To propagate and quantify the impact of these uncertainties on the outputs of the model, we have used the \acrfull{PCE} method implemented in the Coppitters' RHEIA framework \cite{coppittersthesis}.
+The future of a national whole-energy system is subject to a plethora of uncertainties such as the cost of purchasing resources, their availabilities, the \acrfull{CAPEX} of technologies or the level of demands.
+In this thesis, we have used the uncertainty characterisation approach developed by \citet{moret2016strategic} to provide the range of variation of the uncertain parameters.
+To propagate and quantify the impact of these uncertainties on the outputs of the model, we have used the \acrfull{PCE} method has been used as implemented in the Coppitters' RHEIA framework \cite{coppittersthesis}.
 
-To explore the myopic pathways respecting a \ce{CO2}-budget equivalent to 10 years of current emissions, without \ce{CO2}-trajectory, we have used the \acrfull{RL} approach.
-In this framework, an agent took actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
+The \acrfull{RL} approach was adopted to explore the myopic pathways respecting a \ce{CO2}-budget equivalent to 10 years of current emissions, without \ce{CO2}-trajectory.
+In this framework, an agent takes actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
 Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
-Although, these actions are taken every five years, they impact the system, for the next ten years --- their time window.
-Repeating the whole transition with different sequences of actions-states allow the agent to come up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
+Repeating the whole transition with different sequences of actions-states allow the agent coming up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
 
 Finally, we have assessed the robustness of technological roadmaps tested in myopic transitions.
-To do so, we used the approach of \acrfull{PCA} to identify the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.
-Projecting the results of myopic transitions on these principal components of the transition pointed out which roadmap was more robust and less likely to require additional investments along the transition.\\
+To do so, the approach of \acrfull{PCA} was mobilised to identify the technologies that are more sensitive to uncertainties and that are more likely to impact the overall variance of the transition design.
+Projecting the results of myopic transitions on these principal components of the transition pointed out which roadmap is more robust and less likely to require additional investments along the transition.\\
 
 % THE RESULTS WE GOT (WHAT?)
-Given the ambitious \ce{CO2}-budget, the presumably high levels of demands and the limited potential of local renewables, electrofuels found to be the third pillar of the Belgian transition after opting for more efficient technologies and fully deploying local solar and wind energies.
-The uncertainty on the cost of purchasing these energy carriers drives by far (46.8\%) the variation of the total transition cost, before the industrial end-use demands (23.2\%), the interest rate (12.0\%) and the cost of purchasing fossil fuels (5.7\%).
-E-methanol is the key-molecule to import in the mid-term to defossilise the non-energy demand and substitute naphtha-crackers to produce on average 43\,TWh \acrfull{HVC}.
-The uncertainty on this demand is the most impacting parameter (80\%) on the variation of importing e-methanol.
-In the longer-term, importing e-ammonia becomes crucial to supply its own non-energy demand ($\sim$ 10\,TWh) and, before all, \acrfull{CCGT} to provide flexibility in the production of electricity.
-For this reason, after the variation of its cost of purchasing (50\%), the quantity of imported e-ammonia varies depending on the installation of nuclear \acrfull{SMR} (20\%) that substitutes \gls{CCGT}.
-To a smaller extent, e-methane and e-hydrogen help defossilising the industrial sectors and the heavy-duty road transport, respectively.
-E-methane imports is mostly impacted by industrial demands (45\%).
-The CAPEX of fuel cell vehicles (25\%) and its cost of purchasing (20\%) are the key parameters on the variation of the e-hydrogen import.
+Given the \ce{CO2}-budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  the cost-optimum transition of Belgium should be to first opt for more efficiency to reduce by one third the primary energy consumption electrofuels.
+Then comes the full deployment of local solar and wind energies that would represent about 30\% of the mix by 2050.
+To fill the gap left by limited potential of local renewables, the electrofuels are found to be the third pillar of the Belgian transition.
+47\% of the variation of the total transition cost are due to the uncertainty on the cost of purchasing these energy carriers, before the industrial end-use demands (23\%), the interest rate (12\%) and the cost of purchasing fossil fuels (6\%).
 
-Through the \gls{RL}-based exploration of myopic transitions, we pointed out that near-term actions were mandatory to hope succeeding this ambitious transition.
+E-methanol is found to be the key-molecule to import in the mid-term to defossilise the non-energy demand and substitute naphtha-crackers to produce on average 43\,TWh \acrfull{HVC}.
+The uncertainty on this demand is the most impacting parameter (80\%) on the variation of the quantity of imported e-methanol.
+
+In the longer-term, importing e-ammonia becomes crucial to supply its own non-energy demand ($\sim$ 10\,TWh) and, before all, \acrfull{CCGT} to provide flexibility in the production of electricity in competition with nuclear \acrfull{SMR}.
+For this reason, where the variation of the quantity of imported e-ammonia is mostly driven by the uncertainty on its cost of purchasing (50\%), the uncertainty on the availability of nuclear \gls{SMR} by 2040 is the second most impacting parameter (20\%).
+
+To a smaller extent, e-methane help defossilising the industrial boilers and \acrfull{CHP}. E-methane imports is mostly impacted by industrial demands (45\%).
+
+By 2050, e-hydrogen is only imported from abroad in the wide majority of the transition pathways under uncertainties generated during this work.
+This molecule supports the defossilisation of the freight-transport by trucks.
+The variation of the imported quantity of e-hydrogen is mainly caused by the uncertainty on the CAPEX of fuel cell vehicles (25\%) and its cost of purchasing (20\%).
+In 12\% of the transition pathways, up to 10.5\,GW of electrolysers are installed when local renewable electricity is abundant, \gls{SMR} is available and imported e-hydrogen is expensive.
+These electrolysers would locally produce up to 27\,TWh of e-hydrogen.
+
+Through the Reinforcement Learning-based exploration of myopic transitions, we pointed out that near-term actions were mandatory to hope succeeding this ambitious transition.
 Beyond 2030, if the right actions have not been taken, the cumulative emissions are likely to overshoot the \ce{CO2}-budget.
 Besides limiting the use of coal at any cost, limiting the consumption of fossil gas was needed at the early stages.
-Once the fuel switch, from fossil fuels to their renewable equivalents, has been operated, limiting the overall emissions is the most effective action to stick to the direction towards sustainability.
-The system overall efficiency provides less informative insight than the share of renewable energy carriers in the primary energy mix.
+Once the fuel switch has been operated from fossil gas to e-methane, limiting the overall emissions is the most effective action to stick to the direction towards sustainability and encourage the import of renewable energy carriers.
+Regarding the information on which the \gls{RL}-agent determines its actions, the system overall efficiency provides less informative insight than the share of renewable energy carriers in the primary energy mix.
 The latter shows intermediate milestones to reach, 62\%-share by 2030, to have higher chance to succeed the transition by 2050.
-Below this threshold, these chances are much more limited, \ie no-go zones.
-Finally, due to limited foresight into the future and its uncertainties, successful myopic transitions rely more on importing renewable electrofuels than PF approach.
+Below this threshold, these chances of success are much more limited, the so-called no-go zones.
+Finally, due to limited foresight into the future and its uncertainties, successful myopic transitions rely more on importing renewable electrofuels than the perfect foresight approach.
 This is even more the case for e-ammonia that is in general twice more imported by 2050 in myopic than PF transitions.
 
-Finally, in the case of Belgium, the integration of solar \gls{PV} supported by the electrification of the industrial and decentralised heating sectors drive 57\% of the variance of the technological mix through the transition.
+In the case of Belgium, the uncertainty on the maximum installed capacity of solar \gls{PV}, which is supported by the electrification of the industrial and decentralised heating sectors, drives 57\% of the variance of the technological mix through the transition.
 This affects the entire transition.
@@
 In the non-energy sector, the shift between naphtha-crackers and \acrfull{MTO} takes place in 2025.
-Then, testing roadmaps resulting from perfect foresight optimisations in myopic conditions show that relying on \gls{SMR} by 2040 onward was not providing more or less robustness in terms of additional capacities to install.However, despite representing additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \acrfull{CHP} provide significantly better level of robustness (between 33\% and 49\%).\\
+
+Finally, testing roadmaps resulting from perfect foresight optimisations in myopic conditions has shown that cases relying or not on 6\,GW of \gls{SMR} by 2040 onward provide the same level of robustness in terms of additional capacities to install.
+However, despite representing an additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \gls{CHP} provides significantly better level of robustness (between 33\% and 49\%).\\
 
@@
 In the context of the energy transition, this thesis aimed at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
-Even though the following conclusions have been drawn from studies on Belgium, the trends can be transposed to other countries with a high demand and low renewable potentials such as the Netherlands or Germany.
+Even though the following messages result from studies on Belgium, the trends can be transposed to other countries with a high demand and low renewable potentials such as the Netherlands or Germany.
 
 Besides a wider electrification of the system and the integration of more efficient technologies, a robust policy consists in investing in electrofuels.
-Chemically equivalent to their fossil-based equivalents, they can benefit from the current infrastructures and provide flexibility to the mismatch in space and time between the production of renewable electricity and its consumption.
-Already stated via an analysis on the snapshot model by \citet{rixhon2021role}, this work confirms through the optimisation of the whole transition pathway that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
+Chemically similar to their fossil-based equivalents, they can benefit from the current infrastructures and provide flexibility to the mismatch between the production of renewable electricity and its consumption.
+They would be imported from countries ``in excess'' of renewable electricity from wind or solar such as Australia \cite{Australia_efuels}, Chile \cite{Chile_efuels} or Spain \cite{Spain_efuels} to provide seasonal storage of electricity.
+Already stated in the analysis of the snapshot model of \citet{rixhon2021role}, this work confirms through the optimisation of the whole transition pathway that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
 On the road to a robust whole-energy system in Belgium, the policy makers, the industries, and academia should invest in projects to produce and use these fuels or developing the exchange networks with neighbouring countries.
@@
 
-Also brought up as another ``unicorn'' solution to the problem of the energy transition, \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but would still support the deployment of local \acrfull{VRES} such as solar and wind.
-This confirmed that investing as soon as possible in local renewables is one pillar of the transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2}-budget of the transition.
+Also brought up as another ``unicorn'' solution to the problem of the energy transition, nuclear \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but would still support the deployment of local \acrfull{VRES} such as solar and wind.
+This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2}-budget of the transition.
 
@@
 
-The Belgian transition pathways explored in this thesis rely massively on the early deployment of local \gls{VRES} and import of electrofuels.
+The Belgian transition pathways explored in this thesis massively relied on the early deployment of local \gls{VRES} and import of electrofuels.
 In some of these pathways, in a 5 to 10 year time window, the current capacities of \gls{PV} and wind turbines are multiplied by 10 and 6, respectively.
+This is much faster than the multiplication by 2 and 2.5 of PV and wind capacities, respectively, from 2024 and 2034 projected by Elia, Belgium Transmission System Operator (TSO),  in their adequacy analysis \cite{Elia_2024_2034}.
 On their side, the amount of imported electrofuels could be 10 to 20 times more than the potential that Belgium could benefit from exporting countries, given the already signed agreements \cite{lefebvre2022electrofuel}.
@@
 For a growing part of the scientific community \cite{o2018good}, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
-A study could consist in analysing the level of demands, within their respective sector, we should reach to meet the \gls{CO2}-budget and limit the installation of solar PV and wind turbines or the import of electrofuels.
+A study could consist in analysing the level of demands, within their respective sector, we should reach to meet the \ce{CO2}-budget and limit the installation of solar PV, wind turbines or nuclear energy or the import of electrofuels.
 
@@
 
-In this thesis, we have constrained the cumulative \acrfull{GHG} emissions of the transition based on the use of the resources.
+The cumulative \acrfull{GHG} emissions of the transitionhas been constrained based on the use of the resources.
 However, in a system reducing drastically its emissions, indirect emissions such as the ones related to the construction and dismantling of assets, would have a much higher share of the total emissions.
@@
 
-Regarding the \gls{RL} approach, we identify two threads of future development: reward shaping and multi-fidelity.
+Regarding the \gls{RL} approach, two threads of future development have been identified: reward shaping and multi-fidelity.
 Rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}.
@@
 The first step, transfer learning, would consist in starting the learning process on the monthly and faster model, refine the action space, and continue learning on the more accurate hourly model.
-More than this unidirectional transfer of information, \citet{cutler2014reinforcement} developed the Mutli-Fidelity Reinforcement Learning (MFRL) framework that defines rules for when the agent should move up to a higher fidelity environment, as well as moving down in fidelity before over-exploring in a more expensive optimisation.
-This would result in an agent that can, on the one hand, use information from the monthly model to limit the exploration in the hourly model, and, on the other hand, exploit the data of the hourly model to update the policy learned on the monthly one.
+More than this unidirectional transfer of information, \citet{cutler2014reinforcement} developed the Multi-Fidelity Reinforcement Learning (MFRL) framework that defines rules for when the agent should move up to a higher fidelity environment, as well as moving down in fidelity before over-exploring in a more expensive optimisation.
+This would result in an agent that can, on the one hand, use information from the monthly model to limit the exploration in the hourly model, and, on the other hand, exploit the data of the hourly model to update the policy learned on the monthly one.\\
+
+Nowadays, the development on solar and wind energies supported by other technical solutions such as electrofuels exchange market or nuclear energy, is not fast enough to stick to the +1.5$^{\circ}$ C of global warming.
+Even though it represents a growing part of the literature, sufficiency is rarely brought up by decision-makers as a solution.
+At the same time, whole-energy system models and related methodologies do not claim to predict the future but rather aims at quantifying options under several assumptions.
+In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into discussions and decisions that should account for more interdisciplinarity.

### 2024-04-22 chap_RL/chap_RL.tex
RobPol + Methodo PCA: Review by HJ

@@
 
-Given the shape of the reward (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.
-The skewness of the cumulative emissions and costs in 2050 are indications of this shape of reward (see Table \ref{tab:skewness_gwp_cost}).
-When succeeding the transitions, the cumulative emissions and costs have a negative and positive skewness, respectively.
-On the contrary, when it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).
+Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.
+The skewness of the cumulative emissions and costs in 2050 are indications of this reward function (see Table \ref{tab:skewness_gwp_cost}).
+When succeeding the transitions, the cumulative emissions have a negative skewness: the agent successfully stayed within the budget and most of the cases where close to that budget (median at 0.9 Gt$_{\ce{CO2},\text{eq}}$).
+On the contrary, the cumulative cost of successful transitions has a positive skewness: the agent successfully reduces the cost of the system as a secondary objective with 30\% of the cases above the reference transition cost.
+This hierarchy of agent's objective is verified with the failures.
+When it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).
 

### 2024-04-22 chap_RobPol/chap_RobPol.tex
RobPol + Methodo PCA: Review by HJ

@@
 Where resistance consists in planning for the worst-case scenario, resilience aims at a fast recovery whatever the conditions in the future.
-Finally, in static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions.
-Where the adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}, the objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.
+In static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions.
+The adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}.
+The objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.
 \citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}.
-In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economy model, like EnergyScope, that usually assume rational choice within an overall cost minimization.
-However, in the case of the transition pathway of a whole-eenrgy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''.
+In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economic model, like EnergyScope, that usually assume rational choice within an overall cost minimization.
+However, in the case of the transition pathway of a whole-energy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''.
 Between the sole total transition cost and the entire set of installed technologies that give too few and too much information, respectively, the performance metric here is defined through the \gls{PCA} approach.
@@
 The main contributions of this chapter is the application of the methodology proposed in Section \ref{sec:meth:PCA} to the case study of the Belgian energy transition.
-First, we develop the different steps that lead to the principal components of the transition.
+First, we apply the different steps that lead to the principal components of the transition.
 We analyse these big trends of variation and highlight the fact that these variations stand for the entire pathway, a group of consecutive representative years or rather on a tipping-year.
@@
 Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition.
-In other words, the further goes the transition, the more limited are the degrees of freedom to respect the \ce{CO2}-budget.
+In other words, the further the transition goes, the more limited the degrees of freedom are to respect the \ce{CO2}-budget.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Top5_PC1.pdf}
-\includegraphics[width=0.8\textwidth]{Top5_PC2.pdf}
-\includegraphics[width=0.8\textwidth]{Top5_PC3.pdf}
+\includegraphics[width=0.9\textwidth]{Top5_PC1.pdf}
+\includegraphics[width=0.9\textwidth]{Top5_PC2.pdf}
+\includegraphics[width=0.9\textwidth]{Top5_PC3.pdf}
 \caption{Contribution of the top-5 technologies to the first three \gls{PCs} over the different representative years.
@@
 Even though the following observations could be made by analysing the distribution of installed capacities (see Appendix \ref{app:UQ_tech_cap}) or the covariance matrices, the \gls{PCA} decomposition offers a more visual and summarising representation of the main trends of variation.
-Due to their intermittency, integrating more \gls{PV} panels leads to the installation of other technologies to benefit from free and renewable electricity when it exceeds the electrical \gls{EUD}.
+Due to their intermittency, increasing the integration of \gls{PV} panels requires the installation of other technologies to benefit from free and renewable electricity when it exceeds the electrical \gls{EUD}.
 Therefore, we observe that the variation of installed \gls{PV} is directly linked with the variation of installed industrial electrical resistors and, to a smaller extent, of decentralised and \gls{DHN} electrical \gls{HP}.
-These variations cover the whole transition as the main varying factors given their significant contributions to the $\text{PC}_{1}$.
-Where this was an example for correlated technologies, we can also identify some key modal shifts where one technology is either substituted by others or in balance with another one.
+These variations spread over the whole transition and they are the main varying factors to the first $\text{PC}_{\text{transition}}$.
+Where this was an example for correlated technologies, we can also identify some key modal shifts where one technology is either substituted or in balance with others.
 First, about the \gls{LT} heat sector, the early stages of the transition, \ie 2025-2030, sees the shift from mainly decentralised oil and gas boilers towards decentralised and \gls{DHN} electrical \gls{HP}.
 Later in the transition, there seems to be a tight competition between (bio)diesel and \gls{FC} trucks that drive the design variance to a smaller extent as they mostly appear in the second and third \gls{PCs} of the representative years.
-Not capturing a significant share of variance, there are other modal shifts (\eg \gls{BEV} substituting diesel and gasoline cars) that are not visible through the \gls{PCs}.
-Besides these modal shifts spread over several representative years, 2025 is the tipping year concerning the shift from naphtha-cracker and methanol-to-olefins to supply \gls{HVC}.
+Representing a smaller share of the total variance, there are other modal shifts (\eg \gls{BEV} substituting diesel and gasoline cars) that are not visible through the \gls{PCs}.
+Besides these modal shifts spread over several representative years, 2025 is the tipping year concerning the shift from naphtha-cracker to methanol-to-olefins to supply \gls{HVC}.
 Finally, there are also technologies contributing to \gls{PCs} because they are the main producing assets of their respective end-use sector and the demand varies significantly, \eg biomass-to-methanol.
@@
 Summing all of these variances over the different years results in a ``pseudo'' total variance of the transition.
-To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 80\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).
+To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 85\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).
 This results in keeping 14 $\text{PC}_{y}$: the first and second \gls{PCs} of each representative year and the third PC of 2025 and 2035.
-Even though the 80\%-threshold is reached with the first 11 \gls{PCs},  $\text{PC}_{2045,2}$, $\text{PC}_{2050,2}$ and $\text{PC}_{2035,3}$ are accounted for as they are similar to $\text{PC}_{2040,2}$.
-Consequently, aiming at capturing at least 80\% of the total transition variance,
 
@@
 10 & 2025 & $\text{PC}_3$ & 15.0 & 78.2 \\
-11 & 2040 & $\text{PC}_2$ & 8.7 & \textbf{80.2} \\
+11 & 2040 & $\text{PC}_2$ & 8.7 & 80.2 \\
 12 & 2045 & $\text{PC}_2$ & 7.9 & 82.0 \\
 13 & 2050 & $\text{PC}_2$ & 7.7 & 83.7 \\
-14 & 2035 & $\text{PC}_3$ & 7.3 & 85.4 \\
+14 & 2035 & $\text{PC}_3$ & 7.3 & \textbf{85.4} \\
 $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$\\
@@
 Consequently, costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies are considered at their worst values, \ie the upper bound of their respective range.
-In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{LT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
+In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
 
@@
 It is mostly the industrial electric heaters that absorb the abundant and intermittent electricity produced by \gls{PV} panels and wind turbines.
-In the mobility sectors, \gls{BEV} substitutes gasoline cars from 2025 onward and electric trucks make the transition from diesel to \gls{FC} trucks.
+In the mobility sectors, \gls{BEV} substitutes gasoline cars from 2025 onward and electric trucks substitute diesel trucks in the near-term before being substituted by \gls{FC} trucks at the end of the transition.
 On the supply side, besides the local \gls{VRES} and the direct import of electricity from abroad, \gls{CHP} units, mostly industrial, are favoured given their higher efficiency at the expense of ammonia-\gls{CCGT}.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_ROB.pdf}
-\includegraphics[width=0.8\textwidth]{Elec_Layer_ROB.pdf}
+\includegraphics[width=0.82\textwidth]{Elec_Tech_Cap_ROB.pdf}
+\includegraphics[width=0.82\textwidth]{Elec_Layer_ROB.pdf}
 \caption{Installed capacities and layer balance of the power sector in the ROB case.
@@
 
-In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}.More expensive imported energy carriers lead to, on the one hand, an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) supported by a higher electrification of the heating and transport sectors.
+In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}.
+More expensive imported energy carriers lead to, on the one hand, an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) supported by a higher electrification of the heating and transport sectors.
 On the other hand, it increases the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
@@
 
+The ROB roadmap is more robust than the REF case for all PCs but PC 4.
+About PC 1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors.
+In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations.
+For the same reasons, the ROB roadmap is 37\% more robust for PC 2.
+The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers.
+In myopic transitions though, this technology keeps on being installed in the mid-term.
+Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
+Finally, the ROB roadmap is 21\% less robust for PC 5.
+Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller.
+As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.
+
 \begin{table}[htbp!]
@@
 5 &  +34\% & -21\%\\
-
 \bottomrule							
-
 \end{tabular}
 \end{table}
-
-The ROB roadmap is more robust than the REF case all PCs but PC 4.
-About PC 1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors.
-In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations.
-For the same reasons, the ROB roadmap is 37\% more robust for PC 2.
-The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers.
-In myopic transitions though, this technology keeps on being installed in the mid-term.
-Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
-Finally, the ROB roadmap is 21\% less robust for PC 5.
-Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller.
-As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.
 
@@
 \centering
-\includegraphics[width=0.49\textwidth]{Top5_PCtransition_REF_MY.pdf}
-\includegraphics[width=0.49\textwidth]{Top5_PCtransition_SMR_MY.pdf}
-\end{figure}
-
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=0.49\textwidth]{Top5_PCtransition_ROB_MY.pdf}
+\includegraphics[width=0.75\textwidth]{Top5_PCtransition_REF_MY.pdf}
+\includegraphics[width=0.75\textwidth]{Top5_PCtransition_SMR_MY.pdf}
+\includegraphics[width=0.75\textwidth]{Top5_PCtransition_ROB_MY.pdf}
 \caption{Contribution of the top-5 technologies to the \gls{PCs} of myopic transitions for the REF, SMR and ROB cases.
@@
 
+\newpage
 \section{Conclusions}

### 2024-04-22 chap_intro_ccl/intro.tex
RobPol + Methodo PCA: Review by HJ

@@
-Anthropogenic \gls{GHG} emissions must be reduced to mitigate environmental damages \cite{IPCC_CO2_budget}.
+The rate at which climate changes today is mostly due to the concentration
+of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
+Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
 Energy consumption, mainly from fossil fuels, is the main contributor: it is responsible for 75\% of \gls{GHG} emissions, the remaining 25\% originating from agriculture and industrial processes \cite{ourworldindata_CO2_world}.

### 2024-04-22 chap_methodo/chap_methodo.tex
RobPol + Methodo PCA: Review by HJ

@@
 Like the normalization of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude.
-Finally, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
-After this preprocessing, PCs can be computed for each of the year of the transition then aggregated to give a bigger picture over the whole transition.\\
+Then, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
+After this preprocessing, PCs can be computed for each of the representative year of the transition.
+Finally, PCs that indicate similar directions in different years are averaged into one single PCs of transition.
+The different PCs of transition give a bigger picture over the whole transition.\\
 
@@
 To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})).
-Even though these represent only the design part of the result of the optimization, along with the operation, focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:c_inv})) and, more indirectly, the resources to use.
+Even though these represent only the design part of the result of the optimisation focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation.
 In other words, it captures the technological landscape of the transition.
@@
 For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
+Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.
 Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}).
@@
 This gives the whole dataset considered in this \gls{PCA} (see Figure \ref{fig:PCA_data}).
-Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.\\
+\\
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Share_EUD_sectors.pdf}
+\includegraphics[width=1\textwidth]{Share_EUD_sectors.pdf}
 \caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition.
@@
 
-To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal value (see Chapter \ref{chap:atom_mol}).
+To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal values (see Chapter \ref{chap:atom_mol}).
 The end-use-demands as well as the commodity produced for the sector coupling are based on the results of this deterministic REF case.
@@
 Finally, to compare apples with apples, we converted the \gls{EUD} in the mobility sectors, \ie passenger and freight, into the \gls{FEC} they require in the REF case.
-This gives the second weighing factor to scale data, on top of the one of Equation (\ref{eq:PCA_scaling_0}) (see Equation (\ref{eq:PCA_scaling_1})).
+This gives the second weighing factors to scale data, on top of the ones of Equation (\ref{eq:PCA_scaling_0}), see Equation (\ref{eq:PCA_scaling_1}).
 
@@
 
-One would notice that this second scaling factor omits the infrastructure and storage technologies.
+One can notice that this second scaling factor omits the infrastructure and storage technologies.
 In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact.
@@
 These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}).
-Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the rest of the dataset for this specific technology and year across all the samples.
+Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for this specific technology and year.
 There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
@@
 Now that data are selected and preprocessed, principal components are first computed for each representative year separately, using the Python package \textsc{PCA} from \textsc{sklearn.decomposition}.
-As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.} in our case study) which is intractable.
+As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73 in our case study\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.}) which is intractable.
 Moreover, the first PCs keep track of most of the variance of the system whereas the last ones present a smaller interest.
@@
 Retaining too few principal components may result in loss of important information, while retaining too many may lead to unclear analysis.
-In this work is to compute, for each of the representative years of the transition (except 2020),  the PCs capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
-At the end of this step, we have a list of $m$ PCs, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and
+For each of the representative years of the transition (except 2020), we compute the PCs that capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
+At the end of this step, a list of $m$ PCs is obtained, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and
 
@@
 where $m$ is presumably different for each representative year and $p$ is the total number of variables, hence the maximum number of PCs.
-For the entire transition, it gives a total of $M$ $\text{PC}_{y,i}$.\\
+For the entire transition, it gives a total of $M=34$ $\text{PC}_{y,i}$.\\
 
@@
 Therefore, in this work, PCs are considered similar if their cosine similarity, $S_C (A,B)$, is either greater or equal to 90\% or lower or equal to -90\% (see Equation (\ref{eq:cos_similarity})).
-It is important to consider to second specific case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors.
+It is important to consider the second case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors.
 For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$.
@@
 The components of these similar PCs are then averaged to form the first PC of the transition, $\text{PC}_{\text{transition},1}$.
-Then, the process repeats with the $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$.
+Then, the process repeats with the next $\text{PC}_{y,i}$ with the highest absolute variance but that has not been integrated in the construction of $\text{PC}_{\text{transition},1}$, to form $\text{PC}_{\text{transition},2}$.
 This goes on until the sum of the absolute variance of the $N$ $\text{PC}_{y,i}$ used to construct these $\text{PC}_{\text{transition}}$ is greater or equal to $\alpha$\% of the sum of the absolute variance of all the $M$ $\text{PC}_{y,i}$, \ie the ``total transition variance'', generated at the previous step:
@@
 
-In our case, we have decided to set the value of $\alpha$ to 80\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{y}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).
-For instance, capturing up to 90\% of the the total transition variance would mean almost doubling the number of $\text{PC}_{y}$ to consider and, consequently, the number of directions of variation to assess.
+In our case, we have decided to set the value of $\alpha$ to 85\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{\text{transition}}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).
+Besides being the metrics to assess the robustness of roadmap, these $\text{PC}_{\text{transition}}$ also point out the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.
 
@@
 \includegraphics[width=0.5\textwidth]{PC_y_share_tot_var.pdf}
-\caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{y}$.
-Passing a certain threshold, here 80\%, the number of $\text{PC}_{y}$ to consider increase significantly compared to the share of transition variance that they capture.}
+\caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{\text{transition}}$.
+Passing a certain threshold, here 85\%, the number of $\text{PC}_{\text{transition}}$ to consider increase significantly compared to the share of transition variance that they capture.}
 \label{fig:PC_y_share_tot_var}
@@
 \noindent
-Similarly to the work of \citet{moret2020overcapacity}, roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimizations.
-The $\text{PC}_{\text{transition}}$ are the ``direction/metrics'' on which are projected the results from the myopic pathway optimization subject to minimal installed capacities set by these roadmaps (see Figure \ref{fig:Methodo_Robustness}).
+In their analysis, \citet{moret2020overcapacity} assessed the additional capacities after the realisation of uncertainties, the so-called \textit{possibility of recourse} \cite{grossmann2016recent}.
+These adaptations from the initial investments lead to overcapacity where assets are built but not used because the values taken by uncertain parameters make them cost inefficient.
+To address the same question of overcapacity in the context of energy transition pathway with progressively revealed uncertainties, we have adapted the method of \citet{moret2020overcapacity} (see Figure \ref{fig:Methodo_Robustness}).
+Roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimisations.
+The $\text{PC}_{\text{transition}}$ are the metrics on which the results from the myopic pathway optimisation subject to minimal installed capacities set by these roadmaps are projected.
 

### 2024-04-23 chap_RL/chap_RL.tex
RobPol + RL: Review by FC (part 1)

@@
 Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}.
-For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track with the 1.5°C (if not, 2°C) increase of global temperature by the end of the century.
+For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century.
 On the contrary, using their top-down model DICE, \citet{nordhaus2014question} state that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.
-This is even more valid when models assess a myopic transition pathway subject, with limited foresight through the future with progressively unveiled uncertainties.
 
@@
 At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}.
-Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the 5-year steps transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
+Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
 The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
-However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10,and relies on probability distribution that are often difficult to define properly.
-Increasing the number of these uncertainties in stochastic programming usually leads to a computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
+However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly.
+Increasing the number of these uncertainties in stochastic programming increases exponentially the computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
 Based on the approach of \citet{bertsimas2004price}, and similarly to \citet{Moret2017PhDThesis}, \citet{nicolas2021robust} rather opted for the robust optimisation of the global pathway up to 2200 given different temperature deviation targets, \ie 2 or 3°C by 2200 via the use of uncertainty budget, $\Gamma$, in the TIAM-World model \cite{loulou2005documentation}.
 Considering 9 climate parameters and their respective lower and upper bounds, the idea behind the uncertainty budget stems from the improbability of all parameters simultaneously reaching one of their two extreme values.
+However, robust optimisation fails to provide a comprehensive hedging strategy, that can help guide near-term actions, like stochastic programming.
 
-In the exploration of the myopic transition pathway under uncertainties, we decided to investigate the \gls{RL} approach to benefit from its policy optimisation mechanism.
+To keep the best of both worlds, \ie stochastic programming and robust optimisation, we decided to investigate the \gls{RL} approach, and its policy optimisation mechanism, to explore myopic transition pathways under uncertainties.
 Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
@@
 Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of system \cite{perera2021applications}.
-This thesis rather investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.
+This thesis also investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.
 
@@
 
-Finally, comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight, if not myopic, optimisation without learning process, highlights the added-value brought by the optimised policy.
+Finally, comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.
 
@@
 As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}).
-In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment.
+In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment, \ie that affects parameters of the Linear Programming in EnergyScope Pathway.
 Then, the window 2020-2030 is optimised via EnergyScope.
 Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state he took it.
-Based on these two pieces of information, \ie the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised.
-This goes on until eventually reaching 2050.
-The main purpose of this section is to define the shape of the reward as well as the sets of actions and states.\\
+Based on the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised.
+This goes on until 2050.
 
-\myparagraph{Actions}\\
+\subsection{Actions}
+\label{subsec:RL:act_states_rew:act}
 
-\noindent
 Defining the levers of action, the core of the policy, to support the transition of a country-size whole-energy system is challenging, especially when accounting for political and socio-technical aspects \cite{castrejon2020making}.
-In our work, focusing only on the techno-economic aspect, we assume that the actions taken by the agent are directly implemented and impacting the environment, without ``misfire''.
+In our work, focusing only on the techno-economic aspect, we assume that the actions taken by the agent are directly implemented and impacting the environment.
 In other words, considering only the techno-economic lens, there is no moderation nor contest towards the agent's actions, as the objective is to assess how far and when within the transition to push the different levers of action.
@@
 \label{eq:RL:act_gwp}
-&\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})\leq \mathrm{act}_{\mathrm{gwp}} \cdot \textbf{GWP\textsubscript{tot}}(2020) &
+&\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})\leq \mathrm{act}_{\mathrm{gwp}} \cdot \textbf{GWP\textsubscript{tot}}(2020). &
 \end{flalign}
@@
 
-Out of the total \gls{GHG} emissions in Belgium in 2020,oil (\ie so-called \gls{LFO} in the model), on the one hand and, on the other hand, fossil gas, account for roughly 40\% and 31\%, respectively.
-Then, even though its use in 2020 is much more limited compared to the two formers, \ie 28\,TWh of solid fossil fuels (\ie so-called COAL in the model) versus 159 and 142\,TWh for oil and fossil gas, respectively, coal is a cheap, 17€/MWh, and highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
-For these reasons, three additional actions support the strict limitation of overall emissions of the first action: limiting the consumption of these three fossil resources up to the level of consumption in 2020, $\textbf{Cons\textsubscript{fossil gas}}(2020)$, $\textbf{Cons\textsubscript{LFO}}(2020)$ and $\textbf{Cons\textsubscript{coal}}(2020)$,  over the entire concerned time window, except the first one as this year is the initial condition of the time window and cannot be optimised any more:
+\noindent
+This action is equivalent to setting a national \ce{CO2}-quota.
+
+Three additional actions support the strict limitation of yearly emissions: limiting the consumption of oil, fossil gas and coal.
+Out of the total \gls{GHG} emissions in Belgium in 2020, oil (\ie so-called ``\gls{LFO}'' in the model) and fossil gas account for roughly 40\% and 31\%, respectively.
+In 2020, solid fossil fuels (\ie so-called ``coal'' in the model) is much less consumed than oil and gas: \ie 28\,TWh of solid fossil fuels versus 159 and 142\,TWh for oil and fossil gas, respectively.
+Even though its cost (17€/MWh) makes coal cost-competitive, it is a highly-emitting resource, 0.40\,kt$_{\ce{CO2},\text{eq}}$/GWh.
+For these reasons, three independent actions limit the consumption of these three fossil resources up to the level of consumption in 2020, $\textbf{Cons\textsubscript{fossil gas}}(2020)$, $\textbf{Cons\textsubscript{LFO}}(2020)$ and $\textbf{Cons\textsubscript{coal}}(2020)$,  over the entire concerned time window, except the first one as this year is the initial condition of the time window and cannot be optimised any more:
 
@@
 where $\mathrm{act}\textsubscript{fossil gas}$, $\mathrm{act}\textsubscript{LFO}$ and $\mathrm{act}\textsubscript{coal}$ can take values between 0 and 1.
-These complete the action space of the agent, $A\in \mathbb{R}^4_{[0,1]}$.\\
+These complete the action space of the agent, $A\in \mathbb{R}^4_{[0,1]}$.
 
-\myparagraph{Reward}\\
+\subsection{Reward}
+\label{subsec:RL:act_states_rew:rew}
 
-\noindent
-Properly defined the reward fed by the environment to the agent is crucial in \gls{RL} for several reasons.If the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
+When the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
 Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
@@
 
+The reward function is defined in three steps.
 First of all, taking a set of actions at a certain state might lead to an infeasible optimisation problem.
@@
 In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300.
-If EnergyScope is able to provide a solution to the time window to optimise and the end of the transition, \ie 2050, is not reached, a test on the cumulative emissions so far.
+If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far is evaluated.
 On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent.
@@
 On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded.
-Eventually, if reaching 2050, we decided to tweak the reward function in capping the share of the cumulative emissions and integrating the transition cost.
-Given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
-Moreover, to make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans.
+Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
+On top of that, the reward function includes a secondary objective: the cumulative transition cost.
+To make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans.
 cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€.
@@
 The values of these weights are the results of a trial and error.
-This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.\\
+This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.
 
-\myparagraph{States}\\
+\subsection{States}
+\label{subsec:RL:act_states_rew:states}
 
-\noindent
 Besides the reward, states are the other piece of information provided by the environment to the agent.
@@
 
-Consequently, after another process of trial and error, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window.
+Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window.
 The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050.
@@
 Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment.
-Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified by in Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency.
+Finally, to enrich the level of details, we have added two other dimensions representative of the key-to-the-transition indicators identified in the Renewable Energy Directive (RED) III of the European Commission \cite{REDIII}: the share of renewables in the primary energy mix and, the energy efficiency.
 The former is computed as the share of local renewables (\ie wind, solar, hydro and biomass) and imported renewable energy carriers (\ie biofuels and electrofuels) in the total consumption of primary energy.
@@
 For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium.
-At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
+At the beginning of each episode, the agent starts with the Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
 Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
-This sample will remain valid for the following time windows.
-In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).
+This single draw will affect the parameters for all the subsequent time windows (see Figure \ref{fig:ranges_transition}).
 Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward.
@@
 Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
-\\
 
-\myparagraph{Reward and success}\\
+\subsection{Reward and success}
+\label{subsec:RL:learning:rew_succ}
 
 The learning phase has been split into batches of 500 steps.
-For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
+%For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
 This makes sure replay buffer is full enough for useful updates.
@@
 The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal.
-Right hand side of Figure \ref{fig:reward_success} shows the success rate as the share of transitions meeting the \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) over the total number of attempted transitions, \ie episodes.
-Even though this success rate is not what drives the agent's optimisation, it shows that the shape of the reward (see Figure \ref{fig:Reward}) leads towards more and more successes.
+As this successes drive indirectly the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes.
 
@@
 The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view.
-The evolution of the success rate also shows the shape of the reward aims at more and more successful transitions.}
+The evolution of the success rate also shows the reward function aims at more and more successful transitions.}
 \label{fig:reward_success}
@@
 During the learning process, the algorithm explores numerous transition pathways: 2037 successful transitions out of 10,751 attempts.
-Each pathway provides valuable insight into the best course of actions --- the primary goal of reinforcement learning.
+Each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning.
 As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken.
@@
 Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive.
-This is another indication that the shape of the reward is appropriate in this exploration of successful transition pathways.
+This is another indication that the reward function is appropriate in this exploration of successful transition pathways.
 
@@
 
-\myparagraph{States}\\
+\subsection{States}
+\label{subsec:RL:learning:states}
 
@@
 Per definition, the threshold of 1.2\,Gt$_{\ce{CO2},\text{eq}}$ splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
-In the successful transitions, the median cumulative emissions, $\text{P}_{50}$,are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
-This is due to the combination of efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
+In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
+Reaching cumulative emissions significantly lower than the \ce{CO2}-budget are possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
 Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$, 12\% higher than the \ce{CO2}-budget.
@@
 
-\myparagraph{Actions}\\
+\subsection{Actions}
+\label{subsec:RL:learning:actions}
 
@@
 Besides this observation, no specific range of values for the different actions at the different timing seems to lead to more successes.
-In other words, there is no clear set of actions that would support more effectively the transition.
+Looking at action individually, there does not seem to be any that supports more effectively the transition.
+The success come from the combination of these actions.
 
@@
 In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}).
-When the problem is bounded and feasible, these lines are the edges of a convex polygon, the domain of feasibility.
+When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility.
 The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function.
@@
 After filtering out failures of the learning episodes and keeping only the successful transitions, only a limited set of the actions are binding and have an actual impact on the result of the optimisation in EnergyScope Pathway (see Figure \ref{fig:Binding_learning}).
-This allows identifying key actions to support the myopic transition.Limiting the \gls{GWP} in the near-term is a key-factor for success.
+This allows identifying key actions to support the myopic transition.
+Limiting the \gls{GWP} in the near-term is a key-factor for success.
 However, this action has an effective impact on the environment only when it forces the system to be close to carbon-neutrality.
@@
 When it comes to limiting the use of \gls{LFO} and coal, the conclusions are more straightforward.
-At the beginning of the transition, most of the 159TWh of \gls{LFO} are consumed by naphtha-cracker (46\%) and decentralised boilers (45\%).
+At the beginning of the transition, most of the 159\,TWh of \gls{LFO} are consumed by naphtha-cracker (46\%) and decentralised boilers (45\%).
 The remaining 10\% are consumed by industrial boilers.
@@
 On the contrary, limiting coal is always binding.
-Before all, this is due because coal is a cheap resource (17€/MWh).
+Before all, this is due because coal is a cheap resource (17\,€/MWh).
 In other words, the cost-driven environment will favour it.
@@
 \label{sec:RL:conclusions}
-In the literature, two options are investigated to explore transition pathways of a whole-energy system.
-On the one hand, in perfect foresight optimisation, a full knowledge of the parameters is assumed over the entire time horizon.
-On the other hand, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon, to add more realism \cite{poncelet2016myopic}.
-To meet objective of climate change mitigation, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
+In the literature, two options are investigated to explore transition pathways of a whole-energy system: perfect foresight and myopic.
+In perfect foresight optimisation, a full knowledge of the parameters is assumed over the entire time horizon.
+To add more realism, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon \cite{poncelet2016myopic}.
+To respect a \ce{CO2}-budget over the transition, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
 However, since the effect of \ce{CO2}-emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself.
@@
 
-In \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
+%In \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
 The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt$_{\ce{CO2},\text{eq}}$ equivalent to 10 years at the current level of emissions.

### 2024-04-23 chap_RobPol/chap_RobPol.tex
RobPol + RL: Review by FC (part 1)

@@
 However, in the case of the transition pathway of a whole-energy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''.
-Between the sole total transition cost and the entire set of installed technologies that give too few and too much information, respectively, the performance metric here is defined through the \gls{PCA} approach.
+Between the total transition cost giving too few information and the entire set of installed technologies giving too much, the performance metric here is defined through the \gls{PCA} approach.
 Then, when comes the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps.
-In other terms, one roadmap will not be robust or not in itself but rather more or less robust than another one.
+Each roadmap will be ranked in terms of robustness.
 
@@
 These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.
-As introduced in Section \ref{subsec:meth:PCA:transition}, the data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}).
+Data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}).
 This gave 1260 different transitions resulting, for each of them, from the pathway optimisation subject to a sample of uncertain parameters (see Section \ref{subsec:uncert_charac}).
@@
 \label{subsec:RobPol:PC_year}
-Before computing the \gls{PCs} of each representative year of the transition, we pre-preprocessed the raw data though scaling and managing the outliers detailed in Section \ref{subsec:meth:PCA:transition}.
-The second step discarded decentralised \gls{HP} running on methane.
-Without managing the outliers, this technology has a contribution in the \gls{PCs} of transition for several reasons.
+Before computing the \gls{PCs} of each representative year, we pre-preprocessed the raw data through scaling and managing the outliers (see Section \ref{subsec:meth:PCA:transition}).
+Managing the outliers is needed when assessing the main directions of variation of the transition.
+For example, decentralised \gls{HP} running on methane is discarded.
+This technology would have a contribution in the \gls{PCs} of transition for several reasons.
 First of all, the \gls{LT} heating sector represents between 22\% and 27\% of the \gls{EUD} in the case of Belgium.
@@
 
-First of all, before investigating the \gls{PCs}, it is worth looking at the total variance of each representative year (see Table \ref{tab:design_variance}).
+The total variance decreases by more than 50\% between the earlier stages and the final ones (see Table \ref{tab:design_variance}).
 Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition.
@@
 \label{subsec:RobPol:PC_transition}
-Based on the \gls{PCs} of each representative year, $\text{PC}_{y}$, the \gls{PCs} of the transition, \ie the metrics to assess the robustness of roadmaps, can be computed.
-Before aggregating and averaging similar $\text{PC}_{y}$, it is necessary to rank them to ensure capturing most of the transition variance in the subsequent $\text{PC}_{\text{transition}}$.
+Based on the \gls{PCs} of each representative year (34 in total), this section identifies a sufficient subset to characterise the roadmaps (and assess their robustness in Section \ref{subsec:RobPol:Projection}).
 
-Based on the design variance captured by each $\text{PC}_{y}$ in their respective representative year, we can rank them (see Table \ref{tab:ranking_PCs}).
+Before aggregating and averaging similar $\text{PC}_{y}$, it is necessary to rank them to ensure capturing most of the transition variance in the subsequent $\text{PC}_{\text{transition}}$.This ranking is based on the design variance captured by each $\text{PC}_{y}$ in their respective representative year (see Table \ref{tab:ranking_PCs}).
 Summing all of these variances over the different years results in a ``pseudo'' total variance of the transition.
@@
 On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''.
-Then, we assess the robustness of these three roadmaps via a similar ``three-step rolling horizon approach'' as in the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.
+Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.
 
@@
 In practice, $\Gamma_{\text{obj}}= 1$ means considering the cost of purchasing electrofuels at its worst value.
-$\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value in the lot, and so on.
-In the present analysis, we consider the ``robust'' roadmap by setting $\Gamma_{\text{obj}}= 6$.
-Consequently, costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies are considered at their worst values, \ie the upper bound of their respective range.
+$\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value, and so on.
+In the present analysis, we label the roadmap as ``robust'' (ROB) by setting $\Gamma_{\text{obj}}= 6$.
+Consequently, the following six parameters are set at their worst value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies.
 In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
@@
 
-In the transport of freight, the ROB roadmap transitions from diesel to \gls{FC} trucks with the temporal use of electric trucks from 2025 to 2035.
-Similarly to the industrial electric heaters, it is a way to integrate the early installation of solar \gls{PV} and absorb part of its intermittent electricity.
+In the ROB roadmap, the transport of freight by trucks first transitions from diese to electric trucks and then, after 2035, to \gls{FC} trucks.
+Used as a way to support the the early installation of solar \gls{PV}, electric trucks are then replaced by \gls{FC} ones given the limited availability of renewable electricity and its more cost-effective use in other sectors such as private mobility.
 Freight transport via trains and boats are not different from the REF roadmap.
@@
 In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}.
-More expensive imported energy carriers lead to, on the one hand, an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) supported by a higher electrification of the heating and transport sectors.
-On the other hand, it increases the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
+More expensive imported energy carriers lead to an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) and an increase of the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
 This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}.
@@
 
-No tested roadmap is more robust than the others on all the \gls{PCs} of transition (Table \ref{tab:projection}).
-The SMR roadmap spans over similar ranges as the REF case.
+The SMR roadmap spans over similar ranges as the REF case (see Table \ref{tab:projection}).
 The most significant difference concerns PC 5.
@@
 
-The ROB roadmap is more robust than the REF case for all PCs but PC 4.
-About PC 1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors.
+The ROB roadmap is more robust than the REF case for the first three PCs.
+About PC\,1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors.
 In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations.
-For the same reasons, the ROB roadmap is 37\% more robust for PC 2.
+For the same reasons, the ROB roadmap is 37\% more robust for PC\,2.
 The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers.
 In myopic transitions though, this technology keeps on being installed in the mid-term.
-Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
-Finally, the ROB roadmap is 21\% less robust for PC 5.
+Concerning PC\,3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.
+Finally, the ROB roadmap is 21\% less robust for PC\,5.
 Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller.
@@
 \centering
-\begin{tabular}{c| c c}
+\begin{tabular}{c l| c c}
 \toprule
-\multirow{3}{*}{\textbf{PC}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
-& SMR & ROB \\
+\multirow{2}{*}{\textbf{PC}} & \multirow{2}{*}{\textbf{Main contributors}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
+& & SMR & ROB \\
 \midrule
-1 & $\simeq$ & +49\% \\
-2 &  $\simeq$ & +37\%\\
-3 &  $\simeq$ & +33\%\\
-4 & $\simeq$ & $\simeq$ \\
-5 &  +34\% & -21\%\\
+1 & \gls{PV} and electrification & $\simeq$ & +49\% \\
+2 & \gls{PV} and dec.heating & $\simeq$ & +37\%\\
+3 & Freight transport & $\simeq$ & +33\%\\
+4 & Dec. oil boilers & $\simeq$ & $\simeq$ \\
+5 &  Naphtha-crackers and \gls{MTO} & +34\% & -21\%\\
 \bottomrule							
@@
 Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}.
-From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation the additional capacities needed in myopic optimisations subject to uncertainties.
+From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation of the additional capacities needed in myopic optimisations subject to uncertainties.
 This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation.

### 2024-04-23 chap_case_study/chap_case_study.tex
RL chapter: submission for review
RobPol + RL: Review by FC (part 1)

@@
 Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.
-Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.
+Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C above pre-industrial levels by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.
 
@@
 On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are directly limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use.
+Similarly to \citet{limpens2021generating}, we constrained the endogenous non-renewable wastes to be consumed locally.
 The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh.

### 2024-04-23 chap_intro_ccl/conclusion.tex
RobPol + RL: Review by FC (part 1)

@@
 Regarding the \gls{RL} approach, two threads of future development have been identified: reward shaping and multi-fidelity.
-Rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}.
+First, the reward function could be adapted to incentivise more the successful transitions.
+Then, rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}.
 Reward shaping aims at accelerating the learning or guiding the agent towards achieving the desired behaviour more efficiently.

### 2024-04-24 appendices/app_RL_TD_MO.tex
RobPol + RL: Review by FC (part 2)

The \acrfull{RL}-based exploration of myopic transition pathways can require thousands of runs to converge. To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time-resolution, the so-called transfer learning \cite{mann2013directed}.  Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}). This section aims at comparing results from the learning process on the monthly model with the one on the hourly model. The monthly learning has been carried out by applying the same rules (see Section \ref{sec:RL:act_states_rew}). The purpose is to assess the capacity of the monthly model to be a good proxy of the hourly myopic transitions.

Considering the actions taken by the agent and their binding effect on the environment with a monthly resolution,  we see similar trends as in the hourly model (see Figure \ref{fig:app:Binding_constr_joss}). Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation. Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon-neutrality becomes more crucial at the end of the transition.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Binding_learning_TD_joss.pdf}
\includegraphics[width=0.49\textwidth]{Binding_learning_MO_joss.pdf}
\caption{Binding versus non-binding constraints on the hourly and monthly models. When the agent learns on the monthly model, the actions are binding in the similar timing and similar ranges as when learning on the hourly model. }
\label{fig:app:Binding_constr_joss}
\end{figure}

Besides the sharper decrease of emissions early in the transition, the monthly myopic pathways fit with the trend provided by \gls{RL} optimisation on the hourly model. The difference in the early drop of emissions between the monthly and hourly models comes from the easier integration of local \gls{VRES} as the hourly intermittency is averaged over a month (see Appendix \ref{app:mo_vs_td}).

%\begin{figure}[!htbp]
%\centering
%\includegraphics[width=0.8\textwidth]{Gwp_pathway_app.pdf}
%\caption{Comparison of \ce{CO2}-emissions pathways from the \gls{RL}-based myopic optimisation on the hourly and monthly models.}
%\label{fig:app:Gwp_pathway}
%\end{figure}

Even more than for the emissions, the monthly model is a good proxy to assess the pathways of annual system cost as well as cumulative costs such as CAPEX, OPEX and salvage value by 2050 (see Figures \ref{fig:app:System_cost_pathway} and \ref{fig:app:Opex_Capex_Salvage_comp}).

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{System_cost_pathway_app.pdf}
\caption{Comparison of annual system cost pathways from the \gls{RL}-based myopic optimisation on the hourly and monthly models.}
\label{fig:app:System_cost_pathway}
\end{figure}

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Opex_2050_comp_app.pdf}
\includegraphics[width=0.325\textwidth]{Capex_2050_comp_app.pdf}
\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_app.pdf}
\caption{Comparison of cumulative OPEX (left), CAPEX (right) and salvage value (right) in 2050 from the \gls{RL}-based myopic optimisation on the hourly and monthly models.}
\label{fig:app:Opex_Capex_Salvage_comp}
\end{figure}

\newpage
The monthly model also captures the need to import more electrofuels by 2050 in the myopic transitions (see Figure \ref{fig:app:Mix_2050_comp}). The other main difference concerns the import of electricity from neighbouring countries. As detailed in Appendix \ref{app:mo_vs_td}, this comes from the averaging of the hourly times series of \gls{VRES} that can provide electricity in a more constant way.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.8\textwidth]{Mix_2050_comp_app.pdf}
\caption{Comparison of the primary energy mix in 2050 from the \gls{RL}-based myopic optimisation on the hourly and monthly models. To a lesser extent that the hourly model, the monthly optimisation also captures the need to import more electrofuels when the foresight is limited.}
\label{fig:app:Mix_2050_comp}
\end{figure}

In conclusion, with the advantage of the smaller computational burden, the monthly model myopic model is found to be a good proxy with its equivalent based on hourly time-resolution.

### 2024-04-24 chap_RL/chap_RL.tex
RobPol + RL: Review by FC (part 2)

@@
 
-To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model.
-Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
-
 \subsection{Reward and success}
@@
 Reaching cumulative emissions significantly lower than the \ce{CO2}-budget are possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
-Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$, 12\% higher than the \ce{CO2}-budget.
+Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
 As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Cum_gwp_cost.pdf}
+\includegraphics[width=0.8\textwidth]{Cum_gwp_cost_2.pdf}
 \caption{Exploration of the state space over the learning process: cumulative emissions and costs.
-Successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
+The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
 }
@@
 \label{sec:RL:testing}
-After investigating the exploration of transition pathways on the monthly model, this section compares these results with two references: the \gls{RL}-agent trained on the hourly model and the perfect foresight under uncertainties from the \gls{UQ} analysis presented in Chapter \ref{chap:atom_mol}.
-The first reference has been built by applying the same rules (see Section \ref{sec:RL:act_states_rew}) but on the hourly model over a more limited number of batches, 43 versus 80, leading to 928 versus 2037 successful transitions for the results presented in Section \ref{sec:RL:learning}.
-First, the purpose is to assess the capacity of the monthly model to be a good proxy of the hourly myopic transitions and, then,  compare the explored pathways with the perfect foresight under uncertainties.
-
-Considering the actions taken by the agent and their binding effect on the environment with a hourly resolution,  we see similar trends as in the monthly model (see Figure \ref{fig:Binding_learning_TD}).
-Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation.
-Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon-neutrality becomes more crucial at the end of the transition.
-
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=0.8\textwidth]{Binding_learning_TD.pdf}
-\caption{When the agent learns on the hourly model, the actions are binding in the similar timing and similar ranges as when learning on the monthly model. }
-\label{fig:Binding_learning_TD}
-\end{figure}
+After investigating the exploration of myopic transition, this section compares these results with the perfect foresight under uncertainties from the \gls{UQ} analysis presented in Chapter \ref{chap:atom_mol}.
+To reduce the computational time, the monthly model could also be used to carry out \gls{RL}-based exploration of myopic pathways.
+Appendix \ref{app:app_RL_TD_MO} presents the comparison between the results obtained from the learning process on the hourly and monthly models.
 
@@
 Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition.
-This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 9 or more Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
+This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 10 or more Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
 These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% on average by 2050.
-The difference in the early drop of emissions between the monthly and hourly models comes from the easier integration of local \gls{VRES} as the hourly intermittency is averaged over a month (see Appendix \ref{app:mo_vs_td}).
-Besides this sharper decrease of emissions early in the transition, the monthly myopic pathways fit with the trend provided by \gls{RL} optimisation on the hourly model.
 Finally, the long-term vision of the perfect foresight approach results in smoother reduction of the emissions to end up with less emissions by 2050.
+
+The comparison between the failures and the pathways demonstrates the added-value brought by myopic pathway optimisation.
+In the near-term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed.
+This shows that limited foresight encourages to strongly act at the early stages.
+On top of this, following the initial steps of \ce{CO2}-emissions pathways resulting from PF approach would likely ($\sim$80\%) lead to fail the transition.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Gwp_pathway.pdf}
-\caption{Comparison of \ce{CO2}-emissions pathways from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\includegraphics[width=0.8\textwidth]{Gwp_pathway_core.pdf}
+\caption{Comparison of \ce{CO2}-emissions pathways from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation. }
 \label{fig:Gwp_pathway}
@@
 Via its actions and its cost-based reward, the agent can reach systems that are cheaper than any other solution obtained by the perfect foresight approach.
-Even more than for the emissions, the monthly model is a good proxy to assess the annual system cost of the pathway on the hourly model.
 The perfect foresight approach gives a wider variability in its cost since this method always found a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{System_cost_pathway.pdf}
-\caption{Comparison of annual system costs pathways from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\includegraphics[width=0.8\textwidth]{System_cost_pathway_core.pdf}
+\caption{Comparison of annual system costs pathways from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
 \label{fig:System_cost_pathway}
@@
 \centering
-\includegraphics[width=0.325\textwidth]{Opex_2050_comp.pdf}
-\includegraphics[width=0.325\textwidth]{Capex_2050_comp.pdf}
-\includegraphics[width=0.325\textwidth]{Salvage_2050_comp.pdf}
-\caption{Comparison of cumulative OPEX (top-left), CAPEX (top-right) and salvage value (bottom) in 2050 from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.}
+\includegraphics[width=0.325\textwidth]{Opex_2050_comp_core.pdf}
+\includegraphics[width=0.325\textwidth]{Capex_2050_comp_core.pdf}
+\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
+\caption{Comparison of cumulative OPEX (left), CAPEX (right) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
 \label{fig:Opex_Capex_Salvage_comp}
@@
 The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}.
-The assessment of the primary energy mix by the 2050 highlights where the differences of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
-In the majority of the cases, e-ammonia is twice more imported in the myopic transitions.
-This is even more the case for the myopic transitions carried out on the hourly model.
+The assessment of the primary energy mix by the 2050 highlights that the differences of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
+In the majority of the cases, e-ammonia is more than two times more imported in the myopic transitions.
 Being cheaper than e-methane (see Chapter \ref{chap:case_study}), e-ammonia brings flexibility in the production of electricity via \gls{CCGT} (see Chapter \ref{chap:atom_mol}).
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Mix_2050_comp.pdf}
-\caption{Comparison of the primary energy mix in 2050 from the perfect foresight optimisation under uncertainties ($\text{REF}_{\text{TD-PF}}$), the \gls{RL}-based myopic optimisation on the monthly ($\text{RL}_{\text{MO-MY}}$) and hourly ($\text{RL}_{\text{TD-MY}}$) models.
-The biggest difference is about e-ammonia to supply \gls{CCGT}}
+\includegraphics[width=0.8\textwidth]{Mix_2050_comp_core.pdf}
+\caption{Comparison of the primary energy mix in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.
+The biggest difference is about e-ammonia to supply \gls{CCGT}.}
 \label{fig:Mix_2050_comp}
@@
 
-With the advantage of the smaller computational burden, the monthly model myopic model was found to be a good proxy with its equivalent based on hourly time-resolution.
-
-Finally, we have compared the results coming from this \gls{RL}-based myopic optimisation with monthly time-resolution with the hourly perfect foresight approach.
+We have compared the results coming from this \gls{RL}-based myopic optimisation with the hourly perfect foresight approach.
 These myopic optimisations provided pathways respecting the \ce{CO2}-budget that were more drastic in cutting emissions in the short-term than the perfect.

### 2024-04-25 appendices/app_EnergyScope.tex
SUBMISSION

@@
 % Specific need & criteria
-
+\newpage
 In accordance with the previous classification, models aimed at aiding decision-makers in the energy transition primarily fall under the categories of planning and scenario analysis, with a lower technical resolution than the other classes of model (see Figure \ref{fig:energy_models_classification}).
@@
 In the conclusion, it is claimed `\textit{H2RES offers practically unlimited potential for functionality expansion since it is an open-source program}´ which open the doors for future developments to encompass new features.}
-&  {\color{lightgray} \checkmark}\footref{foot:h2res} & \checkmark & \checkmark \\ 
+&  {\color{lightgray} \checkmark} & \checkmark & \checkmark \\ 
 iHOGA & \cite{ihoga} & \checkmark & {\color{gray} \xmark}\footref{foot:transportnotsaid}\footref{foot:industrynotaccounted}\footref{foot:dhnnotaccounted}\footref{foot:decLTHnotaccounted} & {\color{lightgray} \checkmark} \footref{foot:invopt} & \checkmark & {\color{lightgray} \checkmark} \footref{foot:freeundersomespecial} \\
@@
 
+\newpage
 Finally, to properly capture the inter days dynamics, EnergyScope TD uses the ``Coupling typical days'' method from \citet{Gabrielli2018}.
@@
 
+\newpage
 The proposed formulation relies on representative years, selected every 5 years from 2020 to 2050.  The period between two of them is called `\emph{PHASE}´.
@@
 \includegraphics[width=\textwidth]{path_e.g._tech.pdf}
-\caption{Example of how the technology capacities and associated variables are evolving.
+\caption{Example of how the technology capacities and associated variables are evolving. }
+%\caption{Example of how the technology capacities and associated variables are evolving.
 The example uses a technology with a 20 years lifetime.

### 2024-04-25 appendices/app_case_study.tex
SUBMISSION

@@
 
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{Investment_transition.pdf}
+\caption{The cumulative investments over the transition are unequally spread between the sectors.
+The energy system in 2020 is imposed as the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
+The investments required during the transition represents 150\% the initial investment and mainly in the same three sectors.
+}
+\label{fig:pestd_cumul_inv}
+\end{figure}
+
 Initially, the infrastructure, transport, and electricity sectors each account for approximately one-third of the investments.
@@
 At Belgian level, McKinsey (2023) calculates an average annual additional requirement equivalent to 2 to 3% of GDP.
-
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=\textwidth]{Investment_transition.pdf}
-\caption{The cumulative investments over the transition are unequally spread between the sectors.
-The energy system in 2020 is imposed as the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
-The investments required during the transition represents 150\% the initial investment and mainly in the same three sectors.
-}
-\label{fig:pestd_cumul_inv}
-\end{figure}
 
@@
 
+\newpage
 \section{Myopic versus perfect foresight pathway optimisation}
 \label{app:my_vs_pf}
-This section aims at digging more into the details of the differences observed between the myopic approaches and the reference case (\ie hourly perfect foresight model).
+This section aims at digging more into the details of the differences observed between the myopic approaches and the reference case, the hourly perfect foresight model (see Table \ref{tab:PF_MY}).
 
@@
 
-Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€ in 2030 whereas the overall cumulative investments, so far, are 421.3~b€.
-The significant lifetime (\eg 80 years) and investment cost (\ie 368M€/GW$_\text{VRES}$ \cite{readthedocs_pathway}) of the power grid, and, on a smaller scale, the district heating network, explain why the myopic optimisation opts for a higher investment in these infrastructures, at early stages.
-Similarly to what \citet{keppo2010short} observed in their studies, these early investments consequently lead to more investments, later in the transition, to renew technologies that have become too old before 2050: during the phase between 2045 and 2050, the myopic approach needs to invest in 9.2GW of PVs that have been installed 25 years before whereas the perfect foresight, by smoothing its investments over the entire transition, has to renew only 2.5GW of PVs.
-
 \begin{figure}[!htbp]
@@
 \label{fig:my_pestd_total_cost_cum_diff}
+\end{figure}
+
+Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€ in 2030 whereas the overall cumulative investments, so far, are 421.3~b€.
+The significant lifetime (\eg 80 years) and investment cost (\ie 368M€/GW$_\text{VRES}$ \cite{readthedocs_pathway}) of the power grid, and, on a smaller scale, the district heating network, explain why the myopic optimisation opts for a higher investment in these infrastructures, at early stages.
+Similarly to what \citet{keppo2010short} observed in their studies, these early investments consequently lead to more investments, later in the transition, to renew technologies that have become too old before 2050: during the phase between 2045 and 2050, the myopic approach needs to invest in 9.2GW of PVs that have been installed 25 years before whereas the perfect foresight, by smoothing its investments over the entire transition, has to renew only 2.5GW of PVs.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=\textwidth]{MY_investment_return_area.pdf}
+\caption{Sum of the respective salvage values of each sector, per end year of each time window.
+All together, these salvage values represent 223.9~b€, 53\% of the cumulative investment costs in 2030, and 157.7~b€, 24\% of the cumulative investment costs in 2050.}
+\label{fig:my_pestd_inv_return_area}
 \end{figure}
@@
 
-In general, due to the formulation of the salvage value (see Equation (\ref{eq:salvage})), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition.
-Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}.
-In parallel, in the near term, the system relies on average slightly more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.
-
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=\textwidth]{MY_investment_return_area.pdf}
-\caption{Sum of the respective salvage values of each sector, per end year of each time window.
-All together, these salvage values represent 223.9~b€, 53\% of the cumulative investment costs in 2030, and 157.7~b€, 24\% of the cumulative investment costs in 2050.}
-\label{fig:my_pestd_inv_return_area}
-\end{figure}
-
  \begin{figure}[!htbp]
@@
 \end{figure}
+
+In general, due to the formulation of the salvage value (see Equation (\ref{eq:salvage})), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition.
+Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}.
+In parallel, in the near term, the system relies on average slightly more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.
 

### 2024-04-25 chap_RL/chap_RL.tex
SUBMISSION

@@
 For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century.
-On the contrary, using their top-down model DICE, \citet{nordhaus2014question} state that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.
+On the contrary, using his top-down model DICE, \citet{nordhaus2014question} states that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.
 
-To address this issue, several approaches have been used.
-Among them,  multi-stage stochastic programming is often put forward as a promising method.
-Stochastic programming formulates the problem as a mathematical program with probabilistic constraints or objective functions.
-These models explicitly consider the uncertainty by incorporating probability distributions for the uncertain parameters.
-The goal is to find an optimal decision that minimizes/maximizes the expected value of the objective function while satisfying the probabilistic constraints, modelled as a scenario tree.
-At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}.
-Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
-The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
-However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly.
-Increasing the number of these uncertainties in stochastic programming increases exponentially the computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
-Based on the approach of \citet{bertsimas2004price}, and similarly to \citet{Moret2017PhDThesis}, \citet{nicolas2021robust} rather opted for the robust optimisation of the global pathway up to 2200 given different temperature deviation targets, \ie 2 or 3°C by 2200 via the use of uncertainty budget, $\Gamma$, in the TIAM-World model \cite{loulou2005documentation}.
-Considering 9 climate parameters and their respective lower and upper bounds, the idea behind the uncertainty budget stems from the improbability of all parameters simultaneously reaching one of their two extreme values.
-However, robust optimisation fails to provide a comprehensive hedging strategy, that can help guide near-term actions, like stochastic programming.
-
-To keep the best of both worlds, \ie stochastic programming and robust optimisation, we decided to investigate the \gls{RL} approach, and its policy optimisation mechanism, to explore myopic transition pathways under uncertainties.
-Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
-\gls{RL} exhibits two main advantages: its effectiveness to handle uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimize the policy \cite{perera2021applications}.Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.
+To navigate through the myopic transitions not constrained by a prescribed \ce{CO2}-trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium.
+Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.
 Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2}-budget, are very limited.
@@
 \section*{Contributions}
-\label{sec:meth:contributions}
+\label{sec:RL:contributions}
 Applying the \gls{RL} approach to the optimization of the myopic transition pathway of a whole-energy system presents several novelties.
@@
 
-Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of system \cite{perera2021applications}.
+Applying \gls{RL} to the optimisation environment of EnergyScope Pathway myopic,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework: while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.
+
+Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.
+
+Usually, applications of \gls{RL} focus more on the result of the learning process, the optimised policy, for optimising the control of a system \cite{perera2021applications}.
 This thesis also investigates the learning episodes themselves to explore the field of possibilities to succeed the transition.
-
-Then, applying to this optimisation environment, \ie EnergyScope myopic Pathway,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework.
-In this agent, while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.
-
-Finally, comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.
 
@@
 \label{sec:RL:act_states_rew}
-As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}).
+As already introduced in Section \ref{subsec:meth_RL_algo}, the environment with which the \gls{RL}-agent interacts is the optimisation of the transition pathway of a whole-energy system on a specific time window, \eg 2020-2030 then 2025-2035 and so on, until 2040-2050 (see Figure \ref{fig:Schematics_RL}).
 In a nutshell, starting from the initial state of the environment (\ie the whole-energy system in 2020), the agent takes a set of actions that influence the environment, \ie that affects parameters of the Linear Programming in EnergyScope Pathway.
@@
 
-When the reward is not properly defined, the agent may optimize its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
+When the reward is not properly defined, the agent may optimise its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
 Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
@@
 \centering
-\includegraphics[width=0.65\textwidth]{Reward.pdf}
+\includegraphics[width=0.8\textwidth]{Reward.pdf}
 \caption{Reward function, $R$.
@@
 In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300.
-If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far is evaluated.
+If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far are evaluated.
 On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent.
@@
 On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded.
-Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ambitious ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
+Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
 On top of that, the reward function includes a secondary objective: the cumulative transition cost.
@@
 In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost.
-The values of these weights are the results of a trial and error.
+The values of these weights are the results of a trial and error to fine-tune the balance between more expensive successes and cheaper failures.
 This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.
@@
 The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050.
-Therefore, the cumulative emissions emitted so far in the current step of the transition is the first dimension of the states.
+Therefore, the cumulative emissions emitted so far up to the current step of the transition is the first dimension of the states.
 Similarly, the cumulative cost of the transition so far constitutes the second dimension of the states to inform the agent about the cost-impact of its actions on the environment.
@@
 Where electricity, heat and non-energy \gls{EUD} are expressed in terms of energy content, we needed to convert passenger and freight transports into their respective \gls{FEC} to integrate them in the ratio.
+The information of efficiency fed back by the environment to the agent is the ratio between a ``hybrid'' \gls{EUD} and the consumption of primary energy resources.
 
@@
 
-Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the learning of the \gls{NN}, also called ``training''.
+Before comparing the agent's optimised policy with references, the first step consists in assessing the learning of the \gls{NN}, also called ``training''.
 For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium.
@@
 This goes on until the end of the episode.
-For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+For the next episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), another sample of uncertain parameters is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
 
@@
 As this successes drive indirectly the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes.
+However, given the wide range of uncertainty of some parameters and the agent's levers of action, this success rate stays limited at the end of the learning process.
+In other words, there are conditions where it is impossible for the agent to succeed the transition.
 
@@
 The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view.
-The evolution of the success rate also shows the reward function aims at more and more successful transitions.}
+The evolution of the success rate also shows that the reward function aims at more and more successful transitions.}
 \label{fig:reward_success}
@@
 
-During the learning process, the algorithm explores numerous transition pathways: 2037 successful transitions out of 10,751 attempts.
-Each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning.
+\newpage
+During the learning process, the algorithm explored numerous transition pathways: 2,037 successful transitions out of 10,751 attempts.
+2\% of the total attempts led to infeasible optimisation of the initial time window.
+On top of that, each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning.
 As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken.
@@
 The trends provided below are therefore not proven.
-The randomness of the process and the number of explored transition pathways still give us high confidence.
+The randomness of the process and the number of explored transition pathways still give high confidence in the conclusions drawn from these analyses.
 
@@
 This is another indication that the reward function is appropriate in this exploration of successful transition pathways.
-
-Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
-Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
-In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
+More indirectly, this substantiates that the weights between emissions and cost defined in Section \ref{subsec:RL:act_states_rew:rew} could be used and adapted for another case study.
 
@@
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
-Beyond this point, through this learning process, succeeding the transition represents 38\% of the episodes.}
+Beyond this point, through this learning process, succeeding the transition represents 38\% of the episodes.
+}
 \label{fig:reward_status}
 \end{figure}
+
+\newpage
+Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
+Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
+In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Cum_gwp_cost_2.pdf}
+\includegraphics[width=0.8\textwidth]{Cum_gwp_cost_3.pdf}
 \caption{Exploration of the state space over the learning process: cumulative emissions and costs.
@@
 
+\newpage
 Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.
@@
 Cumulative emissions are skewed to the left and to the right for the successes and failures, respectively.
-The skewness of the cumulative costs for successful transitions is higher compared to failures.On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
+The skewness of the cumulative costs for successful transitions is higher compared to failures.
+On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
 \label{tab:skewness_gwp_cost}
@@
 
+\newpage
 Finally, we observe that the majority of the successful transitions are cheaper than the reference transition cost, 1.1\,b€.
-Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is more increased than the one of electrofuels (see Table \ref{tab:param_RL}).
-In other words, to have higher chances to succeed a myopic transition, the key factor is to reduce the uncertainty on the cost of purchasing electrofuels.
+Among the parameters impacting the most the total transition cost, we observe that success occurs when, on average,  the cost of purchasing fossil fuels is increased more than the one of electrofuels (see Table \ref{tab:param_RL}).
+In other words, to have higher chances to succeed a myopic transition, the key factor is to reduce the uncertainty on the cost of purchasing electrofuels or to increase the cost of fossil fuels.
 Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases.
@@
 Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}.
-In 2050, the success-failure threshold has been observed at 82\%.
-In the REF case of Chapter \ref{chap:atom_mol}, the share of renewables reached 86\% by 2050.
-However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another threshold at lower share of renewables in the mix.
+In 2050, the threshold where occurences of success are more numerous than failures was at 82\% of renewable share.
+In the REF case of Chapter \ref{chap:atom_mol}, this share reached 86\% by 2050.
+However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another ``bump'' at lower share of renewables in the mix.
 This area corresponds to the possibility to install \gls{SMR}.
-As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold to 60\% as in the SMR case of Chapter \ref{chap:atom_mol}.
+As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold as in the SMR case of Chapter \ref{chap:atom_mol}.
 Besides these milestones to respect the \ce{CO2}-budget of the transition, one can also look at the other side of the thresholds.
-Below the near-term threshold of 60\%, this the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.
+Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.
 
-The efficiency, as defined in Section \ref{sec:RL:act_states_rew}, gives less valuable information to direct towards successful transition.
-Through the transition, besides the share of success increasing over the failures, these two scenarios indistinguishably spread over the whole range.
+The efficiency, as defined in Section \ref{sec:RL:act_states_rew}, gives less valuable information towards successful transitions.
+Through the transition, besides the share of success increasing over the failures, the distributions of success and failure indistinguishably spread over the whole range.
 Similarly than the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.
@@
 Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
-Below a near-term threshold (about 60\%), the chances of success are limited, \ie no-go zones.
+Below a near-term threshold ($\sim$60\%), the chances of success are limited, \ie no-go zones.
 Efficiency is a less valuable information for the agent to succeed transitions as failures and successes indistinguishably spread over the whole range.}
@@
 Limiting the \gls{GWP} in the near-term is a key-factor for success.
-However, this action has an effective impact on the environment only when it forces the system to be close to carbon-neutrality.
+However, this action has a binding effect on the environment only at the end of the transition.
 The range over which limiting the use of fossil gas binds the optimisation is wider.
 Compared to other non-renewable fuels, this is due to the longer use of this energy carrier favoured by its low \gls{GWP} (the second after uranium) and its versatility (applications in the electricity, heat and mobility sectors).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Binding_learning.pdf}
+\caption{Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
+Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.
+Limiting coal consumption is always constraining, unlike \gls{LFO} that is ``naturally'' substituted by EnergyScope Pathway in the near-term.}
+\label{fig:Binding_learning}
+\end{figure}
 
@@
 
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=0.8\textwidth]{Binding_learning.pdf}
-\caption{Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
-Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.
-Limiting coal consumption is always constraining, unlike \gls{LFO} that is ``naturally'' substituted by EnergyScope Pathway in the near-term.}
-\label{fig:Binding_learning}
-\end{figure}
-
+\newpage
 \section{Comparison with references}
@@
 Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition.
-This buffer is compensated by the end of the transition where the 50\% of the myopic transitions reach 2050 with 10 or more Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
+This buffer is compensated by the end of the transition where 50\% of the myopic transitions reach 2050 with 10 or more remaining Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
 These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% on average by 2050.
@@
 
-The transition pathways of the annual system cost also show the advantage of the \gls{RL}-based exploration (see Figure \ref{fig:System_cost_pathway}).
+The transition pathways of the annual system cost show another impact of the agent in the \gls{RL}-based exploration (see Figure \ref{fig:System_cost_pathway}).
 Via its actions and its cost-based reward, the agent can reach systems that are cheaper than any other solution obtained by the perfect foresight approach.
@@
 
+\newpage
 The analysis of the cumulative costs shows that the \gls{OPEX} is the main difference between myopic and perfect foresight transitions (see Figure \ref{fig:Opex_Capex_Salvage_comp}).
@@
 The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}.
-The assessment of the primary energy mix by the 2050 highlights that the differences of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
+The assessment of the primary energy mix by 2050 highlights that the difference of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
 In the majority of the cases, e-ammonia is more than two times more imported in the myopic transitions.
@@
 
-%In \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
+%In the \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
 The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt$_{\ce{CO2},\text{eq}}$ equivalent to 10 years at the current level of emissions.
-A secondary focus is also put on aiming at minimum transition cost.
-To know its progress through the transition, the agent had the information of the cumulative emissions and costs as well the share of renewable energy carriers in the primary energy mix and the system overall efficiency.
+A secondary focus is also put on aiming at a minimum transition cost.
+To know its progress through the transition, the agent has the information of the cumulative emissions and costs as well the share of renewable energy carriers in the primary energy mix and the system overall efficiency.
 
@@
 Where \gls{LFO} becomes less cost-competitive in the near-future, limiting the use of coal should be done at any cost.
-Then, fossil gas should be replaced by e-methane in the mid-term while limiting the overall emissions becomes more effective by the end of the transition.
+Then, fossil gas should be replaced by e-methane in the mid-term while putting a strict limit on the overall emissions becomes the most effective action by the end of the transition.
 The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances to succeed the transition.
@@
 
+Further analyses should assess more accurately the impact of uncertainties on the agent's capacity to succeed the myopic transition.
+These analyses would aim at identifying the parameters that should necessarily be in specific ranges to allow the agent to succeed.
+On the contrary, there could be ranges of values for which the chances of success would be much more limited.
+
 In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2}-budget target.

### 2024-04-25 chap_RobPol/chap_RobPol.tex
SUBMISSION

@@
 \label{sec:RobPol:PC_transition}
-As detailed in Section \ref{subsec:meth:PCA:transition}, we have decided to define the directions of variation, \ie the robustness metrics, based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol.
+The directions of variation, \ie the robustness metrics, are based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol (see Section \ref{subsec:meth:PCA:transition}).
 These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.
@@
 Highlighting the top-5 technologies for $\text{PC}_{y,1}$, $\text{PC}_{y,2}$ and $\text{PC}_{y,3}$, we observe general trends over the whole transition as well as tipping year where there is a clear trade-off between several technologies (see Figure \ref{fig:Top5_PC_year}).
+The top-5 technologies of $\text{PC}_{y,1}$ represent on average 99.4\% of the norm of their respective PC.
+For $\text{PC}_{y,2}$ and $\text{PC}_{y,3}$, this share decreases to 97.6\% and 90.9\%, respectively.
 As pointed out in Section \ref{subsec:meth:PCA:transition}, \gls{PCA} does not make any distinction between a vector of variation and its opposite.
@@
 
+\newpage
 Averaging the components of similar $\text{PC}_y$ allows constructing the $\text{PC}_{\text{transition}}$ (see Figure \ref{fig:Top5_PC_transition}).

### 2024-04-25 chap_atom_mol/chap_atom_mol.tex
RL chapter: analysis about import of electrofuels
SUBMISSION

@@
 
-In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimize a transition pathway with a small number of time stages, \ie two or three.
-Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Appendix \ref{app:UC_full}).
+\section*{Contributions}
+\label{sec:atom_mol:contributions}
+
+In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimise a transition pathway with a small number of time stages, \ie two or three.
+Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}).
 Accounting for uncertainties presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}, by the end of the transition, \ie 2050.
@@
 
-\section{Deterministic impact of integrating \gls{SMR} in 2040}
+\section[Deterministic impact of integrating SMR in 2040]{Deterministic impact of integrating \gls{SMR} in 2040}
 \label{sec:atom_mol:results_deter} 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_REF.pdf}
-\includegraphics[width=0.8\textwidth]{Elec_Tech_Cap_SMR.pdf}
+\includegraphics[width=\textwidth]{Elec_Tech_Cap_REF.pdf}
+\includegraphics[width=\textwidth]{Elec_Tech_Cap_SMR.pdf}
 \caption{As soon as available (\ie 2040), \acrfull{SMR} is deployed to their maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie gas and ammonia \gls{CCGT}).}
@@
 
+\newpage
 When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Elec_Layer_REF.pdf}
-\includegraphics[width=0.8\textwidth]{Elec_Layer_SMR.pdf}
+\includegraphics[width=\textwidth]{Elec_Layer_REF.pdf}
+\includegraphics[width=\textwidth]{Elec_Layer_SMR.pdf}
 \caption{The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
@@
 
+\newpage
 \subsection{System-level impacts}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Cum_total_cost_diff_REF.pdf}
-\includegraphics[width=0.8\textwidth]{System_cost_DIFF.pdf}
+\includegraphics[width=\textwidth]{Cum_total_cost_diff_REF.pdf}
+\includegraphics[width=\textwidth]{System_cost_DIFF.pdf}
 \caption{(Top) Cumulative transition cost and (down) annual system cost differences between the SMR and the REF cases.
@@
 
-Extending the use of naphtha-cracker and, to a lesser extent the postponing the deployment of PV, lead to higher \ce{CO2}-emissions at the beginning of the transition that are then compensated by the deployment of \gls{SMR} (\Cref{fig:GWP_per_sector_diff}).
+Extending the use of naphtha-cracker and, to a lesser extent, postponing the deployment of PV, lead to higher \ce{CO2}-emissions at the beginning of the transition that are then compensated by the deployment of \gls{SMR} (\Cref{fig:GWP_per_sector_diff}).
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{GWP_per_sector_diff.pdf}
+\includegraphics[width=\textwidth]{GWP_per_sector_diff.pdf}
 \caption{Emissions per energy sector differerence between the SMR and the REF cases.
@@
 \caption{(Top) Primary energy mix over the transition for the REF and SMR cases.
-(Down) Difference of the mix between the two cases after aggregation per category.
+(Bottom) Difference of the mix between the two cases after aggregation per category.
 In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
@@
 
+\newpage
 \subsection{Non-power sectors}
@@
 
-Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
-
-Then, as the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index.
-The model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
-In practice, the interest rate would vary depending on the technology investment risk.
-This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.
-
-Finally, the cost of purchasing fossil fuels is also a key parameter in terms of variation of the total transition cost.
-However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their impact smaller than their renewable alternatives.
-
 \begin{table}[htbp!]
@@
 \end{table}
+
+Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
+
+As the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index.
+The model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
+In practice, the interest rate would vary depending on the technology investment risk.
+This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.
+
+The cost of purchasing fossil fuels is also a key parameter in terms of variation of the total transition cost.
+However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their impact smaller than their renewable alternatives.
 
@@
 It is an indicator of statistical dispersion around the median, $Q2$ or $P_{50}$.} of about 50\,TWh.
-In some extreme cases, e-ammonia is the most imported molecules, \ie up to 216\,TWh or 58\% of the total primary mix.
+In some extreme cases, e-ammonia is the most imported molecules, \ie up to 167\,TWh or 45\% of the total primary mix.
 
@@
 
-As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}).
-This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production.
-Then, considering the smaller-impact parameters, we notice that \gls{SMR} plays a non-negligible role.
-Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives.
-This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}.
-Highly available local biomass also leads to smaller import of e-methane to supply bio-hydrolysis and produce methane-equivalent gas.
-Finally, and surprisingly, costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively.
-In other words, by 2050, more expensive electrofuels result in more e-methane import.
-On the contrary, cheaper electrofuels result in higher import of fossil methane.
-Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}.
-Subject to the \ce{CO2}-budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity.
-First running on fossil gas, these \gls{CHP} consume more e-methane by 2050.
-On the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
-This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
-In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
-Consequently, there is smaller investment in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers.
-Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2}-budget allows it.
-About the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system.
-If \gls{NG} is more expensive, the system will import less of it.
-Subsequently, the investments in methane-\gls{CHP} and boilers are more limited.
-This ends up in smaller need for e-methane by 2050.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-methane in 2050.
-Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
-The grey dashed line gives the value of the output of interest in the REF case.
-}
-\label{fig:results_uq_samples_gas}
-\end{figure}
-
 In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector (Figure \ref{fig:results_uq_samples_H2}).
@@
 \label{fig:results_uq_samples_H2}
+\end{figure}
+
+As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}).
+This parameter directly dictates the demand of industrial high-temperature heat for which industrial gas \gls{CHP}, and industrial gas boilers to a lower extent, represent, on average over all the samples, respectively 25.6\% and 6.1\% of the total production.
+Then, considering the smaller-impact parameters, we notice that \gls{SMR} plays a non-negligible role.
+Indeed, if deployed, \gls{SMR} produces abundant low-emitting electricity for industrial electric heaters that substitute, even completely in some cases, gas alternatives.
+This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}.
+Highly available local biomass also leads to smaller import of e-methane to supply bio-hydrolysis and produce methane-equivalent gas.
+Finally, and surprisingly, costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively.
+In other words, by 2050, more expensive electrofuels result in more e-methane import.
+On the contrary, cheaper electrofuels result in higher import of fossil methane.
+Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}.
+Subject to the \ce{CO2}-budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity.
+First running on fossil gas, these \gls{CHP} consume more e-methane by 2050.
+On the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
+This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
+In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
+Consequently, there are smaller investments in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers.
+Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2}-budget allows it.
+About the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system.
+If \gls{NG} is more expensive, the system will import less of it.
+Subsequently, the investments in methane-\gls{CHP} and boilers are more limited.
+This ends up in smaller need for e-methane by 2050.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
+\caption{Trend lines of the key parameters on the import of e-methane in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_gas}
 \end{figure}
@@
 
+\newpage
 \subsection{Local renewables}

### 2024-04-25 chap_case_study/chap_case_study.tex
SUBMISSION

@@
 \centering
-\includegraphics[width=0.8\textwidth]{NED_hist.pdf}
+\includegraphics[width=\textwidth]{NED_hist.pdf}
 \caption{Historical data for the non-energy use in Belgium \cite{statbel_NED_2019}.
@@
 \centering
-\includegraphics[width=0.49\textwidth]{EUD_sec.pdf}
-\includegraphics[width=0.49\textwidth]{EUD_cat.pdf}
-\includegraphics[width=0.49\textwidth]{EUD_mob.pdf}
+\includegraphics[width=0.43\textwidth]{EUD_sec.pdf}
+\includegraphics[width=0.43\textwidth]{EUD_cat.pdf}
+\includegraphics[width=0.43\textwidth]{EUD_mob.pdf}
 \caption{EnergyScope splits the whole-energy system end-use demands (EUD) into two sets: (non)-energy and transport-related.
@@
 \label{sec:cs:resources}
-To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$, their global warming potential, $\emph{gwp}_{\mathrm{op}}$, as well as their
+To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$ (see \Cref{fig:cs_resources_cost}), their global warming potential, $\emph{gwp}_{\mathrm{op}}$, as well as their
 availability, as detailed by \citet{limpens2024pathway}.
+Regarding the cost of ``renewable electrofuels'', they are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis \cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
 
-First, the evolution of the respective costs of purchasing is presented (see \Cref{fig:cs_resources_cost}).
-Regarding ``renewable electrofuels'', the costs are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis \cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.43\textwidth]{Res_ren.pdf}
+\includegraphics[width=0.43\textwidth]{Res_foss.pdf}
+\includegraphics[width=0.43\textwidth]{Res_others.pdf}
+\caption{Cost of purchasing the different resources.
+Besides the free local renewables (\ie sun, wind and hydro) limited by technical potentials, EnergyScope accounts for renewable energy carriers and their respective fossil counterparts (top-left and top-right).
+These fuels can be imported from abroad without limitation on their availability.
+Other carriers are limited either by their local potentials (\ie biomass and waste) or other considerations like the power grid interconnections or the capacity of nuclear power plants.}
+\label{fig:cs_resources_cost}
+\end{figure}
 
-Then, besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
+Besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
 The limitation in terms of availability can be direct or indirect.
@@
 Besides these, every other resource has its specific \gls{GWP} like coal ($\emph{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), natural gas ($\emph{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh) or the fossil-based molecules equivalent to the electrofuels (\eg $\emph{gwp}_{\mathrm{op,ammonia}}=0.46$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\emph{gwp}_{\mathrm{op,methanol}}=0.41$\,kt$_{\ce{CO2},\text{eq}}$/GWh).
-
-\begin{figure}[htbp!]
-\centering
-
-\includegraphics[width=0.49\textwidth]{Res_ren.pdf}
-\includegraphics[width=0.49\textwidth]{Res_foss.pdf}
-\includegraphics[width=0.49\textwidth]{Res_others.pdf}
-\caption{Cost of purchasing the different resources.
-Besides the free local renewables (\ie sun, wind and hydro) limited by technical potentials, EnergyScope accounts for renewable energy carriers and their respective fossil counterparts (top-left and top-right graphs).
-These fuels can be imported from abroad without limitation on their availability.
-Other carriers are limited either by their local potentials (\ie biomass and waste) or other considerations like the power grid interconnections or the capacity of nuclear power plants.}
-\label{fig:cs_resources_cost}
-\end{figure}
 
@@
 In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined range of uncertainties for the model parameters.
-\Cref{tab:UC_short} gives the uncertainty ranges of some key parameters.
 As the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III, based on the work of \citet{Moret2017PhDThesis}.
@@
 This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
-A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_short}.
+A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.
 As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040.

### 2024-04-25 chap_intro_ccl/conclusion.tex
SUBMISSION

@@
 The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2}-budget without prescribed trajectory for the reduction of the emissions.
-We have pursued this objective on the case study of Belgium with a focus on the role that renewable electrofuels such as e-ammonia, e-methane, e-methanol and e-hydrogen, could play in this densely-populated and highly industrialised country with limited local renewable potentials.
-To have 66\% chance of limiting warming to 1.5°C \cite{IPCC_CO2_budget} and using the ``grandfathering'' sharing principle of the global budget \cite{knight2013grandfathering}, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
-This budget represents 10 years of the current level of emissions.
+We have studied the case study of Belgium with a focus on the role that renewable electrofuels (e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely-populated and highly industrialised country with limited local renewable potentials.
+To have 66\% chance of limiting warming to 1.5°C and using the ``grandfathering'' sharing principle of the global budget, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
 
-To optimise such a system, we have developed EnergyScope Pathway that minimises the total transition cost of the system from 2020 to 2050 and implemented the myopic approach.
-This approach considers a sequence of 10-year time windows with an overlap of 5 years rather than one global optimisation over the 30 years of the transition.
-This allows assessing the realism of the decision making process where knowledge about the future is progressively unfold rather than assumed to be fully known from the beginning of the transition as in the perfect foresight (PF) approach.
+To optimise such a system, we have developed EnergyScope Pathway: a whole-energy system that minimises the total transition cost of the system from 2020 to 2050.
+A myopic approach has been implemented, considering a sequence of 10-year time windows with an overlap of 5 years rather than one global optimisation over the 30 years of the transition.
+It allows assessing the realism of the decision making process where knowledge about the future is progressively unfold, rather than assumed to be fully known from the beginning of the transition as in the perfect foresight (PF) approach.
 
-The future of a national whole-energy system is subject to a plethora of uncertainties such as the cost of purchasing resources, their availabilities, the \acrfull{CAPEX} of technologies or the level of demands.
+The future of a national whole-energy system is subject to a plethora of uncertainties.
+Main examples fo uncertainties are the cost of purchasing resources, their availabilities, the \acrfull{CAPEX} of technologies and the level of demands.
 In this thesis, we have used the uncertainty characterisation approach developed by \citet{moret2016strategic} to provide the range of variation of the uncertain parameters.
-To propagate and quantify the impact of these uncertainties on the outputs of the model, we have used the \acrfull{PCE} method has been used as implemented in the Coppitters' RHEIA framework \cite{coppittersthesis}.
+To propagate uncertainties in the model and quantify their impact on the outputs of interest, we have used the \acrfull{PCE} method as implemented in the Coppitters' RHEIA framework \cite{coppittersthesis}.
 
@@
 In this framework, an agent takes actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
-Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050.
-Repeating the whole transition with different sequences of actions-states allow the agent coming up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
+Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050 where a transition was considered as successful when respecting the \ce{CO2}-budget.
+Repeating the whole transition with different sequences of actions-states allowed the agent coming up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
 
@@
 % THE RESULTS WE GOT (WHAT?)
-Given the \ce{CO2}-budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  the cost-optimum transition of Belgium should be to first opt for more efficiency to reduce by one third the primary energy consumption electrofuels.
+Given the \ce{CO2}-budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  aiming at a cost-optimum transition, Belgium should first opt for more efficiency to reduce by one third the primary energy consumption.
 Then comes the full deployment of local solar and wind energies that would represent about 30\% of the mix by 2050.
-To fill the gap left by limited potential of local renewables, the electrofuels are found to be the third pillar of the Belgian transition.
-47\% of the variation of the total transition cost are due to the uncertainty on the cost of purchasing these energy carriers, before the industrial end-use demands (23\%), the interest rate (12\%) and the cost of purchasing fossil fuels (6\%).
+Due to the limited potential of local renewables, electrofuels are found to be the third pillar of the Belgian transition.
+The uncertainty on the total cost of the transition is mainly due to the uncertainty on the cost of purchasing electrofuels (47\%), on the industrial end-use demand (23\%), on the interest rate (12\%) and on the price of fossil fuels (6\%).
 
-E-methanol is found to be the key-molecule to import in the mid-term to defossilise the non-energy demand and substitute naphtha-crackers to produce on average 43\,TWh \acrfull{HVC}.
+E-methanol is the key molecule to import in the mid-term to defossilise the non-energy demand of \acrfull{HVC}, about 43\,TWh/year, by substituting naphtha-crackers by \acrfull{MTO}.
 The uncertainty on this demand is the most impacting parameter (80\%) on the variation of the quantity of imported e-methanol.
 
-In the longer-term, importing e-ammonia becomes crucial to supply its own non-energy demand ($\sim$ 10\,TWh) and, before all, \acrfull{CCGT} to provide flexibility in the production of electricity in competition with nuclear \acrfull{SMR}.
+In the longer-term, on top of the non-energy demand to further produce fertilizers (about 10\,TWh), importing e-ammonia becomes crucial to supply \acrfull{CCGT} and to provide flexibility in the production of electricity in competition with nuclear \acrfull{SMR}.
 For this reason, where the variation of the quantity of imported e-ammonia is mostly driven by the uncertainty on its cost of purchasing (50\%), the uncertainty on the availability of nuclear \gls{SMR} by 2040 is the second most impacting parameter (20\%).
+To a smaller extent, e-methane helps defossilising the industrial boilers and \acrfull{CHP}.
+E-methane imports is mostly impacted by industrial demands (45\%).
 
-To a smaller extent, e-methane help defossilising the industrial boilers and \acrfull{CHP}. E-methane imports is mostly impacted by industrial demands (45\%).
-
-By 2050, e-hydrogen is only imported from abroad in the wide majority of the transition pathways under uncertainties generated during this work.
+By 2050, in 88\%of the transitions, e-hydrogen is only imported from abroad rather than locally produced.
 This molecule supports the defossilisation of the freight-transport by trucks.
@@
 
-Through the Reinforcement Learning-based exploration of myopic transitions, we pointed out that near-term actions were mandatory to hope succeeding this ambitious transition.
-Beyond 2030, if the right actions have not been taken, the cumulative emissions are likely to overshoot the \ce{CO2}-budget.
-Besides limiting the use of coal at any cost, limiting the consumption of fossil gas was needed at the early stages.
-Once the fuel switch has been operated from fossil gas to e-methane, limiting the overall emissions is the most effective action to stick to the direction towards sustainability and encourage the import of renewable energy carriers.
-Regarding the information on which the \gls{RL}-agent determines its actions, the system overall efficiency provides less informative insight than the share of renewable energy carriers in the primary energy mix.
-The latter shows intermediate milestones to reach, 62\%-share by 2030, to have higher chance to succeed the transition by 2050.
-Below this threshold, these chances of success are much more limited, the so-called no-go zones.
-Finally, due to limited foresight into the future and its uncertainties, successful myopic transitions rely more on importing renewable electrofuels than the perfect foresight approach.
-This is even more the case for e-ammonia that is in general twice more imported by 2050 in myopic than PF transitions.
+The Reinforcement Learning-based exploration of myopic transitions points out that near-term actions are mandatory to hope succeeding this ambitious transition.
+Beyond 2030, if the right actions are not taken, the cumulative emissions are likely to overshoot the \ce{CO2}-budget.
+On top of strongly reducing the use of coal, fossil gas consumption should also be limited within 5 to 10 years.
+Once the fuel switch is operated from fossil gas to e-methane, putting a strong limit on the overall emissions is the most effective action to stick to the direction towards sustainability and encourage the import of renewable energy carriers.
+In the \gls{RL} framework, the agent's actions are mainly driven by the share of renewable energy carriers in the primary energy mix rather than the overall system efficiency.
+This share of renewable energy carriers shows intermediate milestones to reach, 62\%-share by 2030, to have higher chance to succeed the transition by 2050.
+Below this threshold, we find the so-called ``no-go zone'' where the chances of success are much lower.
+Due to limited foresight into the future and uncertainties like the possibility to install \gls{SMR} by 2040, successful myopic transitions rely more on importing renewable electrofuels than the perfect foresight approach.
+An illustrative example is the case of e-ammonia, for which imports by 2050 are in general twice larger in the myopic transition.
+When compared to agent-based myopic transitions, this work also highlighted the major added-value of the perfect foresight approach: the robustness towards uncertainties.
+Given its assumed full knowledge on the future, PF finds a solution to respect the \ce{CO2}-budget whatever the values taken by the uncertain parameters.
+Moreover, the trajectories to decrease the \ce{CO2} emissions are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
 
-In the case of Belgium, the uncertainty on the maximum installed capacity of solar \gls{PV}, which is supported by the electrification of the industrial and decentralised heating sectors, drives 57\% of the variance of the technological mix through the transition.
-This affects the entire transition.
-On the contrary, other key contributions to this variance are more focused on tipping years.
-For instance, the low-temperature heating sectors shift from oil and gas decentralised boilers to electric \acrfull{HP} and \acrfull{DHN} from 2025 to 2035.
-In the non-energy sector, the shift between naphtha-crackers and \acrfull{MTO} takes place in 2025.
+The uncertainty on the maximum installed capacity of solar \gls{PV}, which is supported by the electrification of the industrial and decentralised heating sectors, drives 57\% of the variance of the entire technological mix through the whole transition.
+On the contrary, other key contributions to this variance are more focused on tipping years such as the low-temperature heating sectors shift from oil and gas decentralised boilers to electric Heat Pumps (HP) and Decentralised Heating Networks (DHN) from 2025 to 2035.
+In the non-energy sector, the shift from naphtha-crackers to \acrfull{MTO} takes place in 2025.
 
-Finally, testing roadmaps resulting from perfect foresight optimisations in myopic conditions has shown that cases relying or not on 6\,GW of \gls{SMR} by 2040 onward provide the same level of robustness in terms of additional capacities to install.
-However, despite representing an additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \gls{CHP} provides significantly better level of robustness (between 33\% and 49\%).\\
+Finally, testing technological roadmaps in myopic conditions has shown that cases relying or not on 6\,GW of \gls{SMR} by 2040 onward provide the same level of robustness in terms of additional capacities to install.
+Despite representing an additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \gls{CHP} provides significantly better level of robustness (between 33\% and 49\%).\\
 
@@
 Besides a wider electrification of the system and the integration of more efficient technologies, a robust policy consists in investing in electrofuels.
-Chemically similar to their fossil-based equivalents, they can benefit from the current infrastructures and provide flexibility to the mismatch between the production of renewable electricity and its consumption.
-They would be imported from countries ``in excess'' of renewable electricity from wind or solar such as Australia \cite{Australia_efuels}, Chile \cite{Chile_efuels} or Spain \cite{Spain_efuels} to provide seasonal storage of electricity.
-Already stated in the analysis of the snapshot model of \citet{rixhon2021role}, this work confirms through the optimisation of the whole transition pathway that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
-On the road to a robust whole-energy system in Belgium, the policy makers, the industries, and academia should invest in projects to produce and use these fuels or developing the exchange networks with neighbouring countries.
+Chemically similar to their fossil-based equivalents, they can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.
+Electrofuels should be imported from countries where the potential wind and solar electricity production largely exceeds the local consumption (Australia \cite{Australia_efuels}, Chile \cite{Chile_efuels} or Spain \cite{Spain_efuels}).
+The optimisation of the whole transition pathway indicates that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
+On the road to a robust whole-energy system in Belgium, the policy makers, the industries and academia should invest in projects to produce and use these fuels or developing the exchange networks with foreign countries.
 This would provide significant reduction in the variation of the total transition cost and robustness towards uncertainties progressively unfolding with the future.
 
-About the use of the electrofuels, the two molecules imported in largest quantity, e-methanol and e-ammonia, would aim at defossilising the non-energy and providing flexibility in the production of electricity, respectively.
+About the use of the electrofuels, there are two molecules imported in larger quantities: e-methanol and e-ammonia.
+E-ammonia would aim at defossilising the production of \acrfull{HVC} in the non-energy sector and e-ammonia would dominantly provide flexibility in the production of electricity, respectively.
 The main use of e-hydrogen is in the freight transport via trucks and e-methane would supply the industrial sectors through \gls{CHP} and boilers.
 
-Also brought up as another ``unicorn'' solution to the problem of the energy transition, nuclear \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but would still support the deployment of local \acrfull{VRES} such as solar and wind.
+Often brought up as a ``unicorn'' solution to the energy transition, nuclear \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but not against the deployment of local \acrfull{VRES} such as solar and wind.
 This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2}-budget of the transition.
 
-On a methodological point of view, this thesis developed novel approaches to explore myopic pathway transitions, highlight intermediate milestones and no-go zones towards sustainability and assess the robustness of transition technological roadmaps.
+Regarding methodologies, this thesis developed novel approaches to explore myopic pathway transitions, highlight intermediate milestones and no-go zones towards sustainability and assess the robustness of transition technological roadmaps.
 Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them on their specific case study.
@@
 
-As the research in the field of the energy transition is a never-ending story, we conclude with perspectives that could benefit from the methods and the results developed in this thesis.
+We conclude with perspectives that could benefit from the methods and the results developed in this thesis.
 Future studies could focus on different scenario analyses, the development of EnergyScope Pathway or the approaches applied to it such as \gls{RL}.
@@
 The Belgian transition pathways explored in this thesis massively relied on the early deployment of local \gls{VRES} and import of electrofuels.
-In some of these pathways, in a 5 to 10 year time window, the current capacities of \gls{PV} and wind turbines are multiplied by 10 and 6, respectively.
-This is much faster than the multiplication by 2 and 2.5 of PV and wind capacities, respectively, from 2024 and 2034 projected by Elia, Belgium Transmission System Operator (TSO),  in their adequacy analysis \cite{Elia_2024_2034}.
-On their side, the amount of imported electrofuels could be 10 to 20 times more than the potential that Belgium could benefit from exporting countries, given the already signed agreements \cite{lefebvre2022electrofuel}.
-As the objective of this work was to explore as much as possible transition pathways, we have decided not to put further constraints on these energy carriers.
-One future work would be to constrain the potential of local renewables to a more progressive deployment or the availability of the electrofuels.
-This could provide different pathways where electrofuels would be produced more locally using carbon-capture technologies.
-This could also show the limit of the Belgian energy to meet such an ambitious \ce{CO2}-budget in its energy transition.
+In some pathways, the current capacities of \gls{PV} and wind turbines are multiplied by 10 and 6 over a 5 to 10 year time window.
+This is much faster than the multiplication by 2 and 2.5 of PV and wind capacities from 2024 and 2034 projected by Elia, Belgium Transmission System Operator (TSO), in their adequacy analysis \cite{Elia_2024_2034}.
+The necessary amount of imported electrofuels could be 10 to 20 times higher than the potential that Belgium could benefit from considering the existing agreements with exporting countries \cite{lefebvre2022electrofuel}.
+A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
+This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
+This could also show the limits of the Belgian energy system to meet such an ambitious \ce{CO2}-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
+One could then question the level of demands and address the question of the sufficiency.
+For a growing part of the scientific community, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
+A study could consist in analysing the level of demands, within their respective sector, we should reach to stay within the \ce{CO2}-budget and limit the installation of solar PV, wind turbines, nuclear energy or the import of electrofuels.
 
@@
 
-Rather than limiting the availability of renewable resources, one could question the level of demands and address the question of the sufficiency.
-For a growing part of the scientific community \cite{o2018good}, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
-A study could consist in analysing the level of demands, within their respective sector, we should reach to meet the \ce{CO2}-budget and limit the installation of solar PV, wind turbines or nuclear energy or the import of electrofuels.
-
 On the methodological side, several perspectives could be considered about the whole-energy system model or the implementation of \gls{RL}.
-
-The cumulative \acrfull{GHG} emissions of the transitionhas been constrained based on the use of the resources.
-However, in a system reducing drastically its emissions, indirect emissions such as the ones related to the construction and dismantling of assets, would have a much higher share of the total emissions.
-Depending on the level of emission reduction, these indirect emissions could be one to three times higher than the direct emissions \cite{blanco2020life}.
+The cumulative \acrfull{GHG} emissions of the transition has been constrained based on the use of the resources.
+However, in a system reducing drastically its emissions, indirect emissions such as the ones related to the construction and dismantling of assets would have a much higher share of the total emissions.
+Depending on the level of emission reduction, these indirect emissions could be one to three times higher than the direct emissions.
 To accurately integrate these emissions in the model and account for them in the \ce{CO2}-budget, \gls{LCA} can compute these indirect emissions.
@@
 Even though it represents a growing part of the literature, sufficiency is rarely brought up by decision-makers as a solution.
-At the same time, whole-energy system models and related methodologies do not claim to predict the future but rather aims at quantifying options under several assumptions.
-In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into discussions and decisions that should account for more interdisciplinarity.
+Whole-energy system models and related methodologies do not claim to predict the future but rather aims at quantifying options under several assumptions.
+In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into the intrinsically political and interdisciplinary discussions and decisions that must be made in the coming years.

### 2024-04-25 chap_intro_ccl/intro.tex
SUBMISSION

@@
 The rate at which climate changes today is mostly due to the concentration
-of anthropogenic \gls{GHG} in the environment \cite{IPCC_CO2_budget}.
+of anthropogenic \gls{GHG} in the atmosphere \cite{IPCC_CO2_budget}.
 Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
 Energy consumption, mainly from fossil fuels, is the main contributor: it is responsible for 75\% of \gls{GHG} emissions, the remaining 25\% originating from agriculture and industrial processes \cite{ourworldindata_CO2_world}.
-The Kaya identity is an economic formulation stating that the total emission level of the \gls{GHG} can be expressed as the product of four factors: human population, \gls{GDP} per capita, energy intensity (per unit of GDP), and carbon intensity (emissions per unit of energy consumed) \cite{kaya1997environment}.From an energy point of view, this identity can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
+The Kaya identity is an economic formulation stating that the total emission level of the \gls{GHG} can be expressed as the product of four factors: human population, \gls{GDP} per capita, energy intensity (per unit of GDP), and carbon intensity (emissions per unit of energy consumed) \cite{kaya1997environment}.
+From an energy point of view, this identity can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
 \gls{EUD} is the energy service required by the the final consumer.
-In Equation (\ref{eq:equality_GHG}), the first term represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
+In Equation (\ref{eq:equality_GHG}), the first factor on the right hand side represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
 
@@
 Explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, it is defined as ``\emph{a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people}'' \cite{lage2023citizens}.
-Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving sociology, psychology and political science, as it is outside the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
-One of the objectives of this manuscript is to support interdisciplinary projects in the assessment of sufficiency policy.
+Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving economy, sociology, psychology and political science, and is outside the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
+One of the objectives of this work is to support interdisciplinary projects in the assessment of sufficiency policy.
 \\
@@
 To ensure the energy supply of a growing and more demanding population in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
-Similarly to the RED III, the transformations identified by \citet{iea2020world} focus on technical aspects, disregarding the sufficiency.
+Similarly to the RED III, the transformations identified by the \citet{iea2020world} focus on technical aspects, disregarding the sufficiency.
 The technical transformations that can be brought to the energy system concern both the primary energy sources and the technologies used to convert more efficiently these resources into the \gls{EUD} \cite{iea2020world,luderer2018residual}.
@@
 \centering
-\includegraphics[width=\textwidth]{IEA_WEO_2019.png}
-\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario of \citet{iea2020world}.}
+\includegraphics[width=\textwidth]{IEA_WEO_2019.pdf}
+\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario of the \citet{iea2020world}.}
 \label{fig:intro:IEA_WEO_2019}
@@
 \gls{VRES} like wind and solar are the keystones to defossilise the energy system but they face challenges related to their intermittency and space disparity.
-To address this issue,  limitations like range, power or  costs make \gls{DC} lines or batteries not the most relevant solutions.
-The transport and long-term storage of the renewable electricity produced in excess should be optimised.
-This is why, to reach the objective of sustainable development, more renewables is associated with ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}).
+With a growing share of \gls{VRES}, sector coupling and electrification, with electrical heat pumps or Battery Electric Vehicles (BEV), are essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them cost-effectively \cite{brown2018response, limpensECOS2021}.
 
-The fuel switch consists in replacing traditional fossil fuels with new energy carriers.
+However, direct electrification is not foreseen to be cost-competitive for some sectors such as marine, aviation and heavy-duty transport \cite{horvath2018techno, brynolf2018,IEA2021}.
+Moreover, transport and long-term storage of the renewable electricity is not optimal with electricity-based solutions, \eg {DC} lines or batteries.
+This is why, to reach the objective of sustainable development, after increasing the efficiency of the system, more renewables is associated with ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}).
+
+The fuel switch consists in replacing fossil fuels with new energy carriers.
 Among them, \textit{electrofuels} are promising solutions to tackle the challenges related to renewable electricity \cite{rozzi2020}.
@@
 
-Electrofuels offer three main advantages: infrastructure compatibility, storage and capacity to link sectors (\ie from electricity to mobility, heat, or industry).
-Development on electrofuels aims at getting them more and more compatible with existing and mature technologies \cite{Ahlgren2012}.
+Electrofuels offer three main advantages: infrastructure compatibility, capacity to link sectors (\ie from electricity to mobility, heat, or industry) and storagr.
+Given their similar physicochemical properties, electrofuels could substitute fossil in technologies that are already in place today.
 An example is ammonia-hydrogen blends burned in spark ignition engines \cite{lhuillier2020experimental} or \gls{CHP} applications \cite{pochet202022}.
-Mobility applications such as marine, aviation and heavy-duty transport are hard to electrify and require high-density energy carriers \cite{horvath2018techno, brynolf2018}.
+
+Electrofuels have also the ability to couple energy and non-energy sectors \cite{Stancin2020}.
+For instance, electricity from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser, coupling the power and agro-industrial sectors \cite{verleysen2020can}.
 
@@
 On the contrary, electrofuels are promising solutions for high capacity (from 100 GWh) and long-term storage of energy \cite{child2018role, dias2020energy} (Figure \ref{fig:intro:Storage_electrofuels}).
-In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent a better usage of \ce{CO2} than \gls{CCS} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
-gas networks present much more storage potential than electrical networks: 50 times more in Germany and 300 times more in France \cite{Rosa2017}.
+In their analysis of the German transport sector in 2050, \citet{millinger2021electrofuels} highlighted that producing electrofuels can represent an efficient usage of the ambient \ce{CO2} to supply hydrocarbon fuels while limiting the curtailment of \gls{VRES}.
+Moreover, gas networks present much more storage potential than electrical networks: 50 times more in Germany and 300 times more in France \cite{Rosa2017}.
 
@@
 
-Finally, with a growing share of \gls{VRES}, sector coupling is essential to absorb the surplus of electricity from these intermittent production means \cite{robinius2017linking} and integrate them more cost-effectively \cite{brown2018response, limpensECOS2021}.
-Besides direct electrification of other sectors with electrical heat pumps, Battery Electric Vehicles (BEV),  converting power to hydrogen and methane is advantageous at high shares of renewables for the optimisation of the European energy system \cite{brown2018synergies}.
-Electrofuels have also the ability to couple energy and non-energy sectors \cite{Stancin2020}.
-For instance, electricity from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser, coupling the power and agro-industrial sectors \cite{verleysen2020can}.
-
+\newpage
 Even if the share of electricity will increase in the energy system through the electrification of the end-use demand, gaseous and liquid fuels will keep on being big players during, and after, the energy transition \cite{Ahlgren2012}.
@@
 
-To harvest the maximum potential of synthetic energy carriers in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a whole-energy system which is multi-sector and account for multiple energy carriers \cite{contino2020whole}.
+To harvest the maximum potential of synthetic fuels in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a whole-energy system which is multi-sector and account for multiple energy carriers \cite{contino2020whole}.
 To reach this goal, an energy system optimisation model (ESOM) can optimise the design and the operation of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
@@
 In the field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most ESOMs use a deterministic approach (75\% of the 134 reviewed studies).
-However, the model structures are inherently uncertain as well as their numerous composing parameters, especially when it comes to define an energy transition strategy for a large-scale system such as a country.
+However, the models and their numerous composing parameters are inherently uncertain especially when it comes to define an energy transition strategy for a large-scale system such as a country.
 Given the lifetime of energy conversion technologies, such a strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
@@
 
-On top of dealing with these uncertainties, the reality of decision-makers includes a limited foresight into the future \cite{poncelet2016myopic}.
+On top of dealing with these uncertainties, the reality of decision-makers leads to a limited foresight into the future \cite{poncelet2016myopic}.
 In a perfect foresight approach, decision-makers would be able to, from now on, see the ``finish-line of the transition'' in 2050 and accordingly make the planning decisions once and for all.
 On the contrary, they uncover the realisation of these uncertainties step-by-step, in a ``myopic'' way, and progressively act on them to, hopefully, meet the set target to reduce the anthropogenic \gls{GHG}.
-In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there was a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
-On top of the ``what to do?'', this \gls{RL}-based framework aims at helping the policymakers to answer the  question ``how to do it?''.
+In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there is a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
+On top of the ``what to do?'', this framework should aim at helping the policymakers to answer the  question ``how to do it?''.
 
@@
 This thesis aims at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
-Rather than trying, in vain, to answer the question ``What could possibly happen in the future?'', this work rather addresses the ``What could or should we do to make the future possible?''.
+Rather than trying to answer the question ``What could possibly happen in the future?'', this work addresses the ``What could or should we do to make the future possible?''.
 In that spirit, the research questions are described as follows:
@@
 \item What can be the role of electrofuels in the transition pathways of a whole-energy system considering uncertainties,  a limited foresight, and a given \ce{CO2}-budget?
-What are the key uncertainties driving their import?
+What are the key uncertainties driving their local production or import?
 \item How to explore the multiple possible pathways through the optimisation of a sequence of actions to support an energy transition?
@@
 To explore the different pathway trajectories in this myopic optimisation process, we have applied the \acrfull{RL} method.
-In this branch of machine learning, an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie sequence of actions to take to support the transition.
+In this branch of machine learning, an ``agent'' is trained through its interactions with its environment, EnergyScope Pathway, to optimise its policy, \ie the sequence of actions to take to support the transition.
 
 Finally, in the objective to assess the sensibility to uncertainties of different transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
-This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a frame.
+This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a reference frame.
 Roadmaps resulting from perfect foresight optimisation have been tested in a myopic and uncertain pathways.
@@
 
-All this work has been applied to the case of Belgium, a densely populated country with a limited amount of renewable energies, which represents roughly one third of the forecast energy demand \cite{Limpens2020}.
+All this work has been applied to the case of Belgium, a densely populated country with a limited potential of renewable energies, which represents roughly half of the forecast energy demand \cite{Limpens2020}.
 This makes it a challenging case to go from a highly fossil-dominated system in 2020 (73\% of the primary energy mix \cite{spf_economy_2022}) to carbon-neutrality by 2050.\\
@@
 Chapter \ref{chap:case_study} presents the case study of this work: Belgium and its energy transition.
-Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the data-related \gls{SMR} as an option to produce nuclear-based electricity in the future.
+Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the data related to \gls{SMR} as an option to produce nuclear-based electricity in the future.
 
@@
 
-Finally, we draw the main conclusions of this work: near-term actions are needed to respect ambitious \ce{CO2}-budget and investing into electrofuels provide robustness to uncertainties and the limited foresight into the future.
-Perspectives are suggested for future research on uses and further developments of the methodological tools.
+Finally, we draw the main conclusions of this work: near-term actions are needed to respect ambitious \ce{CO2}-budget and investing into electrofuels provides robustness to uncertainties and the limited foresight into the future.
+Perspectives are suggested for future research on further developments and uses of the methodological tools.

### 2024-04-25 chap_methodo/chap_methodo.tex
SUBMISSION

@@
 Assessing the robustness of a whole-energy system transition pathway calls a variety of methodological tools.
-First and foremost, such an extensive system needs to be represented, \ie modelled, to further be optimized.
+First and foremost, such an extensive system needs to be represented, \ie modelled, to further be optimised.
 This model requires some characteristics to capture the peculiarities of this system such as the intermittency of \gls{VRES}, the coupling between the different energy and non-energy sectors and a pathway vision to pave the way from where we are to where we want to go.
@@
 The latter allows assessing the impact that such uncertainties can have on the output of the model.
-Finally, meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
+Finally, meeting the environmental objectives while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
 This work encompasses the optimisation of the policy,\ie set of actions to take along the transition with a specific methodology to assess its robustness.
@@
 \label{sec:meth:contributions}
-The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimize the behaviour of an artificial agent interacting with its environment, \ie the the whole-energy system through its transition.
-Given the uncertainties and potential shocks of the future, this approach allows the agent to play the transition in a sequential, \ie myopic, way and optimize the choice and the timing of its actions.
+The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimise the behaviour of an artificial agent interacting with its environment, \ie the the whole-energy system through its transition.
+Given the uncertainties and potential shocks of the future, this approach allows the agent to play the transition in a sequential, \ie myopic, way and optimise the choice and the timing of its actions.
 This optimization is done through trial and error where the agent repeats the transition with a new set of uncertainties.
@@
 Then, to support this step-by-step transition with limited foresight in the future, we have extended the EnergyScope Pathway model \cite{limpens2021generating}.
-Originally developed to optimize the transition in one global optimisation up to 2050, \ie perfect foresight, part of this thesis consisted in making this model able to optimize the same transition but with sequential more limited time windows, \ie myopic approach.
+Originally developed to optimise the transition in one global optimisation up to 2050, \ie perfect foresight, part of this thesis consisted in making this model able to optimise the same transition but with sequential more limited time windows, \ie myopic approach.
 Besides its shorter computational time, this approach is more representative an actual decision process with limited foresight in the future \cite{babrowski2014reducing}.
@@
 
-\section{Whole-energy system transition model optimisation: EnergyScope Pathway}
+\section[EnergyScope Pathway]{Whole-energy system transition model optimisation: EnergyScope Pathway}
 \label{sec:meth:ES}
@@
 \label{subsec:meth:PF}
+
+The whole-energy system model developed in this work originates from the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core}) of EnergyScope---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050 \cite{limpens2024pathway}.
+Each representative year is represented via the the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
+Then, to draw a consistent pathway between these years, additional constraints aim at linking them, \eg limiting the modal shifts within some energy sectors.
 
@@
 
-The whole-energy system model developed in this work originates from the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core}) of EnergyScope---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050 \cite{limpens2024pathway}.
-Each representative year is represented via the the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
-Then, to draw a consistent pathway between these years, additional constraints aim at linking them, \eg limiting the modal shifts within some energy sectors.
-
 The objective function of the pathway model, \ie the total transition cost $\textbf{C\textsubscript{tot,trans}}$, is computed as the sum of the total \gls{CAPEX}, \textbf{C\textsubscript{tot,capex}}, and the \gls{OPEX}, \textbf{C\textsubscript{tot,opex}} (see Equation (\ref{eq:obj_func_v2})).
@@
 % adding 25pt space, otherwise flalign with two "&" would flush to the extreme 
-\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}&%\\
+\hspace{0pt} \min \text{  } & \textbf{C\textsubscript{tot,trans}} = \textbf{C\textsubscript{tot,capex}} + \textbf{C\textsubscript{tot,opex}}.&%\\
 \end{flalign}
@@
 \textbf{F\textsubscript{decom}}(p2,p,j)\right)&
-\forall j \in \emph{TECH}
+\forall j \in \emph{TECH}.
 \end{flalign}
@@
 EnergyScope proposes to account for the embodied emissions of the technologies based on a \gls{LCA}.
-These stand for extraction of materials, refining, construction and end of life \cite{schnidrig2023integration}.
+These stand for extraction of materials, refining, construction and end of life \cite{schnidrig2023integration} and can be one to three times higher than the direct emissions depending on the level of emission reduction \cite{blanco2020life}.
 However, this work is still in progress and the database is not yet complete.
@@
 \begin{flalign} 
- &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), j \in \text{\emph{TECH}}
+ &&  \forall p \in \text{\emph{PHASE}}, \emph{y\textsubscript{stop}} \in \emph{Y\_STOP}(p), \emph{y\textsubscript{start}} \in \emph{Y\_START}(p), j \in \text{\emph{TECH}},
  \end{flalign}
@@
 
-\subsection{Myopic: Sequential optimisation of the transition with limited foresight}
+\subsection[Myopic: Sequential optimisation of the transition]{Myopic: Sequential optimisation of the transition with limited foresight}
 \label{subsec:meth:MY}
@@
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=\textwidth]{MY_schematic.pdf}
-\caption{The myopic approach (in pink) uses several instances of the pathway model (illustrated in Figure \ref{fig:meth_path_methodology}).
-In this example, the pathway instance has a time horizon of 10 years ($N\textsubscript{year,opti}=10$) with a 5 year-overlap ($N\textsubscript{year,overlap}=5$).
-As a comparison the Perfect foresight (in blue) has a time horizon of 30 years.}
-\label{fig:my_schematic}
-\end{figure}
-
-\noindent
 Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has three main advantages: shorter computational time, more realistic representation of the short-sightedness of decision-makers and the ability to capture shocks and unexpected events \cite{mccollum2020energy}.
@@
 
-As illustrated in Figure \ref{fig:my_schematic_2}, after optimising, in design and operation, one time window (\eg from 2020 to 2030), the intermediate system design (\ie the installed capacities) is set as initial conditions for the start of the next time window (\eg from 2025 to 2035) as well as the historical investment decisions (\ie $\textbf{F\textsubscript{new}}$, $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$).
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{MY_schematic.pdf}
+\caption{The myopic approach (in pink) uses several instances of the pathway model (illustrated in Figure \ref{fig:meth_path_methodology}).
+In this example, the pathway instance has a time horizon of 10 years ($N\textsubscript{year,opti}=10$) with a 5 year-overlap ($N\textsubscript{year,overlap}=5$).
+As a comparison the Perfect foresight (in blue) has a time horizon of 30 years.}
+\label{fig:my_schematic}
+\end{figure}
+
+After optimising, in design and operation, one time window (\eg from 2020 to 2030), the intermediate system design (\ie the installed capacities) is set as initial conditions for the start of the next time window (\eg from 2025 to 2035) as well as the historical investment decisions (\ie $\textbf{F\textsubscript{new}}$, $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$) (see Figure \ref{fig:my_schematic_2}).
 Consequently, the solution obtained at the end of the first time window (\eg 2030) as well as potential investment decisions between the start of the second time window and this end-year are discarded.
@@
 \midrule
-$\text{YEARS}\textsubscript{WND}$ 	&	$y\in$Y	& 	\pbox{20cm}{\vspace{1mm} Representative years of the \\ time window to optimize}\\
+$\text{YEARS}\textsubscript{WND}$ 	&	$y\in$Y	& 	\pbox{20cm}{\vspace{1mm} Representative years of the \\ time window to optimise}\\
 $\text{YEARS}\textsubscript{UP TO}$ &	$y\in$Y	& 	\pbox{20cm}{\vspace{1mm} Representative years including the \\ years already optimised, \ie the history}\\
-$\text{PHASE}\textsubscript{WND}$ &  $\emph{p}\in$P & 	\pbox{20cm}{\vspace{1mm} Phases of the time window to optimize}\\
+$\text{PHASE}\textsubscript{WND}$ &  $\emph{p}\in$P & 	\pbox{20cm}{\vspace{1mm} Phases of the time window to optimise}\\
 $\text{PHASE}\textsubscript{UP TO}$ &  $\emph{p}\in$P & 	\pbox{20cm}{\vspace{1mm} Phases including the phases \\ already optimised, \ie the history}\\
@@
 Therefore, optimising again this first year could lead to rounding errors preventing from the optimization to converge.
-For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimize that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
-This remark stays valid for any time window except the first one of the transition where the year 2020 is optimized even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
+For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimise that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
+This remark stays valid for any time window except the first one of the transition where the year 2020 is optimised even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
 Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Equation (\ref{eq:salvage}).\\
@@
 In a nutshell, this API allows the pre/post-processing of an ampl optimisation problem by accessing its features (\eg constraints,parameters, variables, objective function) from within Python.
-Using this\gls{API}, this updated version of the model interacts with the AMPL problem representing the optimization of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
+Using this \gls{API}, this updated version of the model interacts with the AMPL problem representing the optimisation of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
+\\
 
@@
 \includegraphics[width=10cm]{MY_process_code.pdf}
-\caption{Schematic of the iterative optimization of the whole-energy system transition pathway.}
+\caption{Schematic of the iterative optimisation of the whole-energy system transition pathway.}
 \label{fig:MY_process_code}
@@
 
+\newpage
 Finally, the model accounts for thousands of parameters.
@@
 
-\noindent where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (\autoref{app:UC_full}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$.
+\noindent where the vector $\bm{\xi} = (\xi_1,\xi_2, \dots \xi_d)$ comprises the independent random input parameters (see Section \ref{sec:cs:uncertainty}), $d$ corresponds to the number of input distributions and $\bm{\alpha}$ is a multi-index, \ie a vector of non-negative indices of length $d$, where each index corresponds to the degree of each univariate polynomial that forms the basis of the multivariate polynomial $\bm{\Psi_{\bm{\alpha}}}$.
 The coefficients ($u_0, u_1, \dots, u_{P+1}$) are quantified using a regression method applied to orthonormal polynomials~\cite{Sudret2014}.
@@
 On top of this selection, we have added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
-The exhaustive list of these 34 parameters is presented in Appendix \ref{app:UC_full}.
-
-\section{Agent-based reinforcement learning for energy transition support}
+The exhaustive list of these 34 parameters is presented in Section \ref{sec:cs:uncertainty}.
+
+\section{Agent-based reinforcement learning to support myopic energy transition}
 \label{sec:meth:RL}
 
-% See chatGPT
-% 1. Introduction to Reinforcement Learning --> Methodo --> DONE
-% 2. Problem Formulation --> Methodo --> DONE
-% 3. Markov Decision Process (MDP) --> Methodo --> DONE
-% 4. RL Algorithms and Techniques --> Methodo --> DONE
-% 5. Training Setup
-% 6. Evaluation Metrics
-% 7. Baseline and Comparisons --> Results
-% 8. Implementation Details --> Chap_RL --> DONE
-% 9. Experimental/Testing Setup
-
-The transition towards carbon-neutrality of a whole-energy system (i.e.
-including all streams of energy carriers and demands) is uncertain.
+The transition towards carbon-neutrality of a whole-energy system (\ie including all streams of energy carriers and demands) is uncertain.
 Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), policy makers rather go through multiple rolling-horizon short-term decisions.
 Yet, these decisions can have long-term impacts, 20 to 50 years.
-This long-term future is intrinsically uncertain and could be the place for potential sudden unexpected events.
-Meeting the environmental objectives while minimizing the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
-To navigate through the transition and investigate the efficiency of different policies, this work implements the reinforcement learning approach.
+The long-term future is intrinsically uncertain and could be the place for potential sudden unexpected events.
+Meeting the environmental objectives without prescribed \ce{CO2}-pathway while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crises, requires a framework to assess the relevance and the timing of the decisions throughout the transition.
+
+To address this issue, several approaches have already been used.
+Among them,  multi-stage stochastic programming is often put forward as a promising method.
+Stochastic programming formulates the problem as a mathematical program with probabilistic constraints or objective functions.
+These models explicitly consider the uncertainty by incorporating probability distributions for the uncertain parameters.
+The goal is to find an optimal decision that minimises/maximises the expected value of the objective function while satisfying the probabilistic constraints, modelled as a scenario tree.
+At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}.
+Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
+The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
+However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly.
+Increasing the number of these uncertainties in stochastic programming increases exponentially the computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
+Based on the approach of \citet{bertsimas2004price}, and similarly to \citet{Moret2017PhDThesis}, \citet{nicolas2021robust} rather opted for the robust optimisation of the global pathway up to 2200 given different temperature deviation targets, \ie 2 or 3°C by 2200 via the use of uncertainty budget, $\Gamma$, in the TIAM-World model \cite{loulou2005documentation}.
+Considering 9 climate parameters and their respective lower and upper bounds, the idea behind the uncertainty budget stems from the improbability of all parameters simultaneously reaching one of their two extreme values.
+However, robust optimisation fails to provide a comprehensive hedging strategy, that can help guide near-term actions, like stochastic programming.
+
+To keep the best of both worlds, \ie stochastic programming and robust optimisation, we decided to investigate the \gls{RL} approach, and its policy optimisation mechanism, to explore myopic transition pathways under uncertainties.
+Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
+\gls{RL} exhibits two main advantages: its effectiveness to handle uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimise the policy \cite{perera2021applications}.
+
 This section aims at presenting the general concepts of this approach.
@@
 
-\subsection{Reinforcement learning fundamentals and application to energy systems}
+\subsection[Reinforcement Learning fundamentals]{Reinforcement Learning fundamentals and application to energy systems}
 \label{subsec:meth_RL_fundamentals}
@@
 The agent takes some action that has an impact on the environment which feeds back the agent with a reward and the new state.
-The objective of the agent is to optimize its policy, \ie mapping between the state it is at and the action to take, by maximizing its cumulative reward.}
+The objective of the agent is to optimise its policy, \ie mapping between the state it is at and the action to take, by maximising its cumulative reward.}
 \label{fig:RL_Fundamentals}
@@
 
-The agent learns to optimize its policy by maximizing a notion of cumulative reward over time.
+The agent learns to optimise its policy by maximising the cumulative reward over time.
 This policy refers to the strategy or mapping from states to actions that the agent employs to make decisions.
 Essentially, it defines the behavior of the agent in the environment.
-The ultimate goal of the agent is often to find an optimal policy, which maximizes the expected cumulative reward over time.
+The ultimate goal of the agent is often to find an optimal policy, which maximises the expected cumulative reward over time.
 All these concepts and interactions between the agent and its environment are formalized as a \gls{MDP} \cite{sutton2018reinforcement}, represented by the tuple $<s,a,T,r,\pi,\gamma>$.
 The Markov property of such a decision process states that a decision is made only based on this tuple and not on the history/path that has led to it.
-In this tuple, $s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions; $a\in A$ is the action among the action space, $A$; $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$; $r$ is the reward received by the agent when taking the action $a$ from state $s$, $R(s,a)$; $\pi$ is the policy telling the action to take depending on the current state and; $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
-During the learning/optimization process, the agent acts according to the exploitation-exploration trade-off.
+In this tuple, $s\in S$ is the state defined in a certain state space, $S$, that represents the observable parts of the environment that the agent uses to make decisions; $a\in A$ is the action among the action space, $A$; $T$ is the probability of transitioning from one state $s$ to another state $s'$ given a specific action, $a$: $T(s,a,s'): \text{Pr}\left(s'|s,a\right)$; $r$ is the reward received by the agent when taking the action $a$ from state $s$, $r(s,a)$; $\pi$ is the policy telling the action to take depending on the current state and; $\gamma$ is the discount factor that controls the importance of future rewards versus immediate rewards.
+During the learning/optimisation process, the agent acts according to the exploitation-exploration trade-off.
 In the exploitation, the action $a$ is directly given by the mapping provided by the current policy $\pi$,  depending on the state $s$.
@@
 Due to the increasing complexity of the systems and the integration of uncertainties, the last decades have seen the emergence of publications where \gls{RL} is applied to energy systems \cite{cao2020reinforcement,perera2021applications}.
-In their respective reviews, \citet{cao2020reinforcement} and \citet{perera2021applications} highlighted groups of problems addressed with \gls{RL} in the research field of energy systems: building energy management system (BEMS), optimization of dispatch and operational control closely linked with the energy market and the optimal power flow problem in the grid, micro-grid management, electro-mobility or even demand-side management or optimal control of energy system devices like maximum power point tracking (MPPT) of wind turbines and \gls{PV} panels.
-The major novelty of this thesis is the application of \gls{RL} to a new kind of energy system problem: the optimization of the transition pathway of a whole-energy system.
-In this sense, the objective is to optimize and provide a policy to support this transition subject to uncertainties.
+In their respective reviews, \citet{cao2020reinforcement} and \citet{perera2021applications} highlighted groups of problems addressed with \gls{RL} in the research field of energy systems: building energy management system (BEMS), optimisation of dispatch and operational control closely linked with the energy market and the optimal power flow problem in the grid, micro-grid management, electro-mobility or even demand-side management or optimal control of energy system devices like maximum power point tracking (MPPT) of wind turbines and \gls{PV} panels.
+The major novelty of this thesis is the application of \gls{RL} to a new kind of energy system problem: the optimisation of the transition pathway of a whole-energy system.
+In this sense, the objective is to optimise a policy to support this transition subject to uncertainties.
 
@@
 Finally, the ``natural'' end is the prescribed end of the transition, \ie 2050.
-Consequently, the maximum value of steps for an episode is equal to $N_{ep,max}=5$.
-
+Consequently, the maximum value of steps for an episode is equal to $N=5$.
+
+\newpage
 Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of \gls{RL} with \gls{DNN}, called \gls{DRL}.
@@
 
-The goal of a \gls{RL} approach is to optimise the mapping between inputs (\ie the observations) and output (\ie the actions), called the policy $\pi\left(\bm{a}_n | \bm{o}_n\right)$.
-To do so, an objective function, $\bm{J}(\pi)$, is built on the cumulative rewards collected during each episode.
+To optimise the mapping between the observations and the actions, the policy $\pi\left(\bm{a}_n | \bm{o}_n\right)$, an objective function, $\bm{J}(\pi)$, is built on the cumulative rewards collected during each episode.
 Finally, a back-propagation process updates the weights and biases of the \gls{NN} during the learning of the agent.
@@
 
-\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented\footnote{The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Equation (\ref{eq:SAC_objective}).} objective function (see Equation (\ref{eq:SAC_objective})).
-Where entropy represents the amount of energy in a system not available to produce work in thermodynamics, this term, also called Shannon entropy in the \gls{RL} context, stands for the randomness or stochasticity of the policy.
+\gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function (see Equation (\ref{eq:SAC_objective})).
+The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Equation (\ref{eq:SAC_objective}).
+In the \gls{RL} context, the entropy, also called ``Shannon entropy'', stands for the randomness or stochasticity of the policy.
 
@@
 \label{sec:meth:PCA}
-When optimizing a transition pathway of a whole-energy system, including its uncertainties, capturing the variance of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
+When optimising a transition pathway of a whole-energy system, including its uncertainties, capturing the variance of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
 For the case study detailed in Chapter \ref{chap:case_study}, this consists of 113 possibles technologies over 7 representative years.
@@
 
-From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximize their spread, \ie variance (see \ref{fig:PCA_intro_proj}).
+From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximise their spread, \ie variance (see Figure \ref{fig:PCA_intro_proj}).
 Another way to understand this is to look at the other side of the same coin.
 Indeed, \gls{PCA} also looks for properties that allow reconstructing the original features as accurately as possible.
-In practice, it aims at minimizing the total reconstruction error that is the average squared distance between the original observations and their respective projection.
+In practice, it aims at minimising the total reconstruction error that is the average squared distance between the original observations and their respective projection.
 
@@
 Overall, the original dataset is $\textbf{x}(y,j,s)$ where, on top of $y$ and $j$ previously defined, $s \in [1,2,...,S]$ stands for the sample number of the uncertainty quantification method.
-For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimization of the transition pathway under uncertainties.
+For the investigated case detailed in Chapter \ref{chap:atom_mol}, this represents $S=1260$ samples resulting from the perfect foresight optimisation of the transition pathway under uncertainties.
 Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.
@@
 \caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition.
-These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimized through the perfect foresight approach.
+These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimised through the perfect foresight approach.
 Over the transition, sectors like electricity (\ie from 22\% in 2025 to 33\% in 2050) or ammonia (\ie from 2\% to 8\%) become more important due to sector coupling, \eg e-mobility or ammonia-\gls{CCGT}.}
@@
 
-To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimized according to the perfect foresight approach and considering all the uncertain parameters to their nominal values (see Chapter \ref{chap:atom_mol}).
+To do so, we have arbitrarily considered these shares from the REF case, where the pathway is optimised according to the perfect foresight approach and considering all the uncertain parameters to their nominal values (see Chapter \ref{chap:atom_mol}).
 The end-use-demands as well as the commodity produced for the sector coupling are based on the results of this deterministic REF case.
@@
 There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
-Since all the data points correspond to a result provided by the optimization, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
+Since all the data points correspond to a result provided by the optimisation, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
 In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively.
@@
 
+\newpage
 \myparagraph{Robustness assessment of roadmaps}\\
@@
 \centering
-\includegraphics[width=0.8\textwidth]{Projection_PC_transition.pdf}
+\includegraphics[width=0.65\textwidth]{Projection_PC_transition.pdf}
 \caption{Projection on the first PC of the transition, $\text{PC}_{\text{transition},1}$, of the different myopic runs under uncertainties based on different roadmaps.

### 2024-06-05 rebuttals/rebuttals.tex
Typos + Init Rebuttals

\documentclass[12pt,a4paper]{article}
\usepackage{longtable}
\usepackage[LGR,T1]{fontenc}
\usepackage{lscape}
%\usepackage{pdfsync}
\usepackage{multirow}
\usepackage{amsmath,bm} 
\usepackage{amsfonts}
\usepackage{amsthm} % Extended theorem environments
%\usepackage{amssymb} % Math symbols %%redundant with stix package
%\usepackage{esint} % Intégrales multiples
%\usepackage{esvect} % Vecteurs
\usepackage{mathtools}
\usepackage{pifont}

\usepackage{fancyhdr}
\usepackage{graphicx}

\graphicspath{{../frontend/img/}{../chap_intro_ccl/img/}{../chap_case_study/img/}{../chap_ES_PCE/img/}{../chap_electro_uq/img/}{../chap_methodo/img/}{../chap_myopic/img/}{../chap_RL/img/}{../chap_atom_mol/img/}{../chap_RobPol/img/}{../appendices/img/}} % Figures folder different for each input
%\graphicspath{{img/}} %
\DeclareGraphicsExtensions{.eps,.pdf,.png,.jpg}

\usepackage{lastpage}
\usepackage{afterpage}
\usepackage{lettrine}
\usepackage{color,soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{colortbl}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{titlesec}
\def\eg{e.g.,\ }
\def\ie{i.e.,\ }

%Palatino font
%\usepackage{pxfonts}
%\usepackage{libertine}
\usepackage[scaled=0.88]{beraserif}
\usepackage[scaled=0.85]{berasans}
\usepackage[scaled=0.84]{beramono}
\usepackage{mathpazo}
%\linespread{1.05}
\usepackage[T1,small,euler-digits]{eulervm}

\usepackage[nomessages]{fp}

\definecolor{bleuUCLclair}{rgb}{.09, 0.569, 1}
\definecolor{bleuUCLfonce}{rgb}{ .13, .52, .86}
\definecolor{redBurn}{rgb}{.91, 0.29, 0.08}

\usepackage{tocloft}
\renewcommand{\cftsecpresnum}{Reviewer \#}
\renewcommand{\cftsecnumwidth}{6em}
\renewcommand{\cftsubsecpresnum}{Comment \#}
\renewcommand{\cftsubsecnumwidth}{6em}

\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
\usepackage[square,numbers,sort&compress]{natbib}

\bibliographystyle{biblio/elsarticle-num-names}  %Ordered by appearance in the text, with DOI and URL
%\usepackage[backend=biber, natbib=true, style=numeric-comp, citestyle=numeric-comp, sorting=none, giveninits=true, maxcitenames=1]{biblatex}

\usepackage[]{tocbibind}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
%    linkcolor=black,
    bookmarks=true,
    pdfpagemode=FullScreen,
}
\setcounter{tocdepth}{1}

\addtolength{\topmargin}{-1.5cm}
\addtolength{\textheight}{1.5cm}
\addtolength{\textwidth}{2cm}
\addtolength{\footskip}{2cm}
\setlength{\evensidemargin}{-0.5cm}
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\arrayrulewidth}{0.25pt}

\renewcommand{\baselinestretch}{1.1} % Interligne

\newenvironment{maliste}%
{ \begin{list}%
	{\textcolor{bleuUCLfonce}{$\bullet$}\hspace{0.5cm}}%
	{\setlength{\labelwidth}{50pt}%
	 \setlength{\leftmargin}{25pt}%
	 \setlength{\itemsep}{30pt}}}%
{ \end{list} }

%\renewcommand{\headrulewidth}{0.0pt}
%\newcommand{\clearemptydoublepage}{%
%	\newpage{\pagestyle{empty}\cleardoublepage}}

%section like title in longtable
\newcommand{\seclong}[1]{\multicolumn{2}{@{}l}{{\Large\sffamily #1}} 
\vspace{0.5cm}
\\}

%enumerate on two columns
\newcounter{listlong}
\newcommand{\newlistlong}{\setcounter{listlong}{1}}
\newcommand{\iteml}[1]{%
\hspace{4.5cm}\textcolor{redBurn}{\arabic{listlong}}\stepcounter{listlong}%
&%
#1%

\\%
}

%left in column
\newcommand{\lcol}[1]{%
\begin{minipage}[t]{.35\textwidth}%

#1%

\end{minipage}%
}

\title{\vspace{-1cm}
\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members}\end{flushleft}}
\date{\vspace{-1.7cm}\begin{flushleft}\sffamily Exploration of uncertainty-aware energy transition pathways - Reinforcement learning and principal component analysis-based methods\end{flushleft}}
%
%Xavier Rixhon, Gauthier Limpens, Diederik Coppitters, Hervé Jeanmart and Francesco Contino\end{flushleft}}

\pagestyle{fancy} 
\fancyhf{}
\fancyfoot[R]{\sffamily\thepage\ / \pageref{LastPage}}
  \fancyfoot[L]{ }

\fancypagestyle{plain}{%
  \fancyhf{}%
  \fancyfoot[R]{\sffamily\thepage\ / \pageref{LastPage}}
  \fancyfoot[L]{ }
}

\renewcommand{\headrulewidth}{0.0pt}

\newcommand{\hlc}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }

\titleformat{\section}
  {\bfseries\scshape}{}{1em}{}

\titleformat{\subsection}
  {\normalfont\scshape}{}{1em}{}

\usepackage[framemethod=default]{mdframed}
%\mdfsetup{skipabove=\topskip,skipbelow=\topskip}

\global\mdfdefinestyle{comment}{%
     linecolor=red,linewidth=0.1cm,%
     leftmargin=-0.5cm,rightmargin=-0.5cm, innerleftmargin=0.4cm,innerrightmargin=0.4cm,
     topline=false,bottomline=false
}

\global\mdfdefinestyle{manuscript}{%
     linecolor=gray!20,linewidth=0.05cm,backgroundcolor=gray!20,%
     leftmargin=-0.5cm,rightmargin=-0.5cm, innerleftmargin=0.4cm,innerrightmargin=0.4cm
}

  \renewcommand{\subsectionautorefname}{Comment}
\begin{document}
\maketitle
%\emph{Start with some thank you note. For example:}

I would like to thank the jury members for the comments, which significantly helped improving the manuscript and substantiating the novelty of my work. Based on the notes taken during the private defense, I hereby transcribed, as accurately as possible, the jury's comments.

I believe I have addressed all the issues raised in the following answers. Some of them required adaptations of the text. These adaptations are either directly in the thesis manuscript or left for further developments in subsequent papers. For each comment, I have first highlighted the issue, then provided an answer, and finally described how the manuscript was adjusted, if needed.\\

When a comment explicitly came from specific members of the jury, they are listed at the beginning of the comment with the following color code: {\color{orange} \textbf{Stefano} Moret}, {\color{teal} \textbf{Stefan} Pfenninger}, {\color{purple} \textbf{Sylvain} Quoilin} and {\color{violet} \textbf{Christophe} De Vleeschouwer}. Here is how an answer to a comment is structured:

\begin{mdframed}[style=comment] % Comment from the reviewer
Text of the comment.
\end{mdframed}

\noindent The answer provided to the comment and, based on this, where the potential modification brought to the manuscript is located in the thesis manuscript using {\color{blue} blue font}.

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
New version of the text in the manuscript.
\end{mdframed}

\clearpage

%
% Reviewer 1
%
\section{Introduction}
\label{Introduction}
%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal. In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly. The following replies include as well all the changes we made to our manuscript based on your comments.

\subsection{Literature review}
\label{Intro_lit_review}

\section{Reinforcement Learning}
\label{RL}
%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal. In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly. The following replies include as well all the changes we made to our manuscript based on your comments.

\subsection{Binding}
\label{RL_binding}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} - The concept of binding constraint was not clear.
\end{mdframed}

\noindent
Besides the information given in Section 4.2.3 of the manuscript (and reminded here below), I do not see further explanations that could clarify the concept of binding constraint in a Linear Programming (LP) problem. Consequently, regarding this comment, there has not been further modification brought to the manuscript.

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
To identify the actions that have an actual impact on the environment, we can check if they are binding or not. In a LP problem, constraints represent hyperplanes in the domain of variables. In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}). When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility. The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function. Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge. In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}. The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal. In other words, binding constraints, when tightened, aggravate the objective value function. If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.
\end{mdframed}

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
\caption{Binding versus non-binding constraints. In LP where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$). The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution. These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
\label{fig:Binding_constr}
\end{figure}

\section{Principal Component Analysis}
\label{PCA}

\subsection{Confusion with the word PC}
\label{Confusion_PC_wording}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} When using the word ``component'', there seems to be a confusion and it is not always easy to understand if you refer to the vector or the coefficient related to one of the original variable.
\end{mdframed}

\noindent The confusion probably comes from the fact a Principal Component (PC) actually represents an eigenvector of the covariance matrix. This vector is, by definition, composed of \textbf{components}, each of them being a coefficient related to a specific original variable. Here below the parts where there could have been misleading confusion are listed and, potentially, modifications have been brought to the manuscript:

{\color{blue} End of second paragraph of section 1.4.1 - no modification}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
Moreover, this means that $\alpha_{ki}$, \ie the \textbf{component} of $\bm{\alpha}_{\mathbf{k}}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$. 
\end{mdframed}

\clearpage
\def\bibfont{\scriptsize}
\bibliography{../bib_thesis.bib}
\normalsize

\end{document}

### 2024-06-05 chap_RL/chap_RL.tex
Typos + Init Rebuttals

@@
 
-Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}.
+Uncertainties about the future along with a large variety of \gls{IAMs} yield an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}.
 For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century.
@@
 
-To identify the actions that have an actual impact on the environment, we can check if they were binding or not.
+To identify the actions that have an actual impact on the environment, we can check if they are binding or not.
 In a \gls{LP} problem, constraints represent hyperplanes in the domain of variables.
@@
 \includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
-\caption{Comparison of cumulative OPEX (left), CAPEX (right) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
+\caption{Comparison of cumulative OPEX (left), CAPEX (center) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
 \label{fig:Opex_Capex_Salvage_comp}

### 2024-06-05 chap_RobPol/chap_RobPol.tex
Typos + Init Rebuttals

@@
 On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''.
-Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.
+Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of additional investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.
+Finally, we assess the variation of the transition costs resulting from these myopic transitions under uncertainties.
 
@@
 1 & \gls{PV} and electrification & $\simeq$ & +49\% \\
-2 & \gls{PV} and dec.heating & $\simeq$ & +37\%\\
+2 & \gls{PV} and dec. heating & $\simeq$ & +37\%\\
 3 & Freight transport & $\simeq$ & +33\%\\
@@
 
+\subsection{What about the costs over the transition?}
+\label{subsec:RobPol:costs_comparison}
+Besides the additional capacities needed through the myopic transitions, this section takes a step back to assess the variations of costs over the transition: the total transition cost, the cumulative \gls{OPEX} by 2050 and the cumulative \gls{CAPEX} by 2050 (see Figure \ref{fig:Total_Opex_Capex_REF_ROB}).
+We observe that the differences are marginal.
+This validates the initial rationale to go beyond these costs to assess the robustness of roadmaps.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.325\textwidth]{Transition_cost_REF_ROB.pdf}
+\includegraphics[width=0.325\textwidth]{OPEX_2050_REF_ROB.pdf}
+\includegraphics[width=0.325\textwidth]{CAPEX_2050_REF_ROB.pdf}
+\caption{Comparison of total transition cost (left), cumulative OPEX (center) and CAPEX (right) by 2050 from the REF and ROB myopic transitions.
+The differences are marginal.
+However, the ROB roadmap lead to cheaper and less varying total transition cost.
+Requiring more investments at earlier stages, this roadmap is less affected by the variation of the operational costs.}
+\label{fig:Total_Opex_Capex_REF_ROB}
+\end{figure}
+
+Considering the median values, the ROB roadmap leads to a 1.6\% more expensive total transition.
+This comes from the earlier investments in \gls{PV} panels together with electrified solutions in the heating sector that need to be renewed before the end of the transition.
+The extra-\gls{CAPEX} also comes from more efficient (and more expensive) technologies such as \gls{CHP}.
+These decisions allow being less affected by the variation of the cumulative operational costs that are dominated ($\sim$\,70\%) by the cost of purchasing energy carriers.
+This makes the investment-to-operation ratio slightly higher for the ROB case (46\%) than for the REF case (44\%).
+
+We can assess the variation of these costs with the quartile coefficient of dispersion.
+This coefficient of dispersion is defined as $\left(Q_3-Q_1 \right)/\left(Q_3+Q_1 \right)$ where $Q_1=P_{25}$ and $Q_3=P_{75}$ are the first and third quartiles of the data, respectively.
+We observe that the ROB case has a smaller quartile coefficient of dispersion for the total transition cost and cumulative \gls{OPEX} but higher for the cumulative \gls{CAPEX} (see Table \ref{tab:quartile_coeff_dispersion}).The biggest decrease of these coefficients of dispersion between the REF and the ROB cases concerns the total transition cost.
+This means that the ROB case ensures to decrease more the uncertainty on this cost.
+
+\begin{table}[htbp!]
+\caption{Quartile coefficient of dispersion for the total transition cost, cumulative OPEX and CAPEX by 2050 from the REF and ROB myopic transitions.
+The ROB case leads to more certain total transition cost and cumulative \gls{OPEX} than the REF case.}
+\label{tab:quartile_coeff_dispersion}
+\centering
+\begin{tabular}{l| c c}
+\toprule
+\textbf{Quartile coefficient of dispersion} &  REF  & ROB \\
+\midrule
+Total transition cost & 12.9\% & 12.2\%  \\
+Cumulative \gls{OPEX} by 2050 & 21.3\% & 21.1\% \\
+Cumulative \gls{CAPEX} by 2050 & 5.9\% & 6.0\%\\
+\bottomrule							
+\end{tabular}
+\end{table}
+
+All these conclusions are similar to the ones observed on the study of overcapacity of \citet{moret2020overcapacity}).
+Working on the snapshot model and fixing the installed capacities only in the power sector, they although observed bigger gain to opt for a robust strategy rather than the deterministic solution.
+
 \newpage
@@
 However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario.
-Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
+Even though this roadmap represents 7.4\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
+On top of decreasing the risk to invest into additional capacities, the ROB is also less sensitive to the variation of the costs of purchasing the energy carriers that are the main contributors of uncertainty.
 

### 2024-06-05 chap_atom_mol/chap_atom_mol.tex
Typos + Init Rebuttals

@@
 Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand in methanol.
-Its \gls{NED} represents, on average, 3\% of the total consumption of methnaol.
+Its \gls{NED} represents, on average, 3\% of the total consumption of methanol.
 The variation of imported e-methanol is due to its role in the industrial production of \gls{HVC} through the \acrfull{MTO} process as it represents 95\% of the total consumption.
@@
 The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050.
-We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transiton.Next to the name of a parameter, one can read its Sobol' index versus the output of interest.
+We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transition.
+Next to the name of a parameter, one can read its Sobol' index versus the output of interest.
 For these outputs of interest, different from the total transition cost, the \gls{LOO} error is around 20\%.

### 2024-06-05 chap_case_study/chap_case_study.tex
Typos + Init Rebuttals

@@
 In practice, it means that their \gls{GWP} is assumed to be zero in the model.
-Compared to equivalent electrofuels that would be locally produced, this assumption does not account account for the emissions related to their transport of imported electrofuels from abroad to Belgium.
+Compared to equivalent electrofuels that would be locally produced, this assumption does not account for the emissions related to the transport of imported electrofuels from abroad to Belgium.
 However, the share of these emissions is limited (between 5 and 10\%) compared to the whole cycle of production and use of imported electrofuels \cite{coppitters2024towards}.
@@
 Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability.
-Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
+Finally, the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
 

### 2024-06-05 chap_intro_ccl/conclusion.tex
Typos + Init Rebuttals

@@
 About the use of the electrofuels, there are two molecules imported in larger quantities: e-methanol and e-ammonia.
-E-ammonia would aim at defossilising the production of \acrfull{HVC} in the non-energy sector and e-ammonia would dominantly provide flexibility in the production of electricity, respectively.
+E-methanol would aim at defossilising the production of \acrfull{HVC} in the non-energy sector and e-ammonia would dominantly provide flexibility in the production of electricity, respectively.
 The main use of e-hydrogen is in the freight transport via trucks and e-methane would supply the industrial sectors through \gls{CHP} and boilers.

### 2024-06-05 chap_intro_ccl/intro.tex
Typos + Init Rebuttals

@@
 From an energy point of view, this identity can be adapted by replacing the economic metric, \gls{GDP}, by the \gls{EUD}, see Equation (\ref{eq:equality_GHG}).
-\gls{EUD} is the energy service required by the the final consumer.
+\gls{EUD} is the energy service required by the final consumer.
 In Equation (\ref{eq:equality_GHG}), the first factor on the right hand side represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
@@
 
-Electrofuels offer three main advantages: infrastructure compatibility, capacity to link sectors (\ie from electricity to mobility, heat, or industry) and storagr.
+Electrofuels offer three main advantages: infrastructure compatibility, capacity to link sectors (\ie from electricity to mobility, heat, or industry) and storage.
 Given their similar physicochemical properties, electrofuels could substitute fossil in technologies that are already in place today.
@@
 
-Electrofuels have also the ability to couple energy and non-energy sectors \cite{Stancin2020}.
+Electrofuels also have the ability to couple energy and non-energy sectors \cite{Stancin2020}.
 For instance, electricity from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser, coupling the power and agro-industrial sectors \cite{verleysen2020can}.
@@
 A variety of techniques are used to assess this robustness including Monte-Carlo analysis, stochastic programming or robust optimisation \cite{yue2018review}.
-However, the robustness is commonly assessed to evaluate the sensibility of the total cost (objective function) to input parameters.
-Given the complexity of whole-energy system models, the time scale of the transition and the large number of uncertainties, only considering the total cost does not give information on the sensibility of the design strategy, \ie the investment decisions.
-For instance, total cost might consist in excessive upfront investments and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable depsite a slightly higher total cost.
+However, the robustness is commonly assessed to evaluate the sensitivity of the total cost (objective function) to input parameters.
+Given the complexity of whole-energy system models, the time scale of the transition and the large number of uncertainties, only considering the total cost does not give information on the sensitivity of the design strategy, \ie the investment decisions.
+For instance, total cost might consist in excessive upfront investments and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable despite a slightly higher total cost.
 \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
 However, their work focused on the power sector only and for a target future year (2035).
-Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost and provide more details on the sensibility of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
+Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost and provide more details on the sensitivity of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
 
@@
 ``\textit{Our task is not to foresee the future, but to enable it}''\footnote{Saint-Exupéry in Citadelle, 1948}.
+Rather than trying to answer the question ``What could possibly happen in the future?'', this work addresses the ``What could or should we do to make the future possible?''.
 This thesis aims at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
-Rather than trying to answer the question ``What could possibly happen in the future?'', this work addresses the ``What could or should we do to make the future possible?''.
 In that spirit, the research questions are described as follows:
@@
 
-Finally, in the objective to assess the sensibility to uncertainties of different transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
+Finally, in the objective to assess the sensitivity to uncertainties of different transition roadmaps, we have defined an approach to come up with a ``robustness metric''.
 This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a reference frame.

### 2024-06-05 chap_methodo/chap_methodo.tex
Typos + Init Rebuttals

@@
 Finally, meeting the environmental objectives while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
-This work encompasses the optimisation of the policy,\ie set of actions to take along the transition with a specific methodology to assess its robustness.
+This work encompasses the optimisation of the policy, \ie set of actions to take along the transition with a specific methodology to assess its robustness.
 
@@
 
-he \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Equation (\ref{eq:gwp_tot_transition})).
+The \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Equation (\ref{eq:gwp_tot_transition})).
 Similarly to initial operational costs to account for the system in place in 2020 (see Equation (\ref{eq:Copex_tot_v2})), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
@@
 
-One of the main methodological contributions of this work regarding the development of the whole-energy system model consists in giving it the possibility to optimise the transition pathway in a myopic approach.After introducing the general concept of it, this section details more the additions brought to the model in terms of implementation.\\
+One of the main methodological contributions of this work regarding the development of the whole-energy system model consists in giving it the possibility to optimise the transition pathway in a myopic approach.
+After introducing the general concept, this section details more the additions brought to the model in terms of implementation.\\
 
@@
 After characterising the uncertainty ranges, \citet{Moret2017} quantified the impact of these uncertainties on the snapshot model of EnergyScope,  \ie ranking them, using the Morris method \cite{morris_factorial_1991}.
-This method, as a statistical analysis, relies on individually randomized one-factor-at-a-time designs.
+This method, as a statistical analysis, relies on individually randomised one-factor-at-a-time designs.
 Given the $d$ model parameters $\vec{\xi}=(\xi_1,\xi_2,...,\xi_d)$, the first step of the method consists in generating independent random samples of $\vec{\xi}$ in a standardised and discretised $p$-level \textit{region of experimentation}, $\omega$.
@@
 
-Beyond the original Morris method, we used the standardized elementary effects, $SEE_{i}$, formulation \cite{Sin2009}, given by
+Beyond the original Morris method, we used the standardised elementary effects, $SEE_{i}$, formulation \cite{Sin2009}, given by
 \begin{equation}
@@
 Among other things, the $SEE$ allows comparing the influence of different inputs on the same output or compare the influence of a same parameter on different outputs, even if these parameters or outputs are significantly different in terms of variation range or average amplitude.
-Moreover, this standardized analysis does not require any additional model evaluations.\\
+Moreover, this standardised analysis does not require any additional model evaluations.\\
 Therefore, in the following results, we rather use
@@
 
-to rank parameters among each other. In Equation (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardized elementary effects, as done in \citet{Moret2017PhDThesis}.
+to rank parameters among each other. In Equation (\ref{eq:mu*}), $SF_{i}$ is the distribution formed by the $r$ standardised elementary effects, as done in \citet{Moret2017PhDThesis}.
 
@@
 The ultimate goal of the agent is often to find an optimal policy, which maximises the expected cumulative reward over time.
-All these concepts and interactions between the agent and its environment are formalized as a \gls{MDP} \cite{sutton2018reinforcement}, represented by the tuple $<s,a,T,r,\pi,\gamma>$.
+All these concepts and interactions between the agent and its environment are formalised as a \gls{MDP} \cite{sutton2018reinforcement}, represented by the tuple $<s,a,T,r,\pi,\gamma>$.
 The Markov property of such a decision process states that a decision is made only based on this tuple and not on the history/path that has led to it.
@@
 To tackle this challenge, we have developed a methodology based on the \acrfull{PCA}.
-The philosophy behind this approach is to identify key-combinations of design variables capturing most of the system variance, beyond their sole objective function, \ie total transition cost, but without having to compare each design variable individually.
+The philosophy behind this approach is to identify key-combinations of design variables capturing most of the system variance, beyond the sole objective function, \ie total transition cost, but without having to compare each design variable individually.
 This methodology provides two main outputs.
@@
 Born in the early 20th century \cite{pearson1901on,hotelling1933analysis}, the \acrfull{PCA} finds its fundamentals from the \gls{SVD}.
-\gls{SVD} is a generalization, to an arbitrary (\ie not necessary square) matrix, of the spectral theorem stating that a normal matrix can be diagonalized by an orthonormal basis of eigenvectors.
+\gls{SVD} is a generalisation, to an arbitrary (\ie not necessary square) matrix, of the spectral theorem stating that a normal matrix can be diagonalised by an orthonormal basis of eigenvectors.
 The core concept of principal component analysis (PCA) involves simplifying the original dataset with numerous interconnected variables by reducing its dimensionality while preserving as much variability within the data.
@@
 These components are uncorrelated and arranged in such a way that the first ones retain the majority of the variability found in the original dataset.
-On the other hand, the final principal components (PCs) pinpoint directions where there is minimal variation, indicating nearly constant linear relationships among the original variables \cite{jolliffe2002principal}.
+On the other hand, the final PCs pinpoint directions where there is minimal variation, indicating nearly constant linear relationships among the original variables \cite{jolliffe2002principal}.
 
@@
 These eigenvectors are orthogonal to each other so that they represent independent directions in the original feature space.
-These eigenvectors are also normalized such that $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$ \cite{jolliffe2002principal}.
+These eigenvectors are also normalised such that $\mathbf{\alpha_k}^{T}\mathbf{\alpha_k}=1$ \cite{jolliffe2002principal}, where $\mathbf{^T}$ means the transpose vector.
 This makes it easier to interpret the relative importance of each principal component in explaining the variability of the data.
@@
   \label{eq:PCA_z1}%7
- z_1=\sum_{j=1}^{p} \alpha_{1j}x_j=\mathbf{\alpha_1^{T}x}=\alpha_{11}x_1+\alpha_{12}x_2,
+ z_1=\sum_{j=1}^{p} \alpha_{1j}x_j=\mathbf{\alpha_1^{T}x}=\alpha_{11}x_1+\alpha_{12}x_2.
 \end{flalign}
@@
 \noindent
-where $\mathbf{^T}$ means the transpose vector.
-Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$,uncorrelated with $z_1$ and capturing the remaining variance.
+Then, $z_2=\mathbf{\alpha_2^{T}x}$, is another linear function of $\mathbf{x}$, uncorrelated with $z_1$ and capturing the remaining variance.
 These linear transformations can be seen as projection of the original data on the principal direction, \ie PCs (see Figure \ref{fig:PCA_intro_PCA}).
-In more general cases, one can write $z_k=\mathbf{\alpha_k^{T}x}$ as the $k^{\text{th}}$ PC.
+
+In the general cases of $p$ original variable, one can write $z_k=\mathbf{\alpha_k^{T}x}$ as the $k^{\text{th}}$ PC.
 There can be up to $p$ PCs even though, usually, most of the variance of the original data can be captured by $m$ PCs where $m\ll p$.
@@
 \noindent
-To characterize the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})).
-Even though these represent only the design part of the result of the optimisation focusing on the installed capacities give a direct information regarding the required capital investment (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation.
+To characterise the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})).
+Even though these represent only the design part of the result of the optimisation, focusing on the installed capacities gives a direct information regarding the required capital investments (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation.
 In other words, it captures the technological landscape of the transition.
@@
 Appendix \ref{app:UQ_tech_cap} gives the distribution of the installed capacities among the different end-use sectors from the \gls{GSA}.
-Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place (see Appendix \ref{app:bel_2020}).
+Finally, among the seven representative years of the transition, we do not consider 2020 as it is the initialisation year for which the design of the system is fixed, to be representative of the actual design that was in place in Belgium in 2020 (see Appendix \ref{app:bel_2020}).
 In other words, we focus here only on the years 2025, 2030, 2035, 2040, 2045 and 2050.

### 2024-06-12 rebuttals/rebuttals.tex
Integration of comments in rebuttals

@@
 
+% {\color{orange} \textbf{Stefano}}
+% {\color{purple} \textbf{Sylvain}}
+% {\color{teal} \textbf{Stefan}}
+% {\color{violet} \textbf{Christophe}}
+
 \clearpage
 
-%
-% Reviewer 1
-%
+\section{General comments/questions}
+\label{General}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Overall, many methods: PCE + RL + NN + PCA. At each stage, some assumption is made. Why not a simpler approach or a simpler model? Do you need all these steps? 
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{Assumptions - Fine tuning}
+\label{fine_tuning}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Many assumptions, e.g., alpha = 85\% (p.
+40) or reward = -300 (Chapter 4).
+How sensitive are the results to these assumptions?
+It looks like all these parameters have been tuned to obtain the desired results.
+How robust are they?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{Graphs}
+\label{general_graphs}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} \& {\color{teal} \textbf{Stefan}} - Graphs should be better explained. Many axes labels are missing, and the captions are often insufficient to read the figures. 
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{Structure}
+\label{structure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - The thesis structure makes it quite difficult to follow.
+Having all the methodology summarised in Chapter 1, makes it difficult to link it to the different chapters.
+As changing the structure may be now a major work, I would invite you to think how to better connect the different parts.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+There are various typos in the thesis, e.g.
+``storagR'' (page 3), ``therefore, therefore'' (p.
+11), ``The'' (p.14), ``Moleclues'' (p.
+61), etc..
+Also, the use English language can be improved, e.g.
+``calls FOR a variety'' (p.
+9), ``cumulative emissions ARE'' (p.
+14), ``than to'' (p.
+34), etc.
+The document will benefit from a spelling/grammar check.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{References}
+\label{references}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - I found it unusual to see many references to own work (Rixhon et al.) throughout the thesis.
+Unless the articles are unrelated to the thesis, these references should be removed and the associated content should be included in the manuscript.
+As an example, at p.
+25, results are mentioned which are not reported in the thesis.
+The reader currently must refer to [48] to see these results, which should instead be included in the thesis.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{violet} \textbf{Christophe}} - State of the art mainly European. Why not reference to other regions of the world.
+\end{mdframed}
+
+\noindent
+It is an interesting topic for further investigations, especially how they address the subject in other parts of the world like Australia/Asia.
+This could be linked with the political perception of energy system optimisation models and the subsequent analyses.
+However, no modification has been brought to the manuscript regarding this topic.
+
+\subsection{Writing style}
+\label{writing_style}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} -While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc.
+I recommend reducing them to the bare minimum, or completely getting rid of them.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{Interest rate}
+\label{irate}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Wouldn't it be more appropriate to talk about ``discount rate'' rather than ``interest rate''?
+\end{mdframed}
+
+\noindent Indeed, as we want to capture the time preference associated with obtaining finance for the project, effectively quantifying the trade-off between present and future financial value, it is more appropriate to refer to ``discount rate'' when considering the parameter $i_{\text{rate}}$.
+We have changed every occurrence of ``interest rate'' into ``discount rate'' {\color{blue}in the entire manuscript}.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Differentiated discount rates could be useful? 
+\end{mdframed}
+
+\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
+(1.3) when discussing the annualised phase factor}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
+In practice, the discount rate would vary depending on the technology investment risk.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+For this reason, we have decided to keep an identical value of discount rate for all the technologies.
+\end{mdframed}
+
 \section{Introduction}
 \label{Introduction}
-%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal.
-In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly.
-The following replies include as well all the changes we made to our manuscript based on your comments.
 
-\subsection{Literature review}
-\label{Intro_lit_review}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} \& {\color{teal} \textbf{Stefan}} - At page 1, it is mentioned that you focus on the technical levers of the transition, ``renewables'' and ``efficiency''.
+From this I understand that you are not focusing on ``sufficiency''.
+This seems to contradict a statement at p.
+2, where it is mentioned that one objective of the thesis is to ``support interdisciplinary projects in the assessment of sufficiency policy''.
+I suggest clarifying this potential misunderstanding.
+\end{mdframed}
 
-\section{Reinforcement Learning}
-\label{RL}
-%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal.
-In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly.
-The following replies include as well all the changes we made to our manuscript based on your comments.
+\noindent I totally agree with this remark, especially because I do not want to overlook the necessary third pillar of the transition which is ``sufficiency'', even though my thesis does not address directly this aspect.
+Consequently, I have adapted {\color{blue} page 2 of the introduction}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Among all the lenses through which it is necessary to assess sufficiency policies, one of the objectives of this work is to support these interdisciplinary projects by providing informed techno-economic guidelines.
+\end{mdframed}
+
+\noindent This technical contribution to the interdisciplinary debate of sufficiency is reminded in {\color{blue} the last sentence of the conclusion}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into the intrinsically political and interdisciplinary discussions and decisions that must be made in the coming years.
+\end{mdframed}
+
+\section{Chapter 1 - Methodology}
+\label{methodo}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.11 : The stated computational time seems low for a pathway model with hourly resolution.
+Were there approximations (e.g., disabling crossover) added and what was their impact?\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Can you explain the logic of the annualization factor in the pathway model and the choice of N=5 for the uncertainty ranges?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Often, multi-stage models are affected by the end-of-time-horizon effects. Did you observe those and how did you deal with them? See, e.g. Figure 3.3 p. 66.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.17-19: is the myopic version modeled as a single LP or as a sequence of LPs?
+This should be better clarified.
+The diagram at p.
+19 focuses on the code, it would be better replaced by a code showcasing the logic of the approach.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 20: “inspired by Guevara et al.” $\rightarrow$ how was their approach used in your work?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+21: footnote.
+The statement is qualitative, it should be quantitative.
+What is the minimum number of runs needed to validate your experiments?
+Did you look into that?
+How was the surrogate model validated?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsection{Uncertainty quantification}
+\label{methodo_UQ}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+23: is PCE an appropriate method for LP / MILP models.
+There is a good fit for the total cost, which is often quite linear, but how about technology choice / sizing?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 24: Sobol indices. Can you explain the choice wrt other methods? Difference between factor prioritization and factor fixing?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Why is PCE needed is EnergyScope is already very fast?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsection{Reinforcement Learning}
+\label{methodo_RL}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 31: Can you explain the need of using NN here? Couldn’t you just run the energyscope model?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsection{Principal Component Analysis}
+\label{methodo_PCA}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 32: Why is PCA needed? Wouldn’t it be easier to just aggregate similar outputs of interest?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+42: Figure 1.17.
+What is your definition of robustness?
+Having a tight distribution but all shifted towards very suboptimal costs seems to fit your definition.
+Please clarify.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\section{Chapter 2 - Case study}
+\label{methodo}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - I am a bit underwhelmed by the many nitty-gritty data in this Chapter.
+While it is surely important to document everything, couldn’t at least part of this be moved to the Appendix?\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Figure 2.5: What is the purpose of the violet line between ``biomass gasification'' and ``methanolation''?
+\end{mdframed}
+
+\noindent These were actually two different arrows, one pointing from ``biomass gasification for CH3OH'' to ``Methanol'' and the other from ``Methanolation'' to ``Methanol''.
+To avoid this confusion, the figure has been updated and the origin of these arrows are now on different vertical levels {\color{blue} see Section 2.3}.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{NED_tech.pdf}
+\label{fig:NED_tech}
+\end{figure}
+
+\section{Chapter 4 - Reinforcement Learning}
+\label{Chap_RL}
 
@@
 This vector is, by definition, composed of \textbf{components}, each of them being a coefficient related to a specific original variable.
-Here below the parts where there could have been misleading confusion are listed and, potentially, modifications have been brought to the manuscript:
+Parts where the word ``component'' is not directly linked to ``vector'', ``eigenvector'' or ``PC'' are those that could mislead the reader.
+Modifications have been brought to these parts.
 
-{\color{blue} End of second paragraph of section 1.4.1 - no modification}:
+{\color{blue} End of second paragraph of section 1.4.1}:
 
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Moreover, this means that $\alpha_{ki}$, \ie the \textbf{component} of $\bm{\alpha}_{\mathbf{k}}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
+Moreover, this means that \textbf{the coefficient} $\alpha_{ki}$, \ie the component of $\bm{\alpha}_{\mathbf{k}}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
 \end{mdframed}

### 2024-06-12 appendices/app_EnergyScope.tex
Integration of comments in rebuttals

@@
 The objective, Equation (\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
-The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Equation (\ref{eq:tau_app}).
+The annualised factor $\tau$ is computed \emph{a priori} based on the discount rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Equation (\ref{eq:tau_app}).
 
@@
 
-As a consequence, a new dimension `\emph{year}' is added to all \textbf{Variables} and parameters, except the interest rate (\emph{i\textsubscript{rate}}) assumed constant during the transition.
+As a consequence, a new dimension `\emph{year}' is added to all \textbf{Variables} and parameters, except the discount rate (\emph{i\textsubscript{rate}}) assumed constant during the transition.
 This new dimension is necessary to represent the changes of technology and resource characteristics over the representative years.
@@
 the product of the annualised phase factor, defined in Equation (\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase.
-The annualised phase factor is defined based on an average interest rate during the transition.
+The annualised phase factor is defined based on an average discount rate during the transition.
 %assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor_app}).

### 2024-06-12 appendices/app_results_UQ.tex
SUBMISSION
Integration of comments in rebuttals

@@
 Next, comes, naturally, the industrial \gls{EUD}, representing, at the nominal value, 60\% of the total demands by 2050.
-The top-3 is completed by the variation of the interest rate, directly impacting the annualisation and the salvage values of the assets.
+The top-3 is completed by the variation of the discount rate, directly impacting the annualisation and the salvage values of the assets.
 Finally, since the current Belgian whole-energy system deeply relies on fossil resources, and would still do in the near future, the cost of purchasing fossil fuels is part of the impacting parameters.
@@
 Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) \\
-Interest rate & 3 & (12.0\%) & 3 & (11.0\%) \\
+Discount rate & 3 & (12.0\%) & 3 & (11.0\%) \\
 Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) \\
@@
 %Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) & 2 & (26.4\%)  \\
-%Interest rate & 3 & (12.0\%) & 3 & (11.0\%) &  3 & (13.2\%)  \\
+%Discount rate & 3 & (12.0\%) & 3 & (11.0\%) &  3 & (13.2\%)  \\
 %Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) & 4 & (6.9\%)   \\
@@
 \caption{Distribution of the installed capacities among the end-use sectors of electricity, \gls{HT} and \gls{LT} heat from the \acrfull{GSA}.}
-\label{fig:results_uq_tech_cap}
+\label{fig:results_uq_tech_cap_ELEC_HT_LT}
 \end{figure}
@@
 \caption{Distribution of the installed capacities among the end-use sectors of ammonia, methanol and \gls{HVC} from the \acrfull{GSA}.}
-\label{fig:results_uq_tech_cap}
+\label{fig:results_uq_tech_cap_NED}
 \end{figure}
@@
 \caption{Distribution of the installed capacities among the end-use sectors of passenger and freight mobility from the \acrfull{GSA}.}
-\label{fig:results_uq_tech_cap}
+\label{fig:results_uq_tech_cap_MOB}
 \end{figure}

### 2024-06-12 chap_RL/chap_RL.tex
Integration of comments in rebuttals

@@
 Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases.
-On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and interest rate.
+On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and discount rate.
 These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2}-budget.
@@
 Industry EUD & 49.8\% & 0.026 \\
-Interest rate & 48.4\% & 0.089\\
+Discount rate & 48.4\% & 0.089\\
 Purchase fossil fuels  & 55.0\% & -0.068\\

### 2024-06-12 chap_RobPol/chap_RobPol.tex
Integration of comments in rebuttals

@@
 In the present analysis, we label the roadmap as ``robust'' (ROB) by setting $\Gamma_{\text{obj}}= 6$.
-Consequently, the following six parameters are set at their worst value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies.
-In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
+Consequently, the following six parameters are set at their worst value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the discount rate and the variable OPEX of technologies.
+In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the discount rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
 
@@
 This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}.
-The increased interest rate and variable OPEX of technologies have a more negligible impact on the roadmap strategy as they identically affect the entire set of technologies.
+The increased discount rate and variable OPEX of technologies have a more negligible impact on the roadmap strategy as they identically affect the entire set of technologies.
 

### 2024-06-12 chap_atom_mol/chap_atom_mol.tex
Integration of comments in rebuttals

@@
 Often compared, if not opposed, to local renewables like wind and solar \cite{suna2016nuclear,khatib2016economics}, this chapter rather assesses the integration of nuclear energy in the future versus the need to import renewable molecules from abroad, in a country where the local potential of \gls{VRES} is limited compared to its \gls{EUD}.
-
-Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
-
-\section*{Contributions}
-\label{sec:atom_mol:contributions}
-
-In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimise a transition pathway with a small number of time stages, \ie two or three.
-Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}).
-Accounting for uncertainties presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}, by the end of the transition, \ie 2050.
-
 Relying on nuclear energy for some is highly controversial.
@@
 This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in the Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
+
+Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters).
+Accounting for uncertainties presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}, by the end of the transition, \ie 2050.
+
+\section*{Contributions}
+\label{sec:atom_mol:contributions}
+
+In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimise a transition pathway with a small number of time stages, \ie two or three.
+Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}).
 
@@
 Industry EUD & 2 & 23.2\% \\
-Interest rate & 3 & 12.0\% \\
+Discount rate & 3 & 12.0\% \\
 Purchase fossil fuels  & 4 & 5.7\% \\
@@
 
-As the driving factor for the annualisation and the salvage value of the assets, the interest rate has a 12\% Sobol' index.
-The model considers an overall interest rate for the entire system (\ie 1.5\% as a nominal value).
-In practice, the interest rate would vary depending on the technology investment risk.
+As the driving factor for the annualisation and the salvage value of the assets, the discount rate has a 12\% Sobol' index.
+As detailed in Chapter \ref{chap:chap_methodo}, the model considers an overall discount rate for the entire system (\ie 1.5\% as a nominal value).
+In practice, the discount rate would vary depending on the technology investment risk.
 This variation would have, for instance, a major impact on the \gls{LCOE} of technologies like nuclear power plants \cite{world_nuclear_asso}, given the important capital needs and long time horizons \cite{IEA_Nuclear_2022}.

### 2024-06-12 chap_case_study/chap_case_study.tex
Integration of comments in rebuttals

@@
 
-This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
+This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the modal share change in different key sectors.
 A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.
@@
 
-\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Discount rate & I & -46.2\% & 46.2\% \\
 & $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\
@@
 %
-%\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+%\multirow{4}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Discount rate & I & -46.2\% & 46.2\% \\
 %& $\Delta_{\mathrm{change,freight}}$ & Modal share change freight mobility & - & -30\% & 30\% \\

### 2024-06-12 chap_intro_ccl/conclusion.tex
Integration of comments in rebuttals

@@
 Due to the limited potential of local renewables, electrofuels are found to be the third pillar of the Belgian transition.
-The uncertainty on the total cost of the transition is mainly due to the uncertainty on the cost of purchasing electrofuels (47\%), on the industrial end-use demand (23\%), on the interest rate (12\%) and on the price of fossil fuels (6\%).
+The uncertainty on the total cost of the transition is mainly due to the uncertainty on the cost of purchasing electrofuels (47\%), on the industrial end-use demand (23\%), on the discount rate (12\%) and on the price of fossil fuels (6\%).
 

### 2024-06-12 chap_intro_ccl/intro.tex
Integration of comments in rebuttals

@@
 Explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, it is defined as ``\emph{a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people}'' \cite{lage2023citizens}.
-Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving economy, sociology, psychology and political science, and is outside the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
-One of the objectives of this work is to support interdisciplinary projects in the assessment of sufficiency policy.
-\\
+Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving economy, sociology, psychology and political science, and is beyond the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
+Among all the lenses through which it is necessary to assess sufficiency policies, one of the objectives of this work is to support these interdisciplinary projects by providing informed techno-economic guidelines.\\
 
@@
 The SMR case is the one introduced in Chapter \ref{chap:atom_mol} where we allow the model to install \gls{SMR} from 2040 onward.
-Eventually, the ROB case accounts for the highest values of parameters having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, interest rate and variable \gls{OPEX} of technologies).
+Eventually, the ROB case accounts for the highest values of parameters having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, discount rate and variable \gls{OPEX} of technologies).
 

### 2024-06-12 chap_methodo/chap_methodo.tex
Integration of comments in rebuttals

@@
 \noindent where $\tau\textsubscript{\emph{phase}}$ is the annualised phase factor and $\emph{c\textsubscript{inv}}(p,j)$ is the arithmetic average of the investment cost of the technology $j$ at the beginning and the end of the phase, $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j)$ and $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)$.
+The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
+In practice, the discount rate would vary depending on the technology investment risk.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+For this reason, we have decided to keep an identical value of discount rate for all the technologies.
+
 Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed (see Equation (\ref{eq:salvage}))
@@
 Industry EUD & 2 & (23.2\%) & 2 & (18.3\%) \\
-Interest rate & 3 & (12.0\%) & 3 & (12.1\%) \\
+Discount rate & 3 & (12.0\%) & 3 & (12.1\%) \\
 Purchase fossil fuels & 4 & (5.7\%) & 5 & (3.5\%) \\
@@
 In other words, given this normalization, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
-Moreover, this means that $\alpha_{ki}$, \ie the component of $ \mathbf{\alpha_k}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
+Moreover, this means that the coefficient $\alpha_{ki}$, \ie the component of $ \mathbf{\alpha_k}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
 This PC captures $\lambda_k$ variance of the original data.
@@
 In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact.
-Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies (\eg \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}).
+Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies.
+For instance, \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}.
 \\
@@
 There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
-Since all the data points correspond to a result provided by the optimisation, we have decided to keep these points but carry out a modification of them \cite{aguinis2013best}.
+Since all the data points correspond to a result provided by the optimisation, we have decided to keep these points but carry out a ``modification'' of them \cite{aguinis2013best}.
 In practice,  the value of ``high outliers'' or ``low outliers'' is set to the upper or lower fence, respectively.
@@
 For each of the representative years of the transition (except 2020), we compute the PCs that capture 90\% of the total variance of this year \cite{jolliffe2002principal}.
-At the end of this step, a list of $m$ PCs is obtained, \ie $\text{PC}_{y,i}$ where $y$ stands for the year between 2025 and 2050 and
+At the end of this step, a list of $m$ PCs is obtained, \ie $\text{PC}_{y,i}$ where $y$ stands for the representative year between 2025 and 2050 and
 

### 2024-06-13 rebuttals/rebuttals.tex
Rebuttals filled in with comments
... now is the time to address them

@@
 \maketitle
-%\emph{Start with some thank you note. For example:}
 
@@
 
-When a comment explicitly came from specific members of the jury, they are listed at the beginning of the comment with the following color code: {\color{orange} \textbf{Stefano} Moret}, {\color{teal} \textbf{Stefan} Pfenninger}, {\color{purple} \textbf{Sylvain} Quoilin} and {\color{violet} \textbf{Christophe} De Vleeschouwer}.
+When a comment explicitly comes from specific members of the jury, they are listed at the beginning of the comment with the following color code: {\color{orange} \textbf{Stefano} Moret}, {\color{teal} \textbf{Stefan} Pfenninger}, {\color{purple} \textbf{Sylvain} Quoilin} and {\color{violet} \textbf{Christophe} De Vleeschouwer}.
 Here is how an answer to a comment is structured:
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-Text of the comment.
+Text of the comment. When a page number is mentioned at the beginning of the comment, it refers to the page number of the manuscript submitted before the private defense, not its final version.
 \end{mdframed}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Overall, many methods: PCE + RL + NN + PCA. At each stage, some assumption is made. Why not a simpler approach or a simpler model? Do you need all these steps? 
+{\color{orange} \textbf{Stefano}} - Overall, many methods: PCE + RL + NN + PCA.
+At each stage, some assumption is made.
+Why not a simpler approach or a simpler model?
+Do you need all these steps?
+Explain the role of the different methods and why PCE instead of monthly version for uncertainty?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{violet} \textbf{Christophe}} - The variables and the uncertainties should be presented in a compact and mathematical way.
 \end{mdframed}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Many assumptions, e.g., alpha = 85\% (p.
+{\color{orange} \textbf{Stefano}}, {\color{purple} \textbf{Sylvain}} \& {\color{teal} \textbf{Stefan}} - Many assumptions, e.g., alpha = 85\% (p.
 40) or reward = -300 (Chapter 4).
@@
 How robust are they?
+Can the methodology be generalized to other models?
+Can we keep the same values for other applications?
+You need a model that runs pretty fast, right?
+\end{mdframed}
+
+\noindent Perhaps there is a need to better prove or show that it is not a magic number.
+How to generalise it and hook it to something physical (even if it is done a posteriori).
+Guidelines to apply it to some other models
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Make sure to discuss the hypotheses because one might not agree with all of them.
+So, it could provide more context but also some limitations and indications of the alternatives (and the spectrum of what is possible).
 \end{mdframed}
@@
 
-\subsection{Graphs}
-\label{general_graphs}
+\subsection{Graphs and results}
+\label{general_graphs_results}
 
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} \& {\color{teal} \textbf{Stefan}} - Graphs should be better explained. Many axes labels are missing, and the captions are often insufficient to read the figures. 
+{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 19, 22 and 24) should be better explained.
+Many axes labels are missing, and the captions are often insufficient to read the figures.
+In general, improve the presentation of the results, especially for the RL chapter.
 \end{mdframed}
@@
 \noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Work on the graph of slide 19 (there is a misunderstanding on the fact that the grey curve is labelled as failure but they are failures in the future)
+\end{mdframed}
+
+\noindent Make the link with the success rate (during the training).  To be discussed to keep most of the attemps even though they are discarded
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Some big topics are missing like storage. Why is storage not more discussed? 
+\end{mdframed}
+
+\noindent Storage is a by-product of the results and they’re barely affected by the different studies. It’s mostly gas storage (on average 18TWh) and, to a smaller extent, seasonal thermal storage.
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods?
+We could go further in the analysis as we know we have to phase out coal, install renewables.
+\end{mdframed}
+
+\noindent The outcome is a bit too obvious (e.g.
+getting rid of the coal) $\rightarrow$ go a bit in the details.
+Explain why the method is taking better decisions but explain why.
+$\rightarrow$ go beyond what we know
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - The thesis structure makes it quite difficult to follow.
+{\color{orange} \textbf{Stefano}} \& {\color{teal} \textbf{Stefan}} - The thesis structure makes it quite difficult to follow.
 Having all the methodology summarised in Chapter 1, makes it difficult to link it to the different chapters.
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-There are various typos in the thesis, e.g.
-``storagR'' (page 3), ``therefore, therefore'' (p.
-11), ``The'' (p.14), ``Moleclues'' (p.
-61), etc..
-Also, the use English language can be improved, e.g.
-``calls FOR a variety'' (p.
-9), ``cumulative emissions ARE'' (p.
-14), ``than to'' (p.
-34), etc.
-The document will benefit from a spelling/grammar check.
-\end{mdframed}
-
-\noindent
+\noindent Initially, I wanted to group methodology and results per kind of analysis, \ie UQ on Pathway (Chapter 3), RL (Chapter 4) and PCA (Chapter 5) but that would have made the RL, and even more PCA, methodology parts come too far in the manuscript.
+For this reason, I made the choice to group all the methodological approaches together in Chapter 1 and the results in their respective chapter.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - I found it unusual to see many references to own work (Rixhon et al.) throughout the thesis.
+{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - I found it unusual to see many references to own work (Rixhon et al.) throughout the thesis.
 Unless the articles are unrelated to the thesis, these references should be removed and the associated content should be included in the manuscript.
@@
 
-\subsection{Interest rate}
-\label{irate}
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - Wouldn't it be more appropriate to talk about ``discount rate'' rather than ``interest rate''?
+There are various typos in the thesis, e.g.
+``storagR'' (page 3), ``therefore, therefore'' (p.
+11), ``The'' (p.14), ``Moleclues'' (p.
+61), etc..
+Also, the use English language can be improved, e.g.
+``calls FOR a variety'' (p.
+9), ``cumulative emissions ARE'' (p.
+14), ``than to'' (p.
+34), etc.
+The document will benefit from a spelling/grammar check.
 \end{mdframed}
 
-\noindent Indeed, as we want to capture the time preference associated with obtaining finance for the project, effectively quantifying the trade-off between present and future financial value, it is more appropriate to refer to ``discount rate'' when considering the parameter $i_{\text{rate}}$.
-We have changed every occurrence of ``interest rate'' into ``discount rate'' {\color{blue}in the entire manuscript}.
+\noindent
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\subsection{Documentation}
+\label{documentation}
 
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - Differentiated discount rates could be useful? 
+{\color{purple} \textbf{Sylvain}} - Where is the repository for the code? Don't forget the documentation?
 \end{mdframed}
 
-\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''.
-Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
-These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
-(1.3) when discussing the annualised phase factor}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
-In practice, the discount rate would vary depending on the technology investment risk.
-Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
-For this reason, we have decided to keep an identical value of discount rate for all the technologies.
-\end{mdframed}
+\noindent As discussed during the private defense, this is a task that is already planned for the year to come, after the public defense.
+A better reference towards these repository and documentation will be given in the subsequent paper on RL.
+Consequently, there has not been further work done in this regard in the manuscript by the public defense.
 
@@
 
+\subsection{EnergyScope Pathway}
+\label{ESPathway}
 \begin{mdframed}[style=comment] % Comment from the reviewer
@@
 
+\subsubsection{End-of-time-horizon and salvage value}
+
 \begin{mdframed}[style=comment] % Comment from the reviewer
@@
 
-\noindent
+% \noindent Better highlight the importance, discuss the finetuning, discuss  alternatives (e.g. adding time windows at the end that are thrown afterwards).
 
@@
 \end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - The salvage value plays a large role in the decision of the model. Give more details to give more confidence in the choice.
+\end{mdframed}
+
+\noindent See what's been written in appendix paper Pathway
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsubsection{Myopic pathway}
 
@@
 \end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - With the myopic approach, are you predicting what will come or drawing a path that is the best?
+Is that a desirable feature to mimic the way we proceed now.
+why artificially limit the model as we know we have to go to zero in 2050.
+\end{mdframed}
+
+\noindent Link to salvage value
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - p.18 : Surprising that myopic leads to a sooner investment to the transition ?
+Is it an effect of the salvage value?
+was there a sensitivity analysis on this salvage value?
+Perhaps include more details on this.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsubsection{Discount rate and annualization}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Wouldn't it be more appropriate to talk about ``discount rate'' rather than ``interest rate''?
+\end{mdframed}
+
+\noindent Indeed, as we want to capture the time preference associated with obtaining finance for the project, effectively quantifying the trade-off between present and future financial value, it is more appropriate to refer to ``discount rate'' when considering the parameter $i_{\text{rate}}$.
+We have changed every occurrence of ``interest rate'' into ``discount rate'' {\color{blue}in the entire manuscript}.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Value of the discount rate is generally above 6\% in other works. You should have a better justification of selecting 1.5\%.
+\end{mdframed}
+
+\noindent  {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.13 - Why no annualizing over the lifetime of the unit? Would that lead to different results? 
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsubsection{Power grid and transport infrastructure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - How do you model the power grid interconnections with the neighbouring countries as well as the transport network of the different energy carriers within the country?
+\end{mdframed}
+
+\noindent Copper-plate assumption $\rightarrow$ no representation of the transport sector of either electricity or gaseous and liquid energy carriers.
+However, one technology called ``GRID'' accounts for the additional investments in the power grid needed when deploying more VRES like PV panels and wind turbines but there is not an actual spatial representation of this grid.
+{\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\subsection{Uncertainty quantification}
+\label{methodo_UQ}
 
@@
 
-\subsection{Uncertainty quantification}
-\label{methodo_UQ}
-
 \begin{mdframed}[style=comment] % Comment from the reviewer
@@
 23: is PCE an appropriate method for LP / MILP models.
-There is a good fit for the total cost, which is often quite linear, but how about technology choice / sizing?
+There is a good fit (1\% LLO error) for the total cost, which is often quite linear, but how about technology choice / sizing?
 \end{mdframed}
 
-\noindent
+\noindent Expand on this in the chapter while acknowledging the limitations and saying it is not the main topic of the thesis.
 
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.21 - How are the groups of uncertainties defined? Is there a table?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - Table 1.2 - what are the other parameters?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
 \subsection{Reinforcement Learning}
@@
 \noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 27 When presenting the advantages of RL, you mention model-free approach. Ok, but you have a representation of the real world available to train the model!
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Justification of the actions selected? why no other actions like the cost of some technologies? 
+\end{mdframed}
+
+\noindent These actions have a direct translation in constraints and we can therefore assess their effectiveness through the fact they are binding or not.
+Give examples of actions tried during the work (incentivize renewables) or other from Assemblée citoyenne in France and explain why they’re not as good as those chosen in the thesis.
+Explain history of what we tried/consider?
+ {\color{blue} }.
 
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 39 Why are their outliers in the first case? Isn't a bit arbitrary to set them to a min/max value?
+\end{mdframed}
+
+\noindent I should insist on those elements in the text {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
 \section{Chapter 2 - Case study}
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 45 How is the range accounted for the characterisation of private mobility technologies?
+\end{mdframed}
+
+\noindent Indeed, the increase of the range is a consequence of the increased efficiency and battery capacity.
+Consequently, I have removed the mention to the range {\color{blue}at the end of the last paragraph in the contributions of Chapter 2}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Regarding BEV, while the CAPEX has been kept unchanged, the efficiency and the battery capacity have been increased. 
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51 Offshore wind max capacity is too low and PV is lower than Bregilab
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51 Offshore wind max capacity is too low and PV is lower than Bregilab
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51 Why investigating SMR?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51 Last paragraph of p. 51 is not clear. Are we talking about the purchase cost or the CO2 content?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 53 4850€/kW is at the lower end of the EnergyVille scenario for SMR (4500€/kW).
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 53 What about fuel cost? This is a very optimistic price forecast! 
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 53 Do the parameters of the uncertainty analysis arrive later on?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 54 Is there a differentiation between ground mounted and rooftop PV?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.
+54 Current tenders for 3rd generation are higher than 100€/MWh.
+There are papers showing that SMRs are not really expected to be cheaper than 3rd generation.
+How realistic is 41 €/MWh?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 54 LCOE of SMRs would be much lower if: - the capital cost was more realistic and, - the Capacity factor was lowered to account for their flexible use.
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 55 - Figure 2.5 - Coupled with heat for endogenous and exogenous reactions?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 56 Are modal shares exogenously set?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 57 Why a uniform distribution? A gaussian woul clearly be more appropriate. Maybe some uncertainties are underestimated (because clipped by uniform distribution)
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\section{Chapter 3 - Atom-vs-molecules}
+\label{Chap_atom_vs_molecules}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Would it be fair to say that the scope of this Chapter is adding one technology (SMR) to the model? Is this a significant contribution?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - What is the learning of this chapter? I was underwhelmed by the long list of results. What is the take-away?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 70: ``in other words, the variation of the parameter f$\_$max,SMR…''. I am not sure that this sentence is equivalent to the previous one. Please clarify.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 71, Table 3.1: isn’t the low impact simply due to the low assumed potential? Is this finding meaningful?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 74, Figure 3.8: this graphs looks potentially interesting but it is not sufficiently well explained.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 63 What does efficiency change in the installed capacity?
+\end{mdframed}
+
+\noindent Indeed, the efficiency only has an impact on the amount of primary uranium needed to produce the electricity.
+Consequently, I have removed this part of the sentence {\color{blue}in the first paragraph of Section 3.1.1}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Overall, given the restriction on yearly availability and the slightly higher electrification [...]
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 64 - 15 GW of PV in 2025. Realistic?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 65 - Decides to use ammonia instead of SNG => is the transport infrastructure accounted for?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 67 - It really depends on what is considered primary energy in the case of nuclear
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 69 - No curtailment at all? It would be nice to have the exact number.
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 69 - Very strange. Should be the same as for industrial boilers. No smart charging?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.71 - Differentiated discount rates could be useful.
+\end{mdframed}
+
+\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
+(1.3) when discussing the annualised phase factor}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
+In practice, the discount rate would vary depending on the technology investment risk.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+For this reason, we have decided to keep an identical value of discount rate for all the technologies.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 72 - Battery trucks are gaining momentum. Is the hydrogen route really the way to go?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 73 - How is the transport infrastructure accounted for?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 78 - Why this threshold? Shouldn't it follow the potential capacity?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Lot of the results hinge on the import of e-fuels with the assumption that they have GWP = 0; is it justified? Those fuels are directly available knowing they are no.
+\end{mdframed}
+
+\noindent Perhaps reinforce the message as you answered the question $\rightarrow$ We need to go for radical changes, a “unicorn” solution, and one option would be the import of renewable molecules {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - SMR / interest rate must be revisited if a paper is published.
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
 \section{Chapter 4 - Reinforcement Learning}
@@
 
-\subsection{Binding}
-\label{RL_binding}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Explain more in details the different steps in the RL approach
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 84: Why -300? Seems like fine-tuned to obtain desired results?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+85, Figure 4.1: Does it mean we can run ``infinite energy transitions''?
+Is this realistic?
+Isn’t the ``learning'' supposed to happen during the different stages?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - How can you account for pivotal events / shocks in the system?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+89: Figure 4.3 seems to deliver an important message (early action is needed to ensure we meet climate goals), but it is not clearly explained.
+Improve captions / add axes.
+	Where do I see the ``tipping points'' in this figure?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Overall, I suggest adding a diagram to Chapter 4, detailing what are the actions available to the decision-maker.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - How does this approach compare, for example, to stochastic programming?
+\end{mdframed}
+
+\noindent This work will be done for the subsequent paper on RL.
+The strategy would to implement a simplified case study where it would be ``computationally affordable'' to test the stochastic programming approach and compare it with the optimal policy delivered by the RL approach.
+Consequently, regarding this comment, there has not been further modification brought to the manuscript.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.91, Table 4.2: how do you differentiate between actions that depend on the agent and exogenous uncertainties?
+More in general: how is uncertainty integrated in the RL framework?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+94: Figure 4.6 seems to be less revealing than other figures.
+Also the figure caption seems to contradict the first statement following the figure.
+It would be important to clarify: which are the actions that emerge to actually matter?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?
+\end{mdframed}
+
+\noindent {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 87 - What is a step?
+\end{mdframed}
+
+\noindent To define in Section 1.3.2 and remind here {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 96 - The early constraints on generation avoid lock-in situations correct?
+\end{mdframed}
+
+\noindent To define in Section 1.3.2 and remind here {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
+\end{mdframed}
+
+\noindent To explain in the definition of the action {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - p.
+97 - The agent can reach systems that are cheaper than any other solution obtained by the perfect foresight approach.
+How can this be possible?
+\end{mdframed}
+
+\noindent To explain in the definition of the action {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 100 - What/where are the no-go zones? Is there a narrative to describe them ?
+\end{mdframed}
+
+\noindent  {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
 
@@
 
-\section{Principal Component Analysis}
+\section{Chapter 5 - Principal Component Analysis}
 \label{PCA}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 101: you mention different definitions of “robustness”: which one do you use in your work?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 109: how is Gamma-robustness implemented here? It seems like it is in some sense “manually” forced? Explain better how the ROB solution is derived.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
 
@@
 
+\section{Conclusion}
+\label{Conclusion}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - p.
+119: the statement mentioning the ``major added value of the perfect foresight'' approach seems surprising and needs to be clarified.
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - The conclusion is a summary and should rather be a discussion (\eg discard coal as soon as possible is obvious).
+\end{mdframed}
+
+\noindent To explain in the definition of the action {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
 \clearpage

### 2024-06-13 chap_atom_mol/chap_atom_mol.tex
Rebuttals filled in with comments
... now is the time to address them

@@
 To a lesser extent, the last 2\% deployment of solar-\gls{PV} is slightly delayed as the capacity in 2025 is 1.3\,GW smaller than in the REF case.
-Overall, given the smaller efficiency of \gls{SMR}, \ie 40\% versus 51\% for ammonia-\gls{CCGT}, the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.
+Overall, given the restriction on yearly availability and the slightly higher electrification (\Cref{fig:results_deter_layer_elec}), the total power capacity installed by 2050 is 3.5\% higher for the SMR case.
 In their work, \citet{PATHS2050} also showed that \gls{SMR} first substitutes ``e-fuels/hydrogen'' turbines before reducing the need for \gls{PV}.

### 2024-07-05 rebuttals/rebuttals.tex
Rebuttals: Keep going...

@@
 \usepackage{pifont} 
+\usepackage[version=4]{mhchem} % Chemical equations
 
@@
 
-\graphicspath{{../frontend/img/}{../chap_intro_ccl/img/}{../chap_case_study/img/}{../chap_ES_PCE/img/}{../chap_electro_uq/img/}{../chap_methodo/img/}{../chap_myopic/img/}{../chap_RL/img/}{../chap_atom_mol/img/}{../chap_RobPol/img/}{../appendices/img/}} % Figures folder different for each input
+\graphicspath{{../frontend/img/}{../chap_intro_ccl/img/}{../chap_case_study/img/}{../chap_ES_PCE/img/}{../chap_electro_uq/img/}{../chap_methodo/img/}{../chap_myopic/img/}{../chap_RL/img/}{../chap_atom_mol/img/}{../chap_RobPol/img/}{../appendices/img/}{../rebuttals/img/}} % Figures folder different for each input
 %\graphicspath{{img/}} %
@@
 
+\newcommand{\clearemptydoublepage}{\newpage{\pagestyle{empty}\cleardoublepage}} %pour effacer les en-tetes sur la page vierge avant chaque chapitre
+
 %Palatino font
@@
 \usepackage{tocloft}
-\renewcommand{\cftsecpresnum}{Reviewer \#}
+
+%\renewcommand{\cftsecpresnum}{Reviewer \#}
+\renewcommand{\cftsecpresnum}{}
 \renewcommand{\cftsecnumwidth}{6em}
-\renewcommand{\cftsubsecpresnum}{Comment \#}
+\renewcommand{\cftsubsecpresnum}{}
 \renewcommand{\cftsubsecnumwidth}{6em}
 
+\setcounter{tocdepth}{3}
+\setcounter{secnumdepth}{3}
+
+\usepackage[nottoc]{tocbibind}
 \usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
@@
 }
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-Text of the comment. When a page number is mentioned at the beginning of the comment, it refers to the page number of the manuscript submitted before the private defense, not its final version.
+Text of the comment.
+When a page number is mentioned at the beginning of the comment, it refers to the page number of the manuscript submitted before the private defense, which might differ from the one in its final version.
 \end{mdframed}
@@
 \clearpage
+%\tableofcontents
+%
+%\clearemptydoublepage
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\noindent In my opinion, the main variables of EnergyScope Pathway are presented {\color{blue} in Section 1.1 of the core of the manuscript}.
+Further details are also given {\color{blue}in Appendix B.2}.
+Finally, these variables are even exhaustively detailed in \cite{limpens2019energyscope} and \cite{limpens2024pathway} that are cited several times in the thesis manuscript.
+Regarding the uncertainties, the way the ranges are defined and the way they affect the value of some parameters in the model are explained {\color{blue}in Section 1.2 of the core of the manuscript}.
+Consequently, no modification has been brought to the manuscript regarding this topic.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 19, 22 and 24) should be better explained.
-Many axes labels are missing, and the captions are often insufficient to read the figures.
+{\color{orange} \textbf{Stefano}} - Work on the graph of slide 19 (there is a misunderstanding on the fact that the grey curve is labelled as failure but they are failures in the future)
+\end{mdframed}
+
+\noindent To clarify this point, I have extended the caption of the related figures (see previous comments) and added two sentences in {\color{blue} Section 4.2.2}:
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+Since infeasible cases or those that overshoot the \ce{CO2}-budget are discarded before reaching 2050 (see Section 4.1.2), the number of attempts that reach further steps in the transition progressively decreases.
+Consequently, the share of successful transitions compared to failures progressively increases with time.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained.
 In general, improve the presentation of the results, especially for the RL chapter.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Work on the graph of slide 19 (there is a misunderstanding on the fact that the grey curve is labelled as failure but they are failures in the future)
-\end{mdframed}
-
-\noindent Make the link with the success rate (during the training).  To be discussed to keep most of the attemps even though they are discarded
+{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Many axes labels are missing, and the captions are often insufficient to read the figures. 
+\end{mdframed}
+
+\noindent I agree and have gone through all the figures of the manuscript to clarify them by adding axes labels or extending the caption where it was needed.
+Figures where y-axis (and its label) is missing are distribution (or density) plots (\ie Figures 3.6, 4.4, 4.5, 4.6, 4.8 and E.1).
+Consequently, similarly to an histogram, the feature of interest and its span of values are represented on the x-axis.
+The y-axis has no real interest in terms of information.
+What matters is the shape of the curve to be able to point out for which values/ranges of values, the feature of interest occurred more or less often.
+For this reason, I have not added y-axis for the aforementioned figures but rather improved their respective caption, emphasising, among other things, that these figures represent distribution of occurrences.
+For instance, the caption of {\color{blue}Figure 4.4} is now more explicit:
 
 \begin{mdframed}[style=manuscript] % Comment from the reviewer
-
+Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
 \end{mdframed}
@@
 
-\noindent Storage is a by-product of the results and they’re barely affected by the different studies. It’s mostly gas storage (on average 18TWh) and, to a smaller extent, seasonal thermal storage.
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\noindent As discussed during the private defense, storage is not more discussed in the manuscript for two reasons.
+First, the storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}).
+Analyses under uncertainties have not shown major variations neither.
+Second, storage assets are intrinsically linked to the conversion technologies and, consequently, can be seen as by-products of the optimisation results.
+For these two reasons, I have rather focused my analyses on the use of resources and the installation and operation of conversion technologies.
+Given the discussion during the private defense and these further explanations, I have not brought modification to the manuscript in this regard.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.7\textwidth]{STORAGE_2.pdf}
+\caption{Storage installed capacities for the REF case}
+\label{fig:STORAGE_2}
+\end{figure}
 
@@
 \noindent Initially, I wanted to group methodology and results per kind of analysis, \ie UQ on Pathway (Chapter 3), RL (Chapter 4) and PCA (Chapter 5) but that would have made the RL, and even more PCA, methodology parts come too far in the manuscript.
-For this reason, I made the choice to group all the methodological approaches together in Chapter 1 and the results in their respective chapter.
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+For this reason, I made the choice to group all the methodological approaches together in Chapter 1 and the results in their respective chapters.
+Given this explanation, I have added one sentence {\color{blue}at the end of the introductory paragraph of the RL method in Chapter 1}:
+
+\begin{mdframed}[style=manuscript] % Modification to manuscript
+As Chapter 2 presents the input of the case study and Chapter 3 details the results of the GSA on the pathway model, the reader interested by the results of the RL method is invited to go to Chapter 4.
+\end{mdframed}
+
+\noindent and {\color{blue}at the end of the introductory paragraph of the PCA-based method in Chapter 1}:
+
+\begin{mdframed}[style=manuscript] % Modification to manuscript
+The reader interested by the results of the PCA-based method is invited to go to Chapter 5.
+\end{mdframed}
+
+\noindent Finally, I have added a schematic at the end of the Introduction of the thesis to clarify the structure and highlight that Chapter 4 and 5 can be read after the methodological sections 1.3 and 1.4, respectively (see Figure \ref{fig:intro:Thesis_Structure}).
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.7\textwidth]{Thesis_Structure.pdf}
+\caption{Structure of the thesis.
+Beyond Chapter 2 that describes the case study of Belgium, Chapters 3, 4 and 5 collect the analyses resulting from the application of the different methodologies developed in Chapter 1.
+Given the (quasi-)independence of the analyses presented in the last three chapters, they can be read separately.}
+\label{fig:intro:Thesis_Structure}
+\end{figure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{violet} \textbf{Christophe}} - Formal description of the problem is missing even in described in the text to make the link between the ideas and the implementation.
+Add some flow charts to introduce key elements
+\end{mdframed}
+
+\noindent This point is addressed with the structure chart presented in the previous comment as well as the improvement of Figure 1.4 presenting the implementation of the myopic approach as well as the way uncertainties and RL change the values of some parameters (see Figure \ref{fig:MY_process_code} of the present document).
 
@@
 
-\noindent
+\noindent This choice to sometimes refer to own works was made to focus only on the main contributions of the thesis.
+However, I agree that some content should be included in the thesis to enhance its reading.
+I hereby list the references to works I have been involved in (with the reference number of the manuscript version submitted before the private defense) and detail the potential modifications brought to the manuscript for each of them.
+
+\begin{itemize}
+\item \emph{[40] Limpens et al., Energyscope pathway: an open-source model to optimise the energy transition pathways of a regional whole-energy system}: Mostly cited to give credits to Gauthier Limpens' work in developing the Pathway version of EnergyScope, I have decided to refer to the paper about the formulation choices rather than including them in the thesis manuscript.
+Consequently, no further modification regarding this reference.
+\item \emph{[48] Limpens et al., The impact of uncertainties on the Belgian energy system: Application of the Polynomial Chaos Expansion to the EnergyScope model}: The table gathering the results of the comparison between Morris' and PCE methods has been included {\color{blue}at the end of Section 1.2.2}:
+\end{itemize}
 
 \begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+In \cite{limpens2020impact}, we have assessed the PCE approach on the cost-optimum Belgian energy system in 2035, using EnergyScope TD.
+To do so, we compared the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$.
+Even if the output of each method does not have the same physical meaning, both methods can rank the parameters by their impact on the total annual cost of the energy system.
+Both rankings were very similar which validated the use of PCE in the rest of this work (see Table \ref{tab:GSA:Comparison}).
+\end{mdframed}
+
+\begin{table}[htbp]
+	\caption{Comparison of the Top-14 rankings for the improved Morris method (left) and total-order PCE method (right).
+We used the DTU's implementation of Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$).
+}
+    \label{tab:GSA:Comparison}
+    \centering
+    \begin{tabular}{lcc}
+    \hline
+		\textbf{Parameter} & \textbf{Morris Ranking} ($\mu^*_{i}$) & \textbf{PCE Ranking} ($S_i^T$) \\
+        \hline
+        Prices of hydrocarbons & 1 (0.7141) & 1 (0.6434) \\
+        Cost of increased efficiency & 2 (0.3046) & 2 (0.1014) \\
+        Cost of maintenance & 3 (0.2371) & 4 (0.0534) \\
+        Discount rate & 4 (0.2019) & 5 (0.0467) \\
+        Price of imported electricity & 5 (0.1974) & 3 (0.0809) \\
+        CAPEX of the grid & 6 (0.1498) & 6 (0.0252) \\
+        CAPEX of PV & 7 (0.1270) & 7 (0.0187) \\
+        Increase of electricity demand & 8 (0.1168) & 8 (0.0154) \\
+        CAPEX of nuclear power plant & 9 (0.0963) & 9 (0.0119) \\
+        Increase of space heating demand & 10 (0.0856) & 11 (0.0088) \\
+        Price of coal & 11 (0.0856) & 12 (0.0087)\\
+        Price of renewable fuels & 12 (0.0794) & 10 (0.0104)\\
+        Price of uranium & 13 (0.0607) & 13 (0.0047)\\
+        Onshore wind load factor & 14 (0.0600) & 14 (0.0046)\\
+       \hline
+    \end{tabular}
+\end{table}
+
+\begin{itemize}
+\item \emph{[49] Rixhon et al., The role of electrofuels under uncertainties for the Belgian energy transition}: This paper is cited to highlight the GSA performed on the snapshot model with different GWP limits.
+The main contribution of this work in the context of the thesis was to provide a preliminary screening and selection of parameters to carry out the GSA on the pathway model.
+The result of this selection is given in Section 2.4 where the uncertain parameters and their respective ranges are listed.
+Consequently, no further modification regarding this reference.
+\item \emph{[104] Rixhon et al., Integration of non-energy among the end-use demands of bottom-up whole-energy system models}: Reference used to highlight the fact that whole-energy systems usually do not account for the non-energy demand.
+Consequently, no further modification regarding this reference.
+\item \emph{[116] Rixhon et al., Comprehensive
+integration of the non-energy demand within a whole-energy system: towards a defossilisation of the chemical industry in Belgium}: Reference to graphs that have either been adapted from the paper or directly taken from it.
+Consequently, no further modification regarding this reference.
+\item \emph{[128] Rixhon et al., Taxonomy of the fuels in a whole-energy system}: Where the content of this paper is the Appendix A of the thesis, some references to this paper aimed at clarifying the meaning of ``renewable fuels'' (Section 2.2) or the fact that uranium is considered neither fossil nor renewable (Sections 3.1.2 and 4.2.2).
+Consequently, no further modification regarding this reference.
+\end{itemize}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} -While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc.
+{\color{orange} \textbf{Stefano}} - While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc.
 I recommend reducing them to the bare minimum, or completely getting rid of them.
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\noindent The listed typos have been corrected and, overall, the manuscript has gone through Grammarly to improve the spelling and the grammar.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Can you explain the logic of the annualization factor in the pathway model and the choice of N=5 for the uncertainty ranges?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\subsubsection{End-of-time-horizon and salvage value}
+\noindent To decrease the computational time by one order of magnitude, besides opting for Gurobi to solve the LP, I disabled crossover.
+Besides this significant reduction of computational time, I observed no other impact regarding the values of the objective function or the design variables at the optimum.
+As disabling crossover is a very LP-specific aspect, I added this information {\color{blue}in a footnote in the first paragraph of Section 1.1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+To keep this low computational time, crossover was disabled to solve the optimisation problem.
+This had no observed impact regarding the values of the objective function or the design variables at the optimum.
+\end{mdframed}
+
+\subsubsection{End-of-time-horizon}
 
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - The salvage value plays a large role in the decision of the model. Give more details to give more confidence in the choice.
-\end{mdframed}
-
-\noindent See what's been written in appendix paper Pathway
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\subsubsection{Myopic pathway}
+\subsubsection{Myopic pathway and salvage value}
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Even though this sequence of LPs was already mentioned {\color{blue}in the caption of Figure 1.3}, I have complemented the text {\color{blue} in the paragraph ``Myopic pathway implementation'' of Section 1.1.2},
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+[...] The myopic optimisation of the pathway consists in a sequence of smaller LP problems limited to their respective 10-year time-window, until eventually reaching 2050.
+\end{mdframed}
+
+\noindent reminded it in the caption of Figure 1.2
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+The myopic approach (in pink) uses several instances of the pathway model, illustrated in Figure 1.1, as a sequence of different and linked LP problems. 
+\end{mdframed}
+
+\noindent and extended the caption of {\color{blue} Figure 1.4}.
+However, besides highlighting the key step 3.3 where there is an update of parameters value, I have decided to keep the figure as is since, on the one hand, the extension of the caption provides now a better understanding and, on the other hand, the code itself is exactly what is schematically represented on the diagram.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=10cm]{MY_process_code.pdf}
+\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
+Myopic optimisation of the pathway consists in a sequence of smaller LP problems limited to their respective 10-year time-window.
+If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters are changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
+\label{fig:MY_process_code}
+\end{figure}
 
@@
 Perhaps include more details on this.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\subsubsection{Discount rate and annualization}
+The salvage value plays a large role in the decision of the model.
+Give more details to give more confidence in the choice.
+\end{mdframed}
+
+\noindent Indeed, it is an effect of the salvage value as already written in the last paragraph ``Impact of myopic formulation on the system'' of Section 1.1: ``The main difference lies in the myopic transition itself and especially in the earlier
+deployment of PVs and offshore wind turbines.
+These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset.
+This is mostly due to impact of the salvage value, Equation (1.4), in the objective function.''.
+
+Sensitivity analysis has been performed on the formulation of the salvage value by \citet{goffauxpathway}.
+The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon.
+The methodology and results of this analysis is now summed up {\color{blue}in Appendix B.2}
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+In the work of \citet{goffauxpathway}, we carried out a sensitivity analysis on the formulation of the salvage value.
+First and foremost, this analysis confirmed the need to account for the salvage value to avoid penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}.
+Then, the sensitivity analysis focused on two elements in the expression of the salvage value: $\tau_{phase}(p)$ and $\textbf{F\textsubscript{decom}}(p2,p,i)$.
+
+In Equation (B.22), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
+With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost.
+However, one could wonder if a technology that is still operational for half of its lifetime is not worth less than half of its initial investment cost, given the depreciation of the asset.
+In that case, the annualisation factor should account for the last phase of the optimisation (\ie 2045\_2050) when the residual value of the technologies are calculated and subtracted from the total transition cost.
+
+On top of this, we also investigated the impact of accounting for the decommissioned technologies or not in the expression of the salvage value.
+In Equation (B.22), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0.
+This is relevant since a technology decommissioned before 2050 would not be available after 2050.
+The issue is that the model could keep unnecessary technologies to subtract their salvage value from the total transition cost in order to decrease it.
+This would be done if the fixed OPEX of an unused technology is lower than its salvage value.
+Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value.
+In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a salvage value.
+
+The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon. 
+\end{mdframed}
+
+\noindent and a reference to this appendix is present {\color{blue}in Section 1.1 where Equation (1.4) is detailed}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+After assessing the sensitivity of the formulation of the salvage value (see Appendix B.2), we have kept the one given by Equation (1.4), being also similar to the expression used by \citet{prina2019transition}.
+\end{mdframed}
+
+\subsubsection{Discount rate and annualisation}
 
@@
 
-\noindent  {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p.13 - Why no annualizing over the lifetime of the unit? Would that lead to different results? 
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Indeed, the discount rate is usually between 7.5\% and 12\% in other studies.
+The main justification for this choice was to be in line with G.
+Limpens' work \cite{limpens2021generating}.
+However, assessing the impact of $i_{\text{rate}}$ on $\tau_{phase}$, we see that a lower value of $i_{\text{rate}}$ will favour less late expenses compared to a high value.
+In other words, considering $i_{\text{rate}}=$1.5\% rather than 10\% will less postpone investments (see Figure \ref{fig:tau_phase}).
+I have added some explanations in this regard {\color{blue}in Section 1.1.1}:
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=10cm]{tau_phase.png}
+\caption{$\tau_{phase}$ versus the timing of the expenses (columns) and the discount rate, $i_rate$ (rows).}
+\label{fig:tau_phase}
+\end{figure}
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+In EnergyScope Pathway, $\tau\textsubscript{\emph{phase}}$ aims at depreciating the value of expenses done in the future.
+The discount rate accounted for in this annualisation factor, $i_{\text{rate}}$, is considered as identical for all the technologies of the system.
+In line with \citet{limpens2021generating}, we set the nominal value to $i_{\text{rate}}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}.
+Having a low value of discount rate favours less late expenses compared to a higher value of discount rate.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.13: Why no annualizing over the lifetime of the unit? Would that lead to different results?
+
+{\color{orange} \textbf{Stefano}} - Can you explain the logic of the annualization factor in the pathway model?
+\end{mdframed}
+
+\noindent Accounting for the depreciation of money with time (see previous comment), we considered that the assets are fully paid at the moment they are installed.
+For this reason, we consider the arithmetic average between the investment cost of the assets at the year starting the phase of installation and the year ending it.
+This choice was done for two main reasons: potential decommissioning and lack of data beyond 2050.
+In the Pathway version of EnergyScope, the asset will be used over its whole lifetime unless it is prematurely decommissioned.
+Therefore, annualising over the entire lifetime of the unit might induce a bias in case of this premature decommissioning.
+Then, for assets with longer lifetime and/or installed later in the transition, annualising over the lifetime would require to have data beyond 2050.
+Effort that I have not decided to make.
+
+Regarding the impact on the results, my educated guess is that this impact would be limited as the technologies of a same sector are assumed to have similar trends in terms of the evolution with time of their respective investment costs.
+Consequently, the major impact would be to end up with a lower total transition costs but the design of the system would be marginally impacted.
+
+Given these further explanations, I have not brought any modification to manuscript in this regard.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 20: “inspired by Guevara et al.” $\rightarrow$ how was their approach used in your work?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+{\color{orange} \textbf{Stefano}} - Can you explain the choice of N=5 for the uncertainty ranges?
+\end{mdframed}
+
+\noindent
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 20: ``inspired by Guevara et al.'' $\rightarrow$ how was their approach used in your work?
+\end{mdframed}
+
+\noindent \citet{guevara2022modeling} implemented a stochastic process to maintain realism in the evolution of uncertainty through the transition.
+In this sense, they avoid a parameter to go from the upper bound at year $y$ to the lower bound 5 years later.
+It allows zigzag from one time step to another.
+(see Figure \ref{fig:Guevara}).
+In my case, once the sample of uncertain parameters is drawn, there is no zigzag ($\alpha=0$ in \cite{guevara2022modeling}) but without especially be at the nominal value.
+Besides this, the inspiration does not go further.
+Given this explanation, I have not brought modification to the manuscript in this regard.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.5\textwidth]{Guevara.png}
+\label{fig:Guevara}
+\end{figure}
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+\noindent The footnote was intentionally qualitative as the quantitative information came later on (in Section 1.2.2).
+To improve this, I have moved the content of this footnote into the core of the text and after this quantitative information has been brought {\color{blue}in Section 1.2.2, after ``to achieve a LOO error below 1\% for the total transition cost.''}
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+As second order PCE is the minimum to ensure accuracy of the surrogate model, selecting and grouping the parameters as presented in Section 1.2.3 is compulsory to alleviate the computational burden.
+Otherwise, considering thousands of independent uncertain parameters would lead to millions of EnergyScope Pathway runs, if no more.
+Another alternative to reduce the computational cost of the GSA would be to consider the monthly (instead of hourly) time resolution for the runs of EnergyScope Pathway.
+However, despite its shorter computation time (\ie couple of seconds), the main drawback of the monthly model is its poor representation of the integration of VRES \cite{limpens2024pathway}.
 
@@
 23: is PCE an appropriate method for LP / MILP models.
-There is a good fit (1\% LLO error) for the total cost, which is often quite linear, but how about technology choice / sizing?
-\end{mdframed}
-
-\noindent Expand on this in the chapter while acknowledging the limitations and saying it is not the main topic of the thesis.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+Why is PCE needed is EnergyScope is already very fast?
+There is a good fit (1\% LOO error) for the total cost, which is often quite linear, but how about technology choice / sizing?
+\end{mdframed}
+
+\noindent Like in Diederik Coppitters' works \cite{coppitters2021robust,coppittersthesis}, the ``ultimate'' objective of PCE is to build up a surrogate model, $\hat{M}$, to serve a robust design optimisation (RDO) process.
+For an original non-linear model $M$, this surrogate model, formed of as a series of polynomials,  has the main advantage to run much faster than the original model.
+In my case, I used PCE as an uncertainty quantification (UQ) tool without going up to the RDO step.
+As written {\color{blue}in the first paragraph of Section 1.2.2}, I (only) aim at getting the statistical moments of the quantity of interest and determine Sobol' indices.
+Similarly, in Chapter 5, I used PCA to extract main direction of variation when it comes to the design of the system without performing runs using these PCs that are another representation of the original model.
+Consequently, regarding this objective, PCE is an appropriate method for EnergyScope Pathway.
+
+\noindent When the LOO error is higher than 1\% (but limited to 20-25\%), which is the case when performing the PCE on technology size/choice (\ie installed capacity of SMR by 2050) or the import of renewable molecules (see Section 3.2.2), the approach remains valid to qualitatively rank the parameters, {\color{blue}as discussed in the paragraph ``Qualitative ranking based on Sobol' indices'' of Section 1.2.2}.
+
+\noindent Given these explanations and the elements already present in the text, I have not brought further modifications in the manuscript.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Why is PCE needed is EnergyScope is already very fast?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent This comparison with a similar approach like factor fixing and factor prioritization (based on Morris method \cite{morris_factorial_1991}) is detailed {\color{blue}in the paragraph ``Comparison with a proven method'' of Section 1.2.2}.
+I have extended this paragraph by integrating the table of comparison between the Morris method and PCE.
+This also avoids the reference to our own work and the reference [48], page 25 (see previous comment).
+
+\noindent Besides being an in-house used method, the choice to use PCE is to be able to quickly extract statistical moments of the output of interest, as explained {\color{blue}in the first paragraph of Section 1.2.2}.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent Even though the uncertain parameters, their groups and their ranges of uncertainties might be the same for other case studies, they have been especially adapted for the present case study, \ie Belgium.
+Consequently, I decided to exhaustively list these uncertain parameters in the chapter dedicated to the case study (see Section 2.4).
+However, to not leave the reader handing there and wondering if there other parameters and how we have decided to group them, I have added a reference to Section 2.4 {\color{blue}at the end of Section 1.2.1} focusing on the characterisation of the uncertainties:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Section 2.4 gathers the list of the uncertain parameters considered in this work.
 \end{mdframed}
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent See previous comment.
 
@@
 \section{Chapter 2 - Case study}
-\label{methodo}
+\label{case_study}
+
+\subsection{General and introduction}
+\label{methodo_general}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 45 How is the range accounted for the characterisation of private mobility technologies?
+\end{mdframed}
+
+\noindent Indeed, the increase of the range is a consequence of the increased efficiency and battery capacity.
+Consequently, I have removed the mention to the range {\color{blue}at the end of the last paragraph in the contributions of Chapter 2}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Regarding BEV, while the CAPEX has been kept unchanged, the efficiency and the battery capacity have been increased. 
+\end{mdframed}
+
+\subsection{End-use demands}
+\label{methodo_eud}
+
+\subsection{Resources}
+\label{methodo_resources}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51 Last paragraph of p. 51 is not clear. Are we talking about the purchase cost or the CO2 content?
+\end{mdframed}
+
+\noindent There was indeed a confusion in the information to deliver in this paragraph.
+For the sake of clarity, I have restructured Section 2.2.
+After the introductory paragraph, the graphs and the following paragraph focus on the cost of purchasing.
+The two last paragraphs address the aspects of availabilities and global warming potentials, respectively.
+
+\subsection{Conversion technologies}
+\label{methodo_technologies}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{orange} \textbf{Stefano}} - Figure 2.5: What is the purpose of the violet line between ``biomass gasification'' and ``methanolation''?
@@
 \noindent These were actually two different arrows, one pointing from ``biomass gasification for CH3OH'' to ``Methanol'' and the other from ``Methanolation'' to ``Methanol''.
-To avoid this confusion, the figure has been updated and the origin of these arrows are now on different vertical levels {\color{blue} see Section 2.3}.
+To avoid this confusion, the figure has been updated and the origin of these arrows are now on different vertical levels ({\color{blue} see Section 2.3}).
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 45 How is the range accounted for the characterisation of private mobility technologies?
-\end{mdframed}
-
-\noindent Indeed, the increase of the range is a consequence of the increased efficiency and battery capacity.
-Consequently, I have removed the mention to the range {\color{blue}at the end of the last paragraph in the contributions of Chapter 2}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Regarding BEV, while the CAPEX has been kept unchanged, the efficiency and the battery capacity have been increased. 
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 51 Offshore wind max capacity is too low and PV is lower than Bregilab
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 51 Offshore wind max capacity is too low and PV is lower than Bregilab
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+{\color{purple} \textbf{Sylvain}} - p. 54 Is there a differentiation between ground mounted and rooftop PV?
+\end{mdframed}
+
+\noindent In the continuity of Gauthier Limpens' works \cite{limpens2021generating}, we consider here only rooftop PV.
+This potential of 59.2\,GW corresponds to 250 km$^2$ of available roof well oriented that exist today.
+Accounting for ground mounted PV would almost double this potential, as shown in Paolo Thiran's thesis.
+In future works, it would be interesting to also implement this additional potential, especially for the sake of harmonisation of input data.
+On the contrary, Elia, the Belgian TSO, foresees a much more limited increase of installed PV, \ie 18\,GW by 2034 \cite{Elia_2024_2034}.
+Consequently, for this thesis, I have decided to stick to this intermediate potential of 59.2\,GW.
+For the sake of clarity, I only added ``rooftop'' {\color{blue}in the second paragraph of Section 2.2}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of rooftop PV panels (59.2\,GW) [...]
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 51: Offshore wind max capacity is too low and PV is lower than Bregilab.
+\end{mdframed}
+
+\noindent Concerning the potential of PV, see previous comment.
+Regarding offshore wind, we have assumed a conservative potential of 6\,GW as the lower bound of the forecast done by the Belgian offshore platform\footnote{https://www.belgianoffshoreplatform.be/fr/}.
+Similarly to EnergyVille \cite{PATHS2050}, we could have gone up to the upper bound of 8\,GW.
+Going much higher, such as 24\,GW for the ``Electrification'' scenario of EnergyVille, would account for access to a part of the 150\,GW offshore wind park project in the North Sea signed betwee Belgium, Germany, Denmark and the Netherlands (the ``Esbjerg Offshore Wind Declaration'').
+
+\noindent Since these potentials have been discussed and detailed in previous works and are not the main topic of my thesis, I have not brought any further modification to the manuscript in this regard.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 51 Last paragraph of p. 51 is not clear. Are we talking about the purchase cost or the CO2 content?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 53 4850€/kW is at the lower end of the EnergyVille scenario for SMR (4500€/kW).
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 53 What about fuel cost? This is a very optimistic price forecast! 
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 53 Do the parameters of the uncertainty analysis arrive later on?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 54 Is there a differentiation between ground mounted and rooftop PV?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent As discussed during the private defense, we have decided to investigate to role of nuclear energy in the future for two main reasons: hot topic in Belgium and similar study by EnergyVille \cite{PATHS2050}.
+First, when presenting the results of our work at conferences/meetings, we have been asked many times how our results, massively relying on imports of renewable molecules, were influenced by the installation of nuclear power plants by 2050.
+Second, during my thesis, EnergyVille delivered a report PATHS2050 where, among others, they investigated the impact of SMR.
+As they observed that installing SMR would significantly reduce the need to import renewable molecules, we also wanted to assess this impact in our work.
+Given these explanations, no further modification has been brought to the manuscript in this regard.
 
@@
 {\color{purple} \textbf{Sylvain}} - p.
-54 Current tenders for 3rd generation are higher than 100€/MWh.
+53 4850€/kW is at the lower end of the EnergyVille scenario for SMR (4500€/kW).
+This is a very optimistic price forecast!
+Current tenders for 3rd generation are higher than 100€/MWh.
 There are papers showing that SMRs are not really expected to be cheaper than 3rd generation.
 How realistic is 41 €/MWh?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 54 LCOE of SMRs would be much lower if: - the capital cost was more realistic and, - the Capacity factor was lowered to account for their flexible use.
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+LCOE of SMRs would be much higher if: - the capital cost was more realistic and, - the Capacity factor was lowered to account for their flexible use.
+\end{mdframed}
+
+\noindent As discussed during the private defense, similarly to EnergyVille \cite{PATHS2050}, we have considered the same CAPEX for SMR as the one considered for conventional nuclear power plants.
+Where EnergyVille considers a nominal CAPEX of 7500€/kW (like the current large Gen III design of Hinkley Point in the UK), we considered 4850€/kW like the CAPEX of conventional nuclear units in EnergyScope.
+
+On top of the uncertainty applied to the CAPEX of SMR [-40\%, 44\%] considered in the thesis, I carried out a test on these assumptions by plotting the LCOE of electricity-production units considering different assumptions for SMR (see Figure \ref{fig:LCOE_comparison}).
+Even considering 50\% more expensive SMR subject to a higher discount rate (5\% versus 1.5\% for the other technologies), we notice that SMR keeps on being more cost-competitive than other flexible production units, like e-ammonia CCGTs, that are substituted by SMR when the latter can be installed (see Chapter 3).
+In other words, it means that the optimistic values assumed for SMR in the thesis lead to a cheaper whole-energy system but does not affect its design, \ie the installed technologies.
+Consequently, no further modification has been brought to the thesis regarding this comment.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.49\textwidth]{LCOE_line_2.pdf}
+\includegraphics[width=0.49\textwidth]{LCOE_line_3.pdf}
+\label{fig:LCOE_comparison}
+\caption{LCOE per technology producing exclusively electricity considered in the thesis (left) and same LCOE but, this time, with a higher CAPEX (7270€/kW versus 4850€/kW in the thesis and a discount rate of 5\% on this technology.
+Even considering these less favourable assumptions for SMR, it still outcompetes other flexible production units like CCGT running on e-ammonia.}
+\end{figure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 53 What about fuel cost?
+\end{mdframed}
+
+\noindent Even though the cost of purchasing uranium was already presented in Figure 2.3 of the thesis, I have added this value {\color{blue}in Table 2.1}:
+
+\begin{table}[htbp!]
+\caption{Nominal features of the SMRs in EnergyScope.
+SMR exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year.}
+\label{tab:SMR_features}
+\begin{minipage}{\linewidth}
+\centering
+\begin{tabular}{l c c}
+\hline
+\textbf{Feature} & \textbf{Value} & \textbf{Unit}\\
+\hline
+CAPEX & 4850 & €/kW \\
+Annual OPEX & 103 & €/kW/year \\
+Lifetime & 60 & year \\
+\textbf{Cost of purchasing uranium} & \textbf{4} & \textbf{€/MWh}\\
+Efficiency & 40\% & -\\
+Maximum capacity & 6 & GW \\
+Annual availability & 85\% & -\\
+Operational year & 2040 & - \\
+Flexibility & Full & - \\
+\hline
+
+\end{tabular}
+\end{minipage}
+\end{table}
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent Indeed, the potential consumption/production of heat and/or electricity for each conversion process is considered in the model.
+For instance, the Methanol-To-Olefins (MTO) process consumes 1.24\,GWh of methanol and 0.33\,GWh of high-temperature heat to produce 1\,GWh of HVC, which makes a 64\% efficiency.
+To clarify this point, {\color{blue}the caption of Figure 2.5} has been extended:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Schematic view of the different resources able to produce HVC, ammonia and methanol with their related conversion technologies (including energy efficiency and their CAPEX - in €/kW of final molecules).
+\textbf{The efficiencies account for the potential consumption/production of heat and/or electricity within each conversion process.} Values stand for 2035.
+Graph from \cite{rixhon2021comprehensive}.
 \end{mdframed}
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 57 Why a uniform distribution? A gaussian woul clearly be more appropriate. Maybe some uncertainties are underestimated (because clipped by uniform distribution)
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent In line with \citet{limpens2019energyscope}, modal shares are indeed exogenously set.
+This aims at keeping some realism and ensures the need of less cost-competitive options like private versus public transport or decentralised versus DHN heating.
+To clarify this point, it is now explicitly written {\color{blue}in Section 2.4}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+[...] (ix) other parameters like the discount rate or the \textbf{exogenous} modal share change in different key sectors.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.
+57: Why a uniform distribution?
+A gaussian would clearly be more appropriate.
+Maybe some uncertainties are underestimated (because clipped by uniform distribution).
+\end{mdframed}
+
+\noindent As already written in the first paragraph of Section 1.2.1, given the scarcity of data to actually define a distribution of uncertainty for the parameters, the common practice is to consider uniform distribution.
+As detailed by \citet{coppittersthesis}: ``Distributions are typically defined based on large datasets, which are not always at hand for each parameter.
+Therefore, when no meaningful information can be extracted on the distribution from a limited dataset, a uniform distribution is typically assumed, which assigns an equal probability to each value within a range.
+This approach is similar to assigning an interval to a parameter, but takes advantage of the central tendency when uniform distributions are propagated through a system model (\ie central limit theorem)''.
+Given what is already in the thesis manuscript and these further explanation, no modification has been brought to the manuscript in this regard.
 
@@
 \label{Chap_atom_vs_molecules}
+
 \begin{mdframed}[style=comment] % Comment from the reviewer
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent Besides having added SMR to the model which is one of the contributions regarding the case study (Chapter 2), the main contributions of this chapter is the global sensitivity analysis (GSA) performed on the optimisation of the transition pathway of a whole-energy system with 34 uncertain parameters.
+When applied to the case study of Belgium, this analysis also allowed to highlight the competition between SMR and some of the imported electrofuels (mostly e-ammonia and e-methane) by 2050.
+
+To make it clearer from the beginning of the Chapter, I have extended {\color{blue}the contributions of Chapter 3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+This GSA applied to a model optimising the transition pathway of a whole-energy system with these many uncertain parameters is the main contribution of this chapter.
+Through this analysis, we also managed to point out the impact of integrating SMR in the Belgian energy system as well as the main drivers of the import of e-hydrogen, e-methane, e-ammonia and e-methanol by 2050.
 \end{mdframed}
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 70: ``in other words, the variation of the parameter f$\_$max,SMR…''. I am not sure that this sentence is equivalent to the previous one. Please clarify.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent The results are deeply detailed in Sections 3.1 and 3.2.
+These sections address the impact of adding SMR in the deterministic solution (\ie uncertain parameters at their respective nominal value) of the Belgian energy transition and global sensitivity analyses on different outputs of interest, respectively.
+The purpose of these two sections is to give numerical substance to the main conclusions and take-aways that are listed in Section 3.3:
+\begin{itemize}
+\item If available, SMR is installed and mostly affect the electricity and high-temperature sectors;
+\item To minimise the variation of the total transition cost, the key parameter is the cost of purchasing electrofuels where parameters related to SMR (availability and CAPEX) have much lower impact;
+\item E-ammonia and, to a lesser extent, e-methane imports by 2050 are impacted by the availability of SMR whereas e-hydrogen and e-methanol are rather influenced by variation related to transport technologies and non-energy demand, respectively;
+\item Even though SMR and electrofuels are in competition, they do not limit the deployment of VRES in Belgium.
+In other words, investing in wind and solar is a must-do for the Belgian energy transition;
+\item Given the set of assumptions, especially considering the end-use-demands and the local potential of VRES, Belgium will need to import molecules in the near future.
+On top of investing into VRES, it seems reasonable to invest in the imports of these electrofuels as betting on SMR means letting the short-term emissions go up and, potentially, requiring significant (and quick) adjustments in case the technology is not ready on time to still respect the \ce{CO2}-budget.
+\end{itemize}
+
+Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+70: ``in other words, the variation of the parameter $f_{\mathrm{max,SMR}}$…''.
+I am not sure that this sentence is equivalent to the previous one.
+Please clarify.
+\end{mdframed}
+
+\noindent The Sobol' index is the ratio between the variance of the total transition cost due to the variation of one parameter and the total variance of the total transition cost.
+If $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the impact on the system, and, consequently, on its total transition cost, is the same: no SMR installed.
+A wide range of variation of the parameter leading to zero variation of the output of interest is one of the reasons to explain why the Sobol' index of this parameter is low.
+To clarify this point, I have rephrased the sentence {\color{blue}in Section 3.2.1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Given the uncertainty characterisation presented in Section 2.4, there are 60\% chance that no SMR could be installed.
+\textbf{When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
+}
 \end{mdframed}
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent This low assumed potential is indeed one of the reasons why the SMR-related uncertain parameters have a limited impact on the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
+However, this is not the only reasons as detailed earlier in the same paragraph: the late potential availability (2040 at the soonest) as well as the ``cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.''.
+
+Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent The objective of this set of figures is to point out the key parameters that drive the variation of some outputs of interest.
+Taking the case of Figure 3.8, we focus on the import of e-methanol in 2050.
+On the x-axis, we represent the value taken by the uncertain parameters in the different samples.
+As the range is not the same for all the parameters, we have limited this axis from ``Min'' to ``Max''.
+The y-axis represents the feature of interest.
+The drawn lines represent average trend lines between the output of interest and one specific parameter.
+For the case of Industry EUD, for samples of the GSA where this uncertain parameter is at 35\% of its range, the mean value of methanol imported in 2050 is roughly 50\,TWh.
+The distributions present on the extreme left and right of the graphs correspond to the distributions of the output of interest for the samples of the GSA where the considered parameter takes a value in the bottom (left) or top (right) 15\% of its range.
+
+To improve the readability of this figure, I have updated the text zones to specify that these lines and distributions represent results of some specific samples of the GSA.
+Additionally, I have added a text zone to remind that the value between brackets is the Sobol' index of the considered parameter regarding this specific output of interest, and added this information in the caption of the figure:
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-methanol in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_methanol}
+\end{figure}
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent As specified in the table about SMR features (see previous comment), the considered fuel for SMR is uranium, as it is the case for conventional nuclear power plants.
+Given this, I have not brought modification to the manuscript in this regard.
 
@@
 \noindent This work will be done for the subsequent paper on RL.
-The strategy would to implement a simplified case study where it would be ``computationally affordable'' to test the stochastic programming approach and compare it with the optimal policy delivered by the RL approach.
+The strategy would to implement a simplified case study (\ie smaller system, fewer uncertain parameters and more limited time horizon) where it would be ``computationally more affordable'' to test the stochastic programming approach and compare it with the optimal policy delivered by the RL approach.
 Consequently, regarding this comment, there has not been further modification brought to the manuscript.
@@
 
-\noindent To define in Section 1.3.2 and remind here {\color{blue} }.
+\noindent {\color{blue} }.
 
@@
 
-\noindent To explain in the definition of the action {\color{blue} }.
+\noindent {\color{blue} }.
 
@@
 In a LP problem, constraints represent hyperplanes in the domain of variables.
-In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}).
+In a two-dimension space, these are straight lines (see Figure 4.7).
 When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility.
@@
 In other words, binding constraints, when tightened, aggravate the objective value function.
-If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.
-\end{mdframed}
-
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
-\caption{Binding versus non-binding constraints.
-In LP where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$).
-The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution.
-These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
-\label{fig:Binding_constr}
-\end{figure}
+If these are inequality constraints, as represented in Figure 4.7, it means that their left and right sides of the equation are equal.
+\end{mdframed}
 
@@
 \end{mdframed}
-
-\subsection{Confusion with the word PC}
-\label{Confusion_PC_wording}
 

### 2024-07-05 appendices/app_EnergyScope.tex
Rebuttals: Keep going...

@@
 
-``\textit{Only when single-model results are contextualized by the model's position in the larger ensemble, the reader would be able to have a complete and correct interpretation of the output}" \cite{dekker2023identifying}.
+``\textit{Only when single-model results are contextualised by the model's position in the larger ensemble, the reader would be able to have a complete and correct interpretation of the output}'' \cite{dekker2023identifying}.
 Energy system models of varying complexity are valuable tools for guiding policymakers and projecting future trends.
@@
 \includegraphics[width=\textwidth]{Models_classification.png}
-\caption{Model can be classified by their core focus: \textbf{Operation} or \textbf{Design}.
+\caption{Models can be classified by their core focus: \textbf{Operation} or \textbf{Design}.
 These categories can be broken down into subcategories.
@@
 Besides the behavioral and social factors, they also highlighted the challenges related to mutli-sectorial systems, time and space resolution or the open-accessibility of data and models and their ability to account for uncertainties.
-In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv).
+In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}'' models, focusing on criteria (i-ii) and (iv).
 This review was followed by a classification where criteria (i-iv) were taken into account \cite{prina2020classification}.
@@
 Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic.
-In fact, the model is typically employed for snapshot analysis, \ie optimization of a target future year.
+In fact, the model is typically employed for snapshot analysis, \ie optimisation of a target future year.
 Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}.
@@
 This choice was substantiated by the work of \citet{welsch2014incorporating}, which demonstrated that this level of temporal granularity yields consistent results in comparison to hourly time resolution over a year.
-However, it is noteworthy that the utilization of a limited number of time slices may oversimplify the optimization of storage technologies, especially those designed for inter-month energy storage.
+However, it is noteworthy that the utilisation of a limited number of time slices may oversimplify the optimisation of storage technologies, especially those designed for inter-month energy storage.
 This simplification can be viewed as a pragmatic approach to reduce the computational burden while over-simplifying the challenge of accurately integrating intermittent renewable energy sources.
-Furthermore, PyPSA, a modeling framework recognized for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}.
+Furthermore, PyPSA, a modeling framework recognised for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}.
 \\
@@
 It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration.
-This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute.
+This model has a widespread adoption and has been utilised by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute.
 Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}.
@@
 Recent developments by the IEA-ETSAP have resulted in a version that is compatible with the open-source solver CBC.
-In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}.
+In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilise 12 to 32 time-slices annually \cite{prina2020classification}.
 To highlight the sensitivity of results to time resolution, \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using up to 12 typical days with an hourly resolution.
@@
 The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost and greenhouse gas emissions of the energy system.
-Typically, the two objectives are integrated by placing a limit on emissions while simultaneously minimizing the costs.
+Typically, the two objectives are integrated by placing a limit on emissions while simultaneously minimising the costs.
 
@@
 \includegraphics[width=\textwidth]{chp_estd_overview.pdf}
-\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimizing model (Optimisation) and results (Energy strategy). }
+\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimising model (Optimisation) and results (Energy strategy). }
 \label{fig:estd_overview}
@@
 This approach results in seven instances of EnergyScope TD -- called representative years -- spanning the 30-year transition period, covering the years from 2020 to 2050.
-To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition (see Figure \ref{fig:meth_path_methodology}).
+To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition (see Figure \ref{fig:meth_path_methodology_app}).
 Overall, these constraints are integrated into a linear framework, ensuring computational efficiency, with an approximate computational time of 14 minutes on a personal laptop (2.4 GHz Intel Core i5 quad-core).
@@
 The overall problem is the pathway model.}
-\label{fig:meth_path_methodology}
+\label{fig:meth_path_methodology_app}
 \end{figure}
@@
 The residual value is not applied to technologies that are removed prematurely.
-This differ from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).
+This is similar to the work of \citet{prina2019transition} but differs from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).
+
+In the work of \citet{goffauxpathway}, we carried out a sensitivity analysis on the formulation of the salvage value.
+First and foremost, this analysis confirmed the need to account for the salvage value to avoid penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}.
+Then, the sensitivity analysis focused on two elements in the expression of the salvage value: $\tau_{phase}(p)$ and $\textbf{F\textsubscript{decom}}(p2,p,i)$.
+
+In Equation (eq:salvage_app), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
+With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost.
+However, one could wonder if a technology that is still operational for half of its lifetime is not worth less than half of its initial investment cost, given the depreciation of the asset.
+In that case, the annualisation factor should account for the last phase of the optimisation (\ie 2045\_2050) when the residual value of the technologies are calculated and subtracted from the total transition cost.
+
+On top of this, we also investigated the impact of accounting for the decommissioned technologies or not in the expression of the salvage value.
+In Equation (eq:salvage_app), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0.
+This is relevant since a technology decommissioned before 2050 would not be available after 2050.
+The issue is that the model could keep unnecessary technologies to subtract their salvage value from the total transition cost in order to decrease it.
+This would be done if the fixed OPEX of an unused technology is lower than its salvage value.
+Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value.
+In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a salvage value.
+
+The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon.
 

### 2024-07-05 appendices/app_case_study.tex
Integration of comments in rebuttals
Rebuttals: Keep going...

@@
 \includegraphics[width=\textwidth]{Gwp_per_sector.pdf}
-\caption{Energy sectors have different rate to reduce \gls{GHG} emissions over the transition.
+\caption{Energy sectors have different rates to reduce \gls{GHG} emissions over the transition.
 The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon-neutrality in 2050.}
@@
 \includegraphics[width=\textwidth]{System_cost.pdf}
-\caption{The yearly system cost shows the shift from non-renewable to renewable resources (mainly electrofuels). Operation cost and maintenance represents almost 50\% of the expenses.}
+\caption{The yearly system cost shows the shift from non-renewable to renewable resources (mainly electrofuels). Operation and maintenance costs represents almost 50\% of the expenses.}
 \label{fig:pestd_cumul_op}
@@
 For instance, in the myopic approach, 0.8GW of industrial coal boilers will remain installed in 2045 and 2050 or 3.6GW of naphtha-crackers to produce HVC in 2040, whereas these technologies are not used.
-This could be compare to the ``lock-ins" detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
+This could be compare to the ``lock-ins'' detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
 However, in our case, keeping in place unused technologies is due to the formulation of the salvage value.
@@
 \includegraphics[width=\textwidth]{MY_resources_category_diff.pdf}
-\caption{Primary energy resources difference between the myopic and perfect foresight (PF) approaches. Positive values mean that the myopic approach is higher than perfect foresight.}
+\caption{Difference of primary energy resources between the myopic and perfect foresight (PF) approaches. Positive values mean that the myopic approach is higher than perfect foresight.}
 \label{fig:my_pestd_res_cat_diff}
@@
 Then, the generation mix in heating sectors or non-energy demands is less diversified, leading the sharper and later switch of resources and technologies.
-For instance, lignocellulosic biomass is ``cannibalized" by industrial boilers to the detriment of biomass-to-methanol or \gls{HVC}.
+For instance, lignocellulosic biomass is ``cannibalised'' by industrial boilers to the detriment of biomass-to-methanol or \gls{HVC}.
 Besides this, in general, we observe an overestimation of conventional technologies (\eg running on gas or oil) generation, either by installing more capacities or having a higher load factor for similar capacities.
@@
 
-%This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors.
+%This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the modal share change in different key sectors.
 For the specific case of \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ will influence the maximum capacity (\ie 6~GW) to install to translate somehow the readiness of this technology.
@@
 %
-%\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Interest rate & I & -46.2\% & 46.2\% \\
+%\multirow{6}{*}{\textbf{Miscellaneous}} &$i_{\mathrm{rate}}$  & Discount rate & I & -46.2\% & 46.2\% \\
 %& $\%_{\mathrm{pub,max}}$ & Max share of public transport & I & -10\% & 10\% \\

### 2024-07-05 appendices/app_taxonomy.tex
RobPol + RL: Review by FC (part 2)
Rebuttals: Keep going...

@@
-\section{Introduction}
-\label{sec:intro}
 Since stone age, humans use fuels, defined as any energy carriers intended for energy conversion \cite{FAO_biofuel, ISO16559}.
@@
 The objective of this terminology is to avoid the confusion between these fuels and thereby reduce misunderstanding in political or academic discussions.
-To a further extent, such an harmonisation may help ``to facilitate comparison between national fuel market and enhance trading" \cite{FAO_biofuel}.
+To a further extent, such an harmonisation may help ``to facilitate comparison between national fuel market and enhance trading'' \cite{FAO_biofuel}.
 It also aims at promoting and increasing the use of biomass as energy carriers and electricity transformed into fuel.
@@
 \label{sec:lit_review}
-\citet{ridjan2016} performed a first literature review between 2006 and 2014, looking in Scopus for the three keywords: (i) ``synthetic fuels", (ii) ``electrofuels" and (iii)``alternative fuels".
-Based on this extensive review work, they proposed the following definitions: (i) \textit{Synthetic fuels} were defined as x-To-Liquid (xTL) processed fuels, within the scope of ``Fischer-Tropsch fuels that are produced by gasification of either coal, natural gas or biomass".
+\citet{ridjan2016} performed a first literature review between 2006 and 2014, looking in Scopus for the three keywords: (i) ``synthetic fuels'', (ii) ``electrofuels'' and (iii)``alternative fuels''.
+Based on this extensive review work, they proposed the following definitions: (i) \textit{Synthetic fuels} were defined as x-To-Liquid (xTL) processed fuels, within the scope of ``Fischer-Tropsch fuels that are produced by gasification of either coal, natural gas or biomass''.
 Comparatively, similar definitions are given in \cite{eia2006} and \cite{iea2007}, except that the latter excludes biomass from the potential feedstocks.
@@
 
-More recently, \citet{Stancin2020} has shown that scientific research about ``\textit{alternative fuels}" has experienced a significant impulse since the 2000s.
+More recently, \citet{Stancin2020} has shown that scientific research about ``\textit{alternative fuels}'' has experienced a significant impulse since the 2000s.
 Therefore, in the continuity of \cite{ridjan2016}, we performed a similar literature review in order to update the terminology used by the scientific community.
@@
 The non-relevant documents usually contained information that fall out of the scope of the current work, \eg information on specific catalysts.
-Out of the relevant documents, the main outcomes concern the relative majority of ``synthetic fuels", the emergence of ``electrofuels" and the scattered diversity of other terms.
+Out of the relevant documents, the main outcomes concern the relative majority of ``synthetic fuels'', the emergence of ``electrofuels'' and the scattered diversity of other terms.
 
-Where terms like ``synthetic gas", ``electrofuels" or ``synthetic natural gas" account for 19\%, 13\% and 8\% of the additional relevant literature, respectively, ``synthetic fuels" is used in 35\% of the case to cover fuels produced from a wide variety of primary resources (renewable or not) and through various conversion processes.
+Where terms like ``synthetic gas'', ``electrofuels'' or ``synthetic natural gas'' account for 19\%, 13\% and 8\% of the additional relevant literature, respectively, ``synthetic fuels'' is used in 35\% of the case to cover fuels produced from a wide variety of primary resources (renewable or not) and through various conversion processes.
 To pick only a few examples, its usage ranged from liquid hydrocarbons produced from synthesis gas through the Fischer-Tropsch process \cite{haarlemmer2014,Trieb2018} to ammonia from the Haber-Bosch process \cite{bargiacchi2019} or biomass gasification or pyrolysis \cite{monaco2018, rao2015}.
-As a consequence of an increasing share of \gls{VRES} in the energy system, ``electrofuels" are gaining more popularity: where this term was used in 4\% of the relevant literature between 2006 and 2014 \cite{ridjan2016}, it is now present in 13\% of the additional relevant literature reviewed for this work.
+As a consequence of an increasing share of \gls{VRES} in the energy system, ``electrofuels'' are gaining more popularity: where this term was used in 4\% of the relevant literature between 2006 and 2014 \cite{ridjan2016}, it is now present in 13\% of the additional relevant literature reviewed for this work.
 These fuels are considered as being produced mainly based on the electricity provided by \gls{VRES} \cite{brynolf2018}.
 Besides water electrolysis, this electricity also supplies processes like \ce{CO2} hydrogenation \cite{brynolf2018,Decker2019,Pearson2012} or co-electrolysis of \ce{CO2} and \ce{H2O} \cite{hanggi2019,larsson2015}.
-Aside the two aforementioned keywords, the literature makes use of a scattered diversity of terms that aim at precising the features of the considered fuel: for instance, ``solar fuel", ``power gas" or ``green synthetic fuel".
-However, out of context, words like ``green fuel", ``advanced fuel" or ``solar fuel" may be misunderstood because of their multiple interpretations \cite{ridjan2016}.
+Aside the two aforementioned keywords, the literature makes use of a scattered diversity of terms that aim at precising the features of the considered fuel: for instance, ``solar fuel'', ``power gas'' or ``green synthetic fuel''.
+However, out of context, words like ``green fuel'', ``advanced fuel'' or ``solar fuel'' may be misunderstood because of their multiple interpretations \cite{ridjan2016}.
 
@@
 %To meet climate objectives while ensuring energy supply, any energy carrier or conversion technologies of an energy system aims at being sustainable.
-Thereon, European Commission defines ``sustainable development as a development that meets the needs of current generations without compromising the ability of future generations to meet theirs" \cite{eu_sustainable}.
+Thereon, European Commission defines ``sustainable development as a development that meets the needs of current generations without compromising the ability of future generations to meet theirs'' \cite{eu_sustainable}.
 Given the difficulty to quantify and circumscribe the sustainability of a fuel, we decided here to suggest classification and definition of the fuels regardless of this feature.
@@
 Renewable energy sources are non-fossil sources that are naturally replenished on a human timescale (wind, solar, geothermal, wave, tidal, hydro-power, biomass) \cite{ellabban2014}.
-A fuel can be defined as ``renewable" only when it is based on renewable sources.
-To be ``sustainable", a ``renewable fuel" should not increase the concentration of \ce{CO2} in the atmosphere \cite{Dincer2020}, among other things \cite{eu_sustainable}.
+A fuel can be defined as ``renewable'' only when it is based on renewable sources.
+To be ``sustainable'', a ``renewable fuel'' should not increase the concentration of \ce{CO2} in the atmosphere \cite{Dincer2020}, among other things \cite{eu_sustainable}.
 %and components as well as the energy necessary to its production come from renewable sources.
@@
 Therefore, besides the aforementioned renewable/non-renewable distinction, we propose to categorise fuels in three groups whose definitions are broad enough to capture the diversity of the fuels while maintaining a simplified representation: (i) \textit{Biofuels} as the absolute majority of the energy balance of their production consists of biomass; (ii) \textit{Electrofuels}, similarly to biofuels, but with electricity instead of biomass.
-\textit{A priori}, it is complicated to give an exact number to quantify this ``majority" of the energy balance but we propose a solution in \Cref{sec:discussion} to remove this ambiguity.
+\textit{A priori}, it is complicated to give an exact number to quantify this ``majority'' of the energy balance but we propose a solution in \Cref{sec:discussion} to remove this ambiguity.
 (iii) \textit{Synthetic fuels} when the end-use fuel results from the upgrade of another fuel to improve physical and/or chemical characteristics (\eg to increase the volumetric energy density).
@@
     \caption{Overview of the main pathways to produce fuels.
-The boxes in the ``conversion" category illustrates the terminology to use in the different cases: first, a fuel can be defined as \emph{bio} or \emph{electro} if the absolute majority of the energy balance of its production consists of biomass or electricity, respectively.
+The boxes in the ``conversion'' category illustrates the terminology to use in the different cases: first, a fuel can be defined as \emph{bio} or \emph{electro} if the absolute majority of the energy balance of its production consists of biomass or electricity, respectively.
 Then, a fuel can be defined as \emph{synthetic} if it has been obtained through a synthesis process from another fuel, regardless of the feedstock.
-The box "Fuels" gives some examples to illustrate the terminology used and applied to very common fuels.}
+The box ''Fuels'' gives some examples to illustrate the terminology used and applied to very common fuels.}
     \label{fig:synfuels_paths}
@@
 They are mainly refined from fossil hydrocarbons.
-Thus, we define ``\textit{fossil} fuels" as: \textit{fuel based on fossil resources.}
+Thus, we define ``\textit{fossil} fuels'' as: \textit{fuel based on fossil resources.}
 
@@
 Biomass is mainly harvested in a solid form and has a lower energy density than fossil hydrocarbons.
-Thus, the ``bio" processes aim at concentrating the energy and converting it into a fuel more convenient to use (\eg pellet from woody wastes).
+Thus, the ``bio'' processes aim at concentrating the energy and converting it into a fuel more convenient to use (\eg pellet from woody wastes).
 %liquid or gaseous fuels.
 Similarly to the \citet{OEERE_biofuel}, in the literature, biofuels mainly target liquid or gaseous transport fuels (\eg biodiesel and bioethanol) and imply a sustainable production \cite{eu_biofuel, pottering2009directive}.
-The definition given by official institutions (\eg the FAO or the European Commission) is aligned with the ISO 16559:2014: ``biofuel is a solid, liquid or gaseous fuel produced directly or indirectly from biomass" \cite{eu_biofuels, FAO_biofuel, ISO16559}.
+The definition given by official institutions (\eg the FAO or the European Commission) is aligned with the ISO 16559:2014: ``biofuel is a solid, liquid or gaseous fuel produced directly or indirectly from biomass'' \cite{eu_biofuels, FAO_biofuel, ISO16559}.
 To add a quantitative and measurable aspect in the terminology, we propose the following definition: \textit{biofuels are fuels produced from biomass as the major component of their energy balance}.
-Even if our terminology lacks the subtlety brought by terms like ``woodfuels", ``agrofuels", or ``municipal by-products", that allow to highlight the source of biomass at stake, provided by the FAO \cite{FAO_biofuel}, this terminology aims at harmonising and desambiguating the meaning of the general term ``biofuel".
+Even if our terminology lacks the subtlety brought by terms like ``woodfuels'', ``agrofuels'', or ``municipal by-products'', that allow to highlight the source of biomass at stake, provided by the FAO \cite{FAO_biofuel}, this terminology aims at harmonising and desambiguating the meaning of the general term ``biofuel''.
 
 %According to the European Commission, biofuels are ``liquid or gaseous transport fuels such as biodiesel and bioethanol which are made from biomass.
-For biofuels to reduce greenhouse gas emissions without adversely affecting the environment or social sustainability, they must be produced in a sustainable way" \cite{eu_biofuel}.
-This is also stated in the directive of the European Parliament: ``(10) Biofuel production should be sustainable" \cite{pottering2009directive}.
+For biofuels to reduce greenhouse gas emissions without adversely affecting the environment or social sustainability, they must be produced in a sustainable way'' \cite{eu_biofuel}.
+This is also stated in the directive of the European Parliament: ``(10) Biofuel production should be sustainable'' \cite{pottering2009directive}.
 However, this definition is too restrictive as it does not include solid biofuels (\eg bio-char that could replace coke in steel production \cite{mousa2016biomass}) nor biofuels used for other purposes than transport (\ie heat and power).
@@
 Electrofuels group all the fuels where the absolute majority of the energy contained into the fuel comes from electricity.
-In this sense, it is important to point out that ``electrofuels" do not especially imply ``renewable fuels" (\Cref{fig:synfuels_paths}).
+In this sense, it is important to point out that ``electrofuels'' do not especially imply ``renewable fuels'' (\Cref{fig:synfuels_paths}).
 For instance, the electricity needed to produce \ce{H2} through an electrolyser can be generated from renewable or non-renewable sources \cite{bhandari2014}, even if the most sustainable way to produce electrofuels remains through the electricity produced in excess from \gls{VRES}.
@@
 It encompasses upgrading processes, such as methane produced through the Sabatier process or longer hydrocarbons based on syngas through the Fischer-Tropsch process.
-We extend the definition of the European Commission and Energy Information Agency who consider synthetic fuels as ``any liquid fuel obtained from coal, natural gas or biomass" \cite{eia2006, eu_synthetic}.
+We extend the definition of the European Commission and Energy Information Agency who consider synthetic fuels as ``any liquid fuel obtained from coal, natural gas or biomass'' \cite{eia2006, eu_synthetic}.
 By extending this definition and focusing on the synthesis operation, similarly to \citet{speight2020synthetic}, we are able to include emerging fuels like ammonia from the Haber-Bosch process.
@@
 Another example could be the production of ammonia via the synthesis of \ce{H2} from methane-reforming.
-Consequently, defining a fuel only as ``synthetic" gives a reduced insight about the fuel, especially about its ``renewability".
-Therefore, we recommend to avoid using ``synthetic fuel".
-In the former example, one would then rather define the produced methanol as a ``bio-electrofuel" given that its main sources are electricity and biomass.
+Consequently, defining a fuel only as ``synthetic'' gives a reduced insight about the fuel, especially about its ``renewability''.
+Therefore, we recommend to avoid using ``synthetic fuel''.
+In the former example, one would then rather define the produced methanol as a ``bio-electrofuel'' given that its main sources are electricity and biomass.
 % To summarise, we define \textit{synthetic as a generic, and therefore less specific, term that gathers all the aforementioned ones (\ie fossil, bio or electro).
@@
 
-Finally, besides these recommended terms in the context of the energy transition, others are mainly misleading: ``alternative" or ``non-conventional" fuels.
+Finally, besides these recommended terms in the context of the energy transition, others are mainly misleading: ``alternative'' or ``non-conventional'' fuels.
 These terms are defined in opposition to another kind of fuels, mostly fossil.
 However, alternatives of today will be conventions of tomorrow.
-This means that depending on the ``conventional" of the context, these terms will represent different fuels.
+This means that depending on the ``conventional'' of the context, these terms will represent different fuels.
 Because of this implied ambiguity, we recommend not to use these words as their interpretation is case and time sensitive.
@@
 In this context, we suggest comprehensive terminology and classification to characterise the fuels.
-Given the limited insight provided by the widely-used term ``synthetic fuels", we recommend to use more specific terms like ``biofuels", ``electrofuels", or even hybridised terms (\eg ``bio-electrofuels").
+Given the limited insight provided by the widely-used term ``synthetic fuels'', we recommend to use more specific terms like ``biofuels'', ``electrofuels'', or even hybridised terms (\eg ``bio-electrofuels'').
 These allow to highlight the contribution of biomass and electricity in the energy balance of the fuel.
@@
 Therefore, to emphasise the effort to step away from fossil fuels, we suggest to specify the contribution of biomass (or electricity) within the energy balance of the production of the fuel looking at both the feedstock and the energy to supply the conversion or production process.
-This would give, for instance, ``bio(25) methanol" if 25\% of the energy balance to produce the methanol comes from biomass (\eg biomass gasification).
-Another example could be ``electro(40) ammonia" if 40\% of the energy balance comes from electricity (\eg water electrolysis, air-captured nitrogen and power supply of the Haber-Bosch process).
+This would give, for instance, ``bio(25) methanol'' if 25\% of the energy balance to produce the methanol comes from biomass (\eg biomass gasification).
+Another example could be ``electro(40) ammonia'' if 40\% of the energy balance comes from electricity (\eg water electrolysis, air-captured nitrogen and power supply of the Haber-Bosch process).
 As electricity can be produced from a wide variety of sources (\Cref{fig:synfuels_paths}), we recommend to give insights on the composition of this electricity (\eg from wind, solar or the grid mix).
@@
 
-% terms like ``renewable fuels", ``electrofuels", ``biofuels" or ``synthetic fuels" are more and more popular in the literature and used for a vast variety of fuels.
+% terms like ``renewable fuels'', ``electrofuels'', ``biofuels'' or ``synthetic fuels'' are more and more popular in the literature and used for a vast variety of fuels.
 These terms imply specific use of primary resources or process.
@@
 
-% Through a comprehensive literature review, we first identified the top-used term (\ie ``synthetic fuel"), the emergence of new ones like ``electrofuel" and the wide variety of other terms.
+% Through a comprehensive literature review, we first identified the top-used term (\ie ``synthetic fuel''), the emergence of new ones like ``electrofuel'' and the wide variety of other terms.
 Then, on the purpose for setting a comprehensive terminology and classification, we pointed out different categories of fuels.
 
-% First of all, the definition of ``renewable fuels" was clearly reminded as fuels produced from non-fossil sources that are naturally replenished on a human timescale.
+% First of all, the definition of ``renewable fuels'' was clearly reminded as fuels produced from non-fossil sources that are naturally replenished on a human timescale.
 As we aim at a pathway towards sustainability where fuels will keep playing a significant role, the renewable fuels shall gain more and more shares in the global energy mix.
-Therefore, we recommend to use ``renewable fuel" when it is not produced from fossil sources.
-Although, we also point out that ``renewable" does not directly mean ``sustainable" as this term implies many more aspects than ``Affordable and Clean Energy" \cite{eu_sustainable}, especially when we consider biofuels (\eg use of lands, indirect impacts on environment).
+Therefore, we recommend to use ``renewable fuel'' when it is not produced from fossil sources.
+Although, we also point out that ``renewable'' does not directly mean ``sustainable'' as this term implies many more aspects than ``Affordable and Clean Energy'' \cite{eu_sustainable}, especially when we consider biofuels (\eg use of lands, indirect impacts on environment).
 
-% Then, besides this first distinction between renewable and non-renewable, mostly fossil fuels, we identified three terms to describe them: ``biofuels", ``electrofuels" and ``synthetic fuels".
+% Then, besides this first distinction between renewable and non-renewable, mostly fossil fuels, we identified three terms to describe them: ``biofuels'', ``electrofuels'' and ``synthetic fuels''.
 On one hand, the first two really focus on the resource and the energy balance to produce the fuel.
-Fuels are defined as ``bio" or ``electro" if a majority of their energy balance (\ie feedstock and energy supply for the conversion processes) consists respectively of biomass or electricity.
+Fuels are defined as ``bio'' or ``electro'' if a majority of their energy balance (\ie feedstock and energy supply for the conversion processes) consists respectively of biomass or electricity.
 On the other hand, the last one targets the synthesis process from which a fuel can be produced based on another (or other) one(s).
 As observed in the literature, this includes many different fuels, from many different processes.
-Therefore, we encourage to use specific terms like ``biofuels" or ``electrofuels", even hybridised term (\eg ``bio-electrofuels") to give more insight to the reader about the technical implications behind the concerned fuel.
+Therefore, we encourage to use specific terms like ``biofuels'' or ``electrofuels'', even hybridised term (\eg ``bio-electrofuels'') to give more insight to the reader about the technical implications behind the concerned fuel.
 
-% Finally, we recommended not to use case and time sensitive terms like ``alternative" or ``non-conventional" fuels, given their ambiguous dependency on the alternatives or conventions which these fuels are opposed to.
+% Finally, we recommended not to use case and time sensitive terms like ``alternative'' or ``non-conventional'' fuels, given their ambiguous dependency on the alternatives or conventions which these fuels are opposed to.
 

### 2024-07-05 chap_RL/chap_RL.tex
Rebuttals: Keep going...

@@
 \caption{Reward distribution between successes and failures.
-Right hand side details at the end of which time-window the failure occurred.
+Graph on the right hand side details at the end of which time-window the failure occurred.
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
@@
 Per definition, the threshold of 1.2\,Gt$_{\ce{CO2},\text{eq}}$ splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
+Since infeasible cases or those that overshoot the \ce{CO2}-budget are discarded before reaching 2050 (see Section \ref{subsec:RL:act_states_rew:rew}), the number of attempts that reach further steps in the transition progressively decreases.
+Consequently, the share of successful transitions compared to failures progressively increases with time.
 In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
@@
 \includegraphics[width=0.8\textwidth]{Cum_gwp_cost_3.pdf}
-\caption{Exploration of the state space over the learning process: cumulative emissions and costs.
+\caption{Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
 The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
@@
 \includegraphics[width=0.8\textwidth]{RE_in_mix_Efficiency.pdf}
-\caption{Exploration of the state space over the learning process: share of renewable energy carriers in the primary energy mix and efficiency.
+\caption{Exploration of the state space over the learning process: distribution of occurrence of share of renewable energy carriers in the primary energy mix (left) and efficiency (right).
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
 Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
@@
 \includegraphics[width=0.8\textwidth]{Actions_learning.pdf}
-\caption{Besides the wide exploration of the action space, successes and failures indistinguishably spread over the whole ranges of actions.
+\caption{Exploration of the action space over the learning process: distribution of occurrence of the different actions at the time they are taken by the agent.
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Besides this wide exploration, successes and failures indistinguishably spread over the whole ranges of actions.
 In other words, it does not seem to be any clear set of actions to support successful transitions.}
@@
 \includegraphics[width=0.8\textwidth]{Binding_learning.pdf}
-\caption{Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
+\caption{Keeping only the successful transitions, distribution of occurrence of binding and not binding actions.
+Depending on the action and its timing, it is actually constraining the optimisation through EnergyScope Pathway or not.
 Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.

### 2024-07-05 chap_RobPol/chap_RobPol.tex
Rebuttals: Keep going...

@@
 \caption{Contribution of the top-5 technologies to the first three \gls{PCs} over the different representative years.
-Where the variation of \gls{PV} panels supported by the electrification of the \gls{HT} heat via industrial electric resistors is the main varying drivers over the whole transition, there are tipping years like 2025 for the production of \gls{HVC} (\ie naphtha-cracker to \gls{MTO} or 2025-2030 for the production of \gls{LT} heat (\ie from oil and gas boilers to electric \gls{HP}.}
+Where the variation of \gls{PV} panels supported by the electrification of the \gls{HT} heat via industrial electric resistors are the main varying drivers over the whole transition, there are tipping years like 2025 for the production of \gls{HVC} (\ie naphtha-cracker to \gls{MTO}) or 2025-2030 for the production of \gls{LT} heat (\ie from oil and gas boilers to electric \gls{HP}).}
 \label{fig:Top5_PC_year}

### 2024-07-05 chap_atom_mol/chap_atom_mol.tex
Rebuttals: Keep going...

@@
 For instance, in their work, \citet{My2050} assessed different scenarios for a climate neutral Belgium by 2050.
-Depending on the scenarios, the emphasis is put on a higher electrification of the whole-energy system, a higher consumption of hydrogen or more complex moleclules (\ie electrofuels) or a bigger reliance on biomass and the related \gls{BECCS}.
+Depending on the scenarios, the emphasis is put on a higher electrification of the whole-energy system, a higher consumption of hydrogen or more complex molecules (\ie electrofuels) or a bigger reliance on biomass and the related \gls{BECCS}.
 Besides these, ``unicorn'' technologies are also investigated \cite{heuberger2018impact}.
@@
 Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}).
+This \gls{GSA} applied to a model optimising the transition pathway of a whole-energy system with these many uncertain parameters is the main contribution of this chapter.
+Through this analysis, we also managed to point out the impact of integrating \gls{SMR} in the Belgian energy system as well as the main drivers of the import of e-hydrogen, e-methane, e-ammonia and e-methanol by 2050.
 
@@
 \includegraphics[width=\textwidth]{Elec_Tech_Cap_SMR.pdf}
-\caption{As soon as available (\ie 2040), \acrfull{SMR} is deployed to their maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie gas and ammonia \gls{CCGT}).}
+\caption{Installed capacities in the power sector in the REF (top) and SMR (down) scenarios.
+As soon as available (\ie 2040), \acrfull{SMR} is deployed to their maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie methane and ammonia \gls{CCGT}).}
 \label{fig:results_deter_tech_cap_elec}
@@
 \includegraphics[width=\textwidth]{Elec_Layer_SMR.pdf}
-\caption{The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
+\caption{Balance between the production and the consumption of electricity in the REF (top) and SMR (down) scenarios.
+The production of electricity from \gls{SMR} substitutes more efficient technologies (\ie \gls{CCGT} and \gls{CHP}) and boosts the electrification of the rest of the system, mostly the industrial high-temperature heat sector.}
 \label{fig:results_deter_layer_elec}
@@
 \includegraphics[width=\textwidth]{GWP_per_sector_diff.pdf}
-\caption{Emissions per energy sector differerence between the SMR and the REF cases.
+\caption{Emissions per energy sector difference between the SMR and the REF cases.
 The deployment of \gls{SMR} from 2040 onward allows maintaining longer the use of \acrfull{LFO} for the production of \acrfull{HVC}.}
@@
 \caption{(Top) Primary energy mix over the transition for the REF and SMR cases.
-(Bottom) Difference of the mix between the two cases after aggregation per category.
+(Down) Difference of the mix between the two cases after aggregation per category.
 In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
@@
 The main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
-In the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production" of the 59\,GW solar-PV when fully deployed, in the sunny days.
+In the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days.
 With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity.
@@
 Given the uncertainty characterisation presented in Section \ref{sec:cs:uncertainty}, there are 60\% chance that no \gls{SMR} could be installed.
-In other words, the variation of the parameter $f_{\mathrm{max,SMR}}$ has zero impact on the variation of the total transition cost in 60\% of the samples.
+When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
 Then, in perspective with the scenario analysis of Section \ref{sec:atom_mol:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
@@
 
-\Cref{fig:UQ_PDF_total_transition_cost} shows the \gls{PDF} of the total transition cost given the 1260 samples.
+\Cref{fig:UQ_PDF_total_transition_cost} shows the distribution of the total transition cost given the 1260 samples.
 Stretching between 660\,b€ and 2050\,b€, the mean, the median and the nominal value (\ie REF case) are close to each other, respectively 1180\,b€, 1160\,b€ and 1080\,b€.
 Similarly to the analysis carried out by \citet{coppitters2023optimizing}, one observes here that the distribution is right-skewed.
-It could then be qualified as ``fragile" as the top 50\% of the samples cover a bigger range (\ie between 1160 and 2050) than the bottom 50\% of the samples (\ie between 660 and 1160).
+It could then be qualified as ``fragile'' as the top 50\% of the samples cover a bigger range (\ie between 1160 and 2050) than the bottom 50\% of the samples (\ie between 660 and 1160).
 In other words, the bad scenarios, resulting in a total transition cost higher than the median, have a bigger effect on this cost.
@@
 \includegraphics[width=0.8\textwidth]{UQ_PDF_total_transition_cost.pdf}
-\caption{\acrfull{PDF} of the total transition cost.
+\caption{Distribution of the total transition cost over the 1260 runs of the \acrfull{GSA}.
 The mean, $\mu=1.18\cdot10^3$\,b€, is slightly higher than the median ($P_{50}=1.16\cdot10^3$\,b€) and the nominal cases cost, $1.08\cdot10^3$\,b€ and $1.04\cdot10^3$\,b€ respectively for the REF and SMR cases.
@@
 \caption{Distribution of the imported renewable electrofuels over the transition.
-Starting from no electrofuel in 2020, their respective import rises progressively along the transition (\ie increasing median) at different growth rates and with different ranges of values.
+Starting from no electrofuel in 2020, their respective import rises progressively along the transition at different growth rates and with different ranges of values.
 }
@@
 \includegraphics[width=0.8\textwidth]{UQ_Methanol_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-methanol in 2050.
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-methanol in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 \includegraphics[width=0.8\textwidth]{UQ_H2_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-hydrogen in 2050.
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-hydrogen in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 \includegraphics[width=0.8\textwidth]{UQ_Gas_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-methane in 2050.
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-methane in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 \includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
-\caption{Trend lines of the key parameters on the import of e-ammonia in 2050.
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-ammonia in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 \includegraphics[width=0.8\textwidth]{UQ_SMR_samples_2.pdf}
-\caption{Trend lines of the key parameters on the installed capacity of \gls{SMR} in 2050.
+\caption{Trend lines of the key parameters (and their Sobol' index) on the installed capacity of \gls{SMR} in 2050.
 Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
@@
 However, given the need to import molecules at earlier stages of the transition, with or without \gls{SMR}, it seems reasonable to keep on investing in the transport and the infrastructure to support these imports.
-This would allow covering ourselves against the risk of eventually not having \gls{SMR} available by the end of the first half of this century.
+In practice, Fluxys, the manager of the gas network in Belgium, already presents significant investments by 2032 (\ie 1.3\,b€) to support this transition in the near future \cite{Fluxys_2023}.
+The investments into the molecules would allow covering ourselves against the risk of eventually not having \gls{SMR} available by the end of the first half of this century.
 The perfect foresight approach does not model this risk as the possibility to install or not \gls{SMR} is known in advance.
 In the EnergyScope Pathway model, assuming the same technology mix as in the SMR case up to 2040 but without installing \gls{SMR} for the rest of the transition, \ie between 2040 and 2050, would require drastic (and quick) changes to still meet the \ce{CO2}-budget: around 44\,b€ of additional cumulative OPEX in purchasing electrofuels, mostly e-ammonia (+32\,TWh by 2050) and e-methane (+31\,TWh) to supply \gls{CCGT} and industrial \gls{CHP}.
-As a more concrete example, Fluxys, the manager of the gas network in Belgium, already presents significant investments by 2032 (\ie 1.3\,b€) to support this transition in the near future \cite{Fluxys_2023}.
-
-Chapters \ref{chap:chap_RL} and \ref{chap:chap_RobPol} put in practice the methodology detailed in Chapter \ref{chap:chap_methodo} to come up with a policy to get a system robust to this possibility of finally not having access to \gls{SMR}, as well as other uncertainties, especially when considering a myopic progression through the transition.
+%
+%Chapters \ref{chap:chap_RL} and \ref{chap:chap_RobPol} put in practice the methodology detailed in Chapter \ref{chap:chap_methodo} to come up with a policy to get a system robust to this possibility of finally not having access to \gls{SMR}, as well as other uncertainties, especially when considering a myopic progression through the transition.

### 2024-07-05 chap_case_study/chap_case_study.tex
Rebuttals filled in with comments
... now is the time to address them
Rebuttals: Keep going...

@@
 To align with other similar works on the modelling of whole-energy systems \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
-Regarding \gls{BEV}, while the CAPEX has been kept unchanged, the efficiency and the battery capacity, \ie the range, have been increased.
+Regarding \gls{BEV}, while the CAPEX has been kept unchanged, the efficiency and the battery capacity have been increased.
 As seen in the results, this change of data made \gls{BEV} often more competitive than its hydrogen-based equivalent.
@@
 \caption{Historical data for the non-energy use in Belgium \cite{statbel_NED_2019}.
-``Other oil products" account for tar and sulphur.
+``Other oil products'' account for tar and sulphur.
 This group also includes by-products of the oil-refining industry like aromatics (\eg \acrfull{BTX}) and olefins (\eg propylene).
@@
 \label{sec:cs:resources}
-To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$ (see \Cref{fig:cs_resources_cost}), their global warming potential, $\emph{gwp}_{\mathrm{op}}$, as well as their
-availability, as detailed by \citet{limpens2024pathway}.
+To supply the aforementioned demands, EnergyScope Pathway implements a variety of resources defined by their cost of purchasing, $\mathit{c}_{\mathrm{op}}$ (see \Cref{fig:cs_resources_cost}), their availability, as well as their global warming potential, $\emph{gwp}_{\mathrm{op}}$, as detailed by \citet{limpens2024pathway}.
 Regarding the cost of ``renewable electrofuels'', they are in line with the recent study of \citet{genge2023supply} who carried out an extensive review and ``meta-analysis \cite{grant2009typology,page2021prisma} of 30 studies on the supply costs of chemical energy carriers''.
@@
 
-Besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
+Regarding the imported renewable electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco).
+As the amount to import from each of these locations is hard to forecast, the current work considers the average cost between the different locations.
+
+Besides the costs, the resources are either limited or unlimited in terms of availability and either renewable or not.
 The limitation in terms of availability can be direct or indirect.
@@
 The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh.
-On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
+On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of rooftop \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
 In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}.
@@
 
-In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
+In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable'' in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
 In practice, it means that their \gls{GWP} is assumed to be zero in the model.
@@
 Therefore, the conclusions drawn in this thesis are not affected by this assumption.
-
-Regarding these electrofuels, the \citet{h2coalition} has carried out an extensive techno-economic analysis to estimate their respective cost of purchasing, after having identified some key locations from which importing these energy carriers (\eg Chile, Australia or Morocco).
-As the amount to import from each of these locations is hard to forecast, the current work considers the average cost between the different locations.
 Besides these, every other resource has its specific \gls{GWP} like coal ($\emph{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), natural gas ($\emph{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh) or the fossil-based molecules equivalent to the electrofuels (\eg $\emph{gwp}_{\mathrm{op,ammonia}}=0.46$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\emph{gwp}_{\mathrm{op,methanol}}=0.41$\,kt$_{\ce{CO2},\text{eq}}$/GWh).
@@
 Lifetime & 60 & year \\
+Cost of purchasing uranium & 4 & €/MWh\\
 Efficiency & 40\% & -\\
@@
 \caption{Schematic view of the different resources able to produce \gls{HVC}, ammonia and methanol with their related conversion technologies (including energy efficiency and their CAPEX - in €/kW of final molecules).
+The efficiencies account for the potential consumption/production of heat and/or electricity within each conversion process.
 Values stand for 2035.
@@
 Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}.
-The uncertainty characterisation gives the uncertainty ranges per parameter or group of parameters (category).
 
-This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the modal share change in different key sectors.
+This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the exogenous modal share change in different key sectors.
 A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.

### 2024-07-05 chap_intro_ccl/intro.tex
Rebuttals: Keep going...

@@
 \includegraphics[width=\textwidth]{IEA_WEO_2019.pdf}
-\caption{Energy-related CO2 emissions and reductions by sources in the Sustainable Development Scenario of the \citet{iea2020world}.}
+\caption{Energy-related \ce{CO2} emissions and reductions by sources in the Sustainable Development Scenario of the \citet{iea2020world}.}
 \label{fig:intro:IEA_WEO_2019}
@@
 \includegraphics[width=0.9\textwidth]{Storage_electrofuels.pdf}
-\caption{Energy carriers and technologies to store electricity. Electrofuels provide a solution for high capacity and long-term storage of energy. Graph adapted from \cite{ISPT2017}.}
+\caption{Energy carriers and technologies to store electricity. Electrofuels provide a solution for high-capacity and long-term storage of energy. Graph adapted from \cite{ISPT2017}.}
 \label{fig:intro:Storage_electrofuels}
@@
 \noindent
-This thesis is composed of five chapters to provide answers to the different research questions.
+This thesis is composed of five chapters to provide answers to the different research questions (see Figure \ref{fig:intro:Thesis_Structure}).
 Chapter \ref{chap:chap_methodo} details the tools and methods we used.
@@
 Perspectives are suggested for future research on further developments and uses of the methodological tools.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=\textwidth]{Thesis_Structure.pdf}
+\caption{Structure of the thesis.
+Beyond Chapter 2 that describes the case study of Belgium, Chapters 3, 4 and 5 collect the analyses resulting from the application of the different methodologies developed in Chapter 1.
+Given the (quasi-)independence of the analyses presented in the last three chapters, they can be read separately.}
+\label{fig:intro:Thesis_Structure}
+\end{figure}

### 2024-07-05 chap_methodo/chap_methodo.tex
Rebuttals: Keep going...

@@
 
-Assessing the robustness of a whole-energy system transition pathway calls a variety of methodological tools.
+Assessing the robustness of a whole-energy system transition pathway calls for a variety of methodological tools.
 First and foremost, such an extensive system needs to be represented, \ie modelled, to further be optimised.
@@
 Following an approach similar to \citet{guevara2022modeling}, we have extended the ranges of uncertainty developed by \citet{Moret2017} to the pathway optimisation.
-After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model subject to different emission-constraints \citet{rixhon2021role} as well as the pathway model.
+After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model subject to different emission-constraints \cite{rixhon2021role} as well as the pathway model (see Chapter \ref{chap:atom_mol}).
 Eventually, starting from the initial investigation of \citet{goffauxpathway}, this work has converged to the most relevant formulation of the salvage value for the model EnergyScope.
 This aims at considering the residual value of assets that would still be in place after the end of the optimisation.
-It avoids penalising capital-intensive and long-lasting asset.In this formulation, the capacities that have been anticipatively decommissioned are removed from the total installed capacities.
-Therefore,  Therefore, this penalises decisions that would lead to investments that are later decommissioned before having reached the end of their lifetime.
+It avoids penalising capital-intensive and long-lasting asset.
+In this formulation, the capacities that have been anticipatively decommissioned are removed from the total installed capacities.
+Therefore, this penalises decisions that would lead to investments that are later decommissioned before having reached the end of their lifetime.
 
@@
 According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is an open-source documented model.
-Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation).
+Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation)\footnote{To keep this low computational time, crossover was disabled to solve the optimisation problem.
+This had no observed impact regarding the values of the objective function or the design variables at the optimum.}.
 From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the former approach to further dig into more details about the latter.
-The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from, EnergyScope TD, the a snapshot model, optimising a target future year with a greenfield approach \cite{geidl2006greenfield}.
+The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from, EnergyScope TD, the snapshot model, optimising a target future year with a greenfield approach \cite{geidl2006greenfield}.
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
@@
 \includegraphics[width=\textwidth]{ES_Pathway.pdf}
-\caption{Illustration of the pathway methodology based on an existing energy system model.
+\caption{Illustration of the pathway methodology based on an existing energy system model, here EnergyScope TD \cite{limpens2019energyscope}.
 The methodology spans from 2020 to 2050, with one representative year every five years.
@@
 The formulation includes additional constraints (black boxes) that link the years together.
-The pathway's initialisation assumes that all capacities installed in 2020 were built during the pseudo-phase of 2015-2020 (grey box).
+The initialisation of the pathway assumes that all capacities installed in 2020 were built during the pseudo-phase of 2015-2020 (grey box).
 The overall problem is defined as the pathway model.}
@@
 \label{eq:PhaseInv}%5
-&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \emph{c\textsubscript{inv}}(p,j)&\forall p \in \emph{PHASE},
+&\textbf{C\textsubscript{inv,phase}}(p) = \sum_{\mathclap{j \in \emph{TECH}}} \textbf{F\textsubscript{new}}(p,j)\cdot \tau\textsubscript{\emph{phase}}(p)\cdot \emph{c\textsubscript{inv}}(p,j)&\forall p \in \emph{PHASE},\\
+\label{eq:path_annu_factor}
+&\tau\textsubscript{\emph{phase}}(p) = 1/(1+\emph{i\textsubscript{rate}})^{\emph{diff\_2015\_year(p)}}, &
 \end{flalign}
@@
 \noindent where $\tau\textsubscript{\emph{phase}}$ is the annualised phase factor and $\emph{c\textsubscript{inv}}(p,j)$ is the arithmetic average of the investment cost of the technology $j$ at the beginning and the end of the phase, $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{start}},j)$ and $\emph{c\textsubscript{inv}}(\emph{y\textsubscript{stop}},j)$.
-The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
+
+In EnergyScope Pathway, $\tau\textsubscript{\emph{phase}}$ aims at depreciating the value of expenses done in the future (see Equation (\ref{eq:path_annu_factor})).
+The discount rate accounted for in this annualisation factor, $i_{rate}$, is considered identical for all the technologies of the system.
+In line with \citet{limpens2021generating}, we set the nominal value to $i_{rate}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}.
+Having a low value of discount rate favours less late expenses compared to a higher value of discount rate.
 In practice, the discount rate would vary depending on the technology investment risk.
@@
 
-Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed (see Equation (\ref{eq:salvage}))
+Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed.
+After assessing the sensitivity of the formulation of the salvage value (see Appendix \ref{sec:path_methodo}), we have kept the one given by Equation (\ref{eq:salvage}), being also similar to the expression used by \citet{prina2019transition}.
 
@@
 Similarly to initial operational costs to account for the system in place in 2020 (see Equation (\ref{eq:Copex_tot_v2})), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
-Then, as detailed in Section \ref{sec:cs:CO2-budget}, these cumulative emissions is constrained by a budget (see Equation (\ref{eq:limit_gwp_trans})).
+Then, as detailed in Section \ref{sec:cs:CO2-budget}, these cumulative emissions are constrained by a budget (see Equation (\ref{eq:limit_gwp_trans})).
 
@@
 This approach is used by \citet{heuberger2018impact} who assessed the speed of integration of technologies due to these events.
-In their analysis of the overcapacity in European power systems, \citet{moret2020overcapacity} emphasised that such a ``possibility of \textit{recourse}" is very appropriate to address uncertainty gradually unfolding over time.
+In their analysis of the overcapacity in European power systems, \citet{moret2020overcapacity} emphasised that such a ``possibility of \textit{recourse}'' is very appropriate to address uncertainty gradually unfolding over time.
 Consequently, the development of the myopic approach with an overlap between two successive time windows has been implemented.
@@
 \includegraphics[width=\textwidth]{MY_schematic.pdf}
-\caption{The myopic approach (in pink) uses several instances of the pathway model (illustrated in Figure \ref{fig:meth_path_methodology}).
+\caption{The myopic approach (in pink) uses several instances of the pathway model, illustrated in Figure \ref{fig:meth_path_methodology_core}, as a sequence of different and linked \gls{LP} problems.
 In this example, the pathway instance has a time horizon of 10 years ($N\textsubscript{year,opti}=10$) with a 5 year-overlap ($N\textsubscript{year,overlap}=5$).
-As a comparison the Perfect foresight (in blue) has a time horizon of 30 years.}
+As a comparison the perfect foresight (in blue) has a time horizon of 30 years.}
 \label{fig:my_schematic}
@@
 \includegraphics[width=\textwidth]{MY_schematic_2.pdf}
-\caption{Sequential optimisation of the transition pathway in the myopic approach: (i) first time-window optimisation, (ii) set-up of the initial conditions of the second time-window, (iii) second time-window optimisation discarding intermediate results}
+\caption{Sequential optimisation of the transition pathway in the myopic approach: (i) first time-window optimisation, (ii) set-up of the initial conditions of the second time-window, (iii) second time-window optimisation discarding intermediate results from the first time-window optimisation.}
 \label{fig:my_schematic_2}
@@
 \begin{table}[htbp]
-\caption[New SETs for myopic pathway formulation.]{New SETs for myopic pathway formulation.} 
+\caption[New SETs for the myopic pathway formulation.]{New SETs for the myopic pathway formulation.} 
 \label{tab:path_my_sets}
@@
 Using this \gls{API}, this updated version of the model interacts with the AMPL problem representing the optimisation of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
-\\
+The myopic optimisation of the pathway consists in a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window, until eventually reaching 2050.\\
 
@@
 \includegraphics[width=10cm]{MY_process_code.pdf}
-\caption{Schematic of the iterative optimisation of the whole-energy system transition pathway.}
+\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
+Myopic optimisation of the pathway consists in a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window.
+If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters are changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
 \label{fig:MY_process_code}
@@
 Moreover, some ranges have been added to account for new parameters coming from the pathway formulation described in Section \ref{sec:meth:ES} like the society inertia.
-Like other works \cite{li2019renewables,coppitters2021robust} and given the scarcity of information, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
+Like other works \cite{li2019renewables,coppitters2021robust,Moret2017PhDThesis} and given the scarcity of information, the uncertain parameters are assumed to be independent and uniformly distributed between their respective lower and upper bounds.
 Alternatives like PERT or Gaussian distributions could also have been considered \cite{coppittersthesis}.
@@
 Finally, the model accounts for thousands of parameters.
-The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs\footnote{As detailed in Section \ref{subsec:pce}, the number of runs required for the \gls{GSA} is proportional to the factorial of the number of uncertain parameters.
-As second order \gls{PCE} is the minimum to ensure accuracy of the surrogate model, considering thousands of independent uncertain parameters would lead to millions of runs, if no more.}).
+The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs).
 Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped to one single uncertain parameter.
@@
 
+Section \ref{sec:cs:uncertainty} gathers the list of the uncertain parameters considered in this work.
+
 \subsection{Polynomial Chaos Expansion}
@@
 
-To avoid the avoid the computational burden of well-known method like Monte-Carlo analysis \cite{yue2018review}, we used \gls{PCE} to carry out a \gls{GSA}.
+To avoid the computational burden of well-known method like Monte-Carlo analysis \cite{yue2018review}, we used \gls{PCE} to carry out a \gls{GSA}.
 \gls{PCE} is an approach for surrogate-assisted \gls{UQ} that propagates uncertainties in input parameters through the system model.
-This allowed us to assess statistical moments on the quantity of interest and determine Sobol' indices~\cite{coppitters2020robust}.
+This allowed us to assess statistical moments of the quantity of interest and determine Sobol' indices~\cite{coppitters2020robust}.
 To construct a PCE of the EnergyScope Pathway model, we employed the open-source Python framework RHEIA~\cite{coppitters2022rheia,readthedocs_rheia}.
@@
 For the specific study of this work, a polynomial order of 2 is necessary (with 1260 training samples as per Equation\,(\ref{eq:pce:nterms})) to achieve a \gls{LOO} error below \SI{1}{\%} for the total transition cost.
+As second order \gls{PCE} is the minimum to ensure accuracy of the surrogate model, selecting and grouping the parameters as presented in Section \ref{subsec:screening} is compulsory to alleviate the computational burden.
+Otherwise, considering thousands of independent uncertain parameters would lead to millions of EnergyScope Pathway runs, if no more.
+Another alternative to reduce the computational cost of the \gls{GSA} would be to consider the monthly (instead of hourly) time resolution for the runs of EnergyScope Pathway.
+However, despite its shorter computation time (\ie couple of seconds), the main drawback of the monthly model is its poor representation of the integration of \gls{VRES} \cite{limpens2024pathway}.
 
@@
 
-In \cite{limpens2020impact}, we have assessed the \gls{PCE} approach, comparing the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$.
+In \cite{limpens2020impact}, we have assessed the \gls{PCE} approach on the cost-optimum Belgian energy system in 2035, using EnergyScope TD.
+To do so, we compared the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$.
 Even if the output of each method does not have the same physical meaning, both methods can rank the parameters by their impact on the total annual cost of the energy system.
-Both rankings were very similar which validates the use of \gls{PCE} in the rest of this work.
+Both rankings were very similar which validated the use of \gls{PCE} in the rest of this work (see Table \ref{tab:GSA:Comparison}).
+
+\begin{table}[htbp]
+	\caption{Comparison of the Top-14 rankings for the improved Morris method (left) and total-order \gls{PCE} method (right).
+We used the DTU's implementation of Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$).
+}
+    \label{tab:GSA:Comparison}
+    \centering
+    \begin{tabular}{lcc}
+    \toprule
+		\textbf{Parameter} & \textbf{Morris Ranking} ($\mu^*_{i}$) & \textbf{PCE Ranking} ($S_i^T$) \\
+        \midrule
+        Prices of hydrocarbons & 1 (0.7141) & 1 (0.6434) \\
+        Cost of increased efficiency & 2 (0.3046) & 2 (0.1014) \\
+        Cost of maintenance & 3 (0.2371) & 4 (0.0534) \\
+        Discount rate & 4 (0.2019) & 5 (0.0467) \\
+        Price of imported electricity & 5 (0.1974) & 3 (0.0809) \\
+        CAPEX of the grid & 6 (0.1498) & 6 (0.0252) \\
+        CAPEX of PV & 7 (0.1270) & 7 (0.0187) \\
+        Increase of electricity demand & 8 (0.1168) & 8 (0.0154) \\
+        CAPEX of nuclear power plant & 9 (0.0963) & 9 (0.0119) \\
+        Increase of space heating demand & 10 (0.0856) & 11 (0.0088) \\
+        Price of coal & 11 (0.0856) & 12 (0.0087)\\
+        Price of renewable fuels & 12 (0.0794) & 10 (0.0104)\\
+        Price of uranium & 13 (0.0607) & 13 (0.0047)\\
+        Onshore wind load factor & 14 (0.0600) & 14 (0.0046)\\
+       \bottomrule
+    \end{tabular}
+\end{table}
 
@@
 Then, its application to the myopic optimisation of a whole-energy system is introduced as well as the policy optimisation algorithm.
+As Chapter \ref{chap:case_study} presents the input of the case study and Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} on the pathway model, the reader interested by the results of the \gls{RL} method is invited to go to Chapter \ref{chap:chap_RL}.
 
@@
 The agent takes some action that has an impact on the environment which feeds back the agent with a reward and the new state.
-The objective of the agent is to optimise its policy, \ie mapping between the state it is at and the action to take, by maximising its cumulative reward.}
+The objective of the agent is to optimise its policy, \ie the mapping between the state it is at and the action to take, by maximising its cumulative reward.}
 \label{fig:RL_Fundamentals}
@@
 \includegraphics[width=0.8\textwidth]{Schematics_RL.pdf}
-\caption{\Acrfull{RL} framework made of the agent interacting with its environment, \ie the energy-system model on a limited decision window of 10 years.}
+\caption{The \Acrfull{RL} framework applied to the myopic optimisation of the energy transition pathway between 2020 and 2050.
+Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.}
 \label{fig:Schematics_RL}
@@
 After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of policies.
+The reader interested by the results of the \gls{PCA}-based method is invited to go to Chapter \ref{chap:chap_RobPol}.
 
@@
 At the same time, the average squared distance between the original observations and their respective projection is smaller for the projection on the left.
-Consequently, the left projection is closer to be the first PC than to the right projection.}
+Consequently, the left projection is closer than the right projection to be the first PC.}
 \label{fig:PCA_intro_proj}
@@
 \caption{Cumulative share of the sum of the transition variance versus the number of $\text{PC}_{\text{transition}}$.
-Passing a certain threshold, here 85\%, the number of $\text{PC}_{\text{transition}}$ to consider increase significantly compared to the share of transition variance that they capture.}
+Passing a certain threshold, here 85\%, the number of $\text{PC}_{\text{transition}}$ to consider increases significantly compared to the share of transition variance that they capture.}
 \label{fig:PC_y_share_tot_var}

### 2024-07-08 rebuttals/rebuttals.tex
Go, go, go...

@@
 Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
 The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Interest rate must be revisited if a paper is published.
+\end{mdframed}
+
+\noindent See previous comments.
+
 \subsubsection{Power grid and transport infrastructure}
@@
 {\color{purple} \textbf{Sylvain}} - How do you model the power grid interconnections with the neighbouring countries as well as the transport network of the different energy carriers within the country?
-\end{mdframed}
-
-\noindent Copper-plate assumption $\rightarrow$ no representation of the transport sector of either electricity or gaseous and liquid energy carriers.
-However, one technology called ``GRID'' accounts for the additional investments in the power grid needed when deploying more VRES like PV panels and wind turbines but there is not an actual spatial representation of this grid.
-{\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+p.63 - The model decides to use ammonia instead of SNG $\rightarrow$ is the transport infrastructure accounted for?
+p.73 - How is the transport infrastructure accounted for?
+\end{mdframed}
+
+\noindent Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes.
+The whole-energy system of the country is represented as a single-node and the imported energy commodities are available ``at the front door'' of the system.
+It goes the same way for the infrastructures to transport energy carriers within the system.
+As it is represented assuming a copper-plate approach, all the demands/resources/conversion technologies are concentrated in one single node.
+This is now explicitly written {\color{blue}in a new paragraph of Section 2.2}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes.
+The imported energy commodities are available ``at the front door'' of the system.
+In terms of spatial resolution of the country itself, it is modelled as a single-node.
+This is similar to the copper plate model of a power grid.
+In other words, the infrastructures to transport the energy carriers within the country are not considered.
+It is assumed that the demands have to be supplied by the production, regardless of the flows between the producers and the consumers.
+Yet, adapting the networks is accounted for in terms of the required investments.
+For instance, a high share of VRES requires an investment to reinforce the power grid (\ie 368\,M€/GW of additional installed capacity of VRES).
 \end{mdframed}
@@
 \end{figure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - SMR must be revisited if a paper is published.
+\end{mdframed}
+
+\noindent See previous comments.
 
@@
 {\color{orange} \textbf{Stefano}} - p. 71, Table 3.1: isn’t the low impact simply due to the low assumed potential? Is this finding meaningful?
-\end{mdframed}
-
-\noindent This low assumed potential is indeed one of the reasons why the SMR-related uncertain parameters have a limited impact on the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
+
+\noindent {\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
+\end{mdframed}
+
+\noindent These assumed low potential and CAPEX are indeed some reasons why the SMR-related uncertain parameters have a limited impact on the variation of the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
 However, this is not the only reasons as detailed earlier in the same paragraph: the late potential availability (2040 at the soonest) as well as the ``cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.''.
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 65 - Decides to use ammonia instead of SNG => is the transport infrastructure accounted for?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent In the continuity of Gauthier Limpens' thesis  and in line with the rest of the research field, the objective of this thesis is to ``illustrate the priorities to be taken in each of the sectors and the efforts to make in one sector to facilitate the integration of certain technologies into other sectors'' \cite{limpens2021generating}.
+This work highlights the ``must-do'' or ``no-regret investments'' to respect the (very) ambitious \ce{CO2} budget imposed to the model.
+In this sense, besides the maximum potential of 59\,GW solar-PV, we have not made assumption on the rate of deployment.
+15\,GW of PV in 2025 can be considered as very optimistic and Elia forecasts to have these 15\,GW by 2030 \cite{Elia_2024_2034}.
+I agree that not limiting the rate of deployment of PV (and wind turbines) is a strong assumption.
+For this reason, in the perspectives of this work (see Conclusion), I have already pointed this out as ``a future work would be to constrain the deployment of local renewables.
+
+Given these explanations, I have not brought further modification to the manuscript in this regard.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent There is indeed some curtailment, which remains very limited compared to the total production of solar-PV: 0.3\,TWh versus 61\,TWh.
+This information has been added {\color{blue}in the paragraph dedicated to ``High-temperature heat'' of Section 3.1.3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
 \end{mdframed}
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent The difference between BEV and industrial electric heaters is that BEV is already a preferred option for electrification, even the REF case.
+BEV completely substitute diesel and gasoline-cars as soon as possible in the transition and there cannot be more cars as private mobility is limited to supply up to 50\% of the demand of passengers mobility.
+Unlike high-temperature heat, there is no modal swap due to the installation of SMR.
+
+Given these explanations, I have not brought further modification to the manuscript in this regard.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 73 - How is the transport infrastructure accounted for?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent In general, the model gives fuel-cell trucks as the favoured option for the future of road freight transport given more cost-competitive options for electrification like private mobility and low-temperature heat.
+However, as illustrated in the ROB roadmap (see Section 5.2.1), battery trucks seem to be the cost-optimal solution to transition between diesel and either hydrogen trucks, when fossil and electrofuels are assumed to be expensive.
+
+In the Global Sensitivity Analysis, electric trucks are the favoured option for a limited set of samples, about 10\% (see Appendix D.3).
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent The threshold locates at 60\% of the uncertain parameter $f_{\mathrm{max,SMR}}$ which represents the potential capacity of SMR.
+As presented in Section 2.4, this threshold corresponds to the possibility to install up to 6\,GW of SMR in 2050.
+As already written in Section 3.2.2: ``For the installed capacity of SMR, it is the availability of the technology that drives its installation (Figure 3.12).
+Not shown here but all the samples of the GSA highlight that SMR is installed to its maximum capacity, \ie 6GW, as soon as possible.'' To smooth the trend curves of the different uncertain parameters, I have applied a rolling average.
+For this reason, the 60\%-threshold does not appear on the Figure as sharp as it actually is.
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
 
@@
 \noindent Perhaps reinforce the message as you answered the question $\rightarrow$ We need to go for radical changes, a “unicorn” solution, and one option would be the import of renewable molecules {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - SMR / interest rate must be revisited if a paper is published.
-\end{mdframed}
-
-\noindent {\color{blue} }.
 
@@
 
-\noindent
+\noindent Indeed, Figure 4.6 is less revealing as it shows no big trend in terms of neither impacting actions nor their timing in the transition.
+It is detailed in the first paragraph of Section 4.2.3 after which the figure should appear in the manuscript.
+I have worked to make Figures as close as possible to their related paragraph.
+However, in some cases, several other paragraphs come between them.
+To identify the actions that emerge to actually matter, we need to check if they are binding or not, as detailed in the subsequent paragraphs and in Figure 4.7.
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard, except minor adjustment to get Figure 4.6 as close as possible to its related paragraph.
 
@@
 
-\noindent To define in Section 1.3.2 and remind here {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent In the context of Reinforcement Learning, a step consists in taking an action, collecting a reward based on the previous state and the taken action and ending in a new state.
+Even though it was already mentioned in Chapter 1, the fact that the results come further from the methodology part makes it harder for the reader to remember the meaning of a step.
+It is now more explicitly defined {\color{blue}in the first paragraph of Section 1.3.2}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+From this state, the agent takes a step: action, reward, new state.
+\end{mdframed}
+
+\noindent and reminded {\color{blue}in the first sentence of Section 4.2.1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+The learning phase has been split into batches of 500 steps,\ie 500 sequences of state-action-reward-new state. 
 \end{mdframed}
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent Correct. This confirms what \citet{vogt2018starting} detailed in their works. To emphasise on this point, I have added one sentence {\color{blue}in Section 4.2.3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+In line with \citet{vogt2018starting}, the early constraints on generation and the sharper decrease of the emissions avoid lock-in situations.
 \end{mdframed}
@@
 
-\noindent  {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent No-go zones are introduced in the second paragraph of Chapter 4: ``[...] as well as no-go zones where the chances of succeeding the transition, i.e., respecting the \ce{CO2}-budget, are very limited.'' It is then detailed at the end of Section 4.2.2: ``Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except
+if betting on the future installation of SMR. '' and in the caption of the related figure. Given the information already present in the manuscript, no modification has been brought in this regard.
 
@@
 \noindent
-Besides the information given in Section 4.2.3 of the manuscript (and reminded here below), I do not see further explanations that could clarify the concept of binding constraint in a Linear Programming (LP) problem.
+Besides the information given in Section 4.2.3 of the manuscript, I do not see further explanations that could clarify more the concept of binding constraint in a Linear Programming (LP) problem.
 Consequently, regarding this comment, there has not been further modification brought to the manuscript.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-To identify the actions that have an actual impact on the environment, we can check if they are binding or not.
-In a LP problem, constraints represent hyperplanes in the domain of variables.
-In a two-dimension space, these are straight lines (see Figure 4.7).
-When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility.
-The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function.
-Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge.
-In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}.
-The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal.
-In other words, binding constraints, when tightened, aggravate the objective value function.
-If these are inequality constraints, as represented in Figure 4.7, it means that their left and right sides of the equation are equal.
-\end{mdframed}
 
@@
 119: the statement mentioning the ``major added value of the perfect foresight'' approach seems surprising and needs to be clarified.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - The conclusion is a summary and should rather be a discussion (\eg discard coal as soon as possible is obvious).
-\end{mdframed}
-
-\noindent To explain in the definition of the action {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+The decreasing \ce{CO2} emissions trajectories shouldn't be a S-shaped (logistic) curve?
+\end{mdframed}
+
+\noindent I agree that this sentence and what came next focused on the perfect foresight approach whereas the rest of the paragraph rather targets the RL-based myopic results.
+Therefore, I have reformulated {\color{blue}the end of this paragraph} and focused on the shape of the decreasing \ce{CO2} trajectories.
+These are indeed S-shaped curves but that are sharper for the myopic approach than in the PF results.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+This work also highlighted a major difference between the perfect foresight (PF) approach and the agent-based myopic transitions.
+Given the assumed full knowledge on the future, the trajectories to decrease the \ce{CO2} emissions resulting from perfect foresight are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - The conclusion is a summary and should rather be a discussion.
+\end{mdframed}
+
+\noindent I decided to split the conclusion into four blocks of paragraphs.
+The first two (from page 117 to the top of page 120, in the version of the manuscript submitted before the private defense) are indeed a summary of (1) what we have done in this thesis (case study and methodology) and (2) the results we have got.
+The two last blocks, however, take a step back from the results and aim at drawing main conclusions, discussing them and providing decision-makers with further insightful information and methods to plan the energy transition.
+Consequently, I have not brought modification to the manuscript in this regard.
 

### 2024-07-08 appendices/app_EnergyScope.tex
Go, go, go...

@@
 
-In Equation (eq:salvage_app), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
+In Equation (\ref{eq:salvage_app}), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
 With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost.
@@
 On top of this, we also investigated the impact of accounting for the decommissioned technologies or not in the expression of the salvage value.
-In Equation (eq:salvage_app), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0.
+In Equation (\ref{eq:salvage_app}), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0.
 This is relevant since a technology decommissioned before 2050 would not be available after 2050.

### 2024-07-08 chap_RL/chap_RL.tex
Go, go, go...

@@
 
-The learning phase has been split into batches of 500 steps.
+The learning phase has been split into batches of 500 steps,\ie 500 sequences of state-action-reward-new state.
 %For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
@@
 Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
 Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
@@
 Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
 Besides this wide exploration, successes and failures indistinguishably spread over the whole ranges of actions.
@@
 Compared to other non-renewable fuels, this is due to the longer use of this energy carrier favoured by its low \gls{GWP} (the second after uranium) and its versatility (applications in the electricity, heat and mobility sectors).
+In line with \citet{vogt2018starting}, the early constraints on generation and the sharper decrease of the emissions avoid lock-in situations.
 

### 2024-07-08 chap_atom_mol/chap_atom_mol.tex
Go, go, go...

@@
 The main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
-In the REF case, industrial electric heaters are mainly used to absorb, instead of curtailing, the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days.
+In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
 With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity.

### 2024-07-08 chap_case_study/chap_case_study.tex
Go, go, go...

@@
 
+Regarding the power grid interconnections with the neighbouring countries, similarly to the import of other energy carriers, there is no representation of those physical routes.
+The imported energy commodities are available ``at the front door'' of the system.
+In terms of spatial resolution of the country itself, it is modelled as a single-node.
+This is similar to the copper plate model of a power grid.
+In other words, the infrastructures to transport the energy carriers within the country are not considered.
+It is assumed that the demands have to be supplied by the production, regardless of the flows between the producers and the consumers.
+Yet, adapting the networks is accounted for in terms of the required investments.
+For instance, a high share of VRES requires an investment to reinforce the power grid (\ie 368\,M€/GW of additional installed capacity of \gls{VRES}).
+
 In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable'' in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}.
@@
 They encompass, for instance, the power grid, the \gls{DHN} or technologies to produce intermediate energy carriers (\eg wood pyrolysis, biomethanolation or steam methane reforming to produce hydrogen).
-Not digging into too much details about the exhaustive list of these technologies presented in a previous work \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} and the technologies to supply the \glsxtrfull{NED}.
+
+Not digging into too much more details about the exhaustive list of these technologies presented in a previous work \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} and the technologies to supply the \glsxtrfull{NED}.
 

### 2024-07-08 chap_intro_ccl/conclusion.tex
Rebuttals: Keep going...
Go, go, go...

@@
 
-By 2050, in 88\%of the transitions, e-hydrogen is only imported from abroad rather than locally produced.
+By 2050, in 88\% of the transitions, e-hydrogen is only imported from abroad rather than locally produced.
 This molecule supports the defossilisation of the freight-transport by trucks.
@@
 An illustrative example is the case of e-ammonia, for which imports by 2050 are in general twice larger in the myopic transition.
-When compared to agent-based myopic transitions, this work also highlighted the major added-value of the perfect foresight approach: the robustness towards uncertainties.
-Given its assumed full knowledge on the future, PF finds a solution to respect the \ce{CO2}-budget whatever the values taken by the uncertain parameters.
-Moreover, the trajectories to decrease the \ce{CO2} emissions are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
+This work also highlighted a major difference between the perfect foresight (PF) approach and the agent-based myopic transitions.
+Given the assumed full knowledge on the future, the trajectories to decrease the \ce{CO2} emissions resulting from perfect foresight are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
 

### 2024-07-08 chap_methodo/chap_methodo.tex
Go, go, go...

@@
 Even though an observation is a subset of the state, this work uses these two words interchangeably.
-Then, it takes an action, $\bm{a}_0$, impacting its environment, \ie the energy system limited transition over the first decision window (2020-2030).
+From this state, the agent takes a step: action, reward, new state.
+The action, $\bm{a}_0$, impacts the environment, \ie the energy system limited transition over the first decision window (2020-2030).
 Through this interaction with its environment, the agent is given a reward, $r_1=r\left(\bm{a}_0 | \bm{o}_0 \right)$, and ends up in a new state, \ie the energy system in 2025, characterised by a new observation, $\bm{o}_1$, and so on (see Figure \ref{fig:Schematics_RL}).

### 2024-07-09 rebuttals/rebuttals.tex
Allez, allez,...

@@
 
-\subsubsection{End-of-time-horizon}
+\subsubsection{End-of-time-horizon, myopic pathway and salvage value}
 
@@
 
-% \noindent Better highlight the importance, discuss the finetuning, discuss  alternatives (e.g. adding time windows at the end that are thrown afterwards).
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\subsubsection{Myopic pathway and salvage value}
+ \noindent To cope with these ``end-of-time-horizon effects'', we have implemented the salvage value, $\textbf{C\textsubscript{inv,return}}$.
+This represents the residual investment that would remain after 2050.
+As detailed in Section 1.1.1: ``the salvage value avoids penalising the capital intensive technologies towards the end of the transition'' \cite{poncelet2016myopic}.
+The sensitivity analysis on the formulation of the salvage value has been added in Appendix B.2.2 (see further comments on the salvage value).
+
+As discussed during the private defense, instead of implementing a salvage value another alternative would be to implement artificial representative years after 2050 that are then excluded from the total transition cost and emissions in while post-processing the results of the optimisation.
+Even though the choice of the formulation of the salvage value accounts for assumptions (discussed in Appendix B.2.2), we opted for this rather than ``artificial future years'' for several reasons.
+First of all, it allows mitigating the computational burden of the pathway problem.
+Indeed, the more there are representative years, the bigger are the matrices and the longer it takes to solve the problem.
+Second, the salvage value-approach is also used in other works \cite{poncelet2016myopic,prina2019transition}.
+Finally, implementing ``artificial  future years'' would require to either collect more data, which is quite time-consuming, or make other kind of assumptions on the evolution of costs, GWP and other features beyond 2050.
+
+Since the implementation of the salvage value and the formulation choices is the work developed by \citet{limpens2024pathway} (partly integrated in Appendix B) and given the above explanations, I have not brought further modification to the manuscript in this regard.
 
@@
 Is that a desirable feature to mimic the way we proceed now.
-why artificially limit the model as we know we have to go to zero in 2050.
-\end{mdframed}
-
-\noindent Link to salvage value
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+Why artificially limit the model as we know we have to go to zero in 2050?
+\end{mdframed}
+
+\noindent As written in the manuscript and discussed during the private defense, the purpose of the myopic approach is to challenge the assumption of the complete knowledge about the uncertain future and mimic the short-sightedness of the decision-making process.
+This sequential optimisation allows showcasing robust policies to support myopic transitions (see Chapter 4) and the robustness of roadmaps (see Chapter 5).
+The myopic approach is also the methodological basement to support the implementation of the RL and PCA-based methods.
+
+Given these explanations and what is already in the manuscript, I have not brought modification in this regard.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent As detailed by \citet{Moret2017PhDThesis}: ``the ranges resulting from the uncertainty characterisation, are defined for a specific point in time in the planning horizon, usually its last year.
+Therefore, they need to be scaled accordingly when used in the optimisation model.
+Without this scaling, it would mean assuming that the uncertainty for year $N$ stands for all the years in the planning horizon''.
+In his thesis, Moret considered $N=20$ as the duration of the planning horizon between 2015, the year of Moret's work, and the snapshot optimisation of the Swiss case study in 2035.
+In my case, the planning horizon is 5-year long as the duration of a phase between two representative years of the transition.
+Referring to \citet{Moret2017PhDThesis} for the interested reader, I have explained a bit more the reason for this scaling factor {\color{blue}in the second paragraph of Section 2.4}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined ranges of uncertainties for the model parameters.
+For the operation-type parameters (types II and III), a scaling factor needs to be considered to avoid assuming that the uncertainty for the last year of the planning horizon ($N$) is representative for all years in the planning horizon \citet{Moret2017PhDThesis}.
+In EnergyScope Pathway, as the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III.
 \end{mdframed}
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent The need of using NN is in the learning phase of the RL approach, not in running EnergyScope.
+The justification to use NN is already detailed in Section 1.3.2: ``Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of RL with Deep Neural Network (DNN), called Deep Reinforcement Learning (DRL).
+Among others, one of the main drawbacks of traditional RL algorithms, \ie without the use of Neural Network (NN),  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter 4).
+By approximating the state-action function with its parameters (\ie weights and biases), DNN can address this difficulty.
+''
+
+Given what is already written in the manuscript, I have not brought modification in this regard.
 
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Indeed, we know the input parameters, the constraints and the objective function implemented in EnergyScope to optimise the transition pathway of the whole-energy system.
+In the RL framework, a model is representation of how the environment will respond to the agent's actions.
+Unlike other cases such as grid world\footnote{\url{https://towardsdatascience.com/reinforcement-learning-implement-grid-world-from-scratch-c5963765ebff}} or chess, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
+Going for a model-free approach, we avoid building a model within the agent to inform it about this (complex) dynamics.
+
+This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' Given what is already written in the manuscript, I have not brought modification in this regard.
 
@@
 {\color{orange} \textbf{Stefano}} - I am a bit underwhelmed by the many nitty-gritty data in this Chapter.
-While it is surely important to document everything, couldn’t at least part of this be moved to the Appendix?\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+While it is surely important to document everything, couldn’t at least part of this be moved to the Appendix?
+\end{mdframed}
+
+\noindent Most of the data about the Belgian case study have been collected by \citet{limpens2021generating}.
+The objective of this Chapter was to focus on my contributions: non-energy demand (NED), electrofuels, SMR, uncertain ranges for the pathway model and \ce{CO2}-budget.
+However, to put these contributions in perspective with similar features in the model, I have decided to keep general information about the end-use demands, the resources and the conversion technologies.
+Consequently, I have kept the same level of information in Chapter 2 rather than in appendix.
 

### 2024-07-09 chap_case_study/chap_case_study.tex
Allez, allez,...

@@
 
-In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined range of uncertainties for the model parameters.
-As the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III, based on the work of \citet{Moret2017PhDThesis}.
+In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined ranges of uncertainties for the model parameters.
+For the operation-type parameters (types II and III), a scaling factor needs to be considered to avoid assuming that the uncertainty for the last year of the planning horizon ($N$) is representative for all years in the planning horizon \citet{Moret2017PhDThesis}.
+In EnergyScope Pathway, as the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III.
 For type III uncertainties (\ie uncertainty ranges increasing with time), a 50\% increase has been set arbitrarily between the ranges for 2025 and these same ranges for 2050.

### 2024-07-09 chap_methodo/chap_methodo.tex
Allez, allez,...

@@
 Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed.
+To cope with the end-of-time-horizon effects of multi-stages models, the salvage value avoids penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}.
 After assessing the sensitivity of the formulation of the salvage value (see Appendix \ref{sec:path_methodo}), we have kept the one given by Equation (\ref{eq:salvage}), being also similar to the expression used by \citet{prina2019transition}.
@@
 Similarly, the uncertainties impacting the industrial demand, $industry\_\emph{EUD}$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand.
-
 Section \ref{sec:cs:uncertainty} gathers the list of the uncertain parameters considered in this work.
@@
 Starting from an initial state, the agent takes an action that impacts its environment.
-The latter feeds back the agent with a reward and the new state (see Figure \ref{fig:RL_Fundamentals}).
+The latter feeds back the agent with a reward and the new state.
 This goes on until reaching the end of the episode.

### 2024-07-11 rebuttals/rebuttals.tex
Almost there

@@
 \begin{itemize}
-\item \emph{[40] Limpens et al., Energyscope pathway: an open-source model to optimise the energy transition pathways of a regional whole-energy system}: Mostly cited to give credits to Gauthier Limpens' work in developing the Pathway version of EnergyScope, I have decided to refer to the paper about the formulation choices rather than including them in the thesis manuscript.
+\item \emph{\cite{limpens2024pathway} Limpens et al., Energyscope pathway: an open-source model to optimise the energy transition pathways of a regional whole-energy system}: Mostly cited to give credits to Gauthier Limpens' work in developing the Pathway version of EnergyScope, I have decided to refer to the paper about the formulation choices rather than including them in the thesis manuscript.
 Consequently, no further modification regarding this reference.
-\item \emph{[48] Limpens et al., The impact of uncertainties on the Belgian energy system: Application of the Polynomial Chaos Expansion to the EnergyScope model}: The table gathering the results of the comparison between Morris' and PCE methods has been included {\color{blue}at the end of Section 1.2.2}:
+\item \emph{\cite{limpens2020impact} Limpens et al., The impact of uncertainties on the Belgian energy system: Application of the Polynomial Chaos Expansion to the EnergyScope model}: The table gathering the results of the comparison between Morris' and PCE methods has been included {\color{blue}at the end of Section 1.2.2}:
 \end{itemize}
@@
 \begin{itemize}
-\item \emph{[49] Rixhon et al., The role of electrofuels under uncertainties for the Belgian energy transition}: This paper is cited to highlight the GSA performed on the snapshot model with different GWP limits.
+\item \emph{\cite{rixhon2021role} Rixhon et al., The role of electrofuels under uncertainties for the Belgian energy transition}: This paper is cited to highlight the GSA performed on the snapshot model with different GWP limits.
 The main contribution of this work in the context of the thesis was to provide a preliminary screening and selection of parameters to carry out the GSA on the pathway model.
@@
 Consequently, no further modification regarding this reference.
-\item \emph{[104] Rixhon et al., Integration of non-energy among the end-use demands of bottom-up whole-energy system models}: Reference used to highlight the fact that whole-energy systems usually do not account for the non-energy demand.
+\item \emph{\cite{rixhon2022integration} Rixhon et al., Integration of non-energy among the end-use demands of bottom-up whole-energy system models}: Reference used to highlight the fact that whole-energy systems usually do not account for the non-energy demand.
 Consequently, no further modification regarding this reference.
-\item \emph{[116] Rixhon et al., Comprehensive
+\item \emph{\cite{rixhon2021comprehensive} Rixhon et al., Comprehensive
 integration of the non-energy demand within a whole-energy system: towards a defossilisation of the chemical industry in Belgium}: Reference to graphs that have either been adapted from the paper or directly taken from it.
 Consequently, no further modification regarding this reference.
-\item \emph{[128] Rixhon et al., Taxonomy of the fuels in a whole-energy system}: Where the content of this paper is the Appendix A of the thesis, some references to this paper aimed at clarifying the meaning of ``renewable fuels'' (Section 2.2) or the fact that uranium is considered neither fossil nor renewable (Sections 3.1.2 and 4.2.2).
+\item \emph{\cite{rixhon2021terminology} Rixhon et al., Taxonomy of the fuels in a whole-energy system}: Where the content of this paper is the Appendix A of the thesis, some references to this paper aimed at clarifying the meaning of ``renewable fuels'' (Section 2.2) or the fact that uranium is considered neither fossil nor renewable (Sections 3.1.2 and 4.2.2).
 Consequently, no further modification regarding this reference.
@@
 In the RL framework, a model is representation of how the environment will respond to the agent's actions.
-Unlike other cases such as grid world\footnote{\url{https://towardsdatascience.com/reinforcement-learning-implement-grid-world-from-scratch-c5963765ebff}} or chess, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
+Unlike other cases such as grid world, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
 Going for a model-free approach, we avoid building a model within the agent to inform it about this (complex) dynamics.
@@
 This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' Given what is already written in the manuscript, I have not brought modification in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - Justification of the actions selected? why no other actions like the cost of some technologies? 
-\end{mdframed}
-
-\noindent These actions have a direct translation in constraints and we can therefore assess their effectiveness through the fact they are binding or not.
-Give examples of actions tried during the work (incentivize renewables) or other from Assemblée citoyenne in France and explain why they’re not as good as those chosen in the thesis.
-Explain history of what we tried/consider?
- {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - How can you account for pivotal events / shocks in the system?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Even though Reinforcement Learning can be applied to problems with infinite time-horizon discounting reward coming further in the future, it is not the case in my thesis.
+Indeed, each episode (\ie the transition) is finite.
+The purpose of Figure 4.1 was to focus on the reward function and how it articulates in the action-state-reward process.
+As already represented in this figure, three situations can terminate an episode: (i) infeasible optimisation, (ii) overshot \ce{CO2}-budget before 2050 or, (iii) reaching 2050.
+This is now clarified {\color{blue}in the caption of Figure 4.1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Reward function, $R$.
+Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded.
+If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues.
+Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates.
+After terminating an episode, the process starts over at the initial state, \ie 2020.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - How can you account for pivotal events/shocks in the system?
+\end{mdframed}
+
+\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/shocks can be implemented through uncertain parameters.
+For instance, this is how we have implemented the potential to install SMR in 2040 at the soonest.
+Given the probability described in Section 2.4, the technology can (or not) be installed.
+However, as detailed in Section 1.2.1, for the other uncertain parameters, they keep the same value relatively to their type and range for the whole transition.
+In other words, we do not allow zig-zag where, for instance, a cost of purchasing an energy carrier was low for the first half of the transition and, suddenly, in 2040, becomes very high.
+
+Given these explanations and what is already written, I have not brought modification to the manuscript in this regard.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Like other figures in this Chapter, Figure 4.3 represents distributions of features spreading over the x-axis.
+In this case, these are distributions of the rewards collected by the agent at the end of each episode of its learning process.
+For this reason, labelling the axis does not seem obligatory as the x-axis is nothing else than the feature (here, the reward) and the y-axis has no physical meaning expect representing how often the feature of interest took a certain value/range of values.
+Compared to the other similar figures of the chapter, I have not labelled the x-axis as, for Figure 4.3, the reward is the only feature represented whereas two or more features are gathered on the same figure.
+
+With this in mind and what is already written in the caption of the figure and the core of the text, we can infer from the right hand side that 2040 is the ``tipping year'' as it represents the year at which the failures occur more often, compared to the other representative years.
+
+Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Complementary to the text in the core of the manuscript, I have added a diagram to detail the agent's actions and how they impact the model: taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Schematic_actions.pdf}
+\caption{Actions available to the decision-maker.
+Taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.}
+\label{fig:Schematic_actions}
+\end{figure}
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2}-budget) by 2050 at the latest.
+The number of remaining attempts reduce as prematurely ended episode are removed from the further counts.
+The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves.
+These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
+\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
 \end{mdframed}
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Both, agent's actions and exogenous uncertainties, change the value of some parameters of EnergyScope Pathway.
+However, where the sample of uncertainties is drawn once and for all at the beginning of the episode as explained in Section 1.2.1, the agent's actions are taken at the beginning of each 10-year time window.
+This is now clarified {\color{blue}at the beginning of Section 1.3.2 and the updated version of Figure 1.7, and its caption} (see Figure \ref{fig:Schematics_RL} in the present document):
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Before starting an episode, a sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1. 
+\end{mdframed}
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Schematics_RL.pdf}
+\caption{The Reinforcement Learning (RL) framework applied to the myopic optimisation of the energy transition pathway between 2020 and 2050.
+Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.
+At the beginning of each episode, one different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1.}
+\label{fig:Schematics_RL}
+\end{figure}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Justification of the actions selected? why no other actions like the cost of some technologies? 
+\end{mdframed}
+
+\noindent These actions have a direct translation in constraints and we can therefore assess their effectiveness through the fact they are binding or not.
+Give examples of actions tried during the work (incentivize renewables) or other from Assemblée citoyenne in France and explain why they’re not as good as those chosen in the thesis.
+Explain history of what we tried/consider?
+ {\color{blue} }.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - p.

### 2024-07-11 chap_RL/chap_RL.tex
Almost there

@@
 where $\mathrm{act}\textsubscript{fossil gas}$, $\mathrm{act}\textsubscript{LFO}$ and $\mathrm{act}\textsubscript{coal}$ can take values between 0 and 1.
-These complete the action space of the agent, $A\in \mathbb{R}^4_{[0,1]}$.
+These complete the action space of the agent, $A\in \mathbb{R}^4_{[0,1]}$ (see Figure \ref{fig:Schematic_actions}).
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{Schematic_actions.pdf}
+\caption{Actions available to the decision-maker.
+Taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.}
+\label{fig:Schematic_actions}
+\end{figure}
 
@@
 If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues.
-Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
+Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates.
+After terminating an episode, the process starts over at the initial state, \ie 2020.}
 \label{fig:Reward}
@@
 \begin{table}[htbp!]
-\caption{Uncertain parameters impacting the most the total transition cost, their Sobol' index and, for the successful transitions, their mean of their values between 0 and 100\%, $\mu$, and their skewness, $\gamma$.
+\caption{Uncertain parameters impacting the most the total transition cost and, for the successful transitions, the mean of their values between 0 and 100\%, $\mu$, and their skewness, $\gamma$.
 On top of being supported by the agent's actions, successful transitions occur when the cost of purchasing fossil fuels is more increased than the one of electrofuels.}

### 2024-07-11 chap_methodo/chap_methodo.tex
Almost there

@@
 \label{subsec:meth_RL_algo}
+Before starting an episode, a sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section \ref{subsec:uncert_charac}.
 At the initial state, \ie the energy system in 2020, the agent gets an initial observation, $\bm{o}_0$.
@@
 \caption{The \Acrfull{RL} framework applied to the myopic optimisation of the energy transition pathway between 2020 and 2050.
-Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.}
+Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.
+At the beginning of each episode, one different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section \ref{subsec:uncert_charac}.}
 \label{fig:Schematics_RL}

### 2024-07-16 rebuttals/rebuttals.tex
Getting there...

@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\noindent The purpose of each methods and the assumptions are discussed in the ``Assumptions - Fine tuning'' section of the present document.
+Regarding the choice of PCE instead of monthly version for uncertainty, as discussed during the private defense, monthly version of the model does not provide enough detail to provide sufficient confidence in the results given the need to massively integrate VRES for Belgium to succeed its transition.
+Given the hourly/daily intermittency, properly assessing the integration of wind and solar requires a fine time resolution that is not provided by the monthly model.
+
+Given what is detailed further in this document and what has been discussed during the private defense, I have not brought modification to the manuscript in this regard.
 
@@
 You need a model that runs pretty fast, right?
-\end{mdframed}
-
-\noindent Perhaps there is a need to better prove or show that it is not a magic number.
-How to generalise it and hook it to something physical (even if it is done a posteriori).
-Guidelines to apply it to some other models
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
+
+{\color{teal} \textbf{Stefan}} - Justification of the RL actions selected? why no other actions like the cost of some technologies?
+
+{\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?
+
 {\color{purple} \textbf{Sylvain}} - Make sure to discuss the hypotheses because one might not agree with all of them.
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
+\noindent I have not carried out analysis to assess the sensitivity of the results to these assumptions.
+They result either from trial and error or observations made \textit{a posteriori}.
+However, to discuss these assumptions and the methodological choices, I have added a specific section to each methodological parts of Chapter 1 to provide future users and developers of the methods proposed in this work with guidelines.
+These aim at helping other researchers in either using the same methodologies and assumptions on their own case study or keep on developing and improving these methodologies.
+
+\begin{mdframed}[style=manuscript]
+\textbf{1.1.3 Discussion and guidelines for future researchers} [EnergyScope Pathway]
+
+Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time-resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix B.1).
+
+However, depending on the research questions to address, other pathway energy system models might be more appropriate given their better spatial resolution or higher techno-economical details, for instance.
+Such models can be used and combined with the other methodological tools presented in this chapter.
+Although, the main (and only) requirement is to keep a low computational time as getting relevant results from the following approaches usually requires $10^3\sim 10^4$ runs.
+In our case, we have implemented the model to keep the time to run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
+\end{mdframed}
+
+\begin{mdframed}[style=manuscript]
+\textbf{1.2.4 Discussion and guidelines for future researchers} [Uncertainty quantification]
+
+Carrying out a Global Sensitivity Analysis (GSA) requires two steps: uncertainty characterisation and uncertainty quantification.
+Characterising the distribution followed by uncertain parameters require a lot of data that are either seldom available or poorly documented in the literature.
+Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distribution.
+Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further researches to refine the characterisation of these parameters.
+
+About the method to quantify the uncertainty, PCE has the advantages to provide Sobol' indices as the impact indicators of the uncertain parameters and extract statistical moments of the output of interest.
+Based on our experience, considering a second-order PCE is the minimum to get a reasonably low LOO error (below 1\%) and reliable results.
+However, as presented in Section 1.2.2, starting with a first-order PCE is a good proxy to get a initial insight into the impact of the parameters and discard the negligible ones before moving to a higher order PCE to increase the accuracy of the results.
+\end{mdframed}
+
+\begin{mdframed}[style=manuscript]
+\textbf{1.3.3 Discussion and guidelines for future researchers} [Agent-based reinforcement learning to support myopic energy transitions]
+
+Reinforcement Learning (RL) was found to be an appropriate approach to explore myopic transition pathways under uncertainties and assess the robustness of policies to support such pathways.
+As a novice user of a RL framework, we would recommend to opt for SAC as it is sample efficient, ensures a wide exploration and has a low sensitivity to hyper-parameters \cite{haarnoja2018soft}.
+Using the SAC package developed by \textsc{Stable-Baselines3} allows a handy introduction to apply RL.
+
+Besides the choice of the algorithm, most of the work when using a RL framework consists in properly define the rules of the games following which the agent interacts with its environment: actions, states and reward.
+These are defined and discussed in Chapter 4.
+\end{mdframed}
+
+\begin{mdframed}[style=manuscript]
+\textbf{1.4.3 Discussion and guidelines for future researchers} [Robustness assessment via PCA]
+
+Principal Component Analysis (PCA) was found to be an appropriate method to provide a metric that capture a major part of the system design variance through the transition while keeping the amount of information to a ``graspable'' level of details.
+Besides selecting the installed capacities of end-use-type categories (electricity, heat, mobility and non-energy), one could discuss the scaling of these data and the management of the outliers.
+Where scaling the data is fundamental before computing the PCs, future researchers could adapt the scaling factor.
+To avoid singularities in our PCs, we have decided \textit{a posteriori} to put high and low outliers to the upper and lower fences, respectively.
+However,before managing the outliers, we would recommend to keep them and assess the provided results in terms of number of PCs and the key technologies that characterise them.
+
+Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{text{transition}}$ (85\%).
+These thresholds have been defined based on good practices discussed in the literature and \textit{a posteriori} observations specific to our case study.
+We would recommend starting with these values and adapting them in a second time if needed to increase or decrease the level of details captured by the PCs.
+\end{mdframed}
+
+Moreover, to discuss the rules of the RL games, presented in Section 4.2, I have added {\color{blue}a Section 4.2.4} introduced {\color{blue}at the end of the introductory paragraph of Section 4.2}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+We end this section with justifications of actions, reward and states implemented in this work as well as guidelines for researchers that would like to apply the RL approach on their own case study with their own model.
+\end{mdframed}
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+\textbf{4.2.4 Discussion and guidelines for future researchers}
+
+As introduced in Chapter 1, most of the work in applying RL is the definition of the interactions between the agent and its environment (\ie actions, reward and states) that are very dependent on the case study and the research questions to answer.
+The elements presented in this work result from several trial and errors to end up with meaningful results according to our research questions.
+
+Besides mimicking potential actual policies, the actions chosen in this work have a direct translation into constraints and we can therefore assess their effectiveness through the fact they are binding or not.
+In this work, we have investigated other actions like incentivising solar PV panels and wind turbines by ``artificially'' reducing their CAPEX.
+The result was not conclusive as these technologies need to take part to the Belgian energy transition and because it was harder to assess the impact of this action.
+
+The reward function was designed to first aiming at respecting the \ce{CO2}-budget then minimising the total transition cost.
+The -300 penalty given in case of an infeasible optimisation problem was arbitrarily set.
+\textit{A posteriori}, it seems to be a well-defined penalty given its significant relative difference with the values taken by the reward otherwise, between -120 and 44 (see Figure 4.4).
+Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
+This requires to define the \ce{CO2}-budget according to a certain sharing principle (see Section 2.5) and to compute reference total transition cost.
+However, other researches focusing on reaching carbon-neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
+
+Finally, states aim at representing the information relevant to the agent to efficiently learn and progress through the transitions.
+For this reason, on top reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assessing their policies to reach their targets (share of renewables in the mix and the system overall efficiency).
+For other studies, one might consider other information like the metrics considered by \citet{pickering2022diversity} (\eg heat electrification, average national import or level of curtailment).
+
+In conclusion, future studies might start from the actions-reward-states defined in this work and adapt these rules depending on the research questions to answer, the case study and the energy system optimisation model.
+\end{mdframed}
+
+Finally, I have added a sentence {\color{blue}in the conclusion of the manuscript} to remind that the methods and the assumptions presented in this work can serve as a good basis but will need to be adapted depending on the model and the case study:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further researches but will need to be adapted according to model that is used or the case study that is studied.
 \end{mdframed}
@@
 Consequently, the share of successful transitions compared to failures progressively increases with time.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained.
-In general, improve the presentation of the results, especially for the RL chapter.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
 \end{mdframed}
@@
 \end{figure}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods?
-We could go further in the analysis as we know we have to phase out coal, install renewables.
-\end{mdframed}
-
-\noindent The outcome is a bit too obvious (e.g.
-getting rid of the coal) $\rightarrow$ go a bit in the details.
-Explain why the method is taking better decisions but explain why.
-$\rightarrow$ go beyond what we know
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
 
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-[...] The myopic optimisation of the pathway consists in a sequence of smaller LP problems limited to their respective 10-year time-window, until eventually reaching 2050.
+[...] The myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time-window, until eventually reaching 2050.
 \end{mdframed}
@@
 \caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
-Myopic optimisation of the pathway consists in a sequence of smaller LP problems limited to their respective 10-year time-window.
-If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters are changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
+Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time-window.
+If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
 \label{fig:MY_process_code}
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent Aggregate similar outputs of interest, without prior knowledge about these ``similarities'' would be challenging and require assumptions to carry out this aggregation, especially given the size of the problem (113 technologies over 7 representative years).
+The objective of PCA is to alleviate this effort and automatise this aggregation in order to capture the main variance of the design through the whole transition in a reduced number of principal components.
+It is reminded in Section 5.1.1: ``Even though the following observations could be made by analysing the distribution of installed capacities (see Appendix D.3) or the covariance matrices, the PCA decomposition offers a more visual and summarising representation of the main trends of variation.''
+
+Given what is already in the text and these explanations, I have not brought modification to the manuscript in this regard.
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent As discussed during the private defense, it is more the spread of variation of the distribution that matters to characterise the robustness rather than the average position.
+This is now clarified {\color{blue}at the end of Section 1.4.2}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+It is really the range over which the distribution spans, $2\cdot \mathrm{MOE}$, that matters to characterise the variability of the projections of a roadmap.
+Indeed, as $\text{PC}_{\text{transition}}$ are vectors composed of positive and negative coefficients, the average position of the distribution, $\mu$, does not bring insight about the robustness of a roadmap.
 \end{mdframed}
@@
 
-\noindent I should insist on those elements in the text {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent As detailed in Section 1.4.2, even though these points result from the optimisation and make sense mathematically, they are defined as outliers as they fall out of the whiskers of their respective box plot.
+The way we managed these outliers is arbitrary but is fair trade-off between leaving them as they are which would mislead the direction of PCs and completely removing them.
+As extensively discussed in Section 5.1.1, this pre-processing step will mostly affect the LT-heat sector where heat pumps running on methane are installed in rare cases, \ie less than 10\% of the 1260 transitions.
+Given the size of this sector, \ie between 22\% and 27\% of the EUD for Belgium, and the LP approach where there is a complete substitution of a technology by another one under certain conditions, these gas-HP can have a major weight in the system.
+However, to keep the main trends of variation without being ``polluted'' by outlying results (here, less than 10\% of the cases), we proceeded that way.
+
+Given what is already in the text and these explanations, I have not brought modification to the manuscript in this regard.
 
@@
 
-\noindent Perhaps reinforce the message as you answered the question $\rightarrow$ We need to go for radical changes, a “unicorn” solution, and one option would be the import of renewable molecules {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+\noindent To succeed its energy transition, Belgium will need radical changes, ``unicorn'' solutions, and one of them would be the import of electrofuels.
+As presented in the introduction and reminded in the conclusion of the manuscript: ``Chemically similar to their fossil-based equivalents, they [the electrofuels] can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.'' Besides these intrinsic advantages, considering these fuels as infinitely available (like their fossil equivalents) and with 0-GWP allow finding a solution for the Belgian case study, without considering a decrease of the demands.
+These are strong assumptions that can be challenged as suggested in the conclusion: ``A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
+This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
+This could also show the limits of the Belgian energy system to meet such an ambitious CO2-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
+''
+
+The assumption about 0-GWP molecules is similar to the one considered \cite{PATHS2050} which is the most recent study that addressed the case of the Belgian energy transition.
+
+As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.
+This need for this radical change and the fact that solutions rely a lot on assumed to be carbon-neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
+Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
+\end{mdframed}
+
+and reminded {\color{blue}in the conclusions of the manuscript}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
+One robust policy would consist in massively investing in electrofuels.
 \end{mdframed}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Explain more in details the different steps in the RL approach
+{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
+\end{mdframed}
+
+\noindent Because the action limits the emissions of the system at the end of the 10-year time-window whereas limiting the consumption of fossil gas, LFO and coal is done over the whole time-window, as detailed in Section 4.1.1.
+Given what is already written in the manuscript, I have not brought modification to the manuscript in this regard.
+
+\subsection{Graphs and results}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods like RL?
+We could go further in the analysis as we know we have to phase out coal, install renewables.
+\end{mdframed}
+
+\noindent The outcome is a bit too obvious (e.g.
+getting rid of the coal) $\rightarrow$ go a bit in the details.
+Explain why the method is taking better decisions but explain why.
+$\rightarrow$ go beyond what we know
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained.
+In general, improve the presentation of the results, especially for the RL chapter.
 \end{mdframed}
@@
 
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 84: Why -300? Seems like fine-tuned to obtain desired results?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+89: Figure 4.3 seems to deliver an important message (early action is needed to ensure we meet climate goals), but it is not clearly explained.
+Improve captions / add axes.
+	Where do I see the ``tipping points'' in this figure?
+\end{mdframed}
+
+\noindent Like other figures in this Chapter, Figure 4.3 represents distributions of features spreading over the x-axis.
+In this case, these are distributions of the rewards collected by the agent at the end of each episode of its learning process.
+For this reason, labelling the axis does not seem obligatory as the x-axis is nothing else than the feature (here, the reward) and the y-axis has no physical meaning expect representing how often the feature of interest took a certain value/range of values.
+Compared to the other similar figures of the chapter, I have not labelled the x-axis as, for Figure 4.3, the reward is the only feature represented whereas two or more features are gathered on the same figure.
+
+With this in mind and what is already written in the caption of the figure and the core of the text, we can infer from the right hand side that 2040 is the ``tipping year'' as it represents the year at which the failures occur more often, compared to the other representative years.
+
+Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
+\end{mdframed}
+
+\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2}-budget) by 2050 at the latest.
+The number of remaining attempts reduce as prematurely ended episode are removed from the further counts.
+The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves.
+These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
+\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+94: Figure 4.6 seems to be less revealing than other figures.
+Also the figure caption seems to contradict the first statement following the figure.
+It would be important to clarify: which are the actions that emerge to actually matter?
+\end{mdframed}
+
+\noindent Indeed, Figure 4.6 is less revealing as it shows no big trend in terms of neither impacting actions nor their timing in the transition.
+It is detailed in the first paragraph of Section 4.2.3 after which the figure should appear in the manuscript.
+I have worked to make Figures as close as possible to their related paragraph.
+However, in some cases, several other paragraphs come between them.
+To identify the actions that emerge to actually matter, we need to check if they are binding or not, as detailed in the subsequent paragraphs and in Figure 4.7.
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard, except minor adjustment to get Figure 4.6 as close as possible to its related paragraph.
+
+\subsection{Others}
 
@@
 
-\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/shocks can be implemented through uncertain parameters.
+\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/ shocks can be implemented through uncertain parameters.
 For instance, this is how we have implemented the potential to install SMR in 2040 at the soonest.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-89: Figure 4.3 seems to deliver an important message (early action is needed to ensure we meet climate goals), but it is not clearly explained.
-Improve captions / add axes.
-	Where do I see the ``tipping points'' in this figure?
-\end{mdframed}
-
-\noindent Like other figures in this Chapter, Figure 4.3 represents distributions of features spreading over the x-axis.
-In this case, these are distributions of the rewards collected by the agent at the end of each episode of its learning process.
-For this reason, labelling the axis does not seem obligatory as the x-axis is nothing else than the feature (here, the reward) and the y-axis has no physical meaning expect representing how often the feature of interest took a certain value/range of values.
-Compared to the other similar figures of the chapter, I have not labelled the x-axis as, for Figure 4.3, the reward is the only feature represented whereas two or more features are gathered on the same figure.
-
-With this in mind and what is already written in the caption of the figure and the core of the text, we can infer from the right hand side that 2040 is the ``tipping year'' as it represents the year at which the failures occur more often, compared to the other representative years.
-
-Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{orange} \textbf{Stefano}} - Overall, I suggest adding a diagram to Chapter 4, detailing what are the actions available to the decision-maker.
@@
 \caption{Actions available to the decision-maker.
-Taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.}
+Taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
+Unlike the first action that sets a target for the end of the time-window, the last three aim at limiting the consumption of these fossil resources over the whole time-window.}
 \label{fig:Schematic_actions}
@@
 Consequently, regarding this comment, there has not been further modification brought to the manuscript.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
-\end{mdframed}
-
-\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2}-budget) by 2050 at the latest.
-The number of remaining attempts reduce as prematurely ended episode are removed from the further counts.
-The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves.
-These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
-\end{mdframed}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-94: Figure 4.6 seems to be less revealing than other figures.
-Also the figure caption seems to contradict the first statement following the figure.
-It would be important to clarify: which are the actions that emerge to actually matter?
-\end{mdframed}
-
-\noindent Indeed, Figure 4.6 is less revealing as it shows no big trend in terms of neither impacting actions nor their timing in the transition.
-It is detailed in the first paragraph of Section 4.2.3 after which the figure should appear in the manuscript.
-I have worked to make Figures as close as possible to their related paragraph.
-However, in some cases, several other paragraphs come between them.
-To identify the actions that emerge to actually matter, we need to check if they are binding or not, as detailed in the subsequent paragraphs and in Figure 4.7.
-
-Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard, except minor adjustment to get Figure 4.6 as close as possible to its related paragraph.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?
-\end{mdframed}
-
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+{\color{orange} \textbf{Stefano}} - Explain more in details the different steps in the RL approach.
+\end{mdframed}
+
+\noindent Besides the general concepts/fundamentals detailed in Section 1.3.1, the application of the RL approach on the myopic optimisation of the energy transition of a whole-energy system is detailed in Section 1.3.2.
+Now, Figure 1.7 has been updated to clarify the difference between the uncertainty of parameters and the agent's actions as well as the interactions between the RL code (action, state, reward) and EnergyScope Pathway (optimisation of the concerned 10-year time-window).
+I have also added a diagram to explain the impact that the agent's actions have on some parameters of the model.
+
+Given what was already in the manuscript and the aforementioned modifications, I have not added further information to detail the different steps of the RL approach.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
-\end{mdframed}
-
-\noindent To explain in the definition of the action {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - Justification of the actions selected? why no other actions like the cost of some technologies? 
-\end{mdframed}
-
-\noindent These actions have a direct translation in constraints and we can therefore assess their effectiveness through the fact they are binding or not.
-Give examples of actions tried during the work (incentivize renewables) or other from Assemblée citoyenne in France and explain why they’re not as good as those chosen in the thesis.
-Explain history of what we tried/consider?
- {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - p.
@@
 
-\noindent {\color{blue} }.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
+\noindent This was observed on the annual system cost which is not the objective function of the Pathway model.
+It is more relevant to look at the total transition cost which is, logically, lower for the PF approach compared to myopic when facing the same sample of uncertain parameters.
+To clarify this aspect, I have removed the paragraph (and the graph) about the comparison of the annual system cost and added instead a paragraph (and a graph) about the comparison of the total transition cost {\color{blue}in Section 4.3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Looking at the total transition cost, the combination of the agent's actions and favourable economic conditions (see Section 4.2.2) make the myopic transitions cheaper, on average, than the PF cases (see right side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
+This is also due to the fact that the perfect foresight approach always finds a solution even in worst conditions such as high cost of purchasing resources and high EUD.
+This explains the wider variability of the results too.
+However, with the same sample of uncertain parameters, given the assumed full knowledge over the whole time horizon, PF naturally results in a cheaper transition than its myopic equivalent.
+\end{mdframed}
+
+\begin{figure}[!htbp]
+\centering
+\includegraphics[height=4cm]{Gwp_pathway_core.pdf}
+\includegraphics[height=4cm]{Transition_cost_comp_2.pdf}
+\caption{Comparison of \ce{CO2}-emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the RL-based myopic optimisation.
+Myopic transitions succeed with a more drastic reduction of emissions in the short-term and, on average, more favourable economic conditions.}
+\label{fig:Gwp_pathway_total_tran_cost}
+\end{figure}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 101: you mention different definitions of “robustness”: which one do you use in your work?
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 109: how is Gamma-robustness implemented here? It seems like it is in some sense “manually” forced? Explain better how the ROB solution is derived.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-
+{\color{orange} \textbf{Stefano}} - p. 101: you mention different definitions of ``robustness'': which one do you use in your work?
+\end{mdframed}
+
+\noindent This work assesses the robustness of the decision-making process in two different ways.
+On the one hand, using a Reinforcement Learning (RL) approach, it allows highlighting the key actions to take to support a myopic transition and maximise the chances to respect the emissions-budget.
+On the other hand, after drawing a robustness-metric based on a Principal Component Analysis (PCA) approach, this work investigates how much a planned roadmap is sensitive to additional investments required in a myopic transition with progressively unveiled uncertainties.
+In other words, we use two different definitions of robustness: (i) the ability of a policy to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget, (ii) the ability of a technological roadmap to limit the investment into additional capacities when tested in myopic conditions.
+
+These two definitions are now clearly introduced {\color{blue}in the introduction of the manuscript}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Chapter 4 focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget.
+
+Chapter 5 assesses the robustness of different technological roadmaps as their ability to limit the investment into additional capacities when tested in myopic conditions. 
+\end{mdframed}
+
+{\color{blue}at the end of the introductory paragraph of Chapter 1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Finally, an agent-based Reinforcement Learning (RL) approach is detailed to address the robustness of a policy in uncertain transitions with limited vision in the future.
+This chapter ends with a PCA-based approach to draw a robustness metric and assess the ability of technological roadmaps to limit the investment into additional capacities.
+\end{mdframed}
+
+{\color{blue}in the second to last paragraph of the contributions of Chapter 1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+The third principal methodological contribution is the use of Principal Component Analysis (PCA) to assess the robustness of a technological roadmap.
+Given the uncertainties and the timespan of the transition, this approach allows highlighting the main ``directions'' of variation of the system design (\ie the installed capacities).
+After this step of identification, strategies can be projected on these directions to see how robust they are to the need of additional capacities.
+\end{mdframed}
+
+In the second paragraph of Chapter 5, we list different definitions of robustness.
+In this work, we consider indeed ``static'' robustness as we aim at either getting a policy that maximises the chance to succeed the transition or a technological roadmap that limits the need to invest into additional capacities.
+It is now clarified {\color{blue}in the second paragraph of Chapter 5}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Where Chapter 4 addressed the static robustness of a policy as its ability to maximise the chances to succeed a myopic transition, the objective of this chapter is to apply the method described in Section 1.4 to deal with the static robustness of pathway technological roadmaps, defined as their ability to limit the need to invest into additional capacities.
+\end{mdframed}
+
+Finally, I have gone through the manuscript and made sure that RL-based approach was addressing the robustness of \textbf{policies} (\ie sequences of actions) whereas the PCE-based approach rather focuses on the robustness of \textbf{technological roadmaps}.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 109: how is Gamma-robustness implemented here? It seems like it is in some sense ``manually'' forced? Explain better how the ROB solution is derived.
+\end{mdframed}
+
+\noindent To my understanding, the $\Gamma$-robustness consists in considering a number, $\Gamma$, of uncertain parameters taking their worst-case value simultaneously and ``evaluate how the solution of the optimisation problem changes when varying $\Gamma$'' \cite{moret2020overcapacity}.
+Inspired by this approach, we have arbitrarily set this value to $\Gamma=6$ to consider the six first uncertain parameters impacting the most the variation of the total transition cost.
+Consequently, the ROB technological roadmap is obtained from the deterministic optimisation of the perfect foresight pathway accounting for these six parameters (\ie costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry EUD, the interest rate and the variable OPEX of technologies) at their respective worst-case value.
+As it does not exactly follow the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}, I have changed {\color{blue}the first sentence of Section 5.2.1:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow a \textbf{method inspired by} the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}.
+\end{mdframed}
+
+Moreover, the facts that this $\Gamma=6$ is arbitrarily set and that the ROB roadmap, like the REF and SMR ones, results from a deterministic optimisation of the perfect foresight pathway are now clarified:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+In the present analysis, we label the roadmap as ``robust'' (ROB) by arbitrarily setting $\Gamma_{\text{obj}}= 6$.
+The ROB roadmap results from the deterministic optimisation of the perfect foresight pathway where the following six parameters are set at their worst-case value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry EUD, the discount rate and the variable OPEX of technologies.
 \end{mdframed}

### 2024-07-16 chap_RL/chap_RL.tex
Getting there...

@@
 Then, the window 2020-2030 is optimised via EnergyScope.
-Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state he took it.
+Some of the outputs of this optimisation feed the agent with either the new state of the system or the reward, \ie telling the agent how good the actions were at the state the agent took it.
 Based on the new state and the reward, the agent takes another set of actions and the window 2025-2035 is optimised.
@@
 For the next episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), another sample of uncertain parameters is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+
+We end this section with justifications of actions, reward and states implemented in this work as well as guidelines for researchers that would like to apply the RL approach on their own case study with their own model.
 
@@
 
+\subsection{Discussion and guidelines for future researchers}
+\label{subsec:RL:act_states_rew:discussion}
+As introduced in Chapter \ref{chap:chap_methodo}, most of the work in applying \gls{RL} is the definition of the interactions between the agent and its environment (\ie actions, reward and states) that are very dependent on the case study and the research questions to answer.
+The elements presented in this work result from several trial and errors to end up with meaningful results according to our research questions.
+
+Besides mimicking potential actual policies, the actions chosen in this work have a direct translation into constraints and we can therefore assess their effectiveness through the fact they are binding or not.
+In this work, we have investigated other actions like incentivising solar \gls{PV} panels and wind turbines by ``artificially'' reducing their \gls{CAPEX}.
+The result was not conclusive as these technologies need to take part to the Belgian energy transition and because it was harder to assess the impact of this action.
+
+The reward function was designed to first aiming at respecting the \ce{CO2}-budget then minimising the total transition cost.
+The -300 penalty given in case of an infeasible optimisation problem was arbitrarily set.
+\textit{A posteriori}, it seems to be a well-defined penalty given its significant relative difference with the values taken by the reward otherwise, between -120 and 44 (see Figure \ref{fig:reward_status}).
+Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
+This requires to define the \ce{CO2}-budget according to a certain sharing principle (see Section \ref{sec:cs:CO2-budget}) and to compute reference total transition cost.
+However, other researches focusing on reaching carbon-neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
+
+Finally, states aim at representing the information relevant to the agent to efficiently learn and progress through the transitions.
+For this reason, on top reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assessing their policies to reach their targets (share of renewables in the mix and the system overall efficiency).
+For other studies, one might consider other information like the metrics considered by \citet{pickering2022diversity} (\eg heat electrification, average national import or level of curtailment).
+
+In conclusion, future studies might start from the actions-reward-states defined in this work and adapt these rules depending on the research questions to answer, the case study and the energy system optimisation model.
+
 \newpage
@@
 
-\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2}-budget (see Figure \ref{fig:Gwp_pathway}).
+\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2}-budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
 However, driven first by this \ce{CO2}-budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}).
@@
 
+Looking at the total transition cost, the combination of the agent's actions and favourable economic conditions (see Section \ref{subsec:RL:learning:states}) make the myopic transitions cheaper, on average, than the PF cases (see right side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
+This is also due to the fact that the perfect foresight approach always finds a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}.
+This explains the wider variability of the results too.
+However, with the same sample of uncertain parameters, given the assumed full knowledge over the whole time horizon, PF naturally results in a cheaper transition than its myopic equivalent.
+
 \begin{figure}[!htbp]
 \centering
-\includegraphics[width=0.8\textwidth]{Gwp_pathway_core.pdf}
-\caption{Comparison of \ce{CO2}-emissions pathways from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation. }
-\label{fig:Gwp_pathway}
+\includegraphics[height=3.5cm]{Gwp_pathway_core.pdf}
+\includegraphics[height=3.5cm]{Transition_cost_comp_2.pdf}
+\caption{Comparison of \ce{CO2}-emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.
+Myopic transitions succeed with a more drastic reduction of emissions in the short-term and, on average, more favourable economic conditions.}
+\label{fig:Gwp_pathway_total_tran_cost}
 \end{figure}
 
-The transition pathways of the annual system cost show another impact of the agent in the \gls{RL}-based exploration (see Figure \ref{fig:System_cost_pathway}).
+%The transition pathways of the annual system cost show another impact of the agent in the \gls{RL}-based exploration (see Figure \ref{fig:System_cost_pathway}).
 Via its actions and its cost-based reward, the agent can reach systems that are cheaper than any other solution obtained by the perfect foresight approach.
@@
 
-\begin{figure}[!htbp]
-\centering
-\includegraphics[width=0.8\textwidth]{System_cost_pathway_core.pdf}
-\caption{Comparison of annual system costs pathways from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
-\label{fig:System_cost_pathway}
-\end{figure}
+%\begin{figure}[!htbp]
+%\centering
+%\includegraphics[width=0.8\textwidth]{System_cost_pathway_core.pdf}
+%\caption{Comparison of annual system costs pathways from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
+%\label{fig:System_cost_pathway}
+%\end{figure}
 

### 2024-07-16 chap_RobPol/chap_RobPol.tex
Go, go, go...
Getting there...

@@
 In static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions.
-The adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}.
-The objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.
+Where Chapter \ref{chap:chap_RL} addressed the static robustness of a policy as its ability to maximise the chances to succeed a myopic transition, the objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway technological roadmaps, defined as their ability to limit the need to invest into additional capacities.
+
 \citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}.
@@
 \label{subsec:RobPol:Rob_roadmap}
-Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow here the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}.
+Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow a method inspired by the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}.
 In their works, the authors consider a factor $\Gamma_{\text{obj}}\in [0,d]$ that represents a ``protection parameter'' of the objective function where $d$ is the total number of uncertain parameters (see Section \ref{subsec:pce}).
@@
 $\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value, and so on.
-In the present analysis, we label the roadmap as ``robust'' (ROB) by setting $\Gamma_{\text{obj}}= 6$.
-Consequently, the following six parameters are set at their worst value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the discount rate and the variable OPEX of technologies.
+In the present analysis, we label the roadmap as ``robust'' (ROB) by arbitrarily setting $\Gamma_{\text{obj}}= 6$.
+The ROB roadmap results from the deterministic optimisation of the perfect foresight pathway where the following six parameters are set at their worst-case value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the discount rate and the variable OPEX of technologies.
 In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the discount rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.
@@
 
-In the ROB roadmap, the transport of freight by trucks first transitions from diese to electric trucks and then, after 2035, to \gls{FC} trucks.
+In the ROB roadmap, the transport of freight by trucks first transitions from diesel to electric trucks and then, after 2035, to \gls{FC} trucks.
 Used as a way to support the the early installation of solar \gls{PV}, electric trucks are then replaced by \gls{FC} ones given the limited availability of renewable electricity and its more cost-effective use in other sectors such as private mobility.

### 2024-07-16 chap_atom_mol/chap_atom_mol.tex
Getting there...

@@
 
-Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the biggest impact of the cost of purchasing electrofuels on the variation of the total transition cost, around 45\%.
+Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
+Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
 On the contrary, parameters directly related to \gls{SMR} (\ie its availability and its CAPEX) have limited impact, below 1\%.

### 2024-07-16 chap_intro_ccl/conclusion.tex
Getting there...

@@
 
-Besides a wider electrification of the system and the integration of more efficient technologies, a robust policy consists in investing in electrofuels.
+Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
+One robust policy would consist in massively investing in electrofuels.
 Chemically similar to their fossil-based equivalents, they can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.
@@
 Regarding methodologies, this thesis developed novel approaches to explore myopic pathway transitions, highlight intermediate milestones and no-go zones towards sustainability and assess the robustness of transition technological roadmaps.
-Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them on their specific case study.
-\\
+Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them on their specific case study and energy system optimisation model.
+Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further researches but will need to be adapted according to model that is used or the case study that is studied.\\
 

### 2024-07-16 chap_intro_ccl/intro.tex
Getting there...

@@
 \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
-However, their work focused on the power sector only and for a target future year (2035).
+However, their work focused on the power sector only and for the target future year of 2035.
 Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost and provide more details on the sensitivity of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
@@
 
-To account for uncertainties, uncertainty characterisation and quantification have been applied by adapting the works of \citet{Moret2017PhDThesis} and \citet{coppittersthesis}, respecitvely.
+To account for uncertainties, uncertainty characterisation and quantification have been applied by adapting the works of \citet{Moret2017PhDThesis} and \citet{coppittersthesis}, respectively.
 In his thesis, \citet{Moret2017PhDThesis} developed a framework to obtain uncertainty ranges for a variety of parameters such as cost of purchasing and availability of resources or investment cost and efficiency of technologies.
@@
 In other words, we define the action and state spaces as well as the reward function driving the behaviour of the agent in its quest to optimise its policy.
-Then, we analyse the results of the learning phase before testing --- under uncertainties --- the learned policy versus more ``classic'' myopic optimisation, \ie without the support of such a policy.
+Then, we analyse the results of the agent's learning process and compare these results with the perfect foresight approach.
+Chapter \ref{chap:chap_RL} focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget.
 
-Chapter \ref{chap:chap_RobPol} assesses the robustness of different roadmaps resulting from deterministic perfect foresight optimisation under certain conditions: REF, SMR and ROB.
+Chapter \ref{chap:chap_RobPol} assesses the robustness of different technological roadmaps as their ability to limit the investment into additional capacities when tested in myopic conditions.
+Three roadmaps resulting from different deterministic perfect foresight optimisation, are presented: REF, SMR and ROB.
 The first one, the reference case, considers nominal values for all the uncertain parameters.
 The SMR case is the one introduced in Chapter \ref{chap:atom_mol} where we allow the model to install \gls{SMR} from 2040 onward.
-Eventually, the ROB case accounts for the highest values of parameters having the biggest impact on the total cost of transition (\ie cost of purchasing energy carriers, industrial \gls{EUD}, discount rate and variable \gls{OPEX} of technologies).
+Eventually, the ROB case accounts for the highest values of uncertain parameters having the biggest impact on the total cost of transition (\ie cost of purchasing fossil fuels and renewable electrofuels, industrial \gls{EUD}, discount rate and variable \gls{OPEX} of technologies).
 

### 2024-07-16 chap_methodo/chap_methodo.tex
Getting there...

@@
 Then, it focuses on the uncertainties, their characterisation as well as their quantification.
-Finally, an agent-based \acrfull{RL} approach is detailed to address the sequential decision-making process in the uncertain transition with limited vision in the future.
-The robustness of these policies is assessed via the use of the \gls{PCA}.
+Finally, an agent-based \acrfull{RL} approach is detailed to address the robustness of a policy in the uncertain transition with limited vision in the future.
+This chapter ends with a \gls{PCA}-based approach to draw a robustness metric and assess the ability of technological roadmaps to limit the investment into additional capacities.
+Each section of this chapter ends with some discussion and guidelines for researchers to further use and develop these methodologies.
 
@@
 
-The third principal methodological contribution is the use of \acrfull{PCA} to assess the robustness of a policy, \ie how much the optimal pathway is affected by the uncertainties.
-Given the uncertainties and the timespan of the transition, this approach allows highlighting the main \og directions'' of variation of the system design (\ie the installed capacities).
-After this step of identification, strategies and policies can be projected on these directions to see how robust they are to the overall transition uncertainties.
+The third principal methodological contribution is the use of \acrfull{PCA} to assess the robustness of a technological roadmap.
+Given the uncertainties and the timespan of the transition, this approach allows highlighting the main ``directions'' of variation of the system design (\ie the installed capacities).
+After this step of identification, strategies can be projected on these directions to see how robust they are to the need of additional capacities.
 
@@
 \noindent
-The major add-on from the original EnergyScope Pathway model \cite{limpens2021generating} to the myopic version developed in this thesis, is the possibility to carry out the optimisation on a limited time window, of which the duration is defined by $N\textsubscript{year,opti}$.
-Moreover, there is also the possibility of having an overlap between two consecutive time windows.
+The major add-on from the original EnergyScope Pathway model \cite{limpens2021generating} to the myopic version developed in this thesis is the possibility to carry out the optimisation on a limited time window, of which the duration is defined by $N\textsubscript{year,opti}$.
+Moreover, there is the possibility of having an overlap between two consecutive time windows.
 The timespan of this overlap is defined by the parameter $N\textsubscript{year,overlap}$.
@@
 This way, the newly developed model can easily be used to perform a perfect foresight optimisation by setting the time window to $N\textsubscript{year,opti}=$30 years (\ie between 2020 and 2050) and the overlap between the time windows to $N\textsubscript{year,overlap}=$0.
-Consequently, as it is fundamental to define, on the one hand, the actual time window on which the system is optimised, and on the other hand, the history, \ie what has already been optimised earlier in the transition, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
+Consequently, as it is fundamental to define on the one hand, the actual time window on which the system is optimised, and on the other hand, the history, \ie what has already been optimised earlier in the transition, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
 
@@
 $\text{YEARS}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{WND}$ substitute $\text{YEARS}$ and $\text{PHASE}$ in the constraints defined in the pathway model in Section \ref{subsec:meth:PF}.
-These two sets aim at setting the optimization to a more limited time window.
-Progressing through the transition, $\text{YEARS}\textsubscript{UP TO}$ and $\text{PHASE}\textsubscript{UP TO}$ allow keeping track of the history of the investments (\eg technologies installation, decommissioning or retirement), the consumption of resources, the cumulative amount of emissions, etc.
+These two sets aim to set the optimization to a more limited time window.
+Progressing through the transition, $\text{YEARS}\textsubscript{UP TO}$ and $\text{PHASE}\textsubscript{UP TO}$ allow keeping track of the history of the investments (\eg technology installations, decommissioning or retirement), the consumption of resources, the cumulative amount of emissions, etc.
 
@@
 For instance, the first year of a time window is the result of the optimisation of the previous one.
-Therefore, optimising again this first year could lead to rounding errors preventing from the optimization to converge.
-For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimise that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
-This remark stays valid for any time window except the first one of the transition where the year 2020 is optimised even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
+Therefore, optimising again this first year could lead to rounding errors preventing the optimisation to converge.
+For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimise, that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
+This remark remains valid for any time window except the first one of the transition where the year 2020 is optimised even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
 Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Equation (\ref{eq:salvage}).\\
@@
 Using this \gls{API}, this updated version of the model interacts with the AMPL problem representing the optimisation of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
-The myopic optimisation of the pathway consists in a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window, until eventually reaching 2050.\\
+The myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window, until eventually reaching 2050.\\
 
@@
 \caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
-Myopic optimisation of the pathway consists in a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window.
-If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters are changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
+Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window.
+If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
 \label{fig:MY_process_code}
@@
 
-\myparagraph{Impact of myopic formulation on the system}\\
+\myparagraph{Impact of the myopic formulation on the system}\\
 
@@
 A more detailed comparison between the myopic and perfect foresight approaches is available in Appendix \ref{app:my_vs_pf}.
+
+\subsection{Discussion and guidelines for future researchers}
+\label{subsec:meth:ES:discussion}
+Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time-resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix \ref{app:ESPathway_choice}).
+
+However, depending on the research questions to address, other pathway energy system models might be more appropriate given their better spatial resolution or higher techno-economical details, for instance.
+Such models can be used and combined with the other methodological tools presented in this chapter.
+Although, the main (and only) requirement is to keep a low computational time as getting relevant results from the following approaches usually requires $10^3\sim 10^4$ runs.
+In our case, we have implemented the model to keep the time to run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
 
@@
 
+\subsection{Discussion and guidelines for future researchers}
+\label{subsec:meth:UQ:discussion}
+Carrying out a \acrfull{GSA} requires two steps: uncertainty characterisation and uncertainty quantification.
+Characterising the distribution followed by uncertain parameters require a lot of data that are either seldom available or poorly documented in the literature.
+Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distribution.
+Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further researches to refine the characterisation of these parameters.
+
+About the method to quantify the uncertainty, \gls{PCE} has the advantages to provide Sobol' indices as the impact indicators of the uncertain parameters and extract statistical moments of the output of interest.
+Based on our experience, considering a second-order \gls{PCE} is the minimum to get a reasonably low \gls{LOO} error (below 1\%) and reliable results.
+However, as presented in Section \ref{subsec:pce}, starting with a first-order \gls{PCE} is a good proxy to get a initial insight into the impact of the parameters and discard the negligible ones before moving to a higher order \gls{PCE} to increase the accuracy of the results.
+
 \section{Agent-based reinforcement learning to support myopic energy transition}
@@
 
+\subsection{Discussion and guidelines for future researchers}
+\label{subsec:meth:RL:discussion}
+\acrfull{RL} was found to be an appropriate approach to explore myopic transition pathways under uncertainties and assess the robustness of policies to support such pathways.
+As a novice user of a \gls{RL} framework considering a continuous action and state spaces, we would recommend to opt for \gls{SAC} as it is sample efficient, ensures a wide exploration and has a low sensitivity to hyper-parameters \cite{haarnoja2018soft}.
+Using the \gls{SAC} package developed by \textsc{Stable-Baselines3} allows a handy introduction to apply \gls{RL}.
+
+Besides the choice of the algorithm, most of the work when using a \gls{RL} framework consists in properly define the rules of the games following which the agent interacts with its environment: actions, states and reward.
+These are defined and discussed in Chapter \ref{chap:chap_RL}.
+
 \section{Robustness assessment via PCA}
@@
 This methodology provides two main outputs.
-First and foremost, using the model runs necessary to quantify the impact of the uncertain parameters on the total cost of the transition (see Section \ref{sec:meth:UQ}), it gives a metric on which to assess the robustness of different energy transition policies.
+First and foremost, using the model runs necessary to quantify the impact of the uncertain parameters on the total cost of the transition (see Section \ref{sec:meth:UQ}), it gives a metric on which to assess the robustness of different energy transition technological roadmaps.
 Where the total transition cost is a single value representing the entire system and analysing individually all technologies would be overwhelming, this metric gives the happy medium.
 Second, these ``directions of variation'' can highlight key modal shifts or highly varying design strategies over the transition.
-After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of policies.
+After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of roadmaps.
 The reader interested by the results of the \gls{PCA}-based method is invited to go to Chapter \ref{chap:chap_RobPol}.
@@
 \label{subsec:meth:PCA:PCA}
-Born in the early 20th century \cite{pearson1901on,hotelling1933analysis}, the \acrfull{PCA} finds its fundamentals from the \gls{SVD}.
+Born in the early 20th century, the \acrfull{PCA} finds its fundamentals from the \gls{SVD} \cite{pearson1901on,hotelling1933analysis}.
 \gls{SVD} is a generalisation, to an arbitrary (\ie not necessary square) matrix, of the spectral theorem stating that a normal matrix can be diagonalised by an orthonormal basis of eigenvectors.
@@
 Finally, this ensures a fair comparison between the original features.
-Without normalization, variables with larger scales would dominate the principal components, potentially skewing the results and leading to misinterpretation of the principal components.
-In other words, given this normalization, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
+Without normalisation, variables with larger scales would dominate the principal components, potentially skewing the results and leading to misinterpretation of the principal components.
+In other words, given this normalisation, $\mathrm{var}\left(z_k\right)=\lambda_k$, where $\mathrm{var}\left(z_k\right)$ is the variance of $z_k$.
 Moreover, this means that the coefficient $\alpha_{ki}$, \ie the component of $ \mathbf{\alpha_k}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$.
@@
 Like any other dimension-reduction process, \gls{PCA} has to be supplied with relevant data to reach the stated objective.
-Like the normalization of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude.
+Like the normalisation of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude.
 Then, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
@@
 Preprocessing the dataset before employing a method to reduce dimensionality, like \gls{PCA}, can greatly affect the structure of the simplified representation and the characteristics of the features extracted from the dataset \cite{parente2013principal,peerenboom2015dimension}.
-Scaling the original raw data via normalization, \ie reducing data to $[0, 1]$ interval has a double purposes: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
+Scaling the original raw data via normalisation, \ie reducing data to $[0, 1]$ interval has a double purposes: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
 Consequently, the first part of this data preprocessing consists in scaling the installed capacities versus their respective sector and representative year (see Equation (\ref{eq:PCA_scaling_0})).
@@
 One can notice that this second scaling factor omits the infrastructure and storage technologies.
-In the process to define ``metrics'' to assess the robustness of a policy for the case of Belgium, this has a negligible impact.
+In the process to define ``metrics'' to assess the robustness of a roadmap for the case of Belgium, this has a negligible impact.
 Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies.
@@
 These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}).
-Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for this specific technology and year.
+Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for these specific technology and year.
 There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
@@
 where $\mathrm{PPF}_{\mathrm{t}}$ is the percent point function (\ie inverse of cumulative distribution function) of the Student's distribution.
+It is really the range over which the distribution spans, $2\cdot \mathrm{MOE}$, that matters to characterise the variability of the projections of a roadmap.
+Indeed, as $\text{PC}_{\text{transition}}$ are vectors composed of positive and negative coefficients, the average position of the distribution, $\mu$, does not bring insight about the robustness of a roadmap.
 
@@
 \end{figure}
+
+\subsection{Discussion and guidelines for future researchers}
+\label{subsec:meth:PCA:discussion}
+\acrfull{PCA} was found to be an appropriate method to provide a metric that capture a major part of the system design variance through the transition while keeping the amount of information to a ``graspable'' level of details.
+Besides selecting the installed capacities of end-use-type categories (electricity, heat, mobility and non-energy), one could discuss the scaling of these data and the management of the outliers.
+Where scaling the data is fundamental before computing the PCs, future researchers could adapt the scaling factor.
+To avoid singularities in our PCs, we have decided \textit{a posteriori} to put high and low outliers to the upper and lower fences, respectively.
+However,before managing the outliers, we would recommend to keep them and assess the provided results in terms of number of PCs and the key technologies that characterise them.
+
+Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{text{transition}}$ (85\%).
+These thresholds have been defined based on good practices discussed in the literature and \textit{a posteriori} observations specific to our case study.
+We would recommend starting with these values and adapting them in a second time if needed to increase or decrease the level of details captured by the PCs.

### 2024-07-19 rebuttals/rebuttals.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 \clearpage
-%\tableofcontents
-%
-%\clearemptydoublepage
+\tableofcontents
+
+\clearemptydoublepage
 
@@
 
-{\color{teal} \textbf{Stefan}} - Justification of the RL actions selected? why no other actions like the cost of some technologies?
-
-{\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?
-
-{\color{purple} \textbf{Sylvain}} - Make sure to discuss the hypotheses because one might not agree with all of them.
+\noindent {\color{teal} \textbf{Stefan}} - Justification of the RL actions selected? why no other actions like the cost of some technologies?
+
+\noindent {\color{purple} \textbf{Sylvain}} - p. 86 - A lot of heuristic and trial and error. Can the methodology be generalized to other models?
+
+\noindent {\color{purple} \textbf{Sylvain}} - Make sure to discuss the hypotheses because one might not agree with all of them.
 So, it could provide more context but also some limitations and indications of the alternatives (and the spectrum of what is possible).
@@
 
-Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time-resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix B.1).
+Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix B.1).
 
@@
 
-The reward function was designed to first aiming at respecting the \ce{CO2}-budget then minimising the total transition cost.
+The reward function was designed to first aiming at respecting the \ce{CO2} budget then minimising the total transition cost.
 The -300 penalty given in case of an infeasible optimisation problem was arbitrarily set.
@@
 Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
-This requires to define the \ce{CO2}-budget according to a certain sharing principle (see Section 2.5) and to compute reference total transition cost.
-However, other researches focusing on reaching carbon-neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
+This requires to define the \ce{CO2} budget according to a certain sharing principle (see Section 2.5) and to compute reference total transition cost.
+However, other researches focusing on reaching carbon neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
 
@@
 \label{general_graphs_results}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Work on the graph of slide 19 (there is a misunderstanding on the fact that the grey curve is labelled as failure but they are failures in the future)
-\end{mdframed}
-
-\noindent To clarify this point, I have extended the caption of the related figures (see previous comments) and added two sentences in {\color{blue} Section 4.2.2}:
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-Since infeasible cases or those that overshoot the \ce{CO2}-budget are discarded before reaching 2050 (see Section 4.1.2), the number of attempts that reach further steps in the transition progressively decreases.
-Consequently, the share of successful transitions compared to failures progressively increases with time.
-\end{mdframed}
 
@@
 Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
-The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
+The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case.
 \end{mdframed}
@@
 \end{figure}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - Lot of the results hinge on the import of e-fuels with the assumption that they have GWP = 0; is it justified? Those fuels are directly available knowing they are no.
+\end{mdframed}
+
+\noindent To succeed its energy transition, Belgium will need radical changes, ``unicorn'' solutions, and one of them would be the import of electrofuels.
+As presented in the introduction and reminded in the conclusion of the manuscript: ``Chemically similar to their fossil-based equivalents, they [the electrofuels] can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.'' Besides these intrinsic advantages, considering these fuels as infinitely available (like their fossil equivalents) and with 0-GWP allow finding a solution for the Belgian case study, without considering a decrease of the demands.
+These are strong assumptions that can be challenged as suggested in the conclusion: ``A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
+This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
+This could also show the limits of the Belgian energy system to meet such an ambitious CO2-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
+''
+
+The assumption about 0-GWP molecules is similar to the one considered \cite{PATHS2050} which is the most recent study that addressed the case of the Belgian energy transition.
+
+As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.
+This need for this radical change and the fact that solutions rely a lot on assumed to be carbon neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
+Assumed to be carbon neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
+\end{mdframed}
+
+and reminded {\color{blue}in the conclusions of the manuscript}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
+One robust policy would consist in massively investing in electrofuels.
+\end{mdframed}
 
@@
 
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\noindent The listed colloquial expressions have been transformed and, overall, I have reviewed the manuscript to reduce them.
+For instance, {\color{blue} in the contributions of Chapter 2}, I have changed ``lion's share'' into ``bigger share''.
 
@@
 
-Since the implementation of the salvage value and the formulation choices is the work developed by \citet{limpens2024pathway} (partly integrated in Appendix B) and given the above explanations, I have not brought further modification to the manuscript in this regard.
+Since the implementation of the salvage value and the formulation choices are the work developed by \citet{limpens2024pathway} (partly integrated in Appendix B) and given the above explanations, I have not brought further modification to the manuscript in this regard.
 
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-[...] The myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time-window, until eventually reaching 2050.
+[...] The myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window, until eventually reaching 2050.
 \end{mdframed}
@@
 \caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
-Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time-window.
+Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window.
 If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - p.18 : Surprising that myopic leads to a sooner investment to the transition ?
+{\color{teal} \textbf{Stefan}} - p.18 : Surprising that myopic leads to a sooner investment to the transition?
 Is it an effect of the salvage value?
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p.13: Why no annualizing over the lifetime of the unit? Would that lead to different results?
-
 {\color{orange} \textbf{Stefano}} - Can you explain the logic of the annualization factor in the pathway model?
+
+\noindent {\color{purple} \textbf{Sylvain}} - p.13: Why no annualizing over the lifetime of the unit? Would that lead to different results? 
 \end{mdframed}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.71 - Differentiated discount rates could be useful.
+\end{mdframed}
+
+\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
+(1.3) when discussing the annualised phase factor}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
+In practice, the discount rate would vary depending on the technology investment risk.
+Depending on the technology readiness level, private investors would be more or less risk-averse.
+However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+For this reason, we have decided to keep an identical value of discount rate for all the technologies.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{teal} \textbf{Stefan}} - Interest rate must be revisited if a paper is published.
@@
 It goes the same way for the infrastructures to transport energy carriers within the system.
-As it is represented assuming a copper-plate approach, all the demands/resources/conversion technologies are concentrated in one single node.
+As it is represented assuming a copper-plate approach, all the demands, resources and conversion technologies are concentrated in one single node.
 This is now explicitly written {\color{blue}in a new paragraph of Section 2.2}:
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 20: ``inspired by Guevara et al.'' $\rightarrow$ how was their approach used in your work?
+{\color{orange} \textbf{Stefano}} - p. 20 - ``inspired by Guevara et al.'' $\rightarrow$ how was their approach used in your work?
 \end{mdframed}
@@
 However, despite its shorter computation time (\ie couple of seconds), the main drawback of the monthly model is its poor representation of the integration of VRES \cite{limpens2024pathway}.
-
 \end{mdframed}
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p.
+57: Why a uniform distribution?
+A gaussian would clearly be more appropriate.
+Maybe some uncertainties are underestimated (because clipped by uniform distribution).
+\end{mdframed}
+
+\noindent As already written in the first paragraph of Section 1.2.1, given the scarcity of data to actually define a distribution of uncertainty for the parameters, the common practice is to consider uniform distribution.
+As detailed by \citet{coppittersthesis}: ``Distributions are typically defined based on large datasets, which are not always at hand for each parameter.
+Therefore, when no meaningful information can be extracted on the distribution from a limited dataset, a uniform distribution is typically assumed, which assigns an equal probability to each value within a range.
+This approach is similar to assigning an interval to a parameter, but takes advantage of the central tendency when uniform distributions are propagated through a system model (\ie central limit theorem)''.
+Given what is already in the thesis manuscript and these further explanation, no modification has been brought to the manuscript in this regard.
+
 \subsection{Reinforcement Learning}
@@
 \noindent Most of the data about the Belgian case study have been collected by \citet{limpens2021generating}.
-The objective of this Chapter was to focus on my contributions: non-energy demand (NED), electrofuels, SMR, uncertain ranges for the pathway model and \ce{CO2}-budget.
+The objective of this Chapter was to focus on my contributions: non-energy demand (NED), electrofuels, SMR, uncertain ranges for the pathway model and \ce{CO2} budget.
 However, to put these contributions in perspective with similar features in the model, I have decided to keep general information about the end-use demands, the resources and the conversion technologies.
@@
 
-\subsection{End-use demands}
-\label{methodo_eud}
-
 \subsection{Resources}
@@
 \noindent There was indeed a confusion in the information to deliver in this paragraph.
-For the sake of clarity, I have restructured Section 2.2.
+For the sake of clarity, I have restructured {\color{blue}Section 2.2}.
 After the introductory paragraph, the graphs and the following paragraph focus on the cost of purchasing.
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p.
-57: Why a uniform distribution?
-A gaussian would clearly be more appropriate.
-Maybe some uncertainties are underestimated (because clipped by uniform distribution).
-\end{mdframed}
-
-\noindent As already written in the first paragraph of Section 1.2.1, given the scarcity of data to actually define a distribution of uncertainty for the parameters, the common practice is to consider uniform distribution.
-As detailed by \citet{coppittersthesis}: ``Distributions are typically defined based on large datasets, which are not always at hand for each parameter.
-Therefore, when no meaningful information can be extracted on the distribution from a limited dataset, a uniform distribution is typically assumed, which assigns an equal probability to each value within a range.
-This approach is similar to assigning an interval to a parameter, but takes advantage of the central tendency when uniform distributions are propagated through a system model (\ie central limit theorem)''.
-Given what is already in the thesis manuscript and these further explanation, no modification has been brought to the manuscript in this regard.
-
 \section{Chapter 3 - Atom-vs-molecules}
 \label{Chap_atom_vs_molecules}
+
+\subsection{General}
 
@@
 \item Given the set of assumptions, especially considering the end-use-demands and the local potential of VRES, Belgium will need to import molecules in the near future.
-On top of investing into VRES, it seems reasonable to invest in the imports of these electrofuels as betting on SMR means letting the short-term emissions go up and, potentially, requiring significant (and quick) adjustments in case the technology is not ready on time to still respect the \ce{CO2}-budget.
+On top of investing into VRES, it seems reasonable to invest in the imports of these electrofuels as betting on SMR means letting the short-term emissions go up and, potentially, requiring significant (and quick) adjustments in case the technology is not ready on time to still respect the \ce{CO2} budget.
 \end{itemize}
 
-Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-70: ``in other words, the variation of the parameter $f_{\mathrm{max,SMR}}$…''.
-I am not sure that this sentence is equivalent to the previous one.
-Please clarify.
-\end{mdframed}
-
-\noindent The Sobol' index is the ratio between the variance of the total transition cost due to the variation of one parameter and the total variance of the total transition cost.
-If $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the impact on the system, and, consequently, on its total transition cost, is the same: no SMR installed.
-A wide range of variation of the parameter leading to zero variation of the output of interest is one of the reasons to explain why the Sobol' index of this parameter is low.
-To clarify this point, I have rephrased the sentence {\color{blue}in Section 3.2.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Given the uncertainty characterisation presented in Section 2.4, there are 60\% chance that no SMR could be installed.
-\textbf{When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
-}
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 71, Table 3.1: isn’t the low impact simply due to the low assumed potential? Is this finding meaningful?
-
-\noindent {\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
-\end{mdframed}
-
-\noindent These assumed low potential and CAPEX are indeed some reasons why the SMR-related uncertain parameters have a limited impact on the variation of the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
-However, this is not the only reasons as detailed earlier in the same paragraph: the late potential availability (2040 at the soonest) as well as the ``cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.''.
-
-Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 74, Figure 3.8: this graphs looks potentially interesting but it is not sufficiently well explained.
+Given these explanations and what is already written in the text, I have not brought further modification to the manuscript in this regard.
+
+\subsection{Graphs}
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 74 - Figure 3.8: this graphs looks potentially interesting but it is not sufficiently well explained.
 \end{mdframed}
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 63 What does efficiency change in the installed capacity?
-\end{mdframed}
-
-\noindent Indeed, the efficiency only has an impact on the amount of primary uranium needed to produce the electricity.
-Consequently, I have removed this part of the sentence {\color{blue}in the first paragraph of Section 3.1.1}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Overall, given the restriction on yearly availability and the slightly higher electrification [...]
-\end{mdframed}
+\subsection{SMR}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+70: ``in other words, the variation of the parameter $f_{\mathrm{max,SMR}}$…''.
+I am not sure that this sentence is equivalent to the previous one.
+Please clarify.
+\end{mdframed}
+
+\noindent The Sobol' index is the ratio between the variance of the total transition cost due to the variation of one parameter and the total variance of the total transition cost.
+If $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the impact on the system, and, consequently, on its total transition cost, is the same: no SMR installed.
+A wide range of variation of the parameter leading to zero variation of the output of interest is one of the reasons to explain why the Sobol' index of this parameter is low.
+To clarify this point, I have rephrased the sentence {\color{blue}in Section 3.2.1}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Given the uncertainty characterisation presented in Section 2.4, there are 60\% chance that no SMR could be installed.
+\textbf{When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
+}
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 71, Table 3.1: isn’t the low impact simply due to the low assumed potential? Is this finding meaningful?
+
+\noindent {\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
+\end{mdframed}
+
+\noindent These assumed low potential and CAPEX are indeed some reasons why the SMR-related uncertain parameters have a limited impact on the variation of the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
+However, this is not the only reasons as detailed earlier in the same paragraph: the late potential availability (2040 at the soonest) as well as the ``cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.''.
+
+Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 67 - It really depends on what is considered primary energy in the case of nuclear
+\end{mdframed}
+
+\noindent As specified in the table about SMR features (see previous comment), the considered fuel for SMR is uranium, as it is the case for conventional nuclear power plants.
+Given this, I have not brought modification to the manuscript in this regard.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 78 - Why this threshold? Shouldn't it follow the potential capacity?
+\end{mdframed}
+
+\noindent The threshold locates at 60\% of the uncertain parameter $f_{\mathrm{max,SMR}}$ which represents the potential capacity of SMR.
+As presented in Section 2.4, this threshold corresponds to the possibility to install up to 6\,GW of SMR in 2050.
+As already written in Section 3.2.2: ``For the installed capacity of SMR, it is the availability of the technology that drives its installation (Figure 3.12).
+Not shown here but all the samples of the GSA highlight that SMR is installed to its maximum capacity, \ie 6GW, as soon as possible.'' To smooth the trend curves of the different uncertain parameters, I have applied a rolling average.
+For this reason, the 60\%-threshold does not appear on the Figure as sharp as it actually is.
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
+
+\subsection{PV and wind}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 67 - It really depends on what is considered primary energy in the case of nuclear
-\end{mdframed}
-
-\noindent As specified in the table about SMR features (see previous comment), the considered fuel for SMR is uranium, as it is the case for conventional nuclear power plants.
-Given this, I have not brought modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{purple} \textbf{Sylvain}} - p. 69 - No curtailment at all? It would be nice to have the exact number.
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p.71 - Differentiated discount rates could be useful.
-\end{mdframed}
-
-\noindent As written in Chapter 3: ``In practice, the discount rate would vary depending on the technology investment risk''.
-Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
-These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
-(1.3) when discussing the annualised phase factor}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-The discount rate accounted for in the annualisation factor is considered as identical for all the technologies of the system.
-In practice, the discount rate would vary depending on the technology investment risk.
-Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
-For this reason, we have decided to keep an identical value of discount rate for all the technologies.
+\subsection{Others}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 63 What does efficiency change in the installed capacity?
+\end{mdframed}
+
+\noindent Indeed, the efficiency only has an impact on the amount of primary uranium needed to produce the electricity.
+Consequently, I have removed this part of the sentence {\color{blue}in the first paragraph of Section 3.1.1}.
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
+Overall, given the restriction on yearly availability and the slightly higher electrification [...]
 \end{mdframed}
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 78 - Why this threshold? Shouldn't it follow the potential capacity?
-\end{mdframed}
-
-\noindent The threshold locates at 60\% of the uncertain parameter $f_{\mathrm{max,SMR}}$ which represents the potential capacity of SMR.
-As presented in Section 2.4, this threshold corresponds to the possibility to install up to 6\,GW of SMR in 2050.
-As already written in Section 3.2.2: ``For the installed capacity of SMR, it is the availability of the technology that drives its installation (Figure 3.12).
-Not shown here but all the samples of the GSA highlight that SMR is installed to its maximum capacity, \ie 6GW, as soon as possible.'' To smooth the trend curves of the different uncertain parameters, I have applied a rolling average.
-For this reason, the 60\%-threshold does not appear on the Figure as sharp as it actually is.
-
-Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - Lot of the results hinge on the import of e-fuels with the assumption that they have GWP = 0; is it justified? Those fuels are directly available knowing they are no.
-\end{mdframed}
-
-\noindent To succeed its energy transition, Belgium will need radical changes, ``unicorn'' solutions, and one of them would be the import of electrofuels.
-As presented in the introduction and reminded in the conclusion of the manuscript: ``Chemically similar to their fossil-based equivalents, they [the electrofuels] can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.'' Besides these intrinsic advantages, considering these fuels as infinitely available (like their fossil equivalents) and with 0-GWP allow finding a solution for the Belgian case study, without considering a decrease of the demands.
-These are strong assumptions that can be challenged as suggested in the conclusion: ``A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
-This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
-This could also show the limits of the Belgian energy system to meet such an ambitious CO2-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
-''
-
-The assumption about 0-GWP molecules is similar to the one considered \cite{PATHS2050} which is the most recent study that addressed the case of the Belgian energy transition.
-
-As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.
-This need for this radical change and the fact that solutions rely a lot on assumed to be carbon-neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
-This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
-\end{mdframed}
-
-and reminded {\color{blue}in the conclusions of the manuscript}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
-One robust policy would consist in massively investing in electrofuels.
-\end{mdframed}
-
 \section{Chapter 4 - Reinforcement Learning}
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
-\end{mdframed}
-
-\noindent Because the action limits the emissions of the system at the end of the 10-year time-window whereas limiting the consumption of fossil gas, LFO and coal is done over the whole time-window, as detailed in Section 4.1.1.
-Given what is already written in the manuscript, I have not brought modification to the manuscript in this regard.
-
-\subsection{Graphs and results}
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods like RL?
-We could go further in the analysis as we know we have to phase out coal, install renewables.
-\end{mdframed}
-
-\noindent The outcome is a bit too obvious (e.g.
-getting rid of the coal) $\rightarrow$ go a bit in the details.
-Explain why the method is taking better decisions but explain why.
-$\rightarrow$ go beyond what we know
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained.
-In general, improve the presentation of the results, especially for the RL chapter.
-\end{mdframed}
-
-\noindent
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-
-\end{mdframed}
+\subsection{General}
 
@@
 {\color{orange} \textbf{Stefano}} - p.
-89: Figure 4.3 seems to deliver an important message (early action is needed to ensure we meet climate goals), but it is not clearly explained.
-Improve captions / add axes.
-	Where do I see the ``tipping points'' in this figure?
-\end{mdframed}
-
-\noindent Like other figures in this Chapter, Figure 4.3 represents distributions of features spreading over the x-axis.
-In this case, these are distributions of the rewards collected by the agent at the end of each episode of its learning process.
-For this reason, labelling the axis does not seem obligatory as the x-axis is nothing else than the feature (here, the reward) and the y-axis has no physical meaning expect representing how often the feature of interest took a certain value/range of values.
-Compared to the other similar figures of the chapter, I have not labelled the x-axis as, for Figure 4.3, the reward is the only feature represented whereas two or more features are gathered on the same figure.
-
-With this in mind and what is already written in the caption of the figure and the core of the text, we can infer from the right hand side that 2040 is the ``tipping year'' as it represents the year at which the failures occur more often, compared to the other representative years.
-
-Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
-\end{mdframed}
-
-\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2}-budget) by 2050 at the latest.
-The number of remaining attempts reduce as prematurely ended episode are removed from the further counts.
-The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves.
-These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-94: Figure 4.6 seems to be less revealing than other figures.
-Also the figure caption seems to contradict the first statement following the figure.
-It would be important to clarify: which are the actions that emerge to actually matter?
-\end{mdframed}
-
-\noindent Indeed, Figure 4.6 is less revealing as it shows no big trend in terms of neither impacting actions nor their timing in the transition.
-It is detailed in the first paragraph of Section 4.2.3 after which the figure should appear in the manuscript.
-I have worked to make Figures as close as possible to their related paragraph.
-However, in some cases, several other paragraphs come between them.
-To identify the actions that emerge to actually matter, we need to check if they are binding or not, as detailed in the subsequent paragraphs and in Figure 4.7.
-
-Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard, except minor adjustment to get Figure 4.6 as close as possible to its related paragraph.
-
-\subsection{Others}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-85, Figure 4.1: Does it mean we can run ``infinite energy transitions''?
+85 - Figure 4.1: Does it mean we can run ``infinite energy transitions''?
 Is this realistic?
@@
 The purpose of Figure 4.1 was to focus on the reward function and how it articulates in the action-state-reward process.
-As already represented in this figure, three situations can terminate an episode: (i) infeasible optimisation, (ii) overshot \ce{CO2}-budget before 2050 or, (iii) reaching 2050.
+As already represented in this figure, three situations can terminate an episode: (i) infeasible optimisation, (ii) overshot \ce{CO2} budget before 2050 or, (iii) reaching 2050.
 This is now clarified {\color{blue}in the caption of Figure 4.1}:
@@
 Reward function, $R$.
-Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded.
-If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues.
+Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded.
+If the optimisation provides a solution and the \ce{CO2} budget is not exceeded, the episode continues.
 Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates.
@@
 
-\noindent Complementary to the text in the core of the manuscript, I have added a diagram to detail the agent's actions and how they impact the model: taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
+\noindent Complementary to the text in the core of the manuscript, I have added a diagram to detail the agent's actions and how they impact the model: taken at the beginning of the time window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
 
@@
 \caption{Actions available to the decision-maker.
-Taken at the beginning of the time-window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time-window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
-Unlike the first action that sets a target for the end of the time-window, the last three aim at limiting the consumption of these fossil resources over the whole time-window.}
+Taken at the beginning of the time window to optimise (year $Y$), the four actions impact (i) the emissions of the system at the end of the time window (year $Y+10$) and, (ii-iv) the consumption of fossil gas, LFO and coal at years $Y+5$ and $Y+10$.
+Unlike the first action that sets a target for the end of the time window, the last three aim at limiting the consumption of these fossil resources over the whole time window.}
 \label{fig:Schematic_actions}
@@
 \noindent Besides the general concepts/fundamentals detailed in Section 1.3.1, the application of the RL approach on the myopic optimisation of the energy transition of a whole-energy system is detailed in Section 1.3.2.
-Now, Figure 1.7 has been updated to clarify the difference between the uncertainty of parameters and the agent's actions as well as the interactions between the RL code (action, state, reward) and EnergyScope Pathway (optimisation of the concerned 10-year time-window).
+Now, Figure 1.7 has been updated to clarify the difference between the uncertainty of parameters and the agent's actions as well as the interactions between the RL code (action, state, reward) and EnergyScope Pathway (optimisation of the concerned 10-year time window).
 I have also added a diagram to explain the impact that the agent's actions have on some parameters of the model.
@@
 Given what was already in the manuscript and the aforementioned modifications, I have not added further information to detail the different steps of the RL approach.
+
+\subsection{Graphs and results}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - Work on the graph of slide 19 (there is a misunderstanding on the fact that the grey curve is labelled as failure but they are failures in the future)
+\end{mdframed}
+
+\noindent To clarify this point, I have extended the caption of the related figures (see previous comments) and added two sentences in {\color{blue} Section 4.2.2}:
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+Since infeasible cases or those that overshoot the \ce{CO2} budget are discarded before reaching 2050 (see Section 4.1.2), the number of attempts that reach further steps in the transition progressively decreases.
+Consequently, the share of successful transitions compared to failures progressively increases with time.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+\noindent {\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained.
+In general, improve the presentation of the results, especially for the RL chapter.
+\end{mdframed}
+
+\noindent Beside extending the captions and clarifying the reasons why the number of total attempts decreases when going down the graph, as detailed in the previous comment, I have brought smaller modifications into the text to improve the presentation of the results, especially in the RL chapter.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{teal} \textbf{Stefan}} - What is the potential of what we can learn from the results of the innovative methods like RL?
+We could go further in the analysis as we know we have to phase out coal, install renewables.
+\end{mdframed}
+
+\noindent First, I focus here below on each of the results sections of Chapter 4 highlighting the key conclusions already in the manuscript:
+
+\begin{itemize}
+\item \textbf{4.2.1 Reward and success}: (i) We have convergence of the learning process; (ii) Limiting success rate given the wide uncertainty ranges of some key parameters and the set of actions given to the agent; (iii) Near-term actions are needed.
+\item \textbf{4.2.2 States}: (i) Validation of the formulation of the reward function (first respect the \ce{CO2} budget then minimise the cost); (ii) Reminder of the need for near-term actions; (iii) No-go zones in terms of share of renewables in the primary mix; (iv) Efficiency of the system does not give effective information for the agent to act upon.
+\item \textbf{4.2.3 Actions}: (i) Limiting GWP and consumption of fossil gas are the keys; (ii) Even in a myopic perspective, limiting LFO is ineffective as it is phased out by the cost-driven model; (iii) Unlike coal that would still be part of the mix if the agent did not limit its consumption.
+\item \textbf{4.3 Comparison with PF}: (i) Following \ce{CO2}-trajectory prescribed by PF would likely lead to failing the transitions with limited knowledge into the future; (ii) Succeeding myopic transitions relies more often on importing more electrofuels, mostly e-ammonia.
+\end{itemize}
+
+Phasing out coal is, of course, necessary as it is a highly-polluting resource.
+However, what this work shows is that, whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
+Consistency towards banning coal through the whole transition is needed.
+To insist on this aspect, this has been added {\color{blue}at the end of Section 4.2.3}:
+
+\begin{mdframed}[style=manuscript] 
+Whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
+Consistency towards banning coal is needed.
+\end{mdframed}
+
+About the installation of renewables, not only it is (obviously) necessary to maximise the chance to succeed the transition (see Chapters 3 and 4) but it also provides robustness towards additional investments to make when deploying a roadmap through a (more realistic) myopic transition (see Chapter 5).
+This conclusion was already drawn by \citet{moret2020overcapacity} for the snapshot model and targeting the power sector only.
+Here, we extend this work and validate this conclusion for the whole transition pathway and addressing the additional investments in all the sectors.
+
+Through the application of RL on the optimisation of the transition pathway, we are able to assess the effectiveness of a policy, which could be of interest to other researchers.
+This has been added {\color{blue}at the end of the conclusions of Chapter 4}:
+
+\begin{mdframed}[style=manuscript] 
+This framework allows future research to test other policies on other case studies and identify the key actions and the timing to take them through the transition.
+\end{mdframed}
+
+In conclusion, besides the results coming from the application to the specific case study of Belgium, a major added-value of this work is also to validate innovative approaches (RL and PCA) when applied to explore and optimise transition pathways under uncertainties.
+For instance, Section 4.2.2 shows that the reward function was properly defined to aim first at respecting the \ce{CO2} budget then minimising the cost.
+It also provides innovative ways to visually present results in the field of the optimisation of transition pathway under uncertainties that suffers from the (overwhelming) curse of dimensionality.
+Via distributions or principal components, this work summarises a lof of information into graspable visuals that allow drawing effective conclusions such as the identification of sweet-spots and no-go zones.
+
+Given these explanations and what is already written in the manuscript, I have not brought other modification to the text.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+89: Figure 4.3 seems to deliver an important message (early action is needed to ensure we meet climate goals), but it is not clearly explained.
+Improve captions / add axes.
+	Where do I see the ``tipping points'' in this figure?
+\end{mdframed}
+
+\noindent Like other figures in this Chapter, Figure 4.3 represents distributions of features spreading over the x-axis.
+In this case, these are distributions of the rewards collected by the agent at the end of each episode of its learning process.
+For this reason, labelling the axis does not seem obligatory as the x-axis is nothing else than the feature (here, the reward) and the y-axis has no physical meaning expect representing how often the feature of interest took a certain value/range of values.
+Compared to the other similar figures of the chapter, I have not labelled the x-axis as, for Figure 4.3, the reward is the only feature represented whereas two or more features are gathered on the same figure.
+
+With this in mind and what is already written in the caption of the figure and the core of the text, we can infer from the right hand side that 2040 is the ``tipping year'' as it represents the year at which the failures occur more often, compared to the other representative years.
+
+Given these explanations and what is already in the text, I have not brought modifications to the manuscript in this regard.
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p. 90: What does it mean to succeed/fail in 2030? And why are the number of attempts reduced going down in the graph?
+\end{mdframed}
+
+\noindent As discussed during the private defense, the grey curves represent the distribution of occurrences failing (\ie infeasible optimisation or overshooting \ce{CO2} budget) by 2050 at the latest.
+The number of remaining attempts reduce as prematurely ended episode are removed from the further counts.
+The purpose of displaying these numbers of ``Total attempts'' is to explain why the areas below the orange curves increase relatively to the grey curves.
+These two aspects are now explicitly explained {\color{blue}in the caption of Figure 4.4 and the other similar figures}:
+
+\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
+Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
+\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case.
+\end{mdframed}
+
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{orange} \textbf{Stefano}} - p.
+94: Figure 4.6 seems to be less revealing than other figures.
+Also the figure caption seems to contradict the first statement following the figure.
+It would be important to clarify: which are the actions that emerge to actually matter?
+\end{mdframed}
+
+\noindent Indeed, Figure 4.6 is less revealing as it shows no big trend in terms of neither impacting actions nor their timing in the transition.
+It is detailed in the first paragraph of Section 4.2.3 after which the figure should appear in the manuscript.
+I have worked to make Figures as close as possible to their related paragraph.
+However, in some cases, several other paragraphs come between them.
+To identify the actions that emerge to actually matter, we need to check if they are binding or not, as detailed in the subsequent paragraphs and in Figure 4.7.
+
+Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard, except minor adjustment to get Figure 4.6 as close as possible to its related paragraph.
+
+\subsection{Others}
 
@@
 
-\noindent No-go zones are introduced in the second paragraph of Chapter 4: ``[...] as well as no-go zones where the chances of succeeding the transition, i.e., respecting the \ce{CO2}-budget, are very limited.'' It is then detailed at the end of Section 4.2.2: ``Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except
+\noindent No-go zones are introduced in the second paragraph of Chapter 4: ``[...] as well as no-go zones where the chances of succeeding the transition, i.e., respecting the \ce{CO2} budget, are very limited.'' It is then detailed at the end of Section 4.2.2: ``Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except
 if betting on the future installation of SMR. '' and in the caption of the related figure. Given the information already present in the manuscript, no modification has been brought in this regard.
@@
 
+\begin{mdframed}[style=comment] % Comment from the reviewer
+{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
+\end{mdframed}
+
+\noindent Because the action limits the emissions of the system at the end of the 10-year time window whereas limiting the consumption of fossil gas, LFO and coal is done over the whole time window, as detailed in Section 4.1.1.
+Given what is already written in the manuscript, I have not brought modification to the manuscript in this regard.
+
 \section{Chapter 5 - Principal Component Analysis}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 101: you mention different definitions of ``robustness'': which one do you use in your work?
+{\color{orange} \textbf{Stefano}} - p. 101 - you mention different definitions of ``robustness'': which one do you use in your work?
 \end{mdframed}
@@
 On the other hand, after drawing a robustness-metric based on a Principal Component Analysis (PCA) approach, this work investigates how much a planned roadmap is sensitive to additional investments required in a myopic transition with progressively unveiled uncertainties.
-In other words, we use two different definitions of robustness: (i) the ability of a policy to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget, (ii) the ability of a technological roadmap to limit the investment into additional capacities when tested in myopic conditions.
+In other words, we use two different definitions of robustness: (i) the ability of a policy to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2} budget, (ii) the ability of a technological roadmap to limit the investment into additional capacities when tested in myopic conditions.
 
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Chapter 4 focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget.
+Chapter 4 focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2} budget.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 109: how is Gamma-robustness implemented here? It seems like it is in some sense ``manually'' forced? Explain better how the ROB solution is derived.
+{\color{orange} \textbf{Stefano}} - p. 109 - how is Gamma-robustness implemented here? It seems like it is in some sense ``manually'' forced? Explain better how the ROB solution is derived.
 \end{mdframed}
@@
 Consequently, the ROB technological roadmap is obtained from the deterministic optimisation of the perfect foresight pathway accounting for these six parameters (\ie costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry EUD, the interest rate and the variable OPEX of technologies) at their respective worst-case value.
-As it does not exactly follow the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}, I have changed {\color{blue}the first sentence of Section 5.2.1:
+As it does not exactly follow the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}, I have changed {\color{blue}the first sentence of Section 5.2.1}:
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{violet} \textbf{Christophe}} When using the word ``component'', there seems to be a confusion and it is not always easy to understand if you refer to the vector or the coefficient related to one of the original variable.
+{\color{violet} \textbf{Christophe}} - When using the word ``component'', there seems to be a confusion and it is not always easy to understand if you refer to the vector or the coefficient related to one of the original variable.
 \end{mdframed}
@@
 {\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - p.
-119: the statement mentioning the ``major added value of the perfect foresight'' approach seems surprising and needs to be clarified.
+119 -  the statement mentioning the ``major added value of the perfect foresight'' approach seems surprising and needs to be clarified.
 The decreasing \ce{CO2} emissions trajectories shouldn't be a S-shaped (logistic) curve?

### 2024-07-19 appendices/app_RL_TD_MO.tex
Rebuttals: Keep going...
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 The \acrfull{RL}-based exploration of myopic transition pathways can require thousands of runs to converge.
-To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time-resolution, the so-called transfer learning \cite{mann2013directed}.
+To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time resolution, the so-called transfer learning \cite{mann2013directed}.
 Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
@@
 Limiting the consumption of coal is always binding where \gls{LFO} is ``naturally'' removed from the mix by the cost-minimisation.
-Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon-neutrality becomes more crucial at the end of the transition.
+Where limiting the consumption of fossil gas has a major impact at the early stages, limiting the system emissions up to carbon neutrality becomes more crucial at the end of the transition.
 
@@
 \includegraphics[width=0.49\textwidth]{Binding_learning_MO_joss.pdf}
-\caption{Binding versus non-binding constraints on the hourly and monthly models.
+\caption{Binding versus non-binding constraints on the hourly and monthly models: distribution of occurrences.
 When the agent learns on the monthly model, the actions are binding in the similar timing and similar ranges as when learning on the hourly model.
@@
 
-In conclusion, with the advantage of the smaller computational burden, the monthly model myopic model is found to be a good proxy with its equivalent based on hourly time-resolution.
+In conclusion, with the advantage of the smaller computational burden, the monthly model myopic model is found to be a good proxy with its equivalent based on hourly time resolution.

### 2024-07-19 appendices/app_case_study.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 
-\section{Belgian energy transition pathway towards carbon-neutrality in 2050} 
+\section{Belgian energy transition pathway towards carbon neutrality in 2050} 
 \label{app:bel_PF_TD}
-This section presents the results of the deterministic (\ie all parameters at their nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon-neutrality in 2050.
+This section presents the results of the deterministic (\ie all parameters at their nominal value) perfect foresight optimisation of the Belgian energy transition pathway constrained to a linear decrease of the \gls{GHG} emissions from 2020 (121~MtCO\textsubscript{2,eq}) to carbon neutrality in 2050.
 After performing a technical investigation of the pathway by checking the greenhouse gas breakdown by energy sectors, the primary energy mix is analysed.
@@
 \caption{Energy sectors have different rates to reduce \gls{GHG} emissions over the transition.
-The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon-neutrality in 2050.}
+The system uses all the allowed \gls{GHG} prescribed by the linear decrease from the emissions in 2020 until carbon neutrality in 2050.}
 \label{fig:pestd_ghg}
@@
 \label{app:CO2_trajectories}
-This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon-neutrality in 2050.
-The first comparison is done with the trajectory subject to respect the \ce{CO2}-budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}).
+This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon neutrality in 2050.
+The first comparison is done with the trajectory subject to respect the \ce{CO2} budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}).
 Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions-trajectory.
 
-\subsection{\ce{CO2}-budget versus linear decrease of emissions}
+\subsection{\ce{CO2} budget versus linear decrease of emissions}
 \label{app:CO2_budget}
 
-\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2}-budget) and a case where the \ce{CO2}-trajectory is constrained instead.
-Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050.
+\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2} budget) and a case where the \ce{CO2}-trajectory is constrained instead.
+Interestingly, these two transition pathways end up in a similar carbon neutral whole-energy system in 2050.
 The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat.
@@
 \includegraphics[width=\textwidth]{GWP_per_sector_REF.pdf}
-\caption{Respecting the \ce{CO2}-budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
+\caption{Respecting the \ce{CO2} budget imposed in the REF case drastically cuts the emissions of the system, especially in the production of \glsxtrfull{HVC} and the high-temperature heating sector.}
 \label{fig:app_CO2_REF_lin}
@@
 \begin{table}[htbp]
-\caption{Comparison between the two different time-resolutions: Hourly (PF) and Monthly (MO).
+\caption{Comparison between the two different time resolutions: Hourly (PF) and Monthly (MO).
 Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}

### 2024-07-19 appendices/app_results_UQ.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 However, this does not jeopardize the comparative analysis given the similar Sobol' indices.
-Given their wide range of uncertainty [-64.3\%; 179.8\%] and their significant role to meet the \ce{CO2}-budget, the cost of purchasing electrofuels is the first, by far, impacting parameter.
+Given their wide range of uncertainty [-64.3\%; 179.8\%] and their significant role to meet the \ce{CO2} budget, the cost of purchasing electrofuels is the first, by far, impacting parameter.
 Next, comes, naturally, the industrial \gls{EUD}, representing, at the nominal value, 60\% of the total demands by 2050.
@@
 Given its lower cost of purchasing than its renewable equivalent (\Cref{fig:cs_resources_cost}) and lower \gls{GWP} than other fossil fuels (\ie $\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh versus $\mathit{gwp}_{\mathrm{op,LFO}}=0.31$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), fossil \gls{NG} remains the main source of gas in the system until 2040.
-Besides bio-hydrolysis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2}-budget for the transition.
+Besides bio-hydrolysis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2} budget for the transition.
 As a versatile energy carrier, gas is used by a wide variety of technologies in the different sectors.
@@
 Similarly to ammonia, on top of local production in rare cases, e-methanol is the key source for methanol.
-Besides its own \gls{NED}, methanol is mostly consumed to produce \gls{HVC} via the \gls{MTO} process instead of naphtha-cracking, in order to respect to \ce{CO2}-budget for the transition.
+Besides its own \gls{NED}, methanol is mostly consumed to produce \gls{HVC} via the \gls{MTO} process instead of naphtha-cracking, in order to respect to \ce{CO2} budget for the transition.
 

### 2024-07-19 appendices/app_taxonomy.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 Consequently, in the context of the energy transition where new use of primary resources and conversion technologies are booming, we encompass in this work all the resources and conversion processes one might consider in the production of fuels.
-The objective of this taxonomy is to fit with the long-term perspective of this transition (\eg carbon-neutrality, defossilisation) while being adapted for the current context.
+The objective of this taxonomy is to fit with the long-term perspective of this transition (\eg carbon neutrality, defossilisation) while being adapted for the current context.
 We do not only focus on carbon-based fuels nor on specific end-use sectors (\eg transport), contrary to what is mostly seen in the literature.
@@
 \textit{Electrofuels} are, according to \citet{JRC_alternative}, fuels whose energy content comes from renewable electricity.
-On top of that, \citet{goldmann2018} even add that they should be carbon-neutral and serve as a storage capacity of electricity.
+On top of that, \citet{goldmann2018} even add that they should be carbon neutral and serve as a storage capacity of electricity.
 To be more comprehensive, we define \textit{electrofuels as fuels produced from electricity}, as the resource used to produce this electricity may vary.

### 2024-07-19 chap_RL/chap_RL.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.
-Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2}-budget, are very limited.
+Then, the results of this policy optimisation point out strategies to follow, \ie \textit{sweet spots}, in the transitions under uncertainties as well as \textit{no-go zones} where the chances of succeeding the transition, \ie respecting the \ce{CO2} budget, are very limited.
 Finally, these results are compared with references, \ie the perfect foresight and the myopic optimisation of the transition under the same uncertainties but without the trained \gls{RL}-agent that can support this transition thanks to its learned policy.
@@
 
-Applying \gls{RL} to the optimisation environment of EnergyScope Pathway myopic,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework: while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2}-budget.
+Applying \gls{RL} to the optimisation environment of EnergyScope Pathway myopic,  rather than a simulation environment, allows building a hierarchical multi-objective optimisation framework: while the objective of the environment remains the minimisation of the total ``transition'' cost (on the concerned limited time window), the agent optimises its strategy to respect the \ce{CO2} budget.
 
@@
 In other words, considering only the techno-economic lens, there is no moderation nor contest towards the agent's actions, as the objective is to assess how far and when within the transition to push the different levers of action.
-Given the overall objective of the agent to succeed the transition, \ie respecting the \ce{CO2}-budget by 2050, we have defined the actions in this sense.
-The first action, $\mathrm{act}_{\mathrm{gwp}} \in [0,1]$, aims at limiting the emissions at the representative year ending the concerned time window, $\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})$, between the level of emissions in 2020, \ie $\textbf{GWP\textsubscript{tot}}(2020)=123\,\text{Mt}_{\ce{CO2},\text{eq}}$, and carbon-neutrality:
+Given the overall objective of the agent to succeed the transition, \ie respecting the \ce{CO2} budget by 2050, we have defined the actions in this sense.
+The first action, $\mathrm{act}_{\mathrm{gwp}} \in [0,1]$, aims at limiting the emissions at the representative year ending the concerned time window, $\textbf{GWP\textsubscript{tot}}(y_{\text{end of the window}})$, between the level of emissions in 2020, \ie $\textbf{GWP\textsubscript{tot}}(2020)=123\,\text{Mt}_{\ce{CO2},\text{eq}}$, and carbon neutrality:
 
@@
 \caption{Reward function, $R$.
-Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded.
-If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues.
+Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded.
+If the optimisation provides a solution and the \ce{CO2} budget is not exceeded, the episode continues.
 Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates.
@@
 In other words, as actions have a direct impact on some constraints of the problem, they might limit too much the feasible domain to the point where no solution can be found.
-For instance, the extreme case of aiming at carbon-neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve.
+For instance, the extreme case of aiming at carbon neutrality, \ie $\mathrm{act}_{\mathrm{gwp}}=0$, and forbidding the use of the three aforementioned fossil fuels, \ie $\mathrm{act}\textsubscript{fossil gas}=\mathrm{act}\textsubscript{LFO}=\mathrm{act}\textsubscript{coal}=0$,  from the beginning of the transition makes the optimisation impossible to solve.
 In this case, the episode is prematurely ended and the reward is ``highly'' negative, -300.
 If the optimisation is feasible and the end of the transition, \ie 2050, is not reached, the cumulative emissions so far are evaluated.
-On the one hand, if these cumulative emissions exceed the \ce{CO2}-budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent.
-This penalisation is proportional to the difference between the \ce{CO2}-budget and the actual cumulative emissions.
-On the other hand, the episode continues with a zero reward if the \ce{CO2}-budget is not exceeded.
-Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2}-budget and not to be more ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2}-budget.
+On the one hand, if these cumulative emissions exceed the \ce{CO2} budget, $1.2\,\text{Gt}_{\ce{CO2},\text{eq}}$ (see Section \ref{sec:cs:CO2-budget}), the episode is also ended and a penalisation is given to the agent.
+This penalisation is proportional to the difference between the \ce{CO2} budget and the actual cumulative emissions.
+On the other hand, the episode continues with a zero reward if the \ce{CO2} budget is not exceeded.
+Eventually, when reaching 2050, given the main objective of the agent to respect the \ce{CO2} budget and not to be more ``\ce{CO2}-ambitious'', we cut short the contribution of the cumulative emissions as soon as they are lower or equal to the \ce{CO2} budget.
 On top of that, the reward function includes a secondary objective: the cumulative transition cost.
@@
 The values of these weights are the results of a trial and error to fine-tune the balance between more expensive successes and cheaper failures.
-This way, we observed that the agent first targeted the respect of the \ce{CO2}-budget and then, to a lesser scale, avoided reaching over-costly transitions.
+This way, we observed that the agent first targeted the respect of the \ce{CO2} budget and then, to a lesser scale, avoided reaching over-costly transitions.
 
@@
 Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window.
-The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2}-budget until 2050.
+The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2} budget until 2050.
 Therefore, the cumulative emissions emitted so far up to the current step of the transition is the first dimension of the states.
@@
 For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium.
-At the beginning of each episode, the agent starts with the Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}).
-Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window.
-This single draw will affect the parameters for all the subsequent time windows (see Figure \ref{fig:ranges_transition}).
-Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward.
-This goes on until the end of the episode.
-For the next episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), another sample of uncertain parameters is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.
+During the learning process, the algorithm explored numerous transition pathways: 2,037 successful transitions out of 10,751 attempts.
+2\% of the total attempts led to an infeasible optimisation of the initial time window.
+On top of that, each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning.
+As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken.
+Yet, the exploration during this learning process is not exhaustive.
+The trends provided below are therefore not proven.
+The randomness of the process and the number of explored transition pathways still give high confidence in the conclusions drawn from these analyses.
 
@@
 
-The learning phase has been split into batches of 500 steps,\ie 500 sequences of state-action-reward-new state.
+The learning phase has been split into batches of 500 steps, \ie 500 sequences of state-action-reward-new state.
 %For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts.
@@
 The mean reward increases rapidly at the beginning of the learning process before reaching a plateau where the optimisation of the policy becomes more marginal.
-As this successes drive indirectly the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes.
+As these successes indirectly drive the agent's optimisation, it shows that the reward function (see Figure \ref{fig:Reward}) leads towards more and more successes.
 However, given the wide range of uncertainty of some parameters and the agent's levers of action, this success rate stays limited at the end of the learning process.
@@
 \newpage
-During the learning process, the algorithm explored numerous transition pathways: 2,037 successful transitions out of 10,751 attempts.
-2\% of the total attempts led to infeasible optimisation of the initial time window.
-On top of that, each pathway provides valuable insight into the possible alternatives --- the primary goal of reinforcement learning.
-As a side benefit, the collection of all explored pathways also identifies the intermediate milestones to reach and the range of actions that must be avoided or must be taken.
-Yet, the exploration during this learning process is not exhaustive.
-The trends provided below are therefore not proven.
-The randomness of the process and the number of explored transition pathways still give high confidence in the conclusions drawn from these analyses.
+When assessing the distributions of the values of reward in the failure and success cases, one notices a range where these distributions overlap (see left hand side of Figure \ref{fig:reward_status}).
+This area corresponds to either transitions that exceed the \ce{CO2} budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive.
+Besides this overlap, we observe that successes account for the majority of the cases with higher reward.
+This is another indication that the reward function is appropriate in this exploration of successful transition pathways.
+More indirectly, in case of applying this methodology to another case study, this substantiates that the weights between emissions and cost defined in Section \ref{subsec:RL:act_states_rew:rew} could be used as an initial step and fine-tuned afterwards.
 
-There is a range in the reward where failures and successes overlap (see Figure \ref{fig:reward_status}).
-This area corresponds to either transitions that exceed the \ce{CO2}-budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive.
-Besides this overlap, we observe that successes account for the majority of the cases where the reward is positive.
-This is another indication that the reward function is appropriate in this exploration of successful transition pathways.
-More indirectly, this substantiates that the weights between emissions and cost defined in Section \ref{subsec:RL:act_states_rew:rew} could be used and adapted for another case study.
+Considering the end of the time window where the \ce{CO2} budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
+Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
+In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
 
@@
 \caption{Reward distribution between successes and failures.
-Graph on the right hand side details at the end of which time-window the failure occurred.
+Graph on the right hand side details at the end of which time window the failure occurred.
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
@@
 \end{figure}
-
-\newpage
-Considering the end of the time-window where the \ce{CO2}-budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
-Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
-In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
 
@@
 Per definition, the threshold of 1.2\,Gt$_{\ce{CO2},\text{eq}}$ splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
-Since infeasible cases or those that overshoot the \ce{CO2}-budget are discarded before reaching 2050 (see Section \ref{subsec:RL:act_states_rew:rew}), the number of attempts that reach further steps in the transition progressively decreases.
+Since infeasible cases or those that overshoot the \ce{CO2} budget are discarded before reaching 2050 (see Section \ref{subsec:RL:act_states_rew:rew}), the number of attempts that reach further steps in the transition progressively decreases.
 Consequently, the share of successful transitions compared to failures progressively increases with time.
 In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
-Reaching cumulative emissions significantly lower than the \ce{CO2}-budget are possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
-Considering the failures, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
+Reaching cumulative emissions significantly lower than the \ce{CO2} budget are possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
+Considering the failures in 2050, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
 As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
-Where 98\% of the failures were below the \ce{CO2}-budget in 2035, only 37\% passed this threshold in 2040.
+Where 98\% of the failures were below the \ce{CO2} budget in 2035, only 37\% passed this threshold in 2040.
 This reminds the importance of near-term, 2025-2030, actions to hope for a successful transition.
@@
 \caption{Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
-The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case.
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
+The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case.
 }
@@
 \newpage
-Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.
+Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2} budget.
 The skewness of the cumulative emissions and costs in 2050 are indications of this reward function (see Table \ref{tab:skewness_gwp_cost}).
-When succeeding the transitions, the cumulative emissions have a negative skewness: the agent successfully stayed within the budget and most of the cases where close to that budget (median at 0.9 Gt$_{\ce{CO2},\text{eq}}$).
+When succeeding the transitions, the cumulative emissions have a negative skewness: the agent successfully stayed within the budget and most of the cases were close to that budget (median at 0.9 Gt$_{\ce{CO2},\text{eq}}$).
 On the contrary, the cumulative cost of successful transitions has a positive skewness: the agent successfully reduces the cost of the system as a secondary objective with 30\% of the cases above the reference transition cost.
-This hierarchy of agent's objective is verified with the failures.
+This hierarchy of agent's objectives is verified with the failures.
 When it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).
@@
 The skewness of the cumulative costs for successful transitions is higher compared to failures.
-On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2}-budget.}
+On top of being the results of the optimisation through EnergyScope, these are influenced by the agent's policy that aims only at lowering the total transition cost as soon as it meets the \ce{CO2} budget.}
 \label{tab:skewness_gwp_cost}
@@
 On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and discount rate.
-These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2}-budget.
+These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2} budget.
 
@@
 
-Besides the cumulative emissions and costs, the agent also observes the share of renewable energy carriers in the primary mix and the efficiency of the system The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
+Besides the cumulative emissions and costs, the agent also observes the share of renewable energy carriers in the primary mix and the efficiency of the system.
+The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
 From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition.
@@
 Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}.
-In 2050, the threshold where occurences of success are more numerous than failures was at 82\% of renewable share.
+In 2050, the threshold where successes occur more often than failures was at 82\% of renewable share.
 In the REF case of Chapter \ref{chap:atom_mol}, this share reached 86\% by 2050.
@@
 As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold as in the SMR case of Chapter \ref{chap:atom_mol}.
-Besides these milestones to respect the \ce{CO2}-budget of the transition, one can also look at the other side of the thresholds.
+Besides these milestones to respect the \ce{CO2} budget of the transition, one can also look at the other side of the thresholds.
 Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except if betting on the future installation of \gls{SMR}.
@@
 \caption{Exploration of the state space over the learning process: distribution of occurrence of share of renewable energy carriers in the primary energy mix (left) and efficiency (right).
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
 Integration of local \gls{VRES} at early stages then massive import of electrofuels later are needed to secure successful transitions.
@@
 
-After investigating the intermediate milestones to meet the \ce{CO2}-budget by 2050, this section details the actions the agent has taken during the learning process (see Figure \ref{fig:Actions_learning}).
+After investigating the intermediate milestones to meet the \ce{CO2} budget by 2050, this section details the actions the agent has taken during the learning process (see Figure \ref{fig:Actions_learning}).
 Rows represent the beginning of the time window at which the set of actions is taken.
 Similarly to the state space, we observe a wide exploration of the action space.
-The more the agent was able to progress through transition, without exceeding the \ce{CO2}-budget, the bigger is the share of successes compared to failures.
+The more the agent was able to progress through transition, without exceeding the \ce{CO2} budget, the bigger is the share of successes compared to failures.
 Besides this observation, no specific range of values for the different actions at the different timing seems to lead to more successes.
 Looking at action individually, there does not seem to be any that supports more effectively the transition.
-The success come from the combination of these actions.
+The success comes from the combination of these actions.
 
@@
 \caption{Exploration of the action space over the learning process: distribution of occurrence of the different actions at the time they are taken by the agent.
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050.
-Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest.
+Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
 Besides this wide exploration, successes and failures indistinguishably spread over the whole ranges of actions.
@@
 The remaining 10\% are consumed by industrial boilers.
-Even though \gls{LFO} represents  30\% of the primary energy mix in 2020, the cost-optimum removes it of the mix without requiring the action of the agent.
+Even though \gls{LFO} represents  30\% of the primary energy mix in 2020, the cost-based model removes it from the mix without requiring the action of the agent.
 Naphtha-crackers, decentralised and industrial boilers get substituted by \gls{MTO}, decentralised \gls{HP} and industrial resistors and \gls{CHP}, respectively.
@@
 Then, as the maximum amount of coal (28\,TWh) is much smaller than fossil gas and \gls{LFO}, high value of $\mathrm{act}\textsubscript{coal}$ still represents small consumption of coal.
+Whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
+Consistency towards banning coal is needed.
 
@@
 Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
-This requires to define the \ce{CO2}-budget according to a certain sharing principle (see Section \ref{sec:cs:CO2-budget}) and to compute reference total transition cost.
-However, other researches focusing on reaching carbon-neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
+This requires to define the \ce{CO2} budget according to a certain sharing principle (see Section \ref{sec:cs:CO2-budget}) and to compute reference total transition cost.
+However, other researches focusing on reaching carbon neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
 
@@
 \label{sec:RL:testing}
-After investigating the exploration of myopic transition, this section compares these results with the perfect foresight under uncertainties from the \gls{UQ} analysis presented in Chapter \ref{chap:atom_mol}.
+After investigating the exploration of myopic transitions, this section compares these results with the perfect foresight under uncertainties from the \gls{GSA} presented in Chapter \ref{chap:atom_mol}.
 To reduce the computational time, the monthly model could also be used to carry out \gls{RL}-based exploration of myopic pathways.
@@
 
-\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2}-budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
-However, driven first by this \ce{CO2}-budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}).
+\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2} budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
+However, driven first by this \ce{CO2} budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}).
 This comes from the agent's actions that limit the emissions and/or the consumption of fossil resources at the early stages.
@@
 This buffer is compensated by the end of the transition where 50\% of the myopic transitions reach 2050 with 10 or more remaining Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
-These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% on average by 2050.
+These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% of the primary mix on average by 2050.
 Finally, the long-term vision of the perfect foresight approach results in smoother reduction of the emissions to end up with less emissions by 2050.
 
-The comparison between the failures and the pathways demonstrates the added-value brought by myopic pathway optimisation.
+The comparison between the failures and the successes demonstrates the added-value brought by myopic pathway optimisation.
 In the near-term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed.
@@
 This is also due to the fact that the perfect foresight approach always finds a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}.
-This explains the wider variability of the results too.
+This explains the wider variability of the PF results too.
 However, with the same sample of uncertain parameters, given the assumed full knowledge over the whole time horizon, PF naturally results in a cheaper transition than its myopic equivalent.
@@
 To add more realism, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon \cite{poncelet2016myopic}.
-To respect a \ce{CO2}-budget over the transition, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
+To respect a \ce{CO2} budget over the transition, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
 However, since the effect of \ce{CO2}-emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself.
@@
 %In the \gls{RL} framework, the agent had four actions to support the 2020-2050 transition: (i) limiting the \ce{CO2}-emissions at the end of each successive time window and limiting the consumption of (ii) fossil gas, (iii) \acrfull{LFO} and (iv) coal.
-The reward fed back by the environment to the agent focused first on meeting the \ce{CO2}-budget, 1.2\,Gt$_{\ce{CO2},\text{eq}}$ equivalent to 10 years at the current level of emissions.
+The reward fed back by the environment to the agent focused first on meeting the \ce{CO2} budget, 1.2\,Gt$_{\ce{CO2},\text{eq}}$ equivalent to 10 years at the current level of emissions.
 A secondary focus is also put on aiming at a minimum transition cost.
@@
 We have compared the results coming from this \gls{RL}-based myopic optimisation with the hourly perfect foresight approach.
-These myopic optimisations provided pathways respecting the \ce{CO2}-budget that were more drastic in cutting emissions in the short-term than the perfect.
+These myopic optimisations provided pathways respecting the \ce{CO2} budget that were more drastic in cutting emissions in the short-term than the perfect.
 Supported by the agent's actions, we could reach cheaper energy system than what the perfect foresight could give.
@@
 
-In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2}-budget target.
+In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2} budget target.
 It also highlighted the actions to take to effectively support a whole-energy system transition.
+This framework allows future research to test other policies on other case studies and identify the key actions and the timing to take them through the transition.

### 2024-07-19 chap_RobPol/chap_RobPol.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 The directions of variation, \ie the robustness metrics, are based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol (see Section \ref{subsec:meth:PCA:transition}).
-These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.
+These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2} budget.
 Data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}).
@@
 Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition.
-In other words, the further the transition goes, the more limited the degrees of freedom are to respect the \ce{CO2}-budget.
+In other words, the further the transition goes, the more limited the degrees of freedom are to respect the \ce{CO2} budget.
 

### 2024-07-19 chap_atom_mol/chap_atom_mol.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 In their work, \citet{lefebvre2022electrofuel} investigated this topic for the case of Belgium considering lower and upper bounds in terms of availability based on, on the one hand, the already signed agreements with the exporting countries and, on the other hand, their maximal technological potential of \gls{VRES}, respectively.
-Their lower bound resulted in a total availability of electrofuels that is one order of magnitude lower than the needs provided by the cost-optimum carbon-neutral Belgium in 2050.
+Their lower bound resulted in a total availability of electrofuels that is one order of magnitude lower than the needs provided by the cost-optimum carbon neutral Belgium in 2050.
 Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} as well as the need of importing electrofuels in the Belgian energy landscape, on a strictly techno-economical point of view with a cost-based optimisation.
@@
 Thanks to the perfect foresight approach, the model knows ahead that cheaper and low-emitting \gls{SMR} will be available in the future.
-As the model can freely spend the constrained \ce{CO2}-budget over the transition, cost-savings also occur before 2040.
+As the model can freely spend the constrained \ce{CO2} budget over the transition, cost-savings also occur before 2040.
 These early-stage cost-savings are equally shared between the extended use of cheaper \gls{LFO} to produce \gls{HVC} and the delayed deployment of \gls{PV}.
@@
 Then, like the power sector, the SMR case ends up in a less efficient whole-energy system by 2050 at it consumes 47\,TWh (+12.7\%) primary energy more to supply an unchanged \gls{EUD}.
-Interestingly, in both cases, given the assumptions made on the \gls{GWP} of the resources (\ie $\mathit{gwp}_{\mathrm{op,electrofuels}}=0$\,kt$_{\ce{CO2},\text{eq}}$/GWh and $\mathit{gwp}_{\mathrm{op,uranium}}=0.004$\,kt$_{\ce{CO2},\text{eq}}$/GWh), the constraint on the \ce{CO2}-budget leads to ``carbon-neutrality'' by 2050\footnote{The model being constrained to keep on using all the waste that would keep on being locally produced, the system in 2050 reaches a $\sim 3.5\,Mt_{\ce{CO2},\text{eq}}$/year.}.
+Interestingly, in both cases, given the assumptions made on the \gls{GWP} of the resources (\ie $\mathit{gwp}_{\mathrm{op,electrofuels}}=0$\,kt$_{\ce{CO2},\text{eq}}$/GWh and $\mathit{gwp}_{\mathrm{op,uranium}}=0.004$\,kt$_{\ce{CO2},\text{eq}}$/GWh), the constraint on the \ce{CO2} budget leads to ``carbon neutrality'' by 2050\footnote{The model being constrained to keep on using all the waste that would keep on being locally produced, the system in 2050 reaches a $\sim 3.5\,Mt_{\ce{CO2},\text{eq}}$/year.}.
 
@@
 The cost of purchasing fossil fuels is also a key parameter in terms of variation of the total transition cost.
-However, due to the ambitious \ce{CO2}-budget, phasing out of fossil fuels is urgent and makes their impact smaller than their renewable alternatives.
+However, due to the ambitious \ce{CO2} budget, phasing out of fossil fuels is urgent and makes their impact smaller than their renewable alternatives.
 
@@
 Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}.
-Subject to the \ce{CO2}-budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity.
+Subject to the \ce{CO2} budget for the transition, the system goes towards more efficient technologies, like industrial methane-\gls{CHP} to substitute e-ammonia-\gls{CCGT} in the production of electricity.
 First running on fossil gas, these \gls{CHP} consume more e-methane by 2050.
 On the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
-This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2}-budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
-In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2}-budget.
+This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2} budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
+In these cases, the use of coal in 2050 highlights that, with a sharper decrease of the emissions at early stages in the transition, the model finds a solution including highly-emitting resources (\eg coal) while respecting the \ce{CO2} budget.
 Consequently, there are smaller investments in methane \gls{CHP}, and consequently import of e-methane as more abundant renewable electricity is produced via e-ammonia-\gls{CCGT} and more HT-heat is supplied by industrial coal boilers.
-Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2}-budget allows it.
+Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2} budget allows it.
 About the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system.
@@
 
-Given the ambitious \ce{CO2}-budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
+Assumed to be carbon neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
 This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
@@
 The perfect foresight approach does not model this risk as the possibility to install or not \gls{SMR} is known in advance.
-In the EnergyScope Pathway model, assuming the same technology mix as in the SMR case up to 2040 but without installing \gls{SMR} for the rest of the transition, \ie between 2040 and 2050, would require drastic (and quick) changes to still meet the \ce{CO2}-budget: around 44\,b€ of additional cumulative OPEX in purchasing electrofuels, mostly e-ammonia (+32\,TWh by 2050) and e-methane (+31\,TWh) to supply \gls{CCGT} and industrial \gls{CHP}.
+In the EnergyScope Pathway model, assuming the same technology mix as in the SMR case up to 2040 but without installing \gls{SMR} for the rest of the transition, \ie between 2040 and 2050, would require drastic (and quick) changes to still meet the \ce{CO2} budget: around 44\,b€ of additional cumulative OPEX in purchasing electrofuels, mostly e-ammonia (+32\,TWh by 2050) and e-methane (+31\,TWh) to supply \gls{CCGT} and industrial \gls{CHP}.
 %

### 2024-07-19 chap_case_study/chap_case_study.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 As detailed by \citet{limpens2024pathway}, the analysis carried out in this work can be applied to any regional whole-energy system.
-As a densely-populated and highly-industrialised country with limited local renewable potentials (\ie mainly solar and wind representing up to 50\% of the primary mix by 2050), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon-neutrality in 2050 makes it an intricate case study.
+As a densely-populated and highly-industrialised country with limited local renewable potentials (\ie mainly solar and wind representing up to 50\% of the primary mix by 2050), the transition of Belgium from a fossil-dominated system in 2020 (Appendix \ref{app:bel_2020}) to carbon neutrality in 2050 makes it an intricate case study.
 Moreover, this case study and the subsequent analyses can be transferred - to some extent - to other industrialised countries highly dependent on fossil fuels with limited local renewable potentials (\eg the Netherlands or Germany) \cite{dommisse2020modelling}.
@@
 Then, the uncertainty ranges considered for some of the parameters are detailed.
-Finally, the \ce{CO2}-budget over the 2020-2050 transition is presented.
+Finally, the \ce{CO2} budget over the 2020-2050 transition is presented.
 
@@
 \label{sec:cs:contributions}
-First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy system integrate energy demands (\ie electricity, heat and mobility), the \acrfull{NED} is often not considered.
+First, as pointed out by \citet{rixhon2022integration}, where most of the studies assessing whole-energy systems integrate energy demands (\ie electricity, heat and mobility), the \acrfull{NED} is often not considered.
 The latter is defined as ‘\textit{energy products used as raw materials in the different sectors; that is not consumed as a fuel or transformed into another fuel}’ \cite{Eurostat2019}.
@@
 This work goes one step further and accounts for the \gls{NED} as a demand for three commodities (\ie \gls{HVC}, ammonia and methanol) as well as the associated production technologies.
-This allows bringing the non-energy sector to a similar level of details as the other sectors.
-Keeping the same methodology to define \gls{EUD} in EnergyScope as \citet{Limpens2020}, this work considers updated values given the latest release of the \og EU reference scenario 2020 : energy, transport and GHG emissions: Trends to 2050 \fg by the European Commission \cite{EuropeanCommission2021}.
+This allows bringing the non-energy sector to a similar level of detail as the other sectors.
+Keeping the same methodology to define \gls{EUD} in EnergyScope as \citet{Limpens2020}, this work considers updated values given the latest release of the \og EU reference scenario 2020: energy, transport and GHG emissions: Trends to 2050 \fg by the European Commission \cite{EuropeanCommission2021}.
 
 Second, given the focus of this work on the electrofuels, the case study includes a more detailed representation of them: where the previous definition of the case study considered only renewable hydrogen and methane \cite{limpens2021generating}, now, e-ammonia and e-methanol (as well as their fossil-based equivalents), are implemented in the case study.
-As detailed later on, these electrofuels are considered as renewable in the sense that their \gls{GWP} is zero.
+As detailed later on, these electrofuels are considered renewable in the sense that their \gls{GWP} is zero.
 This more detailed representation of the molecules themselves also comes with a more exhaustive integration of the ways to produce and use them in the system.
@@
 
-Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition.
+Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2} budget for the transition, \ie limiting the total amount of emissions over the transition.
 Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C above pre-industrial levels by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.
@@
 Finally, this work includes updated values for some parameters compared to the work of \citet{limpens2021generating}.
-The main change concerns the cost and performance of private mobility vehicles, which is a key components in the European \cite{biresselioglu2018electric} and Belgian \cite{BFP_mob} energy transitions.
+The main change concerns the cost and performance of private mobility vehicles, which is a key component in the European \cite{biresselioglu2018electric} and Belgian \cite{BFP_mob} energy transitions.
 Based on the work of the \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
-In previous works \cite{limpens2021generating,rixhon2021role}, despite their lower efficiencies (\ie about 50\% less efficient), \gls{FC} cars had the lion share of the private mobility compared to battery electric vehicles(BEV) given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter.
-In these, the more limited potential to import electricity from abroad and produce it locally via \gls{VRES} forced the model to rather electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.
+In previous works \cite{limpens2021generating,rixhon2021role}, despite their lower efficiencies (\ie about 50\% less efficient), \gls{FC} cars had the bigger share of private mobility compared to battery electric vehicles(BEV) given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter.
+In these works, the more limited potential to import electricity from abroad and to produce it locally via \gls{VRES} forced the model to electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.
 To align with other similar works on the modelling of whole-energy systems \cite{schnidrig2021modelling, EuropeanCommission2021}, the CAPEX and efficiency of fuel cell cars have been increased.
@@
 The \gls{NED} currently represents around 20\% of the final energy consumption in Belgium \cite{FPSEconomy2021}.
-This section summarises the rationale of adding a higher level of details to the \gls{NED} compared to what was done in the previous version of the case study \cite{limpens2021generating}.
+This section summarises the rationale for adding a higher level of detail to the \gls{NED} compared to what was done in the previous version of the case study \cite{limpens2021generating}.
 Then, it explains the methodology used to quantify this demand.\\
@@
 \Gls{HVC} gather the light olefins (e.g.
-ethylene, propylene) and aromatics (benzene, toluene, xylene – BTX), mainly for the production of plastics, synthetic fibers or rubber.
+ethylene, propylene) and aromatics (benzene, toluene, xylene – BTX), mainly for the production of plastics, synthetic fibres or rubber.
 Their production today relies mainly on petroleum products such as naphtha, ethane or liquified petroleum gas.
@@
 Its production is dominated by fossil gas via steam methane reforming to produce hydrogen, used as feedstock in the Haber-Bosch process.
-Methanol is mainly converted to formaldehyde (resin) but also used for the production of other chemicals (e.g.
+Methanol is mainly converted to formaldehyde (resin) but is also used for the production of other chemicals (e.g.
 solvents and gasoline-blends).
@@
 
-Naphtha and \gls{LPG} are consumed in a naphtha cracker, which results in ethylene and propylene, what will be considered as \gls{HVC} in the rest of this work.
-Similarly, fossil gas, as non-energy carrier, is used in steam methane reformers to produce the required hydrogen for the synthesis of ammonia.
+Naphtha and \gls{LPG} are consumed in a naphtha cracker, which results in ethylene and propylene, which will be considered as \gls{HVC} in the rest of this work.
+Similarly, fossil gas, as a non-energy carrier, is used in steam methane reformers to produce the required hydrogen for the synthesis of ammonia.
 The small shares of bitumen and coal tar are used for roadworks and to produce synthetic gas through gasification, respectively.
@@
 Because of the new regulations and strategies promoting recycling and limitation of single-use plastics \cite{EU_plastics}.
-Besides this uncertainty, Belgium stays a major exporter as approximately 2/3 of plastic raw materials produced locally are exported abroad \cite{agoria_plastics}.
+Besides this uncertainty, Belgium remains a major exporter as approximately 2/3 of plastic raw materials produced locally are exported abroad \cite{agoria_plastics}.
 Even if a significant part of \gls{HVC} produced in Belgium is not locally consumed, this demand has been set based on the assumption that Belgium will keep its industrial activity in this sector.
 In other words, we do not deduce the part of local production of \gls{HVC} being then exported (and not consumed locally), unlike ammonia and methanol, which will be more traded commodities in the future (as energy carriers and non-energy products).
-Therefore, the actual demand of \gls{HVC} is inferred from the consumption of naphtha and \gls{LPG} as non-energy use as well as energy-carrier in the chemical and petrochemical industries \cite{statbel_NED_2019}.
-This assumption is based on the fact that, in the conversion processes to produce \gls{HVC} from naphtha or \gls{LPG}, these fuels also serve as energy-carrier to supply the processes themselves.
+Therefore, the actual demand for \gls{HVC} is inferred from the consumption of naphtha and \gls{LPG} as non-energy use as well as energy carriers in the chemical and petrochemical industries \cite{statbel_NED_2019}.
+This assumption is based on the fact that, in the conversion processes to produce \gls{HVC} from naphtha or \gls{LPG}, these fuels also serve as energy carrier to supply the processes themselves.
 Then, given the respective efficiencies (1.83t$_{\text{naphtha}}$/t$_{\text{HVC}}$ and 1.67t$_{\text{LPG}}$/t$_{\text{HVC}}$) \cite{IEA2018_petrochemicals}, the current yearly demand of \gls{HVC} is estimated equal to 3069\,kt, without making distinctions between the different chemicals (\ie ethylene, propylene and \gls{BTX}).
@@
 
-Concerning the demand of methanol, similarly to ammonia, this work solely considers the net imports as there is no local production in Belgium.
-To define the actual non-energy demand of methanol, only a 51\%-share of this net import is kept since, according to the Methanol Institute and \gls{MMSA}, this share is used for formaldehyde production in Belgium \cite{MMSA51}.
+Concerning the demand for methanol, similarly to ammonia, this work solely considers the net imports as there is no local production in Belgium.
+To define the actual non-energy demand for methanol, only a 51\%-share of this net import is kept since, according to the Methanol Institute and \gls{MMSA}, this share is used for formaldehyde production in Belgium \cite{MMSA51}.
 Currently, the rest of the methanol is used for energy purposes, mostly as \gls{MTBE} in gasoline blending.
@@
 Given this discrepancy that is unsubstantiated and specific to the case of Belgium, the evolution trend of the \gls{NED} of the current work has been inferred from the previous edition, published in 2016, \cite{EuropeanCommission2016}.
-Between 2020 and 2050, one observes a noteworthy increase of the electricity (+40\%), passenger (+45\%) and freight mobility (+35\%) demands (see \Cref{fig:cs_demands}).
+Between 2020 and 2050, one observes a noteworthy increase in the electricity (+40\%), passenger (+45\%) and freight mobility (+35\%) demands (see \Cref{fig:cs_demands}).
 The rise of the non-energy demand is more limited, \ie +6\%, whereas the heating demands is forecast to decrease: -11\% for the low-temperature heat demand and -3\% for the high-temperature heat demand.
 This is explained by a better insulation of buildings and an improved efficiency of industrial processes.
-Regarding the top-right graph of \Cref{fig:cs_demands}, it is the aggregation of the same data as in the top-left graph but per economic sector, rather than per energy sector, with the non-energy demand being associated with the industry.
+The top-right graph of \Cref{fig:cs_demands} is the aggregation of the same data as in the top-left graph but per economic sector, rather than per energy sector, with the non-energy demand being associated with the industry.
 This illustrates how industrialised Belgium is, compared to households and services, and, consequently, highly energy-intensive.
 The bottom graph of \Cref{fig:cs_demands} gives the passenger and the freight mobility.
-The sharp increase from 2020 to 2025 is due the COVID-crisis that led to significantly reduced demands in 2020.
+The sharp increase from 2020 to 2025 is due to the COVID-crisis that led to significantly reduced demands in 2020.
 As far as the hourly discretisation of these demands is concerned, time series are based on historical values of 2015 for the fluctuating parts of electricity and low-temperature heating demands \cite{Limpens2020}.
-A daily time series is used for the passenger mobility and applied similarly to every days of the year.
-Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and the share of electricity and low-temperature heat demands that are considered as constant over the year), the yearly demand is distributed uniformly over the different hours of the year.
+A daily time series is used for passenger mobility and applied similarly to every day of the year.
+Finally, for the other demands (\ie high-temperature heat, freight mobility, \gls{NED} and the share of electricity and low-temperature heat demands that are considered constant over the year), the yearly demand is distributed uniformly over the different hours of the year.
 
@@
 Similarly to \citet{limpens2021generating}, we constrained the endogenous non-renewable wastes to be consumed locally.
-The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh.
+The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} limiting this potential to about 40\,TWh.
 On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of rooftop \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
@@
 The imported energy commodities are available ``at the front door'' of the system.
-In terms of spatial resolution of the country itself, it is modelled as a single-node.
+In terms of spatial resolution of the country itself, it is modelled as a single node.
 This is similar to the copper plate model of a power grid.
@@
 \label{sec:cs:technologies}
-As the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, for instance, technologies are implemented to convert these resources into the end-use demands.
+As the end-use demands are defined as energy (and non-energy with the \gls{NED}) services rather than a certain quantity of oil or solar irradiance, for instance, technologies are implemented to convert these resources into end-use demands.
 Besides their CAPEX, OPEX and lifetime defined in Section \ref{sec:meth:ES}, production and conversion technologies (\ie \gls{CCGT}, car or boiler) have a conversion efficiency whereas storage technologies (\ie thermal storage, battery or molecule storage) have charge/discharge losses.
@@
 
-Not digging into too much more details about the exhaustive list of these technologies presented in a previous work \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} and the technologies to supply the \glsxtrfull{NED}.
+Not digging into too many more details about the exhaustive list of these technologies presented in previous work \cite{limpens2021generating}, this section rather focuses on the implementation of \glsxtrfull{SMR} and the technologies to supply the \glsxtrfull{NED}.
 
@@
 
-A specific attention is put on the implementation of \gls{SMR} as the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
-Similarly to the analysis of \citet{PATHS2050}, a Belgian consortium for energy research, and in line with the Belgian Nuclear Research Centre (SCK-CEN) \cite{SCK-CEN_SMR}, \gls{SMR} are implemented with the features listed in \Cref{tab:SMR_features}.
+Specific attention is put on the implementation of \gls{SMR} as the 6\,GW of conventional nuclear are assumed to drop to 2\,GW in 2025 and total phase-out by 2035.
+Similarly to the analysis of \citet{PATHS2050}, a Belgian consortium for energy research, and in line with the Belgian Nuclear Research Centre (SCK-CEN) \cite{SCK-CEN_SMR}, \gls{SMR} is implemented with the features listed in \Cref{tab:SMR_features}.
 Where most of the features are similar to conventional nuclear power plants, they differ from these on two main points: their potential year start, 2040, and their flexibility.
 Indeed, unlike the current nuclear power plants, constrained in the model to produce a constant power output at every hour of the year (\ie baseload production as it is actually the case in Belgium), SMRs, are flexible in the sense that their production can vary between 0 and their full capacity at any hour of each representative year.
-Here, we simplify SMRs as only producing electricity and disregard the heat produced by the nuclear reaction.
-This is considered as lost to the atmosphere.
+Here, we simplify SMRs as only producing electricity and we disregard the heat produced by the nuclear reaction.
+This is considered lost to the atmosphere.
 
@@
 
-For the sake of comparison, the \gls{LCOE} of the principal technologies to produce electricity, based on the computation used by \citet{limpens2021generating}, are detailed (see \Cref{fig:LCOE}).
-Not including here the cost of integrating a technology in the system (\eg reinforcement of the grid and storage capacities for \gls{VRES}), the \gls{LCOE} aims at aggregating and normalizing the CAPEX and OPEX of technologies providing a common commodity, \ie electricity.
+For the sake of comparison, the \gls{LCOE} of the principal technologies to produce electricity, based on the computation used by \citet{limpens2021generating}, is detailed (see \Cref{fig:LCOE}).
+Not included here is the cost of integrating a technology in the system (\eg reinforcement of the grid and storage capacities for \gls{VRES}), the \gls{LCOE} aims at aggregating and normalizing the CAPEX and OPEX of technologies providing a common commodity, \ie electricity.
 Compared to the other flexible generation units, \gls{SMR} is significantly more cost-effective.
@@
 In addition, one sees that \gls{CCGT} supplied by e-ammonia outcompetes its e-methane equivalent, unlike their respective fossil-based equivalent.
-This due to the fact that e-ammonia, not requiring carbon capture, is expected to be more cost-effective to produce versus e-methane \cite{h2coalition}.
+This is because e-ammonia, not requiring carbon capture, is expected to be more cost-effective to produce versus e-methane \cite{h2coalition}.
 On the contrary, fossil-based ammonia, mostly relying on steam methane reforming, requires additional steps in the production process compared to fossil gas, as introduced in Section \ref{subsec:cs:NED}.
@@
 \caption{Levelised cost of energy (LCOE) for the main technologies in the power sector.
-Gray and black curves are related to technologies runnning on renewable and fossil resources, respectively.
+Grey and black curves are related to technologies running on renewable and fossil resources, respectively.
 \gls{SMR} is cheaper than the other flexible options.
@@
 Different paths are implemented to produce the final molecules of the NED (see \Cref{fig:NED_tech}).
-Similarly to \citet{tsiropoulos2018emerging}, naphtha, here considered as \gls{LFO}, resulting from refinery operation is modeled as an imported commodity.
+Similarly to \citet{tsiropoulos2018emerging}, naphtha, here considered as \gls{LFO}, resulting from refinery operation is modelled as an imported commodity.
 Presented here for the specific year of 2035, all data and related references can be found in \cite{GIT_NED}.
-To keep the same level of details with other sectors of the model, the implementation of the conversion technologies consists of a single kind of technology per type of resource to produce a certain product.
+To keep the same level of detail as other sectors of the model, the implementation of the conversion technologies consists of a single kind of technology per type of resource to produce a certain product.
 For instance, in the model, there is only one technology to produce \gls{HVC} either from naphtha or from LPG, two liquid fossil hydrocarbons, \ie \gls{NSC}.
@@
 %\noindent
-Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium, using EnergyScope TD (see Appendix \ref{app:ESTD}).
-This snapshot model, \ie optimising the system in a target future year (\ie 2050 in this case) considering a green field approach, investigated the defossilisation of the system.
+Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium in 2050, using EnergyScope TD (see Appendix \ref{app:ESTD}).
 To analyse the whole-energy system at different ``climate targets'', the model forced the total emissions to decrease by reducing their upper limit while optimising the total cost.
 In practice, 10\% steps of \gls{GWP} reduction were made from the ``reference scenario - 100\%''.
-This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon-neutrality).
+This strategy gave the following points of analysis: 100\% (\ie cost-optimum with no limitation on the total \gls{GWP}), 90\%, 80\%, ..., down to 0\% (\ie  carbon neutrality).
 
-First and foremost, when the \gls{NED} is implemented with this higher level of details, we highlighted that woody biomass was ``cannibalised'' to produce methanol, instead of high-temperature heat, in the cost-optimum situation.
+First and foremost, when the \gls{NED} is implemented with this higher level of detail, we highlighted that woody biomass was ``cannibalised'' to produce methanol, instead of high-temperature heat, in the cost-optimum situation.
 At more ambitious ``climate targets'', e-methanol rises as the keystone to defossilise the \gls{NED} sector as \gls{MTO} becomes the favoured option to produce \gls{HVC}, representing the major share of the \gls{NED}.
-Then, including the \gls{NED} affects the selection of the technologies for the satisfaction of the heat and the electricity demands.
+Then, including the \gls{NED} affects the selection of the technologies for the satisfaction of the heat and electricity demands.
 The additional high-temperature demand required by the \gls{NED} forces the system to invest in more efficient technologies like \gls{CHP} instead of \gls{CCGT}.
 This additional heat demand mostly supplies naphtha-cracking substituted by \gls{MTO} at more ambitious ``climate targets'' to produce \gls{HVC}.
-To a lesser extent, to respect the emissions-caps, integrating the \gls{NED} leads to a higher integration of solar-\gls{PV} to support the electrification of the low-temperature heat sector.
+To a lesser extent, to respect the emissions caps, integrating the \gls{NED} leads to a higher integration of solar \gls{PV} to support the electrification of the low-temperature heat sector.
 
@@
 \label{sec:cs:uncertainty}
-As detailed in Section \ref{sec:meth:UQ}, accounting for uncertainty in \gls{ESOMs} is crucial \cite{mavromatidis2018uncertainty}, especially when it comes to optimise several decades in an inherently uncertain future.
+As detailed in Section \ref{sec:meth:UQ}, accounting for uncertainty in \gls{ESOMs} is crucial \cite{mavromatidis2018uncertainty}, especially when it comes to optimising several decades in an inherently uncertain future.
 The fundamental step in this ambition is to characterise these uncertainties.
@@
 
-This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the exogenous modal share change in different key sectors.
+This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for the variability of solar irradiance or wind speed; (vii) the availability of resources considered limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the exogenous modal share change in different key sectors.
 A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.
@@
 Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
-Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% of chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
 }.
@@
 
-\section{\ce{CO2}-budget for the transition}
+\section{\ce{CO2} budget for the transition}
 \label{sec:cs:CO2-budget}
-In most studies carried out on the pathway optimisation of a whole-energy system, a \ce{CO2}-trajectory is \textit{a priori} set to reach carbon-neutrality by 2050.
+In most studies carried out on the pathway optimisation of a whole-energy system, a \ce{CO2}-trajectory is \textit{a priori} set to reach carbon neutrality by 2050.
 \citet{nerini2017myopic} used the emission trajectory indicated by the UK's Committee on Climate Change in their analysis of the impact of limited foresight to achieve the target of 80\% reduction of \gls{GHG} by 2050 in the United Kingdom.
-In their assessment of the impacts of economy-wide emission policies in the water-energy-land nexus, \citet{licandeo2023assessing} analysed different \ce{CO2}-trajectories considering more or less severe water scarcity for the US.
+In their assessment of the impacts of economy-wide emission policies in the water-energy-land nexus, \citet{licandeo2023assessing} analysed different \ce{CO2} trajectories considering more or less severe water scarcity for the US.
 \citet{poncelet2016myopic} with LUSYM (Leuven University SYstem Model) and \citet{PATHS2050} with TIMES-BE also set decreasing emission trajectories in their analysis of respectively the Belgian power sector and whole-energy system.
-Others only set the objective as the carbon-neutrality by 2050.
+Others only set the objective of carbon neutrality by 2050.
 For instance, \citet{heuberger2018impact} investigated the impact of different factors (\eg limit of the foresight in the future, availability of \og unicorn technologies\fg or committed versus market-driven decarbonisation strategies) to reach this ultimate objective in the UK system.
 In their ``near-term to net zero'' (NT2NZ) approach to estimate \ce{CO2} prices, \citet{kaufman2020near} emphasises the need to select, \textit{a priori}, an emissions pathway to the net-zero target.
-Some authors suggest limiting near-term technological disruptions and, consequently, the initial rate of emission reductions \cite{wigley1996economic} where others, to avoid technology lock-in and to benefit from early investments in the transition, encourage a sharper decrease of the emissions at an earlier stage \cite{vogt2018starting}.
+Some authors suggest limiting near-term technological disruptions and, consequently, the initial rate of emission reductions \cite{wigley1996economic} whereas others, to avoid technology lock-in and to benefit from early investments in the transition, encourage a sharper decrease of the emissions at an earlier stage \cite{vogt2018starting}.
 
-In this work, the effect of greenhouse gases is cumulative over time and a constraint is set on the overall emissions of the transition---a \ce{CO2}-budget for the transition.
-This approach is in line with the works defining safe operating spaces within the nine different global planetary boundaries (\ie (i) novel entities, (ii) stratospheric ozone depletion, (iii) atmospheric aerosol loading, (iv) ocean acidification, (v) biogeochemical flows, (vi) fresh water change, (vii) land system change, (viii) biosphere integrity, and, (ix) climate change) \cite{richardson2023earth,steffen2015planetary,rockstrom2009safe}.
+In this work, the effect of greenhouse gases is cumulative over time and a constraint is set on the overall emissions of the transition---a \ce{CO2} budget for the transition.
+This approach is in line with the works defining safe operating spaces within the nine different global planetary boundaries (\ie (i) novel entities, (ii) stratospheric ozone depletion, (iii) atmospheric aerosol loading, (iv) ocean acidification, (v) biogeochemical flows, (vi) freshwater change, (vii) land system change, (viii) biosphere integrity, and, (ix) climate change) \cite{richardson2023earth,steffen2015planetary,rockstrom2009safe}.
 This ``systemic framework for addressing global anthropogenic impacts on Earth system'' gives quantitative recommendations about the \ce{CO2} concentration, among others, to maintain ``the stability and resilience of Earth system as a whole'' \cite{richardson2023earth}.
@@
 Finally, in \textit{acquired rights} principles, also called ``grandfathering'', the sharing is based on \og maintaining that prior emissions increase future emission entitlements\fg  \cite{knight2013grandfathering}.
-In this thesis, we have chosen the latter principle to allocate the \ce{CO2}-budget to the Belgian transition.
+In this thesis, we have chosen the latter principle to allocate the \ce{CO2} budget to the Belgian transition.
 This budget (1.2\,Gt$_{\ce{CO2},\text{eq}}$) corresponds to the proportion of Belgium's emissions in the world energy-related emissions in 2020 (34.8\,Gt$_{\ce{CO2},\text{eq}}$ \cite{ourworldindata_CO2_world}) applied to the global budget to have a 66\% chance of limiting warming to 1.5°C of 420\,Gt$_{\ce{CO2},\text{eq}}$ \cite{IPCC_CO2_budget}.
@@
 In this approach, the residual emissions in 2050 are not explicitly constrained down to zero.
-This carbon-neutrality is rather implicit given the ambitious \ce{CO2}-budget.
+This carbon neutrality is rather implicit given the ambitious \ce{CO2} budget.
 
 Compared to a linear decrease from the current emissions, as done by \citet{limpens2024pathway}, this budget represents a 60\% reduction of the cumulative emissions over the transition.
-Appendix \ref{app:CO2_budget} compares the emissions trajectory between the REF case and a case (without \gls{SMR}) where the linear decrease is imposed between 2020 and carbon-neutrality in 2050.
+Appendix \ref{app:CO2_budget} compares the emissions trajectory between the REF case and a case (without \gls{SMR}) where the linear decrease is imposed between 2020 and carbon neutrality in 2050.
 
@@
 Mainly based on data collected during previous works \cite{Moret2017PhDThesis,limpens2021generating}, this thesis has developed the case study along different axes.
-First, the non-energy demand, that currently represents about 20\% of the final energy consumed in Belgium, has been brought to a level of details similar to the other energy sectors of the system.
+First, the non-energy demand, which currently represents about 20\% of the final energy consumed in Belgium, has been brought to a level of detail similar to the other energy sectors of the system.
 Previously considered as a fixed demand supplied by fossil gas, it has been detailed to account for final-use products (\ie \acrfull{HVC}, ammonia and methanol) and their respective production routes.
 Second, electrofuels are now more detailed in the model as well as their different supply and consumption routes.
-Third, instead of defining \textit{a priori} the \ce{CO2}-trajectory through the transition, this work sets a limit on the overall \ce{CO2}-budget.
+Third, instead of defining \textit{a priori} the \ce{CO2}-trajectory through the transition, this work sets a limit on the overall \ce{CO2} budget.
 Then, based on the methodology of \citet{Moret2017}, uncertainty ranges are defined for the 5-year step transition from 2020 to 2050.
-Finally, on top of adding the possibility to install small modular reactors (SMR), this thesis accounts for updated values for the \gls{EUD} as well as adjusted assumptions regarding the two key drivers of the transition in the private mobility, \ie battery electric vehicles (BEV) and \acrfull{FC} cars.
+Finally, on top of adding the possibility to install small modular reactors (SMR), this thesis accounts for updated values for the \gls{EUD} as well as adjusted assumptions regarding the two key drivers of the transition in private mobility, \ie battery electric vehicles (BEV) and \acrfull{FC} cars.

### 2024-07-19 chap_intro_ccl/conclusion.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 % WHAT WE DID (HOW?)
-The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2}-budget without prescribed trajectory for the reduction of the emissions.
+The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2} budget without prescribed trajectory for the reduction of the emissions.
 We have studied the case study of Belgium with a focus on the role that renewable electrofuels (e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely-populated and highly industrialised country with limited local renewable potentials.
-To have 66\% chance of limiting warming to 1.5°C and using the ``grandfathering'' sharing principle of the global budget, the \ce{CO2}-budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
+To have 66\% chance of limiting warming to 1.5°C and using the ``grandfathering'' sharing principle of the global budget, the \ce{CO2} budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
 
@@
 
-The \acrfull{RL} approach was adopted to explore the myopic pathways respecting a \ce{CO2}-budget equivalent to 10 years of current emissions, without \ce{CO2}-trajectory.
+The \acrfull{RL} approach was adopted to explore the myopic pathways respecting a \ce{CO2} budget equivalent to 10 years of current emissions, without \ce{CO2}-trajectory.
 In this framework, an agent takes actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
-Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050 where a transition was considered as successful when respecting the \ce{CO2}-budget.
+Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050 where a transition was considered as successful when respecting the \ce{CO2} budget.
 Repeating the whole transition with different sequences of actions-states allowed the agent coming up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
@@
 % THE RESULTS WE GOT (WHAT?)
-Given the \ce{CO2}-budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  aiming at a cost-optimum transition, Belgium should first opt for more efficiency to reduce by one third the primary energy consumption.
+Given the \ce{CO2} budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  aiming at a cost-optimum transition, Belgium should first opt for more efficiency to reduce by one third the primary energy consumption.
 Then comes the full deployment of local solar and wind energies that would represent about 30\% of the mix by 2050.
@@
 The Reinforcement Learning-based exploration of myopic transitions points out that near-term actions are mandatory to hope succeeding this ambitious transition.
-Beyond 2030, if the right actions are not taken, the cumulative emissions are likely to overshoot the \ce{CO2}-budget.
+Beyond 2030, if the right actions are not taken, the cumulative emissions are likely to overshoot the \ce{CO2} budget.
 On top of strongly reducing the use of coal, fossil gas consumption should also be limited within 5 to 10 years.
@@
 Often brought up as a ``unicorn'' solution to the energy transition, nuclear \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but not against the deployment of local \acrfull{VRES} such as solar and wind.
-This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2}-budget of the transition.
+This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2} budget of the transition.
 
@@
 This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
-This could also show the limits of the Belgian energy system to meet such an ambitious \ce{CO2}-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
+This could also show the limits of the Belgian energy system to meet such an ambitious \ce{CO2} budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
 One could then question the level of demands and address the question of the sufficiency.
 For a growing part of the scientific community, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
-A study could consist in analysing the level of demands, within their respective sector, we should reach to stay within the \ce{CO2}-budget and limit the installation of solar PV, wind turbines, nuclear energy or the import of electrofuels.
+A study could consist in analysing the level of demands, within their respective sector, we should reach to stay within the \ce{CO2} budget and limit the installation of solar PV, wind turbines, nuclear energy or the import of electrofuels.
 
@@
 Depending on the level of emission reduction, these indirect emissions could be one to three times higher than the direct emissions.
-To accurately integrate these emissions in the model and account for them in the \ce{CO2}-budget, \gls{LCA} can compute these indirect emissions.
+To accurately integrate these emissions in the model and account for them in the \ce{CO2} budget, \gls{LCA} can compute these indirect emissions.
 On top of this, \gls{LCA} could provide data to assess other impact indicators besides climate change \cite{astudillo2018integrating}.

### 2024-07-19 chap_intro_ccl/intro.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 of anthropogenic \gls{GHG} in the atmosphere \cite{IPCC_CO2_budget}.
-Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damages.
+Therefore, the \gls{GHG} emissions from human activities must be mitigated to prevent further environmental damage.
 Energy consumption, mainly from fossil fuels, is the main contributor: it is responsible for 75\% of \gls{GHG} emissions, the remaining 25\% originating from agriculture and industrial processes \cite{ourworldindata_CO2_world}.
@@
 \gls{EUD} is the energy service required by the final consumer.
-In Equation (\ref{eq:equality_GHG}), the first factor on the right hand side represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
+In Equation (\ref{eq:equality_GHG}), the first factor on the right-hand side represents the \gls{GWP} of the primary energy mix, the second is the inverse of the efficiency and the third is the energy intensity per capita.
 
@@
 Such an identity is criticized for the arbitrary choice of variables, the non-independence of them usually leading to the rebound effect and its encompassing approach that does not appropriately mirror the heterogeneity of the situation \cite{IPCC2000}.
-Considering the expected growing trend of the Population, Equation (\ref{eq:equality_GHG}) highlights three levers of action that should be activated to reduce the \gls{GHG} emissions and favour the energy transition \cite{dodson2020population,scovronick2017impact}: renewables, efficiency and sufficiency.
+Considering the expected growing trend of the Population, Equation (\ref{eq:equality_GHG}) highlights three levers of action that should be activated to reduce \gls{GHG} emissions and favour the energy transition \cite{dodson2020population,scovronick2017impact}: renewables, efficiency and sufficiency.
 
-Focusing on the levers that are within grasp of engineering studies, this thesis addresses the two ``technical'' levers of the energy transition identified in Equation (\ref{eq:equality_GHG}): renewables and efficiency.
+Focusing on the levers that are within the grasp of engineering studies, this thesis addresses the two ``technical'' levers of the energy transition identified in Equation (\ref{eq:equality_GHG}): renewables and efficiency.
 This focus is aligned with the current European policies binding the Member States of the European Union.
@@
 Despite the growing echo it finds in the scientific community \cite{o2018good}, sufficiency is lacking in the RED III.
-Explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, it is defined as ``\emph{a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring an adequate social foundation for all people}'' \cite{lage2023citizens}.
+Explicitly mentioned by the IPCC for the first time in 2022 \cite{IPCC2022}, it is defined as ``\emph{a strategy for reducing, in absolute terms, the consumption and production of end-use products and services through changes in social practices in order to comply with environmental sustainability while ensuring adequate social foundation for all people}'' \cite{lage2023citizens}.
 Due to its intrinsically political nature, studying sufficiency requires an interdisciplinary approach involving economy, sociology, psychology and political science, and is beyond the scope and knowledge of engineering research \cite{schmidt2015interdisciplinary}.
@@
 \noindent
-To ensure the energy supply of a growing and more demanding population in a context of environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
+To ensure the energy supply of a growing and more demanding population in the context of an environmental crisis, major transformations are needed (see Figure \ref{fig:intro:IEA_WEO_2019}).
 Similarly to the RED III, the transformations identified by the \citet{iea2020world} focus on technical aspects, disregarding the sufficiency.
@@
 However, direct electrification is not foreseen to be cost-competitive for some sectors such as marine, aviation and heavy-duty transport \cite{horvath2018techno, brynolf2018,IEA2021}.
-Moreover, transport and long-term storage of the renewable electricity is not optimal with electricity-based solutions, \eg {DC} lines or batteries.
-This is why, to reach the objective of sustainable development, after increasing the efficiency of the system, more renewables is associated with ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}).
+Moreover, transport and long-term storage of renewable electricity are not optimal with electricity-based solutions, \eg {DC} lines or batteries.
+This is why, to reach the objective of sustainable development, after increasing the efficiency of the system, more renewables is associated with a ``fuel switch'' (see Figure \ref{fig:intro:IEA_WEO_2019}).
 
@@
 
-Electrofuels also have the ability to couple energy and non-energy sectors \cite{Stancin2020}.
+Electrofuels can also couple energy and non-energy sectors \cite{Stancin2020}.
 For instance, electricity from \gls{VRES} can be converted into ammonia through the Haber-Bosch process and subsequently transformed into fertiliser, coupling the power and agro-industrial sectors \cite{verleysen2020can}.
 
-Concerning storage,  batteries exhibit limited storage capacity (up to 10\,MWh) as well as self-discharge losses.
+Concerning storage,  batteries exhibit limited storage capacity (up to 10\,MWh) and self-discharge losses.
 On the contrary, electrofuels are promising solutions for high capacity (from 100 GWh) and long-term storage of energy \cite{child2018role, dias2020energy} (Figure \ref{fig:intro:Storage_electrofuels}).
@@
 In other words, hydrocarbons, currently produced from fossil resources, will still be composed of carbon in a renewable world.
-This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in the energy transition that aims at carbon-neutrality where there is no additional, if no removal, of \gls{GHG} emissions in the atmosphere \cite{mertens2020carbon}.
+This is why this thesis rather uses ``defossilisation'' rather than ``decarbonisation'' as carbon will still play a key role in the energy transition that aims at carbon neutrality where there is no additional, if no removal, of \gls{GHG} emissions in the atmosphere \cite{mertens2020carbon}.
 
@@
 To avoid the confusion between all these ``alternative'' fuels and thereby reduce misunderstanding in political or academic discussions, we have developed a comprehensive and harmonised taxonomy (see Appendix \ref{app:Taxonomy}).
-In the rest of this thesis, the electro --- and bio --- fuels, imported from the exterior of the considered energy system, are considered as renewable with no associated \gls{GWP}.
+In the rest of this thesis, the electro --- and bio --- fuels, imported from the exterior of the considered energy system, are regarded as renewable with no associated \gls{GWP}.
 
-To harvest the maximum potential of synthetic fuels in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a whole-energy system which is multi-sector and account for multiple energy carriers \cite{contino2020whole}.
+To harvest the maximum potential of synthetic fuels in a sustainable transition and maximise the overall system efficiency \cite{mathiesen2015}, it is necessary to study the integration of these fuels within a whole-energy system which is multi-sector and accounts for multiple energy carriers \cite{contino2020whole}.
 To reach this goal, an energy system optimisation model (ESOM) can optimise the design and the operation of the system to minimise, for instance, its costs or its emissions \cite{zeng2011review}.
@@
 In the field of energy system planning and scenario analysis, \citet{yue2018review} highlighted that most ESOMs use a deterministic approach (75\% of the 134 reviewed studies).
-However, the models and their numerous composing parameters are inherently uncertain especially when it comes to define an energy transition strategy for a large-scale system such as a country.
+However, the models and their numerous composing parameters are inherently uncertain especially when it comes to defining an energy transition strategy for a large-scale system such as a country.
 Given the lifetime of energy conversion technologies, such a strategy implies decisions with long-term impacts (20 to 50 years) where forecasts can be highly unreliable \cite{Moret2017}.
@@
 On the contrary, they uncover the realisation of these uncertainties step-by-step, in a ``myopic'' way, and progressively act on them to, hopefully, meet the set target to reduce the anthropogenic \gls{GHG}.
-In the objective to respect an overall \ce{CO2}-budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there is a need for a framework to explore these multiple transition pathways and provide insight about intermediate milestones not to miss.
+In the objective to respect an overall \ce{CO2} budget rather than to follow a prescribed \ce{CO2}-emissions trajectory, there is a need for a framework to explore these multiple transition pathways and provide insight into intermediate milestones not to miss.
 On top of the ``what to do?'', this framework should aim at helping the policymakers to answer the  question ``how to do it?''.
@@
 Given the complexity of whole-energy system models, the time scale of the transition and the large number of uncertainties, only considering the total cost does not give information on the sensitivity of the design strategy, \ie the investment decisions.
-For instance, total cost might consist in excessive upfront investments and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable despite a slightly higher total cost.
+For instance, total cost might consist of excessive upfront investments and a better balance of \gls{CAPEX} and \gls{OPEX} might be preferable despite a slightly higher total cost.
 \citet{moret2020overcapacity} proposed a method to assess the robustness of a design by investigating the potential overcapacity needed to face uncertainties.
-However, their work focused on the power sector only and for the target future year of 2035.
+However, their work focused on the power sector only and on the target future year of 2035.
 Instead, an objective of this thesis is to tackle these two challenges: go beyond the total transition cost and provide more details on the sensitivity of the design to uncertainties while assessing a whole-energy system over its whole transition.\\
@@
 ``\textit{Our task is not to foresee the future, but to enable it}''\footnote{Saint-Exupéry in Citadelle, 1948}.
-Rather than trying to answer the question ``What could possibly happen in the future?'', this work addresses the ``What could or should we do to make the future possible?''.
+Rather than trying to answer the question ``What could happen in the future?'', this work addresses the ``What could or should we do to make the future possible?''.
 This thesis aims at providing decision-makers with information and new methods considering the intrinsic uncertainties of the future.
@@
 \begin{itemize}
-\item What can be the role of electrofuels in the transition pathways of a whole-energy system considering uncertainties,  a limited foresight, and a given \ce{CO2}-budget?
+\item What can be the role of electrofuels in the transition pathways of a whole-energy system considering uncertainties,  limited foresight, and a given \ce{CO2} budget?
 What are the key uncertainties driving their local production or import?
@@
 First introduced by Limpens in his PhD thesis \cite{limpens2021generating}, this model optimises the design and the operation of a whole-energy system over several decades and accounts for transition pathways from an existing system to a long-term target, \ie 2050.
-Based on this model originally implemented for a perfect foresight approach, we have developed the myopic method in which the whole time horizon (30 years) is optimised through a sequence of 10-year time-windows with 5-year overlaps.
+Based on this model originally implemented for a perfect foresight approach, we have developed the myopic method in which the whole time horizon (30 years) is optimised through a sequence of 10-year time windows with 5-year overlaps.
 
-To address the question of the role of electrofuels, they have been integrated in the pre-existing model, focusing on four molecules: hydrogen, methane, ammonia and methanol.
+To address the question of the role of electrofuels, they have been integrated into the pre-existing model, focusing on four molecules: hydrogen, methane, ammonia and methanol.
 Moreover, given their current and (expected) future role in the sector of the \gls{NED}, we have added this sector to the model with a similar level of detail as for the other sectors of the system, \ie electricity, heat and transport.
@@
 This approach is based on \gls{PCA} where directions capturing the widest design variations are identified and serve as a reference frame.
-Roadmaps resulting from perfect foresight optimisation have been tested in a myopic and uncertain pathways.
-The results of these myopic runs have then been ``projected'' on the aforementioned frame to be able to compare the robustness of different roadmaps between each other.
+Roadmaps resulting from perfect foresight optimisation have been tested in myopic and uncertain pathways.
+The results of these myopic runs have then been ``projected'' on the aforementioned frame to be able to compare the robustness of different roadmaps with each other.
 
 All this work has been applied to the case of Belgium, a densely populated country with a limited potential of renewable energies, which represents roughly half of the forecast energy demand \cite{Limpens2020}.
-This makes it a challenging case to go from a highly fossil-dominated system in 2020 (73\% of the primary energy mix \cite{spf_economy_2022}) to carbon-neutrality by 2050.\\
+This makes it a challenging case to go from a highly fossil-dominated system in 2020 (73\% of the primary energy mix \cite{spf_economy_2022}) to carbon neutrality by 2050.\\
 
@@
 Chapter \ref{chap:case_study} presents the case study of this work: Belgium and its energy transition.
-Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2}-budget rather than a prescribed emissions-trajectory and the data related to \gls{SMR} as an option to produce nuclear-based electricity in the future.
+Without exhaustively detailing the data used from the work of \citet{limpens2021generating}, this chapter focuses on the main contributions of this work regarding the case study: the \gls{NED}, the implementation of electrofuels and their respective routes of production and consumption, limiting  the cumulative emissions of the transition to a certain \ce{CO2} budget rather than a prescribed emissions-trajectory and the data related to \gls{SMR} as an option to produce nuclear-based electricity in the future.
 
@@
 In Chapter \ref{chap:chap_RL}, we detail the rules of the ``\gls{RL} game''.
-In other words, we define the action and state spaces as well as the reward function driving the behaviour of the agent in its quest to optimise its policy.
+In other words, we define the action and state spaces as well as the reward function driving the agent's behaviour in its quest to optimise its policy.
 Then, we analyse the results of the agent's learning process and compare these results with the perfect foresight approach.
-Chapter \ref{chap:chap_RL} focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2}-budget.
+Chapter \ref{chap:chap_RL} focuses on the robustness of a policy as its ability to maximise the chances to succeed a myopic transition under uncertainties to meet a \ce{CO2} budget.
 
@@
 
-Finally, we draw the main conclusions of this work: near-term actions are needed to respect ambitious \ce{CO2}-budget and investing into electrofuels provides robustness to uncertainties and the limited foresight into the future.
+Finally, we draw the main conclusions from this work: near-term actions are needed to respect the ambitious \ce{CO2} budget and investing in electrofuels provides robustness to uncertainties and limited foresight into the future.
 Perspectives are suggested for future research on further developments and uses of the methodological tools.

### 2024-07-19 chap_methodo/chap_methodo.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2

@@
 This model requires some characteristics to capture the peculiarities of this system such as the intermittency of \gls{VRES}, the coupling between the different energy and non-energy sectors and a pathway vision to pave the way from where we are to where we want to go.
-Then, as looking into the future (\ie up to 2050 in this work) comes with its lot of uncertainties, these have to be assessed carefully in terms of characterisation and quantification.
-The former aims at defining the range over which parameters of the model vary.
+Then, as looking into the future (\ie up to 2050 in this work) comes with a lot of uncertainties, these have to be assessed carefully in terms of characterisation and quantification.
+The former aims at defining the range over which the parameters of the model vary.
 The latter allows assessing the impact that such uncertainties can have on the output of the model.
-Finally, meeting the environmental objectives while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crisis, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
-This work encompasses the optimisation of the policy, \ie set of actions to take along the transition with a specific methodology to assess its robustness.
+Finally, meeting the environmental objectives while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crises, require therefore a framework to assess the relevance and the timing of the decisions throughout the transition.
+This work encompasses the optimisation of the policy, \ie a set of actions to take along the transition with a specific methodology to assess its robustness.
 
@@
 \label{sec:meth:contributions}
-The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimise the behaviour of an artificial agent interacting with its environment, \ie the the whole-energy system through its transition.
+The main methodological contribution of this work is the implementation of the \gls{RL} approach to simulate and optimise the behaviour of an artificial agent interacting with its environment, \ie the whole-energy system through its transition.
 Given the uncertainties and potential shocks of the future, this approach allows the agent to play the transition in a sequential, \ie myopic, way and optimise the choice and the timing of its actions.
@@
 Originally developed to optimise the transition in one global optimisation up to 2050, \ie perfect foresight, part of this thesis consisted in making this model able to optimise the same transition but with sequential more limited time windows, \ie myopic approach.
-Besides its shorter computational time, this approach is more representative an actual decision process with limited foresight in the future \cite{babrowski2014reducing}.
+Besides its shorter computational time, this approach is more representative of an actual decision process with limited foresight in the future \cite{babrowski2014reducing}.
 
@@
 Given the uncertainties and the timespan of the transition, this approach allows highlighting the main ``directions'' of variation of the system design (\ie the installed capacities).
-After this step of identification, strategies can be projected on these directions to see how robust they are to the need of additional capacities.
+After this step of identification, strategies can be projected on these directions to see how robust they are to the need for additional capacities.
 
 Finally, more minor methodological developments are part of this thesis.
-Following an approach similar to \citet{guevara2022modeling}, we have extended the ranges of uncertainty developed by \citet{Moret2017} to the pathway optimisation.
-After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model subject to different emission-constraints \cite{rixhon2021role} as well as the pathway model (see Chapter \ref{chap:atom_mol}).
+Following an approach similar to \citet{guevara2022modeling}, we have extended the ranges of uncertainty developed by \citet{Moret2017} to pathway optimisation.
+After assessing the relevance of using \acrfull{PCE} on the optimisation of a whole-energy system \cite{limpens2020impact}, we have applied this uncertainty quantification method on the snapshot model, \ie optimising the system in a target future year considering a green field approach, subject to different emission constraints \cite{rixhon2021role} as well as the pathway model (see Chapter \ref{chap:atom_mol}).
 Eventually, starting from the initial investigation of \citet{goffauxpathway}, this work has converged to the most relevant formulation of the salvage value for the model EnergyScope.
@@
 As developers of the building blocks of the main contributions of this thesis, three main authors are to be mentioned for having brought a significant part of the methodological work.
-Based on Stefano Moret's monthly whole-energy system model (\ie EnergyScope) \cite{moret2016strategic}, Gauthier Limpens has developed the hourly version of the snapshot model (\ie EnergyScope TD) \cite{limpens2019energyscope}, as well as the perfect foresight pathway model \cite{limpens2024pathway}, to which I personally contributed too.
-Diederik Coppitters has developed the RHEIA framework allowing to quantify the impact of uncertainties and carry out robust optimisation of energy systems \cite{coppittersthesis}.
+Based on Stefano Moret's monthly whole-energy system model (\ie EnergyScope) \cite{moret2016strategic}, Gauthier Limpens has developed the hourly version of the snapshot model (\ie EnergyScope TD) \cite{limpens2019energyscope}, as well as the perfect foresight pathway model \cite{limpens2024pathway}, to which I personally contributed.
+Diederik Coppitters has developed the RHEIA framework that allows quantifying the impact of uncertainties and carry out robust optimisation of energy systems \cite{coppittersthesis}.
 The current work used this framework for the first of these functionalities.
-Finally, Stefano Moret extensively assessed the uncertainty characterisation on the Swiss energy system \cite{Moret2017}.
+Finally, Stefano Moret extensively assessed the uncertainty characterisation of the Swiss energy system \cite{Moret2017}.
 This thesis follows the same methodology, updating the uncertainty ranges for the pathway model.
@@
 This work optimises the entire transition pathway from a known system in 2020 up to 2050 thanks to EnergyScope Pathway \cite{limpens2024pathway}.
-According to pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has a hourly time-resolution and is an open-source documented model.
-Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with a hourly discretisation)\footnote{To keep this low computational time, crossover was disabled to solve the optimisation problem.
+According to the pathway models review (see Appendix \ref{app:ESPathway_choice}), EnergyScope Pathway can be categorised as an investment and operation optimisation model that assesses the whole-energy system, has an hourly time resolution and is an open-source documented model.
+Moreover, it maintains a low computational cost (\ie around 15 minutes for a 30-year pathway with an hourly discretisation)\footnote{To keep this low computational time, crossover was disabled to solve the optimisation problem.
 This had no observed impact regarding the values of the objective function or the design variables at the optimum.}.
 From the perfect to the myopic foresight of the transition optimisation, this section presents only the main constraints of the former approach to further dig into more details about the latter.
-The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from, EnergyScope TD, the snapshot model, optimising a target future year with a greenfield approach \cite{geidl2006greenfield}.
+The reader is invited to refer to Appendix \ref{app:ESPathway_full_formulation} for more details about the formulation of the model and its extension from EnergyScope TD, the snapshot model, optimising a target future year with a greenfield approach \cite{geidl2006greenfield}.
 More extensive information about the formulation choices, for instance, can be found in \cite{limpens2024pathway} and the documentation \cite{readthedocs_pathway}.
@@
 The whole-energy system model developed in this work originates from the perfect foresight (PF) formulation (\Cref{fig:meth_path_methodology_core}) of EnergyScope---the entire transition is computed in one optimisation, assuming a complete but uncertain knowledge of the different parameters until 2050 \cite{limpens2024pathway}.
-Each representative year is represented via the the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
+Each representative year is represented via the variables and constraints of the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
 Then, to draw a consistent pathway between these years, additional constraints aim at linking them, \eg limiting the modal shifts within some energy sectors.
@@
 In line with \citet{limpens2021generating}, we set the nominal value to $i_{rate}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}.
-Having a low value of discount rate favours less late expenses compared to a higher value of discount rate.
+Having a low-value discount rate favours less late expenses compared to a higher value of discount rate.
 In practice, the discount rate would vary depending on the technology investment risk.
 Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+However, EnergyScope only considers the vision of a central planner where a single agent makes the investment decisions without differentiating the sources of the capital, \ie private and public.
+Assuming a differentiated discount rate would require making assumptions about the private-public distribution of the capital.
 For this reason, we have decided to keep an identical value of discount rate for all the technologies.
@@
 Similarly, the salvage value is computed in the proportion of the remaining years of life versus the initial lifetime of an installed capacity of a technology from which the anticipatively decommissioned part, $\textbf{F\textsubscript{decom}}$, is removed.
-To cope with the end-of-time-horizon effects of multi-stages models, the salvage value avoids penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}.
+To cope with the end-of-time-horizon effects of multi-stage models, the salvage value avoids penalising the capital-intensive technologies towards the end of the transition \cite{poncelet2016myopic}.
 After assessing the sensitivity of the formulation of the salvage value (see Appendix \ref{sec:path_methodo}), we have kept the one given by Equation (\ref{eq:salvage}), being also similar to the expression used by \citet{prina2019transition}.
@@
 where $\textbf{C\textsubscript{opex}}(p)$ is the arithmetic average of the operational cost at the beginning and the end of the phase $p$,  $\textbf{C\textsubscript{opex}}(y\textsubscript{start})$ and $\textbf{C\textsubscript{opex}}(y\textsubscript{stop})$.
-The operational cost of a year, $y$, $\textbf{C\textsubscript{opex}} (y)$, is the sum of the costs related to maintenance and operation of technologies, $ \textbf{C\textsubscript{maint}}$, and the consumption of resources, $\textbf{C\textsubscript{op}}$ (see Equation (\ref{eq:opex_yearly})).
+The operational cost of a year, $y$, $\textbf{C\textsubscript{opex}} (y)$, is the sum of the costs related to the maintenance and operation of technologies, $ \textbf{C\textsubscript{maint}}$, and the consumption of resources, $\textbf{C\textsubscript{op}}$ (see Equation (\ref{eq:opex_yearly})).
 
@@
 
-The \ce{CO2}-budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Equation (\ref{eq:gwp_tot_transition})).
+The \ce{CO2} budget for the transition, $\textbf{GWP\textsubscript{tot,trans}}$ is equal to the arithmetic average of the representative years at the beginning and the end of each phase (see Equation (\ref{eq:gwp_tot_transition})).
 Similarly to initial operational costs to account for the system in place in 2020 (see Equation (\ref{eq:Copex_tot_v2})), $ \textbf{GWP\textsubscript{tot}}(2020)$ accounts for the entire operational emissions in 2020, as the initial cumulative emissions of the transition.
@@
 where $\textbf{GWP\textsubscript{tot}}(p)$ is the arithmetic average of the yearly emissions at the beginning and the end of the phase $p$,  $\textbf{GWP\textsubscript{tot}}(y\textsubscript{start})$ and $\textbf{GWP\textsubscript{tot}}(y\textsubscript{stop})$.
-The computation of these yearly emissions are based on the \acrfull{GWP} of the resources:
+The computation of these yearly emissions is based on the \acrfull{GWP} of the resources:
 
@@
 However, this work is still in progress and the database is not yet complete.
-Consequently, it is not included in this work and not accounted for.
+Consequently, it is not included in this work and is not accounted for.
 
@@
 where the variables $\textbf{F\textsubscript{old}}$ and $\textbf{F\textsubscript{decom}}$ are the capacities respectively having reached the end of their lifetime and prematurely decommissioned.
-Moreover, to account for the society inertia and to prevent unrealistically fast modal share change, constraints limit this change for the sectors of the low-temperature, the passenger mobility and freight mobility demands.
+Moreover, to account for the inertia of society and to prevent unrealistically fast modal share change, constraints limit this change for the sectors of the low-temperature, passenger mobility and freight mobility demands.
 The interested reader will find more information about the formulation choices related to it in the work of \citet{limpens2024pathway}.
@@
 
-Compared to the perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has three main advantages: shorter computational time, more realistic representation of the short-sightedness of decision-makers and the ability to capture shocks and unexpected events \cite{mccollum2020energy}.
+Compared to perfect foresight, the myopic approach (Figure \ref{fig:my_schematic}) has three main advantages: shorter computational time, more realistic representation of the short-sightedness of decision-makers and the ability to capture shocks and unexpected events \cite{mccollum2020energy}.
 For this reason, several studies are based on this approach \citep{babrowski2014reducing,poncelet2016myopic,nerini2017myopic,heuberger2018impact}.
 \citet{babrowski2014reducing} analysed the benefit of the myopic approach to reduce the computational time.
-\citet{poncelet2016myopic} uses this approach to analyse the expansion planning of the power sector beyond 2050 to assess the realism of the decision making process brought up by the myopic implementation.
+\citet{poncelet2016myopic} use this approach to analyse the expansion planning of the power sector beyond 2050 to assess the realism of the decision-making process brought up by the myopic implementation.
 \citet{nerini2017myopic} analysed the impact of the horizon windows and overlapping time.
@@
 In this example, the pathway instance has a time horizon of 10 years ($N\textsubscript{year,opti}=10$) with a 5 year-overlap ($N\textsubscript{year,overlap}=5$).
-As a comparison the perfect foresight (in blue) has a time horizon of 30 years.}
+As a comparison, the perfect foresight (in blue) has a time horizon of 30 years.}
 \label{fig:my_schematic}
@@
 \includegraphics[width=\textwidth]{MY_schematic_2.pdf}
-\caption{Sequential optimisation of the transition pathway in the myopic approach: (i) first time-window optimisation, (ii) set-up of the initial conditions of the second time-window, (iii) second time-window optimisation discarding intermediate results from the first time-window optimisation.}
+\caption{Sequential optimisation of the transition pathway in the myopic approach: (i) first time window optimisation, (ii) set-up of the initial conditions of the second time window, (iii) second time window optimisation discarding intermediate results from the first time window optimisation.}
 \label{fig:my_schematic_2}
@@
 Moreover, there is the possibility of having an overlap between two consecutive time windows.
-The timespan of this overlap is defined by the parameter $N\textsubscript{year,overlap}$.
-The philosophy followed behind the development of the myopic approach was to add another layer on top of the perfect foresight model in order to make it more modular.
+The period of this overlap is defined by the parameter $N\textsubscript{year,overlap}$.
+The philosophy behind the development of the myopic approach was to add another layer on top of the perfect foresight model to make it more modular.
 For this reason, the already existing constraints are marginally adapted.
 This way, the newly developed model can easily be used to perform a perfect foresight optimisation by setting the time window to $N\textsubscript{year,opti}=$30 years (\ie between 2020 and 2050) and the overlap between the time windows to $N\textsubscript{year,overlap}=$0.
-Consequently, as it is fundamental to define on the one hand, the actual time window on which the system is optimised, and on the other hand, the history, \ie what has already been optimised earlier in the transition, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
+It is fundamental to define on the one hand, the actual time window on which the system is optimised, and on the other hand, the history, \ie what has already been optimised earlier in the transition.
+Consequently, four new sets are implemented: $\text{YEARS}\textsubscript{WND}$, $\text{YEARS}\textsubscript{UP TO}$, $\text{PHASE}\textsubscript{WND}$ and $\text{PHASE}\textsubscript{UP TO}$ (see Table \ref{tab:path_my_sets}).
 
@@
 For instance, the first year of a time window is the result of the optimisation of the previous one.
-Therefore, optimising again this first year could lead to rounding errors preventing the optimisation to converge.
-For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimise, that is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
+Therefore, optimising again this first year could lead to rounding errors preventing the optimisation from converging.
+For this reason,  the set $\text{YEAR}\textsubscript{ONE}$ accounts for the first representative year of the time window to optimise, which is excluded from $\text{YEARS}\textsubscript{WND}$ to avoid these errors.
 This remark remains valid for any time window except the first one of the transition where the year 2020 is optimised even though its technological strategy is set according to the actual system presented in Appendix \ref{app:bel_2020}.
-Finally, as the end of time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Equation (\ref{eq:salvage}).\\
+Finally, as the end of the time windows changes for each of them, the parameter $remaining\_years$ has to be updated accordingly to keep a meaningful definition of $\textbf{C\textsubscript{inv,return}}$ in Equation (\ref{eq:salvage}).\\
 
@@
 has developed a Python \gls{API} called amplpy \cite{amplpy}.
-In a nutshell, this API allows the pre/post-processing of an ampl optimisation problem by accessing its features (\eg constraints,parameters, variables, objective function) from within Python.
+In a nutshell, this API allows the pre/post-processing of an ampl optimisation problem by accessing its features (\eg constraints, parameters, variables, objective function) from within Python.
 Using this \gls{API}, this updated version of the model interacts with the AMPL problem representing the optimisation of the whole-energy system transition pathway as represented in Figure \ref{fig:MY_process_code}.
-The myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window, until eventually reaching 2050.\\
+The myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time window, until eventually reaching 2050.\\
 
@@
 \caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
-Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time-window.
+Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time window.
 If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
@@
 The main difference lies in the myopic transition itself and especially in the earlier deployment of PVs and offshore wind turbines.
-These induce the reinforcement of the grid that is a capital-intensive and long-lifetime asset.
-This is mostly due to impact of the salvage value, Equation (\ref{eq:salvage}), in the objective function.
-Since this is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at early stages of the transition.
+These induce the reinforcement of the grid which is a capital-intensive and long-lifetime asset.
+This is mostly due to the impact of the salvage value, Equation (\ref{eq:salvage}), on the objective function.
+Since this is now the transition cost over a more limited time window (\ie 10 years rather than 30 years), a bigger salvage value, deduced from the total investments, leads to a temporary better optimum at the early stages of the transition.
 A more detailed comparison between the myopic and perfect foresight approaches is available in Appendix \ref{app:my_vs_pf}.
@@
 \label{subsec:meth:ES:discussion}
-Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time-resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix \ref{app:ESPathway_choice}).
+Given the research questions targeted in this work, EnergyScope Pathway was the right model to choose for several reasons: its low computational burden, its hourly time resolution, its operation and design optimisation of a whole-energy system and the openness of the code and availability of the documentation (see Appendix \ref{app:ESPathway_choice}).
 
@@
 \label{subsec:uncert_charac}
-Characterising precisely the uncertainty---ideally with their respective probability density functions (PDFs)---of the thousands of parameters in the model is daunting if not impossible because of lack of data \cite{marnay2006addressing}.
+Characterising precisely the uncertainty---ideally with their respective probability density functions (PDFs)---of the thousands of parameters in the model is daunting if not impossible because of the lack of data \cite{marnay2006addressing}.
 Therefore, we used a workaround developed by \citet{Moret2017} that defines relative ranges of variation for different groups of parameters.
@@
 Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about the availability.
-Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
+Finally, the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
 
@@
 \includegraphics[width=0.7\textwidth]{ranges_transition.pdf}
-\caption{Expansion of the width of uncertainty range for type III parameters.
+\caption{Expansion of the width of the uncertainty range for type III parameters.
 $\mu_{2020}$, $\mu_{2025}$, ...,  $\mu_{2050}$ are the nominal values equal to 0 as the uncertain parameters represent a relative increase/decrease of actual parameters of the model.
@@
 Finally, the model accounts for thousands of parameters.
-The computational burden to consider all of them separately would be completely overwhelming ($\sim 10^7$ model runs).
-Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped to one single uncertain parameter.
+The computational burden of considering all of them separately would be completely overwhelming ($\sim 10^7$ model runs).
+Similarly to other works \cite{Moret2017,limpens2020impact}, the model parameters that would follow the same uncertainty have been grouped into one single uncertain parameter.
 On top of mitigating the computational burden, this aims at grouping parameters that are closely linked with each other.
 For instance, the uncertainty on the cost of purchasing renewable electrofuels, $c_{\mathrm{op,electrofuels}}$, identically affects the cost of e-hydrogen, e-methane, e-ammonia and e-methanol.
-Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation of cost of electricity or the electrolyser, that drive the majority of their cost of purchasing \cite{h2coalition}.
+Indeed, besides their respective specificities, each of these fuels will be similarly affected by the variation in the cost of electricity or the electrolyser, which drive the majority of their cost of purchasing \cite{h2coalition}.
 Similarly, the uncertainties impacting the industrial demand, $industry\_\emph{EUD}$, alters equally the industrial high- and low-temperature and electricity demands as well as the non-energy demand.
@@
 
-To avoid the computational burden of well-known method like Monte-Carlo analysis \cite{yue2018review}, we used \gls{PCE} to carry out a \gls{GSA}.
+To avoid the computational burden of well-known methods like Monte-Carlo analysis \cite{yue2018review}, we used \gls{PCE} to carry out a \gls{GSA}.
 \gls{PCE} is an approach for surrogate-assisted \gls{UQ} that propagates uncertainties in input parameters through the system model.
@@
 To construct a PCE of the EnergyScope Pathway model, we employed the open-source Python framework RHEIA~\cite{coppitters2022rheia,readthedocs_rheia}.
-Where the first part of this section is dedicated to the mathematical definition of this approach, the second details its choice and summarises the comparison made with another approach (\ie Morris method) in a previous work \cite{limpens2020impact}.\\
+Where the first part of this section is dedicated to the mathematical definition of this approach, the second details its choice and summarises the comparison made with another approach (\ie the Morris method) in a previous work \cite{limpens2020impact}.\\
 
@@
 The coefficients ($u_0, u_1, \dots, u_{P+1}$) are quantified using a regression method applied to orthonormal polynomials~\cite{Sudret2014}.
-As uniform distributions are considered, the Legendre polynomials are adopted, as they are the associated family of polynomials that are orthogonal with respect to standard uniform distributions~\cite{Sudret2014}.
+As uniform distributions are considered, the Legendre polynomials are adopted, as they are the associated family of polynomials that are orthogonal concerning standard uniform distributions~\cite{Sudret2014}.
 
@@
 To generate the training samples, the quasi-random Sobol' sampling technique is employed~\cite{bratley2003implementing}.
-As a low-discrepancy sequence, this technique exhibits the main advantage to investigate efficiently and (almost) uniformly the hypercube of uncertainties, unlike uniformly distributed random numbers.
+As a low-discrepancy sequence, this technique exhibits the main advantage of investigating efficiently and (almost) uniformly the hypercube of uncertainties, unlike uniformly distributed random numbers.
 
@@
 For the specific study of this work, a polynomial order of 2 is necessary (with 1260 training samples as per Equation\,(\ref{eq:pce:nterms})) to achieve a \gls{LOO} error below \SI{1}{\%} for the total transition cost.
-As second order \gls{PCE} is the minimum to ensure accuracy of the surrogate model, selecting and grouping the parameters as presented in Section \ref{subsec:screening} is compulsory to alleviate the computational burden.
-Otherwise, considering thousands of independent uncertain parameters would lead to millions of EnergyScope Pathway runs, if no more.
+As second-order \gls{PCE} is the minimum to ensure the accuracy of the surrogate model, selecting and grouping the parameters as presented in Section \ref{subsec:screening} is compulsory to alleviate the computational burden.
+Otherwise, considering thousands of independent uncertain parameters would lead to millions of EnergyScope Pathway runs, if not more.
 Another alternative to reduce the computational cost of the \gls{GSA} would be to consider the monthly (instead of hourly) time resolution for the runs of EnergyScope Pathway.
-However, despite its shorter computation time (\ie couple of seconds), the main drawback of the monthly model is its poor representation of the integration of \gls{VRES} \cite{limpens2024pathway}.
+However, despite its shorter computation time (\ie a couple of seconds), the main drawback of the monthly model is its poor representation of the integration of \gls{VRES} \cite{limpens2024pathway}.
 
@@
 
-Here, $A$ denotes the collection of all PCE coefficients, and $\alpha_i$ corresponds to the coefficient associated with the uncertain parameter $i$.\\
+Here, $A$ denotes the collection of all PCE coefficients and $\alpha_i$ corresponds to the coefficient associated with the uncertain parameter $i$.\\
 
@@
 A first-order \gls{PCE} to assess the uncertainties on the total transition cost has a \gls{LOO} error of 16\%.
-However, it has the main advantage to require significantly fewer runs than more accurate second-order \gls{PCE} (70 versus 1260) while computing similar Sobol' indices for the most impacting parameters, and consequently giving similar ranking.
+However, it has the main advantage of requiring significantly fewer runs than more accurate second-order \gls{PCE} (70 versus 1260) while computing similar Sobol' indices for the most impacting parameters, and consequently giving a similar ranking.
 
@@
 \caption{Total Sobol' indices of the uncertain parameters over the total transition cost via \gls{PCE} of orders 2 and 1 having a \gls{LOO} error of 1\% and 16\%, respectively.
-The similar rankings (and indices) show the validity of using the faster (even though less accurate) lower order \gls{PCE} to rank the most impacting uncertain parameters.}
+The similar rankings (and indices) show the validity of using the faster (even though less accurate) lower-order \gls{PCE} to rank the most impacting uncertain parameters.}
 \label{tab:methodo:Sobol_ranking_order_1_2}
@@
 \noindent
-Besides being an in-house used method, an early step of this thesis consisted in assessing \gls{PCE} with similar approach used in the literature \cite{limpens2020impact}.
+Besides being an in-house method, an early step of this thesis consisted in assessing \gls{PCE} with a similar approach used in the literature \cite{limpens2020impact}.
 
@@
 
-Then, given these random one-factor-at-a-time samples, Morris method defines, for a given set of $\vec{\xi}$, the elementary effect of the $i$th parameter ($EE_i$) as :
+Then, given these random one-factor-at-a-time samples, the Morris method defines, for a given set of $\vec{\xi}$, the elementary effect of the $i$th parameter ($EE_i$) as :
 \begin{equation}
@@
 
-Finally, in order to evaluate the importance of the $i$th parameter over an output, Morris method relies on $F_i$, the distribution of $r$ elementary effects.
+Finally, to evaluate the importance of the $i$th parameter over an output, the Morris method relies on $F_i$, the distribution of $r$ elementary effects.
 Computing the mean, $\mu_{i}=\mu(F_{i})$, and the standard deviation, $\sigma_{i}=\sigma(F_{i})$, of the $F_{i}$ distribution, allows ranking the parameters based on their influence on the concerned output.
-Usually, in Morris method, $p$ and $r$ respectively get values as follows : $p\in \{4,6,8\}$ and $r\in [15;100]$ depending on, $d$, the number of uncertain parameters.
+Usually, in the Morris method, $p$ and $r$ respectively get values as follows: $p\in \{4,6,8\}$ and $r\in [15;100]$ depending on, $d$, the number of uncertain parameters.
 The higher this number is,  the higher shall be, simultaneously, $p$ and $r$.
-In the following comparative analysis, we set $p$ and $r$ to their maximum values, respectively $8$ and $100$ in order to get the most reliable parameters ranking.\\
+In the following comparative analysis, we set $p$ and $r$ to their maximum values, respectively $8$ and $100$ to get the most reliable parameters ranking.\\
 
@@
 \end{equation}
-Among other things, the $SEE$ allows comparing the influence of different inputs on the same output or compare the influence of a same parameter on different outputs, even if these parameters or outputs are significantly different in terms of variation range or average amplitude.
+Among other things, the $SEE$ allows comparing the influence of different inputs on the same output or compare the influence of the same parameter on different outputs, even if these parameters or outputs are significantly different in terms of variation range or average amplitude.
 Moreover, this standardised analysis does not require any additional model evaluations.\\
@@
 	\caption{Comparison of the Top-14 rankings for the improved Morris method (left) and total-order \gls{PCE} method (right).
-We used the DTU's implementation of Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$).
+We used the DTU's implementation of the Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$).
 }
@@
 After the initial phase of grouping (Section \ref{subsec:uncert_charac}), a preliminary screening was necessary to identify the key parameters to account for in this \gls{GSA}.
-\citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2}-limits using the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
+\citet{rixhon2021role} performed a similar sensitivity analysis on the 2050 Belgian whole-energy system under different \ce{CO2} limits using the snapshot model, EnergyScope TD \cite{limpens2019energyscope}.
 Screening the results of this work, we have discarded some parameters with negligible impact such as CAPEX of electrolysers or variation of the freight demand.
-Parameters are considered as ``negligible'' if their Sobol' index is below the threshold $=1/d$, $d$ being the total number of uncertain parameters \cite{Turati2017}.
+Parameters are considered ``negligible'' if their Sobol' index is below the threshold $=1/d$, $d$ being the total number of uncertain parameters \cite{Turati2017}.
 On top of this selection, we have added others that were intrinsic to the pathway formulation, \eg modal share changes, or related to the integration of \gls{SMR}, $f_{\mathrm{max,SMR}}$.
@@
 Carrying out a \acrfull{GSA} requires two steps: uncertainty characterisation and uncertainty quantification.
-Characterising the distribution followed by uncertain parameters require a lot of data that are either seldom available or poorly documented in the literature.
+Characterising the distribution followed by uncertain parameters requires a lot of data that are either seldom available or poorly documented in the literature.
 Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distribution.
-Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further researches to refine the characterisation of these parameters.
-
-About the method to quantify the uncertainty, \gls{PCE} has the advantages to provide Sobol' indices as the impact indicators of the uncertain parameters and extract statistical moments of the output of interest.
+Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further research to refine the characterisation of these parameters.
+
+About the method to quantify the uncertainty, \gls{PCE} has the advantage to provide Sobol' indices as the impact indicators of the uncertain parameters and extract statistical moments of the output of interest.
 Based on our experience, considering a second-order \gls{PCE} is the minimum to get a reasonably low \gls{LOO} error (below 1\%) and reliable results.
-However, as presented in Section \ref{subsec:pce}, starting with a first-order \gls{PCE} is a good proxy to get a initial insight into the impact of the parameters and discard the negligible ones before moving to a higher order \gls{PCE} to increase the accuracy of the results.
+However, as presented in Section \ref{subsec:pce}, starting with a first-order \gls{PCE} is a good proxy to get an initial insight into the impact of the parameters and discard the negligible ones before moving to a higher-order \gls{PCE} to increase the accuracy of the results.
 
@@
 
-The transition towards carbon-neutrality of a whole-energy system (\ie including all streams of energy carriers and demands) is uncertain.
-Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), policy makers rather go through multiple rolling-horizon short-term decisions.
+The transition towards carbon neutrality of a whole-energy system (\ie including all streams of energy carriers and demands) is uncertain.
+Therefore, instead of establishing single-shot definitive plans towards 2050 (and beyond), policymakers rather go through multiple rolling-horizon short-term decisions.
 Yet, these decisions can have long-term impacts, 20 to 50 years.
-The long-term future is intrinsically uncertain and could be the place for potential sudden unexpected events.
-Meeting the environmental objectives without prescribed \ce{CO2}-pathway while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crises, requires a framework to assess the relevance and the timing of the decisions throughout the transition.
+The long-term future is intrinsically uncertain and could be the place for potentially sudden unexpected events.
+Meeting the environmental objectives without a prescribed \ce{CO2} pathway while minimising the cost of the system, accounting for this decision-making process, the uncertainties, and potential shocks/crises, requires a framework to assess the relevance and the timing of the decisions throughout the transition.
 
@@
 At each stage of the problem, here the transition pathway of a whole-energy system, the model has the possibility of recourse, \ie to adapt the decisions made at earlier stages, in the response to unveiled uncertainties \cite{grossmann2016recent}.
-Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plan and high/low growth scenarios.
-The authors found that hedging strategies, adapting with the future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
-However, stochastic programming is usually applied to limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly.
+Using MARKAL model \cite{fishbone1981markal}, \citet{kanudia1998robust} assessed a multi-stage stochastic optimisation of the transition of Quebec between 1995 and 2035 accounting for high/low mitigation action plans and high/low growth scenarios.
+The authors found that hedging strategies, adapting to future uncertainties, were outperforming the perfect foresight and deterministic optimisation of the different scenarios.
+However, stochastic programming is usually applied to a limited number of uncertainties, \ie up to 10, and relies on probability distributions that are often difficult to define properly.
 Increasing the number of these uncertainties in stochastic programming increases exponentially the computational burden that limits the use of such a method in \gls{IAMs} \cite{nicolas2021robust}.
@@
 To keep the best of both worlds, \ie stochastic programming and robust optimisation, we decided to investigate the \gls{RL} approach, and its policy optimisation mechanism, to explore myopic transition pathways under uncertainties.
-Indeed, policymaking for transitioning a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
-\gls{RL} exhibits two main advantages: its effectiveness to handle uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimise the policy \cite{perera2021applications}.
+Indeed, policymaking to transition a whole-energy system can be viewed as an iterative process of learning from policy implementation efforts, involving ongoing analysis of energy policy challenges and experimenting with various solutions \cite{howlett1995studying}.
+\gls{RL} exhibits two main advantages: its effectiveness in handling uncertainties and the model-free approach where an accurate representation of the real world is not needed to optimise the policy \cite{perera2021applications}.
 
@@
 Then, its application to the myopic optimisation of a whole-energy system is introduced as well as the policy optimisation algorithm.
-As Chapter \ref{chap:case_study} presents the input of the case study and Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} on the pathway model, the reader interested by the results of the \gls{RL} method is invited to go to Chapter \ref{chap:chap_RL}.
+Chapter \ref{chap:case_study} presents the input of the case study and Chapter \ref{chap:atom_mol} details the results of the \gls{GSA} on the pathway model, the reader interested by the results of the \gls{RL}-method is invited to go to Chapter \ref{chap:chap_RL}.
 
@@
 The latter feeds back the agent with a reward and the new state.
-This goes on until reaching the end of the episode.
+This goes on until the end of the episode.
 When the episode is done, the agent starts again from an initial state, takes an action and so on.
@@
 This policy refers to the strategy or mapping from states to actions that the agent employs to make decisions.
-Essentially, it defines the behavior of the agent in the environment.
+Essentially, it defines the behaviour of the agent in the environment.
 The ultimate goal of the agent is often to find an optimal policy, which maximises the expected cumulative reward over time.
@@
 In the exploration, the action is randomly picked within the action space.
-For further information, the interested reader is invited to refer to work of \citet{sutton2018reinforcement} or the course given by \citet{davidsilver_RL_online} available online.
+For further information, the interested reader is invited to refer to the work of \citet{sutton2018reinforcement} or the course given by \citet{davidsilver_RL_online} available online.
 
@@
 Even though an observation is a subset of the state, this work uses these two words interchangeably.
-From this state, the agent takes a step: action, reward, new state.
+From this state, the agent takes a step: action, reward,  and new state.
 The action, $\bm{a}_0$, impacts the environment, \ie the energy system limited transition over the first decision window (2020-2030).
@@
 First, if the actions taken by the agent make the optimisation infeasible, the episode is prematurely stopped before reaching 2050.
-Similarly, cumulative emissions of the system over the predefined \ce{CO2}-budget (see Section \ref{sec:cs:CO2-budget}) lead to an anticipated end of the episode.
+Similarly, cumulative emissions of the system over the predefined \ce{CO2} budget (see Section \ref{sec:cs:CO2-budget}) lead to an anticipated end of the episode.
 Finally, the ``natural'' end is the prescribed end of the transition, \ie 2050.
@@
 In practice, in a model-free approach, the agent estimates the optimal policy directly from experience and without estimating the dynamics of the environment.
-However, model-free methods suffer from two major drawbacks: their sample inefficiency and their sensitivity with respect to their hyper-parameters (\eg learning rates, exploration constants) \cite{haarnoja2018soft}.
-The former leads to a too expensive computational burden while the second requires meticulous settings to get good results.
-To overcome these two challenges, we needed to chose between an ``on-policy'' or ``off-policy'' algorithm.
+However, model-free methods suffer from two major drawbacks: their sample inefficiency and their sensitivity concerning their hyper-parameters (\eg learning rates, exploration constants) \cite{haarnoja2018soft}.
+The former leads to a too-expensive computational burden while the second requires meticulous settings to get good results.
+To overcome these two challenges, we needed to choose between an ``on-policy'' or ``off-policy'' algorithm.
 In a nutshell, in on-policy learning, the agent learns the value function or policy based on the data it generates by following its current policy whereas, in off-policy, the agent can learn from data collected by any policy, not just the one it is currently following, which provides greater flexibility and potential for reusing data stored in the so-called replay buffer.
-This makes off-policy algorithms more data efficient and ensuring better exploration by reusing past experiences or even following random exploration \cite{haarnoja2018soft}.
+This makes off-policy algorithms more data efficient and ensures better exploration by reusing past experiences or even following random exploration \cite{haarnoja2018soft}.
 
@@
 \gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function (see Equation (\ref{eq:SAC_objective})).
-The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie first term of Equation (\ref{eq:SAC_objective}).
-In the \gls{RL} context, the entropy, also called ``Shannon entropy'', stands for the randomness or stochasticity of the policy.
+The word ``augmented'' here is in opposition to the conventional \gls{RL} objective function that is only based on the cumulative reward, \ie the first term of Equation (\ref{eq:SAC_objective}).
+In the \gls{RL} context, entropy, also called ``Shannon entropy'', stands for the randomness or stochasticity of the policy.
 
@@
 $\gamma$ determines how much importance we want to give to future rewards within an episode.
-$\zeta$ balances the trade-off between exploitation of proven actions via the return maximisation, \ie $\sum_{n=0}^{N_{ep}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right)$, and exploration through the entropy term, \ie $\log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right)$.
+$\zeta$ balances the trade-off between the exploitation of proven actions via the return maximisation, \ie $\sum_{n=0}^{N_{ep}}\gamma^n r_n\left(\bm{o}_n,\bm{a}_n \right)$, and the exploration through the entropy term, \ie $\log \left(\pi\left(\bm{a}_n | \bm{o}_n\right) \right)$.
 This way, \gls{SAC} ensures sample efficiency while improving exploration \cite{haarnoja2017reinforcement} and robustness \cite{ziebart2010modeling}.
 In their work, \citet{haarnoja2017reinforcement} showed a lower sensitivity of \gls{SAC} to hyper-parameters.
-These make \gls{SAC} a state-of-the-art algorithm and one of the most efficient model-free deep RL method nowadays \cite{haarnoja2017reinforcement}.
+These make \gls{SAC} a state-of-the-art algorithm and one of the most efficient model-free deep RL methods nowadays \cite{haarnoja2017reinforcement}.
 In this thesis, we used the open-source \gls{SAC} package developed by \textsc{Stable-Baselines3} \cite{raffin2021stable} where the policy \gls{NN} is a fully connected multilayer perceptron (MLP) built with \textsc{TensorFlow} \cite{abadi2016tensorflow}.
@@
 \acrfull{RL} was found to be an appropriate approach to explore myopic transition pathways under uncertainties and assess the robustness of policies to support such pathways.
-As a novice user of a \gls{RL} framework considering a continuous action and state spaces, we would recommend to opt for \gls{SAC} as it is sample efficient, ensures a wide exploration and has a low sensitivity to hyper-parameters \cite{haarnoja2018soft}.
+As a novice user of a \gls{RL} framework considering continuous action and state spaces, we would recommend opting for \gls{SAC} as it is sample efficient, ensures a wide exploration and has a low sensitivity to hyper-parameters \cite{haarnoja2018soft}.
 Using the \gls{SAC} package developed by \textsc{Stable-Baselines3} allows a handy introduction to apply \gls{RL}.
 
-Besides the choice of the algorithm, most of the work when using a \gls{RL} framework consists in properly define the rules of the games following which the agent interacts with its environment: actions, states and reward.
+Besides the choice of the algorithm, most of the work when using a \gls{RL} framework consists in properly defining the rules of the games following which the agent interacts with its environment: action, state and reward.
 These are defined and discussed in Chapter \ref{chap:chap_RL}.
@@
 When optimising a transition pathway of a whole-energy system, including its uncertainties, capturing the variance of design (\ie installed capacities of each technology) can become overwhelming due to the curse of dimensionality.
-For the case study detailed in Chapter \ref{chap:case_study}, this consists of 113 possibles technologies over 7 representative years.
+For the case study detailed in Chapter \ref{chap:case_study}, this consists of 113 possible technologies over 7 representative years.
 To tackle this challenge, we have developed a methodology based on the \acrfull{PCA}.
-The philosophy behind this approach is to identify key-combinations of design variables capturing most of the system variance, beyond the sole objective function, \ie total transition cost, but without having to compare each design variable individually.
+The philosophy behind this approach is to identify key combinations of design variables capturing most of the system variance, beyond the sole objective function, \ie total transition cost, but without having to compare each design variable individually.
 This methodology provides two main outputs.
@@
 Second, these ``directions of variation'' can highlight key modal shifts or highly varying design strategies over the transition.
-After introducing the general concept of \gls{PCA}, this section aims at detailing the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of roadmaps.
-The reader interested by the results of the \gls{PCA}-based method is invited to go to Chapter \ref{chap:chap_RobPol}.
+After introducing the general concept of \gls{PCA}, this section aims to detail the methodology proposed to obtain these ``directions of variation'' and to assess the robustness of roadmaps.
+The reader interested in the results of the \gls{PCA}-based method is invited to go to Chapter \ref{chap:chap_RobPol}.
 
@@
 \label{subsec:meth:PCA:PCA}
-Born in the early 20th century, the \acrfull{PCA} finds its fundamentals from the \gls{SVD} \cite{pearson1901on,hotelling1933analysis}.
-\gls{SVD} is a generalisation, to an arbitrary (\ie not necessary square) matrix, of the spectral theorem stating that a normal matrix can be diagonalised by an orthonormal basis of eigenvectors.
+Born in the early 20th century, the \acrfull{PCA} finds its fundamentals in the \gls{SVD} \cite{pearson1901on,hotelling1933analysis}.
+\gls{SVD} is a generalisation, to an arbitrary (\ie not necessarily square) matrix, of the spectral theorem stating that a normal matrix can be diagonalised by an orthonormal basis of eigenvectors.
 The core concept of principal component analysis (PCA) involves simplifying the original dataset with numerous interconnected variables by reducing its dimensionality while preserving as much variability within the data.
@@
 This PC captures $\lambda_k$ variance of the original data.
-In other words, a high absolute value of $\alpha_{ki}$ means that $x_i$ has a significant impact in the direction given by the $k^{\text{th}}$ PC \cite{zdybal2022advancing}.
+In other words, a high absolute value of $\alpha_{ki}$ means that $x_i$ has a significant impact on the direction given by the $k^{\text{th}}$ PC \cite{zdybal2022advancing}.
 
@@
 Indeed, \gls{PCA} also looks for properties that allow reconstructing the original features as accurately as possible.
-In practice, it aims at minimising the total reconstruction error that is the average squared distance between the original observations and their respective projection.
+In practice, it aims at minimising the total reconstruction error which is the average squared distance between the original observations and their respective projection.
 
@@
 \caption{Two different projections (red dots) of the original observations (grey dots).
-Projection on the left captures a bigger variance than the projection on the right.
+The projection on the left captures a bigger variance than the projection on the right.
 At the same time, the average squared distance between the original observations and their respective projection is smaller for the projection on the left.
@@
 Like any other dimension-reduction process, \gls{PCA} has to be supplied with relevant data to reach the stated objective.
-Like the normalisation of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or order of magnitude.
-Then, properly handling the outliers allows reaching the relevant level of metric between the too vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
-After this preprocessing, PCs can be computed for each of the representative year of the transition.
-Finally, PCs that indicate similar directions in different years are averaged into one single PCs of transition.
+Like the normalisation of the eigenvectors (see Section \ref{subsec:meth:PCA:PCA}), data scaling is fundamental to compare features having potentially different units and/or orders of magnitude.
+Then, properly handling the outliers allows for reaching the relevant level of metric between the too-vague information of the sole total transition cost and too many details hidden in the peculiar/outlying cases.
+After this preprocessing, PCs can be computed for each of the representative years of the transition.
+Finally, PCs that indicate similar directions in different years are averaged into one single PC of transition.
 The different PCs of transition give a bigger picture over the whole transition.\\
@@
 To characterise the variations of design within the transition under uncertainties, we have focused on the installed capacities, $\textbf{F}(y,j)$ for all $y \in \text{\emph{YEARS}}$ and $j \in \text{\emph{TECH}}$ (see Equation (\ref{eq:F_newBuilt})).
-Even though these represent only the design part of the result of the optimisation, focusing on the installed capacities gives a direct information regarding the required capital investments (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation.
+Even though these represent only the design part of the result of the optimisation, focusing on the installed capacities gives direct information regarding the required capital investments (see Equation (\ref{eq:CAPEX_v2})) and, more indirectly, the operation.
 In other words, it captures the technological landscape of the transition.
@@
 Preprocessing the dataset before employing a method to reduce dimensionality, like \gls{PCA}, can greatly affect the structure of the simplified representation and the characteristics of the features extracted from the dataset \cite{parente2013principal,peerenboom2015dimension}.
-Scaling the original raw data via normalisation, \ie reducing data to $[0, 1]$ interval has a double purposes: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
+Scaling the original raw data via normalisation, \ie reducing data to $[0, 1]$ interval has a double purpose: to assess variables (i) representing different sorts of features, with different units (\eg installed capacity of electricity and mobility technologies) and, (ii) ranging over the different orders of magnitude (\eg installed capacity of private and public mobility) (see Appendix \ref{app:bel_2020}).
 Consequently, the first part of this data preprocessing consists in scaling the installed capacities versus their respective sector and representative year (see Equation (\ref{eq:PCA_scaling_0})).
@@
 
-Then, to give ``directions/metrics'' representative to the size of each sector within the energy system, we added another weight based on the relative share of commodity produced by each sector.
-For the case study of Belgium detailed in Chapter \ref{chap:case_study}, this gives a higher weight for electricity and low-temperature heat sectors (see Figure \ref{fig:Share_EUD_sectors}).
+Then, to give ``directions/metrics'' representative of the size of each sector within the energy system, we added another weight based on the relative share of commodities produced by each sector.
+For the case study of Belgium detailed in Chapter \ref{chap:case_study}, this gives a higher weight to electricity and low-temperature heat sectors (see Figure \ref{fig:Share_EUD_sectors}).
 
@@
 \includegraphics[width=1\textwidth]{Share_EUD_sectors.pdf}
-\caption{Multiplying factor for each of the end-use sectors in the case of Belgian energy transition.
-These shares are based on results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimised through the perfect foresight approach.
+\caption{Multiplying factors for each of the end-use sectors in the case of the Belgian energy transition.
+These shares are based on the results of the reference scenario (REF) where nominal values are considered for the uncertain parameters and the transition is optimised through the perfect foresight approach.
 Over the transition, sectors like electricity (\ie from 22\% in 2025 to 33\% in 2050) or ammonia (\ie from 2\% to 8\%) become more important due to sector coupling, \eg e-mobility or ammonia-\gls{CCGT}.}
@@
 One can notice that this second scaling factor omits the infrastructure and storage technologies.
-In the process to define ``metrics'' to assess the robustness of a roadmap for the case of Belgium, this has a negligible impact.
-Indeed, the variation of the installed capacity of these technologies are either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies.
-For instance, \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of grid is caused by additional capacities of \gls{VRES}.
+In the process of defining ``metrics'' to assess the robustness of a roadmap for the case of Belgium, this has a negligible impact.
+Indeed, the variation of the installed capacity of these technologies is either limited compared to end-use-type (EUT) technologies, \ie limiting their influence in the definition of PCs, or directly linked to these EUT technologies.
+For instance, \gls{DHN} installed capacity is directly proportional to technologies producing \gls{LT} heat in \gls{DHN} or the additional capacity of the grid is caused by additional capacities of \gls{VRES}.
 \\
@@
 These are defined as data points differing significantly from the rest of the data set.
-Being extreme values, outliers influence the overall dataset variance and, consequently, rotate the PCs directions towards them \cite{stanimirova2007dealing}.
+Being extreme values, outliers influence the overall dataset variance and, consequently, rotate the PC directions towards them \cite{stanimirova2007dealing}.
 In the context of \gls{PCA}, outliers could be defined as ``model fit outliers'' as their presence influences the fit of the model.
@@
 These whiskers are themselves constructed as being 1.5 times the \gls{IQR} (Q3 - Q1) higher or lower than the third (Q3) or first quartile (Q1), respectively (see Figure \ref{fig:Schematic_boxplot_methodo}).
-Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for these specific technology and year.
+Therefore, the installed capacity of a technology in the year $y$ of the sample $s$ is defined as an outlier if it falls out of this range compared to the dataset for this specific technology and year.
 There exist several techniques to handle these outliers depending on their nature or the method used to identify them.
@@
 Now that data are selected and preprocessed, principal components are first computed for each representative year separately, using the Python package \textsc{PCA} from \textsc{sklearn.decomposition}.
-As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73 in our case study\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies nor the 25 storage technologies.}) which is intractable.
+As explained in Section \ref{subsec:meth:PCA:PCA}, the number of PCs per year to retain can go up to the number of considered variables (\ie 73 in our case study\footnote{73 technologies out of the 113 in total as we do not consider the 15 infrastructure technologies or the 25 storage technologies.}) which is intractable.
 Moreover, the first PCs keep track of most of the variance of the system whereas the last ones present a smaller interest.
@@
 It is important to consider the second case as \gls{PCA} gives the magnitude of variation within the original feature space via the components of the eigenvectors, regardless of the direction of these vectors.
-For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$.
-The second vector being the opposite of the first, they are considered as equivalent in the regard of \gls{PCA}.
+For instance, in a two-dimension original space, if the first PC has $(\sqrt{2}/2;\sqrt{2}/2)$ as an eigenvector, this could also be $(-\sqrt{2}/2;-\sqrt{2}/2)$.
+The second vector being the opposite of the first, they are considered \gls{PCA}-wise equivalent.
 
@@
 
-In our case, we have decided to set the value of $\alpha$ to 85\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' with regard to the number of $\text{PC}_{\text{transition}}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).
-Besides being the metrics to assess the robustness of roadmap, these $\text{PC}_{\text{transition}}$ also point out the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.
+In our case, we have decided to set the value of $\alpha$ to 85\% as, beyond this value, the marginal gain of captured variance becomes too ``costly'' in terms of the number of $\text{PC}_{\text{transition}}$, \ie the level of details, to account for (see Figure \ref{fig:PC_y_share_tot_var}).
+Besides being the metrics to assess the robustness of roadmaps, these $\text{PC}_{\text{transition}}$ also point out the technologies that are more sensitive to uncertainties and are more likely to impact the overall variance of the transition design.
 
@@
 In their analysis, \citet{moret2020overcapacity} assessed the additional capacities after the realisation of uncertainties, the so-called \textit{possibility of recourse} \cite{grossmann2016recent}.
-These adaptations from the initial investments lead to overcapacity where assets are built but not used because the values taken by uncertain parameters make them cost inefficient.
-To address the same question of overcapacity in the context of energy transition pathway with progressively revealed uncertainties, we have adapted the method of \citet{moret2020overcapacity} (see Figure \ref{fig:Methodo_Robustness}).
+These adaptations from the initial investments lead to overcapacity where assets are built but not used because the values taken by uncertain parameters make them cost-inefficient.
+To address the same question of overcapacity in the context of the energy transition pathway with progressively revealed uncertainties, we have adapted the method of \citet{moret2020overcapacity} (see Figure \ref{fig:Methodo_Robustness}).
 Roadmaps are defined by setting minimal installed capacities based on the results of different transition pathway optimisations.
@@
 \noindent
-where the margin of error, MOE,  is computed thanks to standard error of the mean, SEM, and assuming a Student's, distribution, t,  of the $N$ projected data:
+where the margin of error, MOE,  is computed thanks to the standard error of the mean, SEM, and assuming a Student's, distribution, t,  of the $N$ projected data:
 $$
@@
 \noindent
-where $\mathrm{PPF}_{\mathrm{t}}$ is the percent point function (\ie inverse of cumulative distribution function) of the Student's distribution.
-It is really the range over which the distribution spans, $2\cdot \mathrm{MOE}$, that matters to characterise the variability of the projections of a roadmap.
-Indeed, as $\text{PC}_{\text{transition}}$ are vectors composed of positive and negative coefficients, the average position of the distribution, $\mu$, does not bring insight about the robustness of a roadmap.
+where $\mathrm{PPF}_{\mathrm{t}}$ is the per cent point function (\ie the inverse of cumulative distribution function) of the Student's distribution.
+It is the range over which the distribution spans, $2\cdot \mathrm{MOE}$, that matters to characterise the variability of the projections of a roadmap.
+Indeed, as $\text{PC}_{\text{transition}}$ are vectors composed of positive and negative coefficients, the average position of the distribution, $\mu$, does not bring insight into the robustness of a roadmap.
 
@@
 \label{subsec:meth:PCA:discussion}
-\acrfull{PCA} was found to be an appropriate method to provide a metric that capture a major part of the system design variance through the transition while keeping the amount of information to a ``graspable'' level of details.
+\acrfull{PCA} was found to be an appropriate method to provide a metric that captures a major part of the system design variance through the transition while keeping the amount of information to a ``graspable'' level of detail.
 Besides selecting the installed capacities of end-use-type categories (electricity, heat, mobility and non-energy), one could discuss the scaling of these data and the management of the outliers.
@@
 To avoid singularities in our PCs, we have decided \textit{a posteriori} to put high and low outliers to the upper and lower fences, respectively.
-However,before managing the outliers, we would recommend to keep them and assess the provided results in terms of number of PCs and the key technologies that characterise them.
+However, before managing the outliers, we would recommend to keep them and assess the provided results in terms of the number of PCs and the key technologies that characterise them.
 
@@
 These thresholds have been defined based on good practices discussed in the literature and \textit{a posteriori} observations specific to our case study.
-We would recommend starting with these values and adapting them in a second time if needed to increase or decrease the level of details captured by the PCs.
+We would recommend starting with these values and adapting them in a second time if needed to increase or decrease the level of detail captured by the PCs.

### 2024-07-19 rebuttals/rebuttals.tex
Grammarly up to 4.2 (included)

@@
 As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.
-This need for this radical change and the fact that solutions rely a lot on assumed to be carbon neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
+This need for this radical change and the fact that solutions rely a lot on assumed to be carbon-neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
 
@@
 Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
 This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.

### 2024-07-19 chap_RL/chap_RL.tex
Grammarly up to 4.2 (included)

@@
 First of all, as introduced in Section \ref{subsec:meth_RL_fundamentals}, when applied to energy systems, \gls{RL} is more dedicated either to smaller scale systems (\eg \gls{BEMS}, vehicles and energy devices) or to sector-specific, often the power sector, problems (\eg dispatch problems, energy markets and grid) \cite{perera2021applications}.
-In our case, the sector-coupling, the long-term goal at the end of a multiple-steps transition and the number of uncertain parameters make this application new for \gls{RL}.
+In our case, the sector-coupling, the long-term goal at the end of a multiple-step transition and the number of uncertain parameters make this application new for \gls{RL}.
 
@@
 Defining the levers of action, the core of the policy, to support the transition of a country-size whole-energy system is challenging, especially when accounting for political and socio-technical aspects \cite{castrejon2020making}.
-In our work, focusing only on the techno-economic aspect, we assume that the actions taken by the agent are directly implemented and impacting the environment.
+In our work, focusing only on the techno-economic aspect, we assume that the actions taken by the agent are directly implemented and impact the environment.
 In other words, considering only the techno-economic lens, there is no moderation nor contest towards the agent's actions, as the objective is to assess how far and when within the transition to push the different levers of action.
@@
 \noindent
-This action is equivalent to setting a national \ce{CO2}-quota.
+This action is equivalent to setting a national \ce{CO2} quota.
 
@@
 
-When the reward is not properly defined, the agent may optimise its policy for an unintended objective, leading to undesired or subotpimal behavior, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
+When the reward is not properly defined, the agent may optimise its policy for an unintended objective, leading to undesired or suboptimal behaviour, \ie the so-called misalignment of the learning objective \cite{christiano2017deep}.
 Even worse, it can lead to reward hacking (or reward tampering) where the agent exploits loopholes in the reward function to achieve higher rewards without actually performing the desired task \cite{amodei2016concrete}.
-On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring less episode to converge to the optimal policy.
+On the contrary, a proper definition of the reward function increases the sample efficiency, \ie requiring fewer episodes to converge to the optimal policy.
 It also makes the policy more stable and able to withstand variations and uncertainties in the environment \cite{henderson2018deep}.
@@
 Through its maximisation of the expected return (see Section \ref{subsec:meth_RL_algo}), a \gls{RL}-agent is as sensitive to positive reward, \ie the carrot, as negative reward, \ie the stick.
-When the former encourages desired behaviours, the latter can be seen as a penalty or a punishment and discourages the undesirable behaviours \cite{sutton2018reinforcement}.
+When the former encourages desired behaviours, the latter can be seen as a penalty or a punishment and discourages undesirable behaviours \cite{sutton2018reinforcement}.
 In our case, we have decided to combine these two approaches (see Figure \ref{fig:Reward}).
@@
 \caption{Reward function, $R$.
-Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded.
+Before 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded.
 If the optimisation provides a solution and the \ce{CO2} budget is not exceeded, the episode continues.
@@
 On top of that, the reward function includes a secondary objective: the cumulative transition cost.
-To make the agent sensitive to the cost-impact of its policy, we added the total transition cost in the reward function where the \emph{Trans.
+To make the agent sensitive to the cost impact of its policy, we added the total transition cost in the reward function where the \emph{Trans.
 cost$_{\text{ref}}$} on Figure \ref{fig:Reward} is equal to $1.1\cdot10^3$\,b€.
 This value comes from the mean of the total transition costs obtained through the \gls{GSA} performed on the perfect foresight transition pathway optimisation (see Section \ref{subsec:atom_mol:results_uq_cost}).
-In this final form of the reward, one will notice that overshooting cumulative emissions are more penalising than an overshooting transition cost, \ie weight of 200 for the emissions versus 100 for the cost.
+In this final form of the reward, one will notice that overshooting cumulative emissions is more penalising than an overshooting transition cost, \ie a weight of 200 for the emissions versus 100 for the cost.
 The values of these weights are the results of a trial and error to fine-tune the balance between more expensive successes and cheaper failures.
@@
 
-Besides the reward, states are the other piece of information provided by the environment to the agent.
-In \gls{RL}, the purpose of states are to represent the current situation or configuration of the environment in which the agent operates.
+Besides the reward, the states are the other piece of information provided by the environment to the agent.
+In \gls{RL}, the purpose of states is to represent the current situation or configuration of the environment in which the agent operates.
 The primary function of states in RL is to provide the necessary context for the agent to choose appropriate actions based on its current observations and goals \cite{sutton2018reinforcement}.
@@
 
-Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimension state space characterizing the energy system at the end of the optimised time window.
+Consequently, after testing several state spaces and observing the convergence of the reward, we have converged to a four-dimensional state space characterizing the energy system at the end of the optimised time window.
 The first dimension is directly related to the main objective of the agent: respecting the \ce{CO2} budget until 2050.
@@
 
-We end this section with justifications of actions, reward and states implemented in this work as well as guidelines for researchers that would like to apply the RL approach on their own case study with their own model.
+We end this section with justifications of actions, reward and states implemented in this work as well as guidelines for researchers who would like to apply the RL approach on their own case study and/or with their own model.
 
@@
 \caption{Mean reward and success rate of the different learning batches.
-The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view.
+The stabilisation of the reward curve shows a convergence of the learning process from the agent's point of view.
 The evolution of the success rate also shows that the reward function aims at more and more successful transitions.}
@@
 \newpage
-When assessing the distributions of the values of reward in the failure and success cases, one notices a range where these distributions overlap (see left hand side of Figure \ref{fig:reward_status}).
+When assessing the distributions of the values of reward in the failure and success cases, one notices a range where these distributions overlap (see left-hand side of Figure \ref{fig:reward_status}).
 This area corresponds to either transitions that exceed the \ce{CO2} budget in 2050 but are cheaper than the total transition cost of reference (see Section \ref{sec:RL:act_states_rew}) or successful transitions that are more expensive.
-Besides this overlap, we observe that successes account for the majority of the cases with higher reward.
+Besides this overlap, we observe that successes account for the majority of the cases with higher rewards.
 This is another indication that the reward function is appropriate in this exploration of successful transition pathways.
@@
 
-Considering the end of the time window where the \ce{CO2} budget is exceeded, the right hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
+Considering the end of the time window where the \ce{CO2} budget is exceeded, the right-hand side of Figure \ref{fig:reward_status} shows that 2040 is the ``tipping year'' for the agent.
 Beyond this point, through this learning process, the chances to succeed the transition were 38\%.
-In other words, near-term (2025-2030) actions are necessary to hope succeeding the transition.
+In other words, near-term (2025-2030) actions are necessary to hope to succeed the transition.
 
@@
 \caption{Reward distribution between successes and failures.
-Graph on the right hand side details at the end of which time window the failure occurred.
+Graph on the right-hand side details at the end of which time window the failure occurred.
 The ``tipping year'' is 2040 as failing the transition by 2040 represents 57\% of all the failures.
@@
 Per definition, the threshold of 1.2\,Gt$_{\ce{CO2},\text{eq}}$ splits the episodes reaching 2050 into successes and failures (see Figure \ref{fig:Cum_gwp_cost}).
-Since infeasible cases or those that overshoot the \ce{CO2} budget are discarded before reaching 2050 (see Section \ref{subsec:RL:act_states_rew:rew}), the number of attempts that reach further steps in the transition progressively decreases.
+Since infeasible cases or those that overshoot the \ce{CO2} budget are discarded before 2050 (see Section \ref{subsec:RL:act_states_rew:rew}), the number of attempts that reach further steps in the transition progressively decreases.
 Consequently, the share of successful transitions compared to failures progressively increases with time.
 In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
-Reaching cumulative emissions significantly lower than the \ce{CO2} budget are possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
+Reaching cumulative emissions significantly lower than the \ce{CO2} budget is possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
 Considering the failures in 2050, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
@@
 \caption{Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+The number of remaining attempts decreases with time since the infeasible problems and the solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before 2050.
 Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
@@
 On the contrary, the cumulative cost of successful transitions has a positive skewness: the agent successfully reduces the cost of the system as a secondary objective with 30\% of the cases above the reference transition cost.
-This hierarchy of agent's objectives is verified with the failures.
+The hierarchy of the agent's objectives is verified with the failures.
 When it failed the transition, the agent aimed at reducing the emissions (skewness of 0.61) before minimising the total transition cost (skewness of 0.24).
@@
 Given the skewness that is positive and negative for the electrofuels and the fossil fuels, respectively, these cases represent more than the majority of the successful cases.
-On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and discount rate.
+On top of this, total transition costs of successful episodes are lower due to lower industrial \gls{EUD} and discount rates.
 These favourable conditions combined with the right agent's actions led to transitions respecting the \ce{CO2} budget.
@@
 The share of renewable energy carriers in the primary mix allows identifying intermediate milestones along successful transitions (see Figure \ref{fig:RE_in_mix_Efficiency}).
-From the initial state of 10\% in 2020, a boost of integration of renewables in the near-term is needed to hope for a successful transition.
+From the initial state of 10\% in 2020, a boost of integration of renewables in the near term is needed to hope for a successful transition.
 For the successful occurrences to exceed failures, this share increases to 54\% in 2025.
 Along the transitions, this increase goes with the import of electrofuels and the full deployment of local \gls{VRES}.
-In 2050, the threshold where successes occur more often than failures was at 82\% of renewable share.
+In 2050, the threshold where successes occur more often than failures was at 82\% renewable share.
 In the REF case of Chapter \ref{chap:atom_mol}, this share reached 86\% by 2050.
-However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another ``bump'' at lower share of renewables in the mix.
-This area corresponds to the possibility to install \gls{SMR}.
+However, by 2050, Figure \ref{fig:RE_in_mix_Efficiency} shows another ``bump'' at lower shares of renewables in the mix.
+This area corresponds to the possibility of installing \gls{SMR}.
 As uranium is considered as a non-renewable resource \cite{rixhon2021terminology}, installing \gls{SMR} allows lowering the threshold as in the SMR case of Chapter \ref{chap:atom_mol}.
@@
 Through the transition, besides the share of success increasing over the failures, the distributions of success and failure indistinguishably spread over the whole range.
-Similarly than the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.
+Similarly to the emissions, we observe a bump at lower efficiencies by 2050 due to the installation of \gls{SMR}.
 
@@
 \caption{Exploration of the state space over the learning process: distribution of occurrence of share of renewable energy carriers in the primary energy mix (left) and efficiency (right).
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+The number of remaining attempts decreases with time since infeasible problems and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before 2050.
 Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
@@
 Below a near-term threshold ($\sim$60\%), the chances of success are limited, \ie no-go zones.
-Efficiency is a less valuable information for the agent to succeed transitions as failures and successes indistinguishably spread over the whole range.}
+Efficiency is less valuable information for the agent to succeed transitions as failures and successes indistinguishably spread over the whole range.}
 \label{fig:RE_in_mix_Efficiency}
@@
 The more the agent was able to progress through transition, without exceeding the \ce{CO2} budget, the bigger is the share of successes compared to failures.
-Besides this observation, no specific range of values for the different actions at the different timing seems to lead to more successes.
+Besides this observation, no specific range of values for the different actions at the different timings seems to lead to more successes.
 Looking at action individually, there does not seem to be any that supports more effectively the transition.
@@
 \caption{Exploration of the action space over the learning process: distribution of occurrence of the different actions at the time they are taken by the agent.
-Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+The number of remaining attempts decreases with time since infeasible problems and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before 2050.
 Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.
-Besides this wide exploration, successes and failures indistinguishably spread over the whole ranges of actions.
-In other words, it does not seem to be any clear set of actions to support successful transitions.}
+Besides this wide exploration, successes and failures indistinguishably spread over the whole range of actions.
+In other words, there does not seem to be any clear set of actions to support successful transitions.}
 \label{fig:Actions_learning}
@@
 In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}.
-The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal.
+The constraints intersecting at this vertex are considered binding, actually limiting the objective function to be more optimal.
 In other words, binding constraints, when tightened, aggravate the objective value function.
-If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.
+If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that the left and right sides of the equations are equal.
 
@@
 The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution.
-These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
+These constraints are considered binding, \ie having a limiting impact on the optimal solution.}
 \label{fig:Binding_constr}
@@
 This allows identifying key actions to support the myopic transition.
-Limiting the \gls{GWP} in the near-term is a key-factor for success.
+Limiting the \gls{GWP} in the near term is a key factor for success.
 However, this action has a binding effect on the environment only at the end of the transition.
@@
 Sweet spots can be identified when considering the limits of \gls{GWP} and fossil gas consumption.
-Limiting coal consumption is always constraining, unlike \gls{LFO} that is ``naturally'' substituted by EnergyScope Pathway in the near-term.}
+Limiting coal consumption is always constraining, unlike \gls{LFO} which is ``naturally'' substituted by EnergyScope Pathway in the near term.}
 \label{fig:Binding_learning}
@@
 When it comes to limiting the use of \gls{LFO} and coal, the conclusions are more straightforward.
-At the beginning of the transition, most of the 159\,TWh of \gls{LFO} are consumed by naphtha-cracker (46\%) and decentralised boilers (45\%).
+At the beginning of the transition, most of the 159\,TWh of \gls{LFO} are consumed by naphtha-crackers (46\%) and decentralised boilers (45\%).
 The remaining 10\% are consumed by industrial boilers.
@@
 In other words, the cost-driven environment will favour it.
-Then, as the maximum amount of coal (28\,TWh) is much smaller than fossil gas and \gls{LFO}, high value of $\mathrm{act}\textsubscript{coal}$ still represents small consumption of coal.
+Then, as the maximum amount of coal (28\,TWh) is much smaller than fossil gas and \gls{LFO}, high values of $\mathrm{act}\textsubscript{coal}$ still represent small consumptions of coal.
 Whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
@@
 As introduced in Chapter \ref{chap:chap_methodo}, most of the work in applying \gls{RL} is the definition of the interactions between the agent and its environment (\ie actions, reward and states) that are very dependent on the case study and the research questions to answer.
-The elements presented in this work result from several trial and errors to end up with meaningful results according to our research questions.
+The elements presented in this work result from several trials and errors to end up with meaningful results according to our research questions.
 
@@
 In this work, we have investigated other actions like incentivising solar \gls{PV} panels and wind turbines by ``artificially'' reducing their \gls{CAPEX}.
-The result was not conclusive as these technologies need to take part to the Belgian energy transition and because it was harder to assess the impact of this action.
+The result is not conclusive as these technologies need to take part in the Belgian energy transition and because it was harder to assess the impact of this action.
 
-The reward function was designed to first aiming at respecting the \ce{CO2} budget then minimising the total transition cost.
+The reward function was designed to first aim at respecting the \ce{CO2} budget and then minimising the total transition cost.
 The -300 penalty given in case of an infeasible optimisation problem was arbitrarily set.
 \textit{A posteriori}, it seems to be a well-defined penalty given its significant relative difference with the values taken by the reward otherwise, between -120 and 44 (see Figure \ref{fig:reward_status}).
-Besides this penalty and given the observed results, we recommend to start with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
-This requires to define the \ce{CO2} budget according to a certain sharing principle (see Section \ref{sec:cs:CO2-budget}) and to compute reference total transition cost.
-However, other researches focusing on reaching carbon neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
+Besides this penalty and given the observed results, we recommend starting with the same reward function if the objective is similar, first target cumulative emissions then cumulative costs.
+This requires defining the \ce{CO2} budget according to a certain sharing principle (see Section \ref{sec:cs:CO2-budget}) and computing the reference total transition cost.
+However, other research focusing on reaching carbon neutrality by 2050 could define a binary reward function as +1 for reaching the objective and -1 otherwise.
 
 Finally, states aim at representing the information relevant to the agent to efficiently learn and progress through the transitions.
-For this reason, on top reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assessing their policies to reach their targets (share of renewables in the mix and the system overall efficiency).
+For this reason, on top of reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assess their policies to reach their targets (share of renewables in the mix and the overall efficiency of the system).
 For other studies, one might consider other information like the metrics considered by \citet{pickering2022diversity} (\eg heat electrification, average national import or level of curtailment).
@@
 The comparison between the failures and the successes demonstrates the added-value brought by myopic pathway optimisation.
-In the near-term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed.
+In the near term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed.
 This shows that limited foresight encourages to strongly act at the early stages.

### 2024-07-19 chap_atom_mol/chap_atom_mol.tex
Grammarly up to 4.2 (included)

@@
 
-On top of scenarios with more profound behavioural changes, variety of technological pathways are often investigated to meet the ambitions of climate change mitigation.
-For instance, in their work, \citet{My2050} assessed different scenarios for a climate neutral Belgium by 2050.
+On top of scenarios with more profound behavioural changes, a variety of technological pathways are often investigated to meet the ambitions of climate change mitigation.
+For instance, in their work, \citet{My2050} assessed different scenarios for a climate-neutral Belgium by 2050.
 Depending on the scenarios, the emphasis is put on a higher electrification of the whole-energy system, a higher consumption of hydrogen or more complex molecules (\ie electrofuels) or a bigger reliance on biomass and the related \gls{BECCS}.
@@
 In the future, deeper integration of \gls{VRES} will come with the bigger electrification of sectors like mobility, low-temperature and industry \cite{IEA2023electrification}.
-In such a context, there is a case for nuclear energy to produce a reliable low-\ce{CO2}emission electricity \cite{IAEA2008}.
-This is also in line with the willingness to phase out of imported Russian fossil fuels anchored in the European REPowerEU Plan \cite{REPowerEU}.
-Unlike the consumption of fossil gas that represents about 80\% of the \gls{LCOE} from \gls{CCGT}, the price of uranium accounts for 25-30\% of \gls{LCOE} from nuclear power plants.
+In such a context, there is a case for nuclear energy to produce reliable low-\ce{CO2} emission electricity \cite{IAEA2008}.
+This is also in line with the willingness to phase out imported Russian fossil fuels anchored in the European REPowerEU Plan \cite{REPowerEU}.
+Unlike the consumption of fossil gas which represents about 80\% of the \gls{LCOE} from \gls{CCGT}, the price of uranium accounts for 25-30\% of \gls{LCOE} from nuclear power plants.
 This favours nuclear energy against price volatility and for the security of electricity supply, in case of a conflict like the invasion of Ukraine, for instance.
@@
 In their work, \citet{lefebvre2022electrofuel} investigated this topic for the case of Belgium considering lower and upper bounds in terms of availability based on, on the one hand, the already signed agreements with the exporting countries and, on the other hand, their maximal technological potential of \gls{VRES}, respectively.
-Their lower bound resulted in a total availability of electrofuels that is one order of magnitude lower than the needs provided by the cost-optimum carbon neutral Belgium in 2050.
-Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} as well as the need of importing electrofuels in the Belgian energy landscape, on a strictly techno-economical point of view with a cost-based optimisation.
-This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in the Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
+Their lower bound resulted in a total availability of electrofuels that is one order of magnitude lower than the needs provided by the cost-optimum carbon-neutral Belgium in 2050.
+Like the rest of this thesis, the purpose here is only to expose the impact of integrating \gls{SMR} as well as the need to import electrofuels in the Belgian energy landscape, from a strictly techno-economical point of view with a cost-based optimisation.
+This is why, for the sake of transparency, the model and the data are documented and openly available online \cite{readthedocs_pathway} and in Appendices \ref{app:EnergyScope} and \ref{app:case_study}.
 
@@
 
-In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimise a transition pathway with a small number of time stages, \ie two or three.
+In their review, \citet{yue2018review} pointed out that uncertainties were accounted for either snapshot models (\ie optimizing a single target future year), to assess a single energy sector (\ie the power sector, most of the time) or with a limited number of uncertain parameters, \ie about ten, in a stochastic programming approach to optimise a transition pathway with a small number of time stages, \ie two or three.
 Here, the uncertainty quantification addresses the total transition pathway of the whole-energy system, with 34 uncertain parameters (see Section \ref{sec:cs:uncertainty}).
@@
 \label{sec:atom_mol:results_deter} 
-In this section, like in the rest of the thesis, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition and whereas in the \textbf{SMR} case this technology is available, up to 6\,GW (see Chapter \ref{chap:case_study}), from 2040 onward.
+In this section, like in the rest of the thesis, the \textbf{REF} case is without any deployment of \gls{SMR} anytime during the transition whereas in the \textbf{SMR} case, this technology is available, up to 6\,GW (see Chapter \ref{chap:case_study}), from 2040 onward.
 After investigating the deployment of \gls{SMR} through the power sector, the first part of this section focuses on this impact on the whole-system level considerations (\ie overall transition costs, primary energy mix and yearly emissions per sector).
@@
 \label{subsec:atom_mol:results_deter_power_sector}
-\gls{SMR} is deployed as soon as available, \ie 2040, to their maximum capacity, 6\,GW, substituting other flexible power generation units (see \Cref{fig:results_deter_tech_cap_elec}).
+\gls{SMR} is deployed as soon as available, \ie 2040, to its maximum capacity, 6\,GW, substituting other flexible power generation units (see \Cref{fig:results_deter_tech_cap_elec}).
 There is no ammonia-\gls{CCGT} at the end of the transition and an anticipatory reduction of methane \gls{CCGT} (\ie 2.1\,GW in 2040 versus 3.7\,GW for the REF case).
@@
 \caption{Installed capacities in the power sector in the REF (top) and SMR (down) scenarios.
-As soon as available (\ie 2040), \acrfull{SMR} is deployed to their maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie methane and ammonia \gls{CCGT}).}
+As soon as available (\ie 2040), \acrfull{SMR} is deployed to its maximum potential (\ie 6\,GW) to substitute more expensive flexible generation units (\ie methane and ammonia \gls{CCGT}).}
 \label{fig:results_deter_tech_cap_elec}
@@
 Given the 15\% maintenance off-time assumed in this work, this represents 44.6\,TWh by 2050.
-In comparison, in 2020, conventional nuclear power plants produced 34.4\,TWh Belgium.
+In comparison, in 2020, conventional nuclear power plants produced 34.4\,TWh in Belgium.
 This resurgence of nuclear electricity occurs at the expense of other, although more efficient, technologies: \gls{CCGT} and industrial \gls{CHP}.
-Besides the unchanged end-use-demand compared with the REF case, we observe a slight increase of the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.
+Besides the unchanged end-use-demand compared with the REF case, we observe a slight increase in the electrification of the rest of the system: +9.4\% which corresponds to +5.8\,TWh, mostly consumed by electric heaters in industry (+48\%) to produce industrial high-temperature heat.
 
@@
 
-Extending the use of naphtha-cracker and, to a lesser extent, postponing the deployment of PV, lead to higher \ce{CO2}-emissions at the beginning of the transition that are then compensated by the deployment of \gls{SMR} (\Cref{fig:GWP_per_sector_diff}).
+Extending the use of naphtha-crackers and, to a lesser extent, postponing the deployment of PV, lead to higher \ce{CO2} emissions at the beginning of the transition that are then compensated by the deployment of \gls{SMR} (\Cref{fig:GWP_per_sector_diff}).
 
@@
 (Down) Difference of the mix between the two cases after aggregation per category.
-In line with \citet{rixhon2021terminology},uranium is considered as a non-renewable resource.
+In line with \citet{rixhon2021terminology}, uranium is considered as a non-renewable resource.
 It is nevertheless dissociated from other non-renewable resources for the sake of clarity.
@@
 The main impact of including \gls{SMR} from 2040 onward on the high-temperature heat sector is (i) its higher direct electrification and (ii) the reduction of more efficient heat-and-power co-generation in the benefit of single-output industrial boilers.
-In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
+In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, on sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
 With \gls{SMR}, by 2050, an additional 1.7\,GW (+13\%) of these heaters can rely on a more constant supply of defossilised electricity.
@@
 In both cases, the major shift of supply from decentralised to centralised productions operates early in the transition, to hit the constraint that \gls{DHN} cannot supply more than 37\% of the \gls{LT}-heat production.
-Then, from a mix between oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system shifts towards electric \gls{HP} only by 2035.
-Similarly for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economic option.\\
+Then, from a mix of oil (53\%), gas (43\%) and wood (4\%) boilers for the decentralised production of \gls{LT}-heat in 2020, the system shifts towards electric \gls{HP} only by 2035.
+Similarly, for the centralised production of \gls{LT}-heat, electric \gls{HP} remains the most efficient and economical option.\\
 
@@
 
-The passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} substituting \gls{ICE} cars for the private sector by 2030.
-About the public mobility, trains and tramways supply their \textit{a priori} set maximum share, respectively, 50\% and 30\% complemented by \gls{CNG} buses substituting diesel-driven buses.
+Passenger mobility is not affected either as the electrification of the system is preferentially done in this sector with \gls{BEV} substituting \gls{ICE} cars for the private sector by 2030.
+Regarding public mobility, trains and tramways supply their \textit{a priori} set maximum share, respectively, 50\% and 30\% complemented by \gls{CNG} buses substituting diesel-driven buses.
 Similarly, considering the freight transport, technology shifts (\ie from diesel to \gls{FC} trucks) or modal shares (\ie 53\%-47\% split between \gls{NG} and (bio)-diesel boats) are identical between the two cases.\\
@@
 After briefly assessing the \acrfull{GSA} on the total transition cost (Section \ref{subsec:atom_mol:results_uq_cost}), this section investigates more deeply the atom-molecules dilemma (Section \ref{subsec:atom_mol:results_atom_mol}).
-The Sobol' indices are computed for import of renewable molecules and installed capacities of \gls{SMR}.
+The Sobol' indices are computed for the import of renewable molecules and installed capacities of \gls{SMR}.
 
@@
 \Cref{tab:UQ_short} shows that the cost of purchasing electrofuels is the most impacting parameter.
-On the contrary, the potentiality to install \gls{SMR} and its CAPEX have much lower influence on the variation of the total transition cost.
+On the contrary, the potentiality to install \gls{SMR} and its CAPEX have a much lower influence on the variation of the total transition cost.
 Given the uncertainty characterisation presented in Section \ref{sec:cs:uncertainty}, there are 60\% chance that no \gls{SMR} could be installed.
-When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
+When $f_{\mathrm{max,SMR}}$ takes a value below 0.6, the variation of the total transition cost is only due to the variation of the other parameters.
 Then, in perspective with the scenario analysis of Section \ref{sec:atom_mol:results_deter}, the 3.3\% reduction has been observed when \gls{SMR} is installed from 2040 onward.
 This represents only 10\% of the samples.
-Moreover, when it is possible, \gls{SMR} is always installed thanks to its characteristics: cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.
+Moreover, when it is possible, \gls{SMR} is always installed thanks to its characteristics: cheap and low-emitting fuel, a long lifetime leading to lower annualised CAPEX and higher salvage value.
 However, given its limited deployment, up to 6\,GW, \gls{SMR}-related parameters have a lower impact on the variation of the total transition cost.
@@
 \caption{Total Sobol' indices of the uncertain parameters over the total transition cost.
-Where the cost of purchasing electrofuels is the top-1 parameter, \gls{SMR}-related parameters have a negligigble impact on this cost.}
+Where the cost of purchasing electrofuels is the top-1 parameter, \gls{SMR}-related parameters have a negligible impact on this cost.}
 \label{tab:UQ_short}
@@
 
-Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share among the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
+Given the relatively wide uncertainty range (\ie up to [-30.8\%; +24.0\%] by 2050) and, above all, the major share of the total demand, between 53\% and 60\%, the industrial \gls{EUD} is the second most impacting parameter.
 
@@
 
-E-hydrogen becomes rapidly the main stream of hydrogen in the system, to reach a median and a maximum values of, respectively 13.0\,TWh and 42.1\,TWh in 2050.
+E-hydrogen becomes rapidly the main stream of hydrogen in the system, reaching median and maximum values of 13.0\,TWh and 42.1\,TWh in 2050, respectively.
 Hydrogen is more frequently used in the mobility sector.
@@
 Then, the imported e-ammonia rapidly becomes cost-competitive against its fossil alternative (see Chapter \ref{chap:case_study}).
-At early stage of the transition, e-ammonia substitutes fossil ammonia and the Haber-Bosch process.
+At the early stage of the transition, e-ammonia substitutes fossil ammonia and the Haber-Bosch process.
 Where the initial purpose of ammonia is to satisfy a relatively limited \acrfull{NED} (\ie 10 $\pm$ 3\,TWh by 2050), the variation of its import is mostly due to the higher or lower need for ammonia-\gls{CCGT} as a flexible option to produce electricity.
@@
 It is an indicator of statistical dispersion around the median, $Q2$ or $P_{50}$.} of about 50\,TWh.
-In some extreme cases, e-ammonia is the most imported molecules, \ie up to 167\,TWh or 45\% of the total primary mix.
+In some extreme cases, e-ammonia is the most imported molecule, \ie up to 167\,TWh or 45\% of the total primary mix.
 
-Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand in methanol.
+Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand for methanol.
 Its \gls{NED} represents, on average, 3\% of the total consumption of methanol.
@@
 
-In relation to e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector (Figure \ref{fig:results_uq_samples_H2}).
+About e-hydrogen, the sensitivity analysis highlights its dependence on various driving parameters, particularly those linked to the transport sector (Figure \ref{fig:results_uq_samples_H2}).
 The utilisation of e-hydrogen is most prevalent in \gls{FC} trucks, followed by \gls{FC} cars and buses to a lesser extent (Appendix \ref{app:UQ_electrofuels}).
 The adoption of fuel cell engines in trucks contributes, on average, to 63.5\% of the total road freight transport, thereby affecting the level of e-hydrogen imports.
-Consequently, the smaller is the CAPEX of fuel cell engines, the more the system imports e-hydrogen.
+Consequently, the smaller the CAPEX of fuel cell engines, the more the system imports e-hydrogen.
 Similarly, the cost of purchasing electrofuels influences e-hydrogen imports.
 Subsequently, the cost of purchasing biofuels emerges as the third most influential parameter.
-Indeed, biodiesel trucks are the mostly picked alternative to \gls{FC} trucks to provide, on average, 27.6\% of the total.
+Indeed, biodiesel trucks are the most picked alternative to \gls{FC} trucks to provide, on average, 27.6\% of the total.
 Additionally, \gls{CNG} buses are preferred in public road transport (34.9\%), followed by \gls{FC} buses (11.2\%) competing with biodiesel and hybrid biodiesel buses, accounting for 27.8\% and 26.1\%, respectively.
@@
 This confirms the observation made in Section \ref{subsec:atom_mol:results_deter_others}.
-Highly available local biomass also leads to smaller import of e-methane to supply bio-hydrolysis and produce methane-equivalent gas.
-Finally, and surprisingly, costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively.
-In other words, by 2050, more expensive electrofuels result in more e-methane import.
-On the contrary, cheaper electrofuels result in higher import of fossil methane.
+Highly available local biomass also leads to smaller imports of e-methane to supply bio-hydrolysis and produce methane-equivalent gas.
+Finally, and surprisingly, the costs of purchasing electrofuels and fossil fuels have a positive and negative correlation with the amount of e-methane, respectively.
+In other words, by 2050, more expensive electrofuels result in more e-methane imports.
+On the contrary, cheaper electrofuels result in higher imports of fossil methane.
 Given the techno-economic optimum sought by EnergyScope, if electrofuels are more expensive, the system will, overall, import less of them, especially e-ammonia, mainly used by \gls{CCGT}.
@@
 First running on fossil gas, these \gls{CHP} consume more e-methane by 2050.
-On the contrary, if electrofuels are cheaper, there is more import of them, and especially of e-ammonia.
+On the contrary, if electrofuels are cheaper, there is more import of them, especially of e-ammonia.
 This leaves room for more emitting and cheaper resources to be used while respecting the \ce{CO2} budget, \ie coal in industrial boilers that produce, in these cases, on average 24\% of the high-temperature (HT) heat in 2050.
@@
 Even though we might expect that no more coal will be consumed in Belgium by 2050, the model still has the opportunity to use it if the \ce{CO2} budget allows it.
-About the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system.
+Regarding the cost of purchasing fossil fuels, the parameter has mainly an impact on the import of fossil \gls{NG} as the most versatile energy carrier in the whole-energy system.
 If \gls{NG} is more expensive, the system will import less of it.
 Subsequently, the investments in methane-\gls{CHP} and boilers are more limited.
-This ends up in smaller need for e-methane by 2050.
+This ends up in a smaller need for e-methane by 2050.
 
@@
 In practice, we observe that as soon as this parameter is equal or higher than 0.9, 0.8 and 0.6, 6\,GW \gls{SMR} is installed from 2040, 2045 and 2050, respectively.
-Large range (-40\%; +44\%) of its CAPEX has a negligible impact on the installed capacity, with a Sobol' index of 0.9\%.
+The large range of uncertainty of its CAPEX (-40\%; +44\%) has a negligible impact on the installed capacity, with a Sobol' index of 0.9\%.
 This is explained by the long lifetime of the technology (60 years) and the salvage value recovered by 2050.
@@
 \label{subsec:atom_mol:results_uq_VRES}
-In line with results given in Section \ref{sec:atom_mol:results_deter}, the \gls{GSA} shows that \gls{SMR} has negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines).
+In line with the results given in Section \ref{sec:atom_mol:results_deter}, the \gls{GSA} shows that \gls{SMR} has a negligible impact on the deployment of local \gls{VRES} (\ie \gls{PV}, onshore and offshore wind turbines).
 Whereas the installed capacities of onshore wind is totally driven by the uncertainty on its maximum potential, Sobol' indices change for the most impacting parameters on the deployment of \gls{PV} and offshore wind (see Figure \ref{fig:results_uq_pdf_local_ren}).
 The key factor that drives the installed capacities of these two technologies is mostly their respective maximum potential, especially at the end of the transition, much more than their CAPEX.
-Given its higher \gls{LCOE} (\Cref{fig:LCOE}), \gls{PV} is more impacted in the short-term by the variation on the cost of purchasing electrofuels supplying e-methane (and e-ammonia to a lesser extent) \gls{CCGT}.
-However, this impact gets negligible by 2050.
+Given its higher \gls{LCOE} (\Cref{fig:LCOE}), \gls{PV} is more impacted in the short term by the variation in the cost of purchasing electrofuels supplying e-methane (and e-ammonia to a lesser extent) \gls{CCGT}.
+However, this impact becomes negligible by 2050.
 
@@
 \includegraphics[width=0.49\textwidth]{Sobol_WIND_OFFSHORE.pdf}
-\caption{Impacting parameters on the deployment of \gls{PV} and offshore wind over the transition.
+\caption{Parameters impacting the deployment of \gls{PV} and offshore wind over the transition.
 Progressively, the impact of the uncertainty on the maximum potential increases, unlike the one on the cost of purchasing electrofuels.}
@@
 
-Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+Given the ambitious \ce{CO2} budget (\ie a 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
+Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solutions for the case of Belgium is the import of renewable electrofuels, even at the early stages of the transition.
 This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
-On the contrary, parameters directly related to \gls{SMR} (\ie its availability and its CAPEX) have limited impact, below 1\%.
+On the contrary, parameters directly related to \gls{SMR} (\ie its availability and its CAPEX) have a limited impact, below 1\%.
 This \gls{GSA} also points out the key drivers for the import of renewable electrofuels and the installation of \gls{SMR} by 2050.
-Besides the cost of purchasing electrofuels (\ie the lower this cost, the bigger the imports), it shows that available \gls{SMR} would have mainly a direct impact on e-ammonia, by substituting the ammonia \gls{CCGT} that is the biggest consumer of this molecule.
+Besides the cost of purchasing electrofuels (\ie the lower this cost, the bigger the imports), it shows that available \gls{SMR} would have mainly a direct impact on e-ammonia, by substituting the ammonia \gls{CCGT} which is the biggest consumer of this molecule.
 Indirectly, this parameter will also reduce the import of e-methane by reducing the need for gas \gls{CHP} and gas boilers.
@@
 In conclusion, this work puts under the spotlight the ``competition'' between \gls{SMR} and imported electrofuels while both of them support the integration of local \gls{VRES}.
-Betting on \gls{SMR} means letting the emissions go up at early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are 3.3\% and 8.8\% cheaper than in the REF case, respectively.
+Betting on \gls{SMR} means letting the emissions go up at the early stages of the transition (\ie \gls{LFO} still used in 2025 to produce \gls{HVC}) to end up with an overall transition and a system by 2050 that are 3.3\% and 8.8\% cheaper than in the REF case, respectively.
 However, given the need to import molecules at earlier stages of the transition, with or without \gls{SMR}, it seems reasonable to keep on investing in the transport and the infrastructure to support these imports.

### 2024-07-19 options/symbols_and_cmd.tex
(Almost) all comments from FC on atom_molecules review
Still have to implement the one about the graphs on UQ and parameters of molecules and SMR
Rebuttals: Keep going...
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2
Grammarly up to 4.2 (included)

@@
 \newabbreviation[group={Acronyms}]{BEMS}{BEMS}{Building Energy Management System}
-\newabbreviation[group={Acronyms}]{BEV}{BEV}{Battery Electric Vehicle}
+\newabbreviation[group={Acronyms}]{BEV}{BEV}{Battery Electric Vehicles}
 \newabbreviation[group={Acronyms}]{BTX}{BTX}{Benzene, Toluene and Xylene}
@@
 \newabbreviation[group={Acronyms}]{ESOMs}{ESOMs}{Energy System Optimisation Models}
+\newabbreviation[group={Acronyms}]{ESTD}{ESTD}{EnergyScope Typical Days}
 \newabbreviation[group={Acronyms}]{EUD}{EUD}{End-Use Demand}
@@
 \newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
-\newabbreviation[group={Acronyms}]{PDF}{PDF}{Probability Density Function}
 \newabbreviation[group={Acronyms}]{PV}{PV}{Photovoltaic}
@@
 %% S
-\newabbreviation[group={Acronyms}]{SAC}{SAC}{Soft Actor Critic}
+\newabbreviation[group={Acronyms}]{SAC}{SAC}{Soft Actor-Critic}
 \newabbreviation[group={Acronyms}]{SMR}{SMR}{Small Modular Reactor}

### 2024-07-24 chap_RL/chap_RL.tex
Grammarly done for the core of the text

@@
 
-To navigate through the myopic transitions not constrained by a prescribed \ce{CO2}-trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium.
+To navigate through the myopic transitions not constrained by a prescribed \ce{CO2} trajectory and investigate the efficiency of different policies, we apply the reinforcement learning approach detailed in Section \ref{sec:meth:RL} on the case study of Belgium.
 Besides the environment, \ie the myopic transition pathway of the whole-energy system via EnergyScope Pathway (see Chapter \ref{chap:chap_methodo}), the first part of this chapter presents the three key features of interaction between the agent, optimizing its policy, and the environment: actions, states and reward.
@@
 
-Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added-value brought by this approach in the exploration of myopic transition pathways.
+Comparing the \gls{RL}-based results with more conventional approaches, \ie perfect foresight optimisation, highlights the added value brought by this approach in the exploration of myopic transition pathways.
 
@@
 To reduce the computational time, the monthly model could also be used to carry out \gls{RL}-based exploration of myopic pathways.
-Appendix \ref{app:app_RL_TD_MO} presents the comparison between the results obtained from the learning process on the hourly and monthly models.
+Appendix \ref{app:app_RL_TD_MO} compares the results obtained from the learning process on the hourly and monthly models.
 
-\gls{RL}-based myopic optimisation provides \ce{CO2}-emissions pathways different from the perfect foresight approach to respect the same \ce{CO2} budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
+\gls{RL}-based myopic optimisation provides \ce{CO2} emissions pathways different from the perfect foresight approach to respect the same \ce{CO2} budget (see left side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
 However, driven first by this \ce{CO2} budget, the agent often reaches much lower cumulative emissions when succeeding the transition (see Figure \ref{fig:Cum_gwp_cost}).
 This comes from the agent's actions that limit the emissions and/or the consumption of fossil resources at the early stages.
-Thanks to the bigger reduction of emissions at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2}-buffer'' at the end of the transition.
+Thanks to the bigger emission reduction at these early stages, the \gls{RL}-based optimisation can benefit from a ``\ce{CO2} buffer'' at the end of the transition.
 This buffer is compensated by the end of the transition where 50\% of the myopic transitions reach 2050 with 10 or more remaining Mt$_{\ce{CO2},\text{eq}}$ compared to 4 for the perfect foresight approach.
-These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal that account for 3.5\% and 2.4\% of the primary mix on average by 2050.
-Finally, the long-term vision of the perfect foresight approach results in smoother reduction of the emissions to end up with less emissions by 2050.
+These remaining emissions by 2050 come from the consumption in industrial boilers of waste and coal accounting for 3.5\% and 2.4\% of the primary mix on average by 2050.
+Finally, the long-term vision of the perfect foresight approach results in a smoother reduction of emissions to end up with less emissions by 2050.
 
-The comparison between the failures and the successes demonstrates the added-value brought by myopic pathway optimisation.
+The comparison between the failures and the successes demonstrates the added value brought by myopic pathway optimisation.
 In the near term (2025-2030), levels of emission are similar between perfect foresight and myopic cases that have failed.
 This shows that limited foresight encourages to strongly act at the early stages.
-On top of this, following the initial steps of \ce{CO2}-emissions pathways resulting from PF approach would likely ($\sim$80\%) lead to fail the transition.
+On top of this, following the initial steps of \ce{CO2} emissions pathways resulting from the PF approach would likely ($\sim$80\%) lead to failure of the transition.
 
 Looking at the total transition cost, the combination of the agent's actions and favourable economic conditions (see Section \ref{subsec:RL:learning:states}) make the myopic transitions cheaper, on average, than the PF cases (see right side of Figure \ref{fig:Gwp_pathway_total_tran_cost}).
-This is also due to the fact that the perfect foresight approach always finds a solution even in worst conditions such as high cost of purchasing resources and high \gls{EUD}.
+This is also because the perfect foresight approach always finds a solution even in the worst conditions such as the high cost of purchasing resources and high \gls{EUD}.
 This explains the wider variability of the PF results too.
@@
 \includegraphics[height=3.5cm]{Transition_cost_comp_2.pdf}
-\caption{Comparison of \ce{CO2}-emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.
-Myopic transitions succeed with a more drastic reduction of emissions in the short-term and, on average, more favourable economic conditions.}
+\caption{Comparison of \ce{CO2} emissions pathways (left) and total transition cost (right) from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.
+Myopic transitions succeed with a more drastic reduction of emissions in the short term and, on average, more favourable economic conditions.}
 \label{fig:Gwp_pathway_total_tran_cost}
@@
 \includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
-\caption{Comparison of cumulative OPEX (left), CAPEX (center) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
+\caption{Comparison of cumulative OPEX (left), CAPEX (centre) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
 \label{fig:Opex_Capex_Salvage_comp}
@@
 The cost of purchasing the energy carriers represents about 70\% of the total cumulative \gls{OPEX}.
-The assessment of the primary energy mix by 2050 highlights that the difference of OPEX between the perfect foresight and the myopic pathways come from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
+The assessment of the primary energy mix by 2050 highlights that the difference in OPEX between the perfect foresight and the myopic pathways comes from the import of electrofuels, and especially of e-ammonia (see Figure \ref{fig:Mix_2050_comp}).
 In the majority of the cases, e-ammonia is more than two times more imported in the myopic transitions.
 Being cheaper than e-methane (see Chapter \ref{chap:case_study}), e-ammonia brings flexibility in the production of electricity via \gls{CCGT} (see Chapter \ref{chap:atom_mol}).
-Besides the slightly favourable economical conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge in the future, and, among others, the availability of \gls{SMR}.
+Besides the slightly favourable economic conditions (see Table \ref{tab:param_RL}), the myopic optimisations opt to invest massively into importing renewable molecules because of the limited knowledge of the future, and, among others, the availability of \gls{SMR}.
 This explains why 50\% of the successful transitions reached cumulative emissions below 900\,Mt$_{\ce{CO2},\text{eq}}$ (see Figure \ref{fig:Cum_gwp_cost}).
@@
 To add more realism, myopic optimisation considers a sequence of more limited time windows leading to the end of the time horizon \cite{poncelet2016myopic}.
-To respect a \ce{CO2} budget over the transition, this case requires a prescribed \ce{CO2}-trajectory \cite{fais2016impact}.
-However, since the effect of \ce{CO2}-emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself.
+To respect a \ce{CO2} budget over the transition, this case requires a prescribed \ce{CO2} trajectory \cite{fais2016impact}.
+However, since the effect of \ce{CO2} emissions in climate change is cumulative, the total amount of these emissions matters more than the trajectory itself.
 Consequently, we have applied the \acrfull{RL} approach on the environment of the Belgian myopic pathway optimisation under uncertainties.
@@
 
-This \gls{RL}-based exploration pointed out that short-term actions were needed to hope succeeding such a transition, as also demonstrated by \citet{luderer2018residual}.
-Where \gls{LFO} becomes less cost-competitive in the near-future, limiting the use of coal should be done at any cost.
+This \gls{RL}-based exploration pointed out that short-term actions were needed to hope to succeed such a transition, as also demonstrated by \citet{luderer2018residual}.
+Where \gls{LFO} becomes less cost-competitive in the near future, limiting the use of coal should be done at any cost.
 Then, fossil gas should be replaced by e-methane in the mid-term while putting a strict limit on the overall emissions becomes the most effective action by the end of the transition.
-The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances to succeed the transition.
-Below 54\% of renewables in the mix in the near-future, these chances become much more limited, \ie the no-go zone.
+The analysis of the share of renewables in the primary energy mix highlighted intermediate milestones to have higher chances of succeeding the transition.
+Below 54\% of renewables in the mix in the near future, these chances become much more limited, \ie the no-go zone.
 
 We have compared the results coming from this \gls{RL}-based myopic optimisation with the hourly perfect foresight approach.
-These myopic optimisations provided pathways respecting the \ce{CO2} budget that were more drastic in cutting emissions in the short-term than the perfect.
-Supported by the agent's actions, we could reach cheaper energy system than what the perfect foresight could give.
-This comparative analysis also pointed out that investing more into renewable electrofuels was the option when knowledge about the future is limited.
+These myopic optimisations provided pathways respecting the \ce{CO2} budget that were more drastic in cutting emissions in the short term than the perfect foresight.
+This comparative analysis also pointed out that investing more in renewable electrofuels was the option when knowledge about the future is limited.
 To do so, it requires reducing the uncertainties of their cost of purchasing.
@@
 
-In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet-spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2} budget target.
+In conclusion, via the application of the \gls{RL} approach, we have widely explored the different myopic transition pathways and identified sweet spots (and no-go zones) to succeed a transition with an ambitious \ce{CO2} budget target.
 It also highlighted the actions to take to effectively support a whole-energy system transition.

### 2024-07-24 chap_RobPol/chap_RobPol.tex
Grammarly done for the core of the text

@@
 
-\citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}.
-In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economic model, like EnergyScope, that usually assume rational choice within an overall cost minimization.
+\citet{castrejon2020making} assessed the policy mix following the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}.
+In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioural aspects of policy implementation, enriching purely techno-economic models, like EnergyScope, that usually assume rational choice within an overall cost minimization.
 However, in the case of the transition pathway of a whole-energy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''.
-Between the total transition cost giving too few information and the entire set of installed technologies giving too much, the performance metric here is defined through the \gls{PCA} approach.
-Then, when comes the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps.
+Between the total transition cost giving too little information and the entire set of installed technologies giving too much, the performance metric here is defined through the \gls{PCA} approach.
+Then, when it comes to the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps.
 Each roadmap will be ranked in terms of robustness.
@@
 First, we apply the different steps that lead to the principal components of the transition.
-We analyse these big trends of variation and highlight the fact that these variations stand for the entire pathway, a group of consecutive representative years or rather on a tipping-year.
+We analyse these big trends of variation and highlight the fact that these variations stand for the entire pathway, a group of consecutive representative years or rather on a tipping year.
 Then, and most importantly, we assess the robustness of different technological roadmaps by projecting their resulting myopic pathway against these directions of variation.
-The application of \gls{PCA} to provide a new metric for robustness applied to the case of Belgium is the added-value of this chapter.
+The application of \gls{PCA} to provide a new metric for robustness applied to the case of Belgium is the added value of this chapter.
 
@@
 For example, decentralised \gls{HP} running on methane is discarded.
-This technology would have a contribution in the \gls{PCs} of transition for several reasons.
+This technology would have a contribution to the \gls{PCs} of transition for several reasons.
 First of all, the \gls{LT} heating sector represents between 22\% and 27\% of the \gls{EUD} in the case of Belgium.
@@
 Then, given the \gls{LP} implementation of EnergyScope Pathway, the gas heat pumps completely substitute electric heat pumps in specific conditions (less electricity and cheaper methane).
-However, theses cases represent less than 10\% of the 1260 runs.
+However, these cases represent less than 10\% of the 1260 runs.
 For these reasons, we have decided that managing the outliers was needed to assess the main directions of variation of the transition.
@@
 Then, keeping the \gls{PCs} capturing at least 90\% of the total variance of each representative year, this gives between four, in 2035, and seven, in 2050, \gls{PCs} depending on the year (see Figure \ref{fig:PC_over_the_years}), and a total of 34 \gls{PCs}.
-At later stages of the transition, the increasing number of required \gls{PCs}, in line with their smaller share of explained variance, is another indication that the variance of the system design is more spread over a wider range of technologies and with a more limited amplitude.
+In the later stages of the transition, the increasing number of required \gls{PCs}, in line with their smaller share of explained variance, is another indication that the variance of the system design is more spread over a wider range of technologies and with a more limited amplitude.
 
@@
 
-Finally, we consider the respective contribution of the different technologies in the different $\text{PC}_{y}$, \ie their corresponding component in the different eigenvectors.
-Highlighting the top-5 technologies for $\text{PC}_{y,1}$, $\text{PC}_{y,2}$ and $\text{PC}_{y,3}$, we observe general trends over the whole transition as well as tipping year where there is a clear trade-off between several technologies (see Figure \ref{fig:Top5_PC_year}).
+Finally, we consider the respective contributions of the different technologies in the different $\text{PC}_{y}$, \ie their corresponding component in the different eigenvectors.
+Highlighting the top-5 technologies for $\text{PC}_{y,1}$, $\text{PC}_{y,2}$ and $\text{PC}_{y,3}$, we observe general trends over the whole transition as well as the tipping year where there is a clear trade-off between several technologies (see Figure \ref{fig:Top5_PC_year}).
 The top-5 technologies of $\text{PC}_{y,1}$ represent on average 99.4\% of the norm of their respective PC.
@@
 As pointed out in Section \ref{subsec:meth:PCA:transition}, \gls{PCA} does not make any distinction between a vector of variation and its opposite.
-This is why $\text{PC}_{2025,1}$ and $\text{PC}_{2035,1}$ are actually very similar even though there are mostly on the opposite sides of the 0-axis.
+This is why $\text{PC}_{2025,1}$ and $\text{PC}_{2035,1}$ are very similar even though they are mostly on the opposite sides of the 0-axis.
 
@@
 \includegraphics[width=0.9\textwidth]{Top5_PC3.pdf}
-\caption{Contribution of the top-5 technologies to the first three \gls{PCs} over the different representative years.
+\caption{Contributions of the top-5 technologies to the first three \gls{PCs} over the different representative years.
 Where the variation of \gls{PV} panels supported by the electrification of the \gls{HT} heat via industrial electric resistors are the main varying drivers over the whole transition, there are tipping years like 2025 for the production of \gls{HVC} (\ie naphtha-cracker to \gls{MTO}) or 2025-2030 for the production of \gls{LT} heat (\ie from oil and gas boilers to electric \gls{HP}).}
@@
 Therefore, we observe that the variation of installed \gls{PV} is directly linked with the variation of installed industrial electrical resistors and, to a smaller extent, of decentralised and \gls{DHN} electrical \gls{HP}.
-These variations spread over the whole transition and they are the main varying factors to the first $\text{PC}_{\text{transition}}$.
-Where this was an example for correlated technologies, we can also identify some key modal shifts where one technology is either substituted or in balance with others.
+These variations spread over the whole transition and they are the main varying factors in the first $\text{PC}_{\text{transition}}$.
+Where this was an example of correlated technologies, we can also identify some key modal shifts where one technology is either substituted or in balance with others.
 First, about the \gls{LT} heat sector, the early stages of the transition, \ie 2025-2030, sees the shift from mainly decentralised oil and gas boilers towards decentralised and \gls{DHN} electrical \gls{HP}.
@@
 \includegraphics[width=0.8\textwidth]{Top5_PCtransition.pdf}
-\caption{Contribution of the top-5 technologies to the different $\text{PC}_{\text{transition}}$. }
+\caption{Contributions of the top-5 technologies to the different $\text{PC}_{\text{transition}}$. }
 \label{fig:Top5_PC_transition}
@@
 Besides the additional operational costs due to more expensive energy carriers, this case presents cumulative investments over the transition that are 2.6\% (17\,b€) higher than the REF case.
-The following paragraphs investigate the differences in the different sectors between the ROB and the REF cases, similarly to the comparison carried out with the SMR case in Chapter \ref{chap:atom_mol}.\\
+The following paragraphs investigate the differences in the different sectors between the ROB and the REF cases, similar to the comparison carried out with the SMR case in Chapter \ref{chap:atom_mol}.\\
 
@@
 It is mostly the industrial electric heaters that absorb the abundant and intermittent electricity produced by \gls{PV} panels and wind turbines.
-In the mobility sectors, \gls{BEV} substitutes gasoline cars from 2025 onward and electric trucks substitute diesel trucks in the near-term before being substituted by \gls{FC} trucks at the end of the transition.
+In the mobility sectors, \gls{BEV} substitute gasoline cars from 2025 onward and electric trucks substitute diesel trucks in the near term before being substituted by \gls{FC} trucks at the end of the transition.
 On the supply side, besides the local \gls{VRES} and the direct import of electricity from abroad, \gls{CHP} units, mostly industrial, are favoured given their higher efficiency at the expense of ammonia-\gls{CCGT}.
@@
 Compared to the REF case, local \gls{VRES} are deployed as soon as possible supported by an earlier electrification of the heat and transport sectors.
-\gls{CHP} units also represents a higher share of the electricity supply given their higher efficiency at the expense of other flexible production units like ammonia-\gls{CCGT}.}
+\gls{CHP} units also represent a higher share of the electricity supply given their higher efficiency at the expense of other flexible production units like ammonia-\gls{CCGT}.}
 \label{fig:rob_pol_ROB_elec}
@@
 However, in the mid-term, 27\% of these heaters are prematurely decommissioned where 24\% are not renewed after reaching the end of their lifetime.
-Instead, up to +3.5~GW of industrial \gls{CHP} units are installed to efficiently use e-methane that is considered more expensive in the ROB case to supply electricity and 52\% of the \gls{HT} heat demand.
+Instead, up to +3.5~GW of industrial \gls{CHP} units are installed to efficiently use e-methane which is considered more expensive in the ROB case to supply electricity and 52\% of the \gls{HT} heat demand.
 About the \gls{LT} heat sector, the principal impact is the bigger installation of \gls{DHN} and decentralised electric \gls{HP}, up to +2.1~GW and +0.8~GW respectively.
@@
 Used as a way to support the the early installation of solar \gls{PV}, electric trucks are then replaced by \gls{FC} ones given the limited availability of renewable electricity and its more cost-effective use in other sectors such as private mobility.
-Freight transport via trains and boats are not different from the REF roadmap.
-About the passenger mobility, compared to the REF case, the shift from \gls{ICE} cars to \gls{BEV} occurs earlier in the transition, in line with the increased production of intermittent electricity from \gls{VRES}.
+Freight transport via trains and boats is not different from the REF roadmap.
+Regarding passenger mobility, compared to the REF case, the shift from \gls{ICE} cars to \gls{BEV} occurs earlier in the transition, in line with the increased production of intermittent electricity from \gls{VRES}.
 \\
@@
 In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}.
-More expensive imported energy carriers lead to an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) and an increase of the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
+More expensive imported energy carriers lead to an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) and an increase in the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case.
 This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}.
@@
 For these, in case the installed capacity prescribed by the roadmap is higher than the maximum potential affected by the value of the uncertain parameter, the actual installed capacity is set to this new limit.
-In reality, this would be similar to revise to a smaller extent the expected roadmap due to external events (\eg smaller public acceptance \cite{zoellner2008public,sam2014small}, lower \gls{TRL} or longer installation time than expected).
+In reality, this would be similar to revising to a smaller extent the expected roadmap due to external events (\eg smaller public acceptance \cite{zoellner2008public,sam2014small}, lower \gls{TRL} or longer installation time than expected).
 
@@
 As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}.
-As naphtha-cracker is the more economical option in the near-term of a myopic transition, this smoother modal shift leads to less variations of installed capacities.
+As naphtha-crackers are the more economical option in the near term of a myopic transition, this smoother modal shift leads to fewer variations of installed capacities.
 This makes the SMR roadmap more robust than the REF case.
@@
 Finally, the ROB roadmap is 21\% less robust for PC\,5.
-Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller.
+Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long term to supply the potential additional demand of \gls{HVC}, the near term capacities of naphtha-crackers are smaller.
 As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.
@@
 When computing the \gls{PCs} of the different myopic transitions, the ROB case has a variance that is 10\% smaller than the one of the REF case.
-Moreover, we observe that the variations of \gls{PV}, industrial resistors and \gls{LT}-heat technologies still have a major contribution on the variation of the system roadmaps (see Figure \ref{fig:Top5_PCtransition_cases}).
+Moreover, we observe that the variations of \gls{PV}, industrial resistors and \gls{LT}-heat technologies still have a major contribution to the variation of the system roadmaps (see Figure \ref{fig:Top5_PCtransition_cases}).
 More surprising is the contribution of the variation of the technology to produce methanol from methane.
-Where methanol is preferably imported from abroad (or produced from biomass) in the perfect foresight approach, the myopic optimisation opts for this technology in the near-term (2025) to compete with the imports as the main supplier of this energy carrier.
+Where methanol is preferably imported from abroad (or produced from biomass) in the perfect foresight approach, the myopic optimisation opts for this technology in the near term (2025) to compete with the imports as the main supplier of this energy carrier.
 
@@
 \includegraphics[width=0.75\textwidth]{Top5_PCtransition_ROB_MY.pdf}
-\caption{Contribution of the top-5 technologies to the \gls{PCs} of myopic transitions for the REF, SMR and ROB cases.
+\caption{Contributions of the top-5 technologies to the \gls{PCs} of myopic transitions for the REF, SMR and ROB cases.
 The variation of \gls{PV}, industrial electric resistors and \gls{LT}-heat technologies still have a major contribution of the variation in the system roadmaps.
@@
 \includegraphics[width=0.325\textwidth]{CAPEX_2050_REF_ROB.pdf}
-\caption{Comparison of total transition cost (left), cumulative OPEX (center) and CAPEX (right) by 2050 from the REF and ROB myopic transitions.
+\caption{Comparison of total transition cost (left), cumulative OPEX (centre) and CAPEX (right) by 2050 from the REF and ROB myopic transitions.
 The differences are marginal.
-However, the ROB roadmap lead to cheaper and less varying total transition cost.
+However, the ROB roadmap leads to cheaper and less varying total transition costs.
 Requiring more investments at earlier stages, this roadmap is less affected by the variation of the operational costs.}
@@
 \caption{Quartile coefficient of dispersion for the total transition cost, cumulative OPEX and CAPEX by 2050 from the REF and ROB myopic transitions.
-The ROB case leads to more certain total transition cost and cumulative \gls{OPEX} than the REF case.}
+The ROB case leads to a more certain total transition cost and cumulative \gls{OPEX} than the REF case.}
 \label{tab:quartile_coeff_dispersion}
@@
 
-All these conclusions are similar to the ones observed on the study of overcapacity of \citet{moret2020overcapacity}).
-Working on the snapshot model and fixing the installed capacities only in the power sector, they although observed bigger gain to opt for a robust strategy rather than the deterministic solution.
+All these conclusions are similar to the ones observed in the study of overcapacity by \citet{moret2020overcapacity}).
+Although, working on the snapshot model and fixing the installed capacities only in the power sector, they observed a bigger gain to opt for a robust strategy rather than the deterministic solution.
 
@@
 In this chapter, we applied the \gls{PCA}-based approach detailed in Chapter \ref{chap:chap_methodo} to assess the robustness of different transition roadmaps for the case of Belgium.
-First, starting from the principal components of each representative year, we have computed the PCs of the transition to serve as robustness metric.
+First, starting from the principal components of each representative year, we have computed the PCs of the transition to serve as a robustness metric.
 This showed that the variation of installed capacities of \gls{PV}, industrial electric resistors and \gls{LT}-heat technologies drove the major share (75\%) of the variation of the transition design strategy.
@@
 Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB.
-Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}.
-From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation of the additional capacities needed in myopic optimisations subject to uncertainties.
-This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation.
+Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case considered the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}.
+From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation in the additional capacities needed in myopic optimisations subject to uncertainties.
+This showed that counting on the installation of \gls{SMR} by 2040 provided the same level of robustness as the reference case in the key directions of variation.
 However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario.
 Even though this roadmap represents 7.4\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).
-On top of decreasing the risk to invest into additional capacities, the ROB is also less sensitive to the variation of the costs of purchasing the energy carriers that are the main contributors of uncertainty.
+On top of decreasing the risk of investing in additional capacities, the ROB is also less sensitive to the variation in the costs of purchasing the energy carriers that are the main contributors to uncertainty.
 
-Overall, this \gls{PCA}-based approach allows evaluating the variation of multiple-year transition pathway of a whole-energy system and assessing its robustness.
+Overall, this \gls{PCA}-based approach allows evaluation of the variation of multiple-year transition pathway of a whole-energy system and assessing its robustness.
 Doing so, it brings more information than when considering only the variation of the objective function.

### 2024-07-24 chap_intro_ccl/conclusion.tex
Grammarly done for the core of the text

@@
 % WHAT WE DID (HOW?)
-The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2} budget without prescribed trajectory for the reduction of the emissions.
-We have studied the case study of Belgium with a focus on the role that renewable electrofuels (e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely-populated and highly industrialised country with limited local renewable potentials.
-To have 66\% chance of limiting warming to 1.5°C and using the ``grandfathering'' sharing principle of the global budget, the \ce{CO2} budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
+The general objective of this thesis has been to investigate the impact of uncertainties and limited foresight on the transition pathway of a whole-energy system, accounting for multi-energy carriers and multi-sectors, subject to \ce{CO2} budget without a prescribed trajectory for the reduction of the emissions.
+We have studied the case study of Belgium with a focus on the role that renewable electrofuels (e-ammonia, e-methane, e-methanol and e-hydrogen) could play in this densely populated and highly industrialised country with limited local renewable potentials.
+To have a 66\% chance of limiting warming to 1.5°C and using the ``grandfathering'' sharing principle of the global budget, the \ce{CO2} budget of the 2020-2050 transition of Belgium has been set to 1.2\,Gt$_{\ce{CO2},\text{eq}}$.
 
@@
 A myopic approach has been implemented, considering a sequence of 10-year time windows with an overlap of 5 years rather than one global optimisation over the 30 years of the transition.
-It allows assessing the realism of the decision making process where knowledge about the future is progressively unfold, rather than assumed to be fully known from the beginning of the transition as in the perfect foresight (PF) approach.
+It allows assessing the realism of the decision-making process where knowledge about the future is progressively unfolded, rather than assumed to be fully known from the beginning of the transition as in the perfect foresight (PF) approach.
 
 The future of a national whole-energy system is subject to a plethora of uncertainties.
-Main examples fo uncertainties are the cost of purchasing resources, their availabilities, the \acrfull{CAPEX} of technologies and the level of demands.
+The ain examples of uncertainties are the cost of purchasing resources, their availabilities, the \acrfull{CAPEX} of technologies and the levels of demand.
 In this thesis, we have used the uncertainty characterisation approach developed by \citet{moret2016strategic} to provide the range of variation of the uncertain parameters.
@@
 In this framework, an agent takes actions at the beginning of each time window to limit the emissions of the system, the \textit{environment}, and the consumption of fossil fuels.
-Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until reaching 2050 where a transition was considered as successful when respecting the \ce{CO2} budget.
-Repeating the whole transition with different sequences of actions-states allowed the agent coming up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
+Starting from the initial state of the energy system in 2020, the agent takes every five years a set of actions until 2050 when a transition was considered successful when respecting the \ce{CO2} budget.
+Repeating the whole transition with different sequences of actions and states allowed the agent to come up with an effective policy towards sustainability, considering the variation of the parameters of its environment.
 
@@
 % THE RESULTS WE GOT (WHAT?)
-Given the \ce{CO2} budget to stick to a +1.5$^{\circ}$C increase of global temperature and the presumably high levels of demands,  aiming at a cost-optimum transition, Belgium should first opt for more efficiency to reduce by one third the primary energy consumption.
+Given the \ce{CO2} budget to stick to a +1.5$^{\circ}$C increase in global temperature and the presumably high levels of demand,  aiming at a cost-optimum transition, Belgium should first opt for more efficiency to reduce by one third the primary energy consumption.
 Then comes the full deployment of local solar and wind energies that would represent about 30\% of the mix by 2050.
@@
 E-methanol is the key molecule to import in the mid-term to defossilise the non-energy demand of \acrfull{HVC}, about 43\,TWh/year, by substituting naphtha-crackers by \acrfull{MTO}.
-The uncertainty on this demand is the most impacting parameter (80\%) on the variation of the quantity of imported e-methanol.
+The uncertainty on this demand is the most impacting parameter (80\%) on the variation in the quantity of imported e-methanol.
 
-In the longer-term, on top of the non-energy demand to further produce fertilizers (about 10\,TWh), importing e-ammonia becomes crucial to supply \acrfull{CCGT} and to provide flexibility in the production of electricity in competition with nuclear \acrfull{SMR}.
-For this reason, where the variation of the quantity of imported e-ammonia is mostly driven by the uncertainty on its cost of purchasing (50\%), the uncertainty on the availability of nuclear \gls{SMR} by 2040 is the second most impacting parameter (20\%).
-To a smaller extent, e-methane helps defossilising the industrial boilers and \acrfull{CHP}.
-E-methane imports is mostly impacted by industrial demands (45\%).
+In the longer term, on top of the non-energy demand to further produce fertilizers (about 10\,TWh), importing e-ammonia becomes crucial to supply \acrfull{CCGT} and to provide flexibility in the production of electricity in competition with nuclear \acrfull{SMR}.
+For this reason, where the variation in the quantity of imported e-ammonia is mostly driven by the uncertainty on its cost of purchasing (50\%), the uncertainty on the availability of nuclear \gls{SMR} by 2040 is the second most impacting parameter (20\%).
+To a smaller extent, e-methane helps defossilise industrial boilers and \acrfull{CHP}.
+E-methane imports are mostly impacted by industrial demands (45\%).
 
@@
 This molecule supports the defossilisation of the freight-transport by trucks.
-The variation of the imported quantity of e-hydrogen is mainly caused by the uncertainty on the CAPEX of fuel cell vehicles (25\%) and its cost of purchasing (20\%).
+The variation in the imported quantity of e-hydrogen is mainly caused by the uncertainty on the CAPEX of fuel cell vehicles (25\%) and its cost of purchasing (20\%).
 In 12\% of the transition pathways, up to 10.5\,GW of electrolysers are installed when local renewable electricity is abundant, \gls{SMR} is available and imported e-hydrogen is expensive.
@@
 In the \gls{RL} framework, the agent's actions are mainly driven by the share of renewable energy carriers in the primary energy mix rather than the overall system efficiency.
-This share of renewable energy carriers shows intermediate milestones to reach, 62\%-share by 2030, to have higher chance to succeed the transition by 2050.
+This share of renewable energy carriers shows intermediate milestones to reach, 62\%-share by 2030, to have a higher chance to succeed the transition by 2050.
 Below this threshold, we find the so-called ``no-go zone'' where the chances of success are much lower.
-Due to limited foresight into the future and uncertainties like the possibility to install \gls{SMR} by 2040, successful myopic transitions rely more on importing renewable electrofuels than the perfect foresight approach.
-An illustrative example is the case of e-ammonia, for which imports by 2050 are in general twice larger in the myopic transition.
+Due to limited foresight into the future and uncertainties like the possibility of installing \gls{SMR} by 2040, successful myopic transitions rely more on importing renewable electrofuels than the perfect foresight approach.
+An illustrative example is the case of e-ammonia, for which imports by 2050 are in general twice as large as in the myopic transition.
 This work also highlighted a major difference between the perfect foresight (PF) approach and the agent-based myopic transitions.
-Given the assumed full knowledge on the future, the trajectories to decrease the \ce{CO2} emissions resulting from perfect foresight are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
+Given the assumed full knowledge of the future, the trajectories to decrease the \ce{CO2} emissions resulting from perfect foresight are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
 
 The uncertainty on the maximum installed capacity of solar \gls{PV}, which is supported by the electrification of the industrial and decentralised heating sectors, drives 57\% of the variance of the entire technological mix through the whole transition.
-On the contrary, other key contributions to this variance are more focused on tipping years such as the low-temperature heating sectors shift from oil and gas decentralised boilers to electric Heat Pumps (HP) and Decentralised Heating Networks (DHN) from 2025 to 2035.
+On the contrary, other key contributions to this variance are more focused on tipping years such as the low-temperature heating sectors that shift from oil and gas decentralised boilers to electric Heat Pumps (HP) and Decentralised Heating Networks (DHN) from 2025 to 2035.
 In the non-energy sector, the shift from naphtha-crackers to \acrfull{MTO} takes place in 2025.
@@
 Finally, testing technological roadmaps in myopic conditions has shown that cases relying or not on 6\,GW of \gls{SMR} by 2040 onward provide the same level of robustness in terms of additional capacities to install.
-Despite representing an additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \gls{CHP} provides significantly better level of robustness (between 33\% and 49\%).\\
+Despite representing an additional 2.6\% cumulative CAPEX over the transition, investing as soon as possible in local \gls{PV}, the electrification of heating sectors and more efficient technologies such as \gls{CHP} provides a significantly better level of robustness (between 33\% and 49\%).\\
 
@@
 The optimisation of the whole transition pathway indicates that we should aim at reducing the uncertainty on the cost of purchasing these energy carriers.
-On the road to a robust whole-energy system in Belgium, the policy makers, the industries and academia should invest in projects to produce and use these fuels or developing the exchange networks with foreign countries.
-This would provide significant reduction in the variation of the total transition cost and robustness towards uncertainties progressively unfolding with the future.
+On the road to a robust whole-energy system in Belgium, the policymakers, the industries and academia should invest in projects to produce and use these fuels or develop exchange networks with foreign countries.
+This would provide a significant reduction in the variation in the total transition cost and robustness towards uncertainties progressively unfolding in the future.
 
@@
 Often brought up as a ``unicorn'' solution to the energy transition, nuclear \gls{SMR} would be a cost-competitive option against the import of electrofuels, especially e-ammonia and e-methane, but not against the deployment of local \acrfull{VRES} such as solar and wind.
-This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition, especially when accounting for uncertainties and with the goal to respect an ambitious \ce{CO2} budget of the transition.
+This confirmed that investing as soon as possible in local renewables is one pillar of the energy transition respecting an ambitious \ce{CO2} budget, especially when accounting for uncertainties and
 
 Regarding methodologies, this thesis developed novel approaches to explore myopic pathway transitions, highlight intermediate milestones and no-go zones towards sustainability and assess the robustness of transition technological roadmaps.
-Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them on their specific case study and energy system optimisation model.
-Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further researches but will need to be adapted according to model that is used or the case study that is studied.\\
+Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them to their specific case study and energy system optimisation model.
+Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further research but will need to be adapted according tothe model that is used or the case study that is studied.\\
 
@@
 
-The Belgian transition pathways explored in this thesis massively relied on the early deployment of local \gls{VRES} and import of electrofuels.
+The Belgian transition pathways explored in this thesis massively relied on the early deployment of local \gls{VRES} and the import of electrofuels.
 In some pathways, the current capacities of \gls{PV} and wind turbines are multiplied by 10 and 6 over a 5 to 10 year time window.
@@
 The necessary amount of imported electrofuels could be 10 to 20 times higher than the potential that Belgium could benefit from considering the existing agreements with exporting countries \cite{lefebvre2022electrofuel}.
-A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
+Future works would be to constrain the deployment of local renewables or the availability of electrofuels.
 This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
 This could also show the limits of the Belgian energy system to meet such an ambitious \ce{CO2} budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
-One could then question the level of demands and address the question of the sufficiency.
+One could then question the level of demands and address the question of sufficiency.
 For a growing part of the scientific community, sufficiency is seen as a no-regret lever of action to reduce the anthropogenic \gls{GHG}.
-A study could consist in analysing the level of demands, within their respective sector, we should reach to stay within the \ce{CO2} budget and limit the installation of solar PV, wind turbines, nuclear energy or the import of electrofuels.
+A study could consist in analysing the levels of demand to reach, within their respective sector, to stay within the \ce{CO2} budget and limit the installation of solar PV, wind turbines, nuclear energy or the import of electrofuels.
 
-To assess more accurately the boundary conditions of the Belgium in its transition pathway, a future work could be to merge the EnergyScope Pathway with EnergyScope Multi-Cells \cite{thiran2023validation}.
+To assess more accurately the boundary conditions of Belgium in its transition pathway, future works could be to merge the EnergyScope Pathway with EnergyScope Multi-Cells \cite{thiran2023validation}.
 Rather than model a single country and the rest of the world as the exterior, the Multi-Cells optimises multiple interconnected regions.
@@
 On the methodological side, several perspectives could be considered about the whole-energy system model or the implementation of \gls{RL}.
-The cumulative \acrfull{GHG} emissions of the transition has been constrained based on the use of the resources.
+The cumulative \acrfull{GHG} emissions of the transition have been constrained based on the use of the resources.
 However, in a system reducing drastically its emissions, indirect emissions such as the ones related to the construction and dismantling of assets would have a much higher share of the total emissions.
 Depending on the level of emission reduction, these indirect emissions could be one to three times higher than the direct emissions.
-To accurately integrate these emissions in the model and account for them in the \ce{CO2} budget, \gls{LCA} can compute these indirect emissions.
+To accurately integrate these emissions into the model and account for them in the \ce{CO2} budget, \gls{LCA} can compute these indirect emissions.
 On top of this, \gls{LCA} could provide data to assess other impact indicators besides climate change \cite{astudillo2018integrating}.
@@
 Given the similarities between the policy learned on the monthly (coarse) and the hourly (fine) model, another way to improve the efficiency of the \gls{RL} process would be to implement transfer knowledge \cite{mann2013directed}, or even multi-fidelity \cite{cutler2014reinforcement}.
-The first step, transfer learning, would consist in starting the learning process on the monthly and faster model, refine the action space, and continue learning on the more accurate hourly model.
+The first step, transfer learning, would consist in starting the learning process on the monthly and faster model, refine the action space and continue learning on the more accurate hourly model.
 More than this unidirectional transfer of information, \citet{cutler2014reinforcement} developed the Multi-Fidelity Reinforcement Learning (MFRL) framework that defines rules for when the agent should move up to a higher fidelity environment, as well as moving down in fidelity before over-exploring in a more expensive optimisation.
@@
 
-Nowadays, the development on solar and wind energies supported by other technical solutions such as electrofuels exchange market or nuclear energy, is not fast enough to stick to the +1.5$^{\circ}$ C of global warming.
+Nowadays, the development of solar and wind energies supported by other technical solutions such as the electrofuels exchange market or nuclear energy, is not fast enough to stick to the +1.5$^{\circ}$ C of global warming.
 Even though it represents a growing part of the literature, sufficiency is rarely brought up by decision-makers as a solution.
-Whole-energy system models and related methodologies do not claim to predict the future but rather aims at quantifying options under several assumptions.
-In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into the intrinsically political and interdisciplinary discussions and decisions that must be made in the coming years.
+Whole-energy system models and related methodologies do not claim to predict the future but rather aim at quantifying options under several assumptions.
+In this sense, this work provided insight into possible transition pathways for Belgium to bring the technical dimension into the intrinsically political and interdisciplinary discussions and decisions that must be made in the coming years.

### 2024-07-24 appendices/app_case_study.tex
Grammarly up to 4.2 (included)
Grammarly up to App. C.4.1 (included))

@@
 \midrule
-\multirow{2}{*}{Private mobility\footnote{\label{foot:privatemob_80}The private mobility accounts for 80\% of the passengers mobility.}}
+\multirow{2}{*}{Private mobility\footnote{\label{foot:privatemob_80}Private mobility accounts for 80\% of the passengers mobility.}}
  & Diesel car & 48\% & 93.5\,Mpass.-km/h\\
@@
 This is a combination of efficiency and substitution of fossil fuels with electricity.
-In particular, the private mobility drops to near-zero emissions at early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift.
-First, while accounting for the society inertia in this sector (see Equation (\ref{eq:limit_reno_passmob_app})), battery electric vehicles (BEV) substitute \gls{ICE} cars.
+In particular, private mobility drops to near-zero emissions at the early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift.
+First, while accounting for the societal inertia in this sector (see Equation (\ref{eq:limit_reno_passmob_app})), battery electric vehicles (BEV) substitute \gls{ICE} cars.
 This reduces the overall emissions of the private vehicles as electric motors are more efficient than \gls{ICE} and they are driven by a less-emitting resource (\ie electricity versus diesel and gasoline).
@@
 The energy supply decreases from 554~TWh/y in 2020 down to 368~TWh/y in 2050 (\ie -34\%) whereas, in the meantime, the demands have increased by 19\%, on average.
-This drop of primary energy consumption reflects the penetration of efficient measures and technologies, such as the previously mentioned public mobility, \gls{DHN} or heat pumps.
+This drop in primary energy consumption reflects the penetration of efficient measures and technologies, such as the previously mentioned public mobility, \gls{DHN} or heat pumps.
 The results in 2050 are aligned with other studies, such as \citet{Devogelaer2013}\footnote{This study was ordered by the National Planning Bureau in 2013.
@@
 The Belgian calculator has been improved since then into a recent expert version called \textbf{My2050} \cite{My2050}.
-From this study, the results of two scenarios will be used: one based on an optimistic evolution of technologies (Technology), and one focusing on an increased dependence on neighbouring countries (EU integration).} \cite{My2050} which estimates respectively a range of 305-417 TWh/y and 307-364~TWh/y for their central scenarios.
+From this study, the results of two scenarios will be used: one based on an optimistic evolution of technologies (Technology), and one focusing on an increased dependence on neighbouring countries (EU integration).} \cite{My2050} which estimate a range of 305-417 TWh/y and 307-364~TWh/y for their central scenarios, respectively.
 
 The first fossil energy to phase out is gasoline, which is exclusively used for private cars.
-Indeed, private mobility is partially replaced by public one\footnote{Given the major role played by private cars in the Belgian passenger mobility nowadays (\ie around 80\% \cite{BFP_mob}), public transport (\eg tramways, buses and trains) is assumed to be able to supply only half of it.}; and the cars are switching from gasoline and diesel to electricity.
+Indeed, private mobility is partially replaced by public one\footnote{Given the major role played by private cars in the Belgian passenger mobility nowadays (\ie around 80\% \cite{BFP_mob}), public transport (\eg tramways, buses and trains) is assumed to be able to supply only half of it.}; and cars are switching from gasoline and diesel to electricity.
 Then, diesel and \gls{LFO} are decreasing.
-As diesel is used for trucks and buses mobility, it is harder to phase out compared to gasoline exclusively burned in cars.
-The first drop of \gls{LFO} reflects the switch from oil boilers to other technologies: heat pumps and gas cogeneration mainly.
+As diesel is used for truck and bus mobility, it is harder to phase out compared to gasoline exclusively burned in cars.
+The first drop of \gls{LFO} reflects the switch from oil boilers to other technologies: mainly heat pumps and gas cogeneration.
 Then, it is mainly used for the production of \gls{HVC}, this reflects that \gls{HVC} is a feedstock hard to defossilise.
 Finally, coal is kept mainly for industrial usage because it is a cheap fossil fuel (mainly for industrial usage).
-To phase it out beforehand, a penalty mechanism, such as a carbon tax, would be required, or its strict ban should be put in place.
+To phase it out beforehand, a penalty mechanism, such as a carbon tax, would be required, or a strict ban should be put in place.
 The last fossil energy present in the system is fossil gas, used for the production of electricity and heat, through cogeneration mainly.
-Indeed, gas plays a key role to balance the intermittency of solar and wind.
+Indeed, gas plays a key role in balancing the intermittency of solar and wind.
 
@@
 
-In 2020, Belgium has been a net-exporter of electricity, however with the shut-down of nuclear power plants and the increase of electricity consumption, Belgium will become a net importer of electricity.
+In 2020, Belgium has been a net exporter of electricity, however with the shut-down of nuclear power plants and the increase of electricity consumption, Belgium will become a net importer of electricity.
 These imports reach their maximal allowed capacity by 2035 (i.e.
@@
          \includegraphics[width=\textwidth]{ELEC_tech_cap.pdf}
-         \caption{The electrical production capacity will experience  massive expansion of wind turbines (onshore and then offshore) and a soaring installed capacity of \gls{PV}.
+         \caption{The electrical production capacity will experience a massive expansion of wind turbines (onshore and then offshore) and a soaring installed capacity of \gls{PV}.
 Ammonia \gls{CCGT} are installed at the end of the transition to provide a flexible capacity as gas \gls{CCGT} are phased out.}
@@
          \includegraphics[width=\textwidth]{ELEC_layer.pdf}
-         \caption{The electricity supply (positive values) will remain a mix of different technologies where backup is first mainly provided by gas-\gls{CCGT} and then imported electricity, heat and power cogeneration and later ammonia-\gls{CCGT}.
+         \caption{The electricity supply (positive values) will remain a mix of different technologies where the backup is first mainly provided by gas-\gls{CCGT} and then imported electricity, heat and power cogeneration and later ammonia-\gls{CCGT}.
 The electricity demand (negative values) is led by the electricity end-use demand, but the share used to electrify heat (heat pumps), vehicles (cars, trains, trams...) and industrial heaters drastically increases.
@@
 By 2050, wind and solar technologies deployments are 60~GW of \gls{PV}, 10~GW of onshore wind turbines and 6~GW of offshore wind turbines.
-To compensate the intermittency, the system relies on imported electricity, gas \gls{CCGT}, sector coupling and storage.
-This is in line with the work of \citet{Devogelaer2020} that ends up with about 80~GW of total installed power generation capacity by 2050, with less \gls{PV} (\ie 39~GW) and more wind capacities (\ie 25~GW of which offshore takes 8.3~GW).
+To compensate for the intermittency, the system relies on imported electricity, gas \gls{CCGT}, sector coupling and storage.
+This is in line with the work of \citet{Devogelaer2020} who ends up with about 80~GW of total installed power generation capacity by 2050, with less \gls{PV} (\ie 39~GW) and more wind capacities (\ie 25~GW of which offshore takes 8.3~GW).
 As an illustration, in 2050, 176.8~TWh of electricity transit on the grid which includes 32.4~TWh of electricity imported and 15.4~TWh of electricity from \gls{CCGT}.
-This result is aligned with other studies that estimates different ranges: 180-310 TWh/y \cite{Devogelaer2013}, about 250~TWh/y \cite{Devogelaer2020}, 126-140~TWh/y \cite{My2050} and in a more recent study
+This result is aligned with other studies that estimate different ranges: 180-310 TWh/y \cite{Devogelaer2013}, about 250~TWh/y \cite{Devogelaer2020}, 126-140~TWh/y \cite{My2050} and in a more recent study
 %\footnote{The EnergyVille consortium, made of belgian universities and research institutes, applied the TIMES-BE model to the case of Belgium to assess the energy transition pathways \cite{PATHS2050}.
@@
 Higher values from \citet{Devogelaer2013} illustrate an almost exclusively electrified energy system.
-The differences between the study ranges reflect the different assumptions in terms of renewable potentials and availability of nuclear energy.
+The differences between the study ranges reflect the different assumptions in terms of renewable potentials and the availability of nuclear energy.
 A general trend is that Belgium should maximise its use of endogenous renewable resources, which \citet{dubois2023multi} identified as a cheaper option than importing additional renewable energies from abroad.
 Demand management reflects the flexible use of electricity, mainly through heat pumps that uncouple the heat demand and the electricity consumption when combined with thermal storage.
-Gas \gls{CCGT} is also a useful asset to compensate intermittent renewables.
+Gas \gls{CCGT} is also a useful asset to compensate for intermittent renewables.
 However, its capacity remains the same as the one installed in 2020.
-These results are verifying an hourly adequacy of the power demand.
+These results verify the hourly adequacy of the power demand.
 Moreover, in a previous study by \citet{pavivcevic2022bidirectionnal}, the snapshot version of the model has been coupled with Dispa-SET, a dispatch optimisation model.
-Results showed that the backup capacity was underestimated by less than 20\% to respect reserve capacity, mainly due the lack of reserve capacity for grid stability.
+Results showed that the backup capacity was underestimated by less than 20\% to respect reserve capacity, mainly due to the lack of reserve capacity for grid stability.
 
-From 2025, the electricity mix has a strong renewable share that rises up to 60\% in 2050.
+From 2025, the electricity mix has a strong renewable share that rises to 60\% in 2050.
 The remaining 40\% are mainly gas (or ammonia) in \gls{CCGT} and cogeneration and imported electricity.
@@
 This makes 12.6~b€ on average per year, which represents 2.1\% of the current Belgian \gls{GDP}.
-This in the range of other studies on climate-neutral scenarios concluding additional investment needs of the order of 2 to 3\% of the \gls{GDP} on a global scale \cite{IEA2021,IRENA2021} or 2.7 to 3.8\% at the European level \cite{Widuto2023}.
+This is in the range of other studies on climate-neutral scenarios concluding additional investment needs of the order of 2 to 3\% of the \gls{GDP} on a global scale \cite{IEA2021,IRENA2021} or 2.7 to 3.8\% at the European level \cite{Widuto2023}.
 
@@
 \caption{The cumulative investments over the transition are unequally spread between the sectors.
-The energy system in 2020 is imposed as the existing energy system and its expenses are split in three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
-The investments required during the transition represents 150\% the initial investment and mainly in the same three sectors.
+The energy system in 2020 is imposed as the existing energy system and its expenses are split into three main categories: mobility (mainly vehicles), infrastructure (mainly grids) and electricity (mainly thermal power plants).
+The investments required during the transition represent 150\% of the initial investment and are mainly in the same three sectors.
 }
@@
 Apart from the investment required to replace all private vehicles (accounting for 44\% of the overall investments), the remaining sectors represent a total of 212 b€.
-To mitigate the cost of the transition, My2050 suggests deploying a fleet of no more than one million vehicles and implementing a car sharing system, distinct from car-pooling, as an inevitable measure \cite{My2050}.
+To mitigate the cost of the transition, My2050 suggests deploying a fleet of no more than one million vehicles and implementing a car-sharing system, distinct from car-pooling, as an inevitable measure \cite{My2050}.
 
@@
 Investments in infrastructures, public mobility, storage and other long-lifetime technologies experience an important salvage value on the contrary, investments in private mobility will not be recovered as vehicles have a short lifetime.
-All together, these salvage values represent 160.1 b€, 25\% of the cumulative investment costs in 2050.}
+Altogether, these salvage values represent 160.1 b€, 25\% of the cumulative investment costs in 2050.}
 \label{fig:pestd_inv_return}
@@
 \includegraphics[width=\textwidth]{System_cost.pdf}
-\caption{The yearly system cost shows the shift from non-renewable to renewable resources (mainly electrofuels). Operation and maintenance costs represents almost 50\% of the expenses.}
+\caption{The yearly system cost shows the shift from non-renewable to renewable resources (mainly electrofuels). Operation and maintenance costs represent almost 50\% of the expenses.}
 \label{fig:pestd_cumul_op}
@@
 
-The annualised cost of the energy system in 2020 is estimated to 44.3~b€/y and increases by 5.5~b€/y to reach 49.8~b€/y by 2050.
+The annualised cost of the energy system in 2020 is estimated to be 44.3~b€/y and increases by 5.5~b€/y to reach 49.8~b€/y by 2050.
 The work of \citet{My2050} estimates the annualised cost in 2050 between 63 and 82~b€/y, while other studies just indicate the cost increase compared to 2015 (+11.7 to +21) \citep{Devogelaer2013,PATHS2050}.
 The differences come from the scope of the energy system: as an example \citet{My2050} also account for the agriculture sector.
-These differences highlight the difficulty to compare different studies due to difference of scope and partial availability of used data.
+These differences highlight the difficulty of comparing different studies due to differences in scope and partial availability of used data.
 Overall, comparing with existing studies shows the consistency of the results provided by EnergyScope Pathway.
@@
 
-Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity-technologies account respectively for 83.2 and 61.2~b€ in 2030 whereas the overall cumulative investments, so far, are 421.3~b€.
+Figure \ref{fig:my_pestd_inv_return_area} highlights this as infrastructures and the electricity technologies account respectively for 83.2 and 61.2~b€ in 2030 whereas the overall cumulative investments, so far, are 421.3~b€.
 The significant lifetime (\eg 80 years) and investment cost (\ie 368M€/GW$_\text{VRES}$ \cite{readthedocs_pathway}) of the power grid, and, on a smaller scale, the district heating network, explain why the myopic optimisation opts for a higher investment in these infrastructures, at early stages.
-Similarly to what \citet{keppo2010short} observed in their studies, these early investments consequently lead to more investments, later in the transition, to renew technologies that have become too old before 2050: during the phase between 2045 and 2050, the myopic approach needs to invest in 9.2GW of PVs that have been installed 25 years before whereas the perfect foresight, by smoothing its investments over the entire transition, has to renew only 2.5GW of PVs.
+Similar to what \citet{keppo2010short} observed in their studies, these early investments consequently lead to more investments, later in the transition, to renew technologies that have become too old before 2050: during the phase between 2045 and 2050, the myopic approach needs to invest in 9.2GW of PVs that have been installed 25 years before whereas the perfect foresight, by smoothing its investments over the entire transition, has to renew only 2.5GW of PVs.
 
@@
 \caption{Sum of the respective salvage values of each sector, per end year of each time window.
-All together, these salvage values represent 223.9~b€, 53\% of the cumulative investment costs in 2030, and 157.7~b€, 24\% of the cumulative investment costs in 2050.}
+Altogether, these salvage values represent 223.9~b€, 53\% of the cumulative investment costs in 2030, and 157.7~b€, 24\% of the cumulative investment costs in 2050.}
 \label{fig:my_pestd_inv_return_area}
@@
 
-In 2050, the capacities installed in the different sectors, in other words the design of the system, are very similar between the two approaches.
+In 2050, the capacities installed in the different sectors, in other words, the design of the system, are very similar between the two approaches.
 The passenger and freight mobility sectors are the same and differences smaller than 1GW are observed in other sectors.
-More interestingly, the myopic optimisation tends to postpone the decommissioning of capacities when the loss of salvage values at the end of an optimisation window would be bigger than the maintenance cost.
+More interestingly, myopic optimisation tends to postpone the decommissioning of capacities when the loss of salvage values at the end of an optimisation window would be bigger than the maintenance cost.
 For instance, in the myopic approach, 0.8GW of industrial coal boilers will remain installed in 2045 and 2050 or 3.6GW of naphtha-crackers to produce HVC in 2040, whereas these technologies are not used.
-This could be compare to the ``lock-ins'' detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at early stages of the transition remain in place.
+This could be compared to the ``lock-ins'' detailed in other studies \cite{heuberger2018impact,keppo2010short} where technologies installed at the early stages of the transition remain in place.
 However, in our case, keeping in place unused technologies is due to the formulation of the salvage value.
@@
 For instance, in 2035, 3.7GW (+75\%) more of industrial electric heaters produce 5.1TWh/year (+130\%) of additional industrial heat.
-In the low-temperature heat sector, decentralised and centralised electric heat pumps capacities are, respectively, 2.2GW (+19\%) and 1GW (+8\%) higher for each of the representative years between 2030 and 2045, to produce, around 7.8TWh/year (+23\%) and 0.8TWh/year (+1\%), at the expense of other technologies such as gas heat pumps.
+In the low-temperature heat sector, decentralised and centralised electric heat pump capacities are, respectively, 2.2GW (+19\%) and 1GW (+8\%) higher for each of the representative years between 2030 and 2045, to produce, around 7.8TWh/year (+23\%) and 0.8TWh/year (+1\%), at the expense of other technologies such as gas heat pumps.
 Finally, public trains substitute from 2035 a higher share of the CNG-buses.
@@
 \label{app:CO2_trajectories}
-This section compare results of the optimisation subject to different emissions-trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon neutrality in 2050.
+This section compares the results of the optimisation subject to different emissions trajectories with the case where the emissions are constrained to decrease linearly from the level in 2020 to carbon neutrality in 2050.
 The first comparison is done with the trajectory subject to respect the \ce{CO2} budget prescribed in the \gls{RL}-based pathway optimisation, \ie 1.2\,Gt$_{\ce{CO2},\text{eq}}$ (Section \ref{sec:cs:CO2-budget}).
-Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions-trajectory.
+Then, to assess the impact of the myopic approach on the optimisation of the transition, it is compared to perfect foresight results, removing the constraint on the emissions trajectory.
 
@@
 
-\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2} budget) and a case where the \ce{CO2}-trajectory is constrained instead.
-Interestingly, these two transition pathways end up in a similar carbon neutral whole-energy system in 2050.
-The two main sectors that significantly reduce their emissions in the REF case are the production of \gls{HVC} and the high-temperature heat.
-In the former, this is linked to the earlier phase-out of oil products through naphtha-cracking.
-In the latter, it is the use of coal in industrial coal boilers that is phased-out.
-Overall, ending up to the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.
+\Cref{fig:app_CO2_REF_lin} shows the yearly emissions attributed to each sector in the REF case (\ie imposed \ce{CO2} budget) and a case where the \ce{CO2} trajectory is constrained instead.
+Interestingly, these two transition pathways end up in a similar carbon-neutral whole-energy system in 2050.
+The two main sectors that significantly reduce their emissions in the REF case are the productions of \gls{HVC} and high-temperature heat.
+In the former, this is linked to the earlier phase-out of oil products through naphtha cracking.
+In the latter, it is the use of coal in industrial coal boilers that is phased out.
+Overall, ending up with the same level of emissions in 2050, the REF case represents a 60\% reduction of the cumulative emissions compared to the linear decrease, for a 7.5\% more expensive transition.
 
@@
 
-In the following, we perform a similar comparison of the transition pathway without restricting the \gls{GHG} emissions. \\
+In the following, we perform a similar comparison of the transition pathway without restricting \gls{GHG} emissions. \\
 
@@
 
-Similar trends that for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
-However, some changes reflects the cheapest option that the system could utilise such as using as little electrofuels as possible.
+Similar trends that the ones for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
+However, some changes reflect the cheapest option that the system could utilise such as using as little electrofuels as possible.
 In this case, only e-ammonia is used for its end use demand.

### 2024-08-22 rebuttals/rebuttals.tex
Including remarks from HJ (Rebuttals)

@@
 
-I would like to thank the jury members for the comments, which significantly helped improving the manuscript and substantiating the novelty of my work.
-Based on the notes taken during the private defense, I hereby transcribed, as accurately as possible, the jury's comments.
+I would like to thank the jury members for their comments, which significantly helped improving the manuscript and substantiating the novelty of my work.
+Based on the notes taken during the private defense, I hereby answered, as accurately as possible, the jury's comments.
 
 I believe I have addressed all the issues raised in the following answers.
-Some of them required adaptations of the text.
+Some of them required adaptations either to the text of the PhD thesis or to the subsequent papers.
 These adaptations are either directly in the thesis manuscript or left for further developments in subsequent papers.
@@
 
-\noindent The answer provided to the comment and, based on this, where the potential modification brought to the manuscript is located in the thesis manuscript using {\color{blue} blue font}.
+\noindent The answer provided to the comment. Based on this, if a modification was brought to the manuscript, it is located in the thesis manuscript using {\color{blue} blue font}.
 
@@
 
-\noindent The purpose of each methods and the assumptions are discussed in the ``Assumptions - Fine tuning'' section of the present document.
-Regarding the choice of PCE instead of monthly version for uncertainty, as discussed during the private defense, monthly version of the model does not provide enough detail to provide sufficient confidence in the results given the need to massively integrate VRES for Belgium to succeed its transition.
+\noindent The purpose of each method and the assumptions are discussed in the ``Assumptions - Fine tuning'' section of the present document.
+Regarding the choice of PCE instead of monthly version for uncertainty, as discussed during the private defense, monthly version of the model does not provide enough detail to build sufficient confidence in the results given the need to massively integrate VRES for Belgium to succeed its transition.
 Given the hourly/daily intermittency, properly assessing the integration of wind and solar requires a fine time resolution that is not provided by the monthly model.
@@
 Further details are also given {\color{blue}in Appendix B.2}.
-Finally, these variables are even exhaustively detailed in \cite{limpens2019energyscope} and \cite{limpens2024pathway} that are cited several times in the thesis manuscript.
+Finally, these variables are exhaustively detailed in \cite{limpens2019energyscope} and \cite{limpens2024pathway} that are cited several times in the thesis manuscript.
 Regarding the uncertainties, the way the ranges are defined and the way they affect the value of some parameters in the model are explained {\color{blue}in Section 1.2 of the core of the manuscript}.
@@
 Although, the main (and only) requirement is to keep a low computational time as getting relevant results from the following approaches usually requires $10^3\sim 10^4$ runs.
-In our case, we have implemented the model to keep the time to run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
+In our case, EnergyScope Pathway can run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
 \end{mdframed}
@@
 Characterising the distribution followed by uncertain parameters require a lot of data that are either seldom available or poorly documented in the literature.
-Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distribution.
+Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distributions.
 Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further researches to refine the characterisation of these parameters.
@@
 To avoid singularities in our PCs, we have decided \textit{a posteriori} to put high and low outliers to the upper and lower fences, respectively.
-However,before managing the outliers, we would recommend to keep them and assess the provided results in terms of number of PCs and the key technologies that characterise them.
-
-Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{text{transition}}$ (85\%).
+However, before managing the outliers, we would recommend to keep them and assess the provided results in terms of number of PCs and the key technologies that characterise them.
+
+Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{\text{transition}}$ (85\%).
 These thresholds have been defined based on good practices discussed in the literature and \textit{a posteriori} observations specific to our case study.
@@
 In this work, we have investigated other actions like incentivising solar PV panels and wind turbines by ``artificially'' reducing their CAPEX.
-The result was not conclusive as these technologies need to take part to the Belgian energy transition and because it was harder to assess the impact of this action.
+The result was not conclusive as these technologies must take part to the Belgian energy transition and because it was harder to assess the impact of this action.
 
@@
 Finally, states aim at representing the information relevant to the agent to efficiently learn and progress through the transitions.
-For this reason, on top reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assessing their policies to reach their targets (share of renewables in the mix and the system overall efficiency).
+For this reason, on top of reward-related features (cumulative emissions and costs), we added other indicators that are actually monitored to help decision-makers assessing their policies to reach their targets (share of renewables in the mix and the system overall efficiency).
 For other studies, one might consider other information like the metrics considered by \citet{pickering2022diversity} (\eg heat electrification, average national import or level of curtailment).
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further researches but will need to be adapted according to model that is used or the case study that is studied.
+Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further researches but will need to be adapted according to the model that is used or the case study that is studied.
 \end{mdframed}
@@
 
-\noindent As discussed during the private defense, storage is not more discussed in the manuscript for two reasons.
-First, the storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}).
-Analyses under uncertainties have not shown major variations neither.
-Second, storage assets are intrinsically linked to the conversion technologies and, consequently, can be seen as by-products of the optimisation results.
-For these two reasons, I have rather focused my analyses on the use of resources and the installation and operation of conversion technologies.
+\noindent As discussed during the private defense, storage is not more discussed in the manuscript for one main reason: the lack of variation through the pathway and under uncertainties.
+The storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}).
+Analyses under uncertainties have not shown major variations either.
+Moreover, storage assets are intrinsically linked to the conversion technologies and, consequently, can be seen as by-products of the optimisation results.
+For these reasons, I have rather focused my analyses on the use of resources and the installation and operation of conversion technologies.
 Given the discussion during the private defense and these further explanations, I have not brought modification to the manuscript in this regard.
@@
 These are strong assumptions that can be challenged as suggested in the conclusion: ``A future work would be to constrain the deployment of local renewables or the availability of electrofuels.
-This could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
+Combined with a reduced end-use-demand, this could provide different pathways where electrofuels would be produced more locally using carbon capture technologies.
 This could also show the limits of the Belgian energy system to meet such an ambitious CO2-budget in its energy transition when solely relying on technological progress, highlighting the need for demand reduction.
@@
 
-The assumption about 0-GWP molecules is similar to the one considered \cite{PATHS2050} which is the most recent study that addressed the case of the Belgian energy transition.
+The assumption about 0-GWP molecules is similar to the one considered in \cite{PATHS2050} which is the most recent study that addressed the case of the Belgian energy transition.
 
 As discussed during the private defense, the first key-message of this work is the need for this (mix of) ``unicorn'' solution(s) given the ambitious objectives set to mitigate climate change.
-This need for this radical change and the fact that solutions rely a lot on assumed to be carbon-neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
+The need for this radical change and the fact that solutions rely a lot on assumed to be carbon-neutral and directly available electrofuels are explicitly mentioned {\color{blue}in the conclusions of Chapter 3}:
 
@@
 Given the ambitious \ce{CO2} budget (\ie 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
+Assumed to be carbon-neutral and directly available by a large amount, one of the ``unicorn'' solution for the case of Belgium is the import of renewable electrofuels, even at early stages of the transition.
 This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.
@@
 Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
-One robust policy would consist in massively investing in electrofuels.
+One robust policy would consist in massively investing in importing electrofuels.
 \end{mdframed}
@@
 
-\noindent Finally, I have added a schematic at the end of the Introduction of the thesis to clarify the structure and highlight that Chapter 4 and 5 can be read after the methodological sections 1.3 and 1.4, respectively (see Figure \ref{fig:intro:Thesis_Structure}).
+\noindent Finally, I have added a schematic at the end of the Introduction of the thesis to clarify the structure and highlight that Chapters 4 and 5 can be read after the methodological sections 1.3 and 1.4, respectively (see Figure \ref{fig:intro:Thesis_Structure}).
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{violet} \textbf{Christophe}} - Formal description of the problem is missing even in described in the text to make the link between the ideas and the implementation.
+{\color{violet} \textbf{Christophe}} - Formal description of the problem is missing even if described in the text to make the link between the ideas and the implementation.
 Add some flow charts to introduce key elements
@@
 
-\noindent This point is addressed with the structure chart presented in the previous comment as well as the improvement of Figure 1.4 presenting the implementation of the myopic approach as well as the way uncertainties and RL change the values of some parameters (see Figure \ref{fig:MY_process_code} of the present document).
+\noindent Besides the mathematical details brought in the description of the optimisation problem, this point is addressed with the structure chart presented in the previous comment as well as the improvement of Figure 1.4 presenting the implementation of the myopic approach as well as the way uncertainties and RL change the values of some parameters (see Figure \ref{fig:MY_process_code} of the present document).
 
@@
 Finally, implementing ``artificial  future years'' would require to either collect more data, which is quite time-consuming, or make other kind of assumptions on the evolution of costs, GWP and other features beyond 2050.
+For instance, we could artificially replicate the year 2050 and reach a steady state.
 
@@
 Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window.
-If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
+If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
 \label{fig:MY_process_code}
@@
 This sequential optimisation allows showcasing robust policies to support myopic transitions (see Chapter 4) and the robustness of roadmaps (see Chapter 5).
-The myopic approach is also the methodological basement to support the implementation of the RL and PCA-based methods.
+The myopic approach is also the methodological basis to support the implementation of the RL and PCA-based methods.
 
@@
 In Equation (B.22), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
-With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost.
-However, one could wonder if a technology that is still operational for half of its lifetime is not worth less than half of its initial investment cost, given the depreciation of the asset.
-In that case, the annualisation factor should account for the last phase of the optimisation (\ie 2045\_2050) when the residual value of the technologies are calculated and subtracted from the total transition cost.
+With this expression, if a technology is still operational for half of its lifetime after 2050, then half of its initial investment is subtracted to the total transition cost.
+However, one could consider the depreciation of the asset where a technology that is still operational for half of its lifetime is worth less than half of its initial investment cost.
 
@@
 Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value.
-In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a salvage value.
-
-The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon. 
+In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a non-0 salvage value.
+
+In a nutshell, the overall conclusion of this sensitivity analysis was that considering a salvage value was fundamental in our pathway formulation.
+However, the expression of the salvage value itself has little influence on the energy system at the end of the time horizon.
+The interested reader is invited to refer to the work of \citet{goffauxpathway} for the numerical results of this analysis.
 \end{mdframed}
@@
 \noindent Indeed, as we want to capture the time preference associated with obtaining finance for the project, effectively quantifying the trade-off between present and future financial value, it is more appropriate to refer to ``discount rate'' when considering the parameter $i_{\text{rate}}$.
+EnergyScope considers a state that needs to decide when investing in which technology.
+There is no consideration of return on investment as it would be the case for private investors.
 We have changed every occurrence of ``interest rate'' into ``discount rate'' {\color{blue}in the entire manuscript}.
@@
 
-\noindent Indeed, the discount rate is usually between 7.5\% and 12\% in other studies.
-The main justification for this choice was to be in line with G.
+\noindent Unlike snapshot models where the discount rate aims at annualizing investments, discounting in pathway models aims at evaluating the value of future expenses compared to present expenses.
+In other words, the higher the discount rate, the lower the impact of future extra costs.
+In their assessment of the role of the discount rates in energy systems optimisation models, \citet{garcia2016role} showed that higher ``social discount rate'', affecting the whole-energy system, including the operation (in opposition to technology-specific ``hurdle rates''), favours fossil-based technologies and postpones investments in more capital-intensive assets like solar PV and wind turbines.
+Where this is visible in the perfect foresight (PF) approach, the myopic optimisation of transition pathways is less affected by this given the shorter time-horizon considered in the sequential optimisations, 10 years versus 30 years in the PF approach.
+Even though the discount rate is usually between 7.5\% and 12\% in other studies, we have set its nominal value in line with G.
 Limpens' work \cite{limpens2021generating}.
-However, assessing the impact of $i_{\text{rate}}$ on $\tau_{phase}$, we see that a lower value of $i_{\text{rate}}$ will favour less late expenses compared to a high value.
-In other words, considering $i_{\text{rate}}=$1.5\% rather than 10\% will less postpone investments (see Figure \ref{fig:tau_phase}).
 I have added some explanations in this regard {\color{blue}in Section 1.1.1}:
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=10cm]{tau_phase.png}
-\caption{$\tau_{phase}$ versus the timing of the expenses (columns) and the discount rate, $i_rate$ (rows).}
-\label{fig:tau_phase}
-\end{figure}
+%\begin{figure}[htbp!]
+%\centering
+%\includegraphics[width=10cm]{tau_phase.png}
+%\caption{$\tau_{phase}$ versus the timing of the expenses (columns) and the discount rate, $i_{rate}$ (rows).}
+%\label{fig:tau_phase}
+%\end{figure}
 
@@
 In line with \citet{limpens2021generating}, we set the nominal value to $i_{\text{rate}}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}.
-Having a low value of discount rate favours less late expenses compared to a higher value of discount rate.
+In the 30-year time window of the perfect foresight approach, having a higher value of discount rate would postpone investments in more capital-intensive assets.
+However, in a myopic vision, this effect is more limited given the shorter time window.
 \end{mdframed}
@@
 Then, for assets with longer lifetime and/or installed later in the transition, annualising over the lifetime would require to have data beyond 2050.
-Effort that I have not decided to make.
 
@@
 Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+However, EnergyScope only considers a vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capital, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capital.
 These further explanations for keeping a single value of discount rate have been added {\color{blue} after Eq.
@@
 Depending on the technology readiness level, private investors would be more or less risk-averse.
-However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capitals, \ie private and public.
-Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capitals.
+However, EnergyScope only considers the vision of a central-planner where a single agent makes the investment decisions without differentiating the sources of the capital, \ie private and public.
+Assuming a differentiated discount rate would require to make assumptions about the private-public distribution of the capital.
 For this reason, we have decided to keep an identical value of discount rate for all the technologies.
@@
 In other words, the infrastructures to transport the energy carriers within the country are not considered.
-It is assumed that the demands have to be supplied by the production, regardless of the flows between the producers and the consumers.
+It is assumed that the demands have to be supplied by the production, without considering the flows between the producers and the consumers.
 Yet, adapting the networks is accounted for in terms of the required investments.
@@
 23: is PCE an appropriate method for LP / MILP models.
-Why is PCE needed is EnergyScope is already very fast?
+Why is PCE needed if EnergyScope is already very fast?
 There is a good fit (1\% LOO error) for the total cost, which is often quite linear, but how about technology choice / sizing?
@@
 Consequently, I decided to exhaustively list these uncertain parameters in the chapter dedicated to the case study (see Section 2.4).
-However, to not leave the reader handing there and wondering if there other parameters and how we have decided to group them, I have added a reference to Section 2.4 {\color{blue}at the end of Section 1.2.1} focusing on the characterisation of the uncertainties:
+However, to not leave the reader handing there and wondering if there are other parameters and how we have decided to group them, I have added a reference to Section 2.4 {\color{blue}at the end of Section 1.2.1} focusing on the characterisation of the uncertainties:
 
@@
 \noindent Indeed, we know the input parameters, the constraints and the objective function implemented in EnergyScope to optimise the transition pathway of the whole-energy system.
-In the RL framework, a model is representation of how the environment will respond to the agent's actions.
-Unlike other cases such as grid world, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
+In the RL framework, a model is a representation of how the environment will respond to the agent's actions.
+Unlike other cases such as grid world \cite{grid_world}, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
 Going for a model-free approach, we avoid building a model within the agent to inform it about this (complex) dynamics.
 
-This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' Given what is already written in the manuscript, I have not brought modification in this regard.
+This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' One sentence of clarification has been added {\color{blue}in Section 1.3.2}:
+
+\begin{mdframed}[style=manuscript] % Comment from the reviewer
+In Reinforcement Learning, the ``model'' stands for the dynamics ``action-state-reward'' between the environment and the agent. 
+\end{mdframed}
 
@@
 \noindent As detailed in Section 1.4.2, even though these points result from the optimisation and make sense mathematically, they are defined as outliers as they fall out of the whiskers of their respective box plot.
-The way we managed these outliers is arbitrary but is fair trade-off between leaving them as they are which would mislead the direction of PCs and completely removing them.
+The way we managed these outliers is arbitrary but is a fair trade-off between leaving them as they are which would mislead the direction of PCs and completely removing them.
 As extensively discussed in Section 5.1.1, this pre-processing step will mostly affect the LT-heat sector where heat pumps running on methane are installed in rare cases, \ie less than 10\% of the 1260 transitions.
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of rooftop PV panels (59.2\,GW) [...]
+On the other hand, solar, wind, hydro and uranium are limited by the technical potentials of rooftop PV panels (59.2\,GW) [...]
 \end{mdframed}
@@
 Similarly to EnergyVille \cite{PATHS2050}, we could have gone up to the upper bound of 8\,GW.
-Going much higher, such as 24\,GW for the ``Electrification'' scenario of EnergyVille, would account for access to a part of the 150\,GW offshore wind park project in the North Sea signed betwee Belgium, Germany, Denmark and the Netherlands (the ``Esbjerg Offshore Wind Declaration'').
+Going much higher, such as 24\,GW for the ``Electrification'' scenario of EnergyVille, would account for access to a part of the 150\,GW offshore wind park project in the North Sea signed between Belgium, Germany, Denmark and the Netherlands (the ``Esbjerg Offshore Wind Declaration'').
 
@@
 Even considering 50\% more expensive SMR subject to a higher discount rate (5\% versus 1.5\% for the other technologies), we notice that SMR keeps on being more cost-competitive than other flexible production units, like e-ammonia CCGTs, that are substituted by SMR when the latter can be installed (see Chapter 3).
-In other words, it means that the optimistic values assumed for SMR in the thesis lead to a cheaper whole-energy system but does not affect its design, \ie the installed technologies.
+In other words, it means that the optimistic values assumed for SMR in the thesis lead to a cheaper whole-energy system but do not affect its design, \ie the installed technologies.
 Consequently, no further modification has been brought to the thesis regarding this comment.
@@
 \noindent The results are deeply detailed in Sections 3.1 and 3.2.
-These sections address the impact of adding SMR in the deterministic solution (\ie uncertain parameters at their respective nominal value) of the Belgian energy transition and global sensitivity analyses on different outputs of interest, respectively.
+These sections address the impact of adding SMR in the deterministic solution (\ie uncertain parameters at their respective nominal values) of the Belgian energy transition and global sensitivity analyses on different outputs of interest, respectively.
 The purpose of these two sections is to give numerical substance to the main conclusions and take-aways that are listed in Section 3.3:
@@
 
-\noindent The difference between BEV and industrial electric heaters is that BEV is already a preferred option for electrification, even the REF case.
+\noindent The difference between BEV and industrial electric heaters is that BEV is already a preferred option for electrification, even in the REF case.
 BEV completely substitute diesel and gasoline-cars as soon as possible in the transition and there cannot be more cars as private mobility is limited to supply up to 50\% of the demand of passengers mobility.
@@
 
-\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/ shocks can be implemented through uncertain parameters.
+\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/ shocks through uncertain parameters.
 For instance, this is how we have implemented the potential to install SMR in 2040 at the soonest.
@@
 Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.
-At the beginning of each episode, one different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1.}
+At the beginning of each episode, a different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section 1.2.1.}
 \label{fig:Schematics_RL}
@@
 Phasing out coal is, of course, necessary as it is a highly-polluting resource.
-However, what this work shows is that, whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
-Consistency towards banning coal through the whole transition is needed.
+However, what this work shows is that, whatever the stage in the transition, a policy limiting the use of coal will always be effective.
+However, to maximise the chance in succeeding the transition, the sooner the better.
 To insist on this aspect, this has been added {\color{blue}at the end of Section 4.2.3}:
@@
 \begin{mdframed}[style=manuscript] 
-Whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
-Consistency towards banning coal is needed.
+Whatever the stage in the transition, a policy limiting the use of coal will always be effective. However, to maximise the chance in succeeding the transition, the sooner the better.
 \end{mdframed}
@@
 Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right).
-\textbf{Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
+\textbf{Number of remaining attempts decrease with time since infeasible problems and solutions overshooting the \ce{CO2} budget are discarded prematurely, \ie before reaching 2050.
 Besides infeasible problems, distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2} budget by 2050 at the latest.} The majority of successful transitions have cumulative emissions much lower than the \ce{CO2} budget and are cheaper than the REF case.
@@
 \begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-The learning phase has been split into batches of 500 steps,\ie 500 sequences of state-action-reward-new state. 
+The learning phase has been split into batches of 500 steps, \ie 500 sequences of state-action-reward-new state. 
 \end{mdframed}

### 2024-08-22 appendices/app_EnergyScope.tex
First version of rebuttals completed + Grammarly Intro + Chap. 1 & 2
Including remarks from HJ (Rebuttals)

@@
 
-\myparagraph{Society inertia}\\
+\myparagraph{Inertia of society}\\
 
@@
 Even if this result is mathematically and physically correct, (\ie fuels are expensive and investing in more efficient technology is economically and environmentally more profitable), this swap of technology cannot occur in one phase (\ie 5 years).
-Indeed, society inertia to change, available manpower, supply chains and manufacturers limit the change.
+Indeed, inertia of society to change, available manpower, supply chains and manufacturers limit the change.
 
@@
 In Equation (\ref{eq:salvage_app}), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built.
-With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost.
-However, one could wonder if a technology that is still operational for half of its lifetime is not worth less than half of its initial investment cost, given the depreciation of the asset.
-In that case, the annualisation factor should account for the last phase of the optimisation (\ie 2045\_2050) when the residual value of the technologies are calculated and subtracted from the total transition cost.
+With this expression, if a technology is still operational for half of its lifetime after 2050, then half of its initial investment is subtracted to the total transition cost.
+However, one could consider the depreciation of the asset where a technology that is still operational for half of its lifetime is worth less than half of its initial investment cost.
 
@@
 Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value.
-In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a salvage value.
+In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a non-0 salvage value.
 
-The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon.
+In a nutshell, the overall conclusion of this sensitivity analysis was that considering a salvage value was fundamental in our pathway formulation.
+However, the expression of the salvage value itself has little influence on the energy system at the end of the time horizon.
+The interested reader is invited to refer to the work of \citet{goffauxpathway} for the numerical results of this analysis.
 

### 2024-08-22 chap_RL/chap_RL.tex
Including remarks from HJ (Rebuttals)

@@
 Then, as the maximum amount of coal (28\,TWh) is much smaller than fossil gas and \gls{LFO}, high values of $\mathrm{act}\textsubscript{coal}$ still represent small consumptions of coal.
-Whatever the stage in the transition, a policy limiting the use of coal will always be effective, whereas one could think that coal should be limited only at early stages to get rid of coal-driven technologies once and for all for the rest of the transition.
-Consistency towards banning coal is needed.
+Whatever the stage in the transition, a policy limiting the use of coal will always be effective.
+However, to maximise the chance in succeeding the transition, the sooner the better.
 
@@
 In this work, we have investigated other actions like incentivising solar \gls{PV} panels and wind turbines by ``artificially'' reducing their \gls{CAPEX}.
-The result is not conclusive as these technologies need to take part in the Belgian energy transition and because it was harder to assess the impact of this action.
+The result is not conclusive as these technologies must take part in the Belgian energy transition and because it was harder to assess the impact of this action.
 

### 2024-08-22 chap_atom_mol/chap_atom_mol.tex
Including remarks from HJ (Rebuttals)

@@
 Given the ambitious \ce{CO2} budget (\ie a 30-year budget representing 10 years of the current emissions), the global sensitivity analysis highlights the need for radical changes.
-Assumed to be carbon-neutral and directly available to a large amount, one of the ``unicorn'' solutions for the case of Belgium is the import of renewable electrofuels, even at the early stages of the transition.
+Assumed to be carbon-neutral and directly available by a large amount, one of the ``unicorn'' solutions for the case of Belgium is the import of renewable electrofuels, even at the early stages of the transition.
 This is why the most impacting parameter on the variability of the total transition cost (around 45\%) is the cost of purchasing these energy carriers.

### 2024-08-22 chap_case_study/chap_case_study.tex
Including remarks from HJ (Rebuttals)

@@
 The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} limiting this potential to about 40\,TWh.
-On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of rooftop \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
+On the other hand, solar, wind, hydro and uranium are limited by the technical potentials of rooftop \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively.
 In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}.

### 2024-08-22 chap_intro_ccl/conclusion.tex
Including remarks from HJ (Rebuttals)

@@
 Besides a wider electrification of the system and the integration of more efficient technologies, Belgium needs to go for radical changes, ``unicorn'' solutions.
-One robust policy would consist in massively investing in electrofuels.
+One robust policy would consist in massively investing in importing electrofuels.
 Chemically similar to their fossil-based equivalents, they can benefit from the current infrastructures and provide seasonal storage and flexibility against the intrinsic intermittency of solar and wind electricity.
@@
 Keeping this work open-source allows the scientific community and the decision-makers to use these methods and adapt them to their specific case study and energy system optimisation model.
-Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further research but will need to be adapted according tothe model that is used or the case study that is studied.\\
+Besides the formulated recommendations, the assumptions and the formulation choices made in this work can serve as a basis for further research but will need to be adapted according to the model that is used or the case study that is studied.\\
 

### 2024-08-22 chap_methodo/chap_methodo.tex
Including remarks from HJ (Rebuttals)

@@
 In line with \citet{limpens2021generating}, we set the nominal value to $i_{rate}=1.5\%$ which is lower than values taken in other studies, between 7.5\% and 12\% \cite{meinke2017energy,simoes2013jrc,EuropeanCommission2016}.
-Having a low-value discount rate favours less late expenses compared to a higher value of discount rate.
+In the 30-year time window of the perfect foresight approach, having a higher value of discount rate would postpone investments in more capital-intensive assets.
+However, in a myopic vision, this effect is more limited given the shorter time window.
 In practice, the discount rate would vary depending on the technology investment risk.
@@
 Myopic optimisation of the pathway consists of a sequence of smaller \gls{LP} problems limited to their respective 10-year time window.
-If the optimisation is not deterministic, step 3.3 is the one where the value of some parameters is changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
+If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section \ref{sec:meth:UQ}) or the actions taken by the Reinforcement Learning-agent (see Chapter \ref{chap:chap_RL})}
 \label{fig:MY_process_code}
@@
 Although, the main (and only) requirement is to keep a low computational time as getting relevant results from the following approaches usually requires $10^3\sim 10^4$ runs.
-In our case, we have implemented the model to keep the time to run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
+In our case, EnergyScope Pathway can run a pathway composed of 7 representative years below 15 minutes on a personal laptop.
 
@@
 Characterising the distribution followed by uncertain parameters requires a lot of data that are either seldom available or poorly documented in the literature.
-Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distribution.
+Consequently, once lower and upper bounds are computed following a method like the one developed by \citet{Moret2017}, a good practice is to consider uniform distributions.
 Preliminary analyses revealing key parameters impacting the most the feature of interest could direct further research to refine the characterisation of these parameters.
@@
 Here, the agent interacts with its environment, \ie the energy-system model on a limited decision window of 10 years.
-At the beginning of each episode, one different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section \ref{subsec:uncert_charac}.}
+At the beginning of each episode, a different sample of uncertain parameters is drawn and affects the environment, EnergyScope Pathway, according to the methodology detailed in Section \ref{subsec:uncert_charac}.}
 \label{fig:Schematics_RL}
@@
 Given the assumed absence of knowledge of the agent about the dynamics of the environment, \ie its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.
+In Reinforcement Learning, the ``model'' stands for the dynamics ``action-state-reward'' between the environment and the agent.
 In practice, in a model-free approach, the agent estimates the optimal policy directly from experience and without estimating the dynamics of the environment.
@@
 
-Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{text{transition}}$ (85\%).
+Finally, we have assumed thresholds to limit the number of principal components of each representative year, $\text{PC}_y$ (90\%), the similarity between two of these $\text{PC}_y$ (90\%) and to limit the number of $\text{PC}_{\text{transition}}$ (85\%).
 These thresholds have been defined based on good practices discussed in the literature and \textit{a posteriori} observations specific to our case study.

### 2024-08-22 frontend/merci.tex
Including remarks from HJ (Rebuttals)

@@
 
-Thank you, thank you, far too kind
+C’est une bonne situation, ça, doctorant?
+Moi je ne crois pas qu’il y ait de bonne ou de mauvaise situation.
+Moi, si je devais résumer ma thèse aujourd’hui avec vous, je dirais que c’est d’abord des rencontres, des gens qui m’ont tendu la main, peut-être à un moment où je ne pouvais pas, où j’étais seul chez moi.
+Et c’est assez curieux de se dire que les hasards, les rencontres forgent un docteur… Parce que quand on a le goût de la chose, quand on a le goût de la chose bien faite, le beau geste, parfois on ne trouve pas l’interlocuteur en face, je dirais, le miroir qui vous aide à avancer.
+Alors ce n’est pas mon cas, comme je le disais là, puisque moi au contraire, j’ai pu ; et je dis merci aux personnes qui m'ont aidé dans cette aventure, je leur dis merci, je chante ces personnes, je danse ces personnes… Je ne suis que reconnaissance !
+
+% Promoteurs
+Bien qu'un seul nom figure en couverture de cet ouvrage, les pages qui suivent seraient tristement restées vierges sans tout un environnement humain propice à la réflexion scientifique et à mon épanouissement personnel.
+Merci avant tout à mes deux promoteurs de thèse, mes deux ``pères académiques'', Hervé Jeanmart et Francesco Contino.
+Merci de m'avoir fait confiance dès le premier jour, donner la liberté de mener à bien mes recherches en toute autonomie tout en veillant à me prêter vos oreilles attentives et me prodiguer vos conseils avisés lors de nos échanges.
+Merci pour le temps que vous avez consacré lors de nos réunions pour résoudre les casse-têtes de ma thèse ainsi que ceux posés sur la table de Francesco.
+Merci à Hervé pour son pragmatisme et sa spontanéité et à Francesco pour son empathie et son goût du détail pour toujours chercher le dernier carat, à la limite parfois de la diptérophilie.
+
+% Membres de jury
+Merci aux membres de mon jury et encore plus aux membres de mon comité d'accompagnement, Sylvain Quoilin et Stefano Moret, pour le temps qu'ils ont consacré à la lecture de mon manuscrit et surtout pour les remarques toujours bienveillantes et constructives adressées lors de nos réunions et de la défense privée.
+Celles-ci m'ont permis d'apporter des améliorations majeures au texte et de mettre en exergue de futures pistes de recherche.
+
+% UCLouvain + BEST + Véro
+Pour avoir rendu lucrative cette aventure humaine et intellectuelle, je tiens à remercier l'UCLouvain mais aussi le SPF Economie au travers du Fonds de transition énergétique et du projet BEST.
+J'en profite pour dire un énorme merci à la coordinatrice de ce projet, Véronique Dias.
+Outre mes deux pères académiques, tu auras été une vraie mère, toujours aux petits soins pour moi et les autres chercheurs et chercheuses.
+Merci pour ton temps et ton investissement afin de faire de ce projet une franche réussite.
+
+% Collègues en commençant par Gauthier et Diederik
+Merci ensuite à mes parrains, Gauthier et Diederik.
+Vous m'avez catapulté dans cette aventure et m'avez toujours accompagné pour explorer de nouveaux horizons.
+Merci à Marion et Ignace pour votre écolage dans l'univers du Reinforcement Learning.
+Enfin, outre les tontons (parfois) un peu (trop) grincheux, je ne saurais qu'assez remercier tous mes frères et soeurs/cousins et cousines de cette belle et grande famille du TFL.
+Dans le désordre, merci Alice pour ton humour et tes goûts musicaux ``discutables'' et avoir tenté tant bien que mal de m'initier à la science de la musique et à la culture flamande.
+Merci Arnaud pour ton amitié de longue date, ta disponibilité, tes relectures et tes opinions toujours magistralement construites.
+Merci coach PB de m'en avoir fait baver et de m'avoir poussé à dépasser mes limites lors de nos nombreuses courses.
+Merci Youyou d'y avoir supporté nos discussions de darons (et de me botter désormais les fesses).
+Merci Mamasita Aurel pour ta good vibe et ton eau pet', Martinos pour la co-création du label, Paolito pour ta force tranquille, Antoine aka Jabiru pour ton abonnement au MR magazine, p'tit Louis pour avoir toujours si bien veillé à ma roue arrière, Zakito pour ton grand coeur, Nicou pour ton flow et tes serrages.
+Merci Sara de ne pas me détester.
+Xie xie Yanan for the hot pot.
+
+% Amis et famille
+Je remercie aussi du fond du coeur ma famille, ma belle-famille et mes amis. Merci pour votre amour et votre indéfectible soutien qui m'ont permis d'avancer jusqu'ici et de devenir qui je suis.
+
+% Chloé et les enfants
+Last but not least, je remercie ma compagne, Chloé.
+Merci pour ton oreille attentive et bienveillante surtout dans mes moments d'errance quand j'envisageais une reconversion en tant que jardinier ou éleveur de chèvres en Suisse.
+Je te remercie aussi d'être celle qui m'accompagne au quotidien depuis toutes ces années, et, surtout, d'être la formidable maman de nos deux enfants, Victor et Oscar, et de me permettre ainsi d'exercer le plus beau métier du monde.
+
+% Autres
+Ci-dessus n'était que l'humble expression de ma profonde gratitude envers celleux qui m'ont accompagné tout au long de ce chemin tantôt long fleuve tranquille tantôt parcours semé d'embûches.
+J'ai certainement omis de mentionner nombreux d'entre vous.
+Sachez cependant que sans votre présence et votre soutien, ce beau projet professionnel mais, avant tout, cette fabuleuse aventure personnelle n'aurait pas été possible.

### 2024-08-22 options/symbols_and_cmd.tex
Including remarks from HJ (Rebuttals)

@@
 %% A
-\newabbreviation[group={Acronyms}]{API}{API}{Application Programming Interface}
+\newabbreviation{API}{API}{Application Programming Interface}
 
 %% B
-\newabbreviation[group={Acronyms}]{BECCS}{BECCS}{Bioenergy with Carbon Capture and Storage}
-\newabbreviation[group={Acronyms}]{BEMS}{BEMS}{Building Energy Management System}
-\newabbreviation[group={Acronyms}]{BEV}{BEV}{Battery Electric Vehicles}
-\newabbreviation[group={Acronyms}]{BTX}{BTX}{Benzene, Toluene and Xylene}
+\newabbreviation{BECCS}{BECCS}{Bioenergy with Carbon Capture and Storage}
+\newabbreviation{BEMS}{BEMS}{Building Energy Management System}
+\newabbreviation{BEV}{BEV}{Battery Electric Vehicles}
+\newabbreviation{BTX}{BTX}{Benzene, Toluene and Xylene}
 
 %% C
-\newabbreviation[group={Acronyms}]{CAPEX}{CAPEX}{Capital Expenditure}
-\newabbreviation[group={Acronyms}]{CCGT}{CCGT}{Combined Cycle Gas Turbine}
-\newabbreviation[group={Acronyms}]{CCS}{CCS}{Carbon Capture and Storage}
-\newabbreviation[group={Acronyms}]{CHP}{CHP}{Combined Heat and Power}
-\newabbreviation[group={Acronyms}]{CNG}{CNG}{Compressed Natural Gas}
+\newabbreviation{CAPEX}{CAPEX}{Capital Expenditure}
+\newabbreviation{CCGT}{CCGT}{Combined Cycle Gas Turbine}
+\newabbreviation{CCS}{CCS}{Carbon Capture and Storage}
+\newabbreviation{CHP}{CHP}{Combined Heat and Power}
+\newabbreviation{CNG}{CNG}{Compressed Natural Gas}
 
 %% D
-\newabbreviation[group={Acronyms}]{DC}{DC}{Direct Current}
-\newabbreviation[group={Acronyms}]{DHN}{DHN}{District Heating Network}
-\newabbreviation[group={Acronyms}]{DNN}{DNN}{Deep Neural Network}
-\newabbreviation[group={Acronyms}]{DRL}{DRL}{Deep Reinforcement Learning}
+\newabbreviation{DC}{DC}{Direct Current}
+\newabbreviation{DHN}{DHN}{District Heating Network}
+\newabbreviation{DNN}{DNN}{Deep Neural Network}
+\newabbreviation{DRL}{DRL}{Deep Reinforcement Learning}
 
 %% E
-\newabbreviation[group={Acronyms}]{ESOMs}{ESOMs}{Energy System Optimisation Models}
-\newabbreviation[group={Acronyms}]{ESTD}{ESTD}{EnergyScope Typical Days}
-\newabbreviation[group={Acronyms}]{EUD}{EUD}{End-Use Demand}
+\newabbreviation{ESOMs}{ESOMs}{Energy System Optimisation Models}
+\newabbreviation{ESTD}{ESTD}{EnergyScope Typical Days}
+\newabbreviation{EUD}{EUD}{End-Use Demand}
 
 %% F
-\newabbreviation[group={Acronyms}]{FC}{FC}{Fuel Cell}
-\newabbreviation[group={Acronyms}]{FEC}{FEC}{Final Energy Consumed}
+\newabbreviation{FC}{FC}{Fuel Cell}
+\newabbreviation{FEC}{FEC}{Final Energy Consumed}
 
 %% G
-\newabbreviation[group={Acronyms}]{GDP}{GDP}{Gross Domestic Product}
-\newabbreviation[group={Acronyms}]{GHG}{GHG}{Greenhouse Gases}
-\newabbreviation[group={Acronyms}]{GSA}{GSA}{Global Sensitivity Analysis}
-\newabbreviation[group={Acronyms}]{GWP}{GWP}{Global Warming Potential}
+\newabbreviation{GDP}{GDP}{Gross Domestic Product}
+\newabbreviation{GHG}{GHG}{Greenhouse Gases}
+\newabbreviation{GSA}{GSA}{Global Sensitivity Analysis}
+\newabbreviation{GWP}{GWP}{Global Warming Potential}
 
 %% H
-\newabbreviation[group={Acronyms}]{HP}{HP}{Heat Pump}
-\newabbreviation[group={Acronyms}]{HT}{HT}{High-Temperature}
-\newabbreviation[group={Acronyms}]{HVC}{HVC}{High-Value Chemicals}
+\newabbreviation{HP}{HP}{Heat Pump}
+\newabbreviation{HT}{HT}{High-Temperature}
+\newabbreviation{HVC}{HVC}{High-Value Chemicals}
 
 %% I
-\newabbreviation[group={Acronyms}]{IAMs}{IAMs}{Integrated Assessment Models}
-\newabbreviation[group={Acronyms}]{ICE}{ICE}{Internal Combustion Engine}
-\newabbreviation[group={Acronyms}]{IEA}{IEA}{International Energy Agency}
-\newabbreviation[group={Acronyms}]{IPCC}{IPCC}{Intergovernmental Panel on Climate Change}
-\newabbreviation[group={Acronyms}]{IQR}{IQR}{Interquatile Range}
+\newabbreviation{IAMs}{IAMs}{Integrated Assessment Models}
+\newabbreviation{ICE}{ICE}{Internal Combustion Engine}
+\newabbreviation{IEA}{IEA}{International Energy Agency}
+\newabbreviation{IPCC}{IPCC}{Intergovernmental Panel on Climate Change}
+\newabbreviation{IQR}{IQR}{Interquatile Range}
 
 %% L
-\newabbreviation[group={Acronyms}]{LCA}{LCA}{Life Cycle Assessment}
-\newabbreviation[group={Acronyms}]{LCOE}{LCOE}{Levelised Cost of Energy}
-\newabbreviation[group={Acronyms}]{LFO}{LFO}{Light Fuel Oil}
-\newabbreviation[group={Acronyms}]{LOO}{LOO}{Leave-One-Out}
-\newabbreviation[group={Acronyms}]{LP}{LP}{Linear Programming}
-\newabbreviation[group={Acronyms}]{LPG}{LPG}{Liquefied Petroleum Gas}
-\newabbreviation[group={Acronyms}]{LT}{LT}{Low-Temperature}
+\newabbreviation{LCA}{LCA}{Life Cycle Assessment}
+\newabbreviation{LCOE}{LCOE}{Levelised Cost of Energy}
+\newabbreviation{LFO}{LFO}{Light Fuel Oil}
+\newabbreviation{LOO}{LOO}{Leave-One-Out}
+\newabbreviation{LP}{LP}{Linear Programming}
+\newabbreviation{LPG}{LPG}{Liquefied Petroleum Gas}
+\newabbreviation{LT}{LT}{Low-Temperature}
 
 %% M
-\newabbreviation[group={Acronyms}]{MDP}{MDP}{Markov Decision Process}
-\newabbreviation[group={Acronyms}]{MMSA}{MMSA}{Methanol Market Services Asia}
-\newabbreviation[group={Acronyms}]{MTBE}{MTBE}{Methyl Tert-butyl Ether}
-\newabbreviation[group={Acronyms}]{MTO}{MTO}{Methanol-to-olefins}
+\newabbreviation{MDP}{MDP}{Markov Decision Process}
+\newabbreviation{MMSA}{MMSA}{Methanol Market Services Asia}
+\newabbreviation{MTBE}{MTBE}{Methyl Tert-butyl Ether}
+\newabbreviation{MTO}{MTO}{Methanol-to-olefins}
 
 %% N
-\newabbreviation[group={Acronyms}]{NED}{NED}{Non-energy Demand}
-\newabbreviation[group={Acronyms}]{NG}{NG}{Fossil Gas}
-\newabbreviation[group={Acronyms}]{NN}{NN}{Neural Network}
-\newabbreviation[group={Acronyms}]{NRE}{NRE}{Non-renewable Energy}
-\newabbreviation[group={Acronyms}]{NSC}{NSC}{Naphtha Steam Cracker}
+\newabbreviation{NED}{NED}{Non-energy Demand}
+\newabbreviation{NG}{NG}{Fossil Gas}
+\newabbreviation{NN}{NN}{Neural Network}
+\newabbreviation{NRE}{NRE}{Non-renewable Energy}
+\newabbreviation{NSC}{NSC}{Naphtha Steam Cracker}
 
 %% O
-\newabbreviation[group={Acronyms}]{OPEX}{OPEX}{Operational Expenditure}
+\newabbreviation{OPEX}{OPEX}{Operational Expenditure}
 
 %% P
-\newabbreviation[group={Acronyms}]{PC}{PC}{Principal Component}
-\newabbreviation[group={Acronyms}]{PCs}{PCs}{Principal Components}
-\newabbreviation[group={Acronyms}]{PCA}{PCA}{Principal Component Analysis}
-\newabbreviation[group={Acronyms}]{PCE}{PCE}{Polynomial Chaos Expansion}
-\newabbreviation[group={Acronyms}]{PV}{PV}{Photovoltaic}
+\newabbreviation{PC}{PC}{Principal Component}
+\newabbreviation{PCs}{PCs}{Principal Components}
+\newabbreviation{PCA}{PCA}{Principal Component Analysis}
+\newabbreviation{PCE}{PCE}{Polynomial Chaos Expansion}
+\newabbreviation{PV}{PV}{Photovoltaic}
 
 %% R
-\newabbreviation[group={Acronyms}]{RE}{RE}{Renewable Energy}
-\newabbreviation[group={Acronyms}]{RL}{RL}{Reinforcement Learning}
+\newabbreviation{RE}{RE}{Renewable Energy}
+\newabbreviation{RL}{RL}{Reinforcement Learning}
 
 %% S
-\newabbreviation[group={Acronyms}]{SAC}{SAC}{Soft Actor-Critic}
-\newabbreviation[group={Acronyms}]{SMR}{SMR}{Small Modular Reactor}
-\newabbreviation[group={Acronyms}]{SVD}{SVD}{Singular Value Decomposition}
+\newabbreviation{SAC}{SAC}{Soft Actor-Critic}
+\newabbreviation{SMR}{SMR}{Small Modular Reactor}
+\newabbreviation{SVD}{SVD}{Singular Value Decomposition}
 
 %% S
-\newabbreviation[group={Acronyms}]{TRL}{TRL}{Technology Readiness Level}
+\newabbreviation{TRL}{TRL}{Technology Readiness Level}
 
 %% U
-\newabbreviation[group={Acronyms}]{UQ}{UQ}{Uncertainty Quantification}
+\newabbreviation{UQ}{UQ}{Uncertainty Quantification}
 
 %% V
-\newabbreviation[group={Acronyms}]{VRES}{VRES}{Variable Renewable Energy Sources}
+\newabbreviation{VRES}{VRES}{Variable Renewable Energy Sources}
 
@@
 
-\renewcommand\glstreegroupheaderfmt[1]{\begingroup\textbf{#1}\vspace{-.2cm}\par\endgroup}
+%\renewcommand\glstreegroupheaderfmt[1]{\begingroup\textbf{#1}\vspace{-.2cm}\par\endgroup}
 \glsfindwidesttoplevelname
-\glsxtrsetgrouptitle{Acronyms}{Acronyms}
-\glsxtrsetgrouptitle{rs}{Roman Symbols}
-\glsxtrsetgrouptitle{gs}{Greek Symbols}
-\glsxtrsetgrouptitle{ss}{Sub- Superscripts}
-\glsxtrsetgrouptitle{o}{Operators}
+%\glsxtrsetgrouptitle{Acronyms}{Acronyms}
+%\glsxtrsetgrouptitle{rs}{Roman Symbols}
+%\glsxtrsetgrouptitle{gs}{Greek Symbols}
+%\glsxtrsetgrouptitle{ss}{Sub- Superscripts}
+%\glsxtrsetgrouptitle{o}{Operators}
 

### 2024-08-22 appendices/app_RL_TD_MO.tex
Finish with Grammarly

@@
-The \acrfull{RL}-based exploration of myopic transition pathways can require thousands of runs to converge.
+The \gls{RL}-based exploration of myopic transition pathways can require thousands of runs to converge.
 To reduce the computational burden and maximise this exploration, the learning phase of the agent could start on the monthly model before being refined on the one with an hourly time resolution, the so-called transfer learning \cite{mann2013directed}.
-Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
+Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local VRES and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie a couple of seconds versus 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).
 This section aims at comparing results from the learning process on the monthly model with the one on the hourly model.
@@
 The other main difference concerns the import of electricity from neighbouring countries.
-As detailed in Appendix \ref{app:mo_vs_td}, this comes from the averaging of the hourly times series of \gls{VRES} that can provide electricity in a more constant way.
+As detailed in Appendix \ref{app:mo_vs_td}, this comes from the averaging of the hourly times series of \gls{VRES} that can provide electricity more constantly.
 

### 2024-08-22 appendices/app_case_study.tex
Finish with Grammarly

@@
 The outcomes of a model could be limited when the case study is too restrictive.
-Indeed, one could argue that the comparison between myopic, monthly and perfect foresight are very similar as the energy system is strongly constrained in terms of \gls{GHG} emissions.
+Indeed, one could argue that the results of myopic, monthly and perfect foresight are very similar as the energy system is strongly constrained in terms of \gls{GHG} emissions.
 
@@
 
-Similar trends that the ones for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
+Similar trends to the ones for the defossilisation are observed: primary energy mix reduces, renewable energy integration rises and an electro-fuel is imported.
 However, some changes reflect the cheapest option that the system could utilise such as using as little electrofuels as possible.
-In this case, only e-ammonia is used for its end use demand.
+In this case, only e-ammonia is used for its end-use demand.
 
-Instead of analysing the energy system in details, the following paragraphs will investigate if the comparison findings are consistent on a different case study.\\
+Instead of analysing the energy system in detail, the following paragraphs will investigate if the comparison findings are consistent in a different case study.\\
 
@@
 
-First, considering the overall transition cost, the myopic approach keeps on making limited short-term savings, \ie down to -0.2\%, before ending up with a slightly more expensive transition by 2050, \ie +0.2\%.
-Similarly to the case with an imposed \gls{GHG}-emissions trajectory, myopic optimisation invests more, compared to the perfect foresight, at early stages into \gls{VRES} technologies to benefit from the significant salvage values of the related grid infrastructures.
+First, considering the overall transition cost, the myopic approach keeps making limited short-term savings, \ie down to -0.2\%, before ending up with a slightly more expensive transition by 2050, \ie +0.2\%.
+Similar to the case with an imposed \gls{GHG}-emissions trajectory, myopic optimisation invests more, compared to the perfect foresight, at early stages into \gls{VRES} technologies to benefit from the significant salvage values of the related grid infrastructures.
 In 2030, the salvage value of the infrastructures and electricity-generation technologies account for 80.2 and 56.2 b€, respectively, whereas the overall CAPEX are 404.8 b€ by then.
@@
 Then, in the case with a prescribed \gls{GHG}-emissions trajectory, the myopic approach invests more by the end of the transition to renew \gls{PV} installed more massively at early stages and that reached the end of their lifetime before the end of the transition.
-In the case without this emissions trajectory, there is less urgency/need for integrating renewables in the system.
+In the case without this emissions trajectory, there is less urgency/need for integrating renewables into the system.
 Consequently, in the latter case, there is not such an extra-investment to make to renew too old renewable assets.
-On top of this, the slower uprise of \gls{VRES} in the case without emissions trajectory leads to a smaller difference of electrification of the other sectors between the perfect foresight and myopic approaches.
+On top of this, the slower uprise of \gls{VRES} in the case without emissions trajectory leads to a smaller difference in the electrification of the other sectors between the perfect foresight and myopic approaches.
 
 Finally, even though the way to get there differs between the perfect foresight and the myopic approaches, the system designs by 2050 are very similar between these two in most of the sectors.
-The main observed difference is in the freight transport where diesel boats are preferred to gas boats.
-This can be looked as a result of the lock-in effect where choices made at early stages, due to the limited foresight, remain in place in the longer term.
+The main observed difference is in freight transport where diesel boats are preferred to gas boats.
+This can be looked at as a result of the lock-in effect where choices made at early stages, due to limited foresight, remain in place in the longer term.
 
@@
 However, averaging time series of end-use demands and renewable productions brings some discrepancies.
-First and foremost, \gls{VRES} loose their intrinsic power intermittency \cite{haydt2011relevance}.
+First and foremost, \gls{VRES} lose their intrinsic power intermittency \cite{haydt2011relevance}.
 This has a series of consequences.
-Overall, this overestimates the uptake and emergence of solar and wind power generation and, consequently, overestimate their share in the primary electricity mix, especially at intermediate stages of the transition \cite{haydt2011relevance,Poncelet2016} (see Figure \ref{fig:MO_Share_VRES_elec_mix}).
+Overall, this overestimates the uptake and emergence of solar and wind power generation and, consequently, overestimates their share in the primary electricity mix, especially at intermediate stages of the transition \cite{haydt2011relevance,Poncelet2016} (see Figure \ref{fig:MO_Share_VRES_elec_mix}).
 In line with these studies, the more there is \gls{VRES} in the primary electricity mix, the higher is the error\footnote{Like \citet{Poncelet2016}, this error is defined as:$\sum_{i} \frac{|\text{supply share}_i^{MO}-\text{supply share}_i^{PF}|}{2}$, where index i runs over all technologies and import of electricity and the supply shares are expressed as a percentage.} with lower time resolution.
-In 2025 and 2050, where the share of \gls{VRES} in the electricity primary mix is 46\% and 66\%, respectively, this error goes from 10\% to 18\%.
+In 2025 and 2050, when the share of \gls{VRES} in the electricity primary mix is 46\% and 66\%, respectively, this error goes from 10\% to 18\%.
 
@@
 This smaller production of electricity is done to the detriment of direct import of electricity, completely removed from the primary mix, as assumed to keep on having a higher global warming potential than conventional assets (\eg \gls{CCGT} running on gas and, later on, renewable gas and ammonia).
-Then, the generation mix in heating sectors or non-energy demands is less diversified, leading the sharper and later switch of resources and technologies.
+Then, the generation mix in heating sectors or non-energy demands is less diversified, leading to the sharper and later switch of resources and technologies.
 For instance, lignocellulosic biomass is ``cannibalised'' by industrial boilers to the detriment of biomass-to-methanol or \gls{HVC}.
@@
 Finally, in the mobility sectors, there is no difference between low and high time resolutions.
-This is due to the fact that these demands are assumed to stay the same over the different hours of the year (\ie freight mobility) or have a favorite technology to supply them (\eg \gls{BEV} for private mobility).
+This is because that these demands are assumed to stay the same over the different hours of the year (\ie freight mobility) or have a favourite technology to supply them (\eg \gls{BEV} for private mobility).
 

### 2024-08-22 appendices/app_results_UQ.tex
Finish with Grammarly

@@
 The first column shows these indicators for the \gls{GSA} applied on the hourly pathway model.
-For information, the second column gives the same indicators but for an uncertainty quantification carried out on the monthly pathway model that has some limitations \cite{limpens2024pathway} but has the main advantage to run much faster.
-Given the similar rankings of the parameters between these two, this comparison shows that the monthly model can be a computationally efficient proxy to quantify the uncertainties of the actual hourly model and point out the key parameters on optimisation-driving objective, the total transition cost.
+For information, the second column gives the same indicators but for an uncertainty quantification carried out on the monthly pathway model that has some limitations \cite{limpens2024pathway} but has the main advantage of running much faster.
+Given the similar rankings of the parameters between these two, this comparison shows that the monthly model can be a computationally efficient proxy to quantify the uncertainties of the actual hourly model and point out the key parameters of the optimisation-driving objective, the total transition cost.
 
@@
 However, this does not jeopardize the comparative analysis given the similar Sobol' indices.
-Given their wide range of uncertainty [-64.3\%; 179.8\%] and their significant role to meet the \ce{CO2} budget, the cost of purchasing electrofuels is the first, by far, impacting parameter.
+Given their wide range of uncertainty [-64.3\%; 179.8\%] and their significant role in meeting the \ce{CO2} budget, the cost of purchasing electrofuels is the first, by far, impacting parameter.
 Next, comes, naturally, the industrial \gls{EUD}, representing, at the nominal value, 60\% of the total demands by 2050.
-The top-3 is completed by the variation of the discount rate, directly impacting the annualisation and the salvage values of the assets.
-Finally, since the current Belgian whole-energy system deeply relies on fossil resources, and would still do in the near future, the cost of purchasing fossil fuels is part of the impacting parameters.
-On the contrary, due to the very low annualised, cost, long lifetime leading to a significant salvage value and a low-emitting fuel, the parameters related to \gls{SMR} barely impact the total transition cost.
+The top 3 are completed by the variation of the discount rate, directly impacting the annualisation and the salvage values of the assets.
+Finally, since the current Belgian whole-energy system deeply relies on fossil resources, and would still do so in the near future, the cost of purchasing fossil fuels is part of the impacting parameters.
+On the contrary, due to the very low annualised, cost, and long lifetime leading to a significant salvage value and a low-emitting fuel, the parameters related to \gls{SMR} barely impact the total transition cost.
 
@@
 Given its lower cost of purchasing than its renewable equivalent (\Cref{fig:cs_resources_cost}) and lower \gls{GWP} than other fossil fuels (\ie $\mathit{gwp}_{\mathrm{op,NG}}=0.27$\,kt$_{\ce{CO2},\text{eq}}$/GWh versus $\mathit{gwp}_{\mathrm{op,LFO}}=0.31$\,kt$_{\ce{CO2},\text{eq}}$/GWh or $\mathit{gwp}_{\mathrm{op,coal}}=0.40$\,kt$_{\ce{CO2},\text{eq}}$/GWh), fossil \gls{NG} remains the main source of gas in the system until 2040.
-Besides bio-hydrolysis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 in order to respect the \ce{CO2} budget for the transition.
+Besides bio-hydrolysis as the main consumer of wet biomass to consistently produce gas, e-methane eventually substitutes fossil natural gas by 2045-2050 to respect the \ce{CO2} budget for the transition.
 As a versatile energy carrier, gas is used by a wide variety of technologies in the different sectors.
 Initially, in 2020, decentralised gas boilers, \gls{CCGT} and industrial gas boilers represent the biggest consumers of gas with 39\%, 21\% and 16\% of the total consumption, respectively.
-Progressively, in line with the rest of the system shifting towards more efficiency in the mid-term, industrial \gls{CHP} represent the lion's share, next to other usages in the transport or \gls{LT}-heating sectors.
+Progressively, in line with the rest of the system shifting towards more efficiency in the mid-term, industrial \gls{CHP} represents the biggest share, next to other usages in the transport or \gls{LT}-heating sectors.
 
-On the contrary, import of fossil-based hydrogen, largely produced from steam-methane-reforming \cite{spf_economie_h2}, is rarely part of the solution due to the emissions related to the consumption of natural gas.
+On the contrary, the import of fossil-based hydrogen, largely produced from steam-methane-reforming \cite{spf_economie_h2}, is rarely part of the solution due to the emissions related to the consumption of natural gas.
 E-hydrogen is the consistent source of hydrogen in the system, next to local production (\ie steam-methane-reforming, electrolysis or ammonia-cracking) in some rare cases where low industrial \gls{EUD} coincides with more abundant electricity from \gls{SMR} or \gls{PV}.
-In terms of consumption, \gls{FC}-trucks are the more consistent player.
-\gls{FC}-cars are also at stake but in specific cases where their CAPEX and the CAPEX of electric vehicles are in the bottom and the top of their respective uncertainty ranges.
+In terms of consumption, \gls{FC} trucks are the more consistent player.
+\gls{FC} cars are also at stake but in specific cases where their CAPEX and the CAPEX of electric vehicles are at the bottom and the top of their respective uncertainty ranges.
 
-Becoming cheaper than its fossil equivalent at early stages of the transition (\ie from 2030 onward), e-ammonia is the exclusive stream of ammonia in the system, except rare cases.
+Becoming cheaper than its fossil equivalent at the early stages of the transition (\ie from 2030 onward), e-ammonia is the exclusive stream of ammonia in the system, except in rare cases.
 Then, on top of its consistent \gls{NED}, the largest consumption of ammonia is \gls{CCGT} as flexible power generation units, to substitute their e-methane equivalents that have a higher \gls{LCOE} (see \Cref{fig:LCOE}).
 
-Similarly to ammonia, on top of local production in rare cases, e-methanol is the key source for methanol.
-Besides its own \gls{NED}, methanol is mostly consumed to produce \gls{HVC} via the \gls{MTO} process instead of naphtha-cracking, in order to respect to \ce{CO2} budget for the transition.
+Similarly to ammonia, on top of local production in rare cases, e-methanol is the key source of methanol.
+Besides its own \gls{NED}, methanol is mostly consumed to produce \gls{HVC} via the \gls{MTO} process instead of naphtha-cracking, to respect to \ce{CO2} budget for the transition.
 

### 2024-08-22 rebuttals_short/rebuttals_short.tex
Creation of Rebuttals_Short

--- 2024-08-22 rebuttals/rebuttals.tex
@@
 \title{\vspace{-1cm}
-\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members}\end{flushleft}}
+\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members - SHORT}\end{flushleft}}
 \date{\vspace{-1.7cm}\begin{flushleft}\sffamily Exploration of uncertainty-aware energy transition pathways - Reinforcement learning and principal component analysis-based methods\end{flushleft}}
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{teal} \textbf{Stefan}} - Lot of the results hinge on the import of e-fuels with the assumption that they have GWP = 0; is it justified? Those fuels are directly available knowing they are no.
+{\color{teal} \textbf{Stefan}} - Lot of the results hinge on the import of e-fuels with the assumption that they have GWP = 0; is it justified? Those fuels are directly available knowing they are not.
 \end{mdframed}

### 2024-09-02 rebuttals/rebuttals.tex
Ready to print

@@
 \title{\vspace{-1cm}
-\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members}\end{flushleft}}
+\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members - FULL}\end{flushleft}}
 \date{\vspace{-1.7cm}\begin{flushleft}\sffamily Exploration of uncertainty-aware energy transition pathways - Reinforcement learning and principal component analysis-based methods\end{flushleft}}
@@
 The storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}).
+Even though there is no limitation on the yearly quantity of methane (fossil or e-methane), for each representative year, the model is constrained to import the same quantity for each hour of each typical day.
+For this reason, the model decides to install methane storage that is emptied during the winter and refilled during the summer, as methane is mostly used to produce heat.
 Analyses under uncertainties have not shown major variations either.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-There are various typos in the thesis, e.g.
+{\color{orange} \textbf{Stefano}} - There are various typos in the thesis, e.g.
 ``storagR'' (page 3), ``therefore, therefore'' (p.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.91, Table 4.2: how do you differentiate between actions that depend on the agent and exogenous uncertainties?
+{\color{orange} \textbf{Stefano}} \& {\color{violet} \textbf{Christophe}} - p.91, Table 4.2: how do you differentiate between actions that depend on the agent and exogenous uncertainties?
 More in general: how is uncertainty integrated in the RL framework?

### 2024-09-02 rebuttals_short/rebuttals_short.tex
Ready to print

@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - I found it unusual to see many references to own work (Rixhon et al.) throughout the thesis.
-Unless the articles are unrelated to the thesis, these references should be removed and the associated content should be included in the manuscript.
-As an example, at p.
-25, results are mentioned which are not reported in the thesis.
-The reader currently must refer to [48] to see these results, which should instead be included in the thesis.
-\end{mdframed}
-
-\noindent This choice to sometimes refer to own works was made to focus only on the main contributions of the thesis.
-However, I agree that some content should be included in the thesis to enhance its reading.
-I hereby list the references to works I have been involved in (with the reference number of the manuscript version submitted before the private defense) and detail the potential modifications brought to the manuscript for each of them.
-
-\begin{itemize}
-\item \emph{\cite{limpens2024pathway} Limpens et al., Energyscope pathway: an open-source model to optimise the energy transition pathways of a regional whole-energy system}: Mostly cited to give credits to Gauthier Limpens' work in developing the Pathway version of EnergyScope, I have decided to refer to the paper about the formulation choices rather than including them in the thesis manuscript.
-Consequently, no further modification regarding this reference.
-\item \emph{\cite{limpens2020impact} Limpens et al., The impact of uncertainties on the Belgian energy system: Application of the Polynomial Chaos Expansion to the EnergyScope model}: The table gathering the results of the comparison between Morris' and PCE methods has been included {\color{blue}at the end of Section 1.2.2}:
-\end{itemize}
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-In \cite{limpens2020impact}, we have assessed the PCE approach on the cost-optimum Belgian energy system in 2035, using EnergyScope TD.
-To do so, we compared the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$.
-Even if the output of each method does not have the same physical meaning, both methods can rank the parameters by their impact on the total annual cost of the energy system.
-Both rankings were very similar which validated the use of PCE in the rest of this work (see Table \ref{tab:GSA:Comparison}).
-\end{mdframed}
-
-\begin{table}[htbp]
-	\caption{Comparison of the Top-14 rankings for the improved Morris method (left) and total-order PCE method (right).
-We used the DTU's implementation of Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$).
-}
-    \label{tab:GSA:Comparison}
-    \centering
-    \begin{tabular}{lcc}
-    \hline
-		\textbf{Parameter} & \textbf{Morris Ranking} ($\mu^*_{i}$) & \textbf{PCE Ranking} ($S_i^T$) \\
-        \hline
-        Prices of hydrocarbons & 1 (0.7141) & 1 (0.6434) \\
-        Cost of increased efficiency & 2 (0.3046) & 2 (0.1014) \\
-        Cost of maintenance & 3 (0.2371) & 4 (0.0534) \\
-        Discount rate & 4 (0.2019) & 5 (0.0467) \\
-        Price of imported electricity & 5 (0.1974) & 3 (0.0809) \\
-        CAPEX of the grid & 6 (0.1498) & 6 (0.0252) \\
-        CAPEX of PV & 7 (0.1270) & 7 (0.0187) \\
-        Increase of electricity demand & 8 (0.1168) & 8 (0.0154) \\
-        CAPEX of nuclear power plant & 9 (0.0963) & 9 (0.0119) \\
-        Increase of space heating demand & 10 (0.0856) & 11 (0.0088) \\
-        Price of coal & 11 (0.0856) & 12 (0.0087)\\
-        Price of renewable fuels & 12 (0.0794) & 10 (0.0104)\\
-        Price of uranium & 13 (0.0607) & 13 (0.0047)\\
-        Onshore wind load factor & 14 (0.0600) & 14 (0.0046)\\
-       \hline
-    \end{tabular}
-\end{table}
-
-\begin{itemize}
-\item \emph{\cite{rixhon2021role} Rixhon et al., The role of electrofuels under uncertainties for the Belgian energy transition}: This paper is cited to highlight the GSA performed on the snapshot model with different GWP limits.
-The main contribution of this work in the context of the thesis was to provide a preliminary screening and selection of parameters to carry out the GSA on the pathway model.
-The result of this selection is given in Section 2.4 where the uncertain parameters and their respective ranges are listed.
-Consequently, no further modification regarding this reference.
-\item \emph{\cite{rixhon2022integration} Rixhon et al., Integration of non-energy among the end-use demands of bottom-up whole-energy system models}: Reference used to highlight the fact that whole-energy systems usually do not account for the non-energy demand.
-Consequently, no further modification regarding this reference.
-\item \emph{\cite{rixhon2021comprehensive} Rixhon et al., Comprehensive
-integration of the non-energy demand within a whole-energy system: towards a defossilisation of the chemical industry in Belgium}: Reference to graphs that have either been adapted from the paper or directly taken from it.
-Consequently, no further modification regarding this reference.
-\item \emph{\cite{rixhon2021terminology} Rixhon et al., Taxonomy of the fuels in a whole-energy system}: Where the content of this paper is the Appendix A of the thesis, some references to this paper aimed at clarifying the meaning of ``renewable fuels'' (Section 2.2) or the fact that uranium is considered neither fossil nor renewable (Sections 3.1.2 and 4.2.2).
-Consequently, no further modification regarding this reference.
-\end{itemize}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{violet} \textbf{Christophe}} - State of the art mainly European. Why not reference to other regions of the world.
@@
 However, no modification has been brought to the manuscript regarding this topic.
-
-\subsection{Writing style}
-\label{writing_style}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc.
-I recommend reducing them to the bare minimum, or completely getting rid of them.
-\end{mdframed}
-
-\noindent The listed colloquial expressions have been transformed and, overall, I have reviewed the manuscript to reduce them.
-For instance, {\color{blue} in the contributions of Chapter 2}, I have changed ``lion's share'' into ``bigger share''.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-There are various typos in the thesis, e.g.
-``storagR'' (page 3), ``therefore, therefore'' (p.
-11), ``The'' (p.14), ``Moleclues'' (p.
-61), etc..
-Also, the use English language can be improved, e.g.
-``calls FOR a variety'' (p.
-9), ``cumulative emissions ARE'' (p.
-14), ``than to'' (p.
-34), etc.
-The document will benefit from a spelling/grammar check.
-\end{mdframed}
-
-\noindent The listed typos have been corrected and, overall, the manuscript has gone through Grammarly to improve the spelling and the grammar.
 
@@
 \label{ESPathway}
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.11 : The stated computational time seems low for a pathway model with hourly resolution.
-Were there approximations (e.g., disabling crossover) added and what was their impact?\end{mdframed}
-
-\noindent To decrease the computational time by one order of magnitude, besides opting for Gurobi to solve the LP, I disabled crossover.
-Besides this significant reduction of computational time, I observed no other impact regarding the values of the objective function or the design variables at the optimum.
-As disabling crossover is a very LP-specific aspect, I added this information {\color{blue}in a footnote in the first paragraph of Section 1.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-To keep this low computational time, crossover was disabled to solve the optimisation problem.
-This had no observed impact regarding the values of the objective function or the design variables at the optimum.
-\end{mdframed}
 
@@
 Since the implementation of the salvage value and the formulation choices are the work developed by \citet{limpens2024pathway} (partly integrated in Appendix B) and given the above explanations, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.17-19: is the myopic version modeled as a single LP or as a sequence of LPs?
-This should be better clarified.
-The diagram at p.
-19 focuses on the code, it would be better replaced by a code showcasing the logic of the approach.
-\end{mdframed}
-
-\noindent Even though this sequence of LPs was already mentioned {\color{blue}in the caption of Figure 1.3}, I have complemented the text {\color{blue} in the paragraph ``Myopic pathway implementation'' of Section 1.1.2},
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-[...] The myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window, until eventually reaching 2050.
-\end{mdframed}
-
-\noindent reminded it in the caption of Figure 1.2
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-The myopic approach (in pink) uses several instances of the pathway model, illustrated in Figure 1.1, as a sequence of different and linked LP problems. 
-\end{mdframed}
-
-\noindent and extended the caption of {\color{blue} Figure 1.4}.
-However, besides highlighting the key step 3.3 where there is an update of parameters value, I have decided to keep the figure as is since, on the one hand, the extension of the caption provides now a better understanding and, on the other hand, the code itself is exactly what is schematically represented on the diagram.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=10cm]{MY_process_code.pdf}
-\caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
-Myopic optimisation of the pathway consists of a sequence of smaller LP problems limited to their respective 10-year time window.
-If the optimisation is not deterministic, step 3.3 is the one where the values of some parameters are changed due to uncertainties (see Section 1.2) or the actions taken by the Reinforcement Learning-agent (see Chapter 4)}
-\label{fig:MY_process_code}
-\end{figure}
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Can you explain the choice of N=5 for the uncertainty ranges?
-\end{mdframed}
-
-\noindent As detailed by \citet{Moret2017PhDThesis}: ``the ranges resulting from the uncertainty characterisation, are defined for a specific point in time in the planning horizon, usually its last year.
-Therefore, they need to be scaled accordingly when used in the optimisation model.
-Without this scaling, it would mean assuming that the uncertainty for year $N$ stands for all the years in the planning horizon''.
-In his thesis, Moret considered $N=20$ as the duration of the planning horizon between 2015, the year of Moret's work, and the snapshot optimisation of the Swiss case study in 2035.
-In my case, the planning horizon is 5-year long as the duration of a phase between two representative years of the transition.
-Referring to \citet{Moret2017PhDThesis} for the interested reader, I have explained a bit more the reason for this scaling factor {\color{blue}in the second paragraph of Section 2.4}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-In this work, following the approach of \citet{Moret2017PhDThesis}, we have defined ranges of uncertainties for the model parameters.
-For the operation-type parameters (types II and III), a scaling factor needs to be considered to avoid assuming that the uncertainty for the last year of the planning horizon ($N$) is representative for all years in the planning horizon \citet{Moret2017PhDThesis}.
-In EnergyScope Pathway, as the model optimises the system every 5 years, $N=5$ has been selected to get the final ranges of uncertainties of type II and III.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 20 - ``inspired by Guevara et al.'' $\rightarrow$ how was their approach used in your work?
-\end{mdframed}
-
-\noindent \citet{guevara2022modeling} implemented a stochastic process to maintain realism in the evolution of uncertainty through the transition.
-In this sense, they avoid a parameter to go from the upper bound at year $y$ to the lower bound 5 years later.
-It allows zigzag from one time step to another.
-(see Figure \ref{fig:Guevara}).
-In my case, once the sample of uncertain parameters is drawn, there is no zigzag ($\alpha=0$ in \cite{guevara2022modeling}) but without especially be at the nominal value.
-Besides this, the inspiration does not go further.
-Given this explanation, I have not brought modification to the manuscript in this regard.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.5\textwidth]{Guevara.png}
-\label{fig:Guevara}
-\end{figure}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-21: footnote.
-The statement is qualitative, it should be quantitative.
-What is the minimum number of runs needed to validate your experiments?
-Did you look into that?
-How was the surrogate model validated?
-\end{mdframed}
-
-\noindent The footnote was intentionally qualitative as the quantitative information came later on (in Section 1.2.2).
-To improve this, I have moved the content of this footnote into the core of the text and after this quantitative information has been brought {\color{blue}in Section 1.2.2, after ``to achieve a LOO error below 1\% for the total transition cost.''}
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-As second order PCE is the minimum to ensure accuracy of the surrogate model, selecting and grouping the parameters as presented in Section 1.2.3 is compulsory to alleviate the computational burden.
-Otherwise, considering thousands of independent uncertain parameters would lead to millions of EnergyScope Pathway runs, if no more.
-Another alternative to reduce the computational cost of the GSA would be to consider the monthly (instead of hourly) time resolution for the runs of EnergyScope Pathway.
-However, despite its shorter computation time (\ie couple of seconds), the main drawback of the monthly model is its poor representation of the integration of VRES \cite{limpens2024pathway}.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{orange} \textbf{Stefano}} - p.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 24: Sobol indices. Can you explain the choice wrt other methods? Difference between factor prioritization and factor fixing?
-\end{mdframed}
-
-\noindent This comparison with a similar approach like factor fixing and factor prioritization (based on Morris method \cite{morris_factorial_1991}) is detailed {\color{blue}in the paragraph ``Comparison with a proven method'' of Section 1.2.2}.
-I have extended this paragraph by integrating the table of comparison between the Morris method and PCE.
-This also avoids the reference to our own work and the reference [48], page 25 (see previous comment).
-
-\noindent Besides being an in-house used method, the choice to use PCE is to be able to quickly extract statistical moments of the output of interest, as explained {\color{blue}in the first paragraph of Section 1.2.2}.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p.21 - How are the groups of uncertainties defined? Is there a table?
-\end{mdframed}
-
-\noindent Even though the uncertain parameters, their groups and their ranges of uncertainties might be the same for other case studies, they have been especially adapted for the present case study, \ie Belgium.
-Consequently, I decided to exhaustively list these uncertain parameters in the chapter dedicated to the case study (see Section 2.4).
-However, to not leave the reader handing there and wondering if there are other parameters and how we have decided to group them, I have added a reference to Section 2.4 {\color{blue}at the end of Section 1.2.1} focusing on the characterisation of the uncertainties:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Section 2.4 gathers the list of the uncertain parameters considered in this work.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - Table 1.2 - what are the other parameters?
-\end{mdframed}
-
-\noindent See previous comment.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{purple} \textbf{Sylvain}} - p.
@@
 
-\subsection{Reinforcement Learning}
-\label{methodo_RL}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 31: Can you explain the need of using NN here? Couldn’t you just run the energyscope model?
-\end{mdframed}
-
-\noindent The need of using NN is in the learning phase of the RL approach, not in running EnergyScope.
-The justification to use NN is already detailed in Section 1.3.2: ``Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of RL with Deep Neural Network (DNN), called Deep Reinforcement Learning (DRL).
-Among others, one of the main drawbacks of traditional RL algorithms, \ie without the use of Neural Network (NN),  is that it suffers from the ``curse of dimensionality'' when facing problems with continuous action and state spaces (see Chapter 4).
-By approximating the state-action function with its parameters (\ie weights and biases), DNN can address this difficulty.
-''
-
-Given what is already written in the manuscript, I have not brought modification in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 27 When presenting the advantages of RL, you mention model-free approach. Ok, but you have a representation of the real world available to train the model!
-\end{mdframed}
-
-\noindent Indeed, we know the input parameters, the constraints and the objective function implemented in EnergyScope to optimise the transition pathway of the whole-energy system.
-In the RL framework, a model is a representation of how the environment will respond to the agent's actions.
-Unlike other cases such as grid world \cite{grid_world}, the dynamics of the environment, \ie the transition from one state to another given the uncertainties and the actions taken by the agent or the consequent reward function, is not known.
-Going for a model-free approach, we avoid building a model within the agent to inform it about this (complex) dynamics.
-
-This aspect is already explained in Section 1.3.2: ``Given the assumed absence of knowledge of the agent about the dynamics of the environment, i.e., its transition or reward functions, we needed a so-called ``model-free'' learning algorithm.'' One sentence of clarification has been added {\color{blue}in Section 1.3.2}:
-
-\begin{mdframed}[style=manuscript] % Comment from the reviewer
-In Reinforcement Learning, the ``model'' stands for the dynamics ``action-state-reward'' between the environment and the agent. 
-\end{mdframed}
-
 \subsection{Principal Component Analysis}
 \label{methodo_PCA}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 32: Why is PCA needed? Wouldn’t it be easier to just aggregate similar outputs of interest?
-\end{mdframed}
-
-\noindent Aggregate similar outputs of interest, without prior knowledge about these ``similarities'' would be challenging and require assumptions to carry out this aggregation, especially given the size of the problem (113 technologies over 7 representative years).
-The objective of PCA is to alleviate this effort and automatise this aggregation in order to capture the main variance of the design through the whole transition in a reduced number of principal components.
-It is reminded in Section 5.1.1: ``Even though the following observations could be made by analysing the distribution of installed capacities (see Appendix D.3) or the covariance matrices, the PCA decomposition offers a more visual and summarising representation of the main trends of variation.''
-
-Given what is already in the text and these explanations, I have not brought modification to the manuscript in this regard.
 
@@
 
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 45 How is the range accounted for the characterisation of private mobility technologies?
-\end{mdframed}
-
-\noindent Indeed, the increase of the range is a consequence of the increased efficiency and battery capacity.
-Consequently, I have removed the mention to the range {\color{blue}at the end of the last paragraph in the contributions of Chapter 2}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Regarding BEV, while the CAPEX has been kept unchanged, the efficiency and the battery capacity have been increased. 
-\end{mdframed}
-
-\subsection{Resources}
-\label{methodo_resources}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 51 Last paragraph of p. 51 is not clear. Are we talking about the purchase cost or the CO2 content?
-\end{mdframed}
-
-\noindent There was indeed a confusion in the information to deliver in this paragraph.
-For the sake of clarity, I have restructured {\color{blue}Section 2.2}.
-After the introductory paragraph, the graphs and the following paragraph focus on the cost of purchasing.
-The two last paragraphs address the aspects of availabilities and global warming potentials, respectively.
-
 \subsection{Conversion technologies}
 \label{methodo_technologies}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - Figure 2.5: What is the purpose of the violet line between ``biomass gasification'' and ``methanolation''?
-\end{mdframed}
-
-\noindent These were actually two different arrows, one pointing from ``biomass gasification for CH3OH'' to ``Methanol'' and the other from ``Methanolation'' to ``Methanol''.
-To avoid this confusion, the figure has been updated and the origin of these arrows are now on different vertical levels ({\color{blue} see Section 2.3}).
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=\textwidth]{NED_tech.pdf}
-\label{fig:NED_tech}
-\end{figure}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 54 Is there a differentiation between ground mounted and rooftop PV?
-\end{mdframed}
-
-\noindent In the continuity of Gauthier Limpens' works \cite{limpens2021generating}, we consider here only rooftop PV.
-This potential of 59.2\,GW corresponds to 250 km$^2$ of available roof well oriented that exist today.
-Accounting for ground mounted PV would almost double this potential, as shown in Paolo Thiran's thesis.
-In future works, it would be interesting to also implement this additional potential, especially for the sake of harmonisation of input data.
-On the contrary, Elia, the Belgian TSO, foresees a much more limited increase of installed PV, \ie 18\,GW by 2034 \cite{Elia_2024_2034}.
-Consequently, for this thesis, I have decided to stick to this intermediate potential of 59.2\,GW.
-For the sake of clarity, I only added ``rooftop'' {\color{blue}in the second paragraph of Section 2.2}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-On the other hand, solar, wind, hydro and uranium are limited by the technical potentials of rooftop PV panels (59.2\,GW) [...]
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 51: Offshore wind max capacity is too low and PV is lower than Bregilab.
-\end{mdframed}
-
-\noindent Concerning the potential of PV, see previous comment.
-Regarding offshore wind, we have assumed a conservative potential of 6\,GW as the lower bound of the forecast done by the Belgian offshore platform\footnote{https://www.belgianoffshoreplatform.be/fr/}.
-Similarly to EnergyVille \cite{PATHS2050}, we could have gone up to the upper bound of 8\,GW.
-Going much higher, such as 24\,GW for the ``Electrification'' scenario of EnergyVille, would account for access to a part of the 150\,GW offshore wind park project in the North Sea signed between Belgium, Germany, Denmark and the Netherlands (the ``Esbjerg Offshore Wind Declaration'').
-
-\noindent Since these potentials have been discussed and detailed in previous works and are not the main topic of my thesis, I have not brought any further modification to the manuscript in this regard.
 
@@
 \noindent See previous comments.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 53 What about fuel cost?
-\end{mdframed}
-
-\noindent Even though the cost of purchasing uranium was already presented in Figure 2.3 of the thesis, I have added this value {\color{blue}in Table 2.1}:
-
-\begin{table}[htbp!]
-\caption{Nominal features of the SMRs in EnergyScope.
-SMR exhibits the advantage to have a fully flexible production (\ie between 0 to the full capacity) unlike conventional nuclear that is constrained to produce a constant baseload at every hour of the year.}
-\label{tab:SMR_features}
-\begin{minipage}{\linewidth}
-\centering
-\begin{tabular}{l c c}
-\hline
-\textbf{Feature} & \textbf{Value} & \textbf{Unit}\\
-\hline
-CAPEX & 4850 & €/kW \\
-Annual OPEX & 103 & €/kW/year \\
-Lifetime & 60 & year \\
-\textbf{Cost of purchasing uranium} & \textbf{4} & \textbf{€/MWh}\\
-Efficiency & 40\% & -\\
-Maximum capacity & 6 & GW \\
-Annual availability & 85\% & -\\
-Operational year & 2040 & - \\
-Flexibility & Full & - \\
-\hline
-
-\end{tabular}
-\end{minipage}
-\end{table}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 55 - Figure 2.5 - Coupled with heat for endogenous and exogenous reactions?
-\end{mdframed}
-
-\noindent Indeed, the potential consumption/production of heat and/or electricity for each conversion process is considered in the model.
-For instance, the Methanol-To-Olefins (MTO) process consumes 1.24\,GWh of methanol and 0.33\,GWh of high-temperature heat to produce 1\,GWh of HVC, which makes a 64\% efficiency.
-To clarify this point, {\color{blue}the caption of Figure 2.5} has been extended:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Schematic view of the different resources able to produce HVC, ammonia and methanol with their related conversion technologies (including energy efficiency and their CAPEX - in €/kW of final molecules).
-\textbf{The efficiencies account for the potential consumption/production of heat and/or electricity within each conversion process.} Values stand for 2035.
-Graph from \cite{rixhon2021comprehensive}.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 56 Are modal shares exogenously set?
-\end{mdframed}
-
-\noindent In line with \citet{limpens2019energyscope}, modal shares are indeed exogenously set.
-This aims at keeping some realism and ensures the need of less cost-competitive options like private versus public transport or decentralised versus DHN heating.
-To clarify this point, it is now explicitly written {\color{blue}in Section 2.4}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-[...] (ix) other parameters like the discount rate or the \textbf{exogenous} modal share change in different key sectors.
-\end{mdframed}
 
@@
 
-\subsection{SMR}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-70: ``in other words, the variation of the parameter $f_{\mathrm{max,SMR}}$…''.
-I am not sure that this sentence is equivalent to the previous one.
-Please clarify.
-\end{mdframed}
-
-\noindent The Sobol' index is the ratio between the variance of the total transition cost due to the variation of one parameter and the total variance of the total transition cost.
-If $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the impact on the system, and, consequently, on its total transition cost, is the same: no SMR installed.
-A wide range of variation of the parameter leading to zero variation of the output of interest is one of the reasons to explain why the Sobol' index of this parameter is low.
-To clarify this point, I have rephrased the sentence {\color{blue}in Section 3.2.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Given the uncertainty characterisation presented in Section 2.4, there are 60\% chance that no SMR could be installed.
-\textbf{When $f_{\mathrm{max,SMR}}$ takes a value between 0 and 0.59, the variation of the total transition cost is only due to the variation of the other parameters.
-}
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 71, Table 3.1: isn’t the low impact simply due to the low assumed potential? Is this finding meaningful?
-
-\noindent {\color{purple} \textbf{Sylvain}} - p. 70 - But that's because the CAPEX is way too low. With more realistic values, the Sobol would be much higher.
-\end{mdframed}
-
-\noindent These assumed low potential and CAPEX are indeed some reasons why the SMR-related uncertain parameters have a limited impact on the variation of the total transition cost as already stated in the first paragraph of Section 3.2.1: ``However, given its limited deployment, up to 6\,GW, SMR-related parameters have a lower impact on the variation of the total transition cost.''.
-However, this is not the only reasons as detailed earlier in the same paragraph: the late potential availability (2040 at the soonest) as well as the ``cheap and low-emitting fuel, the long lifetime leading to lower annualised CAPEX and higher salvage value.''.
-
-Given these explanations and what is already written in the manuscript, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 67 - It really depends on what is considered primary energy in the case of nuclear
-\end{mdframed}
-
-\noindent As specified in the table about SMR features (see previous comment), the considered fuel for SMR is uranium, as it is the case for conventional nuclear power plants.
-Given this, I have not brought modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 78 - Why this threshold? Shouldn't it follow the potential capacity?
-\end{mdframed}
-
-\noindent The threshold locates at 60\% of the uncertain parameter $f_{\mathrm{max,SMR}}$ which represents the potential capacity of SMR.
-As presented in Section 2.4, this threshold corresponds to the possibility to install up to 6\,GW of SMR in 2050.
-As already written in Section 3.2.2: ``For the installed capacity of SMR, it is the availability of the technology that drives its installation (Figure 3.12).
-Not shown here but all the samples of the GSA highlight that SMR is installed to its maximum capacity, \ie 6GW, as soon as possible.'' To smooth the trend curves of the different uncertain parameters, I have applied a rolling average.
-For this reason, the 60\%-threshold does not appear on the Figure as sharp as it actually is.
-
-Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
-
-\subsection{PV and wind}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 64 - 15 GW of PV in 2025. Realistic?
-\end{mdframed}
-
-\noindent In the continuity of Gauthier Limpens' thesis  and in line with the rest of the research field, the objective of this thesis is to ``illustrate the priorities to be taken in each of the sectors and the efforts to make in one sector to facilitate the integration of certain technologies into other sectors'' \cite{limpens2021generating}.
-This work highlights the ``must-do'' or ``no-regret investments'' to respect the (very) ambitious \ce{CO2} budget imposed to the model.
-In this sense, besides the maximum potential of 59\,GW solar-PV, we have not made assumption on the rate of deployment.
-15\,GW of PV in 2025 can be considered as very optimistic and Elia forecasts to have these 15\,GW by 2030 \cite{Elia_2024_2034}.
-I agree that not limiting the rate of deployment of PV (and wind turbines) is a strong assumption.
-For this reason, in the perspectives of this work (see Conclusion), I have already pointed this out as ``a future work would be to constrain the deployment of local renewables.
-
-Given these explanations, I have not brought further modification to the manuscript in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 69 - No curtailment at all? It would be nice to have the exact number.
-\end{mdframed}
-
-\noindent There is indeed some curtailment, which remains very limited compared to the total production of solar-PV: 0.3\,TWh versus 61\,TWh.
-This information has been added {\color{blue}in the paragraph dedicated to ``High-temperature heat'' of Section 3.1.3}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-In the REF case, industrial electric heaters are mainly used to absorb the ``over-production'' of the 59\,GW solar-PV when fully deployed, in the sunny days, and limit the yearly curtailment to a maximum of 0.3\,TWh, 0.5\% of the yearly 61\,TWh produced by solar-PV.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 69 - Very strange. Should be the same as for industrial boilers. No smart charging?
-\end{mdframed}
-
-\noindent The difference between BEV and industrial electric heaters is that BEV is already a preferred option for electrification, even in the REF case.
-BEV completely substitute diesel and gasoline-cars as soon as possible in the transition and there cannot be more cars as private mobility is limited to supply up to 50\% of the demand of passengers mobility.
-Unlike high-temperature heat, there is no modal swap due to the installation of SMR.
-
-Given these explanations, I have not brought further modification to the manuscript in this regard.
-
-\subsection{Others}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 63 What does efficiency change in the installed capacity?
-\end{mdframed}
-
-\noindent Indeed, the efficiency only has an impact on the amount of primary uranium needed to produce the electricity.
-Consequently, I have removed this part of the sentence {\color{blue}in the first paragraph of Section 3.1.1}.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-Overall, given the restriction on yearly availability and the slightly higher electrification [...]
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 72 - Battery trucks are gaining momentum. Is the hydrogen route really the way to go?
-\end{mdframed}
-
-\noindent In general, the model gives fuel-cell trucks as the favoured option for the future of road freight transport given more cost-competitive options for electrification like private mobility and low-temperature heat.
-However, as illustrated in the ROB roadmap (see Section 5.2.1), battery trucks seem to be the cost-optimal solution to transition between diesel and either hydrogen trucks, when fossil and electrofuels are assumed to be expensive.
-
-In the Global Sensitivity Analysis, electric trucks are the favoured option for a limited set of samples, about 10\% (see Appendix D.3).
-
-Given these explanations and what is already present in the text, I have not brought further modification to the manuscript in this regard.
-
 \section{Chapter 4 - Reinforcement Learning}
@@
 \subsection{General}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p.
-85 - Figure 4.1: Does it mean we can run ``infinite energy transitions''?
-Is this realistic?
-Isn’t the ``learning'' supposed to happen during the different stages?
-\end{mdframed}
-
-\noindent Even though Reinforcement Learning can be applied to problems with infinite time-horizon discounting reward coming further in the future, it is not the case in my thesis.
-Indeed, each episode (\ie the transition) is finite.
-The purpose of Figure 4.1 was to focus on the reward function and how it articulates in the action-state-reward process.
-As already represented in this figure, three situations can terminate an episode: (i) infeasible optimisation, (ii) overshot \ce{CO2} budget before 2050 or, (iii) reaching 2050.
-This is now clarified {\color{blue}in the caption of Figure 4.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Reward function, $R$.
-Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2} budget is exceeded.
-If the optimisation provides a solution and the \ce{CO2} budget is not exceeded, the episode continues.
-Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost, and the episode terminates.
-After terminating an episode, the process starts over at the initial state, \ie 2020.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - How can you account for pivotal events/shocks in the system?
-\end{mdframed}
-
-\noindent Given the limited foresight of the myopic approach, we can implement, pivotal events/ shocks through uncertain parameters.
-For instance, this is how we have implemented the potential to install SMR in 2040 at the soonest.
-Given the probability described in Section 2.4, the technology can (or not) be installed.
-However, as detailed in Section 1.2.1, for the other uncertain parameters, they keep the same value relatively to their type and range for the whole transition.
-In other words, we do not allow zig-zag where, for instance, a cost of purchasing an energy carrier was low for the first half of the transition and, suddenly, in 2040, becomes very high.
-
-Given these explanations and what is already written, I have not brought modification to the manuscript in this regard.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 87 - What is a step?
-\end{mdframed}
-
-\noindent In the context of Reinforcement Learning, a step consists in taking an action, collecting a reward based on the previous state and the taken action and ending in a new state.
-Even though it was already mentioned in Chapter 1, the fact that the results come further from the methodology part makes it harder for the reader to remember the meaning of a step.
-It is now more explicitly defined {\color{blue}in the first paragraph of Section 1.3.2}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-From this state, the agent takes a step: action, reward, new state.
-\end{mdframed}
-
-\noindent and reminded {\color{blue}in the first sentence of Section 4.2.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-The learning phase has been split into batches of 500 steps, \ie 500 sequences of state-action-reward-new state. 
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 96 - The early constraints on generation avoid lock-in situations correct?
-\end{mdframed}
-
-\noindent Correct. This confirms what \citet{vogt2018starting} detailed in their works. To emphasise on this point, I have added one sentence {\color{blue}in Section 4.2.3}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript 
-In line with \citet{vogt2018starting}, the early constraints on generation and the sharper decrease of the emissions avoid lock-in situations.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - p.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 100 - What/where are the no-go zones? Is there a narrative to describe them ?
-\end{mdframed}
-
-\noindent No-go zones are introduced in the second paragraph of Chapter 4: ``[...] as well as no-go zones where the chances of succeeding the transition, i.e., respecting the \ce{CO2} budget, are very limited.'' It is then detailed at the end of Section 4.2.2: ``Below the near-term threshold of $\sim$60\%, this is the ``no-go zone'' where succeeding the transition becomes unlikely, except
-if betting on the future installation of SMR. '' and in the caption of the related figure. Given the information already present in the manuscript, no modification has been brought in this regard.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{violet} \textbf{Christophe}} - The concept of binding constraint was not clear.
@@
 Consequently, regarding this comment, there has not been further modification brought to the manuscript.
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{purple} \textbf{Sylvain}} - p. 96 - Why 10 years if the step is 5?
-\end{mdframed}
-
-\noindent Because the action limits the emissions of the system at the end of the 10-year time window whereas limiting the consumption of fossil gas, LFO and coal is done over the whole time window, as detailed in Section 4.1.1.
-Given what is already written in the manuscript, I have not brought modification to the manuscript in this regard.
 
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} - p. 109 - how is Gamma-robustness implemented here? It seems like it is in some sense ``manually'' forced? Explain better how the ROB solution is derived.
-\end{mdframed}
-
-\noindent To my understanding, the $\Gamma$-robustness consists in considering a number, $\Gamma$, of uncertain parameters taking their worst-case value simultaneously and ``evaluate how the solution of the optimisation problem changes when varying $\Gamma$'' \cite{moret2020overcapacity}.
-Inspired by this approach, we have arbitrarily set this value to $\Gamma=6$ to consider the six first uncertain parameters impacting the most the variation of the total transition cost.
-Consequently, the ROB technological roadmap is obtained from the deterministic optimisation of the perfect foresight pathway accounting for these six parameters (\ie costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry EUD, the interest rate and the variable OPEX of technologies) at their respective worst-case value.
-As it does not exactly follow the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}, I have changed {\color{blue}the first sentence of Section 5.2.1}:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow a \textbf{method inspired by} the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}.
-\end{mdframed}
-
-Moreover, the facts that this $\Gamma=6$ is arbitrarily set and that the ROB roadmap, like the REF and SMR ones, results from a deterministic optimisation of the perfect foresight pathway are now clarified:
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-In the present analysis, we label the roadmap as ``robust'' (ROB) by arbitrarily setting $\Gamma_{\text{obj}}= 6$.
-The ROB roadmap results from the deterministic optimisation of the perfect foresight pathway where the following six parameters are set at their worst-case value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry EUD, the discount rate and the variable OPEX of technologies.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{violet} \textbf{Christophe}} - When using the word ``component'', there seems to be a confusion and it is not always easy to understand if you refer to the vector or the coefficient related to one of the original variable.
@@
 \begin{mdframed}[style=comment] % Comment from the reviewer
-{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - p.
-119 -  the statement mentioning the ``major added value of the perfect foresight'' approach seems surprising and needs to be clarified.
-The decreasing \ce{CO2} emissions trajectories shouldn't be a S-shaped (logistic) curve?
-\end{mdframed}
-
-\noindent I agree that this sentence and what came next focused on the perfect foresight approach whereas the rest of the paragraph rather targets the RL-based myopic results.
-Therefore, I have reformulated {\color{blue}the end of this paragraph} and focused on the shape of the decreasing \ce{CO2} trajectories.
-These are indeed S-shaped curves but that are sharper for the myopic approach than in the PF results.
-
-\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
-This work also highlighted a major difference between the perfect foresight (PF) approach and the agent-based myopic transitions.
-Given the assumed full knowledge on the future, the trajectories to decrease the \ce{CO2} emissions resulting from perfect foresight are more progressive, and foreseen to be more applicable, compared to a sharper drop at the early stages of the transition, like in the myopic approach.
-\end{mdframed}
-
-\begin{mdframed}[style=comment] % Comment from the reviewer
 {\color{teal} \textbf{Stefan}} - The conclusion is a summary and should rather be a discussion.

### 2024-09-02 appendices/app_EnergyScope.tex
Ready to print

@@
 
+\newpage
 From Table \ref{tab:intro_soa}, four models almost check all the boxes (partially the pathway one): Calliope, GENeSYS-MOD, PyPSA and TIMES. \\
@@
 \noindent
-Similarly GENeSYS-MOD presents some limitations.
+Similarly, GENeSYS-MOD presents some limitations.
 This model is an application of the open-source energy modelling system (OSeMOSYS), itself represented as a model with a poor time discretisation and a heavy computational burden according to \cite{prina2019transition}.
@@
 
+\newpage
 \subsection{The starting point: a scenario analysis model}

### 2024-09-02 chap_RL/chap_RL.tex
Ready to print

@@
 
+\newpage
 The reward function is defined in three steps.
@@
 Consequently, the share of successful transitions compared to failures progressively increases with time.
-In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
-Reaching cumulative emissions significantly lower than the \ce{CO2} budget is possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
-Considering the failures in 2050, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
-As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
-Where 98\% of the failures were below the \ce{CO2} budget in 2035, only 37\% passed this threshold in 2040.
-This reminds the importance of near-term, 2025-2030, actions to hope for a successful transition.
 
@@
 
-\newpage
+In the successful transitions, the median cumulative emissions, $\text{P}_{50}$, are about 0.9\,Gt$_{\ce{CO2},\text{eq}}$.
+Reaching cumulative emissions significantly lower than the \ce{CO2} budget is possible thanks to efforts made at earlier stages of the transition and the potential to install \gls{SMR} later on.
+Considering the failures in 2050, half of these episodes ended up with cumulative emissions lower or equal to 1.4\,Gt$_{\ce{CO2},\text{eq}}$.
+As illustrated in Figure \ref{fig:reward_status}, 2040 is identified as the tipping year.
+Where 98\% of the failures were below the \ce{CO2} budget in 2035, only 37\% passed this threshold in 2040.
+This reminds the importance of near-term, 2025-2030, actions to hope for a successful transition.
+
 Given the reward function (see Figure \ref{fig:Reward}), the agent optimises its policy by aiming at lowering the total transition cost as soon as it meets the \ce{CO2} budget.
@@
 
-\newpage
 Finally, we observe that the majority of the successful transitions are cheaper than the reference transition cost, 1.1\,b€.
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!t]
 \centering
@@
 
+\newpage
 After filtering out failures of the learning episodes and keeping only the successful transitions, only a limited set of the actions are binding and have an actual impact on the result of the optimisation in EnergyScope Pathway (see Figure \ref{fig:Binding_learning}).

### 2024-09-02 chap_atom_mol/chap_atom_mol.tex
Ready to print

@@
 However, in their ``Central'' scenario where no \gls{SMR} is installed by 2050, they rely on 96.6\,GW of \gls{PV}, 63\% more than the 59.2\,GW potential considered in our work and about 15 times more than the current installed capacity, \ie 6.5\,GW.
-\\
 
-\begin{figure}[htbp!]
+\begin{figure}[t!]
 \centering
@@
 
-\newpage
 When assessing the electricity production-versus-consumption-balance (\Cref{fig:results_deter_layer_elec}), \gls{SMR}, as a cheaper, flexible and low-emitting power generation system, produces to its full capacity.
@@
 
-\begin{figure}[htbp!]
+\begin{figure}[t!]
 \centering
@@
 
+\newpage
 Then, considering the primary energy mix shown in \Cref{fig:results_deter_energy_mix}, three phases in the transition can be identified.
@@
 
+As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia (Figure \ref{fig:results_uq_samples_ammonia}).
+As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheaper electricity produced by \gls{SMR} (40 versus 151 €/MWh$_{\text{elec}}$) substitutes these \gls{CCGT}.
+With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 2.0\,TWh, 95.4\% less than in the REF case.
+Then, with a 12\%-Sobol' index, the cost of purchasing electricity from abroad, considered as renewable by 2050 and, therefore, a direct competitor to e-ammonia \gls{CCGT}, also affects the need of this molecule, especially when this cost is low.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
+\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-ammonia in 2050.
+Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
+The grey dashed line gives the value of the output of interest in the REF case.
+}
+\label{fig:results_uq_samples_ammonia}
+\end{figure}
+
+\newpage
 As aforementioned, the industrial \gls{EUD} impacts the most the import of e-methane (Figure \ref{fig:results_uq_samples_gas}).
@@
 \label{fig:results_uq_samples_gas}
-\end{figure}
-
-As already pointed out in Section \ref{subsec:atom_mol:results_deter_power_sector}, the installation of \gls{SMR} drastically reduces the import of e-ammonia (Figure \ref{fig:results_uq_samples_ammonia}).
-As ammonia \gls{CCGT} is the biggest consumer of ammonia by the end of the transition, low-emitting and cheaper electricity produced by \gls{SMR} (40 versus 151 €/MWh$_{\text{elec}}$) substitutes these \gls{CCGT}.
-With a higher cost of purchasing electrofuels, this import of e-ammonia drops down to 2.0\,TWh, 95.4\% less than in the REF case.
-Then, with a 12\%-Sobol' index, the cost of purchasing electricity from abroad, considered as renewable by 2050 and, therefore, a direct competitor to e-ammonia \gls{CCGT}, also affects the need of this molecule, especially when this cost is low.
-
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=0.8\textwidth]{UQ_Ammonia_samples_2.pdf}
-\caption{Trend lines of the key parameters (and their Sobol' index) on the import of e-ammonia in 2050.
-Around these lines, box plots point out the distribution of the output of interest for the extreme values (either bottom-15\% or top-15\%) of some parameters.
-The grey dashed line gives the value of the output of interest in the REF case.
-}
-\label{fig:results_uq_samples_ammonia}
 \end{figure}

### 2024-09-02 chap_case_study/chap_case_study.tex
Ready to print

@@
 
-%\noindent
+\newpage
 Works of \citet{rixhon2021comprehensive,rixhon2022integration} assessed the impact of the integration of the \gls{NED} in the case of Belgium in 2050, using EnergyScope TD (see Appendix \ref{app:ESTD}).
@@
 \citet{rixhon2021role} analysed the impact of these parameters on the total cost of the snapshot Belgian whole-energy system in 2050 subject to different \gls{GWP} limits.
-Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full}.
-
-This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for the variability of solar irradiance or wind speed; (vii) the availability of resources considered limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the exogenous modal share change in different key sectors.
-A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.
-As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040.
-Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
-Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
-}.
-Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability.
-Finally, the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
-
-Table \ref{tab:UC_full} summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.
+Based on this work, we have selected a subset of impacting uncertainties, added others due to the pathway formulation (\eg $\Delta_{\mathrm{change,pass}}$), and listed them in Table \ref{tab:UC_full} that summarises the uncertainty ranges for the different groups of technologies and resources, for the year 2025.
 
@@
 \end{table}
+
+This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for the variability of solar irradiance or wind speed; (vii) the availability of resources considered limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the discount rate or the exogenous modal share change in different key sectors.
+A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}.
+As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040.
+Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology.
+Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward.
+}.
+Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability.
+Finally, the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.
 

### 2024-09-02 chap_methodo/chap_methodo.tex
Ready to print

@@
 \centering
-\includegraphics[width=10cm]{MY_process_code.pdf}
+\includegraphics[width=\textwidth]{MY_process_code.pdf}
 \caption{Diagram of the iterative optimisation of the whole-energy system transition pathway.
@@
 
-\newpage
 Finally, the model accounts for thousands of parameters.
@@
 
-\newpage
 Before jumping to the choice of the learning algorithm, it is worth noting that we opted for the combination of \gls{RL} with \gls{DNN}, called \gls{DRL}.
@@
 
+\newpage
 \gls{SAC} is a model-free and off-policy actor-critic deep \gls{RL} algorithm based on the entropy-augmented objective function (see Equation (\ref{eq:SAC_objective})).
@@
 
+\newpage
 From these observations, the main objective of \gls{PCA} is to find the linear combination, \ie projection, of the original variables that maximise their spread, \ie variance (see Figure \ref{fig:PCA_intro_proj}).
@@
 
+\newpage
 \myparagraph{Outliers management}\\
@@
 
-\newpage
 \myparagraph{Robustness assessment of roadmaps}\\

### 2024-09-02 frontend/merci.tex
Ready to print

@@
 % Promoteurs
-Bien qu'un seul nom figure en couverture de cet ouvrage, les pages qui suivent seraient tristement restées vierges sans tout un environnement humain propice à la réflexion scientifique et à mon épanouissement personnel.
+Bien qu'un seul nom figure en couverture de cet ouvrage, les pages qui suivent seraient tristement restées vierges de lettres sans tout un environnement humain propice à la réflexion scientifique et à mon épanouissement personnel.
 Merci avant tout à mes deux promoteurs de thèse, mes deux ``pères académiques'', Hervé Jeanmart et Francesco Contino.
-Merci de m'avoir fait confiance dès le premier jour, donner la liberté de mener à bien mes recherches en toute autonomie tout en veillant à me prêter vos oreilles attentives et me prodiguer vos conseils avisés lors de nos échanges.
-Merci pour le temps que vous avez consacré lors de nos réunions pour résoudre les casse-têtes de ma thèse ainsi que ceux posés sur la table de Francesco.
-Merci à Hervé pour son pragmatisme et sa spontanéité et à Francesco pour son empathie et son goût du détail pour toujours chercher le dernier carat, à la limite parfois de la diptérophilie.
+Merci de m'avoir fait confiance dès le premier jour, donné la liberté de mener à bien mes recherches en toute autonomie tout en veillant à me prêter vos oreilles attentives et me prodiguer vos conseils avisés lors de nos échanges.
+Merci pour le temps que vous avez consacré lors de nos réunions afin de résoudre les casse-têtes de ma thèse ainsi que ceux posés sur la table de Francesco.
+Merci à Hervé pour ton pragmatisme et ta spontanéité et à Francesco pour ton empathie et ton goût du détail pour toujours chercher le dernier carat, à la limite parfois de la diptérophilie.
 
@@
 Enfin, outre les tontons (parfois) un peu (trop) grincheux, je ne saurais qu'assez remercier tous mes frères et soeurs/cousins et cousines de cette belle et grande famille du TFL.
-Dans le désordre, merci Alice pour ton humour et tes goûts musicaux ``discutables'' et avoir tenté tant bien que mal de m'initier à la science de la musique et à la culture flamande.
+Dans le désordre, merci Alice pour ton humour et tes goûts musicaux ``discutables'' et avoir tenté tant bien que mal de m'initier à la science de la musique, à la culture flamande et au pastis.
 Merci Arnaud pour ton amitié de longue date, ta disponibilité, tes relectures et tes opinions toujours magistralement construites.
-Merci coach PB de m'en avoir fait baver et de m'avoir poussé à dépasser mes limites lors de nos nombreuses courses.
+Merci Coach PB de m'en avoir fait baver et de m'avoir poussé à dépasser mes limites lors de nos nombreuses courses.
 Merci Youyou d'y avoir supporté nos discussions de darons (et de me botter désormais les fesses).
@@
 % Amis et famille
-Je remercie aussi du fond du coeur ma famille, ma belle-famille et mes amis. Merci pour votre amour et votre indéfectible soutien qui m'ont permis d'avancer jusqu'ici et de devenir qui je suis.
+Je remercie aussi du fond du coeur ma famille, ma belle-famille et mes amis.
+Merci pour votre amour et votre indéfectible soutien qui m'ont permis d'avancer jusqu'ici et de devenir l'homme qui je suis aujourd'hui et que je deviendrai demain.
 
 % Chloé et les enfants
-Last but not least, je remercie ma compagne, Chloé.
+Last but not least, je remercie ma compagne, ma Tchoup, Chloé.
 Merci pour ton oreille attentive et bienveillante surtout dans mes moments d'errance quand j'envisageais une reconversion en tant que jardinier ou éleveur de chèvres en Suisse.
@@
 J'ai certainement omis de mentionner nombreux d'entre vous.
-Sachez cependant que sans votre présence et votre soutien, ce beau projet professionnel mais, avant tout, cette fabuleuse aventure personnelle n'aurait pas été possible.
+Sachez cependant que sans votre présence et votre soutien, ce beau projet professionnel mais, avant tout, cette fabuleuse aventure personnelle n'auraient été possibles.

### 2024-10-21 chap_case_study/chap_case_study.tex
Final push for THESIS

@@
 Based on the work of the \citet{national2013transitions}, \citet{limpens2021generating} excessively favoured \gls{FC} car versus \gls{BEV}.
-In previous works \cite{limpens2021generating,rixhon2021role}, despite their lower efficiencies (\ie about 50\% less efficient), \gls{FC} cars had the bigger share of private mobility compared to battery electric vehicles(BEV) given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter.
+In previous works \cite{limpens2021generating,rixhon2021role}, despite their lower efficiencies (\ie about 50\% less efficient), \gls{FC} cars had the bigger share of private mobility compared to battery electric vehicles (BEV) given the higher CAPEX (\ie up to 10\% more expensive) and limited range (\ie 24kWh battery) of the latter.
 In these works, the more limited potential to import electricity from abroad and to produce it locally via \gls{VRES} forced the model to electrify the low-temperature heat sector rather than the private mobility running on supposedly infinitely available renewable hydrogen.
